### 2020-11-13 examples/00-Data-Processing-Example.Rmd
v1 Initial commit

---
title: "00-Data-Processing-Example"
output: html_notebook
---

In this notebook we create some examples of processing the data for the analysis of accessibility. Particularly, we're interested in spatial interpolationo of the population for small areas.

## Preliminaries

Load packages used in the notebook:
```{r message = FALSE}
library(pycno)
library(readr)
library(rgrass7)
library(sf)
library(tidyverse)
```

Load data files:
```{r}
load("hamilton_cma.RData")
load("hamilton_da_2016.RData")
```

These two data files are simple features with the dissemination areas according to the 2016 Census, and the boundary of Hamilton CMA.
```{r}
ggplot() +
  geom_sf(data = hamilton_cma)
```

```{r}
ggplot() +
  geom_sf(data = hamilton_da_2016)
```

Read Sobi service area:
```{r}
sobi_service <- st_read("SoBi_Service_Areas.shp")
```

Read parks:
```{r}
parks <- st_read("Parks.shp")
```

Project the Hamilton DA zones to the same projection of the SoBi Service Areas:
```{r}
hamilton_da_2016 <- st_transform(hamilton_da_2016, crs = st_crs(sobi_service))
```

Plot the DAs and the service areas:
```{r}
ggplot() +
  geom_sf(data = hamilton_da_2016,
          color = "lightgray") +
  geom_sf(data = sobi_service,
          fill = NA)
```

We want to work only with the DAs that are within a reasonable buffer of the service areas (e.g., 500 m):
```{r}
ggplot(hamilton_da_2016[sobi_service %>% st_buffer(500),]) +
  geom_sf(color = "lightgray") + 
  geom_sf(data = sobi_service,
          fill = NA)
```

Extract DAs in service areas:
```{r}
hamilton_da_sobi <- hamilton_da_2016[sobi_service %>% st_buffer(500),]
```

Extract parks in service areas:
```{r}
parks_sobi <- parks[sobi_service %>% st_buffer(500),]
```

Read population table (2016 Census of Population, Population and Dwelling counts):
```{r}
population_da_2016 <- read_csv("T1901EN.csv") %>%
  transmute(DA = factor(`Geographic code`), population = `Population, 2016`)
```

Join population table to DAs in service areas:
```{r}
hamilton_da_sobi <- hamilton_da_sobi %>%
  left_join(population_da_2016, by = "DA")
```

Plot population:
```{r}
ggplot() +
  geom_sf(data = hamilton_da_sobi,
          aes(fill = population),
          color = NA) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = parks_sobi,
          fill = "green") +
  scale_fill_distiller(palette = "YlOrRd", direction = 1)
```

Remove the parks from the DAs (use `st_difference()`):
```{r}
# First combine `parks_sobi` to generate a single geography object, and make valid
parks_sobi <- st_combine(parks_sobi) %>% 
  st_make_valid()
# The new simple features object will be the difference of the two objects
hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi, parks_sobi)
```

Plot the new DAs (minus parks) and compare to original DAs:
```{r}
ggplot() +
  geom_sf(data = hamilton_da_sobi_clean,
          color = NA,
          fill = "blue") +
  geom_sf(data = hamilton_da_sobi,
          fill = NA,
          color = "black")
```

Pycnophylactic interpolation needs a `SpatialPolygonsDataFrame` object. Convert the simple features object:
```{r}
hamilton_da_sobi_clean.sp <- as(hamilton_da_sobi_clean, "Spatial")
```

Extract the population from the spatial object:
```{r}
pop <- hamilton_da_sobi_clean.sp@data$population
```

Pycnophylactic interpolation:
```{r}
filter <- stats::filter # Important! `filter()` should be from `stats`, not `dplyer`
interpolated_pop <- pycno(x = hamilton_da_sobi_clean.sp, pops = pop, celldim = 50)
```

Convert to pixels and then simple features:
```{r}
interpolated_pop.pix <- as(interpolated_pop, "SpatialPixelsDataFrame")
interpolated_pop <- data.frame(population = interpolated_pop.pix$dens, 
                               x = interpolated_pop.pix@coords[,1],
                               y = interpolated_pop.pix@coords[,2]) %>%
  st_as_sf(coords = c("x", "y"),
           crs = st_crs(hamilton_da_sobi))
```

Plot interpolated population:
```{r}
ggplot() + 
  geom_sf(data = interpolated_pop,
          aes(color = population)) +
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1) +
  geom_sf(data = hamilton_da_sobi,
          fill = NA,
          color = "lightgray") + 
  geom_sf(data = sobi_service,
          fill = NA)
```

Extract population cells in the service area (and buffer)
```{r}
interpolated_pop <- interpolated_pop[sobi_service %>% st_buffer(500),]
```

Plot interpolated population again:
```{r}
ggplot() + 
  geom_sf(data = interpolated_pop,
          aes(color = population)) +
  geom_sf(data = hamilton_da_sobi,
          fill = NA,
          color = "lightgray") + 
  geom_sf(data = sobi_service,
          fill = NA) +
  coord_sf(xlim = )
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1)
```

Bounding box of population in serviced areas:
```{r}
bounding_box <- st_bbox(interpolated_pop)
```

Plot:
```{r}
ggplot() + 
  geom_sf(data = interpolated_pop,
          aes(color = population)) +
  geom_sf(data = hamilton_da_sobi,
          fill = NA,
          color = "lightgray") + 
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = st_as_sfc(bounding_box),
          linetype = "dashed",
          fill = NA) +
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1)
```

Adjust bounding box to remove service are in park to the east:
```{r}
bounding_box[3] <- 597200
```

Select population within bounding box:
```{r}
interpolated_pop <- interpolated_pop[st_as_sfc(bounding_box),]
```

Plot:
```{r}
ggplot() + 
  geom_sf(data = interpolated_pop,
          aes(color = population)) +
  geom_sf(data = hamilton_da_sobi,
          fill = NA,
          color = "lightgray") + 
  geom_sf(data = parks,
          color = NA,
          fill = "lightgreen") +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = st_as_sfc(bounding_box),
          linetype = "dashed",
          fill = NA) +
  coord_sf(xlim = c(bounding_box[1], bounding_box[3]), # min & max of x values
           ylim = c(bounding_box[2], bounding_box[4])) + # min & max of y values
  scale_color_distiller(palette = "YlOrRd", 
                        direction = 1)
```

### 2020-11-13 examples/junk.R

@@ -1,0 +1,84 @@
+#pycno_interp.R
+# This script demonstrates how to interpolate 
+# attribute values from a polygon dataset tiling a 
+# region of interest to different set of polygons 
+# covering the same area.
+# Written by A. Shortridge, 8/2019, 11/2019
+
+library(sp)
+library(pycno)
+library(raster)
+require(RColorBrewer)
+
+## Load data
+load("ingham_tracts_zips.RData")
+
+## Plot data
+plot(ingTracts, bor='blue', main="Ingham Co. Zips (red) and Tracts (blue)")
+plot(ingZips, bor='red', add=TRUE)
+plot(ingCo, lty=2, lwd=2, add=TRUE)
+#dev.print(png, "ingham_geogs.png", height=400, width=500)
+
+names(ingZips)
+
+# Set up a color ramp
+OR7 <- brewer.pal(7, "OrRd") # A sequential ramp, Orange to Red
+OrRd <- colorRampPalette(OR7)
+
+## The goal is to interpolate data in the Zip Codes polygons - the number 
+## of kids eligible for reduced/free school lunch - to the census tracts.
+
+## 1. Create the variable
+ingZips$numEligible <- ingZips$FreeEligib + ingZips$ReducedEli
+summary(ingZips$numEligible)
+ingZips$numEligible[is.na(ingZips$numEligible)] <- 0   # Convert NAs to 0
+spplot(ingZips, "numEligible", col.regions=OrRd(17), 
+       main="Lunch Eligible by Zip Code")
+#dev.print(png, "ingham_lunch_zip.png", height=400, width=500)
+
+## 2. Crop tracts to the county extent.
+ingTracts2 <- crop(ingTracts, ingCo)
+
+## 3. Use Pycnophylactic Interpolation to convert.
+# First, interpolate numEligible to a raster with 200x200 m cells
+# Each cell is an estimated count of the number of eligible kids in that cell.
+zipEligDens <- pycno(ingZips, ingZips$numEligible, celldim=200)
+spplot(zipEligDens, col.regions=OrRd(17), main="Lunch Eligible (estimated counts)")
+
+# Crop count surface to Ingham County only
+denRast <- raster(zipEligDens)
+zipEligDens2 <- as(crop(denRast, ingCo), 'SpatialPointsDataFrame')
[…]

### 2020-11-13 examples/junk.R
v1 Removed some junk files

@@ -1,84 +1,0 @@
-#pycno_interp.R
-# This script demonstrates how to interpolate 
-# attribute values from a polygon dataset tiling a 
-# region of interest to different set of polygons 
-# covering the same area.
-# Written by A. Shortridge, 8/2019, 11/2019
-
-library(sp)
-library(pycno)
-library(raster)
-require(RColorBrewer)
-
-## Load data
-load("ingham_tracts_zips.RData")
-
-## Plot data
-plot(ingTracts, bor='blue', main="Ingham Co. Zips (red) and Tracts (blue)")
-plot(ingZips, bor='red', add=TRUE)
-plot(ingCo, lty=2, lwd=2, add=TRUE)
-#dev.print(png, "ingham_geogs.png", height=400, width=500)
-
-names(ingZips)
-
-# Set up a color ramp
-OR7 <- brewer.pal(7, "OrRd") # A sequential ramp, Orange to Red
-OrRd <- colorRampPalette(OR7)
-
-## The goal is to interpolate data in the Zip Codes polygons - the number 
-## of kids eligible for reduced/free school lunch - to the census tracts.
-
-## 1. Create the variable
-ingZips$numEligible <- ingZips$FreeEligib + ingZips$ReducedEli
-summary(ingZips$numEligible)
-ingZips$numEligible[is.na(ingZips$numEligible)] <- 0   # Convert NAs to 0
-spplot(ingZips, "numEligible", col.regions=OrRd(17), 
-       main="Lunch Eligible by Zip Code")
-#dev.print(png, "ingham_lunch_zip.png", height=400, width=500)
-
-## 2. Crop tracts to the county extent.
-ingTracts2 <- crop(ingTracts, ingCo)
-
-## 3. Use Pycnophylactic Interpolation to convert.
-# First, interpolate numEligible to a raster with 200x200 m cells
-# Each cell is an estimated count of the number of eligible kids in that cell.
-zipEligDens <- pycno(ingZips, ingZips$numEligible, celldim=200)
-spplot(zipEligDens, col.regions=OrRd(17), main="Lunch Eligible (estimated counts)")
-
-# Crop count surface to Ingham County only
-denRast <- raster(zipEligDens)
-zipEligDens2 <- as(crop(denRast, ingCo), 'SpatialPointsDataFrame')
[…]

### 2020-11-13 examples/00-Data-Processing-Example.Rmd
v1 Updated 00-Data-Processing-Example.Rmd

@@
 
+Bounding box of population in serviced areas:
+```{r}
+bounding_box <- st_bbox(sobi_service %>%
+                          st_buffer(500))
+```
+
+Plot:
+```{r}
+ggplot() + 
+  geom_sf(data = hamilton_da_2016,
+          fill = NA,
+          color = "lightgray") + 
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  geom_sf(data = st_as_sfc(bounding_box),
+          linetype = "dashed",
+          fill = NA)
+```
+
+Adjust bounding box to remove service are in park to the east:
+```{r}
+bounding_box[3] <- 597200
+```
+
+Plot:
+```{r}
+ggplot() + 
+  geom_sf(data = hamilton_da_2016,
+          fill = NA,
+          color = "lightgray") + 
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  geom_sf(data = st_as_sfc(bounding_box),
+          linetype = "dashed",
+          fill = NA)
+```
+
 Read parks:
@@
 parks <- st_read("Parks.shp")
+```
+
+Read cemeteries:
+```{r}
+cemeteries <- st_read("Cemeteries.shp")
+```
+
+Read environmentally sensitive areas:
+```{r}
+esa <- st_read("Environmentally_Sensitive_Areas_Boundaries.shp")
 ```
@@
 
-Plot the DAs and the service areas:
-```{r}
-ggplot() +
-  geom_sf(data = hamilton_da_2016,
-          color = "lightgray") +
-  geom_sf(data = sobi_service,
-          fill = NA)
-```
-
-We want to work only with the DAs that are within a reasonable buffer of the service areas (e.g., 500 m):
-```{r}
-ggplot(hamilton_da_2016[sobi_service %>% st_buffer(500),]) +
-  geom_sf(color = "lightgray") + 
-  geom_sf(data = sobi_service,
-          fill = NA)
-```
-
 Extract DAs in service areas:
 ```{r}
-hamilton_da_sobi <- hamilton_da_2016[sobi_service %>% st_buffer(500),]
+hamilton_da_sobi <- hamilton_da_2016[st_as_sfc(bounding_box),]
 ```
@@
 ```{r}
-parks_sobi <- parks[sobi_service %>% st_buffer(500),]
+parks_sobi <- st_crop(parks, st_as_sfc(bounding_box))
+```
+
+Extract cemeteries in service areas:
+```{r}
+cemeteries_sobi <- st_crop(cemeteries, st_as_sfc(bounding_box))
+```
+
+Extract environmentally sensitive areas in service areas:
+```{r}
+esa_sobi <- st_crop(esa, st_as_sfc(bounding_box))
 ```
@@
           fill = NA) +
+  scale_fill_distiller(palette = "YlOrRd", direction = 1)
+```
+
+Plot "Green" spaces:
+```{r}
+ggplot() +
   geom_sf(data = parks_sobi,
+          fill = "blue") +
+  geom_sf(data = cemeteries_sobi,
+          fill = "gray") +
+  geom_sf(data = esa_sobi,
           fill = "green") +
+  geom_sf(data = sobi_service,
+          fill = NA,
+          color = "black") +
+  geom_sf(data = hamilton_da_sobi,
+          fill = NA,
+          color = "gray") +
   scale_fill_distiller(palette = "YlOrRd", direction = 1)
+```
+
+Union of the "green space" features:
+```{r}
+parks_sobi <- st_combine(parks_sobi)
+cemeteries_sobi <- st_combine(cemeteries_sobi)
+esa_sobi <- st_combine(esa_sobi)
+```
+
+Check that the topology is valid:
+```{r}
+st_is_valid(parks_sobi)
+st_is_valid(cemeteries_sobi)
+st_is_valid(esa_sobi)
+```
+
+Make valid:
+```{r}
+parks_sobi <- st_make_valid(parks_sobi)
+esa_sobi <- st_make_valid(esa_sobi)
 ```
@@
 ```{r}
-# First combine `parks_sobi` to generate a single geography object, and make valid
-parks_sobi <- st_combine(parks_sobi) %>% 
-  st_make_valid()
-# The new simple features object will be the difference of the two objects
 hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi, parks_sobi)
@@
 
-Plot the new DAs (minus parks) and compare to original DAs:
+Remove the cemeteries from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, cemeteries_sobi)
+```
+
+Remove environmentally sensitive areas from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi, esa_sobi)
+```
+
+Plot the new DAs (minus green) and compare to original DAs:
 ```{r}
@@
 
+Check the population totals:
+```{r}
+sum(hamilton_da_sobi_clean$population)
+sum(interpolated_pop$population)
+```
+
 Plot interpolated population:
@@
 ```{r}
-interpolated_pop <- interpolated_pop[sobi_service %>% st_buffer(500),]
+interpolated_pop_clean <- interpolated_pop[sobi_service %>% st_buffer(500),]
 ```
@@
 ggplot() + 
-  geom_sf(data = interpolated_pop,
+  geom_sf(data = interpolated_pop_clean,
           aes(color = population)) +
-  geom_sf(data = hamilton_da_sobi,
-          fill = NA,
-          color = "lightgray") + 
   geom_sf(data = sobi_service,
-          fill = NA) +
-  coord_sf(xlim = )
-  scale_color_distiller(palette = "YlOrRd", 
-                        direction = 1)
-```
-
-Bounding box of population in serviced areas:
-```{r}
-bounding_box <- st_bbox(interpolated_pop)
-```
-
-Plot:
-```{r}
-ggplot() + 
-  geom_sf(data = interpolated_pop,
-          aes(color = population)) +
-  geom_sf(data = hamilton_da_sobi,
-          fill = NA,
-          color = "lightgray") + 
-  geom_sf(data = sobi_service,
-          fill = NA) +
-  geom_sf(data = st_as_sfc(bounding_box),
-          linetype = "dashed",
           fill = NA) +
@@
 ```
-
-Adjust bounding box to remove service are in park to the east:
-```{r}
-bounding_box[3] <- 597200
-```
-
-Select population within bounding box:
-```{r}
-interpolated_pop <- interpolated_pop[st_as_sfc(bounding_box),]
-```
-
-Plot:
-```{r}
-ggplot() + 
-  geom_sf(data = interpolated_pop,
-          aes(color = population)) +
-  geom_sf(data = hamilton_da_sobi,
-          fill = NA,
-          color = "lightgray") + 
-  geom_sf(data = parks,
-          color = NA,
-          fill = "lightgreen") +
-  geom_sf(data = sobi_service,
-          fill = NA) +
-  geom_sf(data = st_as_sfc(bounding_box),
-          linetype = "dashed",
-          fill = NA) +
-  coord_sf(xlim = c(bounding_box[1], bounding_box[3]), # min & max of x values
-           ylim = c(bounding_box[2], bounding_box[4])) + # min & max of y values
-  scale_color_distiller(palette = "YlOrRd", 
-                        direction = 1)
-```

### 2020-11-16 examples/00-Data-Processing-Example.Rmd
v2 processing
Additional data are processed for the analysis of accessibility

@@
 
-In this notebook we create some examples of processing the data for the analysis of accessibility. Particularly, we're interested in spatial interpolationo of the population for small areas.
+## Introduction
+
+In this notebook we create some examples of processing the data for the analysis of accessibility to Hamilton Bike share docking stations.
+Particularly, we're interested in spatial interpolation of the population for small areas.
+
+Data used in this notebook was retrieved from two sources: Open Hamilton and the 2016 Canadian Census.
+Open Hamilton is an online repository of data from the City of Hamilton.
+The following datasets were retrieved:
+
+### SoBi Hub
+
+The dataset was last updated on *November 9, 2020* and was downloaded for this project on *November 10, 2020*.
+This dataset contains the following attributes that are of interest to this research: location of hubs (address, longitude, and latitude) and the number of racks available at each hub.
+The dataset also includes the number of bicycles available at each hub, but since this changes daily, this information is not useful for our purposes.
+The maximum number of bicycles that a hub *could* have (i.e., the baseline) available is important for calculating accessibility in a given area.
+
+We can also retrieve the data weekly over a certain period, for example 4 weeks, to calculate the average number of bicycles that are at each hub.
+Variations in the number of bicycles could also be informative.
+For instance, some hubs may systematically have more or less bicycles than others (i.e., excess of supply around the University campus).
+
+### SoBi Service Area
+
+The dataset was last updated on *November 9, 2020* and was downloaded for this project on *November 10, 2020*.
+Hamilton's bike share program is available only in the lower city from Dundas to Ottawa Street North.
+There is one isolated hub beyond this service area at Van Wagner's Beach for tourism purposes.
+
+### Golf Courses
+
+The dataset was last updated on *November 1, 2020* and was downloaded for this project on *November 10, 2020*. This dataset contains the location of City and privately owned golf courses in Hamilton.
+
+### Parks
+
+The dataset was last updated on *November 8, 2020* and was downloaded for this project on *November 10, 2020*. This dataset contains the location of parks and other green spaces in Hamilton.
+
+### Large Employment Areas
+
+The dataset contains Employment Land boundaries in Hamilton and was last updated on *November 12, 2020*.
+It was downloaded for this project on *November 13, 2020*.
+This dataset contains the location of large business parks and industrial lands in Hamilton.
+
+### Municipal Parking Lots
+
+The dataset contains the location of municipal car parks in Hamilton and was last updated on *November 9, 2020*. It was downloaded for this project on *November 13, 2020*.
+
+### Cemeteries
+
+The dataset contains the location of cemeteries in Hamilton. It was downloaded for this project on *November 13, 2020*.
+
+### Environmentally Sensitive Areas
+
+The dataset contains the location of Environmentally Sensitive Areas (ESAs) in Hamilton.
+ESAs are either land or water areas containing natural features or significant ecological functions in Hamilton.
+The data were downloaded on *November 13, 2020*.
+
+### Street Centreline
+
+The dataset contains the street network in Hamilton.
+Data attributes include road classification so highways can be extracted.
+The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
+
+### Educational Institutions
+
+The dataset contains the location of all educational institutions and schools in Hamilton. The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
+
+### Places of Workshop
+
+The dataset contains the location of buildings used for religious congregations in Hamilton. The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
+
+### Municipal Service Centres
+
+The dataset contains the location of all municipal service centres in Hamilton.
+This includes Hamilton City Hall.
+The dataset was last updated on *November 11, 2020* and downloaded on *November 16, 2020*.
+
+### Recreation and Community Centres
+
+The dataset contains the location of all recreation and community centres in Hamilton.
+This includes Hamilton City Hall.
+The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
+
+### Arenas
+
+The dataset contains the location of all indoor arenas in Hamilton. This includes Hamilton City Hall. The dataset was last updated on *November 12, 2020* and downloaded on *November 16, 2020*.
+
+### EMS Stations
+
+The dataset contains the location of all Emergency Management Services (EMS) Ambulance stations in Hamilton.
+This includes Hamilton City Hall.
+The dataset was last updated on *November 11, 2020* and downloaded on *November 16, 2020*.
+
+### Fire Stations
+
+The dataset contains the location of all fire stations in Hamilton. This includes Hamilton City Hall. The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
+
+### Police Stations
+
+The dataset contains the location of all police stations in Hamilton. This includes Hamilton City Hall. The dataset was last updated on *November 13, 2020* and downloaded on *November 16, 2020*.
+
+### Hospitals
+
+The dataset contains the location of all hospitals in Hamilton. This includes Hamilton City Hall. The dataset was last updated on *April 2, 2019* and downloaded on *November 16, 2020*.
 
 ## Preliminaries
+
+Clear environment:
+```{r}
+rm(list = ls())
+```
 
@@
 
-Adjust bounding box to remove service are in park to the east:
+Adjust bounding box to remove docking station in park, which is for tourism purposes, to the east:
 ```{r}
@@
 
-Project the Hamilton DA zones to the same projection of the SoBi Service Areas:
+Read golf courses:
+```{r}
+golf_courses <- st_read("Golf_Courses.shp")
+```
+
+Read employment land areas, which include business parks and industrial lands:
+```{r}
+employment_lands <- st_read("Employment_Lands.shp")
+```
+
+Read railways:
+```{r}
+railways <- st_read("Railways.shp")
+```
+
+Read municipal parking lots:
+```{r}
+municipal_parking <- st_read("Municipal_Parking_Lots.shp")
+```
+
+Read street network:
+```{r}
+streets <- st_read("Street_Centreline.shp")
+```
+
+Filter for provincial highways:
+```{r}
+streets <- streets %>% dplyr::filter(ROAD_CLASS == "Provincial Highway")
+```
+
+Read educational institutions:
+```{r}
+schools <- st_read("Educational_Institutions.shp")
+```
+
+Read places of worship:
+```{r}
+religious_places <- st_read("Places_of_Worship.shp")
+```
+
+Read municipal service centres:
+```{r}
+service_centres <- st_read("Municipal_Service_Centres.shp")
+```
+
+Read recreation and centres:
+```{r}
+recreation_centres <- st_read("Recreation_and_Community_Centres.shp")
+```
+
+Read arenas:
+```{r}
+arenas <- st_read("Arenas.shp")
+```
+
+Read EMS stations:
+```{r}
+ems <- st_read("EMS_Stations.shp")
+```
+
+Read fire stations:
+```{r}
+fire <- st_read("Fire_Stations.shp")
+```
+
+Read police stations:
+```{r}
+police <- st_read("Police_Stations.shp")
+```
+
+Read hospitals:
+```{r}
+hospitals <- st_read("Hospitals.shp")
+```
+
+Project the Hamilton DA zones to the same projection of the SoBi Service Area:
 ```{r}
@@
 
-Extract DAs in service areas:
+Extract DAs in service area:
 ```{r}
@@
 
-Extract parks in service areas:
+Extract parks in service area:
 ```{r}
@@
 
-Extract cemeteries in service areas:
+Extract cemeteries in service area:
 ```{r}
@@
 
-Extract environmentally sensitive areas in service areas:
+Extract environmentally sensitive areas in service area:
 ```{r}
 esa_sobi <- st_crop(esa, st_as_sfc(bounding_box))
+```
+
+Extract golf courses in service area:
+```{r}
+golf_sobi <- st_crop(golf_courses, st_as_sfc(bounding_box))
+```
+
+Extract employment lands in service area:
+```{r}
+employment_sobi <- st_crop(employment_lands, st_as_sfc(bounding_box))
+```
+
+Extract railways in service area:
+```{r}
+railways_sobi <- st_crop(railways, st_as_sfc(bounding_box))
+```
+Extract municipal parking lots in service area:
+```{r}
+parking_sobi <- st_crop(municipal_parking, st_as_sfc(bounding_box))
+```
+
+Extract provincial highways in service area:
+```{r}
+highways_sobi <- st_crop(streets, st_as_sfc(bounding_box))
+```
+
+Extract educational institutions in service area:
+```{r}
+schools_sobi <- st_crop(schools, st_as_sfc(bounding_box))
+```
+
+Extract places of worship in service area:
+```{r}
+religious_sobi <- st_crop(religious_places, st_as_sfc(bounding_box))
+```
+
+Extract municipal service centres in service area:
+```{r}
+service_sobi <- st_crop(service_centres, st_as_sfc(bounding_box))
+```
+
+Extract recreation and community centres in service area:
+```{r}
+recreation_sobi <- st_crop(recreation_centres, st_as_sfc(bounding_box))
+```
+
+Extract indoor arenas in service area:
+```{r}
+arenas_sobi <- st_crop(arenas, st_as_sfc(bounding_box))
+```
+
+Extract educational institutions in service area:
+```{r}
+schools_sobi <- st_crop(schools, st_as_sfc(bounding_box))
+```
+
+Extract EMS stations in service area:
+```{r}
+ems_sobi <- st_crop(ems, st_as_sfc(bounding_box))
+```
+
+Extract fire stations in service area:
+```{r}
+fire_sobi <- st_crop(fire, st_as_sfc(bounding_box))
+```
+
+Extract police stations in service area:
+```{r}
+police_sobi <- st_crop(police, st_as_sfc(bounding_box))
+```
+
+Extract hospitals in service area:
+```{r}
+hospitals_sobi <- st_crop(hospitals, st_as_sfc(bounding_box))
 ```
@@
           fill = "green") +
+  geom_sf(data = golf_sobi,
+          fill = "lightskyblue") +
+  geom_sf(data = employment_sobi,
+          fill = "purple") +
+  geom_sf(data = railways_sobi,
+          fill = "wheat") +
+  geom_sf(data = parking_sobi,
+          fill = "lightpink") +
+  geom_sf(data = highways_sobi,
+          fill = "orange") +
+  geom_sf(data = schools_sobi,
+          fill = "red") +
+  geom_sf(data = religious_sobi,
+          fill = "pink") +
+  geom_sf(data = service_sobi,
+          fill = "yellow") +
+  geom_sf(data = recreation_sobi,
+          fill = "yellow") +
+  geom_sf(data = arenas_sobi,
+          fill = "yellow") +
+  geom_sf(data = ems_sobi,
+          fill = "yellow") +
+  geom_sf(data = ems_sobi,
+          fill = "yellow") +
+  geom_sf(data = fire_sobi,
+          fill = "yellow") +
+  geom_sf(data = police_sobi,
+          fill = "yellow") +
+  geom_sf(data = hospitals_sobi,
+          fill = "yellow") +
   geom_sf(data = sobi_service,
@@
 esa_sobi <- st_combine(esa_sobi)
+golf_sobi <- st_combine(golf_sobi)
+employment_sobi <- st_combine(employment_sobi)
+railways_sobi <- st_combine(railways_sobi)
+parking_sobi <- st_combine(parking_sobi)
+highways_sobi <- st_combine(highways_sobi)
+schools_sobi <- st_combine(schools_sobi)
+religious_sobi <- st_combine(religious_sobi)
+service_sobi <- st_combine(service_sobi)
+recreation_sobi <- st_combine(recreation_sobi)
+arenas_sobi <- st_combine(arenas_sobi)
+ems_sobi <- st_combine(ems_sobi)
+fire_sobi <- st_combine(fire_sobi)
+police_sobi <- st_combine(police_sobi)
+hospitals_sobi <- st_combine(hospitals_sobi)
 ```
@@
 st_is_valid(esa_sobi)
+st_is_valid(golf_sobi)
+st_is_valid(employment_sobi)
+st_is_valid(railways_sobi)
+st_is_valid(parking_sobi)
+st_is_valid(highways_sobi)
+st_is_valid(schools_sobi)
+st_is_valid(religious_sobi)
+st_is_valid(service_sobi)
+st_is_valid(recreation_sobi)
+st_is_valid(arenas_sobi)
+st_is_valid(ems_sobi)
+st_is_valid(fire_sobi)
+st_is_valid(police_sobi)
+st_is_valid(hospitals_sobi)
 ```
@@
 esa_sobi <- st_make_valid(esa_sobi)
+parking_sobi <- st_make_valid(parking_sobi)
 ```
@@
 ```{r}
-hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi, esa_sobi)
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, esa_sobi)
+```
+
+Remove golf courses from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, golf_sobi)
+```
+
+Remove employment lands from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, employment_sobi)
+```
+
+Remove railways from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, railways_sobi)
+```
+
+Remove municipal parking lots from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, parking_sobi)
+```
+
+Remove provincial highways from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, highways_sobi)
+```
+
+Remove educational institutions from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, schools_sobi)
+```
+
+Remove places of worship from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, religious_sobi)
+```
+
+Remove municipal service centres from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, service_sobi)
+```
+
+Remove recreation and community centres from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, recreation_sobi)
+```
+
+Remove arenas from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, arenas_sobi)
+```
+
+Remove EMS stations from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, ems_sobi)
+```
+Remove fire stations from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, fire_sobi)
+```
+Remove police stations from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, police_sobi)
+```
+Remove hospitals from the DAs (use `st_difference()`):
+```{r}
+hamilton_da_sobi_clean <- st_difference(hamilton_da_sobi_clean, hospitals_sobi)
 ```

### 2020-11-17 examples/00-Data-Processing-Example.Rmd
v1 Added routing process from Chris' accessibility notebooks

@@
 The following datasets were retrieved:
+
+## Provenance of data
 
@@
 ```{r message = FALSE}
+library(disk.frame)
 library(pycno)
 library(readr)
-library(rgrass7)
+library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
 library(sf)
@@
 ```
+
+Define some parameters for `disk.frame` and `r5r`:
+```{r}
+setup_disk.frame()
+options(scipen = 999)
+options(java.parameters = "-Xmx6G")
+options(future.globals.maxSize = Inf)
+```
+
+### Read data
 
@@
 
+Plot service areas:
+```{r}
+ggplot() +
+  geom_sf(data = sobi_service,
+          aes(fill = AREA_NAME))
+```
+
+There is a Core Service Area and a much smaller service area at Van Wagner's. This is a location at the beach with service for beach goers. We will remove this area from further analysis:
+```{r}
+sobi_service <- sobi_service %>%
+  dplyr::filter(AREA_NAME == "Core Service Area")
+```
+
 Bounding box of population in serviced areas:
@@
                           st_buffer(500))
-```
-
-Plot:
-```{r}
-ggplot() + 
-  geom_sf(data = hamilton_da_2016,
-          fill = NA,
-          color = "lightgray") + 
-  geom_sf(data = sobi_service,
-          fill = NA) +
-  geom_sf(data = st_as_sfc(bounding_box),
-          linetype = "dashed",
-          fill = NA)
-```
-
-Adjust bounding box to remove docking station in park, which is for tourism purposes, to the east:
-```{r}
-bounding_box[3] <- 597200
 ```
@@
 
+### Extract features in service area
+
 Project the Hamilton DA zones to the same projection of the SoBi Service Area:
@@
 
+### Read population
+
 Read population table (2016 Census of Population, Population and Dwelling counts):
@@
 ```
+
+### Visualize geographical data
 
@@
 
+### Prepare geographical data for interpolation
+
 Union of the "green space" features:
@@
 ```
+
+### Remove all features that are not population from the DAs
 
@@
 
+## Interpolate population
+
 Pycnophylactic interpolation needs a `SpatialPolygonsDataFrame` object. Convert the simple features object:
@@
 
+## Extract population in service area
+
 Extract population cells in the service area (and buffer)
@@
 interpolated_pop_clean <- interpolated_pop[sobi_service %>% st_buffer(500),]
+```
+
+Note that some cells have population values of zero:
+```{r}
+interpolated_pop_clean %>% select(population) %>% summary()
+```
+
+Filter all cells with non-zero population values:
+```{r}
+interpolated_pop_clean <- interpolated_pop_clean %>%
+            dplyr::filter(population > 0)
+```
+
+Add identifier for the population cells:
+```{r}
+interpolated_pop_clean <- interpolated_pop_clean %>%
+            mutate(UID = 1:n())
 ```
@@
 ```
+
+## Bikeshare system
+
+Read Sobi hubs:
+```{r}
+sobi_hubs <- st_read("SoBi_Hubs.shp")
+```
+
+Plot interpolated population and SoBi hubs:
+```{r}
+ggplot() + 
+  geom_sf(data = interpolated_pop_clean,
+          aes(color = population)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  geom_sf(data = sobi_hubs) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+Remove SoBi hub on beach:
+```{r}
+sobi_hubs <- sobi_hubs %>%
+  dplyr::filter(NAME != "Van Wagners")
+```
+
+Plot again:
+```{r}
+ggplot() + 
+  geom_sf(data = interpolated_pop_clean,
+          aes(color = population)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  geom_sf(data = sobi_hubs,
+          aes(size = RACKS_AMOU)) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  scale_size(range = c(1, 4))
+```
+
+## Routing
+
+I used [BBBike](https://download.bbbike.org/osm/bbbike/) to extract OSM data for Hamilton - not the whole city, but enough to cover SoBI's service area.
+The name of the file is `planet_-80.042,43.183_-79.748,43.309.osm.pbf`.
+Copy to folder `r5_graph`.
+
+Set Up R5 Routing. First define the path to where the graph is located:
+```{r set up r5 path, include=FALSE}
+r5_path <- file.path("./r5_graph")
+```
+
+Build the graph:
+```{r build-graph, include = FALSE}
+r5_sobi <- setup_r5(data_path = r5_path, verbose = FALSE)
+```
+
+Prepare Input Data for `r5r`. The origins are the coordinates of the population cells and the destinations the coordinates of the SoBi hubs:
+```{r prepare-inputs}
+# save origin centroids in format expected by R5R (id, lon, lat)
+origins_i <- data.frame(UID = interpolated_pop_clean$UID, 
+              interpolated_pop_clean %>%
+                st_transform(crs = 4326) %>%
+                st_coordinates()) %>%
+    rename(lon = X, lat = Y, id = UID) %>%
+    select(id, lon, lat)
+
+# now SoBi hubs
+destinations_j <- data.frame(OBJECTID = sobi_hubs$OBJECTID, 
+              sobi_hubs %>%
+                st_transform(crs = 4326) %>%
+                st_coordinates()) %>%
+    rename(lon = X, lat = Y, id = OBJECTID) %>%
+    select(id, lon, lat)
+```
+
+Calculate OD Matrix for Walking:
+```{r calculate walking od matrix, include=FALSE}
+# set up batching according to how many origin rows to process at one time
+chunksize = 2000 # larger chunks for walking will give enough origins in each chunk to allow multiprocessing to spin up with R5R
+num_chunks = ceiling(nrow(origins_i)/chunksize)
+
+# create origin-destination pairs
+origins_chunks <- as.disk.frame(origins_i,
+                          outdir = "./df/origins_i",
+                          nchunks = num_chunks,
+                          overwrite = TRUE)
+
+start.time <- Sys.time()
+pb <- txtProgressBar(0, num_chunks, style = 3)
+
+for (i in 1:num_chunks){
+  origins_i_chunk <- get_chunk(origins_chunks, i)
+  ttm_chunk <- travel_time_matrix(r5_sobi,
+                          origins = origins_i_chunk,
+                          destinations = destinations_j,
+                          mode = c("WALK"),
+                          departure_datetime = as.POSIXct(strptime("2020-03-02 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
+                          max_walk_dist = 10000, # metres
+                          max_trip_duration = 30)
+
+  # export output as disk.frame
+  ifelse(i == 1, output_df <- as.disk.frame(ttm_chunk,
+                                            nchunks = 1,
+                                            outdir = "./df/output_ttm_walk",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_df, ttm_chunk, chunk_id = i))
+  setTxtProgressBar(pb, i)
+}
+end.time <- Sys.time()
+ttm_walk_time <- end.time - start.time
+ttm_walk_time
+```
+
+## Extract travel time matrix
+
+```{r load od matrix for walk and subway, include=FALSE}
+# connect to the walking travel time matrix disk frame
+ttm_walk.disk.frame <- disk.frame("./df/output_ttm_walk")
+#opportunities_j <- as.data.frame(nyc_cb_point) %>% select(GEOID10, total_emp) %>% rename(toId = GEOID10, o_j = total_emp)
+```
+
+Convert disk.frame to data frame:
+```{r}
+ttm_walk <- as.data.frame(ttm_walk.disk.frame) %>%
+  transmute(UID = as.numeric(fromId), OBJECTID = as.numeric(toId), travel_time)
+```
+
+## Save data to disk
+
+Rename data frame:
+```{r}
+population_sobi <- interpolated_pop_clean
+```
+
+Save data files:
+```{r}
+save(population_sobi, file = "population_sobi.RData")
+save(sobi_hubs, file = "sobi_hubs.RData")
+save(sobi_service, file = "sobi_service.RData")
+save(ttm_walk, file = "ttm_walk.RData")
+```

### 2020-11-18 examples/Toy-example-balanced-accessibility.Rmd
v1 Added toy example to illustrate the calculation of balanced FCA accessibility

---
title: "R Notebook"
output: html_notebook
---

Load libraries:
```{r message=FALSE}
library(data.table)
library(dplyr)
library(gdistance)
library(kableExtra)
library(raster)
library(rgdal)
library(sf)
library(tidyverse)
library(units)
```

# Load data

Load geography:
```{r}
load("Simulated Example.RData")
```

Create simple features objects for population centers and clinics:
```{r}
  # Clinics
  sim_clinics <- st_sf(Clinics_data, geometry_clinics, stringsAsFactors = TRUE)

  # Population centers
  sim_population <- st_sf(Population_data, geometry_population, stringsAsFactors = TRUE)
```

Calculate the distance between these points as a measure of cost:
```{r}
distance_matrix <- st_distance(sim_population, sim_clinics)
```

What is in these objects?
```{r}
head(Clinics_data)
head(Population_data)
head(distance_matrix)
```

# Threshold

Calculate quantiles according to distance. This is done to select a threshold:
```{r}
quantiles_d <- quantile(as.vector(distance_matrix), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))

dist_threshold <- quantile(as.vector(distance_matrix), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))[[5]]
```

The distance threshold is equivalent to the _kernel bandwidth_ or _catchment distance_.

# Prepare travel matrix in long format

Convert matrix to data.frame:
```{r}
distance_matrix_long <- as.data.table(distance_matrix)
```

Rename columns:
```{r}
names(distance_matrix_long) <- c('clinic_1', 'clinic_2', 'clinic_3')
distance_matrix_long$id <- Population_data$id
```

Melt distance matrix:
```{r}
distance_matrix_long <- distance_matrix_long %>%
  pivot_longer(cols = starts_with("clinic_"),
                     names_to = "to",
                     values_to = "distance")
```

```{r}
#distance_matrix_long <- data.table::melt(distance_matrix_long, id.vars='from', variable.name = "to", value.name = "dist")
```

Add population information:
```{r}
distance_matrix_long <- left_join(distance_matrix_long, 
                                  Population_data,
                                  by = "id")
```

Add  doctors information:
```{r}
Clinics_data$id <- c('clinic_1', 'clinic_2', 'clinic_3')
#distance_matrix_long$dest <- as.character(distance_matrix_long$dest)
long_matrix <- left_join(distance_matrix_long, Clinics_data, by=c("to" = "id"))
#setDT(long_matrix)
head(long_matrix)
```

Calculate impedance (binary)
```{r eval=TRUE}
# Run only if binary function desired
# Change chunk option eval=TRUE to FALSE if impedance is Gaussian function
long_matrix <- long_matrix %>%
  mutate(distance = drop_units(distance),
         impedance = ifelse(distance <= dist_threshold, 1, 0))
```

Calculate impedance (Gaussian):
```{r eval=FALSE}
# Run only if Gaussian function desired
# Change chunk option eval=FALSE to TRUE if impedance is Gaussian function
long_matrix <- long_matrix %>%
  mutate(distance = drop_units(distance),
         impedance = case_when(distance <= quantiles_d[1] ~ 0.945,
                               (distance > quantiles_d[1]) & (distance <= quantiles_d[3]) ~ 0.600,
                               (distance > quantiles_d[3]) & (distance <= quantiles_d[5]) ~ 0.242,
                               distance > quantiles_d[5] ~ 0))
```

Calculate the sum of the impedance by population for the balancing factors. In the binary case, this should be the same as the number of clinics that each population center can reach; when the impedance is not binary, this is the sum of all impedance values:
```{r}
sum_b1 <- long_matrix %>%
  group_by(id) %>%
  summarize(sum_b1 = sum(impedance),
            .groups = "drop")
sum_b1
```

The population at 1 can only reach one clinic, the population at 2 can reach two clinics, the population at 3 can reach all three clinics, etc.

Calculate the sum of the impedance by clinic for the balancing factors. In the binary case, this should be the same as the number of population centers that each clinic serves; when the impedance is not binary, this is the sum of all impedance values:
```{r}
sum_b2 <- long_matrix %>%
  group_by(to) %>%
  summarize(sum_b2 = sum(impedance))
sum_b2
```

Clinic 1 serves four population centers, clinic 2 serves eight population centers, and clinic 3 serves seven population centers.

Next, join the sum of the impedance to the table to calculate the balanced impedance values:
```{r}
long_matrix <- long_matrix %>%
  left_join(sum_b1, by = "id") %>%
  left_join(sum_b2, by = "to")
```

Calculate balanced impedance values:
```{r}
long_matrix <- long_matrix %>%
  mutate(balanced_impedance_1 = impedance/sum_b1,
         balanced_impedance_2 = impedance/sum_b2)
```

# Calculate level of service

Check the balanced_impedance factors for the population centers:
```{r}
long_matrix %>%
  filter(id == "Population 3") 
```

Notice that the population of each center is sorted among the three clinics proportionally. Thanks to this property, the sum of the population serviced equals the total population in the system:
```{r}
long_matrix %>%
  group_by(id) %>%
  summarize(pop_served = sum(sum(Population * balanced_impedance_1)),
            .groups = "drop") %>%
  pull(pop_served) %>%
  sum()

Population_data %>% 
  pull(Population) %>% 
  sum()
```

The level of service of each clinic is the number of doctors at the clinic, divided by the population that they serve:
```{r}
los <- long_matrix %>%
  filter(impedance > 0) %>% # filter impedance == 0 to avoid divisions by zero
  group_by(to) %>%
  summarize(los = first(Doctors) / sum((Population * balanced_impedance_1)),
            .groups = "drop")
los
```

Join the level of service to the table:
```{r}
long_matrix <- long_matrix %>%
  left_join(los, by = "to")
```

# Calculate accessibility

To calculate accessibility, the level of service of each clinic needs to be distributed proportionally to the population centers. Check the 
```{r}
accessibility <- long_matrix %>%
  group_by(id) %>%
  summarize(accessibility = sum(los * balanced_impedance_2),
            .groups = "drop")
accessibility
```

Check that the sum of the accessibility is equal to the available level of service:
```{r}
accessibility %>% 
  pull(accessibility) %>% 
  sum()

los %>% 
  pull(los) %>%
  sum()
```

# Complete  (or tl;dr version)

The full algorithm can be written in this way.

Obtain the matrix in long form (this is line 85 in this file):
```{r}
long_matrix <- left_join(distance_matrix_long, Clinics_data, by=c("to" = "id"))
```

Calculate the accessibility and level of service:
```{r}
# Calculate impedance (binary)
# long_matrix <- long_matrix %>%
#   mutate(distance = drop_units(distance),
#          impedance = ifelse(distance <= dist_threshold, 1, 0))

# Calculate impedance (Gaussian)
long_matrix <- long_matrix %>%
  mutate(distance = drop_units(distance),
         impedance = case_when(distance <= quantiles_d[1] ~ 0.945,
                               (distance > quantiles_d[1]) & (distance <= quantiles_d[3]) ~ 0.600,
                               (distance > quantiles_d[3]) & (distance <= quantiles_d[5]) ~ 0.242,
                               distance > quantiles_d[5] ~ 0))

# Calculate first set of balancing factors
sum_b1 <- long_matrix %>%
  group_by(id) %>%
  summarize(sum_b1 = sum(impedance),
            .groups = "drop")

# Calculate second set of balancing factors
sum_b2 <- long_matrix %>%
  group_by(to) %>%
  summarize(sum_b2 = sum(impedance),
            .groups = "drop")

# Join the sum of the impedance to the table to calculate the balanced impedance values:
long_matrix <- long_matrix %>%
  left_join(sum_b1, by = "id") %>%
  left_join(sum_b2, by = "to")

# Calculate the balanced impedance
long_matrix <- long_matrix %>%
  mutate(balanced_impedance_1 = impedance/sum_b1,
         balanced_impedance_2 = impedance/sum_b2)

# Calculate the level of service
los <- long_matrix %>%
  filter(impedance > 0) %>% # filter impedance == 0 to avoid divisions by zero
  group_by(to) %>%
  summarize(los = first(Doctors) / sum((Population * balanced_impedance_1)),
            .groups = "drop")

# Join the level of service to the table:
long_matrix <- long_matrix %>%
  left_join(los, by = "to")

# Calculate accessibility
accessibility <- long_matrix %>%
  group_by(id) %>%
  summarize(accessibility = sum(los * balanced_impedance_2),
            .groups = "drop")
```

This is the accessibility:
```{r}
accessibility
```

This is the level of service:
```{r}
los
```

### 2021-01-25 examples/Toy-example-balanced-accessibility.Rmd
Calculate SoBi LOS and accessibility

@@
 ```{r}
-load("Simulated Example.RData")
+load("population_sobi.RData") #interpolated population in SoBi area
+load("sobi_hubs.RData") #location of SoBi hubs, minus Van Wagner's beach
+load("sobi_service.RData") #core service area for SoBi
+load("ttm_walk.RData") #travel time matrix for interpolated population in SoBi area
 ```
 
-Create simple features objects for population centers and clinics:
 ```{r}
-  # Clinics
-  sim_clinics <- st_sf(Clinics_data, geometry_clinics, stringsAsFactors = TRUE)
-
-  # Population centers
-  sim_population <- st_sf(Population_data, geometry_population, stringsAsFactors = TRUE)
+summary(ttm_walk) #time is in minutes
 ```
-
-Calculate the distance between these points as a measure of cost:
-```{r}
-distance_matrix <- st_distance(sim_population, sim_clinics)
-```
-
-What is in these objects?
-```{r}
-head(Clinics_data)
-head(Population_data)
-head(distance_matrix)
-```
-
 # Threshold
@@
 ```{r}
-quantiles_d <- quantile(as.vector(distance_matrix), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))
+quantiles_d <- quantile(as.vector(ttm_walk$travel_time), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))
 
-dist_threshold <- quantile(as.vector(distance_matrix), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))[[5]]
+dist_threshold <- quantile(as.vector(ttm_walk$travel_time), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))[[5]]
 ```
@@
 
-# Prepare travel matrix in long format
+# Set up matrix
 
-Convert matrix to data.frame:
+Add population information:
 ```{r}
-distance_matrix_long <- as.data.table(distance_matrix)
+ttm_walk <- left_join(ttm_walk, 
+                      population_sobi,
+                      by = "UID")
+```
+
+Drop the geometry:
+```{r}
+ttm_walk <- ttm_walk %>% dplyr::select(-geometry)
+```
+
+Add number of bicycle racks per SoBi hub as a measure of the level of supply:
+```{r}
+ttm_walk <- left_join(ttm_walk, sobi_hubs %>% dplyr::select(OBJECTID, RACKS_AMOU), by=c("OBJECTID" = "OBJECTID"))
+ttm_walk <- ttm_walk %>% dplyr::select(-geometry)
+head(ttm_walk)
 ```
@@
 ```{r}
-names(distance_matrix_long) <- c('clinic_1', 'clinic_2', 'clinic_3')
-distance_matrix_long$id <- Population_data$id
+names(ttm_walk) <- c('UID', 'hub', 'travel_time', 'population', 'racks')
 ```
 
-Melt distance matrix:
-```{r}
-distance_matrix_long <- distance_matrix_long %>%
-  pivot_longer(cols = starts_with("clinic_"),
-                     names_to = "to",
-                     values_to = "distance")
-```
+[**Note for Dr. Paez: do we want the binary and Gaussian functions? Or only one of them?**]
 
-```{r}
-#distance_matrix_long <- data.table::melt(distance_matrix_long, id.vars='from', variable.name = "to", value.name = "dist")
-```
-
-Add population information:
-```{r}
-distance_matrix_long <- left_join(distance_matrix_long, 
-                                  Population_data,
-                                  by = "id")
-```
-
-Add  doctors information:
-```{r}
-Clinics_data$id <- c('clinic_1', 'clinic_2', 'clinic_3')
-#distance_matrix_long$dest <- as.character(distance_matrix_long$dest)
-long_matrix <- left_join(distance_matrix_long, Clinics_data, by=c("to" = "id"))
-#setDT(long_matrix)
-head(long_matrix)
-```
-
-Calculate impedance (binary)
+Calculate impedance (binary):
 ```{r eval=TRUE}
@@
 # Change chunk option eval=TRUE to FALSE if impedance is Gaussian function
-long_matrix <- long_matrix %>%
-  mutate(distance = drop_units(distance),
-         impedance = ifelse(distance <= dist_threshold, 1, 0))
+ttm_walk <- ttm_walk %>%
+  mutate(impedance_binary = ifelse(travel_time <= dist_threshold, 1, 0))
 ```
@@
 # Change chunk option eval=FALSE to TRUE if impedance is Gaussian function
-long_matrix <- long_matrix %>%
-  mutate(distance = drop_units(distance),
-         impedance = case_when(distance <= quantiles_d[1] ~ 0.945,
-                               (distance > quantiles_d[1]) & (distance <= quantiles_d[3]) ~ 0.600,
-                               (distance > quantiles_d[3]) & (distance <= quantiles_d[5]) ~ 0.242,
-                               distance > quantiles_d[5] ~ 0))
+ttm_walk <- ttm_walk %>%
+  mutate(impedance_gaussian = case_when(travel_time <= quantiles_d[1] ~ 0.945,
+                               (travel_time > quantiles_d[1]) & (travel_time <= quantiles_d[3]) ~ 0.600,
+                               (travel_time > quantiles_d[3]) & (travel_time <= quantiles_d[5]) ~ 0.242,
+                               travel_time > quantiles_d[5] ~ 0))
 ```
@@
 Calculate the sum of the impedance by population for the balancing factors.
-In the binary case, this should be the same as the number of clinics that each population center can reach; when the impedance is not binary, this is the sum of all impedance values:
+In the binary case, this should be the same as the number of SoBi hubs that each interpolated population unit can reach; when the impedance is not binary, this is the sum of all impedance values:
 ```{r}
-sum_b1 <- long_matrix %>%
-  group_by(id) %>%
-  summarize(sum_b1 = sum(impedance),
+sum_b1 <- ttm_walk %>%
+  group_by(UID) %>%
+  summarize(sum_b1 = sum(impedance_binary),
             .groups = "drop")
@@
 
-The population at 1 can only reach one clinic, the population at 2 can reach two clinics, the population at 3 can reach all three clinics, etc.
+The interpolated population at unit 15 can reach eight SoBi hubs, the population at unit 16 can reach six SoBi hubs, the population at unit 30 can reach zero hubs clinics, etc.
+[**Note for Dr.
+Paez: why do the unit IDs start at 15 and not 1?**
 
-Calculate the sum of the impedance by clinic for the balancing factors.
-In the binary case, this should be the same as the number of population centers that each clinic serves; when the impedance is not binary, this is the sum of all impedance values:
+Calculate the sum of the impedance by SoBi hubs for the balancing factors.
+In the binary case, this should be the same as the number of interpolated population unit that each SoBi hub serves; when the impedance is not binary, this is the sum of all impedance values:
 ```{r}
-sum_b2 <- long_matrix %>%
-  group_by(to) %>%
-  summarize(sum_b2 = sum(impedance))
+sum_b2 <- ttm_walk %>%
+  group_by(hub) %>%
+  summarize(sum_b2 = sum(impedance_binary))
 sum_b2
@@
 
-Clinic 1 serves four population centers, clinic 2 serves eight population centers, and clinic 3 serves seven population centers.
+Hub 1 serves 1502 interpolated population units, hub 2 serves 1393 population units, etc.
 
@@
 ```{r}
-long_matrix <- long_matrix %>%
-  left_join(sum_b1, by = "id") %>%
-  left_join(sum_b2, by = "to")
+ttm_walk <- ttm_walk %>%
+  left_join(sum_b1, by = "UID") %>%
+  left_join(sum_b2, by = "hub")
 ```
@@
 ```{r}
-long_matrix <- long_matrix %>%
-  mutate(balanced_impedance_1 = impedance/sum_b1,
-         balanced_impedance_2 = impedance/sum_b2)
+ttm_walk <- ttm_walk %>%
+  mutate(balanced_impedance_1 = impedance_binary/sum_b1,
+         balanced_impedance_2 = impedance_binary/sum_b2)
 ```
@@
 
-Check the balanced_impedance factors for the population centers:
+Check the balanced_impedance factors for the population units:
 ```{r}
-long_matrix %>%
-  filter(id == "Population 3") 
+ttm_walk %>% 
+   dplyr::filter(hub == "46") 
 ```
 
-Notice that the population of each center is sorted among the three clinics proportionally. Thanks to this property, the sum of the population serviced equals the total population in the system:
+[**Note: not sure if this step is needed for SoBi?**] *Notice that the population of each center is sorted among the three clinics proportionally.
+Thanks to this property, the sum of the population serviced equals the total population in the system:*
 ```{r}
-long_matrix %>%
-  group_by(id) %>%
-  summarize(pop_served = sum(sum(Population * balanced_impedance_1)),
+ttm_walk %>%
+  group_by(UID) %>%
+  summarize(pop_served = sum(sum(population * balanced_impedance_1)),
             .groups = "drop") %>%
@@
 
-Population_data %>% 
-  pull(Population) %>% 
+population_sobi %>% 
+  pull(population) %>% 
   sum()
@@
 
-The level of service of each clinic is the number of doctors at the clinic, divided by the population that they serve:
+The level of service of each hub is the number of bicycle racks at the hub, divided by the population that they serve:
 ```{r}
-los <- long_matrix %>%
-  filter(impedance > 0) %>% # filter impedance == 0 to avoid divisions by zero
-  group_by(to) %>%
-  summarize(los = first(Doctors) / sum((Population * balanced_impedance_1)),
+sobi_los <- ttm_walk %>%
+  dplyr::filter(impedance_binary > 0) %>% # filter impedance == 0 to avoid divisions by zero
+  group_by(hub) %>%
+  summarize(sobi_los = first(racks) / sum((population * balanced_impedance_1)),
             .groups = "drop")
-los
+sobi_los
 ```
@@
 ```{r}
-long_matrix <- long_matrix %>%
-  left_join(los, by = "to")
+ttm_walk <- ttm_walk %>%
+  left_join(sobi_los, by = "hub")
 ```
@@
 
-To calculate accessibility, the level of service of each clinic needs to be distributed proportionally to the population centers. Check the 
+To calculate accessibility, the level of service of each hub needs to be distributed proportionally to the population units. 
 ```{r}
-accessibility <- long_matrix %>%
-  group_by(id) %>%
-  summarize(accessibility = sum(los * balanced_impedance_2),
+sobi_accessibility <- ttm_walk %>%
+  group_by(UID) %>%
+  summarize(accessibility = sum(sobi_los * balanced_impedance_2),
             .groups = "drop")
-accessibility
+sobi_accessibility
 ```
@@
 ```{r}
-accessibility %>% 
+sobi_accessibility %>% 
   pull(accessibility) %>% 
@@
 
-los %>% 
-  pull(los) %>%
+sobi_los %>% 
+  pull(sobi_los) %>%
   sum()
 ```
+[**END FOR SOBI ACCESSIBILITY**]
 

### 2021-01-25 examples/Toy-example-balanced-accessibility.Rmd
Reviewed Elise's initial calculations, made one change to correct some divisions by zero that happened due to a smaller threshold than the 30 min used to calculate the travel time matrix.

@@
 ```
+
+# Check the data
+
+Plot the interpolated population:
+```{r}
+ggplot(data = population_sobi) +
+  geom_sf(aes(color = population)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+Notice that the population units begin with UID 1 and ends with UID 11388:
+```{r}
+summary(population_sobi)
+```
+
+The travel time table, on the other hand, begins with population UID 15 and ends with UID 11305:
+```{r}
+summary(ttm_walk)
+```
+
+What explains this discrepancy? We can compare the population units that are in the population dataset but not in the travel time table:
+```{r}
+ggplot() +
+  geom_sf(data = sobi_service) +
+  geom_sf(data = population_sobi %>%
+            filter(!(UID %in% (distinct(ttm_walk, # filter observations with UIDs that are present in the travel time table
+                                        UID) %>% 
+                                 pull(UID)))))
+
+```
+
+The reason that some population units do not appear in the travel time table is that they are outside of the service area of SoBi and/or cannot reach SoBi hubs by walking.
+
 # Threshold
@@
 
+Check the value of the dist_threshold:
+```{r}
+dist_threshold
+```
+
+Our calculations assume that people do not go further than 24 min to reach a SoBi hub.
+
 Calculate impedance (Gaussian):
@@
 [**Note for Dr.
-Paez: why do the unit IDs start at 15 and not 1?**
+Paez: why do the unit IDs start at 15 and not 1?** ***Note for Elise: Some population units are outside of the service area and/or cannot reach a single sobi hub by walking in 30 min or less, so they disappeared from the travel time table when it was create; see above***].
+
+Sanity check: how many stations are within 24 or less from population UID 15?
+```{r}
+ttm_walk %>%
+  filter(UID == 15, travel_time <= 24) %>%
+  nrow()
+```
+
+**Important:** notice that some values of `sum_b1` are zeros!
+This means that some population units cannot reach a SoBi hub in at most 24 min.
+Since the balancing impedance is calculated by diving the impedance by `sum_b1`, this will lead to NaNs (divisions by zero).
+To avoid issues we will remove any population units that reach zero hubs:
+```{r}
+ttm_walk <- ttm_walk %>%
+  filter(sum_b1 > 0)
+```
 
@@
 Hub 1 serves 1502 interpolated population units, hub 2 serves 1393 population units, etc.
+
+Sanity check: how many population units are served by hub 1?
+```{r}
+ttm_walk %>%
+  filter(hub == 1, travel_time <= 24) %>%
+  nrow()
+```
+
+The values check. Next, are there any hubs that serve zero population units? (same issue with division by zero):
+```{r}
+ttm_walk$sum_b2 %>% 
+  summary()
+```
+
+No zeros or other strange values, we're good.
 
@@
 
-[**Note: not sure if this step is needed for SoBi?**] *Notice that the population of each center is sorted among the three clinics proportionally.
+[**Note: not sure if this step is needed for SoBi?** **Note for Elise: this is only an intermediate check in the analysis, it is not strictly necessary**] *Notice that the population of each center is sorted among the SoBi hubs proportionally.
 Thanks to this property, the sum of the population serviced equals the total population in the system:*
@@
 population_sobi %>% 
+  filter(UID %in% (distinct(ttm_walk,
+                             UID) %>%
+             pull(UID))) %>%
   pull(population) %>% 
@@
 ```
+
+Hub 1 has 0.01538 bicycles per person, hub 2 has 0.02446 bicycles per person, etc.
 
@@
 ```
+
+```{r}
+ggplot(data = population_sobi) +
+  geom_sf(aes(color = population))
+```
+
+Join the accessibility to the population units:
+```{r}
+population_sobi <- population_sobi %>%
+  left_join(sobi_accessibility,
+            by = "UID")
+```
+
+Plot accessibility:
+```{r}
+ggplot(data = population_sobi) +
+  geom_sf(aes(color = accessibility)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
 [**END FOR SOBI ACCESSIBILITY**]
 
+<!--
 # Complete  (or tl;dr version)
@@
 ```
+
+-->

### 2021-01-26 examples/Toy-example-balanced-accessibility.Rmd
Saved the files after rendering the html notebook
Remove notes for clarification

@@
 
-[**Note for Dr. Paez: do we want the binary and Gaussian functions? Or only one of them?**]
-
 Calculate impedance (binary):
@@
 
-The interpolated population at unit 15 can reach eight SoBi hubs, the population at unit 16 can reach six SoBi hubs, the population at unit 30 can reach zero hubs clinics, etc.
-[**Note for Dr.
-Paez: why do the unit IDs start at 15 and not 1?** ***Note for Elise: Some population units are outside of the service area and/or cannot reach a single sobi hub by walking in 30 min or less, so they disappeared from the travel time table when it was create; see above***].
+The interpolated population at unit 15 can reach eight SoBi hubs, the population at unit 16 can reach six SoBi hubs, the population at unit 30 can reach zero hubs clinics, etc..
 
-Sanity check: how many stations are within 24 or less from population UID 15?
+Sanity check: how many stations are within 24 minutes or less from population UID 15?
 ```{r}
@@
 ttm_walk <- ttm_walk %>%
-  filter(sum_b1 > 0)
+  dplyr::filter(sum_b1 > 0)
 ```

### 2021-02-04 examples/Toy-example-balanced-accessibility.Rmd
Added example of equity to the notebook.

--- 2021-01-25 examples/Toy-example-balanced-accessibility.Rmd
@@
 
-**Important:** notice that some values of `sum_b1` are zeros!
-This means that some population units cannot reach a SoBi hub in at most 24 min.
-Since the balancing impedance is calculated by diving the impedance by `sum_b1`, this will lead to NaNs (divisions by zero).
-To avoid issues we will remove any population units that reach zero hubs:
-```{r}
-ttm_walk <- ttm_walk %>%
-  filter(sum_b1 > 0)
-```
-
 Calculate the sum of the impedance by SoBi hubs for the balancing factors.
@@
 
-The values check. Next, are there any hubs that serve zero population units? (same issue with division by zero):
-```{r}
-ttm_walk$sum_b2 %>% 
-  summary()
-```
-
-No zeros or other strange values, we're good.
+The values check.
 
@@
 ```
+
+Check `sum_b1`:
+```{r}
+ttm_walk$sum_b1 %>% 
+  summary()
+```
+
+**Important:** notice that some values of `sum_b1` are zeros!
+This means that some population units cannot reach a SoBi hub in at most 24 min.
+Since the balancing impedance is calculated by diving the impedance by `sum_b1`, this will lead to NaNs (divisions by zero).
+To avoid issues we will remove any population units that reach zero hubs:
+```{r}
+ttm_walk <- ttm_walk %>%
+  filter(sum_b1 > 0)
+```
+
+Are there any hubs that serve zero population units? (same issue with division by zero):
+```{r}
+ttm_walk$sum_b2 %>% 
+  summary()
+```
+
+No zeros or other strange values, we're good.
 
@@
 ttm_walk %>% 
-  dplyr::filter(hub == "46") 
+  dplyr::filter(hub == "46") %>%
+  pull(balanced_impedance_2) %>%
+  sum()
 ```
+
+The sum should be 1.
 
@@
 
+<!--The text that follows was mostly copied from PLoS ONE paper-->
+
+## Disparity Analysis
+
+An advantage of the use of adjusted weights for proportional allocation of demand and level of service it that, after rectifying the inflation artifact, they make it is possible to conduct accessibility disparity analysis in a very intuitive way.
+
+For instance, an analyst interested in equity analysis could allocate the total level of service uniformly to every population unit.
+In other words, the total level of service (which equals the sum of accessibility over the system) can be divided by the number of population units in the system to return the Average Local Population Unit PPR.
+The resulting mean value, call it $L_i^e$ then would be assigned to the population units as their "equitable" share of the total level of service in the system.
+Next, the equitative distribution of the level of service in each population unit is substracted from the estimated accessibility to arrive at a disparity index.
+When the difference between these two quantities is positive, this would indicate that accessibility at that location exceeds its equitable share of level of service.
+On the other hand, when the difference is negative, accessibility at the location is below its equitable share of the level of service.
+
+This approach is reminiscent of the Spatial Access Ratio (SPAR) proposed by Wan et al.
+[-@Wan2012SPAR], which is calculated as the ratio between a population center's accessibility and the mean accessibility across all population centers.
+Wan et al.
+[-@Wan2012] calculate SPAR based on the results of their 3SFCA method, by rescaling the accessibility measures to reflect the percentage difference in each population center's accessibility relative to the mean.
+This measure is designed to overcome the sensitivity of existing FCA metrics to the impedance function.
+In contrast, the approach proposed here, enables more intuitive and interpretable results by preserving the system-wide population and level of service.
+In this way, a disparity index is useful to highlight the absolute difference in accessible provider-to-population ratios across population units.
+
+The disparity for SoBi is shown in the following map.
+The figure reveals the spatial distribution in disparity, with levels of access that are lower than the mean in the periphery of the service area, compared to levels of access that are greater than the mean in the higher-density and more connected urban center.
+
+```{r}
+population_sobi %>%
+  mutate(Lei = accessibility - sum(sobi_los$sobi_los)/nrow(population_sobi %>% drop_na(accessibility))) %>%
+  ggplot() +
+  geom_sf(aes(color = Lei)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
+```
+
+```{r accessibility_disparities, echo=FALSE, include=FALSE}
+#Disparities in this case is defined as the difference between the calculated accessibility (which is interpreted as the local provider-to-population ratio) and the local provider-to-population ratio if the level of service was distributed uniformly
+accessibility_disparities <- transmute(accessibility, DA, 
+                                      Binary = Binary_Adjusted - sum(los_binary_adjusted)/nrow(accessibility_binary_adjusted),
+                                      Stepwise = Stepwise_Adjusted- sum(los_stepwise_adjusted)/nrow(accessibility_stepwise_adjusted))
+```
+
+```{r join-geography, echo=FALSE, include=FALSE}
+#Join accessibility measures to geography.
+hamilton_da.sf_accessibility <- left_join(hamilton_da.sf, accessibility, by = "DA")
+hamilton_da.sf_comparison <- left_join(hamilton_da.sf, accessibility_comparison, by = "DA")
+hamilton_da.sf_disparities <- left_join(hamilton_da.sf, accessibility_disparities, by = "DA")
+```
+
 <!--

### 2021-02-10 examples/00-Data-Processing-Example.Rmd
Calculate disparity index and complete comparative analysis without ERI hubs

@@
 ```{r}
-interpolated_pop_clean %>% select(population) %>% summary()
+interpolated_pop_clean %>% dplyr::select(population) %>% summary()
 ```
@@
 
-## Routing
+Remove ERI hubs that were implemented in 2018 to increase accessibility for under-serviced and disadvantaged areas of the city.
+This reflects the original state of the system without the equity stations:
+```{r}
+sobi_hubs_original <- sobi_hubs %>%
+  dplyr::filter(NAME != "Mars at Wentworth - ERI01") %>%
+  dplyr::filter(NAME != "Sherman at Barton - ERI02") %>%
+  dplyr::filter(NAME != "Barton at Lottridge - ERI03") %>%
+  dplyr::filter(NAME != "Barton and Belview - ERI05") %>%
+  dplyr::filter(NAME != "Barton at Ottawa - ERI06") %>%
+  dplyr::filter(NAME != "Ottawa at Dunsmure - ERI07") %>%
+  dplyr::filter(NAME != "Maple at Rothsay - ERI08") %>%
+  dplyr::filter(NAME != "King at Dunsmure - ERI09") %>%
+  dplyr::filter(NAME != "Dunsmure at Sherman - ERI10") %>%
+  dplyr::filter(NAME != "Gage at Cannon - ERI11") %>%
+  dplyr::filter(NAME != "Belview at Cannon - ERI12") %>%
+  dplyr::filter(NAME != "Westinghouse at Barton - ERI13")
+```
+
+Plot again:
+```{r}
+ggplot() + 
+  geom_sf(data = interpolated_pop_clean,
+          aes(color = population)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  geom_sf(data = sobi_hubs_original,
+          aes(size = RACKS_AMOU)) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  scale_size(range = c(1, 4))
+```
+After removing the twelve ERI hubs, there is much less coverage in the east end of the service area.
+Again, the service area is the area where a cyclist can use a SoBi bike.
+It makes sense then that ERI would want to fill this gap in coverage because individuals who live in the east end of the service area could theoretically use the bike share system but may not live near a hub.
+For this reason, twelve hubs were added in 2018 to expand the system.
+
+## Routing for entire system
 
@@
     rename(lon = X, lat = Y, id = UID) %>%
-    select(id, lon, lat)
+    dplyr::select(id, lon, lat)
 
@@
     rename(lon = X, lat = Y, id = OBJECTID) %>%
-    select(id, lon, lat)
+    dplyr::select(id, lon, lat)
 ```
@@
 
+## Routing for original system without ERI hubs
+
+Build the graph:
+```{r build-graph-original, include = FALSE}
+r5_sobi_original <- setup_r5(data_path = r5_path, verbose = FALSE)
+```
+
+Prepare Input Data for `r5r`. The origins are the coordinates of the population cells and the destinations the coordinates of the SoBi hubs (minus the ERI hubs in the east end of the service area):
+```{r prepare-inputs-original}
+# save origin centroids in format expected by R5R (id, lon, lat)
+origins_io <- data.frame(UID = interpolated_pop_clean$UID, 
+              interpolated_pop_clean %>%
+                st_transform(crs = 4326) %>%
+                st_coordinates()) %>%
+    rename(lon = X, lat = Y, id = UID) %>%
+    dplyr::select(id, lon, lat)
+
+# now SoBi hubs
+destinations_jo <- data.frame(OBJECTID = sobi_hubs_original$OBJECTID, 
+              sobi_hubs_original %>%
+                st_transform(crs = 4326) %>%
+                st_coordinates()) %>%
+    rename(lon = X, lat = Y, id = OBJECTID) %>%
+    dplyr::select(id, lon, lat)
+```
+
+Calculate OD Matrix for Walking:
+```{r calculate walking od matrix original system, include=FALSE}
+# set up batching according to how many origin rows to process at one time
+chunksize_o = 2000 # larger chunks for walking will give enough origins in each chunk to allow multiprocessing to spin up with R5R
+num_chunks_o = ceiling(nrow(origins_io)/chunksize_o)
+
+# create origin-destination pairs
+origins_chunks_o <- as.disk.frame(origins_io,
+                          outdir = "./df/origins_io",
+                          nchunks = num_chunks_o,
+                          overwrite = TRUE)
+
+start.time_o <- Sys.time()
+pb_o <- txtProgressBar(0, num_chunks_o, style = 3)
+
+for (io in 1:num_chunks_o){
+  origins_io_chunk <- get_chunk(origins_chunks_o, io)
+  ttm_chunk_o <- travel_time_matrix(r5_sobi_original,
+                          origins = origins_io_chunk,
+                          destinations = destinations_jo,
+                          mode = c("WALK"),
+                          departure_datetime = as.POSIXct(strptime("2020-03-02 08:00:00", "%Y-%m-%d %H:%M:%S", tz = "EST5EDT")),
+                          max_walk_dist = 10000, # metres
+                          max_trip_duration = 30)
+
+  # export output as disk.frame
+  ifelse(io == 1, output_o_df <- as.disk.frame(ttm_chunk_o,
+                                            nchunks = 1,
+                                            outdir = "./df/output_ttm_walk_o",
+                                            compress = 50,
+                                            overwrite = TRUE),
+         add_chunk(output_o_df, ttm_chunk_o, chunk_id = io))
+  setTxtProgressBar(pb_o, io)
+}
+end.time_o <- Sys.time()
+ttm_walk_time_o <- end.time - start.time
+ttm_walk_time_o
+```
+
+## Extract travel time matrix for original system
+
+```{r load od matrix for walk and sobi, include=FALSE}
+# connect to the walking travel time matrix disk frame
+ttm_walk_o.disk.frame <- disk.frame("./df/output_ttm_walk_o")
+#opportunities_j <- as.data.frame(nyc_cb_point) %>% select(GEOID10, total_emp) %>% rename(toId = GEOID10, o_j = total_emp)
+```
+
+Convert disk.frame to data frame:
+```{r}
+ttm_walk_o <- as.data.frame(ttm_walk_o.disk.frame) %>%
+  transmute(UID = as.numeric(fromId), OBJECTID = as.numeric(toId), travel_time)
+```
+
 ## Save data to disk
@@
 save(sobi_hubs, file = "sobi_hubs.RData")
+save(sobi_hubs_original, file = "sobi_hubs_original.RData")
 save(sobi_service, file = "sobi_service.RData")
 save(ttm_walk, file = "ttm_walk.RData")
+save(ttm_walk_o, file = "ttm_walk_original.RData")
 ```

### 2021-02-10 examples/Toy-example-balanced-accessibility.Rmd
Calculate disparity index and complete comparative analysis without ERI hubs

@@
 ---
+
+Clear workspace:
+```{r}
+rm(list = ls())
+```
 
@@
 load("sobi_hubs.RData") #location of SoBi hubs, minus Van Wagner's beach
+load("sobi_hubs_original.RData") #SoBi hubs, minus ERI hubs
 load("sobi_service.RData") #core service area for SoBi
 load("ttm_walk.RData") #travel time matrix for interpolated population in SoBi area
+load("ttm_walk_original.RData") #travel time matrix for system without ERI hubs
 ```
@@
 summary(ttm_walk) #time is in minutes
+```
+
+Make a copy of the object for original system:
+```{r}
+population_sobi_original <- population_sobi
 ```
@@
 
-The reason that some population units do not appear in the travel time table is that they are outside of the service area of SoBi and/or cannot reach SoBi hubs by walking.
+The reason that some population units do not appear in the travel time table is because they are outside of the service area of SoBi and/or cannot reach SoBi hubs by walking.
 
@@
 
-The interpolated population at unit 15 can reach eight SoBi hubs, the population at unit 16 can reach six SoBi hubs, the population at unit 30 can reach zero hubs clinics, etc..
+The interpolated population at unit 15 can reach eight SoBi hubs in 24 minutes, the population at unit 16 can reach six SoBi hubs, the population at unit 30 can reach zero hubs clinics, etc..
 
@@
 
-No zeros or other strange values, we're good.
+No zeros or other strange values.
 
@@
 
-[**Note: not sure if this step is needed for SoBi?** **Note for Elise: this is only an intermediate check in the analysis, it is not strictly necessary**] *Notice that the population of each center is sorted among the SoBi hubs proportionally.
-Thanks to this property, the sum of the population serviced equals the total population in the system:*
+[**Note: not sure if this step is needed for SoBi?** **Note for Elise: this is only an intermediate check in the analysis, it is not strictly necessary**] Notice that the population of each center is sorted among the SoBi hubs proportionally.
+Thanks to this property, the sum of the population serviced equals the total population in the system:
 ```{r}
@@
 
-[**END FOR SOBI ACCESSIBILITY**]
+# Repeat above for the original SoBi system (no ERI hubs)
+
+Calculate quantiles according to distance. This is done to select a threshold:
+```{r}
+quantiles_d_o <- quantile(as.vector(ttm_walk_o$travel_time), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))
+
+dist_threshold_o <- quantile(as.vector(ttm_walk_o$travel_time), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))[[5]]
+```
+
+The distance threshold is equivalent to the _kernel bandwidth_ or _catchment distance_.
+
+# Set up matrix
+
+Add population information:
+```{r}
+ttm_walk_o <- left_join(ttm_walk_o, 
+                      population_sobi_original,
+                      by = "UID")
+```
+
+Drop the geometry:
+```{r}
+ttm_walk_o <- ttm_walk_o %>% dplyr::select(-geometry)
+```
+
+Add number of bicycle racks per SoBi hub as a measure of the level of supply:
+```{r}
+ttm_walk_o <- left_join(ttm_walk_o, sobi_hubs_original %>% dplyr::select(OBJECTID, RACKS_AMOU), by=c("OBJECTID" = "OBJECTID"))
+ttm_walk_o <- ttm_walk_o %>% dplyr::select(-geometry)
+head(ttm_walk_o)
+```
+
+Rename columns:
+```{r}
+names(ttm_walk_o) <- c('UID', 'hub', 'travel_time', 'population', 'racks')
+```
+
+Calculate impedance (binary):
+```{r eval=TRUE}
+# Run only if binary function desired
+# Change chunk option eval=TRUE to FALSE if impedance is Gaussian function
+ttm_walk_o <- ttm_walk_o %>%
+  mutate(impedance_binary = ifelse(travel_time <= dist_threshold_o, 1, 0))
+```
+
+Check the value of the dist_threshold:
+```{r}
+dist_threshold_o
+```
+
+Our calculations assume that people do not go further than 24 min to reach a SoBi hub.
+
+Calculate impedance (Gaussian):
+```{r eval=FALSE}
+# Run only if Gaussian function desired
+# Change chunk option eval=FALSE to TRUE if impedance is Gaussian function
+ttm_walk_o <- ttm_walk_o %>%
+  mutate(impedance_gaussian = case_when(travel_time <= quantiles_d_o[1] ~ 0.945,
+                                        (travel_time > quantiles_d_o[1]) & (travel_time <= quantiles_d_o[3]) ~ 0.600,
+                                        (travel_time > quantiles_d_o[3]) & (travel_time <= quantiles_d_o[5]) ~ 0.242,
+                                        travel_time > quantiles_d_o[5] ~ 0))
+```
+
+Calculate the sum of the impedance by population for the balancing factors.
+In the binary case, this should be the same as the number of SoBi hubs that each interpolated population unit can reach; when the impedance is not binary, this is the sum of all impedance values:
+```{r}
+sum_b1 <- ttm_walk_o %>%
+  group_by(UID) %>%
+  summarize(sum_b1 = sum(impedance_binary),
+            .groups = "drop")
+sum_b1
+```
+
+The interpolated population at unit 15 can reach eight SoBi hubs in 24 minutes, the population at unit 16 can reach six SoBi hubs, the population at unit 30 can reach zero hubs clinics, etc..
+
+Sanity check: how many stations are within 24 minutes or less from population UID 15?
+```{r}
+ttm_walk_o %>%
+  filter(UID == 15, travel_time <= 24) %>%
+  nrow()
+```
+
+Calculate the sum of the impedance by SoBi hubs for the balancing factors.
+In the binary case, this should be the same as the number of interpolated population unit that each SoBi hub serves; when the impedance is not binary, this is the sum of all impedance values:
+```{r}
+sum_b2 <- ttm_walk_o %>%
+  group_by(hub) %>%
+  summarize(sum_b2 = sum(impedance_binary))
+sum_b2
+```
+
+Hub 1 serves 1502 interpolated population units, hub 2 serves 1393 population units, etc.
+
+Sanity check: how many population units are served by hub 1?
+```{r}
+ttm_walk_o %>%
+  filter(hub == 1, travel_time <= 24) %>%
+  nrow()
+```
+
+The values check.
+
+Next, join the sum of the impedance to the table to calculate the balanced impedance values:
+```{r}
+ttm_walk_o <- ttm_walk_o %>%
+  left_join(sum_b1, by = "UID") %>%
+  left_join(sum_b2, by = "hub")
+```
+
+Check `sum_b1`:
+```{r}
+ttm_walk_o$sum_b1 %>% 
+  summary()
+```
+
+**Important:** notice that some values of `sum_b1` are zeros!
+This means that some population units cannot reach a SoBi hub in at most 24 min.
+Since the balancing impedance is calculated by diving the impedance by `sum_b1`, this will lead to NaNs (divisions by zero).
+To avoid issues we will remove any population units that reach zero hubs:
+```{r}
+ttm_walk_o <- ttm_walk_o %>%
+  filter(sum_b1 > 0)
+```
+
+Are there any hubs that serve zero population units? (same issue with division by zero):
+```{r}
+ttm_walk_o$sum_b2 %>% 
+  summary()
+```
+
+No zeros or other strange values.
+
+Calculate balanced impedance values:
+```{r}
+ttm_walk_o <- ttm_walk_o %>%
+  mutate(balanced_impedance_1 = impedance_binary/sum_b1,
+         balanced_impedance_2 = impedance_binary/sum_b2)
+```
+
+# Calculate level of service
+
+Check the balanced_impedance factors for the population units:
+```{r}
+ttm_walk_o %>% 
+  dplyr::filter(hub == "46") %>%
+  pull(balanced_impedance_2) %>%
+  sum()
+```
+
+The sum should be 1.
+
+The level of service of each hub is the number of bicycle racks at the hub, divided by the population that they serve:
+```{r}
+sobi_los_original <- ttm_walk_o %>%
+  dplyr::filter(impedance_binary > 0) %>% # filter impedance == 0 to avoid divisions by zero
+  group_by(hub) %>%
+  summarize(sobi_los_original = first(racks) / sum((population * balanced_impedance_1)),
+            .groups = "drop")
+sobi_los_original
+```
+
+Hub 1 has 0.01535 bicycles per person, hub 2 has 0.02435 bicycles per person, etc.
+
+Join the level of service to the table:
+```{r}
+ttm_walk_o <- ttm_walk_o %>%
+  left_join(sobi_los_original, by = "hub")
+```
+
+# Calculate accessibility
+
+To calculate accessibility, the level of service of each hub needs to be distributed proportionally to the population units. 
+```{r}
+sobi_accessibility_original <- ttm_walk_o %>%
+  group_by(UID) %>%
+  summarize(accessibility_o = sum(sobi_los_original * balanced_impedance_2),
+            .groups = "drop")
+sobi_accessibility_original
+```
+
+Check that the sum of the accessibility is equal to the available level of service:
+```{r}
+sobi_accessibility_original %>% 
+  pull(accessibility_o) %>% 
+  sum()
+
+sobi_los_original %>% 
+  pull(sobi_los_original) %>%
+  sum()
+```
+
+Compared to the analysis that includes the ERI hubs, the available level of service is lower.
+
+```{r}
+ggplot(data = population_sobi_original) +
+  geom_sf(aes(color = population))
+```
+
+Join the accessibility to the population units:
+```{r}
+population_sobi_original <- population_sobi_original %>% left_join(sobi_accessibility_original, by = "UID")
+```
+
+Plot accessibility:
+```{r}
+ggplot(data = population_sobi_original) +
+  geom_sf(aes(color = accessibility_o)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+## Disparity Analysis
 
 <!--The text that follows was mostly copied from PLoS ONE paper-->
-
-## Disparity Analysis
 
@@
 The resulting mean value, call it $L_i^e$ then would be assigned to the population units as their "equitable" share of the total level of service in the system.
-Next, the equitative distribution of the level of service in each population unit is substracted from the estimated accessibility to arrive at a disparity index.
+Next, the equitable distribution of the level of service in each population unit is subtracted from the estimated accessibility to arrive at a disparity index.
 When the difference between these two quantities is positive, this would indicate that accessibility at that location exceeds its equitable share of level of service.
@@
 Wan et al.
-[-@Wan2012] calculate SPAR based on the results of their 3SFCA method, by rescaling the accessibility measures to reflect the percentage difference in each population center's accessibility relative to the mean.
+[-@Wan2012] calculate SPAR based on the results of their 3SFCA method, by re-scaling the accessibility measures to reflect the percentage difference in each population center's accessibility relative to the mean.
 This measure is designed to overcome the sensitivity of existing FCA metrics to the impedance function.
@@
 The figure reveals the spatial distribution in disparity, with levels of access that are lower than the mean in the periphery of the service area, compared to levels of access that are greater than the mean in the higher-density and more connected urban center.
+It is worth noting that levels of access that are greater than the mean are predominantly concentrated around the University in west Hamilton.
+Areas that have less than their equitable share of the level of service are some of the most socioeconomically disadvantaged areas in the city.
+Our analysis builds upon a previous and recent study [@hosfordWhoArePublic2018] which found that disadvantaged areas in Hamilton are better served by the SoBi system, compared to other Canadian cities [i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau] where advantaged areas have greater access.
+The authors [@hosfordWhoArePublic2018] note "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas.
+Thus, relative to other Canadian programs, Hamilton’s bicycle share program could more easily be established in lower income areas while still being near high population density areas".
+This is an artifact of the 2001 amalgamation of Hamilton with four surrounding communities.	The urban core, which is Hamilton proper, remains predominantly disadvantaged compared to the 4 amalgamated and less densely populated areas (e.g., Dundas, Ancaster, Stoney Creek, Waterdown) that surround the city.
+Interestingly, Dundas is an advantaged area but also has lower accessibility that is less than their equitable share.
+However, by calculating the disparity index, we have shown that more advantaged areas within the SoBi service area have higher levels of service and exceed its equitable share of service.
+# Disparity of system with ERI hubs
 
@@
 
-```{r accessibility_disparities, echo=FALSE, include=FALSE}
-#Disparities in this case is defined as the difference between the calculated accessibility (which is interpreted as the local provider-to-population ratio) and the local provider-to-population ratio if the level of service was distributed uniformly
-accessibility_disparities <- transmute(accessibility, DA, 
-                                      Binary = Binary_Adjusted - sum(los_binary_adjusted)/nrow(accessibility_binary_adjusted),
-                                      Stepwise = Stepwise_Adjusted- sum(los_stepwise_adjusted)/nrow(accessibility_stepwise_adjusted))
+Drop population units that have `NA` for accessibility because they are outside of the service area of SoBi and/or cannot reach SoBi hubs by walking.
+```{r}
+population_sobi <- population_sobi %>% drop_na()
 ```
 
-```{r join-geography, echo=FALSE, include=FALSE}
-#Join accessibility measures to geography.
-hamilton_da.sf_accessibility <- left_join(hamilton_da.sf, accessibility, by = "DA")
-hamilton_da.sf_comparison <- left_join(hamilton_da.sf, accessibility_comparison, by = "DA")
-hamilton_da.sf_disparities <- left_join(hamilton_da.sf, accessibility_disparities, by = "DA")
+This data frame now has the same number of rows as the data frame on `sobi_accessibility`.
+
+Calculate the sum of accessibility over the system:
+```{r}
+sum(population_sobi$accessibility)
+```
+Divide by the number of population units in the system to return the average local population unit provider-to-population ratio:
+```{r}
+population_sobi <- population_sobi %>% mutate(PPR = sum(accessibility/nrow(population_sobi)))
 ```
 
-<!--
-# Complete  (or tl;dr version)
+The "equitable" share of the total level of service in the system is 0.0002025645 per population unit.
 
-The full algorithm can be written in this way.
+Calculate disparities per population unit:
+```{r accessibility_disparities, echo=FALSE, include=FALSE}
+# Disparity in this case is defined as the difference between the calculated accessibility (which is interpreted as the local provider-to-population ratio) and the local provider-to-population ratio if the level of service was distributed uniformly
 
-Obtain the matrix in long form (this is line 85 in this file):
-```{r}
-long_matrix <- left_join(distance_matrix_long, Clinics_data, by=c("to" = "id"))
+population_sobi <- population_sobi %>% mutate(disparity = accessibility - PPR)
 ```
 
-Calculate the accessibility and level of service:
+Check plot:
 ```{r}
-# Calculate impedance (binary)
-# long_matrix <- long_matrix %>%
-#   mutate(distance = drop_units(distance),
-#          impedance = ifelse(distance <= dist_threshold, 1, 0))
-
-# Calculate impedance (Gaussian)
-long_matrix <- long_matrix %>%
-  mutate(distance = drop_units(distance),
-         impedance = case_when(distance <= quantiles_d[1] ~ 0.945,
-                               (distance > quantiles_d[1]) & (distance <= quantiles_d[3]) ~ 0.600,
-                               (distance > quantiles_d[3]) & (distance <= quantiles_d[5]) ~ 0.242,
-                               distance > quantiles_d[5] ~ 0))
-
-# Calculate first set of balancing factors
-sum_b1 <- long_matrix %>%
-  group_by(id) %>%
-  summarize(sum_b1 = sum(impedance),
-            .groups = "drop")
-
-# Calculate second set of balancing factors
-sum_b2 <- long_matrix %>%
-  group_by(to) %>%
-  summarize(sum_b2 = sum(impedance),
-            .groups = "drop")
-
-# Join the sum of the impedance to the table to calculate the balanced impedance values:
-long_matrix <- long_matrix %>%
-  left_join(sum_b1, by = "id") %>%
-  left_join(sum_b2, by = "to")
-
-# Calculate the balanced impedance
-long_matrix <- long_matrix %>%
-  mutate(balanced_impedance_1 = impedance/sum_b1,
-         balanced_impedance_2 = impedance/sum_b2)
-
-# Calculate the level of service
-los <- long_matrix %>%
-  filter(impedance > 0) %>% # filter impedance == 0 to avoid divisions by zero
-  group_by(to) %>%
-  summarize(los = first(Doctors) / sum((Population * balanced_impedance_1)),
-            .groups = "drop")
-
-# Join the level of service to the table:
-long_matrix <- long_matrix %>%
-  left_join(los, by = "to")
-
-# Calculate accessibility
-accessibility <- long_matrix %>%
-  group_by(id) %>%
-  summarize(accessibility = sum(los * balanced_impedance_2),
-            .groups = "drop")
+population_sobi %>%
+  ggplot() +
+  geom_sf(aes(color = disparity)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
 ```
 
-This is the accessibility:
+# Disparity of original system without ERI hubs
+
 ```{r}
-accessibility
+population_sobi_original %>%
+  mutate(Lei = accessibility_o - sum(sobi_los_original$sobi_los_original)/nrow(population_sobi_original %>% drop_na(accessibility_o))) %>%
+  ggplot() +
+  geom_sf(aes(color = Lei)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
 ```
 
-This is the level of service:
+Drop population units that have `NA` for accessibility because they are outside of the service area of SoBi and/or cannot reach SoBi hubs by walking.
 ```{r}
-los
+population_sobi_original <- population_sobi_original %>% drop_na()
 ```
 
--->
+This data frame now has the same number of rows as the data frame on `sobi_accessibility_original`.
+
+Calculate the sum of accessibility over the system:
+```{r}
+sum(population_sobi_original$accessibility_o)
+```
+
+Divide by the number of population units in the system to return the average local population unit provider-to-population ratio:
+```{r}
+population_sobi_original <- population_sobi_original %>% mutate(PPR_o = sum(accessibility_o/nrow(population_sobi_original)))
+```
+
+The "equitable" share of the total level of service in the system is 0.0001927507 per population unit.
+
+Calculate disparities per population unit:
+```{r accessibility_disparities_original, echo=FALSE, include=FALSE}
+# Disparity in this case is defined as the difference between the calculated accessibility (which is interpreted as the local provider-to-population ratio) and the local provider-to-population ratio if the level of service was distributed uniformly
+
+population_sobi_original <- population_sobi_original %>% mutate(disparity_original = accessibility_o - PPR_o)
+```
+
+Check plot:
+```{r}
+population_sobi_original %>%
+  ggplot() +
+  geom_sf(aes(color = disparity_original)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
+```
+
+Join in one data frame:
+```{r}
+sobi_equity <- population_sobi_original %>% st_join(population_sobi, by = "UID")
+```
+
+Drop duplicate population and UID columns:
+```{r}
+sobi_equity <- sobi_equity[ -c(6, 7) ]
+```
+
+Rename columns:
+```{r}
+names(sobi_equity) <- c('population', 'UID', 'accessibility_o', 'PPR_o', 'disparity_o', 'accessibility', 'PPR', 'disparity', 'geometry')
+```
+
+Calculate difference in disparity between system with and without ERI hubs:
+```{r}
+sobi_equity <- sobi_equity %>% mutate(disparity_change = disparity - disparity_o)
+```
+
+Check plot:
+```{r}
+sobi_equity %>%
+  ggplot() +
+  geom_sf(aes(color = disparity_change)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
+```
+
+If you remove the ERI hubs from the east end of the service area, it is evident that those communities are impacted, whereas the rest of the city is not.
+
+Save data file:
+```{r}
+save(sobi_equity, file = "sobi_equity.RData")
+```

### 2021-02-11 examples/Toy-example-balanced-accessibility.Rmd
Reviewed new disparity analysis and made small edits
Added an example of reaggregation of the accessibility indicator and an example of a bivariate choropleth map

@@
 However, by calculating the disparity index, we have shown that more advantaged areas within the SoBi service area have higher levels of service and exceed its equitable share of service.
-# Disparity of system with ERI hubs
+
+## Disparity of system with ERI hubs
 
@@
 
-# Disparity of original system without ERI hubs
+## Disparity of original system without ERI hubs
 
@@
 ```
+
+## Reaggregating accessibility
+
+One advantage of the Balanced FCA approach is that accessibility can be reaggregated while preserving the population and level of service.
+Here, we reaggregate the population to DAs.
+This allows us to then join the DA-level accessibility to DA-level census data for further analysis.
+
+Read the DA-level data:
+```{r}
+load("hamilton_da_2016.RData")
+```
+
+Plot DAs:
+```{r}
+ggplot(data = hamilton_da_2016) +
+  geom_sf()
+```
+
+Project the Hamilton DA zones to the same projection of the SoBi Service Area:
+```{r}
+hamilton_da_2016 <- st_transform(hamilton_da_2016, crs = st_crs(sobi_service))
+```
+
+Bounding box of DAs in serviced areas:
+```{r}
+bounding_box <- st_bbox(sobi_service %>%
+                          st_buffer(500))
+```
+
+Extract DAs in service area:
+```{r}
+hamilton_da_sobi <- hamilton_da_2016[st_as_sfc(bounding_box),]
+```
+
+Plot accessibility and extracted DAs:
+```{r}
+ggplot() +
+  geom_sf(data = population_sobi_original,
+          aes(color = accessibility_o)) +
+  geom_sf(data = hamilton_da_sobi,
+          fill = NA) + 
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+Spatial join:
+```{r}
+accessibility_da_original <- population_sobi_original %>%
+  st_join(hamilton_da_sobi %>% dplyr::select(DA))
+```
+
+Aggregate population and accessibility by DA:
+```{r}
+accessibility_da_original <- accessibility_da_original %>%
+  group_by(DA) %>%
+  summarize(population = sum(population),
+            accessibility_o = sum(accessibility_o),
+            .groups = "drop")
+
+```
+
+Check that accessibility (LOS) is preserved:
+```{r}
+sum(accessibility_da_original$accessibility_o)
+sum(population_sobi_original$accessibility_o)
+```
+
+Check that the population is preserved:
+```{r}
+sum(accessibility_da_original$population)
+sum(population_sobi_original$population)
+```
+
+Ok!
+
+Add the geometry of DAs to the table `accessibility_da` and convert into simple features:
+```{r}
+accessibility_da_original <- accessibility_da_original %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_sobi %>% 
+              dplyr::select(DA, geometry),
+            by = "DA") %>%
+  st_as_sf()
+```
+
+Plot DA-level accessibility:
+```{r}
+ggplot() +
+  geom_sf(data = accessibility_da_original,
+          aes(fill = accessibility_o),
+          color = "black") +
+  scale_fill_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+Same thing for the other accessibility estimates. Spatial join:
+```{r}
+accessibility_da <- population_sobi %>%
+  st_join(hamilton_da_sobi %>% dplyr::select(DA))
+```
+
+Aggregate population and accessibility by DA:
+```{r}
+accessibility_da <- accessibility_da %>%
+  group_by(DA) %>%
+  summarize(population = sum(population),
+            accessibility = sum(accessibility),
+            .groups = "drop")
+
+```
+
+Check that accessibility (LOS) is preserved:
+```{r}
+sum(accessibility_da$accessibility)
+sum(population_sobi$accessibility)
+```
+
+Check that the population is preserved:
+```{r}
+sum(accessibility_da$population)
+sum(population_sobi$population)
+```
+
+Ok!
+
+Add the geometry of DAs to the table `accessibility_da` and convert into simple features:
+```{r}
+accessibility_da <- accessibility_da %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_sobi %>% 
+              dplyr::select(DA, geometry),
+            by = "DA") %>%
+  st_as_sf()
+```
+
+Plot DA-level accessibility:
+```{r}
+ggplot() +
+  geom_sf(data = accessibility_da,
+          aes(fill = accessibility),
+          color = "black") +
+  scale_fill_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+Join two accessibility estimates:
+```{r}
+accessibility_da <- accessibility_da %>%
+  left_join(accessibility_da_original %>% 
+              st_drop_geometry() %>%
+              rename(population_o = population),
+              dplyr::select(DA, accessibility_o, population_o),
+            by = "DA") %>%
+  mutate(accessibility_o = replace_na(accessibility_o, 0), # The population serviced grows with the new stations, and as a result some original population is zero
+         population_o = replace_na(population_o, 0)) # Same for accessibility
+
+```
+
+Calculate the difference in accessibility with and without ERI hubs:
+```{r}
+accessibility_da <- accessibility_da %>%
+  mutate(R_a = accessibility - accessibility_o)
+```
+
+Plot accessibility differences between with and without ERI hubs:
+```{r}
+ggplot() +
+  geom_sf(data = accessibility_da,
+          aes(fill = R_a),
+          color = "black") +
+  scale_fill_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+Retrieve census data using cancensus package (requires API key!)
+```{r}
+library(cancensus)
+```
+
+**NOTE TO ELISE**
+
+You can get an API to use `cancensus` from [CensusMapper](https://censusmapper.ca/). Once you have an API key, you can store it in your system this way:
+```{r eval=FALSE}
+set_api_key("your-key", install = TRUE)
+```
+
+Once you have your API you can use `cancensus` like in this example.
+
+Use `cancensus` to check and download census data. Check the regions:
+```{r}
+list_census_regions('CA16') %>% 
+  filter(level == "CMA", name %in% c("Hamilton"))
+```
+
+We need PR_UID 35.
+
+List census data sets:
+```{r}
+list_census_datasets()
+```
+
+We need CA16.
+
+```{r}
+find_census_vectors("median household income", 
+                    dataset = "CA16", 
+                    type = "total", 
+                    query_type = "keyword", 
+                    interactive = FALSE)
+```
+
+We need v_CA16_2397.
+
+Retrieve data:
+```{r eval=FALSE}
+mhi <- get_census(dataset='CA16', 
+                          regions=list(CMA=c("35537")),
+                          vectors=c("v_CA16_2397"),
+                          level='DA', 
+                          use_cache = FALSE)
+```
+
+Join median household income to accessibility by DA:
+```{r}
+accessibility_da <- accessibility_da %>%
+  left_join(mhi %>%
+              transmute(DA = GeoUID,
+                        median_household_income = `v_CA16_2397: Median total income of households in 2015 ($)`),
+            by = "DA")
+```
+
+Summary of median household income:
+```{r}
+summary(accessibility_da$median_household_income)
+```
+
+There is a missing value (likely a censored unit).
+
+Bivariate choropleth maps of accessibility and income:
+```{r}
+library(biscale)
+library(cowplot)
+```
+
+Create a class for bivariate plotting (drop DA with missing income data):
+```{r}
+data <- bi_class(accessibility_da %>%
+                   drop_na(),
+                 x = accessibility_o, 
+                 y = median_household_income, 
+                 style = "quantile", 
+                 dim = 3) %>%
+  st_as_sf()
+```
+
+Plot:
+```{r}
+ map <- ggplot() +
+  geom_sf(data = data, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", dim = 3) +
+  bi_theme()
+
+map
+```
+
+Create legend:
+```{r}
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+```
+
+Finalize plot:
+```{r}
+ cowplot::ggdraw() +
+  cowplot::draw_plot(map, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```

### 2021-02-11 examples/Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Manuscript

---
title: Examining spatial equity of a public bicycle share program using Floating Catchment Area methods 
author:
  - name: Antonio Paez
    email: paez@mcmaster.ca
    affiliation: Some School
    footnote: 1
  - name: Elise Desjardins
    email: desjae@mcmaster.ca
    affiliation: Some School
    footnote: 2
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliation: Another University
    footnote: 2
address:
  - code: McMaster University
    address: School of Earth, Environment & Society, 1280 Main Street West, Hamilton, ON
  - code: University of Toronto Scarborough
    address: Department, Street, City, State, Zip
footnote:
  - code: 1
    text: "Corresponding Author"
  - code: 2
    text: "Equal contribution"
abstract: |
  This is the abstract.

  It consists of two paragraphs.

journal: "Transportation Research Part D: Transport and Environment"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

```{r echo=FALSE}
library(dplyr)
library(disk.frame)
library(pycno)
library(readr)
library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
library(sf)
library(tidyverse)
library(data.table)
library(gdistance)
library(kableExtra)
library(raster)
library(rgdal)
library(sf)
library(units)
```

```{r echo=FALSE}
load("hamilton_cma.RData")
load("hamilton_da_2016.RData")
load("population_sobi.RData") #interpolated population in SoBi area
load("sobi_hubs.RData") #location of SoBi hubs, minus Van Wagner's beach
load("sobi_hubs_original.RData") #SoBi hubs, minus ERI hubs
load("sobi_service.RData") #core service area for SoBi
load("ttm_walk.RData") #travel time matrix for interpolated population in SoBi area
load("ttm_walk_original.RData") #travel time matrix for system without ERI hubs
```

Introduction
===================

Public bicycle share programs (PBSP)

This paper uses the city of Hamilton, located in Ontario, Canada, as a case study. The city launched the program, called SoBi Hamilton, in March 2015 with 115 stations and 750 bicycles [@Hamilton2020]. The core service area spans 35 km^2 of the city and approximately 180,000 people live there [see Figure 1]. The program was enthusiastically welcomed in the city - within three weeks of launching, 10,000 trips had been made. In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton. To increase spatial equity, an additional 75 bicycles and 15 stations were added to the system, which expanded it to more disadvantaged areas in the city [see Figure 2]. The program also offers subsidized memberships to individuals who identify as low income. As of June 2020, the bike share program has 900 bikes and over 130 stations, and over 26,000 active memberships [@Hamilton2020].

```{r echo=FALSE}
sobi_service <- sobi_service %>%
  dplyr::filter(AREA_NAME == "Core Service Area")
```

```{r echo=FALSE}
bounding_box <- st_bbox(sobi_service %>%
                          st_buffer(500))
```

```{r echo=FALSE}
ggplot() + 
  geom_sf(data = hamilton_da_2016,
          fill = NA,
          color = "gray") + 
  geom_sf(data = sobi_service,
          fill = NA,
          colour = "blue")
```

Methods
===================

Results
===================

Discussion
===================

Conclusion
===================

References {#references .unnumbered}
===================

### 2021-03-12 examples/00-Data-Processing-Example.Rmd
Update data preprocessing

@@
 title: "00-Data-Processing-Example"
-output: html_notebook
+output:
+  pdf_document: default
+  html_notebook: default
 ---
@@
 save(ttm_walk_o, file = "ttm_walk_original.RData")
+save(hamilton_da_sobi, file = "hamilton_da_sobi.RData")
+save(interpolated_pop_clean, file = "interpolated_pop_clean.RData")
 ```

### 2021-03-12 examples/00-Data-Processing-Example.tex
Update data preprocessing

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={00-Data-Processing-Example},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\title{00-Data-Processing-Example}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

In this notebook we create some examples of processing the data for the
analysis of accessibility to Hamilton Bike share docking stations.
Particularly, we're interested in spatial interpolation of the
population for small areas.

Data used in this notebook was retrieved from two sources: Open Hamilton
and the 2016 Canadian Census. Open Hamilton is an online repository of
data from the City of Hamilton. The following datasets were retrieved:

\hypertarget{provenance-of-data}{%
\subsection{Provenance of data}\label{provenance-of-data}}

\hypertarget{sobi-hub}{%
\subsubsection{SoBi Hub}\label{sobi-hub}}

The dataset was last updated on \emph{November 9, 2020} and was
downloaded for this project on \emph{November 10, 2020}. This dataset
contains the following attributes that are of interest to this research:
location of hubs (address, longitude, and latitude) and the number of
racks available at each hub. The dataset also includes the number of
bicycles available at each hub, but since this changes daily, this
information is not useful for our purposes. The maximum number of
bicycles that a hub \emph{could} have (i.e., the baseline) available is
important for calculating accessibility in a given area.

We can also retrieve the data weekly over a certain period, for example
4 weeks, to calculate the average number of bicycles that are at each
hub. Variations in the number of bicycles could also be informative. For
instance, some hubs may systematically have more or less bicycles than
others (i.e., excess of supply around the University campus).

\hypertarget{sobi-service-area}{%
\subsubsection{SoBi Service Area}\label{sobi-service-area}}

The dataset was last updated on \emph{November 9, 2020} and was
downloaded for this project on \emph{November 10, 2020}. Hamilton's bike
share program is available only in the lower city from Dundas to Ottawa
Street North. There is one isolated hub beyond this service area at Van
Wagner's Beach for tourism purposes.

\hypertarget{golf-courses}{%
\subsubsection{Golf Courses}\label{golf-courses}}

The dataset was last updated on \emph{November 1, 2020} and was
downloaded for this project on \emph{November 10, 2020}. This dataset
contains the location of City and privately owned golf courses in
Hamilton.

\hypertarget{parks}{%
\subsubsection{Parks}\label{parks}}

The dataset was last updated on \emph{November 8, 2020} and was
downloaded for this project on \emph{November 10, 2020}. This dataset
contains the location of parks and other green spaces in Hamilton.

\hypertarget{large-employment-areas}{%
\subsubsection{Large Employment Areas}\label{large-employment-areas}}

The dataset contains Employment Land boundaries in Hamilton and was last
updated on \emph{November 12, 2020}. It was downloaded for this project
on \emph{November 13, 2020}. This dataset contains the location of large
business parks and industrial lands in Hamilton.

\hypertarget{municipal-parking-lots}{%
\subsubsection{Municipal Parking Lots}\label{municipal-parking-lots}}

The dataset contains the location of municipal car parks in Hamilton and
was last updated on \emph{November 9, 2020}. It was downloaded for this
project on \emph{November 13, 2020}.

\hypertarget{cemeteries}{%
\subsubsection{Cemeteries}\label{cemeteries}}

The dataset contains the location of cemeteries in Hamilton. It was
downloaded for this project on \emph{November 13, 2020}.

\hypertarget{environmentally-sensitive-areas}{%
\subsubsection{Environmentally Sensitive
Areas}\label{environmentally-sensitive-areas}}

The dataset contains the location of Environmentally Sensitive Areas
(ESAs) in Hamilton. ESAs are either land or water areas containing
natural features or significant ecological functions in Hamilton. The
data were downloaded on \emph{November 13, 2020}.

\hypertarget{street-centreline}{%
\subsubsection{Street Centreline}\label{street-centreline}}

The dataset contains the street network in Hamilton. Data attributes
include road classification so highways can be extracted. The dataset
was last updated on \emph{November 15, 2020} and downloaded on
\emph{November 16, 2020}.

\hypertarget{educational-institutions}{%
\subsubsection{Educational
Institutions}\label{educational-institutions}}

The dataset contains the location of all educational institutions and
schools in Hamilton. The dataset was last updated on \emph{November 15,
2020} and downloaded on \emph{November 16, 2020}.

\hypertarget{places-of-workshop}{%
\subsubsection{Places of Workshop}\label{places-of-workshop}}

The dataset contains the location of buildings used for religious
congregations in Hamilton. The dataset was last updated on
\emph{November 15, 2020} and downloaded on \emph{November 16, 2020}.

\hypertarget{municipal-service-centres}{%
\subsubsection{Municipal Service
Centres}\label{municipal-service-centres}}

The dataset contains the location of all municipal service centres in
Hamilton. This includes Hamilton City Hall. The dataset was last updated
on \emph{November 11, 2020} and downloaded on \emph{November 16, 2020}.

\hypertarget{recreation-and-community-centres}{%
\subsubsection{Recreation and Community
Centres}\label{recreation-and-community-centres}}

The dataset contains the location of all recreation and community
centres in Hamilton. This includes Hamilton City Hall. The dataset was
last updated on \emph{November 15, 2020} and downloaded on
\emph{November 16, 2020}.

\hypertarget{arenas}{%
\subsubsection{Arenas}\label{arenas}}

The dataset contains the location of all indoor arenas in Hamilton. This
includes Hamilton City Hall. The dataset was last updated on
\emph{November 12, 2020} and downloaded on \emph{November 16, 2020}.

\hypertarget{ems-stations}{%
\subsubsection{EMS Stations}\label{ems-stations}}

The dataset contains the location of all Emergency Management Services
(EMS) Ambulance stations in Hamilton. This includes Hamilton City Hall.
The dataset was last updated on \emph{November 11, 2020} and downloaded
on \emph{November 16, 2020}.

\hypertarget{fire-stations}{%
\subsubsection{Fire Stations}\label{fire-stations}}

The dataset contains the location of all fire stations in Hamilton. This
includes Hamilton City Hall. The dataset was last updated on
\emph{November 15, 2020} and downloaded on \emph{November 16, 2020}.

\hypertarget{police-stations}{%
\subsubsection{Police Stations}\label{police-stations}}

The dataset contains the location of all police stations in Hamilton.
This includes Hamilton City Hall. The dataset was last updated on
\emph{November 13, 2020} and downloaded on \emph{November 16, 2020}.

\hypertarget{hospitals}{%
\subsubsection{Hospitals}\label{hospitals}}

The dataset contains the location of all hospitals in Hamilton. This
includes Hamilton City Hall. The dataset was last updated on \emph{April
2, 2019} and downloaded on \emph{November 16, 2020}.

\hypertarget{preliminaries}{%
\subsection{Preliminaries}\label{preliminaries}}

Clear environment:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{rm}\NormalTok{(}\DataTypeTok{list =} \KeywordTok{ls}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

Load packages used in the notebook:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{library}\NormalTok{(disk.frame)}
\KeywordTok{library}\NormalTok{(pycno)}
\KeywordTok{library}\NormalTok{(readr)}
\KeywordTok{library}\NormalTok{(r5r) }\CommentTok{# the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html}
\KeywordTok{library}\NormalTok{(sf)}
\KeywordTok{library}\NormalTok{(tidyverse)}
\end{Highlighting}
\end{Shaded}

Define some parameters for \texttt{disk.frame} and \texttt{r5r}:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{setup_disk.frame}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## The number of workers available for disk.frame is 4
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{options}\NormalTok{(}\DataTypeTok{scipen =} \DecValTok{999}\NormalTok{)}
\KeywordTok{options}\NormalTok{(}\DataTypeTok{java.parameters =} \StringTok{"-Xmx6G"}\NormalTok{)}
\KeywordTok{options}\NormalTok{(}\DataTypeTok{future.globals.maxSize =} \OtherTok{Inf}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{read-data}{%
\subsubsection{Read data}\label{read-data}}

Load data files:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{load}\NormalTok{(}\StringTok{"hamilton_cma.RData"}\NormalTok{)}
\KeywordTok{load}\NormalTok{(}\StringTok{"hamilton_da_2016.RData"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

These two data files are simple features with the dissemination areas
according to the 2016 Census, and the boundary of Hamilton CMA.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ hamilton_cma)}
\end{Highlighting}
\end{Shaded}

\includegraphics{00-Data-Processing-Example_files/figure-latex/unnamed-chunk-5-1.pdf}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ hamilton_da_}\DecValTok{2016}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{00-Data-Processing-Example_files/figure-latex/unnamed-chunk-6-1.pdf}

Read Sobi service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sobi_service <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"SoBi_Service_Areas.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `SoBi_Service_Areas' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\SoBi_Service_Areas.shp' using driver `ESRI Shapefile'
## Simple feature collection with 2 features and 3 fields
## geometry type:  POLYGON
## dimension:      XY
## bbox:           xmin: 583329.4 ymin: 4787385 xmax: 600145.7 ymax: 4793252
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Plot service areas:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sobi_service,}
          \KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill =}\NormalTok{ AREA_NAME))}
\end{Highlighting}
\end{Shaded}

\includegraphics{00-Data-Processing-Example_files/figure-latex/unnamed-chunk-8-1.pdf}

There is a Core Service Area and a much smaller service area at Van
Wagner's. This is a location at the beach with service for beach goers.
We will remove this area from further analysis:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sobi_service <-}\StringTok{ }\NormalTok{sobi_service }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(AREA_NAME }\OperatorTok{==}\StringTok{ "Core Service Area"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Bounding box of population in serviced areas:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{bounding_box <-}\StringTok{ }\KeywordTok{st_bbox}\NormalTok{(sobi_service }\OperatorTok{%>%}
\StringTok{                          }\KeywordTok{st_buffer}\NormalTok{(}\DecValTok{500}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

Plot:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ hamilton_da_}\DecValTok{2016}\NormalTok{,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{,}
          \DataTypeTok{color =} \StringTok{"lightgray"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sobi_service,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =} \KeywordTok{st_as_sfc}\NormalTok{(bounding_box),}
          \DataTypeTok{linetype =} \StringTok{"dashed"}\NormalTok{,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{00-Data-Processing-Example_files/figure-latex/unnamed-chunk-11-1.pdf}

Read parks:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{parks <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Parks.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Parks' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Parks.shp' using driver `ESRI Shapefile'
## Simple feature collection with 622 features and 7 fields
## geometry type:  MULTIPOLYGON
## dimension:      XY
## bbox:           xmin: 564103.5 ymin: 4774053 xmax: 611193.6 ymax: 4809561
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read cemeteries:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cemeteries <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Cemeteries.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Cemeteries' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Cemeteries.shp' using driver `ESRI Shapefile'
## Simple feature collection with 31 features and 10 fields
## geometry type:  MULTIPOLYGON
## dimension:      XY
## bbox:           xmin: 567538.9 ymin: 4785183 xmax: 606126.6 ymax: 4794820
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read environmentally sensitive areas:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{esa <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Environmentally_Sensitive_Areas_Boundaries.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Environmentally_Sensitive_Areas_Boundaries' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Environmentally_Sensitive_Areas_Boundaries.shp' using driver `ESRI Shapefile'
## Simple feature collection with 207 features and 9 fields
## geometry type:  POLYGON
## dimension:      XY
## bbox:           xmin: 561664.6 ymin: 4767888 xmax: 611963.3 ymax: 4814265
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read golf courses:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{golf_courses <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Golf_Courses.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Golf_Courses' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Golf_Courses.shp' using driver `ESRI Shapefile'
## Simple feature collection with 27 features and 4 fields
## geometry type:  POLYGON
## dimension:      XY
## bbox:           xmin: 565041.3 ymin: 4771006 xmax: 608445.4 ymax: 4807785
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read employment land areas, which include business parks and industrial
lands:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{employment_lands <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Employment_Lands.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Employment_Lands' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Employment_Lands.shp' using driver `ESRI Shapefile'
## Simple feature collection with 14 features and 5 fields
## geometry type:  MULTIPOLYGON
## dimension:      XY
## bbox:           xmin: 577850.4 ymin: 4778238 xmax: 611498.4 ymax: 4797555
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read railways:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{railways <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Railways.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Railways' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Railways.shp' using driver `ESRI Shapefile'
## Simple feature collection with 1692 features and 4 fields
## geometry type:  MULTILINESTRING
## dimension:      XY
## bbox:           xmin: 563354.6 ymin: 4780233 xmax: 612351 ymax: 4813272
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read municipal parking lots:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{municipal_parking <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Municipal_Parking_Lots.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Municipal_Parking_Lots' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Municipal_Parking_Lots.shp' using driver `ESRI Shapefile'
## Simple feature collection with 62 features and 4 fields
## geometry type:  MULTIPOLYGON
## dimension:      XY
## bbox:           xmin: 583117.5 ymin: 4785525 xmax: 601281.2 ymax: 4798542
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read street network:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{streets <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Street_Centreline.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Street_Centreline' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Street_Centreline.shp' using driver `ESRI Shapefile'
## Simple feature collection with 19396 features and 31 fields
## geometry type:  LINESTRING
## dimension:      XY
## bbox:           xmin: 560789.8 ymin: 4767178 xmax: 611887.1 ymax: 4813706
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Filter for provincial highways:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{streets <-}\StringTok{ }\NormalTok{streets }\OperatorTok{%>%}\StringTok{ }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(ROAD_CLASS }\OperatorTok{==}\StringTok{ "Provincial Highway"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Read educational institutions:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{schools <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Educational_Institutions.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Educational_Institutions' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Educational_Institutions.shp' using driver `ESRI Shapefile'
## Simple feature collection with 234 features and 14 fields
## geometry type:  POINT
## dimension:      XY
## bbox:           xmin: 564817.6 ymin: 4774480 xmax: 610267.4 ymax: 4806311
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read places of worship:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religious_places <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Places_of_Worship.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Places_of_Worship' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Places_of_Worship.shp' using driver `ESRI Shapefile'
## Simple feature collection with 356 features and 7 fields
## geometry type:  POINT
## dimension:      XY
## bbox:           xmin: 564564 ymin: 4774769 xmax: 607177.1 ymax: 4809573
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read municipal service centres:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{service_centres <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Municipal_Service_Centres.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Municipal_Service_Centres' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Municipal_Service_Centres.shp' using driver `ESRI Shapefile'
## Simple feature collection with 6 features and 6 fields
## geometry type:  POINT
## dimension:      XY
## bbox:           xmin: 583101.7 ymin: 4775812 xmax: 606290.1 ymax: 4797349
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read recreation and centres:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{recreation_centres <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Recreation_and_Community_Centres.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Recreation_and_Community_Centres' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Recreation_and_Community_Centres.shp' using driver `ESRI Shapefile'
## Simple feature collection with 68 features and 4 fields
## geometry type:  POINT
## dimension:      XY
## bbox:           xmin: 564831.5 ymin: 4775169 xmax: 610016 ymax: 4809482
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read arenas:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{arenas <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Arenas.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Arenas' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Arenas.shp' using driver `ESRI Shapefile'
## Simple feature collection with 25 features and 7 fields
## geometry type:  POINT
## dimension:      XY
## bbox:           xmin: 572082.8 ymin: 4775909 xmax: 605573.3 ymax: 4805382
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read EMS stations:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ems <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"EMS_Stations.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `EMS_Stations' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\EMS_Stations.shp' using driver `ESRI Shapefile'
## Simple feature collection with 20 features and 7 fields
## geometry type:  POINT
## dimension:      XY
## bbox:           xmin: 569174.6 ymin: 4775023 xmax: 604223.7 ymax: 4798687
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read fire stations:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fire <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Fire_Stations.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Fire_Stations' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Fire_Stations.shp' using driver `ESRI Shapefile'
## Simple feature collection with 30 features and 9 fields
## geometry type:  POINT
## dimension:      XY
## bbox:           xmin: 569174.6 ymin: 4775023 xmax: 607576.5 ymax: 4805440
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read police stations:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{police <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Police_Stations.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Police_Stations' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Police_Stations.shp' using driver `ESRI Shapefile'
## Simple feature collection with 3 features and 4 fields
## geometry type:  POINT
## dimension:      XY
## bbox:           xmin: 591279.1 ymin: 4783409 xmax: 599816.9 ymax: 4789897
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Read hospitals:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hospitals <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"Hospitals.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `Hospitals' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\Hospitals.shp' using driver `ESRI Shapefile'
## Simple feature collection with 9 features and 7 fields
## geometry type:  POINT
## dimension:      XY
## bbox:           xmin: 587850 ymin: 4786167 xmax: 599582 ymax: 4790541
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

\hypertarget{extract-features-in-service-area}{%
\subsubsection{Extract features in service
area}\label{extract-features-in-service-area}}

Project the Hamilton DA zones to the same projection of the SoBi Service
Area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_}\DecValTok{2016}\NormalTok{ <-}\StringTok{ }\KeywordTok{st_transform}\NormalTok{(hamilton_da_}\DecValTok{2016}\NormalTok{, }\DataTypeTok{crs =} \KeywordTok{st_crs}\NormalTok{(sobi_service))}
\end{Highlighting}
\end{Shaded}

Extract DAs in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi <-}\StringTok{ }\NormalTok{hamilton_da_}\DecValTok{2016}\NormalTok{[}\KeywordTok{st_as_sfc}\NormalTok{(bounding_box),]}
\end{Highlighting}
\end{Shaded}

Extract parks in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{parks_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(parks, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract cemeteries in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{cemeteries_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(cemeteries, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract environmentally sensitive areas in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{esa_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(esa, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract golf courses in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{golf_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(golf_courses, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract employment lands in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{employment_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(employment_lands, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract railways in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{railways_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(railways, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract municipal parking lots in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{parking_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(municipal_parking, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract provincial highways in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{highways_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(streets, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract educational institutions in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{schools_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(schools, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract places of worship in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{religious_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(religious_places, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract municipal service centres in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{service_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(service_centres, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract recreation and community centres in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{recreation_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(recreation_centres, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract indoor arenas in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{arenas_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(arenas, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract educational institutions in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{schools_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(schools, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract EMS stations in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ems_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(ems, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract fire stations in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fire_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(fire, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract police stations in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{police_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(police, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Extract hospitals in service area:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hospitals_sobi <-}\StringTok{ }\KeywordTok{st_crop}\NormalTok{(hospitals, }\KeywordTok{st_as_sfc}\NormalTok{(bounding_box))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

\hypertarget{read-population}{%
\subsubsection{Read population}\label{read-population}}

Read population table (2016 Census of Population, Population and
Dwelling counts):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{population_da_}\DecValTok{2016}\NormalTok{ <-}\StringTok{ }\KeywordTok{read_csv}\NormalTok{(}\StringTok{"T1901EN.csv"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{transmute}\NormalTok{(}\DataTypeTok{DA =} \KeywordTok{factor}\NormalTok{(}\StringTok{`}\DataTypeTok{Geographic code}\StringTok{`}\NormalTok{), }\DataTypeTok{population =} \StringTok{`}\DataTypeTok{Population, 2016}\StringTok{`}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: Missing column names filled in: 'X13' [13]
\end{verbatim}

\begin{verbatim}
## 
## -- Column specification --------------------------------------------------------
## cols(
##   `Geographic code` = col_double(),
##   `Province / territory, english` = col_character(),
##   `Province / territory, french` = col_character(),
##   `Geographic code, Province / territory` = col_double(),
##   `Geographic code, Census division` = col_double(),
##   `Geographic code, Census subdivision` = col_double(),
##   `Population, 2016` = col_double(),
##   `Incompletely enumerated Indian reserves and Indian settlements, 2016` = col_logical(),
##   `Total private dwellings, 2016` = col_double(),
##   `Private dwellings occupied by usual residents, 2016` = col_double(),
##   `Land area in square kilometres, 2016` = col_double(),
##   `Population density per square kilometre, 2016` = col_double(),
##   X13 = col_character()
## )
\end{verbatim}

\begin{verbatim}
## Warning: 56602 parsing failures.
## row col   expected     actual          file
##   1  -- 13 columns 12 columns 'T1901EN.csv'
##   2  -- 13 columns 12 columns 'T1901EN.csv'
##   3  -- 13 columns 12 columns 'T1901EN.csv'
##   4  -- 13 columns 12 columns 'T1901EN.csv'
##   5  -- 13 columns 12 columns 'T1901EN.csv'
## ... ... .......... .......... .............
## See problems(...) for more details.
\end{verbatim}

Join population table to DAs in service areas:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi <-}\StringTok{ }\NormalTok{hamilton_da_sobi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{left_join}\NormalTok{(population_da_}\DecValTok{2016}\NormalTok{, }\DataTypeTok{by =} \StringTok{"DA"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\hypertarget{visualize-geographical-data}{%
\subsubsection{Visualize geographical
data}\label{visualize-geographical-data}}

Plot population:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ hamilton_da_sobi,}
          \KeywordTok{aes}\NormalTok{(}\DataTypeTok{fill =}\NormalTok{ population),}
          \DataTypeTok{color =} \OtherTok{NA}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sobi_service,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_fill_distiller}\NormalTok{(}\DataTypeTok{palette =} \StringTok{"YlOrRd"}\NormalTok{, }\DataTypeTok{direction =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{00-Data-Processing-Example_files/figure-latex/unnamed-chunk-52-1.pdf}

Plot ``Green'' spaces:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ parks_sobi,}
          \DataTypeTok{fill =} \StringTok{"blue"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ cemeteries_sobi,}
          \DataTypeTok{fill =} \StringTok{"gray"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ esa_sobi,}
          \DataTypeTok{fill =} \StringTok{"green"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ golf_sobi,}
          \DataTypeTok{fill =} \StringTok{"lightskyblue"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ employment_sobi,}
          \DataTypeTok{fill =} \StringTok{"purple"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ railways_sobi,}
          \DataTypeTok{fill =} \StringTok{"wheat"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ parking_sobi,}
          \DataTypeTok{fill =} \StringTok{"lightpink"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ highways_sobi,}
          \DataTypeTok{fill =} \StringTok{"orange"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ schools_sobi,}
          \DataTypeTok{fill =} \StringTok{"red"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ religious_sobi,}
          \DataTypeTok{fill =} \StringTok{"pink"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ service_sobi,}
          \DataTypeTok{fill =} \StringTok{"yellow"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ recreation_sobi,}
          \DataTypeTok{fill =} \StringTok{"yellow"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ arenas_sobi,}
          \DataTypeTok{fill =} \StringTok{"yellow"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ ems_sobi,}
          \DataTypeTok{fill =} \StringTok{"yellow"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ ems_sobi,}
          \DataTypeTok{fill =} \StringTok{"yellow"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ fire_sobi,}
          \DataTypeTok{fill =} \StringTok{"yellow"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ police_sobi,}
          \DataTypeTok{fill =} \StringTok{"yellow"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ hospitals_sobi,}
          \DataTypeTok{fill =} \StringTok{"yellow"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sobi_service,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{,}
          \DataTypeTok{color =} \StringTok{"black"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ hamilton_da_sobi,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{,}
          \DataTypeTok{color =} \StringTok{"gray"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_fill_distiller}\NormalTok{(}\DataTypeTok{palette =} \StringTok{"YlOrRd"}\NormalTok{, }\DataTypeTok{direction =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{00-Data-Processing-Example_files/figure-latex/unnamed-chunk-53-1.pdf}

\hypertarget{prepare-geographical-data-for-interpolation}{%
\subsubsection{Prepare geographical data for
interpolation}\label{prepare-geographical-data-for-interpolation}}

Union of the ``green space'' features:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{parks_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(parks_sobi)}
\NormalTok{cemeteries_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(cemeteries_sobi)}
\NormalTok{esa_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(esa_sobi)}
\NormalTok{golf_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(golf_sobi)}
\NormalTok{employment_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(employment_sobi)}
\NormalTok{railways_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(railways_sobi)}
\NormalTok{parking_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(parking_sobi)}
\NormalTok{highways_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(highways_sobi)}
\NormalTok{schools_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(schools_sobi)}
\NormalTok{religious_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(religious_sobi)}
\NormalTok{service_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(service_sobi)}
\NormalTok{recreation_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(recreation_sobi)}
\NormalTok{arenas_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(arenas_sobi)}
\NormalTok{ems_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(ems_sobi)}
\NormalTok{fire_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(fire_sobi)}
\NormalTok{police_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(police_sobi)}
\NormalTok{hospitals_sobi <-}\StringTok{ }\KeywordTok{st_combine}\NormalTok{(hospitals_sobi)}
\end{Highlighting}
\end{Shaded}

Check that the topology is valid:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(parks_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] FALSE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(cemeteries_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(esa_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] FALSE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(golf_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(employment_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(railways_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(parking_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] FALSE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(highways_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(schools_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(religious_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(service_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(recreation_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(arenas_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(ems_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(fire_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(police_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{st_is_valid}\NormalTok{(hospitals_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] TRUE
\end{verbatim}

Make valid:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{parks_sobi <-}\StringTok{ }\KeywordTok{st_make_valid}\NormalTok{(parks_sobi)}
\NormalTok{esa_sobi <-}\StringTok{ }\KeywordTok{st_make_valid}\NormalTok{(esa_sobi)}
\NormalTok{parking_sobi <-}\StringTok{ }\KeywordTok{st_make_valid}\NormalTok{(parking_sobi)}
\end{Highlighting}
\end{Shaded}

\hypertarget{remove-all-features-that-are-not-population-from-the-das}{%
\subsubsection{Remove all features that are not population from the
DAs}\label{remove-all-features-that-are-not-population-from-the-das}}

Remove the parks from the DAs (use \texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi, parks_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove the cemeteries from the DAs (use \texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, cemeteries_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove environmentally sensitive areas from the DAs (use
\texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, esa_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove golf courses from the DAs (use \texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, golf_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove employment lands from the DAs (use \texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, employment_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove railways from the DAs (use \texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, railways_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove municipal parking lots from the DAs (use
\texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, parking_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove provincial highways from the DAs (use \texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, highways_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove educational institutions from the DAs (use
\texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, schools_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove places of worship from the DAs (use \texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, religious_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove municipal service centres from the DAs (use
\texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, service_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove recreation and community centres from the DAs (use
\texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, recreation_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove arenas from the DAs (use \texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, arenas_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove EMS stations from the DAs (use \texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, ems_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove fire stations from the DAs (use \texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, fire_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove police stations from the DAs (use \texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, police_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Remove hospitals from the DAs (use \texttt{st\_difference()}):

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean <-}\StringTok{ }\KeywordTok{st_difference}\NormalTok{(hamilton_da_sobi_clean, hospitals_sobi)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning: attribute variables are assumed to be spatially constant throughout all
## geometries
\end{verbatim}

Plot the new DAs (minus green) and compare to original DAs:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ hamilton_da_sobi_clean,}
          \DataTypeTok{color =} \OtherTok{NA}\NormalTok{,}
          \DataTypeTok{fill =} \StringTok{"blue"}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ hamilton_da_sobi,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{,}
          \DataTypeTok{color =} \StringTok{"black"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{00-Data-Processing-Example_files/figure-latex/unnamed-chunk-74-1.pdf}

\hypertarget{interpolate-population}{%
\subsection{Interpolate population}\label{interpolate-population}}

Pycnophylactic interpolation needs a \texttt{SpatialPolygonsDataFrame}
object. Convert the simple features object:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hamilton_da_sobi_clean.sp <-}\StringTok{ }\KeywordTok{as}\NormalTok{(hamilton_da_sobi_clean, }\StringTok{"Spatial"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Extract the population from the spatial object:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{pop <-}\StringTok{ }\NormalTok{hamilton_da_sobi_clean.sp}\OperatorTok{@}\NormalTok{data}\OperatorTok{$}\NormalTok{population}
\end{Highlighting}
\end{Shaded}

Pycnophylactic interpolation:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{filter <-}\StringTok{ }\NormalTok{stats}\OperatorTok{::}\NormalTok{filter }\CommentTok{# Important! `filter()` should be from `stats`, not `dplyer`}
\NormalTok{interpolated_pop <-}\StringTok{ }\KeywordTok{pycno}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ hamilton_da_sobi_clean.sp, }\DataTypeTok{pops =}\NormalTok{ pop, }\DataTypeTok{celldim =} \DecValTok{50}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in proj4string(x): CRS object has comment, which is lost in output
\end{verbatim}

\begin{verbatim}
## Maximum Change:     41.20295 - will stop at      0.11700
## Maximum Change:     15.53913 - will stop at      0.11700
## Maximum Change:     10.50693 - will stop at      0.11700
## Maximum Change:      6.89863 - will stop at      0.11700
## Maximum Change:      5.30048 - will stop at      0.11700
## Maximum Change:      3.68337 - will stop at      0.11700
## Maximum Change:      2.48480 - will stop at      0.11700
## Maximum Change:      2.01939 - will stop at      0.11700
## Maximum Change:      1.67714 - will stop at      0.11700
## Maximum Change:      1.42249 - will stop at      0.11700
## Maximum Change:      1.19921 - will stop at      0.11700
## Maximum Change:      1.02058 - will stop at      0.11700
## Maximum Change:      0.86826 - will stop at      0.11700
## Maximum Change:      0.74189 - will stop at      0.11700
## Maximum Change:      0.63467 - will stop at      0.11700
## Maximum Change:      0.54407 - will stop at      0.11700
## Maximum Change:      0.46724 - will stop at      0.11700
## Maximum Change:      0.40221 - will stop at      0.11700
## Maximum Change:      0.34688 - will stop at      0.11700
## Maximum Change:      0.29977 - will stop at      0.11700
## Maximum Change:      0.25950 - will stop at      0.11700
## Maximum Change:      0.22496 - will stop at      0.11700
## Maximum Change:      0.20150 - will stop at      0.11700
## Maximum Change:      0.18438 - will stop at      0.11700
## Maximum Change:      0.16802 - will stop at      0.11700
## Maximum Change:      0.15292 - will stop at      0.11700
## Maximum Change:      0.13920 - will stop at      0.11700
## Maximum Change:      0.12700 - will stop at      0.11700
## Maximum Change:      0.11602 - will stop at      0.11700
\end{verbatim}

\begin{verbatim}
## Warning in proj4string(x): CRS object has comment, which is lost in output
\end{verbatim}

Convert to pixels and then simple features:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{interpolated_pop.pix <-}\StringTok{ }\KeywordTok{as}\NormalTok{(interpolated_pop, }\StringTok{"SpatialPixelsDataFrame"}\NormalTok{)}
\NormalTok{interpolated_pop <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{population =}\NormalTok{ interpolated_pop.pix}\OperatorTok{$}\NormalTok{dens, }
                               \DataTypeTok{x =}\NormalTok{ interpolated_pop.pix}\OperatorTok{@}\NormalTok{coords[,}\DecValTok{1}\NormalTok{],}
                               \DataTypeTok{y =}\NormalTok{ interpolated_pop.pix}\OperatorTok{@}\NormalTok{coords[,}\DecValTok{2}\NormalTok{]) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{st_as_sf}\NormalTok{(}\DataTypeTok{coords =} \KeywordTok{c}\NormalTok{(}\StringTok{"x"}\NormalTok{, }\StringTok{"y"}\NormalTok{),}
           \DataTypeTok{crs =} \KeywordTok{st_crs}\NormalTok{(hamilton_da_sobi))}
\end{Highlighting}
\end{Shaded}

Check the population totals:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{sum}\NormalTok{(hamilton_da_sobi_clean}\OperatorTok{$}\NormalTok{population)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 206539
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{sum}\NormalTok{(interpolated_pop}\OperatorTok{$}\NormalTok{population)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 206539
\end{verbatim}

Plot interpolated population:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ interpolated_pop,}
          \KeywordTok{aes}\NormalTok{(}\DataTypeTok{color =}\NormalTok{ population)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_color_distiller}\NormalTok{(}\DataTypeTok{palette =} \StringTok{"YlOrRd"}\NormalTok{, }
                        \DataTypeTok{direction =} \DecValTok{1}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ hamilton_da_sobi,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{,}
          \DataTypeTok{color =} \StringTok{"lightgray"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sobi_service,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{00-Data-Processing-Example_files/figure-latex/unnamed-chunk-80-1.pdf}

\hypertarget{extract-population-in-service-area}{%
\subsection{Extract population in service
area}\label{extract-population-in-service-area}}

Extract population cells in the service area (and buffer)

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{interpolated_pop_clean <-}\StringTok{ }\NormalTok{interpolated_pop[sobi_service }\OperatorTok{%>%}\StringTok{ }\KeywordTok{st_buffer}\NormalTok{(}\DecValTok{500}\NormalTok{),]}
\end{Highlighting}
\end{Shaded}

Note that some cells have population values of zero:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{interpolated_pop_clean }\OperatorTok{%>%}\StringTok{ }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{select}\NormalTok{(population) }\OperatorTok{%>%}\StringTok{ }\KeywordTok{summary}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##    population               geometry    
##  Min.   :  0.000   POINT        :11902  
##  1st Qu.:  4.053   epsg:26917   :    0  
##  Median :  9.415   +proj=utm ...:    0  
##  Mean   : 11.564                        
##  3rd Qu.: 15.243                        
##  Max.   :140.184
\end{verbatim}

Filter all cells with non-zero population values:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{interpolated_pop_clean <-}\StringTok{ }\NormalTok{interpolated_pop_clean }\OperatorTok{%>%}
\StringTok{            }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(population }\OperatorTok{>}\StringTok{ }\DecValTok{0}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Add identifier for the population cells:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{interpolated_pop_clean <-}\StringTok{ }\NormalTok{interpolated_pop_clean }\OperatorTok{%>%}
\StringTok{            }\KeywordTok{mutate}\NormalTok{(}\DataTypeTok{UID =} \DecValTok{1}\OperatorTok{:}\KeywordTok{n}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

Plot interpolated population again:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ interpolated_pop_clean,}
          \KeywordTok{aes}\NormalTok{(}\DataTypeTok{color =}\NormalTok{ population)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sobi_service,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_color_distiller}\NormalTok{(}\DataTypeTok{palette =} \StringTok{"YlOrRd"}\NormalTok{, }
                        \DataTypeTok{direction =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{00-Data-Processing-Example_files/figure-latex/unnamed-chunk-85-1.pdf}

\hypertarget{bikeshare-system}{%
\subsection{Bikeshare system}\label{bikeshare-system}}

Read Sobi hubs:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sobi_hubs <-}\StringTok{ }\KeywordTok{st_read}\NormalTok{(}\StringTok{"SoBi_Hubs.shp"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Reading layer `SoBi_Hubs' from data source `C:\Users\desjae\Documents\GitHub\Accessibility-Sobi-Hamilton\examples\SoBi_Hubs.shp' using driver `ESRI Shapefile'
## Simple feature collection with 134 features and 11 fields
## geometry type:  POINT
## dimension:      XY
## bbox:           xmin: 584909.4 ymin: 4788377 xmax: 600125.2 ymax: 4792103
## projected CRS:  NAD83 / UTM zone 17N
\end{verbatim}

Plot interpolated population and SoBi hubs:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ interpolated_pop_clean,}
          \KeywordTok{aes}\NormalTok{(}\DataTypeTok{color =}\NormalTok{ population)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sobi_service,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sobi_hubs) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_color_distiller}\NormalTok{(}\DataTypeTok{palette =} \StringTok{"YlOrRd"}\NormalTok{, }
                        \DataTypeTok{direction =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{00-Data-Processing-Example_files/figure-latex/unnamed-chunk-87-1.pdf}

Remove SoBi hub on beach:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sobi_hubs <-}\StringTok{ }\NormalTok{sobi_hubs }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "Van Wagners"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Plot again:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ interpolated_pop_clean,}
          \KeywordTok{aes}\NormalTok{(}\DataTypeTok{color =}\NormalTok{ population)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sobi_service,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sobi_hubs,}
          \KeywordTok{aes}\NormalTok{(}\DataTypeTok{size =}\NormalTok{ RACKS_AMOU)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_color_distiller}\NormalTok{(}\DataTypeTok{palette =} \StringTok{"YlOrRd"}\NormalTok{, }
                        \DataTypeTok{direction =} \DecValTok{1}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_size}\NormalTok{(}\DataTypeTok{range =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{4}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\includegraphics{00-Data-Processing-Example_files/figure-latex/unnamed-chunk-89-1.pdf}

Remove ERI hubs that were implemented in 2018 to increase accessibility
for under-serviced and disadvantaged areas of the city. This reflects
the original state of the system without the equity stations:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sobi_hubs_original <-}\StringTok{ }\NormalTok{sobi_hubs }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "Mars at Wentworth - ERI01"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "Sherman at Barton - ERI02"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "Barton at Lottridge - ERI03"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "Barton and Belview - ERI05"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "Barton at Ottawa - ERI06"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "Ottawa at Dunsmure - ERI07"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "Maple at Rothsay - ERI08"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "King at Dunsmure - ERI09"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "Dunsmure at Sherman - ERI10"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "Gage at Cannon - ERI11"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "Belview at Cannon - ERI12"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{filter}\NormalTok{(NAME }\OperatorTok{!=}\StringTok{ "Westinghouse at Barton - ERI13"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Plot again:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{() }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ interpolated_pop_clean,}
          \KeywordTok{aes}\NormalTok{(}\DataTypeTok{color =}\NormalTok{ population)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sobi_service,}
          \DataTypeTok{fill =} \OtherTok{NA}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{geom_sf}\NormalTok{(}\DataTypeTok{data =}\NormalTok{ sobi_hubs_original,}
          \KeywordTok{aes}\NormalTok{(}\DataTypeTok{size =}\NormalTok{ RACKS_AMOU)) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_color_distiller}\NormalTok{(}\DataTypeTok{palette =} \StringTok{"YlOrRd"}\NormalTok{, }
                        \DataTypeTok{direction =} \DecValTok{1}\NormalTok{) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_size}\NormalTok{(}\DataTypeTok{range =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{4}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\includegraphics{00-Data-Processing-Example_files/figure-latex/unnamed-chunk-91-1.pdf}
After removing the twelve ERI hubs, there is much less coverage in the
east end of the service area. Again, the service area is the area where
a cyclist can use a SoBi bike. It makes sense then that ERI would want
to fill this gap in coverage because individuals who live in the east
end of the service area could theoretically use the bike share system
but may not live near a hub. For this reason, twelve hubs were added in
2018 to expand the system.

\hypertarget{routing-for-entire-system}{%
\subsection{Routing for entire system}\label{routing-for-entire-system}}

I used \href{https://download.bbbike.org/osm/bbbike/}{BBBike} to extract
OSM data for Hamilton - not the whole city, but enough to cover SoBI's
service area. The name of the file is
\texttt{planet\_-80.042,43.183\_-79.748,43.309.osm.pbf}. Copy to folder
\texttt{r5\_graph}.

Set Up R5 Routing. First define the path to where the graph is located:

Build the graph:

Prepare Input Data for \texttt{r5r}. The origins are the coordinates of
the population cells and the destinations the coordinates of the SoBi
hubs:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# save origin centroids in format expected by R5R (id, lon, lat)}
\NormalTok{origins_i <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{UID =}\NormalTok{ interpolated_pop_clean}\OperatorTok{$}\NormalTok{UID, }
\NormalTok{              interpolated_pop_clean }\OperatorTok{%>%}
\StringTok{                }\KeywordTok{st_transform}\NormalTok{(}\DataTypeTok{crs =} \DecValTok{4326}\NormalTok{) }\OperatorTok{%>%}
\StringTok{                }\KeywordTok{st_coordinates}\NormalTok{()) }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{rename}\NormalTok{(}\DataTypeTok{lon =}\NormalTok{ X, }\DataTypeTok{lat =}\NormalTok{ Y, }\DataTypeTok{id =}\NormalTok{ UID) }\OperatorTok{%>%}
\StringTok{    }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{select}\NormalTok{(id, lon, lat)}

\CommentTok{# now SoBi hubs}
\NormalTok{destinations_j <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{OBJECTID =}\NormalTok{ sobi_hubs}\OperatorTok{$}\NormalTok{OBJECTID, }
\NormalTok{              sobi_hubs }\OperatorTok{%>%}
\StringTok{                }\KeywordTok{st_transform}\NormalTok{(}\DataTypeTok{crs =} \DecValTok{4326}\NormalTok{) }\OperatorTok{%>%}
\StringTok{                }\KeywordTok{st_coordinates}\NormalTok{()) }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{rename}\NormalTok{(}\DataTypeTok{lon =}\NormalTok{ X, }\DataTypeTok{lat =}\NormalTok{ Y, }\DataTypeTok{id =}\NormalTok{ OBJECTID) }\OperatorTok{%>%}
\StringTok{    }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{select}\NormalTok{(id, lon, lat)}
\end{Highlighting}
\end{Shaded}

Calculate OD Matrix for Walking:

\hypertarget{extract-travel-time-matrix}{%
\subsection{Extract travel time
matrix}\label{extract-travel-time-matrix}}

Convert disk.frame to data frame:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ttm_walk <-}\StringTok{ }\KeywordTok{as.data.frame}\NormalTok{(ttm_walk.disk.frame) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{transmute}\NormalTok{(}\DataTypeTok{UID =} \KeywordTok{as.numeric}\NormalTok{(fromId), }\DataTypeTok{OBJECTID =} \KeywordTok{as.numeric}\NormalTok{(toId), travel_time)}
\end{Highlighting}
\end{Shaded}

\hypertarget{routing-for-original-system-without-eri-hubs}{%
\subsection{Routing for original system without ERI
hubs}\label{routing-for-original-system-without-eri-hubs}}

Build the graph:

Prepare Input Data for \texttt{r5r}. The origins are the coordinates of
the population cells and the destinations the coordinates of the SoBi
hubs (minus the ERI hubs in the east end of the service area):

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# save origin centroids in format expected by R5R (id, lon, lat)}
\NormalTok{origins_io <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{UID =}\NormalTok{ interpolated_pop_clean}\OperatorTok{$}\NormalTok{UID, }
\NormalTok{              interpolated_pop_clean }\OperatorTok{%>%}
\StringTok{                }\KeywordTok{st_transform}\NormalTok{(}\DataTypeTok{crs =} \DecValTok{4326}\NormalTok{) }\OperatorTok{%>%}
\StringTok{                }\KeywordTok{st_coordinates}\NormalTok{()) }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{rename}\NormalTok{(}\DataTypeTok{lon =}\NormalTok{ X, }\DataTypeTok{lat =}\NormalTok{ Y, }\DataTypeTok{id =}\NormalTok{ UID) }\OperatorTok{%>%}
\StringTok{    }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{select}\NormalTok{(id, lon, lat)}

\CommentTok{# now SoBi hubs}
\NormalTok{destinations_jo <-}\StringTok{ }\KeywordTok{data.frame}\NormalTok{(}\DataTypeTok{OBJECTID =}\NormalTok{ sobi_hubs_original}\OperatorTok{$}\NormalTok{OBJECTID, }
\NormalTok{              sobi_hubs_original }\OperatorTok{%>%}
\StringTok{                }\KeywordTok{st_transform}\NormalTok{(}\DataTypeTok{crs =} \DecValTok{4326}\NormalTok{) }\OperatorTok{%>%}
\StringTok{                }\KeywordTok{st_coordinates}\NormalTok{()) }\OperatorTok{%>%}
\StringTok{    }\KeywordTok{rename}\NormalTok{(}\DataTypeTok{lon =}\NormalTok{ X, }\DataTypeTok{lat =}\NormalTok{ Y, }\DataTypeTok{id =}\NormalTok{ OBJECTID) }\OperatorTok{%>%}
\StringTok{    }\NormalTok{dplyr}\OperatorTok{::}\KeywordTok{select}\NormalTok{(id, lon, lat)}
\end{Highlighting}
\end{Shaded}

Calculate OD Matrix for Walking:

\hypertarget{extract-travel-time-matrix-for-original-system}{%
\subsection{Extract travel time matrix for original
system}\label{extract-travel-time-matrix-for-original-system}}

Convert disk.frame to data frame:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ttm_walk_o <-}\StringTok{ }\KeywordTok{as.data.frame}\NormalTok{(ttm_walk_o.disk.frame) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{transmute}\NormalTok{(}\DataTypeTok{UID =} \KeywordTok{as.numeric}\NormalTok{(fromId), }\DataTypeTok{OBJECTID =} \KeywordTok{as.numeric}\NormalTok{(toId), travel_time)}
\end{Highlighting}
\end{Shaded}

\hypertarget{save-data-to-disk}{%
\subsection{Save data to disk}\label{save-data-to-disk}}

Rename data frame:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{population_sobi <-}\StringTok{ }\NormalTok{interpolated_pop_clean}
\end{Highlighting}
\end{Shaded}

Save data files:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{save}\NormalTok{(population_sobi, }\DataTypeTok{file =} \StringTok{"population_sobi.RData"}\NormalTok{)}
\KeywordTok{save}\NormalTok{(sobi_hubs, }\DataTypeTok{file =} \StringTok{"sobi_hubs.RData"}\NormalTok{)}
\KeywordTok{save}\NormalTok{(sobi_hubs_original, }\DataTypeTok{file =} \StringTok{"sobi_hubs_original.RData"}\NormalTok{)}
\KeywordTok{save}\NormalTok{(sobi_service, }\DataTypeTok{file =} \StringTok{"sobi_service.RData"}\NormalTok{)}
\KeywordTok{save}\NormalTok{(ttm_walk, }\DataTypeTok{file =} \StringTok{"ttm_walk.RData"}\NormalTok{)}
\KeywordTok{save}\NormalTok{(ttm_walk_o, }\DataTypeTok{file =} \StringTok{"ttm_walk_original.RData"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\end{document}

### 2021-03-12 examples/Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Changes to draft manuscript

@@
 ---
-title: Examining spatial equity of a public bicycle share program using Floating Catchment Area methods 
+title: Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach 
 author:
-  - name: Antonio Paez
-    email: paez@mcmaster.ca
-    affiliation: Some School
-    footnote: 1
   - name: Elise Desjardins
@@
     affiliation: Some School
-    footnote: 2
+    footnote: 1
   - name: Christopher D. Higgins
@@
     footnote: 2
+  - name: Antonio Paez
+    email: paezha@mcmaster.ca
+    affiliation: Some School
+    footnote: 2
 address:
   - code: McMaster University
-    address: School of Earth, Environment & Society, 1280 Main Street West, Hamilton, ON
+    address: School of Earth, Environment & Society, 1280 Main Street West, Hamilton, ON L8S4L8
   - code: University of Toronto Scarborough
-    address: Department, Street, City, State, Zip
+    address: Department of Geography & Planning, 1265 Military Trail, Toronto, ON M1C1A4
 footnote:
@@
     text: "Equal contribution"
-abstract: |
-  This is the abstract.
+abstract:
 
-  It consists of two paragraphs.
+  Public bicycle share programs are implemented to promote cycling as a convenient and sustainable mode of transportation.
+These systems can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to lower income or disadvantaged populations.
+Recent studies assessing equity in bike sharing have found differences in ridership or membership based on income, education, and trip purpose.
+In Ontario, the City of Hamilton launched a public bicycle share program in 2015 that currently has over 900 operational bicycles and 130 docking stations.
+In 2018, the non-profit organization that operates the program launched an equity initiative to provide subsidized memberships and to expand service by adding twelve docking stations.
+Previous research found that Hamilton’s public bicycle share program targeted well disadvantaged areas in the city compared to programs in other Canadian cities.
+Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program, the location of stations matters.
+Unlike other public amenities with greater ability to adapt to crowding, the utility of stations is limited by the number of bicycles that they can hold, which makes crowding effects critical; the system may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles.
+In this research, we revisit the case of Hamilton and investigate equity differentials in accessibility to bicycle share stations.
+We compare accessibility in the program with and without the equity stations to assess the effect of the initiative.
+Previous research on cycling equity or accessibility has used a macro-level approach with entire neighbourhoods or census tracts as the unit of analysis, although there are recent exceptions.
+In contrast, we implement our analysis parting from microscopic zones to better reflect travel to a docking station, which often happens at the sub-neighborhood level.
+We then reaggregate the estimated accessibility to Dissemination Areas for further analysis using census data.
+Equity analysis of this type is possible thanks to a newly developed approach for accessibility using balanced floating catchment areas.
+We find that the addition of equity stations achieved its goal to increase accessibility for lower income populations, although the increase was relatively modest (2.10 bicycles/person with equity stations vs 1.97 bicycles/person without equity stations).
+Further research is needed to determine whether this encouraged more cycling.
 
@@
 date: "`r Sys.Date()`"
-bibliography: mybibfile.bib
+bibliography: references.bib
 #linenumbers: true
@@
 
-```{r echo=FALSE}
+```{r, include=FALSE}
+rm(list = ls())
+```
+
+```{r, include=FALSE}
+library(data.table)
 library(dplyr)
-library(disk.frame)
-library(pycno)
-library(readr)
-library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
-library(sf)
-library(tidyverse)
-library(data.table)
 library(gdistance)
@@
 library(sf)
+library(tidyverse)
 library(units)
+library(disk.frame)
+library(pycno)
+library(readr)
+library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
+library(tinytex)
+library(ggplot2)
+theme_set(theme_bw())
 ```
 
-```{r echo=FALSE}
+```{r, include=FALSE}
+equity_stations <- data.frame(longitude = c(	
+-79.84364993846138, -79.8362839215856, -79.83144521746212, -79.82217214949738, -79.8160211746482, -79.8192304372373, -79.82500255113169, -79.8268599805827, -79.83888834759689, -79.82740849222654, -79.82342541216097, -79.84240606419081), latitude = c(43.265340158000654, 43.25632467284136, 43.25528943845642, 43.25291050074978, 43.25147804403345, 43.24444659758296, 43.2421627336722, 43.24634382671116, 43.249432741404554, 43.251126444296524, 43.24892849896579, 43.25831920064561))
+```
+
+```{r, include=FALSE}
+#equity_stations <- st_as_sf(equity_stations, coords = c("longitude", "latitude"), 
+    #crs = st_crs(sobi_service), agr = "constant")
+```
+
+```{r, include=FALSE}
 load("hamilton_cma.RData")
@@
 load("ttm_walk_original.RData") #travel time matrix for system without ERI hubs
+load("accessibility.RData") #full accessibility table for SoBi system
+load("accessibility_map_original.RData") #map of accessibility in original SoBi system
+load("accessibility_map_equity.RData") #map of accessibility in current system with ERI
+load("hamilton_da_sobi.RData")
+load("population_sobi_original.RData") #population in original system
+load("mhi.RData") #median household income by DA
 ```
 
-Introduction
+Background
 ===================
 
-Public bicycle share programs (PBSP)
+Public bicycle share programs (PBSP) have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users [@fishmanBikeshareReviewRecent2016].
+In many cities, males use bike share more than females [@breyWantRideMy2017; @fishmanBikeshareReviewRecent2016] as do younger age cohorts [@@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011].
+One study found that bike share users in Washington, DC were more likely to be female [@buckAreBikeshareUsers2013], but the authors note that they relied on self-reported data from the annual member survey, which presents bias in the sample, and other travel survey data collected before the bike share began operations in 2010, which did not provide a comprehensive demographic profile.
+There is some evidence that bike share users are less likely to own a car [@buckAreBikeshareUsers2013; @reillyNoncyclistsFrequentCyclists2020], which suggests that bike share programs could increase transportation options.
+However, the relationship between income or education and bike share use is less clear.
+Neighbourhoods with more bikes from Seattle, WA's dockless bike share system had more college-educated residents, local community resources, and higher incomes [@mooneyFreedomStationSpatial2019].
+On the other hand, stations in disadvantaged communities in Chicago generated most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighbourhoods in Minneapolis-St.
+Paul used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019].
+Being university educated was also a significant correlate of bike share use in Montreal, Canada [@fullerUseNewPublic2011].
+On the other hand, members who reside in minority-concentrated and lower socioeconomic status neighbourhoods use the Minneapolis-St Paul bike share more frequently [@wangNeighborhoodSociodemographicCharacteristics2019].
+Financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016].
+Many studies have found that proximity to a bike share station, either living or working near one, is an important determinant of use or having a membership [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011].
+This makes sense given that individuals are more likely to use services or programs that they can easily access, which highlights the need for bike share systems to be highly accessible for diverse populations in order to increase use.
+The potential of bike share programs to increase cycling levels has been explored recently [@hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019], but more longitudinal evidence is needed to determine their impact on encouraging more cycling and offering more opportunities for physical activity.
+
+Several studies have recently explored the equity of PBSP in North American cities to understand who has access to bike share and where stations are located.
+Equity can be achieved in two different ways: equal spatial distribution across a region (e.g., horizontal equity) or greater access to vulnerable or disadvantaged populations (e.g., vertical equity) [@chenExploringEquityPerformance2019].
+Both are of interest to researchers and transport planners since they are often linked in that advantage, or conversely disadvantage, have spatial patterns.
+Using a negative binomial regression model, Qian and Jaller [-@qianBikesharingEquityDisadvantaged2020] estimated ridership in Chicago's PBSP among disadvantaged communities and found some disparities.
+While annual members in disadvantaged communities have a significantly lower share of trips compared to other areas in the city, they make longer trips.
+This suggests that they may be using PBSP for work commuting, which points to the importance of ensuring equitable access [@qianBikesharingEquityDisadvantaged2020].
+Similar results were found in Philadelphia, where lower income areas generate fewer trips and efforts to increase equity within the program have not been as successful as intended [@caspiBikesharingPhiladelphiaLowerincome2019].
+
+Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program, the location and size (e.g., maximum number of bicycles available) of stations matters.
+Unlike other public amenities with greater ability to adapt to crowding, like health care services, the utility of stations is limited by the number of bicycles that they can hold, which makes crowding effects critical; the system may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles.
+The location of stations is also important to increase the utility of bike share.
+Thus, spatial accessibility is important to consider when investigating equity issues, particularly for services or amenities that can increase transportation options such as PBSPs.
+
+Using floating catchment area methods, a novel approach that has not been used yet in cycling research, we examine accessibility to a public bicycle share program in Hamilton, Ontario.
+Our analysis builds upon a previous and recent study [@hosfordWhoArePublic2018], which found that disadvantaged areas in Hamilton are better served by the city's PBSP compared to other Canadian cities [i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau] with PBSPs where advantaged areas have greater access.
+Hosford and Winters [-@hosfordWhoArePublic2018] acknowledge that "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas".
+Therefore, the core service area for the PBSP in Hamilton is by default in more disadvantaged areas compared to other Canadian cities, but there is also a great deal of variation in income in the city center as well because of the local university and increasing gentrification.
+This paper explores the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities and investigates accessibility across the core service area.
+
+Case Study {#sec:study}
+======================
 
 This paper uses the city of Hamilton, located in Ontario, Canada, as a case study.
-The city launched the program, called SoBi Hamilton, in March 2015 with 115 stations and 750 bicycles [@Hamilton2020].
-The core service area spans 35 km^2 of the city and approximately 180,000 people live there [see Figure 1].
-The program was enthusiastically welcomed in the city - within three weeks of launching, 10,000 trips had been made.
+The city launched a public bicycle share program, called SoBi Hamilton, in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015].
+Stations are spaced between 300 and 600 metres apart.
+The core service area spans 35 km^2 of the city and _approximately 130,000 can reach a bike share hub within 30 minutes of walking_ [see Figure 1 where the core service area is outlined in blue].
+This represents roughly one fifth of the population in Hamilton, according to the 2016 Canadian Census.
+The program was enthusiastically welcomed in the city - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015].
 In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton.
@@
 The program also offers subsidized memberships to individuals who identify as low income.
-As of June 2020, the bike share program has 900 bikes and over 130 stations, and over 26,000 active memberships [@Hamilton2020].
+As of June 2020, the bike share program has 900 bikes and 130 stations, and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
+Hamilton Bike Share has conducted one membership survey to date in 2018 [@civicplanSoBiHamiltonMembership2017], and the findings from 420 members are broadly in line with the trends that we discussed above [see @fishmanBikeshareReviewRecent2016 for a recent review of the literature].
+The majority of respondents live within the core service area and the gender split is typical: 57% of respondents are male and 41% are female.
+The majority of respondents, both male and female, are between 25 and 34 years of age, but the percentage of male respondents is higher in the subsequent age groups.
+Respondents use bike share for commuting (40% of trips) or errands and meetings (24% of trips), and nearly 50% of trips have an average length of 11 to 20 minutes.
+As a result of having a bike share membership, 49% of respondents report that they use their private vehicle less often or much less often but 48% report that their private vehicle use has remained about the same.
+This suggests that SoBi Hamilton has been useful for certain kinds of trips but not all, meaning that some trips continue to require a private vehicle.
 
-```{r echo=FALSE}
+```{r, include=FALSE}
 sobi_service <- sobi_service %>%
@@
 
-```{r echo=FALSE}
-bounding_box <- st_bbox(sobi_service %>%
-                          st_buffer(500))
-```
-
-```{r echo=FALSE}
+```{r, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The core service area of SoBi Hamilton is outlined in blue.
+Dissemination areas in the Hamilton Census Metropolitan Area (CMA) are outlined in grey."}
 ggplot() + 
@@
 
-Methods
-===================
+```{r, include=FALSE}
+eri_hubs <- sobi_hubs %>% 
+             filter(OBJECTID>=117) %>% filter(OBJECTID<=128)
+```
+
+```{r figure-1, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The spatial distribution of PBSP stations in Hamilton, Ontario.
+The service area of the PBSP is outlined in blue, and dissemination areas within the Hamilton CMA that can reach a SoBi hub within 30 minutes of walking are outlined in grey."}
+ggplot() + 
+  geom_sf(fill = "antiquewhite1") +
+  geom_sf(data = hamilton_da_sobi, aes(color = "gray")) +
+  geom_sf(data = sobi_service, fill = NA, aes(color = "blue")) + 
+  geom_sf(data = sobi_hubs_original, size = 1, aes(color = "red")) +
+  geom_sf(data = eri_hubs, size = 1, aes(color = "green"))+
+  scale_color_identity(name = "Station Type",
+                       breaks = c("red", "green"),
+                       labels = c("Original Station", "Equity Station"),
+                       guide = "legend")
+```
+
+Methods {#sec:methods}
+======================
+
+## Floating Catchment Area
+
+Floating catchment area (FCA) methods are an approach commonly used in the accessibility literature.
+More recently, the _balanced_ floating catchment area approach was developed to address issues with demand inflation and supply inflation [@paezDemandLevelService2019].
+Balanced floating catchment area methods have been used to explore accessibility to health care providers [@paezDemandLevelService2019] and COVID-19 health care services [@pereiraGeographicAccessCOVID192021], but have not yet been used in the cycling literature to explore issues of accessibility.
+
+The first step is to allocate the population to be serviced by each PBSP station: 
+$$P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}$$
+In the second step, the level of service at each bicycle share station is the capacity at the location (i.e., the maximum number of bicycle racks) divided by its estimated service population:
+$$L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}$$
+
+Finally, accessibility is calculated:
+$$A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}$$
+
+Weights are then calculated:
+$$w_{ij} = f({c_{ij}})$$
+
+The approach replaces the weights in the calculations above, and uses instead a set of suitably normalized weights as follows:
+$${w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}$$
+and
+$${w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}$$
+
+These weights satisfy the following properties:
+$$\sum_{j = 1}^{J} {w^i_{ji}} = 1$$
+
+and
+$$\sum_{j = 1}^{J} {w^j_{ji}} = 1$$
+
+Finally, accessibility is calculated:
+$$A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}$$
+
+## Pycnophylactic Interpolation
+
+We first plotted the population in all dissemination areas within SoBi Hamilton's core service area.
+Since pycnophylactic interpolation occurs at such a micro scale, we had to ensure that population numbers were not interpolated or allocated to areas where people do not live in Hamilton (i.e., where schools or parks are located).
+To do so, we retrieved shapefiles for the following municipal data from Open Hamilton: Hamilton's public bicycle share program, the public bicycle share program core service area, golf courses, parks, designated large employment areas, municipal parking lots, cemeteries, environmentally sensitive areas, streets, educational institutions, places of worship, municipal service centres, recreation and community centres, arenas, emergency stations, fire stations, police stations, railways, and hospitals.
+Figure 3 shows where these spaces are located within SoBi Hamilton's core service area.
+Next, we extracted these geographic features from the PBSP core service area and used pycnophylactic interpolation methods to reallocate population to smaller polygons.
+
+## Travel Time Matrix
+
+[BBBike](https://download.bbbike.org/osm/bbbike/) is an online cycle route planner that interfaces with OpenStreetMap.
+We extracted OpenStreetMap data for SoBi Hamilton's service area to calculate walking times from each population cell to nearby bicycle share stations, using a walking distance of 10km and a walking time of 30 minutes as a threshold.
+A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bicycle share stations within the selected threshold.
+
+## Data
+
+All data for this research were accessed from Open Hamilton^[https://open.hamilton.ca/], an online repository of data curated by the City of Hamilton, in November 2020.
 
@@
 ===================
+
+```{r figure-2, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Population in all dissemination areas in the SoBi Hamilton core service area and within 30 minutes of walking to a bike share station."}
+ggplot() +
+  geom_sf(data = hamilton_da_sobi,
+          aes(fill = POP2011),
+          color = NA) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_fill_distiller(palette = "YlOrRd", direction = 1)
+```
+
+```{r figure-4, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Interpolated population cells in the SoBi Hamilton core service area (outlined in black) and within 30 minutes walk to the core service area.
+Geographic features have been extracted.
+Dissemination area polygons are outlined in grey."}
+ggplot() + 
+  geom_sf(data = population_sobi,
+          aes(color = population)) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  geom_sf(data = hamilton_da_sobi,
+          fill = NA,
+          color = "lightgray") + 
+  geom_sf(data = sobi_service,
+          fill = NA,
+          colour = "blue")
+```
+
+```{r figure-5, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Interpolated population units in the SoBi Hamilton core service area (outlined in black) and within 30 minutes walk to the core service area.
+Green circles represent the location of the equity stations that were added in 2018.
+Red circles represent the original stations.
+The service area of the PBSP is outlined in blue, and dissemination areas from the 2016 Canadian Census are outlined in light grey."}
+ggplot() + 
+  geom_sf(fill = "antiquewhite1") +
+  geom_sf(data = population_sobi, aes(color = population)) +
+  scale_color_distiller(palette = "YlOrRd", direction = 1) +
+  geom_sf(data = hamilton_da_sobi, fill = NA, color = "lightgray") + 
+  geom_sf(data = sobi_hubs_original, size = 1.5, shape = 21, fill = "red") +
+  geom_sf(data = eri_hubs, size = 1.5, shape = 21, fill = "green") +
+  geom_sf(data = sobi_service, fill = NA, colour = "blue")
+```
+
+## Accessibility Calculations at Different Thresholds
+
+We conduct a sensitivity analysis to calculate accessibility at different thresholds: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
+We did these calculations for both the current system with equity stations and the original system without equity stations to compare how accessibility changes at different thresholds with the addition of the equity stations.
+
+```{r echo=FALSE}
+# current system with equity stations
+ttm_walk <- left_join(ttm_walk, 
+                      population_sobi,
+                      by = "UID")
+```
+
+```{r echo=FALSE} 
+# original system without equity stations
+ttm_walk_o <- left_join(ttm_walk_o, 
+                      population_sobi_original,
+                      by = "UID")
+```
+
+```{r, echo=FALSE}
+ttm_walk_o <- ttm_walk_o[-c(5,7:8)]
+```
+
+```{r echo=FALSE}
+ttm_walk <- ttm_walk %>% dplyr::select(-geometry)
+
+ttm_walk_o <- ttm_walk_o %>% dplyr::select(-geometry)
+```
+
+```{r, include=FALSE}
+# current system with equity stations
+ttm_walk <- left_join(ttm_walk, sobi_hubs %>% dplyr::select(OBJECTID, RACKS_AMOU), by=c("OBJECTID" = "OBJECTID"))
+ttm_walk <- ttm_walk %>% dplyr::select(-geometry)
+head(ttm_walk)
+```
+
+```{r, include=FALSE}
+# original system without equity stations
+ttm_walk_o <- left_join(ttm_walk_o, sobi_hubs_original %>% dplyr::select(OBJECTID, RACKS_AMOU), by=c("OBJECTID" = "OBJECTID"))
+ttm_walk_o <- ttm_walk_o %>% dplyr::select(-geometry)
+head(ttm_walk_o)
+```
+
+```{r echo=FALSE}
+#data frames for current and original systems
+names(ttm_walk) <- c('UID', 'hub', 'travel_time', 'population', 'racks')
+
+names(ttm_walk_o) <- c('UID', 'hub', 'travel_time', 'population', 'racks')
+```
+
+```{r, echo=FALSE}
+population_sobi_original <- population_sobi_original[-c(3,5:6)]
+```
+
+### Threshold of 3 Minutes Walk
+
+```{r, echo=FALSE}
+#minimum threshold for accessibility calculations
+dist_threshold <- 3
+```
+
+```{r echo=FALSE}
+ttm_walk_min <- ttm_walk %>%
+  mutate(impedance_binary = ifelse(travel_time <= dist_threshold, 1, 0))
+
+ttm_walk_o_min <- ttm_walk_o %>%
+  mutate(impedance_binary = ifelse(travel_time <= dist_threshold, 1, 0))
+```
+
+```{r echo=FALSE}
+sum_b1_min <- ttm_walk_min %>%
+  group_by(UID) %>%
+  summarize(sum_b1_min = sum(impedance_binary),
+            .groups = "drop")
+
+sum_b1_o_min <- ttm_walk_o_min %>%
+  group_by(UID) %>%
+  summarize(sum_b1_o_min = sum(impedance_binary),
+            .groups = "drop")
+```
+
+```{r echo=FALSE}
+sum_b2_min <- ttm_walk_min %>%
+  group_by(hub) %>%
+  summarize(sum_b2_min = sum(impedance_binary))
+
+sum_b2_o_min <- ttm_walk_o_min %>%
+  group_by(hub) %>%
+  summarize(sum_b2_o_min = sum(impedance_binary))
+```
+
+```{r echo=FALSE}
+ttm_walk_min <- ttm_walk_min %>%
+  left_join(sum_b1_min, by = "UID") %>%
+  left_join(sum_b2_min, by = "hub")
+
+ttm_walk_o_min <- ttm_walk_o_min %>%
+  left_join(sum_b1_o_min, by = "UID") %>%
+  left_join(sum_b2_o_min, by = "hub")
+```
+
+```{r echo=FALSE}
+ttm_walk_min <- ttm_walk_min %>%
+  dplyr::filter(sum_b1_min > 0)
+
+ttm_walk_o_min <- ttm_walk_o_min %>%
+  dplyr::filter(sum_b1_o_min > 0)
+```
+
+```{r echo=FALSE}
+ttm_walk_min <- ttm_walk_min %>% 
+  dplyr::mutate(balanced_impedance_1 = impedance_binary/sum_b1_min, 
+         balanced_impedance_2 = impedance_binary/sum_b2_min)
+
+ttm_walk_o_min <- ttm_walk_o_min %>% 
+  dplyr::mutate(balanced_impedance_1 = impedance_binary/sum_b1_o_min, 
+         balanced_impedance_2 = impedance_binary/sum_b2_o_min)
+```
+
+```{r echo=FALSE}
+ttm_walk_min %>% 
+  dplyr::filter(hub == "46") %>%
+  pull(balanced_impedance_2) %>%
+  sum()
+
+ttm_walk_o_min %>% 
+  dplyr::filter(hub == "46") %>%
+  pull(balanced_impedance_2) %>%
+  sum()
+```
+
+```{r, include=FALSE}
+sobi_los_min <- ttm_walk_min %>%
+  dplyr::filter(impedance_binary > 0) %>% # filter impedance == 0 to avoid divisions by zero
+  group_by(hub) %>%
+  summarize(sobi_los_min = first(racks) / sum((population * balanced_impedance_1)),
+            .groups = "drop")
+
+sobi_los_o_min <- ttm_walk_o_min %>%
+  dplyr::filter(impedance_binary > 0) %>% # filter impedance == 0 to avoid divisions by zero
+  group_by(hub) %>%
+  summarize(sobi_los_o_min = first(racks) / sum((population * balanced_impedance_1)),
+            .groups = "drop")
+```
+
+```{r, include=FALSE}
+ttm_walk_min <- ttm_walk_min %>%
+  left_join(sobi_los_min, by = "hub")
+
+ttm_walk_o_min <- ttm_walk_o_min %>%
+  left_join(sobi_los_o_min, by = "hub")
+```
+
+```{r, include=FALSE}
+sobi_accessibility_min <- ttm_walk_min %>%
+  group_by(UID) %>%
+  summarize(accessibility_min = sum(sobi_los_min * balanced_impedance_2),
+            .groups = "drop")
+
+sobi_accessibility_o_min <- ttm_walk_o_min %>%
+  group_by(UID) %>%
+  summarize(accessibility_o_min = sum(sobi_los_o_min * balanced_impedance_2),
+            .groups = "drop")
+```
+
+```{r, echo=FALSE}
+#current system with equity stations
+sobi_accessibility_min %>% 
+  pull(accessibility_min) %>% 
+  sum()
+
+sobi_los_min %>% 
+  pull(sobi_los_min) %>%
+  sum()
+```
+
+```{r, echo=FALSE}
+#original system without equity stations
+sobi_accessibility_o_min %>% 
+  pull(accessibility_o_min) %>% 
+  sum()
+
+sobi_los_o_min %>% 
+  pull(sobi_los_o_min) %>%
+  sum()
+```
+
+```{r, include=FALSE}
+population_sobi <- population_sobi %>%
+  left_join(sobi_accessibility_min,
+            by = "UID")
+
+population_sobi_original <- population_sobi_original %>%
+  left_join(sobi_accessibility_o_min,
+            by = "UID")
+```
+
+```{r, echo=FALSE}
+#accessibility plot for current system with equity stations at 3 minutes walk
+ggplot(data = population_sobi,
+       fill = accessibility_min) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+```{r, echo=FALSE}
+#accessibility plot for original system without equity stations at 3 mins walk
+ggplot(data = population_sobi_original) +
+  geom_sf(aes(color = accessibility_o_min)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+```{r, echo=FALSE}
+#level of equitable service in current system at 3 mins walk
+population_sobi %>%
+  mutate(Lei_min = accessibility_min - sum(sobi_los_min$sobi_los_min)/nrow(population_sobi %>% drop_na(accessibility_min))) %>%
+  ggplot() +
+  geom_sf(aes(color = Lei_min)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
+```
+
+```{r, echo=FALSE}
+#level of equitable service in original system at 3 mins walk
+population_sobi_original %>%
+  mutate(Lei_min = accessibility_o_min - sum(sobi_los_o_min$sobi_los_o_min)/nrow(population_sobi_original %>% drop_na(accessibility_o_min))) %>%
+  ggplot() +
+  geom_sf(aes(color = Lei_min)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
+```
+
+```{r, echo=FALSE}
+sum(population_sobi$accessibility_min) # current system with equity
+sum(population_sobi_original$accessibility_o_min) # original system
+```
+
+```{r, echo=FALSE}
+population_sobi <- population_sobi %>% mutate(PPR_min = sum(accessibility_min/nrow(population_sobi)))
+
+# "equitable" share of the total level of service is NA due to division by zero
+```
+
+```{r, echo=FALSE}
+population_sobi_original <- population_sobi_original %>% mutate(PPR_min = sum(accessibility_o_min/nrow(population_sobi_original)))
+
+# "equitable" share of the total level of service is NA due to division by zero
+```
+
+```{r accessibility_disparities, include=FALSE}
+# Disparity in this case is defined as the difference between the calculated accessibility (which is interpreted as the local provider-to-population ratio) and the local provider-to-population ratio if the level of service was distributed uniformly
+
+population_sobi <- population_sobi %>% mutate(disparity_min = accessibility_min - PPR_min)
+
+population_sobi_original <- population_sobi_original %>% mutate(disparity_min = accessibility_o_min - PPR_min)
+```
+
+```{r, echo=FALSE}
+# disparity in current system
+
+population_sobi %>%
+  ggplot() +
+  geom_sf(aes(color = disparity_min)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
+```
+
+```{r, echo=FALSE}
+# disparity in original system
+
+population_sobi_original %>%
+  ggplot() +
+  geom_sf(aes(color = disparity_min)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
+```
+
+```{r, echo=FALSE}
+ggplot() +
+  geom_sf(data = population_sobi,
+          aes(color = accessibility_min)) +
+  geom_sf(data = hamilton_da_sobi,
+          fill = NA) + 
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+```{r, echo=FALSE}
+ggplot() +
+  geom_sf(data = population_sobi_original,
+          aes(color = accessibility_o_min)) +
+  geom_sf(data = hamilton_da_sobi,
+          fill = NA) + 
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+### Threshold of 5 Minutes Walk
+
+### Threshold of 10 Minutes Walk
+
+```{r, echo=FALSE}
+dist_threshold <- 10
+```
+
+```{r echo=FALSE}
+ttm_walk_max <- ttm_walk %>%
+  mutate(impedance_binary = ifelse(travel_time <= dist_threshold, 1, 0))
+
+ttm_walk_o_max <- ttm_walk_o %>%
+  mutate(impedance_binary = ifelse(travel_time <= dist_threshold, 1, 0))
+```
+
+```{r echo=FALSE}
+sum_b1 <- ttm_walk %>%
+  group_by(UID) %>%
+  summarize(sum_b1 = sum(impedance_binary),
+            .groups = "drop")
+
+sum_b1_o <- ttm_walk_o %>%
+  group_by(UID) %>%
+  summarize(sum_b1_o = sum(impedance_binary),
+            .groups = "drop")
+```
+
+```{r echo=FALSE}
+sum_b2 <- ttm_walk %>%
+  group_by(hub) %>%
+  summarize(sum_b2 = sum(impedance_binary))
+
+sum_b2_o <- ttm_walk_o %>%
+  group_by(hub) %>%
+  summarize(sum_b2_o = sum(impedance_binary))
+```
+
+```{r echo=FALSE}
+ttm_walk <- ttm_walk %>%
+  left_join(sum_b1, by = "UID") %>%
+  left_join(sum_b2, by = "hub")
+
+ttm_walk_o <- ttm_walk_o %>%
+  left_join(sum_b1_o, by = "UID") %>%
+  left_join(sum_b2_o, by = "hub")
+```
+
+```{r echo=FALSE}
+ttm_walk <- ttm_walk %>%
+  dplyr::filter(sum_b1 > 0)
+
+ttm_walk_o <- ttm_walk_o %>%
+  dplyr::filter(sum_b1_o > 0)
+```
+
+```{r echo=FALSE}
+ttm_walk <- ttm_walk %>% 
+  dplyr::mutate(balanced_impedance_1 = impedance_binary/sum_b1, 
+         balanced_impedance_2 = impedance_binary/sum_b2)
+
+ttm_walk_o <- ttm_walk_o %>% 
+  dplyr::mutate(balanced_impedance_1 = impedance_binary/sum_b1_o, 
+         balanced_impedance_2 = impedance_binary/sum_b2_o)
+```
+
+```{r echo=FALSE}
+ttm_walk %>% 
+  dplyr::filter(hub == "46") %>%
+  pull(balanced_impedance_2) %>%
+  sum()
+
+ttm_walk_o %>% 
+  dplyr::filter(hub == "46") %>%
+  pull(balanced_impedance_2) %>%
+  sum()
+```
+
+```{r, include=FALSE}
+sobi_los <- ttm_walk %>%
+  dplyr::filter(impedance_binary > 0) %>% # filter impedance == 0 to avoid divisions by zero
+  group_by(hub) %>%
+  summarize(sobi_los = first(racks) / sum((population * balanced_impedance_1)),
+            .groups = "drop")
+
+sobi_los_o <- ttm_walk_o %>%
+  dplyr::filter(impedance_binary > 0) %>% # filter impedance == 0 to avoid divisions by zero
+  group_by(hub) %>%
+  summarize(sobi_los_o = first(racks) / sum((population * balanced_impedance_1)),
+            .groups = "drop")
+```
+
+```{r, include=FALSE}
+ttm_walk <- ttm_walk %>%
+  left_join(sobi_los, by = "hub")
+
+ttm_walk_o <- ttm_walk_o %>%
+  left_join(sobi_los_o, by = "hub")
+```
+
+```{r, include=FALSE}
+sobi_accessibility <- ttm_walk %>%
+  group_by(UID) %>%
+  summarize(accessibility = sum(sobi_los * balanced_impedance_2),
+            .groups = "drop")
+
+sobi_accessibility_o <- ttm_walk_o %>%
+  group_by(UID) %>%
+  summarize(accessibility_o = sum(sobi_los_o * balanced_impedance_2),
+            .groups = "drop")
+```
+
+```{r, echo=FALSE}
+#current system with equity stations
+sobi_accessibility %>% 
+  pull(accessibility) %>% 
+  sum()
+
+sobi_los %>% 
+  pull(sobi_los) %>%
+  sum()
+```
+
+```{r, echo=FALSE}
+#original system without equity stations
+sobi_accessibility_o %>% 
+  pull(accessibility_o) %>% 
+  sum()
+
+sobi_los_o %>% 
+  pull(sobi_los_o) %>%
+  sum()
+```
+
+```{r, echo=FALSE}
+population_sobi_original <- population_sobi_original[-c(3,5:6)]
+```
+
+```{r, include=FALSE}
+population_sobi <- population_sobi %>%
+  left_join(sobi_accessibility,
+            by = "UID")
+
+population_sobi_original <- population_sobi_original %>%
+  left_join(sobi_accessibility_o,
+            by = "UID")
+```
+
+```{r, include=FALSE}
+population_sobi <- population_sobi %>% drop_na()
+population_sobi_original <- population_sobi_original %>% drop_na()
+```
+
+```{r, echo=FALSE}
+#accessibility plot for current system with equity stations
+ggplot(data = population_sobi) +
+  geom_sf(aes(color = accessibility)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+```{r, echo=FALSE}
+#accessibility plot for original system without equity stations
+ggplot(data = population_sobi_original) +
+  geom_sf(aes(color = accessibility_o)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+```{r, echo=FALSE}
+#level of equitable service in current system
+population_sobi %>%
+  mutate(Lei = accessibility - sum(sobi_los$sobi_los)/nrow(population_sobi %>% drop_na(accessibility))) %>%
+  ggplot() +
+  geom_sf(aes(color = Lei)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
+```
+
+```{r, echo=FALSE}
+#level of equitable service in original system
+population_sobi_original %>%
+  mutate(Lei = accessibility_o - sum(sobi_los_o$sobi_los_o)/nrow(population_sobi_original %>% drop_na(accessibility_o))) %>%
+  ggplot() +
+  geom_sf(aes(color = Lei)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
+```
+
+```{r, echo=FALSE}
+sum(population_sobi$accessibility) # current system with equity
+sum(population_sobi_original$accessibility_o) # original system
+```
+
+```{r, echo=FALSE}
+population_sobi <- population_sobi %>% mutate(PPR = sum(accessibility/nrow(population_sobi)))
+
+# "equitable" share of the total level of service in the current system is 0.0004864108 per population cell
+```
+
+```{r, echo=FALSE}
+population_sobi_original <- population_sobi_original %>% mutate(PPR = sum(accessibility_o/nrow(population_sobi_original)))
+
+# "equitable" share of the total level of service in the current system is 0.0005049342 per population cell
+```
+
+```{r accessibility_disparities, include=FALSE}
+# Disparity in this case is defined as the difference between the calculated accessibility (which is interpreted as the local provider-to-population ratio) and the local provider-to-population ratio if the level of service was distributed uniformly
+
+population_sobi <- population_sobi %>% mutate(disparity = accessibility - PPR)
+
+population_sobi_original <- population_sobi_original %>% mutate(disparity = accessibility_o - PPR)
+```
+
+```{r, echo=FALSE}
+# disparity in current system
+
+population_sobi %>%
+  ggplot() +
+  geom_sf(aes(color = disparity)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
+```
+
+```{r, echo=FALSE}
+# disparity in original system
+
+population_sobi_original %>%
+  ggplot() +
+  geom_sf(aes(color = disparity)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_gradient2()
+```
+
+```{r, echo=FALSE}
+ggplot() +
+  geom_sf(data = population_sobi,
+          aes(color = accessibility)) +
+  geom_sf(data = hamilton_da_sobi,
+          fill = NA) + 
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+```{r, echo=FALSE}
+ggplot() +
+  geom_sf(data = population_sobi_original,
+          aes(color = accessibility_o)) +
+  geom_sf(data = hamilton_da_sobi,
+          fill = NA) + 
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+```{r, include=FALSE}
+accessibility_da <- population_sobi %>%
+  st_join(hamilton_da_sobi %>% dplyr::select(DA))
+
+accessibility_da_original <- population_sobi_original %>%
+  st_join(hamilton_da_sobi %>% dplyr::select(DA))
+```
+
+```{r, include=FALSE}
+accessibility_da <- accessibility_da %>%
+  group_by(DA) %>%
+  summarize(population = sum(population),
+            accessibility = sum(accessibility),
+            .groups = "drop")
+
+accessibility_da_original <- accessibility_da_original %>%
+  group_by(DA) %>%
+  summarize(population = sum(population),
+            accessibility_o = sum(accessibility_o),
+            .groups = "drop")
+```
+
+```{r, include=FALSE}
+accessibility_da <- accessibility_da %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_sobi %>% 
+              dplyr::select(DA, geometry),
+            by = "DA") %>%
+  st_as_sf()
+
+accessibility_da_original <- accessibility_da_original %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_sobi %>% 
+              dplyr::select(DA, geometry),
+            by = "DA") %>%
+  st_as_sf()
+```
+
+```{r, echo=FALSE}
+#accessibility plot by DA in current system
+
+ggplot() +
+  geom_sf(data = accessibility_da,
+          aes(fill = accessibility),
+          color = "black") +
+  scale_fill_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+```{r, echo=FALSE}
+#accessibility plot by DA in original system
+
+ggplot() +
+  geom_sf(data = accessibility_da_original,
+          aes(fill = accessibility_o),
+          color = "black") +
+  scale_fill_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+```{r, include=FALSE}
+accessibility_da <- accessibility_da %>%
+  left_join(accessibility_da_original %>% 
+              st_drop_geometry() %>%
+              rename(population_o = population),
+              dplyr::select(DA, accessibility_o, population_o),
+            by = "DA") %>%
+  mutate(accessibility_o = replace_na(accessibility_o, 0), # The population serviced grows with the new stations, and as a result some original population is zero
+         population_o = replace_na(population_o, 0)) # Same for accessibility
+
+```
+
+```{r, include=FALSE}
+accessibility_da <- accessibility_da %>%
+  mutate(R_a = accessibility - accessibility_o)
+```
+
+```{r, echo=FALSE}
+#disparity without equity stations
+ggplot() +
+  geom_sf(data = accessibility_da,
+          aes(fill = R_a),
+          color = "black") +
+  scale_fill_distiller(palette = "YlOrRd", 
+                        direction = 1)
+```
+
+```{r, include=FALSE}
+accessibility_da <- accessibility_da %>%
+  left_join(mhi %>%
+              transmute(DA = GeoUID,
+                        median_household_income = `v_CA16_2397: Median total income of households in 2015 ($)`),
+            by = "DA")
+```
+
+```{r, include=FALSE}
+data <- bi_class(accessibility_da %>%
+                   drop_na(),
+                 x = accessibility_o, 
+                 y = median_household_income, 
+                 style = "quantile", 
+                 dim = 3) %>%
+  st_as_sf()
+```
+
+```{r, echo=FALSE}
+#accessibility of original system at 10 mins walk
+ map <- ggplot() +
+  geom_sf(data = data, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", dim = 3) +
+  bi_theme()
+
+map
+```
+
+```{r, include=FALSE}
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+```
+
+```{r, echo=FALSE}
+#accessibility of original system at 10 mins walk 
+ cowplot::ggdraw() +
+ cowplot::draw_plot(map, 0, 0, 1, 1) +
+ cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r, include=FALSE}
+data_equity <- bi_class(accessibility_da %>%
+                   drop_na(),
+                 x = accessibility, 
+                 y = median_household_income, 
+                 style = "quantile", 
+                 dim = 3) %>%
+  st_as_sf()
+```
+
+```{r, echo=FALSE}
+#accessibility of current system at 10 mins walk
+ map_equity <- ggplot() +
+  geom_sf(data = data_equity, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", dim = 3) +
+  bi_theme()
+
+map_equity
+```
+
+```{r, include=FALSE}
+legend_equity <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+```
+
+```{r, echo=FALSE}
+#accessibility of current system at 10 mins walk
+ cowplot::ggdraw() +
+ cowplot::draw_plot(map_equity, 0, 0, 1, 1) +
+ cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+### Threshold of 15 Minutes Walk
+
+## Comparison
+
+Include table with total # of people who can reach a station within 3 minutes walk, 5 minutes, 10 minutes, and 15 minutes, to compare accessibility at different thresholds for both current and original system.
 

### 2021-03-12 examples/Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Changes to draft manuscript

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{Transportation Research Part D: Transport and Environment} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc citation processing

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{Examining spatial equity and accessibility to a public bicycle share
program using a balanced floating catchment area approach}
    \author[Some School]{Antonio Paez\corref{1}}
   \ead{paez@mcmaster.ca} 
    \author[Some School]{Elise Desjardins\corref{2}}
   \ead{desjae@mcmaster.ca} 
    \author[Another University]{Christopher D. Higgins\corref{2}}
   \ead{cd.higgins@utoronto.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, 1280 Main Street West,
Hamilton, ON}
    \address[University of Toronto Scarborough]{Department, Street, City, State, Zip}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  This is the abstract.

  It consists of two paragraphs.
  \end{abstract}

 \end{frontmatter}

\begin{verbatim}
## 
## Attaching package: 'dplyr'
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:stats':
## 
##     filter, lag
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:base':
## 
##     intersect, setdiff, setequal, union
\end{verbatim}

\begin{verbatim}
## Loading required package: purrr
\end{verbatim}

\begin{verbatim}
## Registered S3 method overwritten by 'pryr':
##   method      from
##   print.bytes Rcpp
\end{verbatim}

\begin{verbatim}
## 
## ## Message from disk.frame:
## We have 1 workers to use with disk.frame.
## To change that, use setup_disk.frame(workers = n) or just setup_disk.frame() to use the defaults.
## 
## 
## It is recommended that you run the following immediately to set up disk.frame with multiple workers in order to parallelize your operations:
## 
## 
## ```r
## # this will set up disk.frame with multiple workers
## setup_disk.frame()
## # this will allow unlimited amount of data to be passed from worker to worker
## options(future.globals.maxSize = Inf)
## ```
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'disk.frame'
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:purrr':
## 
##     imap, imap_dfr, map, map2
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:base':
## 
##     colnames, ncol, nrow
\end{verbatim}

\begin{verbatim}
## Loading required package: sp
\end{verbatim}

\begin{verbatim}
## Loading required package: maptools
\end{verbatim}

\begin{verbatim}
## Checking rgeos availability: TRUE
\end{verbatim}

\begin{verbatim}
## Loading required package: rgeos
\end{verbatim}

\begin{verbatim}
## rgeos version: 0.5-5, (SVN revision 640)
##  GEOS runtime version: 3.8.0-CAPI-1.13.1 
##  Linking to sp version: 1.4-4 
##  Polygon checking: TRUE
\end{verbatim}

\begin{verbatim}
## Loading required namespace: sf
\end{verbatim}

\begin{verbatim}
## Loading required namespace: rJava
\end{verbatim}

\begin{verbatim}
## Please make sure you have already allocated some memory to Java by running:
##   options(java.parameters = '-Xmx2G')
\end{verbatim}

\begin{verbatim}
## Linking to GEOS 3.8.0, GDAL 3.0.4, PROJ 6.3.1
\end{verbatim}

\begin{verbatim}
## -- Attaching packages --------------------------------------- tidyverse 1.3.0 --
\end{verbatim}

\begin{verbatim}
## v ggplot2 3.3.2     v stringr 1.4.0
## v tibble  3.0.4     v forcats 0.5.0
## v tidyr   1.1.2
\end{verbatim}

\begin{verbatim}
## -- Conflicts ------------------------------------------ tidyverse_conflicts() --
## x dplyr::filter()        masks stats::filter()
## x disk.frame::imap()     masks purrr::imap()
## x disk.frame::imap_dfr() masks purrr::imap_dfr()
## x dplyr::lag()           masks stats::lag()
## x disk.frame::map()      masks purrr::map()
## x disk.frame::map2()     masks purrr::map2()
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'data.table'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:purrr':
## 
##     transpose
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:dplyr':
## 
##     between, first, last
\end{verbatim}

\begin{verbatim}
## Loading required package: raster
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'raster'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:data.table':
## 
##     shift
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:tidyr':
## 
##     extract
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:dplyr':
## 
##     select
\end{verbatim}

\begin{verbatim}
## Loading required package: igraph
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'igraph'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:raster':
## 
##     union
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:tidyr':
## 
##     crossing
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:tibble':
## 
##     as_data_frame
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:rgeos':
## 
##     union
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:purrr':
## 
##     compose, simplify
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:dplyr':
## 
##     as_data_frame, groups, union
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:stats':
## 
##     decompose, spectrum
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:base':
## 
##     union
\end{verbatim}

\begin{verbatim}
## Loading required package: Matrix
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'Matrix'
\end{verbatim}

\begin{verbatim}
## The following objects are masked from 'package:tidyr':
## 
##     expand, pack, unpack
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'gdistance'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:igraph':
## 
##     normalize
\end{verbatim}

\begin{verbatim}
## 
## Attaching package: 'kableExtra'
\end{verbatim}

\begin{verbatim}
## The following object is masked from 'package:dplyr':
## 
##     group_rows
\end{verbatim}

\begin{verbatim}
## rgdal: version: 1.5-18, (SVN revision 1082)
## Geospatial Data Abstraction Library extensions to R successfully loaded
## Loaded GDAL runtime: GDAL 3.0.4, released 2020/01/28
## Path to GDAL shared files: C:/Users/desjae/Documents/R/win-library/4.0/rgdal/gdal
## GDAL binary built with GEOS: TRUE 
## Loaded PROJ runtime: Rel. 6.3.1, February 10th, 2020, [PJ_VERSION: 631]
## Path to PROJ shared files: C:/Users/desjae/Documents/R/win-library/4.0/rgdal/proj
## Linking to sp version:1.4-4
## To mute warnings of possible GDAL/OSR exportToProj4() degradation,
## use options("rgdal_show_exportToProj4_warnings"="none") before loading rgdal.
\end{verbatim}

\begin{verbatim}
## udunits system database from C:/Users/desjae/Documents/R/win-library/4.0/units/share/udunits
\end{verbatim}

\hypertarget{background}{%
\section{Background}\label{background}}

Public bicycle share programs (PBSP) have been implemented in over 800
cities worldwide and a great deal has been learned about their typical
users ({\textbf{???}}). Males use bike share more than females
({\textbf{???}}) as do younger age cohorts ({\textbf{???}};
{\textbf{???}}). One study found that bike share users in Washington, DC
were more likely to be female ({\textbf{???}}), but the authors note
that they relied on self-reported data from the annual member survey,
which presents bias in the sample, and other travel survey data
collected before the bike share began operations in 2010, which did not
provide a comprehensive demographic profile. There is some evidence that
bike share users are less likely to own a car ({\textbf{???}};
{\textbf{???}}), which suggests that bike share programs could increase
transportation options. However, the relationship between income or
education and bike share use is not fully understood. Neighbourhoods
with more bikes from Seattle, WA's dockless bike share system had more
college-educated residents, local community resources, and higher
incomes ({\textbf{???}}). Being university educated was also a
significant correlate of bike share use in Montreal, Canada
({\textbf{???}}). On the other hand, members who reside in
minority-concentrated and lower socioeconomic status neighbourhoods use
the Minneapolis-St Paul bike share more frequently ({\textbf{???}}).
Financial savings have been found to motivate those on a low income to
use bike share ({\textbf{???}}). Many studies have found that proximity
to a bike share station, either living or working near one, is an
important determinant of use or having a membership ({\textbf{???}};
{\textbf{???}}). This makes sense given that individuals are more likely
to use services or programs that they can easily access, which
highlights the need for bike share systems to be highly accessible for
diverse populations in order to increase use. The potential of bike
share programs to increase cycling levels has been explored recently
({\textbf{???}}; {\textbf{???}}), but more longitudinal evidence is
needed to determine their impact on encouraging more cycling and
offering more opportunities for physical activity.

This paper uses the city of Hamilton, located in Ontario, Canada, as a
case study. The city launched the program, called SoBi Hamilton, in
March 2015 with 115 stations and 750 bicycles ({\textbf{???}}). The core
service area spans 35 km\^{}2 of the city and \emph{approximately
130,000 can reach a bike share hub within 30 minutes of walking} {[}see
Figure 1 where the core service area is outlined in blue{]}. This
represents roughly one fifth of the population in Hamilton, according to
the 2016 Canadian Census. The program was enthusiastically welcomed in
the city - within three weeks of launching, 10,000 trips had been made
({\textbf{???}}). In 2017, Hamilton Bike Share Inc., the non-profit
organization that operates the program, initiated an equity program,
Everyone Rides Initiative (ERI), to remove barriers that may prevent
individuals from accessing bike share in Hamilton. To increase spatial
equity, an additional 75 bicycles and 15 stations were added to the
system, which expanded it to more disadvantaged areas in the city {[}see
Figure 2{]}. The program also offers subsidized memberships to
individuals who identify as low income. As of June 2020, the bike share
program has 900 bikes and over 130 stations, and over 26,000 active
memberships ({\textbf{???}}). Hamilton Bike Share has conducted one
membership survey to date in 2018 ({\textbf{???}}), and the findings
from 420 members are broadly in line with the trends that we discussed
above (see {\textbf{???}} for a recent review of the literature). The
majority of respondents live within the core service area and the gender
split is typical: 57\% of respondents are male and 41\% are female. The
majority of respondents, both male and female, are between 25 and 34
years, but the percentage of male respondents is higher in the
subsequent age groups. Respondents use bike share for commuting (40\% of
trips) or errands and meetings (24\% of trips), and nearly 50\% of trips
have an average length of 11 to 20 minutes. As a result of having a bike
share membership, 49\% of respondents report that they use their private
vehicle less often or much less often and 48\% report that their private
vehicle use has remained about the same.

Using floating catchment area methods, an approach recently used to
explore accessibility to family physicians in the Hamilton Census
Metropolitan Area (see {\textbf{???}}), we examine accessibility to
Hamilton's public bicycle share program. Our analysis builds upon a
previous and recent study ({\textbf{???}}), which found that
disadvantaged areas in Hamilton are better served by the SoBi Hamilton
compared to other Canadian cities {[}i.e., Toronto, Vancouver, Montreal,
and Ottawa-Gatineau{]} with PBSPs where advantaged areas have greater
access. Hosford and Winters ({\textbf{???}}) acknowledge that ``Hamilton
stands out in that the lower income neighborhoods are located near the
city center and wealthier neighborhoods are in the surrounding suburban
areas''. Therefore, the core service area for the bike share program is
by default in more disadvantaged areas compared to other Canadian
cities, but there is also a great deal of variation in income in the
city center because of the local university and increasing
gentrification. This paper explores the contribution of the ERI hubs to
reducing spatial inequity in the core service area and investigates
accessibility across the core service area.

\includegraphics{Bike-share-spatial-equity_files/figure-latex/unnamed-chunk-4-1.pdf}

\includegraphics{Bike-share-spatial-equity_files/figure-latex/unnamed-chunk-6-1.pdf}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{results}{%
\section{Results}\label{results}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Accessibility (maps, descriptive statistics)
\item
  Comparison with and without ERI hubs (change in accessibility)
\item
\end{enumerate}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Program achieved its stated goal to increase accessibility, only
modestly. It's an open question whether it changed behaviours.

Equity can be achieved in two different ways: equal spatial distribution
across a region (e.g., horizontal equity) or greater access to
vulnerable or disadvantaged populations (e.g., vertical equity)
({\textbf{???}}). By implementing ERI hubs in more disadvantaged areas
in Hamilton, the City of Hamilton has achieved greater horizontal equity
by extending the spatial distribution of bicycles across the city. But
our analysis found that vulnerable or disadvantaged populations do not
necessarily have greater access to the program.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\end{document}

### 2021-03-12 examples/Toy-example-balanced-accessibility.Rmd
Update accessibility calculations

@@
 
-dist_threshold <- quantile(as.vector(ttm_walk$travel_time), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))[[5]]
+# dist_threshold <- quantile(as.vector(ttm_walk$travel_time), probs = c(0, 1/10, 3/10, 5/10, 7/10, 8/10, 9/10))[[5]]
+
+# distance threshold has to be less than 30, maximum distance people are willing to go is distance threshold
+
+dist_threshold <- 10
 ```
@@
 
-Our calculations assume that people do not go further than 24 min to reach a SoBi hub.
+Our calculations assume that 70% of population cells have at least one station within 24 min to reach a SoBi station.
+We set the threshold (what people are willing to travel).
+Gaussian says people are more willing to go 5 than 10, less willing to go 15 than 20.
+Progressively discount, but we don't know the willingness.
+If we had more information about how many people go 5, 10, etc.
+then using the Gaussian function would be possible.
+Otherwise, there are a lot of assumptions.
+Binary requires us to make fewer assumptions.
 
@@
 Calculate the sum of the impedance by SoBi hubs for the balancing factors.
-In the binary case, this should be the same as the number of interpolated population unit that each SoBi hub serves; when the impedance is not binary, this is the sum of all impedance values:
+In the binary case, this should be the same as the number of interpolated population cells that each SoBi hub serves; when the impedance is not binary, this is the sum of all impedance values:
 ```{r}
@@
 
-Hub 1 serves 1502 interpolated population units, hub 2 serves 1393 population units, etc.
+Hub 1 serves 1502 population cells, hub 2 serves 1393 population cells, etc.
 
@@
 
-Same thing for the other accessibility estimates. Spatial join:
+Same thing for the other accessibility estimates for the system with ERI hubs. Spatial join:
 ```{r}
@@
 ```
-
-Ok!
 
@@
 Retrieve data:
-```{r eval=FALSE}
+```{r eval}
 mhi <- get_census(dataset='CA16', 
@@
 
-Plot:
+Plot accessibility of original system without ERI hubs:
 ```{r}
@@
 
-Finalize plot:
+Finalize plot of accessibility with ERI hubs:
 ```{r}
@@
 ```
+
+Create a class for bivariate plotting (drop DA with missing income data):
+```{r}
+data_equity <- bi_class(accessibility_da %>%
+                   drop_na(),
+                 x = accessibility, 
+                 y = median_household_income, 
+                 style = "quantile", 
+                 dim = 3) %>%
+  st_as_sf()
+```
+
+Plot accessibility of original system with ERI hubs:
+```{r}
+ map_equity <- ggplot() +
+  geom_sf(data = data_equity, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", dim = 3) +
+  bi_theme()
+
+map_equity
+```
+
+Create legend:
+```{r}
+legend_equity <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+```
+
+Finalize plot of accessibility with ERI hubs:
+```{r}
+ cowplot::ggdraw() +
+  cowplot::draw_plot(map_equity, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+Save data files:
+```{r}
+save(accessibility_da, file = "accessibility.RData")
+save(map, file = "accessibility_map_original.RData")
+save(map_equity, file = "accessibility_map_equity.RData")
+save(mhi, file = "mhi.RData")
+save(population_sobi_original, file = "population_sobi_original.RData")
+```

### 2021-03-13 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Major reorganization of repository to clean up files and to create (mostly data) package.

--- 2021-03-12 examples/Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
@@
 
-```{r, include=FALSE}
-rm(list = ls())
-```
-
-```{r, include=FALSE}
+```{r load-packages, include=FALSE}
 library(data.table)
@@
 library(sf)
+library(sobiEquity)
 library(tidyverse)
 library(units)
-library(disk.frame)
-library(pycno)
+#library(disk.frame)
+#library(pycno)
 library(readr)
-library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
+#library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
 library(tinytex)
@@
 
-```{r, include=FALSE}
-equity_stations <- data.frame(longitude = c(	
--79.84364993846138, -79.8362839215856, -79.83144521746212, -79.82217214949738, -79.8160211746482, -79.8192304372373, -79.82500255113169, -79.8268599805827, -79.83888834759689, -79.82740849222654, -79.82342541216097, -79.84240606419081), latitude = c(43.265340158000654, 43.25632467284136, 43.25528943845642, 43.25291050074978, 43.25147804403345, 43.24444659758296, 43.2421627336722, 43.24634382671116, 43.249432741404554, 43.251126444296524, 43.24892849896579, 43.25831920064561))
+```{r}
+data("hamilton_cma")
+data("hamilton_da_2016")
+data("population_50x50")
+data("sobi_service")
+data("sobi_hubs")
+data("ttm_walk")
+```
+
+```{r coordinates-of-equity-stations, include=FALSE}
+# define the coordinates of the equity stations
+# equity_stations <- data.frame(longitude = c(-79.84364993846138, 
+#                                             -79.8362839215856,
+#                                             -79.83144521746212, 
+#                                             -79.82217214949738, 
+#                                             -79.8160211746482, 
+#                                             -79.8192304372373, 
+#                                             -79.82500255113169, 
+#                                             -79.8268599805827, 
+#                                             -79.83888834759689, 
+#                                             -79.82740849222654, 
+#                                             -79.82342541216097, 
+#                                             -79.84240606419081), 
+#                               latitude = c(43.265340158000654, 
+#                                            43.25632467284136, 
+#                                            43.25528943845642, 
+#                                            43.25291050074978, 
+#                                            43.25147804403345, 
+#                                            43.24444659758296, 
+#                                            43.2421627336722, 
+#                                            43.24634382671116,
+#                                            43.249432741404554,
+#                                            43.251126444296524, 
+#                                            43.24892849896579, 
+#                                            43.25831920064561))
 ```
@@
 
-```{r, include=FALSE}
-load("hamilton_cma.RData")
-load("hamilton_da_2016.RData")
-load("population_sobi.RData") #interpolated population in SoBi area
-load("sobi_hubs.RData") #location of SoBi hubs, minus Van Wagner's beach
-load("sobi_hubs_original.RData") #SoBi hubs, minus ERI hubs
-load("sobi_service.RData") #core service area for SoBi
-load("ttm_walk.RData") #travel time matrix for interpolated population in SoBi area
-load("ttm_walk_original.RData") #travel time matrix for system without ERI hubs
-load("accessibility.RData") #full accessibility table for SoBi system
-load("accessibility_map_original.RData") #map of accessibility in original SoBi system
-load("accessibility_map_equity.RData") #map of accessibility in current system with ERI
-load("hamilton_da_sobi.RData")
-load("population_sobi_original.RData") #population in original system
-load("mhi.RData") #median household income by DA
+```{r load-data, include=FALSE}
+#load("accessibility.RData") #full accessibility table for SoBi system
+#load("accessibility_map_original.RData") #map of accessibility in original SoBi system
+#load("accessibility_map_equity.RData") #map of accessibility in current system with ERI
+# load("hamilton_cma.RData")
+# load("hamilton_da_2016.RData")
+# load("hamilton_da_sobi.RData")
+# load("mhi.RData") #median household income by DA
+# load("population_sobi.RData") #interpolated population in SoBi area
+#load("population_sobi_original.RData") #population in original system
+# load("sobi_hubs.RData") #location of SoBi hubs, minus Van Wagner's beach
+# load("sobi_hubs_original.RData") #SoBi hubs, minus ERI hubs
+# load("sobi_service.RData") #core service area for SoBi
+# load("ttm_walk.RData") #travel time matrix for interpolated population in SoBi area
+# load("ttm_walk_original.RData") #travel time matrix for system without ERI hubs
 ```
@@
 
-```{r, include=FALSE}
-sobi_service <- sobi_service %>%
-  dplyr::filter(AREA_NAME == "Core Service Area")
-```
-
-```{r, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The core service area of SoBi Hamilton is outlined in blue.
+```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The core service area of SoBi Hamilton is outlined in blue.
 Dissemination areas in the Hamilton Census Metropolitan Area (CMA) are outlined in grey."}
@@
 
-```{r, include=FALSE}
-eri_hubs <- sobi_hubs %>% 
-             filter(OBJECTID>=117) %>% filter(OBJECTID<=128)
-```
-
-```{r figure-1, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The spatial distribution of PBSP stations in Hamilton, Ontario.
-The service area of the PBSP is outlined in blue, and dissemination areas within the Hamilton CMA that can reach a SoBi hub within 30 minutes of walking are outlined in grey."}
-ggplot() + 
-  geom_sf(fill = "antiquewhite1") +
-  geom_sf(data = hamilton_da_sobi, aes(color = "gray")) +
-  geom_sf(data = sobi_service, fill = NA, aes(color = "blue")) + 
-  geom_sf(data = sobi_hubs_original, size = 1, aes(color = "red")) +
-  geom_sf(data = eri_hubs, size = 1, aes(color = "green"))+
-  scale_color_identity(name = "Station Type",
-                       breaks = c("red", "green"),
-                       labels = c("Original Station", "Equity Station"),
-                       guide = "legend")
+```{r filter-eri-hubs, include=FALSE}
+# eri_hubs <- sobi_hubs %>% 
+#              filter(OBJECTID>=117 & OBJECTID<=128)
+```
+
+```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The spatial distribution of PBSP stations in Hamilton, Ontario.
+The service area of the PBSP is outlined in blue, and dissemination areas within the Hamilton CMA that touch the bounding box of the PBSP's service area are outlined in grey."}
+ggplot() + 
+  #geom_sf(fill = "antiquewhite1") +
+  geom_sf(data = hamilton_da_2016 %>% 
+            dplyr::filter(sobi_service == "Yes"),
+          color = "gray") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") + 
+  geom_sf(data = sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "red") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          size = 1.5, shape = 21, 
+          fill = "green")
 ```
@@
 The first step is to allocate the population to be serviced by each PBSP station: 
-$$P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}$$
+$$
+P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}
+$$
 In the second step, the level of service at each bicycle share station is the capacity at the location (i.e., the maximum number of bicycle racks) divided by its estimated service population:
-$$L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}$$
+$$
+L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
+$$
 
 Finally, accessibility is calculated:
-$$A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}$$
+$$
+A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
+$$
 
 Weights are then calculated:
-$$w_{ij} = f({c_{ij}})$$
+$$
+w_{ij} = f({c_{ij}})
+$$
 
 The approach replaces the weights in the calculations above, and uses instead a set of suitably normalized weights as follows:
-$${w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}$$
+$$
+{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
+$$
 and
-$${w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}$$
+$$
+{w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
+$$
 
 These weights satisfy the following properties:
-$$\sum_{j = 1}^{J} {w^i_{ji}} = 1$$
+$$
+\sum_{j = 1}^{J} {w^i_{ji}} = 1
+$$
 
 and
-$$\sum_{j = 1}^{J} {w^j_{ji}} = 1$$
+$$
+\sum_{j = 1}^{J} {w^j_{ji}} = 1
+$$
 
 Finally, accessibility is calculated:
-$$A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}$$
+$$
+A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}
+$$
 
@@
 ggplot() +
-  geom_sf(data = hamilton_da_sobi,
-          aes(fill = POP2011),
+  geom_sf(data = hamilton_da_2016 %>%
+            dplyr::filter(sobi_service == "Yes"),
+          aes(fill = population),
           color = NA) +
@@
           fill = NA) +
-  scale_fill_distiller(palette = "YlOrRd", direction = 1)
+  scale_fill_distiller(palette = "YlOrRd", 
+                       direction = 1)
 ```
@@
 ggplot() + 
-  geom_sf(data = population_sobi,
+  geom_sf(data = population_50x50,
           aes(color = population)) +
@@
                         direction = 1) +
-  geom_sf(data = hamilton_da_sobi,
+  geom_sf(data = hamilton_da_2016 %>%
+            filter(sobi_service == "Yes"),
           fill = NA,
@@
 ggplot() + 
-  geom_sf(fill = "antiquewhite1") +
-  geom_sf(data = population_sobi, aes(color = population)) +
-  scale_color_distiller(palette = "YlOrRd", direction = 1) +
-  geom_sf(data = hamilton_da_sobi, fill = NA, color = "lightgray") + 
-  geom_sf(data = sobi_hubs_original, size = 1.5, shape = 21, fill = "red") +
-  geom_sf(data = eri_hubs, size = 1.5, shape = 21, fill = "green") +
-  geom_sf(data = sobi_service, fill = NA, colour = "blue")
+  #geom_sf(fill = "antiquewhite1") +
+  geom_sf(data = population_50x50, 
+          aes(color = population)) +
+  geom_sf(data = hamilton_da_2016 %>%
+            dplyr::filter(sobi_service == "Yes"), 
+          fill = NA, 
+          color = "lightgray") + 
+  geom_sf(data = sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "red") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          size = 1.5, shape = 21, 
+          fill = "green") +
+  geom_sf(data = sobi_service, 
+          fill = NA, 
+          colour = "blue") +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1)
 ```
@@
 
-```{r echo=FALSE}
+```{r join-population-to-travel-time-data, echo=FALSE}
 # current system with equity stations
 ttm_walk <- left_join(ttm_walk, 
-                      population_sobi,
+                      population_50x50 %>%
+                        st_drop_geometry(),
                       by = "UID")
@@
 
-```{r echo=FALSE} 
-# original system without equity stations
-ttm_walk_o <- left_join(ttm_walk_o, 
-                      population_sobi_original,
-                      by = "UID")
-```
-
-```{r, echo=FALSE}
-ttm_walk_o <- ttm_walk_o[-c(5,7:8)]
-```
-
-```{r echo=FALSE}
-ttm_walk <- ttm_walk %>% dplyr::select(-geometry)
-
-ttm_walk_o <- ttm_walk_o %>% dplyr::select(-geometry)
-```
-
-```{r, include=FALSE}
+```{r join-sobi-station-to-travel-time-data, include=FALSE}
 # current system with equity stations
-ttm_walk <- left_join(ttm_walk, sobi_hubs %>% dplyr::select(OBJECTID, RACKS_AMOU), by=c("OBJECTID" = "OBJECTID"))
-ttm_walk <- ttm_walk %>% dplyr::select(-geometry)
-head(ttm_walk)
-```
-
-```{r, include=FALSE}
-# original system without equity stations
-ttm_walk_o <- left_join(ttm_walk_o, sobi_hubs_original %>% dplyr::select(OBJECTID, RACKS_AMOU), by=c("OBJECTID" = "OBJECTID"))
-ttm_walk_o <- ttm_walk_o %>% dplyr::select(-geometry)
-head(ttm_walk_o)
+ttm_walk <- ttm_walk %>% 
+  left_join(sobi_hubs %>%
+              st_drop_geometry() %>%
+              dplyr::select(OBJECTID, 
+                            RACKS_AMOU),
+            by = c("OBJECTID" = "OBJECTID"))
 ```
@@
 #data frames for current and original systems
-names(ttm_walk) <- c('UID', 'hub', 'travel_time', 'population', 'racks')
-
-names(ttm_walk_o) <- c('UID', 'hub', 'travel_time', 'population', 'racks')
-```
-
-```{r, echo=FALSE}
-population_sobi_original <- population_sobi_original[-c(3,5:6)]
-```
-
-### Threshold of 3 Minutes Walk
-
-```{r, echo=FALSE}
-#minimum threshold for accessibility calculations
-dist_threshold <- 3
-```
-
-```{r echo=FALSE}
-ttm_walk_min <- ttm_walk %>%
-  mutate(impedance_binary = ifelse(travel_time <= dist_threshold, 1, 0))
-
-ttm_walk_o_min <- ttm_walk_o %>%
-  mutate(impedance_binary = ifelse(travel_time <= dist_threshold, 1, 0))
-```
-
-```{r echo=FALSE}
-sum_b1_min <- ttm_walk_min %>%
-  group_by(UID) %>%
-  summarize(sum_b1_min = sum(impedance_binary),
-            .groups = "drop")
-
-sum_b1_o_min <- ttm_walk_o_min %>%
-  group_by(UID) %>%
-  summarize(sum_b1_o_min = sum(impedance_binary),
-            .groups = "drop")
-```
-
-```{r echo=FALSE}
-sum_b2_min <- ttm_walk_min %>%
-  group_by(hub) %>%
-  summarize(sum_b2_min = sum(impedance_binary))
-
-sum_b2_o_min <- ttm_walk_o_min %>%
-  group_by(hub) %>%
-  summarize(sum_b2_o_min = sum(impedance_binary))
-```
-
-```{r echo=FALSE}
-ttm_walk_min <- ttm_walk_min %>%
-  left_join(sum_b1_min, by = "UID") %>%
-  left_join(sum_b2_min, by = "hub")
-
-ttm_walk_o_min <- ttm_walk_o_min %>%
-  left_join(sum_b1_o_min, by = "UID") %>%
-  left_join(sum_b2_o_min, by = "hub")
-```
-
-```{r echo=FALSE}
-ttm_walk_min <- ttm_walk_min %>%
-  dplyr::filter(sum_b1_min > 0)
-
-ttm_walk_o_min <- ttm_walk_o_min %>%
-  dplyr::filter(sum_b1_o_min > 0)
-```
-
-```{r echo=FALSE}
-ttm_walk_min <- ttm_walk_min %>% 
-  dplyr::mutate(balanced_impedance_1 = impedance_binary/sum_b1_min, 
-         balanced_impedance_2 = impedance_binary/sum_b2_min)
-
-ttm_walk_o_min <- ttm_walk_o_min %>% 
-  dplyr::mutate(balanced_impedance_1 = impedance_binary/sum_b1_o_min, 
-         balanced_impedance_2 = impedance_binary/sum_b2_o_min)
-```
-
-```{r echo=FALSE}
-ttm_walk_min %>% 
-  dplyr::filter(hub == "46") %>%
-  pull(balanced_impedance_2) %>%
-  sum()
-
-ttm_walk_o_min %>% 
-  dplyr::filter(hub == "46") %>%
-  pull(balanced_impedance_2) %>%
-  sum()
-```
-
-```{r, include=FALSE}
-sobi_los_min <- ttm_walk_min %>%
-  dplyr::filter(impedance_binary > 0) %>% # filter impedance == 0 to avoid divisions by zero
-  group_by(hub) %>%
-  summarize(sobi_los_min = first(racks) / sum((population * balanced_impedance_1)),
-            .groups = "drop")
-
-sobi_los_o_min <- ttm_walk_o_min %>%
-  dplyr::filter(impedance_binary > 0) %>% # filter impedance == 0 to avoid divisions by zero
-  group_by(hub) %>%
-  summarize(sobi_los_o_min = first(racks) / sum((population * balanced_impedance_1)),
-            .groups = "drop")
-```
-
-```{r, include=FALSE}
-ttm_walk_min <- ttm_walk_min %>%
-  left_join(sobi_los_min, by = "hub")
-
-ttm_walk_o_min <- ttm_walk_o_min %>%
-  left_join(sobi_los_o_min, by = "hub")
-```
-
-```{r, include=FALSE}
-sobi_accessibility_min <- ttm_walk_min %>%
-  group_by(UID) %>%
-  summarize(accessibility_min = sum(sobi_los_min * balanced_impedance_2),
-            .groups = "drop")
-
-sobi_accessibility_o_min <- ttm_walk_o_min %>%
-  group_by(UID) %>%
-  summarize(accessibility_o_min = sum(sobi_los_o_min * balanced_impedance_2),
-            .groups = "drop")
-```
-
-```{r, echo=FALSE}
-#current system with equity stations
-sobi_accessibility_min %>% 
-  pull(accessibility_min) %>% 
-  sum()
-
-sobi_los_min %>% 
-  pull(sobi_los_min) %>%
-  sum()
-```
-
-```{r, echo=FALSE}
-#original system without equity stations
-sobi_accessibility_o_min %>% 
-  pull(accessibility_o_min) %>% 
-  sum()
-
-sobi_los_o_min %>% 
-  pull(sobi_los_o_min) %>%
-  sum()
-```
-
-```{r, include=FALSE}
-population_sobi <- population_sobi %>%
-  left_join(sobi_accessibility_min,
-            by = "UID")
-
-population_sobi_original <- population_sobi_original %>%
-  left_join(sobi_accessibility_o_min,
-            by = "UID")
+names(ttm_walk) <- c('UID', 
+                     'hub', 
+                     'travel_time',
+                     "hub_type",
+                     'population',
+                     'racks')
+```
+
+```{r, echo=FALSE}
+#population_sobi_original <- population_sobi_original[-c(3, 5:6)]
+```
+
+### Varying thresholds
+
+For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility:
+```{r}
+results_3min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional") %>%
+  b2sfca(threshold = 3)
+```
+
+All stations:
+```{r}
+results_3min_all <- ttm_walk %>%
+  b2sfca(threshold = 3)
+```
+
+Sanity check:
+```{r}
+sum(results_3min_conventional$los$los)
+sum(results_3min_conventional$accessibility$accessibility)
+```
+
+Sanity check:
+```{r}
+sum(results_3min_all$los$los)
+sum(results_3min_all$accessibility$accessibility)
+```
+
+For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility:
+```{r}
+results_5min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional") %>%
+  b2sfca(threshold = 5)
+```
+
+All stations:
+```{r}
+results_5min_all <- ttm_walk %>%
+  b2sfca(threshold = 5)
+```
+
+Sanity check:
+```{r}
+sum(results_5min_conventional$los$los)
+sum(results_5min_conventional$accessibility$accessibility)
+```
+
+Sanity check:
+```{r}
+sum(results_5min_all$los$los)
+sum(results_5min_all$accessibility$accessibility)
+```
+
+For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility:
+```{r}
+results_10min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional") %>%
+  b2sfca(threshold = 10)
+```
+
+All stations:
+```{r}
+results_10min_all <- ttm_walk %>%
+  b2sfca(threshold = 10)
+```
+
+Sanity check:
+```{r}
+sum(results_10min_conventional$los$los)
+sum(results_10min_conventional$accessibility$accessibility)
+```
+
+Sanity check:
+```{r}
+sum(results_10min_all$los$los)
+sum(results_10min_all$accessibility$accessibility)
+```
+
+For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility:
+```{r}
+results_15min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional") %>%
+  b2sfca(threshold = 15)
+```
+
+All stations:
+```{r}
+results_15min_all <- ttm_walk %>%
+  b2sfca(threshold = 15)
+```
+
+Sanity check:
+```{r}
+sum(results_15min_conventional$los$los)
+sum(results_15min_conventional$accessibility$accessibility)
+```
+
+Sanity check:
+```{r}
+sum(results_15min_all$los$los)
+sum(results_15min_all$accessibility$accessibility)
+```
+
+Extract results:
+```{r, include=FALSE}
+los <- rbind(data.frame(results_3min_all$los, system = "All stations", threshold = "3 min"),
+             data.frame(results_3min_conventional$los, system = "Conventional", threshold = "3 min"),
+             data.frame(results_5min_all$los, system = "All stations", threshold = "5 min"),
+             data.frame(results_5min_conventional$los, system = "Conventional", threshold = "5 min"),
+             data.frame(results_10min_all$los, system = "All stations", threshold = "10 min"),
+             data.frame(results_10min_conventional$los, system = "Conventional", threshold = "10 min"),
+             data.frame(results_15min_all$los, system = "All stations", threshold = "15 min"),
+             data.frame(results_15min_conventional$los, system = "Conventional", threshold = "15 min"))
+
+accessibility <- rbind(data.frame(results_3min_all$accessibility, system = "All stations", threshold = "3 min"),
+             data.frame(results_3min_conventional$accessibility, system = "Conventional", threshold = "3 min"),
+             data.frame(results_5min_all$accessibility, system = "All stations", threshold = "5 min"),
+             data.frame(results_5min_conventional$accessibility, system = "Conventional", threshold = "5 min"),
+             data.frame(results_10min_all$accessibility, system = "All stations", threshold = "10 min"),
+             data.frame(results_10min_conventional$accessibility, system = "Conventional", threshold = "10 min"),
+             data.frame(results_15min_all$accessibility, system = "All stations", threshold = "15 min"),
+             data.frame(results_15min_conventional$accessibility, system = "Conventional", threshold = "15 min"))
+```
+
+Add geometry:
+```{r}
+los <- los %>%
+  left_join(sobi_hubs %>%
+              dplyr::select(OBJECTID),
+            by = c("hub" = "OBJECTID")) %>%
+  st_as_sf()
+
+accessibility <- accessibility %>%
+  left_join(population_50x50 %>%
+              dplyr::select(UID),
+            by = c("UID" = "UID")) %>%
+  st_as_sf()
 ```
@@
 #accessibility plot for current system with equity stations at 3 minutes walk
-ggplot(data = population_sobi,
-       fill = accessibility_min) +
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_sf(data = accessibility %>%
+            dplyr::filter(threshold == "3 min"),
+       aes(color = accessibility)) +
   geom_sf(data = sobi_service,
@@
   scale_color_distiller(palette = "YlOrRd", 
-                        direction = 1)
-```
-
-```{r, echo=FALSE}
-#accessibility plot for original system without equity stations at 3 mins walk
-ggplot(data = population_sobi_original) +
-  geom_sf(aes(color = accessibility_o_min)) +
+                        direction = 1) +
+  facet_grid(system ~ threshold)
+```
+
+```{r, echo=FALSE}
+#accessibility plot for current system with equity stations at 5 minutes walk
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_sf(data = accessibility %>%
+            dplyr::filter(threshold == "5 min"),
+       aes(color = accessibility)) +
   geom_sf(data = sobi_service,
@@
   scale_color_distiller(palette = "YlOrRd", 
-                        direction = 1)
-```
-
+                        direction = 1) +
+  facet_grid(system ~ threshold)
+```
+
+```{r, echo=FALSE}
+#accessibility plot for current system with equity stations at 10 minutes walk
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_sf(data = accessibility %>%
+            dplyr::filter(threshold == "10 min"),
+       aes(color = accessibility)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  facet_grid(system ~ threshold)
+```
+
+```{r, echo=FALSE}
+#accessibility plot for current system with equity stations at 3 minutes walk
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_sf(data = accessibility %>%
+            dplyr::filter(threshold == "15 min"),
+       aes(color = accessibility)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  facet_grid(system ~ threshold)
+```
+
+<!--
 ```{r, echo=FALSE}
@@
 ```
+-->
 
@@
 ### Threshold of 10 Minutes Walk
-
+<!--
 ```{r, echo=FALSE}

### 2021-03-13 sobiEquity/raw-data-processing/00-Data-Processing-for-sobiEquity.Rmd
Major reorganization of repository to clean up files and to create (mostly data) package.

--- 2020-11-17 examples/00-Data-Processing-Example.Rmd
@@
 ---
-title: "00-Data-Processing-Example"
-output: html_notebook
+title: "00-Data-Processing-for-sobiEquity"
 ---
@@
 Particularly, we're interested in spatial interpolation of the population for small areas.
-
-Data used in this notebook was retrieved from two sources: Open Hamilton and the 2016 Canadian Census.
-Open Hamilton is an online repository of data from the City of Hamilton.
-The following datasets were retrieved:
-
-## Provenance of data
-
-### SoBi Hub
-
-The dataset was last updated on *November 9, 2020* and was downloaded for this project on *November 10, 2020*.
-This dataset contains the following attributes that are of interest to this research: location of hubs (address, longitude, and latitude) and the number of racks available at each hub.
-The dataset also includes the number of bicycles available at each hub, but since this changes daily, this information is not useful for our purposes.
-The maximum number of bicycles that a hub *could* have (i.e., the baseline) available is important for calculating accessibility in a given area.
-
-We can also retrieve the data weekly over a certain period, for example 4 weeks, to calculate the average number of bicycles that are at each hub.
-Variations in the number of bicycles could also be informative.
-For instance, some hubs may systematically have more or less bicycles than others (i.e., excess of supply around the University campus).
-
-### SoBi Service Area
-
-The dataset was last updated on *November 9, 2020* and was downloaded for this project on *November 10, 2020*.
-Hamilton's bike share program is available only in the lower city from Dundas to Ottawa Street North.
-There is one isolated hub beyond this service area at Van Wagner's Beach for tourism purposes.
-
-### Golf Courses
-
-The dataset was last updated on *November 1, 2020* and was downloaded for this project on *November 10, 2020*. This dataset contains the location of City and privately owned golf courses in Hamilton.
-
-### Parks
-
-The dataset was last updated on *November 8, 2020* and was downloaded for this project on *November 10, 2020*. This dataset contains the location of parks and other green spaces in Hamilton.
-
-### Large Employment Areas
-
-The dataset contains Employment Land boundaries in Hamilton and was last updated on *November 12, 2020*.
-It was downloaded for this project on *November 13, 2020*.
-This dataset contains the location of large business parks and industrial lands in Hamilton.
-
-### Municipal Parking Lots
-
-The dataset contains the location of municipal car parks in Hamilton and was last updated on *November 9, 2020*. It was downloaded for this project on *November 13, 2020*.
-
-### Cemeteries
-
-The dataset contains the location of cemeteries in Hamilton. It was downloaded for this project on *November 13, 2020*.
-
-### Environmentally Sensitive Areas
-
-The dataset contains the location of Environmentally Sensitive Areas (ESAs) in Hamilton.
-ESAs are either land or water areas containing natural features or significant ecological functions in Hamilton.
-The data were downloaded on *November 13, 2020*.
-
-### Street Centreline
-
-The dataset contains the street network in Hamilton.
-Data attributes include road classification so highways can be extracted.
-The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
-
-### Educational Institutions
-
-The dataset contains the location of all educational institutions and schools in Hamilton. The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
-
-### Places of Workshop
-
-The dataset contains the location of buildings used for religious congregations in Hamilton. The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
-
-### Municipal Service Centres
-
-The dataset contains the location of all municipal service centres in Hamilton.
-This includes Hamilton City Hall.
-The dataset was last updated on *November 11, 2020* and downloaded on *November 16, 2020*.
-
-### Recreation and Community Centres
-
-The dataset contains the location of all recreation and community centres in Hamilton.
-This includes Hamilton City Hall.
-The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
-
-### Arenas
-
-The dataset contains the location of all indoor arenas in Hamilton. This includes Hamilton City Hall. The dataset was last updated on *November 12, 2020* and downloaded on *November 16, 2020*.
-
-### EMS Stations
-
-The dataset contains the location of all Emergency Management Services (EMS) Ambulance stations in Hamilton.
-This includes Hamilton City Hall.
-The dataset was last updated on *November 11, 2020* and downloaded on *November 16, 2020*.
-
-### Fire Stations
-
-The dataset contains the location of all fire stations in Hamilton. This includes Hamilton City Hall. The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
-
-### Police Stations
-
-The dataset contains the location of all police stations in Hamilton. This includes Hamilton City Hall. The dataset was last updated on *November 13, 2020* and downloaded on *November 16, 2020*.
-
-### Hospitals
-
-The dataset contains the location of all hospitals in Hamilton. This includes Hamilton City Hall. The dataset was last updated on *April 2, 2019* and downloaded on *November 16, 2020*.
 
@@
 ```{r message = FALSE}
+library(cancensus)
 library(disk.frame)
@@
 
+Data used in this notebook was retrieved from two sources: Open Hamilton and the 2016 Canadian Census.
+Open Hamilton is an online repository of data from the City of Hamilton.
+The provenance of the datasets is below.
+
+## Provenance of data: City of Hamilton
+
+Source?
+
+### SoBi Hubs
+
+`SoBi_Hubs.shp`
+
+The dataset was last updated on *November 9, 2020* and was downloaded for this project on *November 10, 2020*.
+This dataset contains the following attributes that are of interest to this research: location of hubs (address, longitude, and latitude) and the number of racks available at each hub.
+The dataset also includes the number of bicycles available at each hub, but since this changes daily, this information is not useful for our purposes.
+The maximum number of bicycles that a hub *could* have (i.e., the baseline) available is important for calculating accessibility in a given area.
+
+We can also retrieve the data weekly over a certain period, for example 4 weeks, to calculate the average number of bicycles that are at each hub.
+Variations in the number of bicycles could also be informative.
+For instance, some hubs may systematically have more or less bicycles than others (i.e., excess of supply around the University campus).
+
+### SoBi Service Area
+
+`SoBi_Service_Areas.shp`
+
+The dataset was last updated on *November 9, 2020* and was downloaded for this project on *November 10, 2020*.
+Hamilton's bike share program is available only in the lower city from Dundas to Ottawa Street North.
+There is one isolated hub beyond this service area at Van Wagner's Beach for tourism purposes.
+
+### Golf Courses
+
+`Golf_Courses.shp`
+
+The dataset was last updated on *November 1, 2020* and was downloaded for this project on *November 10, 2020*. This dataset contains the location of City and privately owned golf courses in Hamilton.
+
+### Parks
+
+`Parks.shp`
+
+The dataset was last updated on *November 8, 2020* and was downloaded for this project on *November 10, 2020*. This dataset contains the location of parks and other green spaces in Hamilton.
+
+### Large Employment Areas
+
+`Employment_Lands.shp`
+
+The dataset contains Employment Land boundaries in Hamilton and was last updated on *November 12, 2020*.
+It was downloaded for this project on *November 13, 2020*.
+This dataset contains the location of large business parks and industrial lands in Hamilton.
+
+### Municipal Parking Lots
+
+`Municipal_Parking_Lots.shp`
+
+The dataset contains the location of municipal car parks in Hamilton and was last updated on *November 9, 2020*. It was downloaded for this project on *November 13, 2020*.
+
+### Cemeteries
+
+`Cemeteries.shp`
+
+The dataset contains the location of cemeteries in Hamilton. It was downloaded for this project on *November 13, 2020*.
+
+### Environmentally Sensitive Areas
+
+`Environmentally_Sensitive_Areas_Boundaries.shp`
+
+The dataset contains the location of Environmentally Sensitive Areas (ESAs) in Hamilton.
+ESAs are either land or water areas containing natural features or significant ecological functions in Hamilton.
+The data were downloaded on *November 13, 2020*.
+
+### Street Centreline
+
+`Street_Centreline.shp`
+
+The dataset contains the street network in Hamilton.
+Data attributes include road classification so highways can be extracted.
+The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
+
+### Educational Institutions
+
+`Educational_Institutions.shp`
+
+The dataset contains the location of all educational institutions and schools in Hamilton. The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
+
+### Places of Worship
+
+`Places_of_Worship.shp`
+
+The dataset contains the location of buildings used for religious congregations in Hamilton. The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
+
+### Municipal Service Centres
+
+`Municipal_Service_Centres.shp`
+
+The dataset contains the location of all municipal service centres in Hamilton.
+This includes Hamilton City Hall.
+The dataset was last updated on *November 11, 2020* and downloaded on *November 16, 2020*.
+
+### Recreation and Community Centres
+
+`Recreation_and_Community_Centres.shp`
+
+The dataset contains the location of all recreation and community centres in Hamilton.
+This includes Hamilton City Hall.
+The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
+
+### Arenas
+
+`Arenas.shp`
+
+The dataset contains the location of all indoor arenas in Hamilton. This includes Hamilton City Hall. The dataset was last updated on *November 12, 2020* and downloaded on *November 16, 2020*.
+
+### EMS Stations
+
+`EMS_Stations.shp`
+
+The dataset contains the location of all Emergency Management Services (EMS) Ambulance stations in Hamilton.
+This includes Hamilton City Hall.
+The dataset was last updated on *November 11, 2020* and downloaded on *November 16, 2020*.
+
+### Fire Stations
+
+`Fire_Stations.shp`
+
+The dataset contains the location of all fire stations in Hamilton. This includes Hamilton City Hall. The dataset was last updated on *November 15, 2020* and downloaded on *November 16, 2020*.
+
+### Police Stations
+
+`Police_Stations.shp`
+
+The dataset contains the location of all police stations in Hamilton. This includes Hamilton City Hall. The dataset was last updated on *November 13, 2020* and downloaded on *November 16, 2020*.
+
+### Hospitals
+
+`Hospitals.shp`
+
+The dataset contains the location of all hospitals in Hamilton. This includes Hamilton City Hall. The dataset was last updated on *April 2, 2019* and downloaded on *November 16, 2020*.
+
+### Railways
+
+## Provenance of data: census
+
+We can get an API key to use `cancensus` from [CensusMapper](https://censusmapper.ca/). Once you have an API key, it can be stored locally this way:
+```{r eval=FALSE}
+set_api_key("your-key", install = TRUE)
+```
+
+Once you have your API you can use `cancensus` like in this example.
+
+Use `cancensus` to check and download census data. Check the regions:
+```{r}
+list_census_regions('CA16') %>% 
+  dplyr::filter(level == "CMA", name %in% c("Hamilton"))
+```
+
+We need PR_UID 35.
+
+List census data sets:
+```{r}
+list_census_datasets()
+```
+
+We need CA16.
+
+Search variables:
+```{r}
+find_census_vectors("population", 
+                    dataset = "CA16", 
+                    type = "total", 
+                    query_type = "keyword", 
+                    interactive = FALSE)
+```
+
+```{r}
+find_census_vectors("median household income", 
+                    dataset = "CA16", 
+                    type = "total", 
+                    query_type = "keyword", 
+                    interactive = FALSE)
+```
+
+We need v_CA16_401 and v_CA16_2397.
+
+Retrieve data:
+```{r}
+population_da_2016 <- get_census(dataset='CA16', 
+                          regions=list(CMA=c("35537")),
+                          vectors=c("v_CA16_401"),
+                          level='DA', 
+                          use_cache = FALSE)
+
+mhi <- get_census(dataset='CA16', 
+                          regions=list(CMA=c("35537")),
+                          vectors=c("v_CA16_2397"),
+                          level='DA', 
+                          use_cache = FALSE)
+```
+
 ### Read data
@@
 ```{r}
-load("hamilton_cma.RData")
-load("hamilton_da_2016.RData")
+load("input-data-files/hamilton_cma.RData")
+load("input-data-files/hamilton_da_2016.RData")
 ```
@@
 ```{r}
-sobi_service <- st_read("SoBi_Service_Areas.shp")
+sobi_service <- st_read("input-data-files/SoBi_Service_Areas.shp")
 ```
@@
 ```{r}
-parks <- st_read("Parks.shp")
+parks <- st_read("input-data-files/Parks.shp")
 ```
@@
 ```{r}
-cemeteries <- st_read("Cemeteries.shp")
+cemeteries <- st_read("input-data-files/Cemeteries.shp")
 ```
@@
 ```{r}
-esa <- st_read("Environmentally_Sensitive_Areas_Boundaries.shp")
+esa <- st_read("input-data-files/Environmentally_Sensitive_Areas_Boundaries.shp")
 ```
@@
 ```{r}
-golf_courses <- st_read("Golf_Courses.shp")
+golf_courses <- st_read("input-data-files/Golf_Courses.shp")
 ```
@@
 ```{r}
-employment_lands <- st_read("Employment_Lands.shp")
+employment_lands <- st_read("input-data-files/Employment_Lands.shp")
 ```
@@
 ```{r}
-railways <- st_read("Railways.shp")
+railways <- st_read("input-data-files/Railways.shp")
 ```
@@
 ```{r}
-municipal_parking <- st_read("Municipal_Parking_Lots.shp")
+municipal_parking <- st_read("input-data-files/Municipal_Parking_Lots.shp")
 ```
@@
 ```{r}
-streets <- st_read("Street_Centreline.shp")
+streets <- st_read("input-data-files/Street_Centreline.shp")
 ```
@@
 ```{r}
-schools <- st_read("Educational_Institutions.shp")
+schools <- st_read("input-data-files/Educational_Institutions.shp")
 ```
@@
 ```{r}
-religious_places <- st_read("Places_of_Worship.shp")
+religious_places <- st_read("input-data-files/Places_of_Worship.shp")
 ```
@@
 ```{r}
-service_centres <- st_read("Municipal_Service_Centres.shp")
+service_centres <- st_read("input-data-files/Municipal_Service_Centres.shp")
 ```
@@
 ```{r}
-recreation_centres <- st_read("Recreation_and_Community_Centres.shp")
+recreation_centres <- st_read("input-data-files/Recreation_and_Community_Centres.shp")
 ```
@@
 ```{r}
-arenas <- st_read("Arenas.shp")
+arenas <- st_read("input-data-files/Arenas.shp")
 ```
@@
 ```{r}
-ems <- st_read("EMS_Stations.shp")
+ems <- st_read("input-data-files/EMS_Stations.shp")
 ```
@@
 ```{r}
-fire <- st_read("Fire_Stations.shp")
+fire <- st_read("input-data-files/Fire_Stations.shp")
 ```
@@
 ```{r}
-police <- st_read("Police_Stations.shp")
+police <- st_read("input-data-files/Police_Stations.shp")
 ```
@@
 ```{r}
-hospitals <- st_read("Hospitals.shp")
+hospitals <- st_read("input-data-files/Hospitals.shp")
 ```
@@
 hamilton_da_sobi <- hamilton_da_2016[st_as_sfc(bounding_box),]
+```
+
+Recreate the bounding box:
+```{r}
+bounding_box <- st_bbox(hamilton_da_sobi)
 ```
@@
 ```
+
 Extract municipal parking lots in service area:
@@
 
-### Read population
+### Prepare population data
 
-Read population table (2016 Census of Population, Population and Dwelling counts):
+Transmute to retain only variables of interest:
 ```{r}
-population_da_2016 <- read_csv("T1901EN.csv") %>%
-  transmute(DA = factor(`Geographic code`), population = `Population, 2016`)
+population_da_2016 <- population_da_2016 %>%
+  transmute(DA = factor(`GeoUID`), population = Population)
 ```
@@
 ```
+
 Remove hospitals from the DAs (use `st_difference()`):
@@
 ```{r}
-sobi_hubs <- st_read("SoBi_Hubs.shp")
+sobi_hubs <- st_read("input-data-files/SoBi_Hubs.shp")
 ```
@@
   geom_sf(data = sobi_hubs,
-          aes(size = RACKS_AMOU)) +
+          aes(shape = hub_type, 
+              size = RACKS_AMOU)) +
   scale_color_distiller(palette = "YlOrRd", 
@@
 
-## Routing
+Identify ERI hubs that were implemented in 2018 to increase accessibility for under-serviced and disadvantaged areas of the city.
+This reflects the original state of the system without the equity stations:
+```{r}
+sobi_hubs <- sobi_hubs %>%
+  dplyr::mutate(hub_type = case_when(NAME == "Mars at Wentworth - ERI01" ~ "ERI",
+                                     NAME == "Sherman at Barton - ERI02" ~ "ERI",
+                                     NAME == "Barton at Lottridge - ERI03" ~ "ERI",
+                                     NAME == "Barton and Belview - ERI05" ~ "ERI",
+                                     NAME == "Barton at Ottawa - ERI06" ~ "ERI",
+                                     NAME == "Ottawa at Dunsmure - ERI07" ~ "ERI",
+                                     NAME == "Maple at Rothsay - ERI08" ~ "ERI",
+                                     NAME == "King at Dunsmure - ERI09" ~ "ERI",
+                                     NAME == "Dunsmure at Sherman - ERI10" ~ "ERI",
+                                     NAME == "Gage at Cannon - ERI11" ~ "ERI",
+                                     NAME == "Belview at Cannon - ERI12" ~ "ERI",
+                                     NAME == "Westinghouse at Barton - ERI13" ~ "ERI",
+                                     TRUE ~ "Conventional"),
+                hub_type = factor(hub_type))
+
+```
+
+Plot again:
+```{r}
+ggplot() + 
+  geom_sf(data = interpolated_pop_clean,
+          aes(color = population)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  geom_sf(data = sobi_hubs,
+          aes(shape = hub_type,
+              size = RACKS_AMOU)) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  scale_size(range = c(1, 4)) + 
+  facet_wrap(~hub_type)
+```
+
+After removing the twelve ERI hubs, there is much less coverage in the east end of the service area.
+Again, the service area is the area where a cyclist can use a SoBi bike.
+It makes sense then that ERI would want to fill this gap in coverage because individuals who live in the east end of the service area could theoretically use the bike share system but may not live near a hub.
+For this reason, twelve hubs were added in 2018 to expand the system.
+
+## Routing for entire system
 
@@
 
+Join hub type information to travel time table:
+```{r}
+ttm_walk <- ttm_walk %>%
+  left_join(sobi_hubs %>%
+              st_drop_geometry() %>%
+              dplyr::select(OBJECTID,
+                            hub_type),
+            by = "OBJECTID")
+```
+
 ## Save data to disk
@@
 ```{r}
-population_sobi <- interpolated_pop_clean
+population_50x50 <- interpolated_pop_clean
+```
+
+Clean DA boundaries:
+```{r}
+hamilton_da_2016 <- hamilton_da_2016 %>%
+  dplyr::select(-POP2011) %>%
+  left_join(population_da_2016,
+            by = "DA") %>%
+  left_join(mhi %>%
+              transmute(GeoUID, median_total_income = `v_CA16_2397: Median total income of households in 2015 ($)`),
+            by = c("DA" = "GeoUID")) %>%
+  left_join(hamilton_da_sobi %>%
+              st_drop_geometry() %>%
+              transmute(DA,
+                        sobi_service = "Yes"),
+            by = "DA") %>%
+  mutate(sobi_service = ifelse(is.na(sobi_service), "No", "Yes") %>%
+           factor())
 ```
@@
 ```{r}
-save(population_sobi, file = "population_sobi.RData")
-save(sobi_hubs, file = "sobi_hubs.RData")
-save(sobi_service, file = "sobi_service.RData")
-save(ttm_walk, file = "ttm_walk.RData")
+save(hamilton_cma, file = "output-data-files/hamilton_cma.RData", compress = "xz")
+save(hamilton_da_2016, file = "output-data-files/hamilton_da_2016.RData", compress = "xz")
+save(population_50x50, file = "output-data-files/population_50x50.RData", compress = "xz")
+save(sobi_hubs, file = "output-data-files/sobi_hubs.RData", compress = "xz")
+save(sobi_service, file = "output-data-files/sobi_service.RData", compress = "xz")
+save(ttm_walk, file = "output-data-files/ttm_walk.RData", compress = "xz")
 ```
+
+Copy files to folder `data` of the package.

### 2021-03-13 sobiEquity/R/b2sfca.R

@@ -1,0 +1,60 @@
+#' Compute balanced floating catchment area accessibility
+#'
+#' @param data A data frame with a column of origin IDs, a column of destination IDs, travel times, type of destinations, population at the origin, and supply at the destination.
+#' @return The level of service \code{los} and accessibility \code{accessibility}.
+#' @export
+#' @examples
+#' #add(1, 1)
+#' #add(10, 1)
+
+b2sfca <- function(ttm = ttm, threshold = threshold){
+  # calculate the impedance
+  ttm <- ttm %>%
+    mutate(impedance_binary = ifelse(travel_time <= threshold, 1, 0))
+
+  # Calculate the sum of the impedance by population for the balancing factors. In the binary case, this should be the same as the number of SoBi hubs that each interpolated population unit can reach
+  sum_b1 <- ttm %>%
+    group_by(UID) %>%
+    summarize(sum_b1 = sum(impedance_binary),
+              .groups = "drop")
+
+  # Calculate the sum of the impedance by SoBi hubs for the balancing factors. In the binary case, this should be the same as the number of interpolated population cells that each SoBi hub serves
+  sum_b2 <- ttm %>%
+    group_by(hub) %>%
+    summarize(sum_b2 = sum(impedance_binary))
+
+  # Join the sum of the impedance to the table to calculate the balanced impedance values:
+  ttm <- ttm %>%
+    left_join(sum_b1, by = "UID") %>%
+    left_join(sum_b2, by = "hub")
+
+  # **Important:** notice that some values of `sum_b1` are zeros! This means that some population units cannot reach a SoBi hub in at most 24 min. Since the balancing impedance is calculated by diving the impedance by `sum_b1`, this will lead to NaNs (divisions by zero). To avoid issues we will remove any population units that reach zero hubs:
+  ttm <- ttm %>%
+    dplyr::filter(sum_b1 > 0,
+                  sum_b2 > 0)
+
+  # Calculate balanced impedance values:
+  ttm <- ttm %>%
+    dplyr::mutate(balanced_impedance_1 = impedance_binary/sum_b1,
+                  balanced_impedance_2 = impedance_binary/sum_b2)
+
+  # The level of service of each hub is the number of bicycle racks at the hub, divided by the population that they serve:
+  los <- ttm %>%
+    group_by(hub) %>%
+    summarize(los = first(racks) / sum((population * balanced_impedance_1)),
+              .groups = "drop")
+
+  # Join the level of service to the table:
[…]

### 2021-03-13 sobiEquity/R/sobiEquity.R

@@ -1,0 +1,23 @@
+#' sobiEquity
+#'
+#' sobiEquity is a companion package for the paper "Examining spatial equity and accessibility
+#'  to a public bicycle share program using a balanced floating catchment area approach".
+#'   The package includes the data used in the analysis.
+#'
+#' @author Elise Desjardins, \email{desjae@@mcmaster.ca}
+#' @author Antonio Paez, \email{paezha@@mcmaster.ca}
+#' @docType package
+#' @name sobiEquity
+NULL
+
+#' Hamilton CMA.
+#'
+#' Boundary of the Hamilton CMA, in Ontario, Canada.
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords transportation
+#' @name hamilton_cma
+#' @usage data(hamilton_cma)
+#' @source Statistics Canada
+"hamilton_cma"

### 2021-03-17 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Revised draft of paper
Background, Methods, Discussion, and Conclusion are written

--- 2021-02-11 examples/Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
@@
     text: "Equal contribution"
-abstract: |
-  This is the abstract.
+abstract:
 
-  It consists of two paragraphs.
+  Public bicycle share programs are implemented to promote cycling as a convenient and sustainable mode of transportation.
+These systems can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to lower income or disadvantaged populations.
+Recent studies assessing equity in bike sharing have found differences in ridership or membership based on income, education, and trip purpose.
+In Ontario, the City of Hamilton launched a public bicycle share program in 2015 that currently has over 900 operational bicycles and 130 docking stations.
+In 2018, the non-profit organization that operates the program launched an equity initiative to provide subsidized memberships and to expand service by adding twelve docking stations.
+Previous research found that Hamilton’s public bicycle share program targeted well disadvantaged areas in the city compared to programs in other Canadian cities.
+Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program, the location of stations matters.
+Unlike other public amenities with greater ability to adapt to crowding, the utility of stations is limited by the number of bicycles that they can hold, which makes crowding effects critical; the system may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles.
+In this research, we revisit the case of Hamilton and investigate equity differentials in accessibility to bicycle share stations.
+We compare accessibility in the program with and without the equity stations to assess the effect of the initiative.
+Previous research on cycling equity or accessibility has used a macro-level approach with entire neighbourhoods or census tracts as the unit of analysis, although there are recent exceptions.
+In contrast, we implement our analysis parting from microscopic zones to better reflect travel to a docking station, which often happens at the sub-neighborhood level, and conduct a sensitivity analysis at several walking thresholds.
+We then reaggregate the estimated accessibility to dissemination areas for further analysis using census data.
+Equity analysis of this type is possible thanks to a newly developed approach for accessibility using balanced floating catchment areas.
+The addition of equity stations achieved its goal to increase accessibility for lower income populations, although the increase was relatively modest (2.10 bicycles/person with equity stations vs 1.97 bicycles/person without equity stations).
+Further research is needed to determine whether this encouraged more cycling.
 
@@
 
-```{r echo=FALSE}
+```{r load-packages, include=FALSE}
+library(data.table)
 library(dplyr)
-library(disk.frame)
-library(pycno)
-library(readr)
-library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
-library(sf)
-library(tidyverse)
-library(data.table)
 library(gdistance)
@@
 library(sf)
+library(sobiEquity)
+library(tidyverse)
 library(units)
+#library(disk.frame)
+#library(pycno)
+library(readr)
+#library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
+library(tinytex)
+library(ggplot2)
+theme_set(theme_bw())
 ```
 
-```{r echo=FALSE}
-load("hamilton_cma.RData")
-load("hamilton_da_2016.RData")
-load("population_sobi.RData") #interpolated population in SoBi area
-load("sobi_hubs.RData") #location of SoBi hubs, minus Van Wagner's beach
-load("sobi_hubs_original.RData") #SoBi hubs, minus ERI hubs
-load("sobi_service.RData") #core service area for SoBi
-load("ttm_walk.RData") #travel time matrix for interpolated population in SoBi area
-load("ttm_walk_original.RData") #travel time matrix for system without ERI hubs
+```{r}
+data("hamilton_cma")
+data("hamilton_da_2016")
+data("population_50x50")
+data("sobi_service")
+data("sobi_hubs")
+data("ttm_walk")
 ```
 
-Introduction
+```{r coordinates-of-equity-stations, include=FALSE}
+# define the coordinates of the equity stations
+# equity_stations <- data.frame(longitude = c(-79.84364993846138, 
+#                                             -79.8362839215856,
+#                                             -79.83144521746212, 
+#                                             -79.82217214949738, 
+#                                             -79.8160211746482, 
+#                                             -79.8192304372373, 
+#                                             -79.82500255113169, 
+#                                             -79.8268599805827, 
+#                                             -79.83888834759689, 
+#                                             -79.82740849222654, 
+#                                             -79.82342541216097, 
+#                                             -79.84240606419081), 
+#                               latitude = c(43.265340158000654, 
+#                                            43.25632467284136, 
+#                                            43.25528943845642, 
+#                                            43.25291050074978, 
+#                                            43.25147804403345, 
+#                                            43.24444659758296, 
+#                                            43.2421627336722, 
+#                                            43.24634382671116,
+#                                            43.249432741404554,
+#                                            43.251126444296524, 
+#                                            43.24892849896579, 
+#                                            43.25831920064561))
+```
+
+```{r, include=FALSE}
+#equity_stations <- st_as_sf(equity_stations, coords = c("longitude", "latitude"), 
+    #crs = st_crs(sobi_service), agr = "constant")
+```
+
+```{r load-data, include=FALSE}
+#load("accessibility.RData") #full accessibility table for SoBi system
+#load("accessibility_map_original.RData") #map of accessibility in original SoBi system
+#load("accessibility_map_equity.RData") #map of accessibility in current system with ERI
+# load("hamilton_cma.RData")
+# load("hamilton_da_2016.RData")
+# load("hamilton_da_sobi.RData")
+# load("mhi.RData") #median household income by DA
+# load("population_sobi.RData") #interpolated population in SoBi area
+#load("population_sobi_original.RData") #population in original system
+# load("sobi_hubs.RData") #location of SoBi hubs, minus Van Wagner's beach
+# load("sobi_hubs_original.RData") #SoBi hubs, minus ERI hubs
+# load("sobi_service.RData") #core service area for SoBi
+# load("ttm_walk.RData") #travel time matrix for interpolated population in SoBi area
+# load("ttm_walk_original.RData") #travel time matrix for system without ERI hubs
+```
+
+Background
 ===================
 
-Public bicycle share programs (PBSP)
+Public bicycle share programs (PBSP) have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users [@fishmanBikeshareReviewRecent2016].
+In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @reillyGenderDisparitiesNew2020; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011].
+However, one study found that bike share users in Washington, DC were more likely to be female [@buckAreBikeshareUsers2013], which suggests that the gender gap among cyclists who use bike share is less disparate than the gap for personal bicycle use [@fishmanBikeshareReviewRecent2016].
+There is some evidence that bike share users are less likely to own a car [@buckAreBikeshareUsers2013; @reillyNoncyclistsFrequentCyclists2020].
+However, the relationship between income or education and bike share use is less clear.
+Stations in disadvantaged communities in Chicago generated most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighbourhoods in Minneapolis-St.
+Paul used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019].
+Being university educated was also a significant correlate of bike share use in Montreal, Canada [@fullerUseNewPublic2011].
+On the other hand, members who reside in minority-concentrated and lower socioeconomic status neighbourhoods use the Minneapolis-St Paul bike share more frequently [@wangNeighborhoodSociodemographicCharacteristics2019].
+Financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016].
+Many studies have found that proximity to a bike share station, either living or working near one, is an important determinant of use or having a membership [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011].
+This makes sense given that individuals are more likely to use services or programs that they can easily access.
+Finally, the potential of bike share programs to increase cycling levels has also been explored recently [@hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019], but more longitudinal evidence is needed to determine their impact on encouraging more cycling and offering more opportunities for physical activity.
+On the whole, these findings highlight the need for bike share systems to be more highly accessible for diverse populations in order to increase use beyond the "typical" users, which has been the focus of recent research [@auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020], and to offer more people the option of using sustainable and active transportation.
+
+Several studies have recently explored the equity of PBSP in North American cities by primarily examining who has access to bike share (e.g., differences by demographics or socioeconomic status) and where stations are located.
+Equity can be achieved in two different ways: equal spatial distribution across a region (e.g., horizontal equity) or greater access for vulnerable or disadvantaged populations (e.g., vertical equity) [@chenExploringEquityPerformance2019].
+Both are of interest to researchers and transport planners since they are often linked in that advantage, or conversely disadvantage, has spatial patterns.
+Using a negative binomial regression model, Qian and Jaller [-@qianBikesharingEquityDisadvantaged2020] estimated ridership in Chicago's PBSP among disadvantaged communities and found some disparities.
+While annual members in disadvantaged communities have a significantly lower share of trips compared to other areas in the city, they make longer trips.
+This suggests that they may be using PBSP for work commuting, which points to the importance of ensuring equitable access [@qianBikesharingEquityDisadvantaged2020].
+Similar results were found in Philadelphia, where lower income areas generate fewer trips and efforts to increase equity within the program have not been as successful as intended [@caspiBikesharingPhiladelphiaLowerincome2019].
+In the case of Seattle, all neighbourhoods had some level of access to dockless bikes but those with higher incomes and more residents of higher education had more bikes [@mooneyFreedomStationSpatial2019].
+Babagoli et al.
+[-@babagoliExploringHealthSpatial2019] found that neighbourhoods in New York City with higher affluence had the greater proportion of Citi Bike stations.
+Overall, these findings suggest that horizontal equity can be achieved while vertical equity is harder to attain.
+Exploring transport equity by investigating where bicycle share stations are located, often using neighbourhoods or census tracts as the spatial unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
+Meaning that, stations may be lacking in certain neighbourhoods but there may be stations accessible within a reasonable walking cutoff time.
+This is where accessibility becomes an important consideration.
+
+Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved since its conceptualization and become linked with recent interest in prioritizing local proximity and modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020].
+As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequities in usage, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options.
+It also addresses some of the challenges of other performance measures such as level of service within a transportation network by measuring person-based indicators and exploring differences in use between population groups [@paezMeasuringAccessibilityPositive2012].
+
+In the context of PBSPs, accessibility refers to the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019].
+Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program and ultimately use of the program, the location and size (e.g., maximum number of bicycles available) of stations matters.
+Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
+Kabra et al.
+[-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300m of stations.
+Furthermore, unlike other public amenities with greater ability to adapt to crowding, like health care services, the utility of stations is limited by the number of bicycles that they can hold, which makes crowding effects critical.
+The program may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles.
+Likewise, more people may not opt to use the program if the supply of bicycles available at the nearest station is insufficient to meet demand.
+The location and size of stations is important to increase the utility of this public transportation option for more people, thus achieving vertical equity.
+Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts.
+For instance, Wang and Lindsey [-@wangNewBikeShare2019] investigated whether new or relocated bicycle share stations increased accessibility and use, which offered important insights to improve the performance of the program.
+
+Several approaches have been commonly used for measuring accessibility: cumulative opportunities, gravity, and utility-based [@handyMeasuringAccessibilityExploration1997].
+Paez et al [-@paezMeasuringAccessibilityPositive2012] provide a recent overview of various formulations and applications of accessibility in transportation research.
+Briefly, the gravity-based approach involves weighting opportunities, for example the quantity of bicycle share stations, as measured by a function of time, using a negative exponential form [@handyMeasuringAccessibilityExploration1997].
+It takes both demand and congestion into account [@paezMeasuringAccessibilityPositive2012].
+Floating catchment area (FCA) methods are a type of gravity measure that have been used in healthcare accessibility research but that are applicable to transportation case studies.
+This approach involves producing flexible catchment areas for populations, recognizing that people may be willing to travel to access particular services [@paezMeasuringAccessibilityPositive2012].
+Thus, this model does a good job of considering potential crowding of services using a binary distance function [@paezMeasuringAccessibilityPositive2012].
+A recent improvement to this approach was achieved through a simple and intuitive adjustment to the allocation of supply and demand, which addressed the effects of demand and service inflation from the conventional approach [see @paezDemandLevelService2019].
+
+Researchers have also taken different approaches when it comes to aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones.
+Previous research on bike share equity has typically used a macro-level approach with aggregate data from entire neighbourhoods or census tracts  [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021].
+This is also true for studies examining correlates of bike share demand [@wangNewBikeShare2019].
+Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals, which is useful for addressing vertical inequities in PBSP usage.
+
+Using balanced floating catchment area methods, a novel approach that has not been used yet in cycling research, we examine accessibility to a public bicycle share program in Hamilton, Ontario.
+Using disaggregate population-level data, we (1) conduct a sensitivity analysis by measuring accessibility and level of service to bike share stations at different thresholds of walking to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median household income of dissemination areas within the core service area.
+
+Case Study {#sec:study}
+======================
 
 This paper uses the city of Hamilton, located in Ontario, Canada, as a case study.
-The city launched the program, called SoBi Hamilton, in March 2015 with 115 stations and 750 bicycles [@Hamilton2020].
-The core service area spans 35 km^2 of the city and approximately 180,000 people live there [see Figure 1].
-The program was enthusiastically welcomed in the city - within three weeks of launching, 10,000 trips had been made.
+The city launched a public bicycle share program, Hamilton Bike Share, in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015].
+Before June 2020, the program was known as SoBi Hamilton.
+Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021].
+The core service area spans 35 km^2 of the city and _approximately 130,000 can reach a bike share hub within 30 minutes of walking_ [see Figure 1].
+This represents roughly one fifth of the population in Hamilton, according to the 2016 Canadian Census.
+The program was enthusiastically welcomed in the city - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015].
 In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton.
-To increase spatial equity, an additional 75 bicycles and 15 stations were added to the system, which expanded it to more disadvantaged areas in the city [see Figure 2].
+An additional 75 bicycles and 15 stations were added to the program, which expanded it to more disadvantaged areas in the core service area [see Figure 2].
 The program also offers subsidized memberships to individuals who identify as low income.
-As of June 2020, the bike share program has 900 bikes and over 130 stations, and over 26,000 active memberships [@Hamilton2020].
+A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019].
+As of June 2020, the bike share program has 900 bikes and 130 stations, and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
 
-```{r echo=FALSE}
-sobi_service <- sobi_service %>%
-  dplyr::filter(AREA_NAME == "Core Service Area")
-```
+Hamilton Bike Share has conducted one membership survey to date in 2018 [@civicplanSoBiHamiltonMembership2017], and the findings from 420 members are broadly in line with the trends that were discussed above [see @fishmanBikeshareReviewRecent2016 for a recent review of the literature].
+The majority of respondents live within the core service area and the gender split is typical: 57% of respondents are male and 41% are female.
+The majority of respondents, both male and female, are between 25 and 34 years of age, but the percentage of male respondents is higher in the subsequent age groups.
+Respondents use bike share for commuting (40% of trips) or errands and meetings (24% of trips), and nearly 50% of trips have an average length of 11 to 20 minutes.
+As a result of having a bike share membership, 49% of respondents report that they use their private vehicle less often or much less often but 48% report that their private vehicle use has remained about the same.
+This suggests that SoBi Hamilton has been useful for certain kinds of trips but not all, meaning that some trips continue to require a private vehicle.
 
-```{r echo=FALSE}
-bounding_box <- st_bbox(sobi_service %>%
-                          st_buffer(500))
-```
+Further research has provided insights about the behaviour and route choice of bike share users in Hamilton.
+The routes most frequently travelled are longer than the shortest distance route from origin to destination [@luUnderstandingBikeShare2018; @scottRouteChoiceBike2021].
+Ridership is influenced by weather conditions, temporal factors such as university terms, employment, and proximity to important destinations like the local university, but is not influenced by population and most transportation infrastructure variables [@scottWhatFactorsInfluence2019].
+Bike share users prefer to deter from the shortest distance route to travel on streets with cycling facilities or lower volumes of traffic [@luUnderstandingBikeShare2018] and avoid steep slopes and busy roads [@scottRouteChoiceBike2021].
+It is important to note that these studies used daily ridership or GPS data from trips taken before the equity stations were implemented, so these findings are informative with respect to the conventional stations.
 
-```{r echo=FALSE}
+Our analysis builds upon a previous and recent study [@hosfordWhoArePublic2018], which found that disadvantaged areas in Hamilton are better served by the city's PBSP compared to other Canadian cities [i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau] with PBSPs where advantaged areas have greater access.
+Hosford and Winters [-@hosfordWhoArePublic2018] acknowledge that "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas".
+Therefore, the core service area for the PBSP in Hamilton is by default in more disadvantaged areas compared to other Canadian cities, but there is also a great deal of variation in income in the city center as well because of the local university and increasing gentrification in the downtown area.
+In their analysis, Hosford and Winters [-@hosfordWhoArePublic2018] did not differentiate between the conventional stations and those added to increase equity, which we refer to as equity stations in this paper.
+Therefore, this paper continues their work and makes an important contribution to the growing body of literature that examines equity issues in PBSPs.
+
+```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The core service area of Hamilton Bike Share is outlined in blue.
+Dissemination areas in the Hamilton Census Metropolitan Area (CMA) are outlined in grey."}
 ggplot() + 
@@
           fill = NA,
-          colour = "blue")
+          colour = "black")
 ```
 
-Methods
-===================
+```{r filter-eri-hubs, include=FALSE}
+# eri_hubs <- sobi_hubs %>% 
+#              filter(OBJECTID>=117 & OBJECTID<=128)
+```
+
+```{r, include=FALSE}
+sobi_downtown <- st_read("Community_Downtown.shp")
+```
+
+```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The spatial distribution of bike share stations in Hamilton, Ontario.
+The service area of Hamilton Bike Share is outlined in blue and the city's downtown core is outlined in light blue.
+Dissemination areas within the Hamilton CMA that touch the bounding box of the core service area are outlined in grey."}
+ggplot() + 
+  geom_sf(data = hamilton_da_2016 %>% 
+            dplyr::filter(sobi_service == "Yes"),
+          color = "gray") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "black") +
+  geom_sf(data = sobi_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = "dark green",
+          color = "dark green") +
+  geom_sf(data = sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "blue") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          size = 1.5, shape = 21, 
+          fill = "green")
+```
+
+Methods {#sec:methods}
+======================
+
+## Floating Catchment Area
+
+Floating catchment area (FCA) methods are an approach commonly used in the healthcare accessibility literature.
+This approach is more appropriate and informative than calculating the provider-to-population ratio (PPR) which simply divides the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019].
+In particular, the Two-Step Floating Catchment Area method produces flexible catchment areas for populations accessing a service rather than using rigid boundaries like PPR, which provides more useful information because it takes into account individual behaviour and doesn't assume that people stay within the pre-defined boundaries [@paezDemandLevelService2019].
+This is an important property that supports our rationale for applying this method to measure accessibility to Hamilton Bike Share.
+The City of Hamilton has positioned stations between 300 and 600 metres apart, which would lead to boundaries around these buffers, but it would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.
+
+More recently, the _balanced_ floating catchment area approach was developed to address issues with demand and supply inflation that are a result of the overlapping catchment areas produced by the Two-Step Floating Catchment Area [@paezDemandLevelService2019].
+This overlap inflates the demand and generates inaccurate or misleading accessibility estimates [@paezDemandLevelService2019].
+By adjusting the impedance weights, both supply and demand are proportionally allocated which eliminates the double-counting of population that leads to demand and supply inflation [@paezDemandLevelService2019].
+Other benefits of this adjusted method include consideration of competition which can occur when catchment areas overlap, as well as the preservation of system-wide population and level of service.
+Balanced floating catchment area methods have been used to explore accessibility to health care providers [@paezDemandLevelService2019] and COVID-19 health care services [@pereiraGeographicAccessCOVID192021], but have not yet been used in the cycling literature to explore issues of accessibility.
+
+In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data.
+This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are more capable of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences.
+Applying the balanced floating catchment area approach to examine accessibility to Hamilton Bike Share constitutes a location-based measure with an individual component by stratifying according to median household income [@geursAccessibilityEvaluationLanduse2004], but conducting a further sensitivity analysis by adjusting distance thresholds would introduce additional spatio-temporal constraints to evaluate equitable accessibility.
+
+The first step in the balanced floating catchment area method is to allocate the population to be serviced by each Hamilton Bike Share station:
+$$
+P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}
+$$
+
+Next, the level of service at each station (i.e., the maximum number of bicycle racks) is divided by its estimated service population within the established catchment area:
+$$
+L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
+$$
+
+Finally, the accessibility of population cell $i$ is calculated as the weighted sum of the level of service of all stations that can be reached from there according to normalized weights:
+$$
+A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
+$$
+
+The approach uses instead a set of suitably normalized weights as follows:
+$$
+{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
+$$
+
+and
+$$
+{w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
+$$
+
+These weights satisfy the following properties:
+$$
+\sum_{j = 1}^{J} {w^i_{ji}} = 1
+$$
+
+and
+$$
+\sum_{i = 1}^{n} {w^j_{ji}} = 1
+$$
+
+Finally, accessibility is calculated without risk of demand or supply inflation:
+$$
+A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}
+$$
+
+## Pycnophylactic Interpolation
+
+We first plotted the population in all dissemination areas within SoBi Hamilton's core service area.
+Since pycnophylactic interpolation occurs at such a micro scale, we had to ensure that population numbers were not interpolated or allocated to areas where people do not live in Hamilton (i.e., where schools or parks are located).
+To do so, we retrieved shapefiles for various geographic features (see Table 1) from Open Hamilton.
+Figure 3 shows where these spaces are located within SoBi Hamilton's core service area.
+Next, we extracted these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population from dissemination areas to smaller polygons.
+These micro-level polygons are 50 by 50 metres.
+
+```{r data-features, echo=FALSE}
+data_tbl <- data.frame(
+  Feature = c("Hamilton Bike Share Stations", "Golf Courses", "Parks", "Designated Large Employment Areas", "Municipal Parking Lots", "Cemeteries", "Environmentally Sensitive Areas", "Streets", "Educational Institutions", "Places of Worship", "Municipal Service Centres", "Recreation and Community Centres", "Arenas", "Emergency Stations", "Fire Stations", "Police Stations", "Railways", "Hospitals"),
+  Description = c(
+    "The location of stations and the number of racks available at each station.",
+    "The location of City and privately owned golf courses.", 
+    "The location of parks and other green spaces.",
+    "The location of large business parks and industrial lands.",
+    "The location of municipal car parks.",
+    "The location of cemeteries.",
+    "The location of either land or water areas containing natural features or significant ecological functions.",
+    "The street network in Hamilton, including road classification for highways.",
+    "The location of all educational institutions and schools.",
+    "The location of buildings used for religious congregations.",
+    "The location of all municipal service centres, including City Hall.",
+    "The location of all recreation and community centres.",
+    "The location of all indoor arenas.",
+    "The location of all Emergency Management Services (EMS) Ambulance stations.",
+    "The location of all fire stations.",
+    "The location of all police stations.",
+    "The railway network in Hamilton.",
+    "The location of all hospitals."
+  )
+)
+
+kbl(data_tbl) %>%
+  kable_paper(full_width = F) %>%
+  column_spec(1, bold = T, border_right = T) %>%
+  column_spec(2, width = "30em") %>%
+  kable_styling(bootstrap_options = "striped")
+```
+
+## Travel Time Matrix
+
+[BBBike](https://download.bbbike.org/osm/bbbike/) is an online cycle route planner that interfaces with OpenStreetMap.
+We extracted OpenStreetMap data for SoBi Hamilton's service area to calculate walking times from each population cell to nearby bike share stations, using a walking distance of 10km and a walking time of 30 minutes as a threshold.
+A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bike share stations within the maximum threshold.
+
+## Data
+
+All data for this research were accessed from Open Hamilton^[https://open.hamilton.ca/], an online repository of data curated by the City of Hamilton.
 
@@
 
+```{r figure-2, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Population in all dissemination areas in the SoBi Hamilton core service area and within 30 minutes of walking to a bike share station."}
+ggplot() +
+  geom_sf(data = hamilton_da_2016 %>%
+            dplyr::filter(sobi_service == "Yes"),
+          aes(fill = population),
+          color = NA) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_fill_distiller(palette = "YlOrRd", 
+                       direction = 1)
+```
+
+```{r figure-4, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Interpolated population cells in the SoBi Hamilton core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
+The downtown area is outlined in dark green.
+Geographic features have been extracted.
+Dissemination area polygons are outlined in grey."}
+ggplot() + 
+  geom_sf(data = population_50x50,
+          aes(color = population)) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  geom_sf(data = hamilton_da_2016 %>%
+            filter(sobi_service == "Yes"),
+          fill = NA,
+          color = "lightgray") + 
+  geom_sf(data = sobi_service,
+          fill = NA,
+          colour = "black") +
+  geom_sf(data = sobi_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green")
+```
+
+```{r figure-5, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Interpolated population cells in the SoBi Hamilton core service area and within 30 minutes of walking to the core service area.
+Green circles represent the location of the equity stations that were added in 2018.
+Blue circles represent the conventional stations.
+The service area of the PBSP is outlined in black, and dissemination areas from the 2016 Canadian Census are outlined in light gray."}
+ggplot() + 
+  geom_sf(data = population_50x50, 
+          aes(color = population)) +
+  geom_sf(data = hamilton_da_2016 %>%
+            dplyr::filter(sobi_service == "Yes"), 
+          fill = NA, 
+          color = "lightgray") + 
+  geom_sf(data = sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "blue") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          size = 1.5, shape = 21, 
+          fill = "green") +
+  geom_sf(data = sobi_service, 
+          fill = NA, 
+          colour = "black") +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  geom_sf(data = sobi_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  facet_grid(rows = vars(hub_type))
+```
+
+## Accessibility Calculations at Different Thresholds
+
+Consensus regarding the distance that individuals are willing to walk to access a PBSP is lacking, therefore the literature on walking behaviour was consulted to determine the thresholds for the sensitivity analysis.
+Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010].
+For this reason, we experiment with various walking thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
+We repeat these calculations for both the current system and the original system to compare how accessibility changes at different thresholds with the addition of the equity stations.
+
+```{r join-population-to-travel-time-data, echo=FALSE}
+# current system with equity stations
+ttm_walk <- left_join(ttm_walk, 
+                      population_50x50 %>%
+                        st_drop_geometry(),
+                      by = "UID")
+```
+
+```{r join-sobi-station-to-travel-time-data, include=FALSE}
+# current system with equity stations
+ttm_walk <- ttm_walk %>% 
+  left_join(sobi_hubs %>%
+              st_drop_geometry() %>%
+              dplyr::select(OBJECTID, 
+                            RACKS_AMOU),
+            by = c("OBJECTID" = "OBJECTID"))
+```
+
+```{r echo=FALSE}
+#data frames for current and original systems
+names(ttm_walk) <- c('UID', 
+                     'hub', 
+                     'travel_time',
+                     "hub_type",
+                     'population',
+                     'racks')
+```
+
+```{r, echo=FALSE}
+#population_sobi_original <- population_sobi_original[-c(3, 5:6)]
+```
+
+### Varying thresholds
+
+For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility in at most 3 minutes:
+```{r}
+results_3min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional") %>%
+  b2sfca(threshold = 3)
+```
+
+All stations:
+```{r}
+results_3min_all <- ttm_walk %>%
+  b2sfca(threshold = 3)
+```
+
+Sanity check:
+```{r}
+sum(results_3min_conventional$los$los)
+sum(results_3min_conventional$accessibility$accessibility)
+```
+
+Sanity check:
+```{r}
+sum(results_3min_all$los$los)
+sum(results_3min_all$accessibility$accessibility)
+```
+
+For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility in at most 5 minutes:
+```{r}
+results_5min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional") %>%
+  b2sfca(threshold = 5)
+```
+
+All stations:
+```{r}
+results_5min_all <- ttm_walk %>%
+  b2sfca(threshold = 5)
+```
+
+Sanity check:
+```{r}
+sum(results_5min_conventional$los$los)
+sum(results_5min_conventional$accessibility$accessibility)
+```
+
+Sanity check:
+```{r}
+sum(results_5min_all$los$los)
+sum(results_5min_all$accessibility$accessibility)
+```
+
+For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility in at most 10 minutes:
+```{r}
+results_10min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional") %>%
+  b2sfca(threshold = 10)
+```
+
+All stations:
+```{r}
+results_10min_all <- ttm_walk %>%
+  b2sfca(threshold = 10)
+```
+
+Sanity check:
+```{r}
+sum(results_10min_conventional$los$los)
+sum(results_10min_conventional$accessibility$accessibility)
+```
+
+Sanity check:
+```{r}
+sum(results_10min_all$los$los)
+sum(results_10min_all$accessibility$accessibility)
+```
+
+For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility in at most 15 minutes:
+```{r}
+results_15min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional") %>%
+  b2sfca(threshold = 15)
+```
+
+All stations:
+```{r}
+results_15min_all <- ttm_walk %>%
+  b2sfca(threshold = 15)
+```
+
+Sanity check:
+```{r}
+sum(results_15min_conventional$los$los)
+sum(results_15min_conventional$accessibility$accessibility)
+```
+
+Sanity check:
+```{r}
+sum(results_15min_all$los$los)
+sum(results_15min_all$accessibility$accessibility)
+```
+
+Extract results:
+```{r, include=FALSE}
+los <- rbind(data.frame(results_3min_all$los, system = "All stations", threshold = "3 min"),
+             data.frame(results_3min_conventional$los, system = "Conventional", threshold = "3 min"),
+             data.frame(results_5min_all$los, system = "All stations", threshold = "5 min"),
+             data.frame(results_5min_conventional$los, system = "Conventional", threshold = "5 min"),
+             data.frame(results_10min_all$los, system = "All stations", threshold = "10 min"),
+             data.frame(results_10min_conventional$los, system = "Conventional", threshold = "10 min"),
+             data.frame(results_15min_all$los, system = "All stations", threshold = "15 min"),
+             data.frame(results_15min_conventional$los, system = "Conventional", threshold = "15 min"))
+
+accessibility <- rbind(data.frame(results_3min_all$accessibility, system = "All stations", threshold = "3 min"),
+             data.frame(results_3min_conventional$accessibility, system = "Conventional", threshold = "3 min"),
+             data.frame(results_5min_all$accessibility, system = "All stations", threshold = "5 min"),
+             data.frame(results_5min_conventional$accessibility, system = "Conventional", threshold = "5 min"),
+             data.frame(results_10min_all$accessibility, system = "All stations", threshold = "10 min"),
+             data.frame(results_10min_conventional$accessibility, system = "Conventional", threshold = "10 min"),
+             data.frame(results_15min_all$accessibility, system = "All stations", threshold = "15 min"),
+             data.frame(results_15min_conventional$accessibility, system = "Conventional", threshold = "15 min"))
+```
+
+Add geometry:
+```{r}
+los <- los %>%
+  left_join(sobi_hubs %>%
+              dplyr::select(OBJECTID),
+            by = c("hub" = "OBJECTID")) %>%
+  st_as_sf()
+
+accessibility <- accessibility %>%
+  left_join(population_50x50 %>%
+              dplyr::select(UID),
+            by = c("UID" = "UID")) %>%
+  st_as_sf()
+```
+
+```{r, echo=FALSE}
+#accessibility plot for current system with equity stations at 3 minutes walk
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_sf(data = accessibility %>%
+            dplyr::filter(threshold == "3 min"),
+       aes(color = accessibility)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  facet_grid(system ~ threshold)
+```
+
+```{r, echo=FALSE}
+#accessibility plot for current system with equity stations at 5 minutes walk
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_sf(data = accessibility %>%
+            dplyr::filter(threshold == "5 min"),
+       aes(color = accessibility)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  facet_grid(system ~ threshold)
+```
+
+```{r, echo=FALSE}
+#accessibility plot for current system with equity stations at 10 minutes walk
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_sf(data = accessibility %>%
+            dplyr::filter(threshold == "10 min"),
+       aes(color = accessibility)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  facet_grid(system ~ threshold)
+```
+
+```{r, echo=FALSE}
+#accessibility plot for current system with equity stations at 15 minutes walk
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_sf(data = accessibility %>%
+            dplyr::filter(threshold == "15 min"),
+       aes(color = accessibility)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  facet_grid(system ~ threshold)
+```
+
+## Comparison
+
+Include table with total # of people who can reach a station within 3 minutes walk, 5 minutes, 10 minutes, and 15 minutes, to compare accessibility at different thresholds for both current and original system.
+
 Discussion
 ===================
+
+Our analysis demonstrates that the stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing accessibility, albeit only modestly.
+
+By implementing equity stations in more areas with lower median household income in Hamilton, the PBSP has achieved greater horizontal equity by extending the spatial distribution of bicycles across the city.
+Under the the conditions of the original system, which did not have equity stations, the level of service was _X_.
+With the addition of twelve equity stations, the level of service increased to _X_.
+
+However, we found that dissemination areas with lower median household income do not necessarily have greater access to the program and that spatial disparities still persist.
+Under the the conditions of the original system, which did not have equity stations, _X%_ of dissemination areas (or _X%_ of the population) within the core service area with median household income less than 40,000 Canadian dollars could access at least one bicycle share station within 10 minutes walk.
+Only _X%_ of dissemination areas (or _X%_ of the population) with median household income less than 40,000 Canadian dollars could access at least one bicycle share station within 3 or 5 minutes walk.
+
+Similar to other studies, we find that areas with higher median household income have greater accessibility to Hamilton Bike Share.
+Our findings align with other studies that have found disparities in station or bike location along income lines in New York City [@babagoliExploringHealthSpatial2019], Tampa [@chenExploringEquityPerformance2019], and Seattle [@mooneyFreedomStationSpatial2019].
+
+# Study Limitations
+
+This paper did not examine or compare ridership data between conventional and equity stations.
+Therefore, further research is needed to determine whether this encouraged more cycling for low-income individuals living near equity stations.
+Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighbourhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of accessible data to conduct similar analyses for Hamilton Bike Share.
 
@@
 
+The addition of specific equity stations to the public bicycle share program in Hamilton, Ontario had the net effect of increasing accessibility and reducing both horizontal and vertical inequities.
+In particular, accessibility improved the most for _X_ income groups at _X_ thresholds, but the gains were only modest as the threshold increased ( _2.10_ bicycles/person with equity stations vs _1.97_ bicycles/person without equity stations).
+Congestion effects were observed at higher thresholds, with accessibility decreasing significantly once the catchment area is increased to 10 minutes walking ( _X_ ).
+We identified specific areas that would benefit from an increased supply of bicycles which make them ideal candidates for new equity stations.
+In this way, our paper has made contributions in a positive way by applying an intuitive and useful approach to measure accessibility, and in a normative way by serving to inform future investments in cycle infrastructure.
+
+Wang and Lindsey [-@wangNewBikeShare2019] have noted that there is a lack of research that examines how bike share users' behaviour changes as a result of program changes to station locations or improvements in accessibility.
+As such, a logical next step to this research is to examine whether Hamilton Bike Share's equity stations increased ridership or resulted in new memberships in areas that were previously under-served.
+An examination of the types of trips undertaken by residents in these areas would also be informative, such as the study undertaken by Caspi and Norland [-@caspiBikesharingPhiladelphiaLowerincome2019] after bike share stations were implemented in low-income Philadelphia neighbourhoods.
+The bulk of cycling facilities that have been built in Hamilton to date located in the core service area near the conventional stations.
+It would be worthwhile to explore the route choice of bike share trips departing or ending at the equity stations and to identify factors that specifically influence trips from these stations, which would extend existing studies conducted by Scott and colleagues [@luUnderstandingBikeShare2018; @scottWhatFactorsInfluence2019; @scottRouteChoiceBike2021].
+This paper, combined with additional studies such as those conceptualized above, would serve as a valuable case study for Hamilton and other cities with PBSPs who wish to evaluate and address spatial inequities in accessibility and transportation options in urban areas.
+
 References {#references .unnumbered}

### 2021-03-17 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Added downtown areas to data package, and the SoBi hubs have a new identifier for status (some hubs have been deactivated). The code in the manuscript now filters non-active stations in the calculation of accessibility.

@@
 
-```{r}
+```{r invoke-data, include=FALSE}
+data("community_downtown")
 data("hamilton_cma")
@@
 data("ttm_walk")
-```
-
-```{r coordinates-of-equity-stations, include=FALSE}
-# define the coordinates of the equity stations
-# equity_stations <- data.frame(longitude = c(-79.84364993846138, 
-#                                             -79.8362839215856,
-#                                             -79.83144521746212, 
-#                                             -79.82217214949738, 
-#                                             -79.8160211746482, 
-#                                             -79.8192304372373, 
-#                                             -79.82500255113169, 
-#                                             -79.8268599805827, 
-#                                             -79.83888834759689, 
-#                                             -79.82740849222654, 
-#                                             -79.82342541216097, 
-#                                             -79.84240606419081), 
-#                               latitude = c(43.265340158000654, 
-#                                            43.25632467284136, 
-#                                            43.25528943845642, 
-#                                            43.25291050074978, 
-#                                            43.25147804403345, 
-#                                            43.24444659758296, 
-#                                            43.2421627336722, 
-#                                            43.24634382671116,
-#                                            43.249432741404554,
-#                                            43.251126444296524, 
-#                                            43.24892849896579, 
-#                                            43.25831920064561))
-```
-
-```{r, include=FALSE}
-#equity_stations <- st_as_sf(equity_stations, coords = c("longitude", "latitude"), 
-    #crs = st_crs(sobi_service), agr = "constant")
-```
-
-```{r load-data, include=FALSE}
-#load("accessibility.RData") #full accessibility table for SoBi system
-#load("accessibility_map_original.RData") #map of accessibility in original SoBi system
-#load("accessibility_map_equity.RData") #map of accessibility in current system with ERI
-# load("hamilton_cma.RData")
-# load("hamilton_da_2016.RData")
-# load("hamilton_da_sobi.RData")
-# load("mhi.RData") #median household income by DA
-# load("population_sobi.RData") #interpolated population in SoBi area
-#load("population_sobi_original.RData") #population in original system
-# load("sobi_hubs.RData") #location of SoBi hubs, minus Van Wagner's beach
-# load("sobi_hubs_original.RData") #SoBi hubs, minus ERI hubs
-# load("sobi_service.RData") #core service area for SoBi
-# load("ttm_walk.RData") #travel time matrix for interpolated population in SoBi area
-# load("ttm_walk_original.RData") #travel time matrix for system without ERI hubs
 ```
@@
 
-```{r filter-eri-hubs, include=FALSE}
-# eri_hubs <- sobi_hubs %>% 
-#              filter(OBJECTID>=117 & OBJECTID<=128)
-```
-
-```{r, include=FALSE}
-sobi_downtown <- st_read("Community_Downtown.shp")
-```
-
 ```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The spatial distribution of bike share stations in Hamilton, Ontario.
@@
           color = "black") +
-  geom_sf(data = sobi_downtown %>%
+  geom_sf(data = community_downtown %>%
             dplyr::filter(NAME == "Hamilton"),
@@
   geom_sf(data = sobi_hubs %>% 
-            dplyr::filter(hub_type == "Conventional"), 
+            dplyr::filter(hub_type == "Conventional" & 
+                            hub_status == "Active"), # Remove deactivated hubs
           size = 1.5, 
@@
           colour = "black") +
-  geom_sf(data = sobi_downtown %>%
+  geom_sf(data = community_downtown %>%
             dplyr::filter(NAME == "Hamilton"),
@@
   geom_sf(data = sobi_hubs %>% 
-            dplyr::filter(hub_type == "Conventional"), 
+            dplyr::filter(hub_type == "Conventional" &
+                            hub_status == "Active"), 
           size = 1.5, 
@@
                         direction = 1) +
-  geom_sf(data = sobi_downtown %>%
+  geom_sf(data = community_downtown %>%
             dplyr::filter(NAME == "Hamilton"),
@@
 #data frames for current and original systems
-names(ttm_walk) <- c('UID', 
-                     'hub', 
-                     'travel_time',
+names(ttm_walk) <- c("UID", 
+                     "hub", 
+                     "travel_time",
                      "hub_type",
-                     'population',
-                     'racks')
-```
-
-```{r, echo=FALSE}
-#population_sobi_original <- population_sobi_original[-c(3, 5:6)]
+                     "hub_status",
+                     "population",
+                     "racks")
 ```
@@
 results_3min_conventional <- ttm_walk %>%
-  dplyr::filter(hub_type == "Conventional") %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
   b2sfca(threshold = 3)
@@
 results_3min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
   b2sfca(threshold = 3)
@@
 results_5min_conventional <- ttm_walk %>%
-  dplyr::filter(hub_type == "Conventional") %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
   b2sfca(threshold = 5)
@@
 results_5min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
   b2sfca(threshold = 5)
@@
 results_10min_conventional <- ttm_walk %>%
-  dplyr::filter(hub_type == "Conventional") %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
   b2sfca(threshold = 10)
@@
 results_10min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
   b2sfca(threshold = 10)
@@
 results_15min_conventional <- ttm_walk %>%
-  dplyr::filter(hub_type == "Conventional") %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
   b2sfca(threshold = 15)
@@
 results_15min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
   b2sfca(threshold = 15)

### 2021-03-17 sobiEquity/raw-data-processing/00-Data-Processing-for-sobiEquity.Rmd
Added downtown areas to data package, and the SoBi hubs have a new identifier for status (some hubs have been deactivated). The code in the manuscript now filters non-active stations in the calculation of accessibility.

@@
 
+INFO
+
+### Community Downtown
+
+INFO
+
 ## Provenance of data: census
@@
 hospitals <- st_read("input-data-files/Hospitals.shp")
+```
+
+Read community downtown:
+```{r}
+community_downtown <- st_read("input-data-files/Community_Downtown.shp")
 ```
@@
   geom_sf(data = sobi_hubs,
-          aes(shape = hub_type, 
-              size = RACKS_AMOU)) +
+          aes(size = RACKS_AMOU)) +
   scale_color_distiller(palette = "YlOrRd", 
@@
 
+Identify hubs that were deactivated:
+```{r}
+sobi_hubs <- sobi_hubs %>%
+  dplyr::mutate(hub_status = case_when(OBJECTID == 114 ~ "Deactivated",
+                                       OBJECTID == 115 ~ "Deactivated",
+                                       OBJECTID == 116 ~ "Deactivated",
+                                       TRUE ~ "Active"),
+                hub_status = factor(hub_status))
+
+```
+
 Plot again:
@@
               dplyr::select(OBJECTID,
-                            hub_type),
+                            hub_type,
+                            hub_status),
             by = "OBJECTID")
@@
 ```{r}
+save(community_downtown, file = "output-data-files/community_downtown.RData", compress = "xz")
 save(hamilton_cma, file = "output-data-files/hamilton_cma.RData", compress = "xz")

### 2021-03-17 sobiEquity/R/sobiEquity.R

@@ -21,3 +21,4 @@ NULL
 #' @usage data(hamilton_cma)
 #' @source Statistics Canada
 "hamilton_cma"
+

### 2021-03-18 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Revised draft

@@
 
+```{r, include=FALSE}
+rm(list = ls())
+```
+
 ```{r load-packages, include=FALSE}
@@
 library(ggplot2)
-theme_set(theme_bw())
+library(biscale)
+library(cowplot)
 ```
@@
 Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021].
-The core service area spans 35 km^2 of the city and _approximately 130,000 can reach a bike share hub within 30 minutes of walking_ [see Figure 1].
+The core service area spans 35 km^2 of the city and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure 1].
 This represents roughly one fifth of the population in Hamilton, according to the 2016 Canadian Census.
@@
 
-## Accessibility Calculations at Different Thresholds
+## Accessibility at Different Thresholds
 
@@
 
-### Varying thresholds
+### Three Minutes
 
-For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility in at most 3 minutes:
-```{r}
+For the system without the equity stations, we remove the equity stations from the travel time matrix before calculating accessibility in at most 3 minutes walk.
+
+```{r min-conventional, include=FALSE}
 results_3min_conventional <- ttm_walk %>%
@@
 
-All stations:
-```{r}
+```{r min-all, include=FALSE}
 results_3min_all <- ttm_walk %>%
@@
 
-Sanity check:
-```{r}
+```{r min-sum-conventional, echo=FALSE}
 sum(results_3min_conventional$los$los)
@@
 
-Sanity check:
-```{r}
+```{r min-sum-all, echo=FALSE}
 sum(results_3min_all$los$los)
@@
 
-For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility in at most 5 minutes:
-```{r}
+### Five Minutes
+
+For the system without the equity stations, we remove the equity stations from the travel time matrix before calculating accessibility in at most 5 minutes walk.
+
+```{r med-conventional, include=FALSE}
 results_5min_conventional <- ttm_walk %>%
@@
 
-All stations:
-```{r}
+```{r med-all, include=FALSE}
 results_5min_all <- ttm_walk %>%
@@
 
-Sanity check:
-```{r}
+```{r, med-sum-conventional, echo=FALSE}
 sum(results_5min_conventional$los$los)
@@
 
-Sanity check:
-```{r}
+```{r med-sum-all, echo=FALSE}
 sum(results_5min_all$los$los)
@@
 
-For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility in at most 10 minutes:
-```{r}
+### Ten Minutes
+
+For the system without the equity stations, we remove the equity stations from the travel time matrix before calculating accessibility in at most 10 minutes.
+
+```{r avg-conventional, include=FALSE}
 results_10min_conventional <- ttm_walk %>%
@@
 
-All stations:
-```{r}
+```{r avg-all, include=FALSE}
 results_10min_all <- ttm_walk %>%
@@
 
-Sanity check:
-```{r}
+```{r avg-sum-conventional, echo=FALSE}
 sum(results_10min_conventional$los$los)
@@
 
-Sanity check:
-```{r}
+```{r, avg-sum-all, echo=FALSE}
 sum(results_10min_all$los$los)
@@
 
-For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility in at most 15 minutes:
-```{r}
+### Fifteen Minutes
+
+For the system without the equity stations, we remove the equity stations from the travel time matrix before calculating accessibility in at most 15 minutes walk.
+
+```{r max-conventional, include=FALSE}
 results_15min_conventional <- ttm_walk %>%
@@
 
-All stations:
-```{r}
+```{r max-all, include=FALSE}
 results_15min_all <- ttm_walk %>%
@@
 
-Sanity check:
-```{r}
+```{r max-sum-conventional, echo=FALSE}
 sum(results_15min_conventional$los$los)
@@
 
-Sanity check:
-```{r}
+```{r max-sum-all, echo=FALSE}
 sum(results_15min_all$los$los)
@@
 
-Extract results:
 ```{r, include=FALSE}
@@
 
-Add geometry:
-```{r}
+```{r, include=FALSE}
 los <- los %>%
@@
 
-```{r, echo=FALSE}
-#accessibility plot for current system with equity stations at 3 minutes walk
+```{r figure-6, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
 ggplot() +
@@
 
-```{r, echo=FALSE}
+```{r figure-7, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 5 minutes walk (average threshold) compared between current system with equity stations and the original system without equity stations."}
 #accessibility plot for current system with equity stations at 5 minutes walk
@@
 
-```{r, echo=FALSE}
+```{r figure-8, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) compared between current system with equity stations and the original system without equity stations."}
 #accessibility plot for current system with equity stations at 10 minutes walk
@@
 
-```{r, echo=FALSE}
+```{r figure-9, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 15 minutes walk (extreme threshold) compared between current system with equity stations and the original system without equity stations."}
 #accessibility plot for current system with equity stations at 15 minutes walk
@@
 Include table with total # of people who can reach a station within 3 minutes walk, 5 minutes, 10 minutes, and 15 minutes, to compare accessibility at different thresholds for both current and original system.
+
+## Accessibility and Median Household Income
+
+```{r, include=FALSE}
+# drop geometry
+test <- accessibility %>% st_drop_geometry()
+
+# join accessibility to population cells
+test <- test %>% left_join(population_50x50, by = "UID")
+```
+
+```{r, include=FALSE}
+test0 <- population_50x50 %>%
+  st_join(hamilton_da_2016 %>% dplyr::select(DA))
+```
+
+```{r, include=FALSE}
+test <- test0 %>% left_join(test, by = "UID")
+```
+
+```{r, include=FALSE}
+# object is a combination of accessibility, population_50x50, with DAs
+test <- test %>% drop_na()
+```
+
+```{r, include=FALSE}
+test <- subset(test, select = -c(1)) %>% dplyr::select(-geometry.y)
+```
+
+```{r, include=FALSE}
+accessibility <- test
+```
+
+```{r, include=FALSE}
+min_all <- accessibility %>%
+            dplyr::filter(threshold == "3 min") %>% 
+            dplyr::filter(system == "All stations")
+
+min_con <- accessibility %>%
+            dplyr::filter(threshold == "3 min") %>%
+            dplyr::filter(system == "Conventional")
+
+avg_all <- accessibility %>%
+            dplyr::filter(threshold == "5 min") %>% 
+            dplyr::filter(system == "All stations")
+
+avg_con <- accessibility %>%
+            dplyr::filter(threshold == "5 min") %>%
+            dplyr::filter(system == "Conventional")
+
+max_all <- accessibility %>%
+            dplyr::filter(threshold == "10 min") %>% 
+            dplyr::filter(system == "All stations")
+
+max_con <- accessibility %>%
+            dplyr::filter(threshold == "10 min") %>%
+            dplyr::filter(system == "Conventional")
+
+ext_all <- accessibility %>%
+            dplyr::filter(threshold == "15 min") %>% 
+            dplyr::filter(system == "All stations")
+
+ext_con <- accessibility %>%
+            dplyr::filter(threshold == "15 min") %>%
+            dplyr::filter(system == "Conventional")
+
+```
+
+```{r, include=FALSE}
+# aggregate population and accessibility by DA
+min_all <- min_all %>%
+           group_by(DA) %>%
+             summarize(population = sum(population.y),
+             accessibility = sum(accessibility)) %>%
+           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
+
+min_con <- min_con %>%
+           group_by(DA) %>%
+             summarize(population = sum(population.y),
+             accessibility = sum(accessibility)) %>%
+           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
+
+avg_all <- avg_all %>%
+           group_by(DA) %>%
+             summarize(population = sum(population.y),
+             accessibility = sum(accessibility)) %>%
+           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
+
+avg_con <- avg_con %>%
+           group_by(DA) %>%
+             summarize(population = sum(population.y),
+             accessibility = sum(accessibility)) %>%
+           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
+
+max_all <- max_all %>%
+           group_by(DA) %>%
+             summarize(population = sum(population.y),
+             accessibility = sum(accessibility)) %>%
+           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
+
+max_con <- max_con %>%
+           group_by(DA) %>%
+             summarize(population = sum(population.y),
+             accessibility = sum(accessibility)) %>%
+           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
+
+ext_all <- ext_all %>%
+           group_by(DA) %>%
+             summarize(population = sum(population.y),
+             accessibility = sum(accessibility)) %>%
+           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
+
+ext_con <- ext_con %>%
+           group_by(DA) %>%
+             summarize(population = sum(population.y),
+             accessibility = sum(accessibility)) %>%
+           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
+```
+
+```{r}
+min_all <- min_all %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_2016 %>% 
+              dplyr::select(DA, geometry),
+            by = "DA") %>%
+  st_as_sf()
+
+min_con <- min_con %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_2016 %>% 
+              dplyr::select(DA, geometry),
+            by = "DA") %>%
+  st_as_sf()
+
+avg_all <- avg_all %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_2016 %>% 
+              dplyr::select(DA, geometry),
+            by = "DA") %>%
+  st_as_sf()
+
+avg_con <- avg_con %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_2016 %>% 
+              dplyr::select(DA, geometry),
+            by = "DA") %>%
+  st_as_sf()
+
+max_all <- max_all %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_2016 %>% 
+              dplyr::select(DA, geometry),
+            by = "DA") %>%
+  st_as_sf()
+
+max_con <- max_con %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_2016 %>% 
+              dplyr::select(DA, geometry),
+            by = "DA") %>%
+  st_as_sf()
+
+ext_all <- ext_all %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_2016 %>% 
+              dplyr::select(DA, geometry),
+            by = "DA") %>%
+  st_as_sf()
+
+ext_con <- ext_con %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_2016 %>% 
+              dplyr::select(DA, geometry),
+            by = "DA") %>%
+  st_as_sf()
+```
+
+```{r}
+data_min_all <- bi_class(min_all,
+                 x = accessibility, 
+                 y = median_total_income, 
+                 style = "quantile", 
+                 dim = 3) %>%
+  st_as_sf()
+
+data_min_con <- bi_class(min_con,
+                 x = accessibility, 
+                 y = median_total_income, 
+                 style = "quantile", 
+                 dim = 3) %>%
+  st_as_sf()
+
+data_avg_all <- bi_class(avg_all,
+                 x = accessibility, 
+                 y = median_total_income, 
+                 style = "quantile", 
+                 dim = 3) %>%
+  st_as_sf()
+
+data_avg_con <- bi_class(avg_con,
+                 x = accessibility, 
+                 y = median_total_income, 
+                 style = "quantile", 
+                 dim = 3) %>%
+  st_as_sf()
+
+data_max_all <- bi_class(max_all,
+                 x = accessibility, 
+                 y = median_total_income, 
+                 style = "quantile", 
+                 dim = 3) %>%
+  st_as_sf()
+
+data_max_con <- bi_class(max_con,
+                 x = accessibility, 
+                 y = median_total_income, 
+                 style = "quantile", 
+                 dim = 3) %>%
+  st_as_sf()
+
+data_ext_all <- bi_class(ext_all,
+                 x = accessibility, 
+                 y = median_total_income, 
+                 style = "quantile", 
+                 dim = 3) %>%
+  st_as_sf()
+
+data_ext_con <- bi_class(ext_con,
+                 x = accessibility, 
+                 y = median_total_income, 
+                 style = "quantile", 
+                 dim = 3) %>%
+  st_as_sf()
+```
+```{r figure-10, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) varying by median total income in the current system with equity stations."}
+map_min_all <- ggplot() +
+  geom_sf(data = data_min_all, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", dim = 3) +
+  bi_theme()
+
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(map_min_all, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-11, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) varying by median total income in the original system with conventional stations."}
+map_min_con <- ggplot() +
+  geom_sf(data = data_min_con, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", dim = 3) +
+  bi_theme()
+
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(map_min_con, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-12, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 5 minutes walk (average threshold) varying by median total income in the current system with equity stations."}
+map_avg_all <- ggplot() +
+  geom_sf(data = data_avg_all, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", dim = 3) +
+  bi_theme()
+
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(map_avg_all, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-13, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 5 minutes walk (average threshold) varying by median total income in the original system with conventional stations."}
+map_avg_con <- ggplot() +
+  geom_sf(data = data_avg_con, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", dim = 3) +
+  bi_theme()
+
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(map_avg_con, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-14, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) varying by median total income in the current system with equity stations."}
+map_max_all <- ggplot() +
+  geom_sf(data = data_max_all, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", dim = 3) +
+  bi_theme()
+
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(map_max_all, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-15, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) varying by median total income in the original system with conventional stations."}
+map_max_con <- ggplot() +
+  geom_sf(data = data_max_con, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", dim = 3) +
+  bi_theme()
+
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(map_max_con, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-16, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 15 minutes walk (extreme threshold) varying by median total income in the current system with equity stations."}
+map_ext_all <- ggplot() +
+  geom_sf(data = data_ext_all, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", dim = 3) +
+  bi_theme()
+
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(map_ext_all, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-17, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 15 minutes walk (extreme threshold) varying by median total income in the original system with conventional stations."}
+map_ext_con <- ggplot() +
+  geom_sf(data = data_ext_con, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", dim = 3) +
+  bi_theme()
+
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(map_ext_con, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r, include=FALSE}
+sum(population_50x50$population)
+sum(min_all$population)
+sum(avg_all$population)
+sum(max_all$population)
+sum(ext_all$population)
+sum(min_con$population)
+sum(avg_con$population)
+sum(max_con$population)
+sum(ext_con$population)
+```
+
+```{r}
+summary(data_max_all)
+```
+
+```{r}
+demographics_minimum <- rbind(data.frame(min_all %>%  
+                                        dplyr::filter(median_total_income >= "38486") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 1", system = "All Stations"),
+             data.frame(min_all %>% dplyr::filter(median_total_income >= "46336") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 2", system = "All Stations"),
+             data.frame(min_all %>% dplyr::filter(median_total_income >= "56640") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 3", system = "All Stations"),
+             data.frame(min_con %>% dplyr::filter(median_total_income >= "38288") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 1", system = "Conventional"),
+             data.frame(min_con %>% dplyr::filter(median_total_income >= "45696") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 2", system = "Conventional"),
+             data.frame(min_con %>% dplyr::filter(median_total_income >= "57664") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 3", system = "Conventional"))
+```
+
+```{r}
+demographics_maximum <- rbind(data.frame(avg_all %>%  
+                                        dplyr::filter(median_total_income >= "38752") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 1", system = "All Stations"),
+             data.frame(avg_all %>% dplyr::filter(median_total_income >= "47040") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 2", system = "All Stations"),
+             data.frame(avg_all %>% dplyr::filter(median_total_income >= "58912") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 3", system = "All Stations"),
+             data.frame(avg_con %>% dplyr::filter(median_total_income >= "38422") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 1", system = "Conventional"),
+             data.frame(avg_con %>% dplyr::filter(median_total_income >= "45696") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 2", system = "Conventional"),
+             data.frame(avg_con %>% dplyr::filter(median_total_income >= "59264") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 3", system = "Conventional"))
+```
+
+```{r}
+demographics_minimum <- rbind(data.frame(min_all %>%  
+                                        dplyr::filter(median_total_income >= "38486") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 1", system = "All Stations"),
+             data.frame(min_all %>% dplyr::filter(median_total_income >= "46336") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 2", system = "All Stations"),
+             data.frame(min_all %>% dplyr::filter(median_total_income >= "56640") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 3", system = "All Stations"),
+             data.frame(min_con %>% dplyr::filter(median_total_income >= "38288") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 1", system = "Conventional"),
+             data.frame(min_con %>% dplyr::filter(median_total_income >= "45696") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 2", system = "Conventional"),
+             data.frame(min_con %>% dplyr::filter(median_total_income >= "57664") %>% summarise(population = sum(population)), accessibility = "min", income = "Quantile 3", system = "Conventional"))
+```
+
+```{r, include=FALSE}
+sum(population_50x50$population)
+sum(min_all$population)
+sum(avg_all$population)
+sum(max_all$population)
+sum(ext_all$population)
+sum(min_con$population)
+sum(avg_con$population)
+sum(max_con$population)
+sum(ext_con$population)
+```
+
+```{r table-2, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Population that can reach at least one bike share station at different thresholds in the current system with equity stations and original system with only conventional stations."}
+data_tbl <- data.frame(
+  System = c("All Stations", "Conventional"),
+  Minimum = c("48,725", "43,238"),
+  Average = c("84,655", "75,026"),
+  Maximum = c("111,680", "104,298"),
+  Extreme = c("118,581", "115,344")
+)
+
+kbl(data_tbl) %>%
+  kable_paper(full_width = F) %>%
+  column_spec(1, bold = T, border_right = T) %>%
+  kable_styling(bootstrap_options = "striped") %>%
+  add_header_above(c(" " = 1, "3 Minutes" = 1, "5 Minutes" = 1, "10 Minutes" = 1, "15 Minutes" = 1))
+```
 

### 2021-03-18 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Revised the code to aggregate accessibility levels by income quintiles

@@
 ```{r med-conventional, include=FALSE}
-results_5min_conventional <- ttm_walk %>%
+results_6min_conventional <- ttm_walk %>%
   dplyr::filter(hub_type == "Conventional" &
                   hub_status == "Active") %>%
-  b2sfca(threshold = 5)
+  b2sfca(threshold = 6)
 ```
@@
 ```{r med-all, include=FALSE}
-results_5min_all <- ttm_walk %>%
+results_6min_all <- ttm_walk %>%
   dplyr::filter(hub_status == "Active") %>%
-  b2sfca(threshold = 5)
+  b2sfca(threshold = 6)
 ```
@@
 ```{r, med-sum-conventional, echo=FALSE}
-sum(results_5min_conventional$los$los)
-sum(results_5min_conventional$accessibility$accessibility)
+sum(results_6min_conventional$los$los)
+sum(results_6min_conventional$accessibility$accessibility)
 ```
@@
 sum(results_5min_all$accessibility$accessibility)
+```
+
+```{r include=FALSE}
+# Testing different thresholds to check how accessibility changes
+accessibility_test <- data.frame(threshold = c(3:15), 
+                                 accessibility = numeric(13))
+for(i in 1:13){
+  results_test <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = i +2)
+  accessibility_test$accessibility[i] <- sum(results_test$accessibility$accessibility)
+}
+
+ggplot(accessibility_test, 
+       aes(x = threshold, y = accessibility)) +
+  geom_line()
 ```
@@
 # drop geometry
-test <- accessibility %>% st_drop_geometry()
+test <- accessibility
 
 # join accessibility to population cells
-test <- test %>% left_join(population_50x50, by = "UID")
-```
-
-```{r, include=FALSE}
-test0 <- population_50x50 %>%
-  st_join(hamilton_da_2016 %>% dplyr::select(DA))
-```
-
-```{r, include=FALSE}
-test <- test0 %>% left_join(test, by = "UID")
-```
-
-```{r, include=FALSE}
-# object is a combination of accessibility, population_50x50, with DAs
-test <- test %>% drop_na()
-```
-
-```{r, include=FALSE}
-test <- subset(test, select = -c(1)) %>% dplyr::select(-geometry.y)
-```
-
-```{r, include=FALSE}
-accessibility <- test
-```
-
-```{r, include=FALSE}
-min_all <- accessibility %>%
-            dplyr::filter(threshold == "3 min") %>% 
-            dplyr::filter(system == "All stations")
-
-min_con <- accessibility %>%
-            dplyr::filter(threshold == "3 min") %>%
-            dplyr::filter(system == "Conventional")
-
-avg_all <- accessibility %>%
-            dplyr::filter(threshold == "5 min") %>% 
-            dplyr::filter(system == "All stations")
-
-avg_con <- accessibility %>%
-            dplyr::filter(threshold == "5 min") %>%
-            dplyr::filter(system == "Conventional")
-
-max_all <- accessibility %>%
-            dplyr::filter(threshold == "10 min") %>% 
-            dplyr::filter(system == "All stations")
-
-max_con <- accessibility %>%
-            dplyr::filter(threshold == "10 min") %>%
-            dplyr::filter(system == "Conventional")
-
-ext_all <- accessibility %>%
-            dplyr::filter(threshold == "15 min") %>% 
-            dplyr::filter(system == "All stations")
-
-ext_con <- accessibility %>%
-            dplyr::filter(threshold == "15 min") %>%
-            dplyr::filter(system == "Conventional")
-
-```
-
-```{r, include=FALSE}
-# aggregate population and accessibility by DA
-min_all <- min_all %>%
-           group_by(DA) %>%
-             summarize(population = sum(population.y),
-             accessibility = sum(accessibility)) %>%
-           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
-
-min_con <- min_con %>%
-           group_by(DA) %>%
-             summarize(population = sum(population.y),
-             accessibility = sum(accessibility)) %>%
-           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
-
-avg_all <- avg_all %>%
-           group_by(DA) %>%
-             summarize(population = sum(population.y),
-             accessibility = sum(accessibility)) %>%
-           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
-
-avg_con <- avg_con %>%
-           group_by(DA) %>%
-             summarize(population = sum(population.y),
-             accessibility = sum(accessibility)) %>%
-           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
-
-max_all <- max_all %>%
-           group_by(DA) %>%
-             summarize(population = sum(population.y),
-             accessibility = sum(accessibility)) %>%
-           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
-
-max_con <- max_con %>%
-           group_by(DA) %>%
-             summarize(population = sum(population.y),
-             accessibility = sum(accessibility)) %>%
-           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
-
-ext_all <- ext_all %>%
-           group_by(DA) %>%
-             summarize(population = sum(population.y),
-             accessibility = sum(accessibility)) %>%
-           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
-
-ext_con <- ext_con %>%
-           group_by(DA) %>%
-             summarize(population = sum(population.y),
-             accessibility = sum(accessibility)) %>%
-           st_join(hamilton_da_2016 %>% dplyr::select(median_total_income))
+test <- test %>% st_join(population_50x50 %>% dplyr::select(-UID)) %>%
+  rename(population50x50 = population)
 ```
@@
 ```{r}
-min_all <- min_all %>%
+test <- test %>%
+  st_join(hamilton_da_2016)
+```
+
+```{r}
+agg_accessibility_da <- test %>%
+  group_by(DA, 
+           system,
+           threshold) %>%
+  summarize(accessibility = sum(accessibility),
+            median_total_income = first(median_total_income), 
+            .groups = "drop") %>%
   st_drop_geometry() %>%
-  left_join(hamilton_da_2016 %>% 
-              dplyr::select(DA, geometry),
-            by = "DA") %>%
+  left_join(hamilton_da_2016 %>% dplyr::select(DA, geometry),
+            by = c("DA" = "DA")) %>%
   st_as_sf()
-
-min_con <- min_con %>%
-  st_drop_geometry() %>%
-  left_join(hamilton_da_2016 %>% 
-              dplyr::select(DA, geometry),
-            by = "DA") %>%
-  st_as_sf()
-
-avg_all <- avg_all %>%
-  st_drop_geometry() %>%
-  left_join(hamilton_da_2016 %>% 
-              dplyr::select(DA, geometry),
-            by = "DA") %>%
-  st_as_sf()
-
-avg_con <- avg_con %>%
-  st_drop_geometry() %>%
-  left_join(hamilton_da_2016 %>% 
-              dplyr::select(DA, geometry),
-            by = "DA") %>%
-  st_as_sf()
-
-max_all <- max_all %>%
-  st_drop_geometry() %>%
-  left_join(hamilton_da_2016 %>% 
-              dplyr::select(DA, geometry),
-            by = "DA") %>%
-  st_as_sf()
-
-max_con <- max_con %>%
-  st_drop_geometry() %>%
-  left_join(hamilton_da_2016 %>% 
-              dplyr::select(DA, geometry),
-            by = "DA") %>%
-  st_as_sf()
-
-ext_all <- ext_all %>%
-  st_drop_geometry() %>%
-  left_join(hamilton_da_2016 %>% 
-              dplyr::select(DA, geometry),
-            by = "DA") %>%
-  st_as_sf()
-
-ext_con <- ext_con %>%
-  st_drop_geometry() %>%
-  left_join(hamilton_da_2016 %>% 
-              dplyr::select(DA, geometry),
-            by = "DA") %>%
-  st_as_sf()
 ```
@@
 ```{r}
-data_min_all <- bi_class(min_all,
-                 x = accessibility, 
+data_map <- agg_accessibility_da %>%
+            dplyr::filter(threshold == "3 min", 
+                          system == "All stations") %>%
+  bi_class(x = accessibility, 
            y = median_total_income, 
@@
 
-data_min_con <- bi_class(min_con,
-                 x = accessibility, 
-                 y = median_total_income, 
-                 style = "quantile", 
-                 dim = 3) %>%
-  st_as_sf()
-
-data_avg_all <- bi_class(avg_all,
-                 x = accessibility, 
-                 y = median_total_income, 
-                 style = "quantile", 
-                 dim = 3) %>%
-  st_as_sf()
-
-data_avg_con <- bi_class(avg_con,
-                 x = accessibility, 
-                 y = median_total_income, 
-                 style = "quantile", 
-                 dim = 3) %>%
-  st_as_sf()
-
-data_max_all <- bi_class(max_all,
-                 x = accessibility, 
-                 y = median_total_income, 
-                 style = "quantile", 
-                 dim = 3) %>%
-  st_as_sf()
-
-data_max_con <- bi_class(max_con,
-                 x = accessibility, 
-                 y = median_total_income, 
-                 style = "quantile", 
-                 dim = 3) %>%
-  st_as_sf()
-
-data_ext_all <- bi_class(ext_all,
-                 x = accessibility, 
-                 y = median_total_income, 
-                 style = "quantile", 
-                 dim = 3) %>%
-  st_as_sf()
-
-data_ext_con <- bi_class(ext_con,
-                 x = accessibility, 
-                 y = median_total_income, 
-                 style = "quantile", 
-                 dim = 3) %>%
-  st_as_sf()
-```
+map_min_all <- ggplot() +
+  geom_sf(data = data_map,
+          mapping = aes(fill = bi_class), 
+          color = "white", 
+          size = 0.1, 
+          show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", 
+                dim = 3) +
+  bi_theme()
+
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(map_min_all, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+Coming back to the question of accessibility by income:
+```{r}
+# Find the income quintiles of the population served by SoBi
+iq <- hamilton_da_2016 %>% 
+  dplyr::filter(sobi_service == "Yes") %>%
+  pull(median_total_income) %>%
+  quantile(c(0, 0.2, 0.4, 0.6, 0.8, 1), na.rm = TRUE)
+```
+
+```{r}
+# Add a variable to identify the income 
+agg_accessibility_income <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"))
+
+inc_stats_All_10 <- agg_accessibility_income %>%
+  st_drop_geometry() %>%
+  dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(accessibility = sum(accessibility),
+            population = sum(population50x50))
+```
+
+```{r}
+# Add a variable to identify the income 
+agg_accessibility_income <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"))
+
+inc_stats_Conv_10 <- agg_accessibility_income %>%
+  st_drop_geometry() %>%
+  dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(accessibility = sum(accessibility),
+            population = sum(population50x50))
+```
+
+```{r}
+rbind(inc_stats_All_10,  
+      inc_stats_Conv_10)
+```
+
 ```{r figure-10, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) varying by median total income in the current system with equity stations."}

### 2021-03-19 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Revised draft with complete calculations

--- 2021-03-17 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
@@
 
-```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The spatial distribution of bike share stations in Hamilton, Ontario.
+```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bike share stations in Hamilton, Ontario.
 The service area of Hamilton Bike Share is outlined in blue and the city's downtown core is outlined in light blue.
@@
                             hub_status == "Active"), # Remove deactivated hubs
-          size = 1.5, 
+          aes(size = RACKS_AMOU),
           shape = 21,
@@
             dplyr::filter(hub_type == "ERI"), 
-          size = 1.5, shape = 21, 
+          aes(size = RACKS_AMOU), 
+          shape = 21,
           fill = "green")
@@
 
-## Accessibility Calculations at Different Thresholds
+## Accessibility by Distance
 
@@
 Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010].
+As seen in Table \ref{tab:accessibility-sensibility}, we find that accessibility increases with a threshold between two and four minutes, but is then maximized at 5 minutes.
+At some stations, Hamilton Bike Share will depict a map to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are anticipated to walk.
+Accessibility decreases significantly after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each bike share station.
+
 For this reason, we experiment with various walking thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
-We repeat these calculations for both the current system and the original system to compare how accessibility changes at different thresholds with the addition of the equity stations.
-
-```{r join-population-to-travel-time-data, echo=FALSE}
+We categorize these thresholds as minimum, average, maximum, and extreme, respectively.
+At each threshold, we compare accessibility between the current system and the original system to examine the contribution of the additional equity stations.
+
+```{r accessibility-sensibility, echo=FALSE}
+# Testing different thresholds to check how accessibility changes
+accessibility_test <- data.frame(threshold = c(3:15), 
+                                 accessibility = numeric(13))
+for(i in 1:13){
+  results_test <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                hub_status == "Active") %>%
+  b2sfca(threshold = i +2)
+  accessibility_test$accessibility[i] <- sum(results_test$accessibility$accessibility)
+}
+
+ggplot(accessibility_test, 
+       aes(x = threshold, y = accessibility)) +
+  geom_line()
+```
+
+```{r join-population-to-travel-time-data, include=FALSE}
 # current system with equity stations
@@
 
-```{r echo=FALSE}
+```{r, include=FALSE}
 #data frames for current and original systems
@@
 
-### Varying thresholds
-
-For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility in at most 3 minutes:
-```{r}
+### Three Minutes
+
+At this threshold, we find that there are 25.2 bicycle racks per person in the original system.
+With the addition of equity stations, there are now 25.4 bicycles per person.
+Figure \ref{fig:figure-6} presents a comparison of accessibility between the systems.
+Accessibility is fairly uniform, with the exception of two small areas near the university and the waterfront park where accessibility is slightly higher.
+Equity stations are shown in green, which helps to demonstrate how they increase accessibility between the systems.
+
+```{r min-conventional, include=FALSE}
 results_3min_conventional <- ttm_walk %>%
@@
 ```
-
-For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility in at most 5 minutes:
-```{r}
+```{r figure-6, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_sf(data = accessibility %>%
+            dplyr::filter(threshold == "3 min"),
+       aes(color = accessibility)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "PuRd", 
+                        direction = 1) +
+  facet_grid(system ~ threshold)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          size = 1.5, shape = 21, 
+          fill = "green")
+```
+
+### Five Minutes
+
+At this threshold, we find that there are 68.6 bicycle racks per person in the original system.
+With the addition of equity stations, there are now 68.8 bicycles per person.
+Figure \ref{fig:figure-7} presents a comparison of accessibility between the systems.
+Again, accessibility is fairly uniform, with the exception of one area near the waterfront park where accessibility is higher.
+
+```{r med-conventional, include=FALSE}
 results_5min_conventional <- ttm_walk %>%
@@
 
-For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility in at most 10 minutes:
-```{r}
+```{r figure-7, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 5 minutes walk (average threshold) compared between current system with equity stations and the original system without equity stations."}
+#accessibility plot for current system with equity stations at 5 minutes walk
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_sf(data = accessibility %>%
+            dplyr::filter(threshold == "5 min"),
+       aes(color = accessibility)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "PuRd", 
+                        direction = 1) +
+  facet_grid(system ~ threshold)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          size = 1.5, shape = 21, 
+          fill = "green")
+```
+
+### Ten Minutes
+
+At this threshold, we find that there are 3.61 bicycle racks per person in the original system.
+With the addition of equity stations, there are now 3.74 bicycles per person.
+Figure \ref{fig:figure-8} presents a comparison of accessibility between the systems.
+We begin to see differences in accessibility across the service area, with users near the university and its adjacent neighbourhoods, as well as neighbourhoods north of the downtown area have slightly higher accessibility.
+While the differences are modest, they are more apparent at this threshold than at shorter walking distances.
+
+```{r avg-conventional, include=FALSE}
 results_10min_conventional <- ttm_walk %>%
@@
 
-For the system without the equity stations we remove the equity stations from the travel time matrix before calculating accessibility in at most 15 minutes:
-```{r}
+```{r figure-8, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) compared between current system with equity stations and the original system without equity stations."}
+#accessibility plot for current system with equity stations at 10 minutes walk
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_sf(data = accessibility %>%
+            dplyr::filter(threshold == "10 min"),
+       aes(color = accessibility)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "PuRd", 
+                        direction = 1) +
+  facet_grid(system ~ threshold)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          size = 1.5, shape = 21, 
+          fill = "green")
+```
+
+### Fifteen Minutes
+
+At this threshold, we find that there are 2.44 bicycle racks per person in the original system.
+With the addition of equity stations, there are now 2.55 bicycles per person.
+Figure \ref{fig:figure-9} presents a comparison of accessibility between the systems.
+Users near the university and the neighbourhoods north of the downtown area (outlined in green) have the highest accessibility, followed by those who live in the city's downtown area.
+Accessibility in the east end of the core service area remains low.
+
+```{r max-conventional, include=FALSE}
 results_15min_conventional <- ttm_walk %>%
@@
 
-Extract results:
+```{r figure-9, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 15 minutes walk (extreme threshold) compared between current system with equity stations and the original system without equity stations."}
+#accessibility plot for current system with equity stations at 15 minutes walk
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_sf(data = accessibility %>%
+            dplyr::filter(threshold == "15 min"),
+       aes(color = accessibility)) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_color_distiller(palette = "PuRd", 
+                        direction = 1) +
+  facet_grid(system ~ threshold)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          size = 1.5, shape = 21, 
+          fill = "green")
+```
+
 ```{r, include=FALSE}
@@
 
-```{r, echo=FALSE}
-#accessibility plot for current system with equity stations at 3 minutes walk
-ggplot() +
-  geom_sf(data = population_50x50,
-          color = "lightgray") +
-  geom_sf(data = accessibility %>%
-            dplyr::filter(threshold == "3 min"),
-       aes(color = accessibility)) +
-  geom_sf(data = sobi_service,
-          fill = NA) +
-  scale_color_distiller(palette = "YlOrRd", 
-                        direction = 1) +
-  facet_grid(system ~ threshold)
-```
-
-```{r, echo=FALSE}
-#accessibility plot for current system with equity stations at 5 minutes walk
-ggplot() +
-  geom_sf(data = population_50x50,
-          color = "lightgray") +
-  geom_sf(data = accessibility %>%
-            dplyr::filter(threshold == "5 min"),
-       aes(color = accessibility)) +
-  geom_sf(data = sobi_service,
-          fill = NA) +
-  scale_color_distiller(palette = "YlOrRd", 
-                        direction = 1) +
-  facet_grid(system ~ threshold)
-```
-
-```{r, echo=FALSE}
-#accessibility plot for current system with equity stations at 10 minutes walk
-ggplot() +
-  geom_sf(data = population_50x50,
-          color = "lightgray") +
-  geom_sf(data = accessibility %>%
-            dplyr::filter(threshold == "10 min"),
-       aes(color = accessibility)) +
-  geom_sf(data = sobi_service,
-          fill = NA) +
-  scale_color_distiller(palette = "YlOrRd", 
-                        direction = 1) +
-  facet_grid(system ~ threshold)
-```
-
-```{r, echo=FALSE}
-#accessibility plot for current system with equity stations at 15 minutes walk
-ggplot() +
-  geom_sf(data = population_50x50,
-          color = "lightgray") +
-  geom_sf(data = accessibility %>%
-            dplyr::filter(threshold == "15 min"),
-       aes(color = accessibility)) +
-  geom_sf(data = sobi_service,
-          fill = NA) +
-  scale_color_distiller(palette = "YlOrRd", 
-                        direction = 1) +
-  facet_grid(system ~ threshold)
-```
-
-## Comparison
-
-Include table with total # of people who can reach a station within 3 minutes walk, 5 minutes, 10 minutes, and 15 minutes, to compare accessibility at different thresholds for both current and original system.
-
-Discussion
+## Accessibility by Median Total Income
+
+Approximately 138,000 people live within Hamilton Bike Share's core service area or within a 10 minute walk to the service area.
+We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in Hamilton Bike Share's core service area [see Table \ref{tab:table-2}].
+While previous research found that neighbourhoods with more disadvantage are better serviced by Hamilton Bike Share [@hosfordWhoArePublic2018], the authors used the Pampalon Deprivation Index from 2011 as a measure of socioeconomic status.
+Instead, we use median total income for each dissemination area to examine whether disparities in accessibility exist within the core service area.
+We divide income by quantiles: bottom 20%, second 20%, third 20%, fourth 20%, and top 20%.
+One of the unique properties of the balanced floating catchment area method is that data can be reaggregated from small to larger polygons while preserving the total population and number of bicycle racks at each station.
+This avoids demand and supply inflation, and also enables us to present findings in a way that is easier to interpret (i.e., at the dissemination area level).
+
+Panels 1 and 2 depict bivariate choropleth maps with the combined spatial distribution of accessibility and median total income per household for the different thresholds.
+Our analysis demonstrates that the stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing accessibility, albeit only modestly.
+By implementing equity stations in more areas with lower median total income in Hamilton, the PBSP has achieved greater horizontal equity by extending the spatial distribution of bicycles across the city.
+This is particularly evident at shorter walking distances of three and five minutes.
+
+However, we found that dissemination areas with lower total household income do not necessarily have greater access to the program [see Table \ref{tab:table-3}].
+Income disparities still persist, however only at certain thresholds.
+With and without equity stations, people in the top 20% of income have the highest level of accessibility at a threshold of ten and fifteen minutes.
+At distance thresholds of three and five minutes, people in the second 20% have the highest level of accessibility by a significant amount.
+This is due to the concentration of stations near or within the city's downtown area and the "lower city" which historically has been more disadvantaged compared to the surrounding suburban neighbourhoods.
+This is an artifact of the city's steel industry which led to more blue collar and working class people living in the city's downtown area and "lower city".
+However, the bottom 20%, who may benefit the most from Hamilton Bike Share's equity initiatives, have the lowest accessibility at three minutes threshold and the second lowest accessibility at all other thresholds.
+
+On the whole, we find that areas that have less than their equitable share of the level of service are some of the most socioeconomically disadvantaged areas in the city.
+This finding aligns with other studies that have found disparities in station or bike location along income lines in New York City [@babagoliExploringHealthSpatial2019], Tampa [@chenExploringEquityPerformance2019], and Seattle [@mooneyFreedomStationSpatial2019].
+
+```{r, include=FALSE}
+# drop geometry
+test <- accessibility
+
+# join accessibility to population cells
+test <- test %>% st_join(population_50x50 %>% dplyr::select(-UID)) %>%
+  rename(population50x50 = population)
+```
+
+```{r, include=FALSE}
+test <- test %>%
+  st_join(hamilton_da_2016)
+```
+
+```{r, include=FALSE}
+agg_accessibility_da <- test %>%
+  group_by(DA, 
+           system,
+           threshold) %>%
+  summarize(accessibility = sum(accessibility),
+            median_total_income = first(median_total_income), 
+            .groups = "drop") %>%
+  st_drop_geometry() %>%
+  left_join(hamilton_da_2016 %>% dplyr::select(DA, geometry),
+            by = c("DA" = "DA")) %>%
+  st_as_sf()
+```
+
+```{r figure-10, echo=FALSE, fig.align = 'center', out.width = "65%"}
+data_map <- agg_accessibility_da %>%
+            dplyr::filter(threshold == "3 min", 
+                          system == "All stations") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+data_map <- ggplot() +
+  geom_sf(data = data_map,
+          mapping = aes(fill = bi_class), 
+          color = "white", 
+          size = 0.1, 
+          show.legend = FALSE) +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "orange") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "purple") +
+  bi_theme() +
+  ggtitle("Accessibility by Income with Equity Stations \n at 3 Minutes Threshold") + 
+  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
+
+legend <- bi_legend(pal = "DkCyan",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-11, echo=FALSE, fig.align = 'center', out.width = "65%"}
+data_map <- agg_accessibility_da %>%
+            dplyr::filter(threshold == "3 min", 
+                          system == "Conventional") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+data_map <- ggplot() +
+  geom_sf(data = data_map,
+          mapping = aes(fill = bi_class), 
+          color = "white", 
+          size = 0.1, 
+          show.legend = FALSE) +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "purple") +
+  bi_theme() +
+  ggtitle("Accessibility by Income without Equity Stations \n at 3 Minutes Threshold") + 
+  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
+
+legend <- bi_legend(pal = "DkCyan",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-12, echo=FALSE, fig.align = 'center', out.width = "65%"}
+data_map <- agg_accessibility_da %>%
+            dplyr::filter(threshold == "5 min", 
+                          system == "All stations") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+data_map <- ggplot() +
+  geom_sf(data = data_map,
+          mapping = aes(fill = bi_class), 
+          color = "white", 
+          size = 0.1, 
+          show.legend = FALSE) +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "orange") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "purple") +
+  bi_theme() +
+  ggtitle("Accessibility by Income with Equity Stations \n at 5 Minutes Threshold") + 
+  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
+
+legend <- bi_legend(pal = "DkCyan",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-13, echo=FALSE, fig.align = 'center', out.width = "65%"}
+data_map <- agg_accessibility_da %>%
+            dplyr::filter(threshold == "5 min", 
+                          system == "Conventional") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+data_map <- ggplot() +
+  geom_sf(data = data_map,
+          mapping = aes(fill = bi_class), 
+          color = "white", 
+          size = 0.1, 
+          show.legend = FALSE) +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "purple") +
+  bi_theme() +
+  ggtitle("Accessibility by Income without Equity Stations \n at 3 Minutes Threshold") + 
+  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
+
+legend <- bi_legend(pal = "DkCyan",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-14, echo=FALSE, fig.align = 'center', out.width = "65%"}
+data_map <- agg_accessibility_da %>%
+            dplyr::filter(threshold == "10 min", 
+                          system == "All stations") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+data_map <- ggplot() +
+  geom_sf(data = data_map,
+          mapping = aes(fill = bi_class), 
+          color = "white", 
+          size = 0.1, 
+          show.legend = FALSE) +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "orange") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "purple") +
+  bi_theme() +
+  ggtitle("Accessibility by Income with Equity Stations \n at 10 Minutes Threshold") + 
+  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
+
+legend <- bi_legend(pal = "DkCyan",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-15, echo=FALSE, fig.align = 'center', out.width = "65%"}
+data_map <- agg_accessibility_da %>%
+            dplyr::filter(threshold == "10 min", 
+                          system == "Conventional") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+data_map <- ggplot() +
+  geom_sf(data = data_map,
+          mapping = aes(fill = bi_class), 
+          color = "white", 
+          size = 0.1, 
+          show.legend = FALSE) +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "purple") +
+  bi_theme() +
+  ggtitle("Accessibility by Income without Equity Stations \n at 10 Minutes Threshold") +
+  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
+
+legend <- bi_legend(pal = "DkCyan",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-16, echo=FALSE, fig.align = 'center', out.width = "65%"}
+data_map <- agg_accessibility_da %>%
+            dplyr::filter(threshold == "15 min", 
+                          system == "All stations") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+data_map <- ggplot() +
+  geom_sf(data = data_map,
+          mapping = aes(fill = bi_class), 
+          color = "white", 
+          size = 0.1, 
+          show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", 
+                dim = 3) +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "orange") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "purple") +
+  bi_theme() +
+  ggtitle("Accessibility by Income with Equity Stations \n at 15 Minutes Threshold") + 
+  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
+
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r figure-17, echo=FALSE, fig.align = 'center', out.width = "65%"}
+data_map <- agg_accessibility_da %>%
+            dplyr::filter(threshold == "15 min", 
+                          system == "Conventional") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+data_map <- ggplot() +
+  geom_sf(data = data_map,
+          mapping = aes(fill = bi_class), 
+          color = "white", 
+          size = 0.1, 
+          show.legend = FALSE) +
+  bi_scale_fill(pal = "DkBlue", 
+                dim = 3) +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional"), 
+          size = 1.5, 
+          shape = 21, 
+          fill = "purple") +
+  bi_theme() +
+  ggtitle("Accessibility by Income without Equity Stations \n at 15 Minutes Threshold") + 
+  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
+
+legend <- bi_legend(pal = "DkBlue",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6)
+
+cowplot::ggdraw() +
+  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
+  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+```
+
+```{r, include=FALSE}
+# Find the income quantiles of the population served by SoBi
+iq <- hamilton_da_2016 %>% 
+  dplyr::filter(sobi_service == "Yes") %>%
+  pull(median_total_income) %>%
+  quantile(c(0, 0.2, 0.4, 0.6, 0.8, 1), na.rm = TRUE)
+```
+
+```{r income-all-3, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_all_3 <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "3 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"))
+
+inc_stats_all_3 <- agg_accessibility_all_3 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(accessibility = sum(accessibility),
+            population = sum(population50x50))
+```
+
+```{r income-con-3, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_con_3 <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "3 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"))
+
+inc_stats_con_3 <- agg_accessibility_con_3 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(accessibility = sum(accessibility),
+            population = sum(population50x50))
+```
+
+```{r income-all-5, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_all_5 <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "5 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"))
+
+inc_stats_all_5 <- agg_accessibility_all_5 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(accessibility = sum(accessibility),
+            population = sum(population50x50))
+```
+
+```{r income-con-5, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_con_5 <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "5 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"))
+
+inc_stats_con_5 <- agg_accessibility_con_5 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(accessibility = sum(accessibility),
+            population = sum(population50x50))
+```
+
+```{r income-all-10, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_all_10 <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"))
+
+inc_stats_all_10 <- agg_accessibility_all_10 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(accessibility = sum(accessibility),
+            population = sum(population50x50))
+```
+
+```{r income-con-10, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_con_10 <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"))
+
+inc_stats_con_10 <- agg_accessibility_con_10 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(accessibility = sum(accessibility),
+            population = sum(population50x50))
+```
+
+```{r income-all-15, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_all_15 <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "15 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"))
+
+inc_stats_all_15 <- agg_accessibility_all_15 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(accessibility = sum(accessibility),
+            population = sum(population50x50))
+```
+
+```{r income-con-15, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_con_15 <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "15 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"))
+
+inc_stats_con_15 <- agg_accessibility_con_15 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(accessibility = sum(accessibility),
+            population = sum(population50x50))
+```
+
+```{r income-3, include=FALSE}
+income_3 <- rbind(inc_stats_all_3,  
+      inc_stats_con_3)
+
+income_3
+```
+
+```{r income-5, include=FALSE}
+income_5 <- rbind(inc_stats_all_5,  
+      inc_stats_con_5)
+
+income_5
+```
+
+```{r income-10, include=FALSE}
+income_10 <- rbind(inc_stats_all_10,  
+      inc_stats_con_10)
+
+income_10
+```
+
+```{r income-3, include=FALSE}
+income_15 <- rbind(inc_stats_all_15,  
+      inc_stats_con_15)
+
+income_15
+```
+
+```{r table-2, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Population that can reach at least one bike share station at different thresholds in the current system with equity stations and original system with only conventional stations."}
+data_tbl <- data.frame(
+  System = c("With Equity Stations", "Without Equity Stations"),
+  Minimum = c("48,725", "43,238"),
+  Average = c("84,655", "75,026"),
+  Maximum = c("111,680", "104,298"),
+  Extreme = c("118,581", "115,344")
+)
+
+kbl(data_tbl) %>%
+  kable_paper(full_width = F) %>%
+  column_spec(1, bold = T, border_right = T) %>%
+  kable_styling(bootstrap_options = "striped") %>%
+  add_header_above(c(" " = 1, "3 Minutes" = 1, "5 Minutes" = 1, "10 Minutes" = 1, "15 Minutes" = 1))
+```
+
+```{r accessibility-income, echo=FALSE}
+accessibility_income <- rbind(
+                        income_3,
+                        income_5,
+                        income_10,
+                        income_15) %>% drop_na()
+
+kbl(accessibility_income,
+    col.names = c("Income Quintile",
+                  "Accessibility",
+                  "Total Population")) %>%
+  kable_paper("striped") %>%
+  pack_rows("With Equity Stations - 3 minutes", 1, 5, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Without Equity Stations - 3 minutes", 6, 10, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("With Equity Stations - 5 minutes", 11, 15, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Without Equity Stations - 5 minutes", 16, 20, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("With Equity Stations - 10 minutes", 21, 25, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Without Equity Stations - 10 minutes", 26, 30, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("With Equity Stations - 15 minutes", 31, 35, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Without Equity Stations - 15 minutes", 36, 40, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  #column_spec(2, color = "white",
+      #background = spec_color(accessibility_income$accessibility[1:40], end = 1)) %>%
+  footnote(general = " ",
+           alphabet = c("With equity stations = Hamilton Bike Share current system (118 conventional stations, 12 equity stations)", "Without equity stations = Hamilton Bike Share original system (118 conventional stations, no equity stations)"))
+```
+
+```{r accessibility-difference-3, echo=FALSE}
+# calculate difference in accessibility with equity stations at 3 min
+inc_stats_all_3 <- inc_stats_all_3 %>%
+    left_join(inc_stats_con_3, by = "income_quintile")
+
+names(inc_stats_all_3) <- c("income_quintile", 
+                            "accessibility", 
+                            "population",
+                            "accessibility_o",
+                            "population_o")
+
+inc_stats_all_3$acc_difference <- (inc_stats_all_3$accessibility - inc_stats_all_3$accessibility_o)
+
+inc_stats_all_3$pop_difference <- (inc_stats_all_3$population - inc_stats_all_3$population_o)
+```
+
+```{r accessibility-difference-5, echo=FALSE}
+# calculate difference in accessibility with equity stations at 3 min
+inc_stats_all_5 <- inc_stats_all_5 %>%
+    left_join(inc_stats_con_5, by = "income_quintile")
+
+names(inc_stats_all_5) <- c("income_quintile", 
+                            "accessibility", 
+                            "population",
+                            "accessibility_o",
+                            "population_o")
+
+inc_stats_all_5$acc_difference <- (inc_stats_all_5$accessibility - inc_stats_all_5$accessibility_o)
+
+inc_stats_all_5$pop_difference <- (inc_stats_all_5$population - inc_stats_all_5$population_o)
+```
+
+```{r accessibility-difference-10, echo=FALSE}
+# calculate difference in accessibility with equity stations at 3 min
+inc_stats_all_10 <- inc_stats_all_10 %>%
+    left_join(inc_stats_con_10, by = "income_quintile")
+
+names(inc_stats_all_10) <- c("income_quintile", 
+                            "accessibility", 
+                            "population",
+                            "accessibility_o",
+                            "population_o")
+
+inc_stats_all_10$acc_difference <- (inc_stats_all_10$accessibility - inc_stats_all_10$accessibility_o)
+
+inc_stats_all_10$pop_difference <- (inc_stats_all_10$population - inc_stats_all_10$population_o)
+```
+
+```{r accessibility-difference-10, echo=FALSE}
+# calculate difference in accessibility with equity stations at 3 min
+inc_stats_all_10 <- inc_stats_all_10 %>%
+    left_join(inc_stats_con_10, by = "income_quintile")
+
+names(inc_stats_all_10) <- c("income_quintile", 
+                            "accessibility", 
+                            "population",
+                            "accessibility_o",
+                            "population_o")
+
+inc_stats_all_10$acc_difference <- (inc_stats_all_10$accessibility - inc_stats_all_10$accessibility_o)
+
+inc_stats_all_10$pop_difference <- (inc_stats_all_10$population - inc_stats_all_10$population_o)
+```
+
+```{r accessibility-difference-15, echo=FALSE}
+# calculate difference in accessibility with equity stations at 3 min
+inc_stats_all_15 <- inc_stats_all_15 %>%
+    left_join(inc_stats_con_10, by = "income_quintile")
+
+names(inc_stats_all_15) <- c("income_quintile", 
+                            "accessibility", 
+                            "population",
+                            "accessibility_o",
+                            "population_o")
+
+inc_stats_all_15$acc_difference <- (inc_stats_all_15$accessibility - inc_stats_all_15$accessibility_o)
+
+inc_stats_all_15$pop_difference <- (inc_stats_all_15$population - inc_stats_all_15$population_o)
+```
+
+```{r accessibility-differences, echo=FALSE}
+income_3 <- inc_stats_all_3[ -c(3,5,7) ]
+income_5 <- inc_stats_all_5[ -c(3,5,7) ]
+income_10 <- inc_stats_all_10[ -c(3,5,7) ]
+income_15 <- inc_stats_all_15[ -c(3,5,7) ]
+```
+
+```{r accessibility-income-difference, echo=FALSE}
+accessibility_difference <- rbind(
+                        income_3,
+                        income_5,
+                        income_10,
+                        income_15) %>% drop_na()
+
+kbl(accessibility_difference,
+    col.names = c("Income Quintile",
+                  "Accessibility with Equity Stations",
+                  "Accessibility without Equity Stations",
+                  "Difference")) %>%
+  kable_paper("striped") %>%
+  pack_rows("Threshold - 3 minutes", 1, 5, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Threshold - 5 minutes", 6, 10, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Threshold - 10 minutes", 11, 15, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Threshold - 15 minutes", 16, 20, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  column_spec(4, color = "white",
+      background = spec_color(accessibility_income$accessibility[1:40], end = 1)) %>%
+  footnote(general = " ",
+           alphabet = c("With equity stations = Hamilton Bike Share current system (118 conventional stations, 12 equity stations)", "Without equity stations = Hamilton Bike Share original system (118 conventional stations, no equity stations)"))
+```
+
+Study Limitations
 ===================
-
-Our analysis demonstrates that the stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing accessibility, albeit only modestly.
-
-By implementing equity stations in more areas with lower median household income in Hamilton, the PBSP has achieved greater horizontal equity by extending the spatial distribution of bicycles across the city.
-Under the the conditions of the original system, which did not have equity stations, the level of service was _X_.
-With the addition of twelve equity stations, the level of service increased to _X_.
-
-However, we found that dissemination areas with lower median household income do not necessarily have greater access to the program and that spatial disparities still persist.
-Under the the conditions of the original system, which did not have equity stations, _X%_ of dissemination areas (or _X%_ of the population) within the core service area with median household income less than 40,000 Canadian dollars could access at least one bicycle share station within 10 minutes walk.
-Only _X%_ of dissemination areas (or _X%_ of the population) with median household income less than 40,000 Canadian dollars could access at least one bicycle share station within 3 or 5 minutes walk.
-
-Similar to other studies, we find that areas with higher median household income have greater accessibility to Hamilton Bike Share.
-Our findings align with other studies that have found disparities in station or bike location along income lines in New York City [@babagoliExploringHealthSpatial2019], Tampa [@chenExploringEquityPerformance2019], and Seattle [@mooneyFreedomStationSpatial2019].
-
-# Study Limitations
 
@@
 The addition of specific equity stations to the public bicycle share program in Hamilton, Ontario had the net effect of increasing accessibility and reducing both horizontal and vertical inequities.
-In particular, accessibility improved the most for _X_ income groups at _X_ thresholds, but the gains were only modest as the threshold increased ( _2.10_ bicycles/person with equity stations vs _1.97_ bicycles/person without equity stations).
-Congestion effects were observed at higher thresholds, with accessibility decreasing significantly once the catchment area is increased to 10 minutes walking ( _X_ ).
-We identified specific areas that would benefit from an increased supply of bicycles which make them ideal candidates for new equity stations.
-In this way, our paper has made contributions in a positive way by applying an intuitive and useful approach to measure accessibility, and in a normative way by serving to inform future investments in cycle infrastructure.
+In particular, accessibility improved the most for the second 20% income groups at all thresholds, but the gains were only modest.
+Congestion effects were observed at higher thresholds, with accessibility decreasing significantly once the catchment area is increased to 10 minutes walking.
+However, We identified specific areas that have both low accessibility and low income which would benefit from an increased supply of bicycles.
+These are ideal candidates for new equity stations.
+In this way, our paper has made contributions in a positive way by applying an intuitive and useful approach to measure accessibility to a PBSP, and in a normative way by serving to inform future investments in cycle infrastructure for Hamilton Bike Share.
 

### 2021-03-19 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Revised in hub in data package and worked with figures and tables for style

@@
 ```{r load-packages, include=FALSE}
+library(biscale)
+library(cowplot)
 library(data.table)
@@
 library(gdistance)
+library(gridExtra)
 library(kableExtra)
+library(patchwork)
+library(readr)
 library(raster)
@@
 library(tidyverse)
+library(tinytex)
 library(units)
@@
 #library(pycno)
-library(readr)
 #library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
-library(tinytex)
-library(ggplot2)
-library(biscale)
-library(cowplot)
 ```
@@
 data("ttm_walk")
+```
+
+```{r crop-das, include=FALSE}
+# Crop DAs using bounding box of service area
+sobi_service_bbox <- sobi_service %>% 
+  st_bbox() %>%
+  st_as_sfc()
+
+cropped_da <- hamilton_da_2016 %>%
+  st_crop(sobi_service_bbox) %>%
+  dplyr::select(DA, geometry)
 ```
@@
   geom_sf(data = hamilton_da_2016,
-          fill = NA,
-          color = "gray") + 
+          color = NA,
+          fill = "gray") + 
   geom_sf(data = sobi_service,
           fill = NA,
-          colour = "black")
+          colour = "black") + 
+  theme(panel.background = element_rect(fill = "white"))
 ```
@@
 ggplot() + 
-  geom_sf(data = hamilton_da_2016 %>% 
-            dplyr::filter(sobi_service == "Yes"),
-          color = "gray") +
+  geom_sf(data = cropped_da,
+          color = NA,
+          fill = "gray") +
   geom_sf(data = sobi_service, 
@@
             dplyr::filter(NAME == "Hamilton"),
-          fill = "dark green",
+          fill = NA,
+          size = 1,
           color = "dark green") +
@@
           shape = 21,
-          fill = "green")
+          fill = "green") +
+  scale_size(name = "Number of racks", range = c(0.5, 4)) +
+  theme(panel.background = element_rect(fill = "white"))
 ```
@@
 
-kbl(data_tbl) %>%
+kable(data_tbl,
+      "latex",
+      booktabs = TRUE,
+      caption = "\\label{tab:landscape-features}Landscape features") %>%
   kable_paper(full_width = F) %>%
@@
   column_spec(2, width = "30em") %>%
-  kable_styling(bootstrap_options = "striped")
+  kable_styling(latex_options = c("striped", "scale_down"))
 ```
@@
           aes(fill = population),
-          color = NA) +
+          color = "white") +
   geom_sf(data = sobi_service,
@@
   scale_fill_distiller(palette = "YlOrRd", 
-                       direction = 1)
+                       direction = 1) +
+  theme(panel.background = element_rect(fill = "white"))
 ```
@@
 ggplot() + 
+  geom_sf(data = cropped_da,
+          fill = "gray",
+          color = NA) + 
   geom_sf(data = population_50x50,
@@
                         direction = 1) +
-  geom_sf(data = hamilton_da_2016 %>%
-            filter(sobi_service == "Yes"),
-          fill = NA,
-          color = "lightgray") + 
   geom_sf(data = sobi_service,
@@
           fill = NA,
-          color = "dark green")
-```
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"))
+```
+
+<!-- Suggest to remove this image: all ithe information it gives is already found in other figures: the location of the SoBi stations and the interpolated popualtion
 
@@
 ggplot() + 
+  geom_sf(data = cropped_da, 
+          color = NA, 
+          fill = "gray") + 
   geom_sf(data = population_50x50, 
           aes(color = population)) +
-  geom_sf(data = hamilton_da_2016 %>%
-            dplyr::filter(sobi_service == "Yes"), 
-          fill = NA, 
-          color = "lightgray") + 
   geom_sf(data = sobi_hubs %>% 
@@
           color = "dark green") +
-  facet_grid(rows = vars(hub_type))
-```
-
-## Accessibility by Distance
+  facet_grid(rows = vars(hub_type)) +
+  theme(panel.background = element_rect(fill = "white"))
+```
+--->
+
+## Accessibility by Distance Thresholds
 
@@
 
-```{r accessibility-sensibility, echo=FALSE}
+```{r accessibility-sensibility, eval=FALSE, include = FALSE}
 # Testing different thresholds to check how accessibility changes
@@
                     hub_status == "Active") %>%
-  b2sfca(threshold = i +2)
+    b2sfca(threshold = i +2 )
   accessibility_test$accessibility[i] <- sum(results_test$accessibility$accessibility)
@@
 
+<!-- Calculate accessibility for different thresholds -->
+
+```{r 3-min-conventional, include=FALSE}
+results_3min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 3)
+```
+
+```{r 3-min-all, include=FALSE}
+results_3min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 3)
+```
+
+```{r 3-min-sum-conventional, include=FALSE}
+sum(results_3min_conventional$los$los)
+sum(results_3min_conventional$accessibility$accessibility)
+```
+
+```{r 3-min-sum-all, include=FALSE}
+sum(results_3min_all$los$los)
+sum(results_3min_all$accessibility$accessibility)
+```
+```{r 5-min-conventional, include=FALSE}
+results_5min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 5)
+```
+
+```{r 5-min-all, include=FALSE}
+results_5min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 5)
+```
+
+```{r, 5-min-sum-conventional, include=FALSE}
+sum(results_5min_conventional$los$los)
+sum(results_5min_conventional$accessibility$accessibility)
+```
+
+```{r 5-min-sum-all, include=FALSE}
+sum(results_5min_all$los$los)
+sum(results_5min_all$accessibility$accessibility)
+```
+
+```{r 10-min-conventional, include=FALSE}
+results_10min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 10)
+```
+
+```{r 10-min-all, include=FALSE}
+results_10min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 10)
+```
+
+```{r 10-min-sum-conventional, include=FALSE}
+sum(results_10min_conventional$los$los)
+sum(results_10min_conventional$accessibility$accessibility)
+```
+
+```{r, 10-min-sum-all, include=FALSE}
+sum(results_10min_all$los$los)
+sum(results_10min_all$accessibility$accessibility)
+```
+
+```{r 15-min-conventional, include=FALSE}
+results_15min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 15)
+```
+
+```{r 15-min-all, include=FALSE}
+results_15min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 15)
+```
+
+```{r 15-min-sum-conventional, include=FALSE}
+sum(results_15min_conventional$los$los)
+sum(results_15min_conventional$accessibility$accessibility)
+```
+
+```{r 15-min-sum-all, include=FALSE}
+sum(results_15min_all$los$los)
+sum(results_15min_all$accessibility$accessibility)
+```
+
+```{r, bind-all-results, include=FALSE}
+los <- rbind(data.frame(results_3min_all$los, system = "All stations", threshold = "3 min"),
+             data.frame(results_3min_conventional$los, system = "Conventional", threshold = "3 min"),
+             data.frame(results_5min_all$los, system = "All stations", threshold = "5 min"),
+             data.frame(results_5min_conventional$los, system = "Conventional", threshold = "5 min"),
+             data.frame(results_10min_all$los, system = "All stations", threshold = "10 min"),
+             data.frame(results_10min_conventional$los, system = "Conventional", threshold = "10 min"),
+             data.frame(results_15min_all$los, system = "All stations", threshold = "15 min"),
+             data.frame(results_15min_conventional$los, system = "Conventional", threshold = "15 min"))
+
+accessibility <- rbind(data.frame(results_3min_all$accessibility, system = "All stations", threshold = "3 min"),
+                       data.frame(results_3min_conventional$accessibility, system = "Conventional", threshold = "3 min"),
+                       data.frame(results_5min_all$accessibility, system = "All stations", threshold = "5 min"),
+                       data.frame(results_5min_conventional$accessibility, system = "Conventional", threshold = "5 min"),
+                       data.frame(results_10min_all$accessibility, system = "All stations", threshold = "10 min"),
+                       data.frame(results_10min_conventional$accessibility, system = "Conventional", threshold = "10 min"),
+                       data.frame(results_15min_all$accessibility, system = "All stations", threshold = "15 min"),
+                       data.frame(results_15min_conventional$accessibility, system = "Conventional", threshold = "15 min"))
+```
+
+```{r, add-geometry-to-results, include=FALSE}
+los <- los %>%
+  left_join(sobi_hubs %>%
+              dplyr::select(OBJECTID),
+            by = c("hub" = "OBJECTID")) %>%
+  st_as_sf()
+
+accessibility <- accessibility %>%
+  left_join(population_50x50 %>%
+              dplyr::select(UID),
+            by = c("UID" = "UID")) %>%
+  st_as_sf()
+```
+
 ### Three Minutes
@@
 
-```{r min-conventional, include=FALSE}
-results_3min_conventional <- ttm_walk %>%
-  dplyr::filter(hub_type == "Conventional" &
-                hub_status == "Active") %>%
-  b2sfca(threshold = 3)
-```
-
-```{r min-all, include=FALSE}
-results_3min_all <- ttm_walk %>%
-  dplyr::filter(hub_status == "Active") %>%
-  b2sfca(threshold = 3)
-```
-
-```{r min-sum-conventional, echo=FALSE}
-sum(results_3min_conventional$los$los)
-sum(results_3min_conventional$accessibility$accessibility)
-```
-
-```{r min-sum-all, echo=FALSE}
-sum(results_3min_all$los$los)
-sum(results_3min_all$accessibility$accessibility)
-```
 ```{r figure-6, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
@@
   geom_sf(data = population_50x50,
-          color = "lightgray") +
+          color = "gray") +
   geom_sf(data = accessibility %>%
@@
           color = "dark green") +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "ERI"), 
-          size = 1.5, shape = 21, 
-          fill = "green")
+  # geom_sf(data =  sobi_hubs %>% 
+  #           dplyr::filter(hub_type == "ERI"), 
+  #         size = 1.5, shape = 21, 
+  #         fill = "green") +
+  theme(panel.background = element_rect(fill = "white"))
 ```
@@
 
-```{r med-conventional, include=FALSE}
-results_5min_conventional <- ttm_walk %>%
-  dplyr::filter(hub_type == "Conventional" &
-                hub_status == "Active") %>%
-  b2sfca(threshold = 5)
-```
-
-```{r med-all, include=FALSE}
-results_5min_all <- ttm_walk %>%
-  dplyr::filter(hub_status == "Active") %>%
-  b2sfca(threshold = 5)
-```
-
-```{r, med-sum-conventional, echo=FALSE}
-sum(results_5min_conventional$los$los)
-sum(results_5min_conventional$accessibility$accessibility)
-```
-
-```{r med-sum-all, echo=FALSE}
-sum(results_5min_all$los$los)
-sum(results_5min_all$accessibility$accessibility)
-```
-
 ```{r figure-7, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 5 minutes walk (average threshold) compared between current system with equity stations and the original system without equity stations."}
@@
 While the differences are modest, they are more apparent at this threshold than at shorter walking distances.
-
-```{r avg-conventional, include=FALSE}
-results_10min_conventional <- ttm_walk %>%
-  dplyr::filter(hub_type == "Conventional" &
-                hub_status == "Active") %>%
-  b2sfca(threshold = 10)
-```
-
-```{r avg-all, include=FALSE}
-results_10min_all <- ttm_walk %>%
-  dplyr::filter(hub_status == "Active") %>%
-  b2sfca(threshold = 10)
-```
-
-```{r avg-sum-conventional, echo=FALSE}
-sum(results_10min_conventional$los$los)
-sum(results_10min_conventional$accessibility$accessibility)
-```
-
-```{r, avg-sum-all, echo=FALSE}
-sum(results_10min_all$los$los)
-sum(results_10min_all$accessibility$accessibility)
-```
 
@@
 Accessibility in the east end of the core service area remains low.
-
-```{r max-conventional, include=FALSE}
-results_15min_conventional <- ttm_walk %>%
-  dplyr::filter(hub_type == "Conventional" &
-                hub_status == "Active") %>%
-  b2sfca(threshold = 15)
-```
-
-```{r max-all, include=FALSE}
-results_15min_all <- ttm_walk %>%
-  dplyr::filter(hub_status == "Active") %>%
-  b2sfca(threshold = 15)
-```
-
-```{r max-sum-conventional, echo=FALSE}
-sum(results_15min_conventional$los$los)
-sum(results_15min_conventional$accessibility$accessibility)
-```
-
-```{r max-sum-all, echo=FALSE}
-sum(results_15min_all$los$los)
-sum(results_15min_all$accessibility$accessibility)
-```
 
@@
           color = "dark green") +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "ERI"), 
-          size = 1.5, shape = 21, 
-          fill = "green")
-```
-
-```{r, include=FALSE}
-los <- rbind(data.frame(results_3min_all$los, system = "All stations", threshold = "3 min"),
-             data.frame(results_3min_conventional$los, system = "Conventional", threshold = "3 min"),
-             data.frame(results_5min_all$los, system = "All stations", threshold = "5 min"),
-             data.frame(results_5min_conventional$los, system = "Conventional", threshold = "5 min"),
-             data.frame(results_10min_all$los, system = "All stations", threshold = "10 min"),
-             data.frame(results_10min_conventional$los, system = "Conventional", threshold = "10 min"),
-             data.frame(results_15min_all$los, system = "All stations", threshold = "15 min"),
-             data.frame(results_15min_conventional$los, system = "Conventional", threshold = "15 min"))
-
-accessibility <- rbind(data.frame(results_3min_all$accessibility, system = "All stations", threshold = "3 min"),
-             data.frame(results_3min_conventional$accessibility, system = "Conventional", threshold = "3 min"),
-             data.frame(results_5min_all$accessibility, system = "All stations", threshold = "5 min"),
-             data.frame(results_5min_conventional$accessibility, system = "Conventional", threshold = "5 min"),
-             data.frame(results_10min_all$accessibility, system = "All stations", threshold = "10 min"),
-             data.frame(results_10min_conventional$accessibility, system = "Conventional", threshold = "10 min"),
-             data.frame(results_15min_all$accessibility, system = "All stations", threshold = "15 min"),
-             data.frame(results_15min_conventional$accessibility, system = "Conventional", threshold = "15 min"))
-```
-
-```{r, include=FALSE}
-los <- los %>%
-  left_join(sobi_hubs %>%
-              dplyr::select(OBJECTID),
-            by = c("hub" = "OBJECTID")) %>%
-  st_as_sf()
-
-accessibility <- accessibility %>%
-  left_join(population_50x50 %>%
-              dplyr::select(UID),
-            by = c("UID" = "UID")) %>%
-  st_as_sf()
+  # geom_sf(data =  sobi_hubs %>% 
+  #           dplyr::filter(hub_type == "ERI"), 
+  #         size = 1.5, shape = 21, 
+  #         fill = "green") +
+  theme(panel.background = element_rect(fill = "white"))
 ```
@@
             median_total_income = first(median_total_income), 
-            .groups = "drop") %>%
-  st_drop_geometry() %>%
-  left_join(hamilton_da_2016 %>% dplyr::select(DA, geometry),
-            by = c("DA" = "DA")) %>%
-  st_as_sf()
+            .groups = "drop")
 ```
@@
 
-data_map <- ggplot() +
-  geom_sf(data = data_map,
+bi_map_3_all <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
           mapping = aes(fill = bi_class), 
-          color = "white", 
-          size = 0.1, 
+          color = NA,
           show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
   bi_scale_fill(pal = "DkCyan", 
                 dim = 3) +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "ERI"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "orange") +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "Conventional"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "purple") +
-  bi_theme() +
-  ggtitle("Accessibility by Income with Equity Stations \n at 3 Minutes Threshold") + 
-  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
 
@@
                     ylab = "Higher income",
-                    size = 6)
-
-cowplot::ggdraw() +
-  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
-  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+                    size = 6) + 
+  theme(plot.background = element_rect(color = "lightgrey"))
 ```
@@
 
-data_map <- ggplot() +
-  geom_sf(data = data_map,
+bi_map_3_conv <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
           mapping = aes(fill = bi_class), 
-          color = "white", 
-          size = 0.1, 
+          color = NA,
           show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
   bi_scale_fill(pal = "DkCyan", 
                 dim = 3) +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "Conventional"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "purple") +
-  bi_theme() +
-  ggtitle("Accessibility by Income without Equity Stations \n at 3 Minutes Threshold") + 
-  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
-
-legend <- bi_legend(pal = "DkCyan",
-                    dim = 3,
-                    xlab = "Higher accessibility ",
-                    ylab = "Higher income",
-                    size = 6)
-
-cowplot::ggdraw() +
-  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
-  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+```
+
+```{r figure-bi-map-threshold-3, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): without equity stations (top panel) and with equity stations (bottom panel)"}
+ggdraw() + draw_plot(bi_map_3_conv, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_3_all, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
 ```
@@
 
-data_map <- ggplot() +
-  geom_sf(data = data_map,
+bi_map_5_all <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
           mapping = aes(fill = bi_class), 
-          color = "white", 
-          size = 0.1, 
+          color = NA,
           show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
   bi_scale_fill(pal = "DkCyan", 
                 dim = 3) +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "ERI"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "orange") +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "Conventional"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "purple") +
-  bi_theme() +
-  ggtitle("Accessibility by Income with Equity Stations \n at 5 Minutes Threshold") + 
-  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
-
-legend <- bi_legend(pal = "DkCyan",
-                    dim = 3,
-                    xlab = "Higher accessibility ",
-                    ylab = "Higher income",
-                    size = 6)
-
-cowplot::ggdraw() +
-  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
-  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
 ```
@@
 
-data_map <- ggplot() +
-  geom_sf(data = data_map,
+bi_map_5_conv <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
           mapping = aes(fill = bi_class), 
-          color = "white", 
-          size = 0.1, 
+          color = NA,
           show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
   bi_scale_fill(pal = "DkCyan", 
                 dim = 3) +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "Conventional"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "purple") +
-  bi_theme() +
-  ggtitle("Accessibility by Income without Equity Stations \n at 3 Minutes Threshold") + 
-  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
-
-legend <- bi_legend(pal = "DkCyan",
-                    dim = 3,
-                    xlab = "Higher accessibility ",
-                    ylab = "Higher income",
-                    size = 6)
-
-cowplot::ggdraw() +
-  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
-  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+```
+
+```{r figure-bi-map-threshold-5, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): without equity stations (top panel) and with equity stations (bottom panel)"}
+ggdraw() + draw_plot(bi_map_5_conv, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_5_all, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
 ```
@@
 
-data_map <- ggplot() +
-  geom_sf(data = data_map,
+bi_map_10_all <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
           mapping = aes(fill = bi_class), 
-          color = "white", 
-          size = 0.1, 
+          color = NA,
           show.legend = FALSE) +
-  bi_scale_fill(pal = "DkCyan", 
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +  bi_scale_fill(pal = "DkCyan", 
                 dim = 3) +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "ERI"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "orange") +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "Conventional"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "purple") +
-  bi_theme() +
-  ggtitle("Accessibility by Income with Equity Stations \n at 10 Minutes Threshold") + 
-  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
-
-legend <- bi_legend(pal = "DkCyan",
-                    dim = 3,
-                    xlab = "Higher accessibility ",
-                    ylab = "Higher income",
-                    size = 6)
-
-cowplot::ggdraw() +
-  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
-  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
 ```
@@
 
-data_map <- ggplot() +
-  geom_sf(data = data_map,
+bi_map_10_conv <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
           mapping = aes(fill = bi_class), 
-          color = "white", 
-          size = 0.1, 
+          color = NA,
           show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
   bi_scale_fill(pal = "DkCyan", 
                 dim = 3) +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "Conventional"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "purple") +
-  bi_theme() +
-  ggtitle("Accessibility by Income without Equity Stations \n at 10 Minutes Threshold") +
-  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
-
-legend <- bi_legend(pal = "DkCyan",
-                    dim = 3,
-                    xlab = "Higher accessibility ",
-                    ylab = "Higher income",
-                    size = 6)
-
-cowplot::ggdraw() +
-  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
-  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+```
+
+```{r figure-bi-map-threshold-10, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): without equity stations (top panel) and with equity stations (bottom panel)"}
+ggdraw() + draw_plot(bi_map_10_conv, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_10_all, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
 ```
@@
                 system == "All stations") %>%
+  drop_na() %>% 
   bi_class(x = accessibility, 
@@
 
-data_map <- ggplot() +
-  geom_sf(data = data_map,
+bi_map_15_all <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
           mapping = aes(fill = bi_class), 
-          color = "white", 
-          size = 0.1, 
+          color = NA,
           show.legend = FALSE) +
-  bi_scale_fill(pal = "DkBlue", 
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  bi_scale_fill(pal = "DkCyan", 
                 dim = 3) +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "ERI"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "orange") +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "Conventional"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "purple") +
-  bi_theme() +
-  ggtitle("Accessibility by Income with Equity Stations \n at 15 Minutes Threshold") + 
-  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
-
-legend <- bi_legend(pal = "DkBlue",
-                    dim = 3,
-                    xlab = "Higher accessibility ",
-                    ylab = "Higher income",
-                    size = 6)
-
-cowplot::ggdraw() +
-  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
-  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
 ```
@@
                 system == "Conventional") %>%
+  drop_na() %>% 
   bi_class(x = accessibility, 
@@
 
-data_map <- ggplot() +
-  geom_sf(data = data_map,
+bi_map_15_conv <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
           mapping = aes(fill = bi_class), 
-          color = "white", 
-          size = 0.1, 
+          color = NA,
           show.legend = FALSE) +
-  bi_scale_fill(pal = "DkBlue", 
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  bi_scale_fill(pal = "DkCyan", 
                 dim = 3) +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "Conventional"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "purple") +
-  bi_theme() +
-  ggtitle("Accessibility by Income without Equity Stations \n at 15 Minutes Threshold") + 
-  theme(plot.title = element_text(face = "bold", colour = "black", size = 12))
-
-legend <- bi_legend(pal = "DkBlue",
-                    dim = 3,
-                    xlab = "Higher accessibility ",
-                    ylab = "Higher income",
-                    size = 6)
-
-cowplot::ggdraw() +
-  cowplot::draw_plot(data_map, 0, 0, 1, 1) +
-  cowplot::draw_plot(legend, 0.05, .6, 0.2, 0.2)
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+```
+
+```{r figure-bi-map-threshold-15, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): without equity stations (top panel) and equity stations (bottom panel)"}
+ggdraw() + draw_plot(bi_map_15_conv, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_15_all, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
 ```
@@
                                      median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
-                                     median_total_income > iq[5] ~ "Top 20%"))
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
 
@@
 
-```{r income-3, include=FALSE}
-income_3 <- rbind(inc_stats_all_3,  
-      inc_stats_con_3)
-
-income_3
-```
-
-```{r income-5, include=FALSE}
-income_5 <- rbind(inc_stats_all_5,  
-      inc_stats_con_5)
-
-income_5
-```
-
-```{r income-10, include=FALSE}
-income_10 <- rbind(inc_stats_all_10,  
-      inc_stats_con_10)
-
-income_10
-```
-
-```{r income-3, include=FALSE}
-income_15 <- rbind(inc_stats_all_15,  
-      inc_stats_con_15)
-
-income_15
-```
-
-```{r table-2, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Population that can reach at least one bike share station at different thresholds in the current system with equity stations and original system with only conventional stations."}
-data_tbl <- data.frame(
-  System = c("With Equity Stations", "Without Equity Stations"),
-  Minimum = c("48,725", "43,238"),
-  Average = c("84,655", "75,026"),
-  Maximum = c("111,680", "104,298"),
-  Extreme = c("118,581", "115,344")
-)
-
-kbl(data_tbl) %>%
-  kable_paper(full_width = F) %>%
-  column_spec(1, bold = T, border_right = T) %>%
-  kable_styling(bootstrap_options = "striped") %>%
-  add_header_above(c(" " = 1, "3 Minutes" = 1, "5 Minutes" = 1, "10 Minutes" = 1, "15 Minutes" = 1))
+```{r income-all-hubs, include=FALSE}
+all_hubs <- rbind(inc_stats_all_3,
+                  inc_stats_all_5,
+                  inc_stats_all_10,
+                  inc_stats_all_15) %>%
+  drop_na() # there is a very small number of population cells without income data
+
+all_hubs
+```
+
+```{r income-conventional-hubs, include=FALSE}
+conventional_hubs <- rbind(inc_stats_con_3,
+                           inc_stats_con_5,
+                           inc_stats_con_10,
+                           inc_stats_con_15) %>%
+  drop_na() # there is a very small number of population cells without income data
+
+conventional_hubs
 ```
@@
 ```{r accessibility-income, echo=FALSE}
-accessibility_income <- rbind(
-                        income_3,
-                        income_5,
-                        income_10,
-                        income_15) %>% drop_na()
-
-kbl(accessibility_income,
+accessibility_income <- cbind(
+  conventional_hubs,
+  all_hubs[, 2:3])
+
+colnames(accessibility_income) <- c("iq", "accessibility_conv", "population_conv", "accessibility_all", "population_all")
+
+accessibility_income <- accessibility_income %>%
+  mutate(accessibility_conv = round(accessibility_conv, 3), 
+         population_conv = round(population_conv),
+         accessibility_all = round(accessibility_all, 3), 
+         population_all = round(population_all),
+         accessibility_diff = accessibility_all - accessibility_conv,
+         population_diff = population_all - population_conv)
+
+kable(accessibility_income,
+      "latex",
+      booktabs = TRUE,
       col.names = c("Income Quintile",
                     "Accessibility",
-                  "Total Population")) %>%
-  kable_paper("striped") %>%
-  pack_rows("With Equity Stations - 3 minutes", 1, 5, 
-            label_row_css = "background-color: #8f9498; color: #fff;") %>%
-  pack_rows("Without Equity Stations - 3 minutes", 6, 10, 
-            label_row_css = "background-color: #8f9498; color: #fff;") %>%
-  pack_rows("With Equity Stations - 5 minutes", 11, 15, 
-            label_row_css = "background-color: #8f9498; color: #fff;") %>%
-  pack_rows("Without Equity Stations - 5 minutes", 16, 20, 
-            label_row_css = "background-color: #8f9498; color: #fff;") %>%
-  pack_rows("With Equity Stations - 10 minutes", 21, 25, 
-            label_row_css = "background-color: #8f9498; color: #fff;") %>%
-  pack_rows("Without Equity Stations - 10 minutes", 26, 30, 
-            label_row_css = "background-color: #8f9498; color: #fff;") %>%
-  pack_rows("With Equity Stations - 15 minutes", 31, 35, 
-            label_row_css = "background-color: #8f9498; color: #fff;") %>%
-  pack_rows("Without Equity Stations - 15 minutes", 36, 40, 
-            label_row_css = "background-color: #8f9498; color: #fff;") %>%
-  #column_spec(2, color = "white",
-      #background = spec_color(accessibility_income$accessibility[1:40], end = 1)) %>%
-  footnote(general = " ",
-           alphabet = c("With equity stations = Hamilton Bike Share current system (118 conventional stations, 12 equity stations)", "Without equity stations = Hamilton Bike Share original system (118 conventional stations, no equity stations)"))
-```
-
-```{r accessibility-difference-3, echo=FALSE}
-# calculate difference in accessibility with equity stations at 3 min
-inc_stats_all_3 <- inc_stats_all_3 %>%
-    left_join(inc_stats_con_3, by = "income_quintile")
-
-names(inc_stats_all_3) <- c("income_quintile", 
-                            "accessibility", 
-                            "population",
-                            "accessibility_o",
-                            "population_o")
-
-inc_stats_all_3$acc_difference <- (inc_stats_all_3$accessibility - inc_stats_all_3$accessibility_o)
-
-inc_stats_all_3$pop_difference <- (inc_stats_all_3$population - inc_stats_all_3$population_o)
-```
-
-```{r accessibility-difference-5, echo=FALSE}
-# calculate difference in accessibility with equity stations at 3 min
-inc_stats_all_5 <- inc_stats_all_5 %>%
-    left_join(inc_stats_con_5, by = "income_quintile")
-
-names(inc_stats_all_5) <- c("income_quintile", 
-                            "accessibility", 
-                            "population",
-                            "accessibility_o",
-                            "population_o")
-
-inc_stats_all_5$acc_difference <- (inc_stats_all_5$accessibility - inc_stats_all_5$accessibility_o)
-
-inc_stats_all_5$pop_difference <- (inc_stats_all_5$population - inc_stats_all_5$population_o)
-```
-
-```{r accessibility-difference-10, echo=FALSE}
-# calculate difference in accessibility with equity stations at 3 min
-inc_stats_all_10 <- inc_stats_all_10 %>%
-    left_join(inc_stats_con_10, by = "income_quintile")
-
-names(inc_stats_all_10) <- c("income_quintile", 
-                            "accessibility", 
-                            "population",
-                            "accessibility_o",
-                            "population_o")
-
-inc_stats_all_10$acc_difference <- (inc_stats_all_10$accessibility - inc_stats_all_10$accessibility_o)
-
-inc_stats_all_10$pop_difference <- (inc_stats_all_10$population - inc_stats_all_10$population_o)
-```
-
-```{r accessibility-difference-10, echo=FALSE}
-# calculate difference in accessibility with equity stations at 3 min
-inc_stats_all_10 <- inc_stats_all_10 %>%
-    left_join(inc_stats_con_10, by = "income_quintile")
-
-names(inc_stats_all_10) <- c("income_quintile", 
-                            "accessibility", 
-                            "population",
-                            "accessibility_o",
-                            "population_o")
-
-inc_stats_all_10$acc_difference <- (inc_stats_all_10$accessibility - inc_stats_all_10$accessibility_o)
-
-inc_stats_all_10$pop_difference <- (inc_stats_all_10$population - inc_stats_all_10$population_o)
-```
-
-```{r accessibility-difference-15, echo=FALSE}
-# calculate difference in accessibility with equity stations at 3 min
-inc_stats_all_15 <- inc_stats_all_15 %>%
-    left_join(inc_stats_con_10, by = "income_quintile")
-
-names(inc_stats_all_15) <- c("income_quintile", 
-                            "accessibility", 
-                            "population",
-                            "accessibility_o",
-                            "population_o")
-
-inc_stats_all_15$acc_difference <- (inc_stats_all_15$accessibility - inc_stats_all_15$accessibility_o)
-
-inc_stats_all_15$pop_difference <- (inc_stats_all_15$population - inc_stats_all_15$population_o)
-```
-
-```{r accessibility-differences, echo=FALSE}
-income_3 <- inc_stats_all_3[ -c(3,5,7) ]
-income_5 <- inc_stats_all_5[ -c(3,5,7) ]
-income_10 <- inc_stats_all_10[ -c(3,5,7) ]
-income_15 <- inc_stats_all_15[ -c(3,5,7) ]
-```
-
-```{r accessibility-income-difference, echo=FALSE}
-accessibility_difference <- rbind(
-                        income_3,
-                        income_5,
-                        income_10,
-                        income_15) %>% drop_na()
-
-kbl(accessibility_difference,
-    col.names = c("Income Quintile",
-                  "Accessibility with Equity Stations",
-                  "Accessibility without Equity Stations",
-                  "Difference")) %>%
-  kable_paper("striped") %>%
+                    "Population",
+                    "Accessibility",
+                    "Population",
+                    "Accessibility",
+                    "Population"),
+      align = c("l", "c", "c", "c", "c", "c", "c"),
+      caption = "\\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile") %>%
+  kable_styling(latex_options = c("striped", "scale_down")) %>%
   pack_rows("Threshold - 3 minutes", 1, 5, 
@@
             label_row_css = "background-color: #8f9498; color: #fff;") %>%
-  column_spec(4, color = "white",
-      background = spec_color(accessibility_income$accessibility[1:40], end = 1)) %>%
+  add_header_above(c(" " = 1, "Without Equity Stations" = 2, "With Equity Stations" = 2, "Difference" = 2)) %>%
+  #column_spec(2, color = "white",
+  #background = spec_color(accessibility_income$accessibility[1:40], end = 1)) %>%
   footnote(general = " ",

### 2021-03-19 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Revised in hub in data package and worked with figures and tables for style

\documentclass[]{elsarticle} %review=doublespace preprint=single 5p=2 column
%%% Begin My package additions %%%%%%%%%%%%%%%%%%%
\usepackage[hyphens]{url}

  \journal{Transportation Research Part D: Transport and Environment} % Sets Journal name

\usepackage{lineno} % add
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{graphicx}
\usepackage{booktabs} % book-quality tables
%%%%%%%%%%%%%%%% end my additions to header

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

% Pandoc citation processing

% Pandoc header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\begin{document}
\begin{frontmatter}

  \title{Examining spatial equity and accessibility to a public bicycle share
program using a balanced floating catchment area approach}
    \author[Some School]{Elise Desjardins\corref{1}}
   \ead{desjae@mcmaster.ca} 
    \author[Another University]{Christopher D. Higgins\corref{2}}
   \ead{cd.higgins@utoronto.ca} 
    \author[Some School]{Antonio Paez\corref{2}}
   \ead{paezha@mcmaster.ca} 
      \address[McMaster University]{School of Earth, Environment \& Society, 1280 Main Street West,
Hamilton, ON L8S4L8}
    \address[University of Toronto Scarborough]{Department of Geography \& Planning, 1265 Military Trail, Toronto, ON
M1C1A4}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

  \begin{abstract}
  Public bicycle share programs are implemented to promote cycling as a
  convenient and sustainable mode of transportation. These systems can
  also play a role in addressing transportation needs and advancing
  transportation equity if they make cycling more accessible to lower
  income or disadvantaged populations. Recent studies assessing equity in
  bike sharing have found differences in ridership or membership based on
  income, education, and trip purpose. In Ontario, the City of Hamilton
  launched a public bicycle share program in 2015 that currently has over
  900 operational bicycles and 130 docking stations. In 2018, the
  non-profit organization that operates the program launched an equity
  initiative to provide subsidized memberships and to expand service by
  adding twelve docking stations. Previous research found that Hamilton's
  public bicycle share program targeted well disadvantaged areas in the
  city compared to programs in other Canadian cities. Since the time or
  distance that members of the population need to reach a bicycle share
  station decreases the potential of accessing the program, the location
  of stations matters. Unlike other public amenities with greater ability
  to adapt to crowding, the utility of stations is limited by the number
  of bicycles that they can hold, which makes crowding effects critical;
  the system may not necessarily be improved if stations are easy to reach
  but offer only a small number of bicycles. In this research, we revisit
  the case of Hamilton and investigate equity differentials in
  accessibility to bicycle share stations. We compare accessibility in the
  program with and without the equity stations to assess the effect of the
  initiative. Previous research on cycling equity or accessibility has
  used a macro-level approach with entire neighbourhoods or census tracts
  as the unit of analysis, although there are recent exceptions. In
  contrast, we implement our analysis parting from microscopic zones to
  better reflect travel to a docking station, which often happens at the
  sub-neighborhood level, and conduct a sensitivity analysis at several
  walking thresholds. We then reaggregate the estimated accessibility to
  dissemination areas for further analysis using census data. Equity
  analysis of this type is possible thanks to a newly developed approach
  for accessibility using balanced floating catchment areas. The addition
  of equity stations achieved its goal to increase accessibility for lower
  income populations, although the increase was relatively modest (2.10
  bicycles/person with equity stations vs 1.97 bicycles/person without
  equity stations). Further research is needed to determine whether this
  encouraged more cycling.
  \end{abstract}

 \end{frontmatter}

\hypertarget{background}{%
\section{Background}\label{background}}

Public bicycle share programs (PBSP) have been implemented in over 800
cities worldwide and a great deal has been learned about their typical
users ({\textbf{???}}). In many cities, males use bike share more than
females ({\textbf{???}}; {\textbf{???}}; {\textbf{???}}; {\textbf{???}})
as do younger age cohorts ({\textbf{???}}; {\textbf{???}};
{\textbf{???}}). However, one study found that bike share users in
Washington, DC were more likely to be female ({\textbf{???}}), which
suggests that the gender gap among cyclists who use bike share is less
disparate than the gap for personal bicycle use ({\textbf{???}}). There
is some evidence that bike share users are less likely to own a car
({\textbf{???}}; {\textbf{???}}). However, the relationship between
income or education and bike share use is less clear. Stations in
disadvantaged communities in Chicago generated most of the average
annual trips ({\textbf{???}}) and individuals from minority or lower
socioeconomic status neighbourhoods in Minneapolis-St.~Paul used the
city's PBSP more ({\textbf{???}}). Being university educated was also a
significant correlate of bike share use in Montreal, Canada
({\textbf{???}}). On the other hand, members who reside in
minority-concentrated and lower socioeconomic status neighbourhoods use
the Minneapolis-St Paul bike share more frequently ({\textbf{???}}).
Financial savings have been found to motivate those on a low income to
use bike share ({\textbf{???}}). Many studies have found that proximity
to a bike share station, either living or working near one, is an
important determinant of use or having a membership ({\textbf{???}};
{\textbf{???}}). This makes sense given that individuals are more likely
to use services or programs that they can easily access. Finally, the
potential of bike share programs to increase cycling levels has also
been explored recently ({\textbf{???}}; {\textbf{???}}), but more
longitudinal evidence is needed to determine their impact on encouraging
more cycling and offering more opportunities for physical activity. On
the whole, these findings highlight the need for bike share systems to
be more highly accessible for diverse populations in order to increase
use beyond the ``typical'' users, which has been the focus of recent
research ({\textbf{???}}; {\textbf{???}}; {\textbf{???}}), and to offer
more people the option of using sustainable and active transportation.

Several studies have recently explored the equity of PBSP in North
American cities by primarily examining who has access to bike share
(e.g., differences by demographics or socioeconomic status) and where
stations are located. Equity can be achieved in two different ways:
equal spatial distribution across a region (e.g., horizontal equity) or
greater access for vulnerable or disadvantaged populations (e.g.,
vertical equity) ({\textbf{???}}). Both are of interest to researchers
and transport planners since they are often linked in that advantage, or
conversely disadvantage, has spatial patterns. Using a negative binomial
regression model, Qian and Jaller ({\textbf{???}}) estimated ridership
in Chicago's PBSP among disadvantaged communities and found some
disparities. While annual members in disadvantaged communities have a
significantly lower share of trips compared to other areas in the city,
they make longer trips. This suggests that they may be using PBSP for
work commuting, which points to the importance of ensuring equitable
access ({\textbf{???}}). Similar results were found in Philadelphia,
where lower income areas generate fewer trips and efforts to increase
equity within the program have not been as successful as intended
({\textbf{???}}). In the case of Seattle, all neighbourhoods had some
level of access to dockless bikes but those with higher incomes and more
residents of higher education had more bikes ({\textbf{???}}). Babagoli
et al.~({\textbf{???}}) found that neighbourhoods in New York City with
higher affluence had the greater proportion of Citi Bike stations.
Overall, these findings suggest that horizontal equity can be achieved
while vertical equity is harder to attain. Exploring transport equity by
investigating where bicycle share stations are located, often using
neighbourhoods or census tracts as the spatial unit of analysis, can
ignore or miss the benefits that may be derived from adjacent zones.
Meaning that, stations may be lacking in certain neighbourhoods but
there may be stations accessible within a reasonable walking cutoff
time. This is where accessibility becomes an important consideration.

Accessibility has been applied in both a positive and normative way to
inform transportation planning ({\textbf{???}}), but its utility to this
field has evolved since its conceptualization and become linked with
recent interest in prioritizing local proximity and modes that are
suitable for local trips like walking and cycling ({\textbf{???}}). As a
measure of the ease of reaching potential destinations spread spatially
in a given area, accessibility is relevant to PBSPs because it can
identify current inequities in usage, as well as guide interventions
that increase access for groups that are under-serviced or address gaps
in transportation options. It also addresses some of the challenges of
other performance measures such as level of service within a
transportation network by measuring person-based indicators and
exploring differences in use between population groups ({\textbf{???}}).

In the context of PBSPs, accessibility refers to the distance an
individual must travel to reach a bicycle share station ({\textbf{???}};
{\textbf{???}}). Since the time or distance that members of the
population need to reach a bicycle share station decreases the potential
of accessing the program and ultimately use of the program, the location
and size (e.g., maximum number of bicycles available) of stations
matters. Indeed, distance to bicycle share stations is associated with
use ({\textbf{???}}; {\textbf{???}}) and can be a barrier to using PBSPs
({\textbf{???}}). Kabra et al.~({\textbf{???}}) found that the majority
of bike share usage in Paris comes from areas within 300m of stations.
Furthermore, unlike other public amenities with greater ability to adapt
to crowding, like health care services, the utility of stations is
limited by the number of bicycles that they can hold, which makes
crowding effects critical. The program may not necessarily be improved
if stations are easy to reach but offer only a small number of bicycles.
Likewise, more people may not opt to use the program if the supply of
bicycles available at the nearest station is insufficient to meet
demand. The location and size of stations is important to increase the
utility of this public transportation option for more people, thus
achieving vertical equity. Accessibility analyses for PBSPs constitute a
positive and evaluation-based approach that also has the potential to
inform equity efforts. For instance, Wang and Lindsey ({\textbf{???}})
investigated whether new or relocated bicycle share stations increased
accessibility and use, which offered important insights to improve the
performance of the program.

Several approaches have been commonly used for measuring accessibility:
cumulative opportunities, gravity, and utility-based ({\textbf{???}}).
Paez et al ({\textbf{???}}) provide a recent overview of various
formulations and applications of accessibility in transportation
research. Briefly, the gravity-based approach involves weighting
opportunities, for example the quantity of bicycle share stations, as
measured by a function of time, using a negative exponential form
({\textbf{???}}). It takes both demand and congestion into account
({\textbf{???}}). Floating catchment area (FCA) methods are a type of
gravity measure that have been used in healthcare accessibility research
but that are applicable to transportation case studies. This approach
involves producing flexible catchment areas for populations, recognizing
that people may be willing to travel to access particular services
({\textbf{???}}). Thus, this model does a good job of considering
potential crowding of services using a binary distance function
({\textbf{???}}). A recent improvement to this approach was achieved
through a simple and intuitive adjustment to the allocation of supply
and demand, which addressed the effects of demand and service inflation
from the conventional approach (see {\textbf{???}}).

Researchers have also taken different approaches when it comes to
aggregation of data, either by using the individual or household as the
smallest unit of analysis or larger spatial zones. Previous research on
bike share equity has typically used a macro-level approach with
aggregate data from entire neighbourhoods or census tracts
({\textbf{???}}; {\textbf{???}}; {\textbf{???}}; {\textbf{???}}),
although there are recent exceptions ({\textbf{???}}; {\textbf{???}}).
This is also true for studies examining correlates of bike share demand
({\textbf{???}}). Handy and Niemeier ({\textbf{???}}) note that using
disaggregated data in accessibility analyses provides a more accurate
estimate for individuals, which is useful for addressing vertical
inequities in PBSP usage.

Using balanced floating catchment area methods, a novel approach that
has not been used yet in cycling research, we examine accessibility to a
public bicycle share program in Hamilton, Ontario. Using disaggregate
population-level data, we (1) conduct a sensitivity analysis by
measuring accessibility and level of service to bike share stations at
different thresholds of walking to reach a station: 3 minutes, 5
minutes, 10 minutes, and 15 minutes; (2) explore the contribution of
specific stations that were added to Hamilton's PBSP to reducing both
horizontal and vertical inequities; and (3) examine whether disparities
in accessibility exist according to median household income of
dissemination areas within the core service area.

\hypertarget{sec:study}{%
\section{Case Study}\label{sec:study}}

This paper uses the city of Hamilton, located in Ontario, Canada, as a
case study. The city launched a public bicycle share program, Hamilton
Bike Share, in March 2015 with 115 stations and 750 bicycles
({\textbf{???}}). Before June 2020, the program was known as SoBi
Hamilton. Stations are spaced between 300 and 600 metres apart
({\textbf{???}}). The core service area spans 35 km\^{}2 of the city and
roughly 138,000 can reach a bike share hub within 30 minutes of walking
{[}see Figure 1{]}. This represents roughly one fifth of the population
in Hamilton, according to the 2016 Canadian Census. The program was
enthusiastically welcomed in the city - within three weeks of launching,
10,000 trips had been made ({\textbf{???}}). In 2017, Hamilton Bike
Share Inc., the non-profit organization that operates the program,
initiated an equity program, Everyone Rides Initiative (ERI), to remove
barriers that may prevent individuals from accessing bike share in
Hamilton. An additional 75 bicycles and 15 stations were added to the
program, which expanded it to more disadvantaged areas in the core
service area {[}see Figure 2{]}. The program also offers subsidized
memberships to individuals who identify as low income. A comparable
program can be found in Philadelphia (see {\textbf{???}}). As of June
2020, the bike share program has 900 bikes and 130 stations, and over
26,000 active memberships ({\textbf{???}}).

Hamilton Bike Share has conducted one membership survey to date in 2018
({\textbf{???}}), and the findings from 420 members are broadly in line
with the trends that were discussed above (see {\textbf{???}} for a
recent review of the literature). The majority of respondents live
within the core service area and the gender split is typical: 57\% of
respondents are male and 41\% are female. The majority of respondents,
both male and female, are between 25 and 34 years of age, but the
percentage of male respondents is higher in the subsequent age groups.
Respondents use bike share for commuting (40\% of trips) or errands and
meetings (24\% of trips), and nearly 50\% of trips have an average
length of 11 to 20 minutes. As a result of having a bike share
membership, 49\% of respondents report that they use their private
vehicle less often or much less often but 48\% report that their private
vehicle use has remained about the same. This suggests that SoBi
Hamilton has been useful for certain kinds of trips but not all, meaning
that some trips continue to require a private vehicle.

Further research has provided insights about the behaviour and route
choice of bike share users in Hamilton. The routes most frequently
travelled are longer than the shortest distance route from origin to
destination ({\textbf{???}}; {\textbf{???}}). Ridership is influenced by
weather conditions, temporal factors such as university terms,
employment, and proximity to important destinations like the local
university, but is not influenced by population and most transportation
infrastructure variables ({\textbf{???}}). Bike share users prefer to
deter from the shortest distance route to travel on streets with cycling
facilities or lower volumes of traffic ({\textbf{???}}) and avoid steep
slopes and busy roads ({\textbf{???}}). It is important to note that
these studies used daily ridership or GPS data from trips taken before
the equity stations were implemented, so these findings are informative
with respect to the conventional stations.

Our analysis builds upon a previous and recent study ({\textbf{???}}),
which found that disadvantaged areas in Hamilton are better served by
the city's PBSP compared to other Canadian cities {[}i.e., Toronto,
Vancouver, Montreal, and Ottawa-Gatineau{]} with PBSPs where advantaged
areas have greater access. Hosford and Winters ({\textbf{???}})
acknowledge that ``Hamilton stands out in that the lower income
neighborhoods are located near the city center and wealthier
neighborhoods are in the surrounding suburban areas''. Therefore, the
core service area for the PBSP in Hamilton is by default in more
disadvantaged areas compared to other Canadian cities, but there is also
a great deal of variation in income in the city center as well because
of the local university and increasing gentrification in the downtown
area. In their analysis, Hosford and Winters ({\textbf{???}}) did not
differentiate between the conventional stations and those added to
increase equity, which we refer to as equity stations in this paper.
Therefore, this paper continues their work and makes an important
contribution to the growing body of literature that examines equity
issues in PBSPs.

\begin{figure}

{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/hamilton-and-sobi-service-area-1}

}

\caption{The core service area of Hamilton Bike Share is outlined in blue. Dissemination areas in the Hamilton Census Metropolitan Area (CMA) are outlined in grey.}\label{fig:hamilton-and-sobi-service-area}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/sobi-stations-in-hamilton-1}

}

\caption{The spatial distribution of bike share stations in Hamilton, Ontario. The service area of Hamilton Bike Share is outlined in blue and the city's downtown core is outlined in light blue. Dissemination areas within the Hamilton CMA that touch the bounding box of the core service area are outlined in grey.}\label{fig:sobi-stations-in-hamilton}
\end{figure}

\hypertarget{sec:methods}{%
\section{Methods}\label{sec:methods}}

\hypertarget{floating-catchment-area}{%
\subsection{Floating Catchment Area}\label{floating-catchment-area}}

Floating catchment area (FCA) methods are an approach commonly used in
the healthcare accessibility literature. This approach is more
appropriate and informative than calculating the provider-to-population
ratio (PPR) which simply divides the level of supply of a service (e.g.,
the number of bicycle racks at a station) by the population who have
access to the service ({\textbf{???}}). In particular, the Two-Step
Floating Catchment Area method produces flexible catchment areas for
populations accessing a service rather than using rigid boundaries like
PPR, which provides more useful information because it takes into
account individual behaviour and doesn't assume that people stay within
the pre-defined boundaries ({\textbf{???}}). This is an important
property that supports our rationale for applying this method to measure
accessibility to Hamilton Bike Share. The City of Hamilton has
positioned stations between 300 and 600 metres apart, which would lead
to boundaries around these buffers, but it would be reasonable to assume
that people are willing to walk beyond this threshold to access other
stations if the ones nearest them have no supply of bicycles.

More recently, the \emph{balanced} floating catchment area approach was
developed to address issues with demand and supply inflation that are a
result of the overlapping catchment areas produced by the Two-Step
Floating Catchment Area ({\textbf{???}}). This overlap inflates the
demand and generates inaccurate or misleading accessibility estimates
({\textbf{???}}). By adjusting the impedance weights, both supply and
demand are proportionally allocated which eliminates the double-counting
of population that leads to demand and supply inflation
({\textbf{???}}). Other benefits of this adjusted method include
consideration of competition which can occur when catchment areas
overlap, as well as the preservation of system-wide population and level
of service. Balanced floating catchment area methods have been used to
explore accessibility to health care providers ({\textbf{???}}) and
COVID-19 health care services ({\textbf{???}}), but have not yet been
used in the cycling literature to explore issues of accessibility.

In their review of accessibility measures, Geurs and van Wee
({\textbf{???}}) highlight the need for greater inclusion of individual
spatio-temporal constraints but acknowledge the challenges of acquiring
and analyzing person-based data. This comes after Kwan's
({\textbf{???}}) work to show that space-time measures are more capable
of capturing interpersonal differences, especially the effect of
space-time constraints on individual behaviour, and are more helpful for
unraveling gender/ethnic differences. Applying the balanced floating
catchment area approach to examine accessibility to Hamilton Bike Share
constitutes a location-based measure with an individual component by
stratifying according to median household income ({\textbf{???}}), but
conducting a further sensitivity analysis by adjusting distance
thresholds would introduce additional spatio-temporal constraints to
evaluate equitable accessibility.

The first step in the balanced floating catchment area method is to
allocate the population to be serviced by each Hamilton Bike Share
station: \[
P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}
\]

Next, the level of service at each station (i.e., the maximum number of
bicycle racks) is divided by its estimated service population within the
established catchment area: \[
L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
\]

Finally, the accessibility of population cell \(i\) is calculated as the
weighted sum of the level of service of all stations that can be reached
from there according to normalized weights: \[
A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
\]

The approach uses instead a set of suitably normalized weights as
follows: \[
{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
\]

and \[
{w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
\]

These weights satisfy the following properties: \[
\sum_{j = 1}^{J} {w^i_{ji}} = 1
\]

and \[
\sum_{i = 1}^{n} {w^j_{ji}} = 1
\]

Finally, accessibility is calculated without risk of demand or supply
inflation: \[
A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}
\]

\hypertarget{pycnophylactic-interpolation}{%
\subsection{Pycnophylactic
Interpolation}\label{pycnophylactic-interpolation}}

We first plotted the population in all dissemination areas within SoBi
Hamilton's core service area. Since pycnophylactic interpolation occurs
at such a micro scale, we had to ensure that population numbers were not
interpolated or allocated to areas where people do not live in Hamilton
(i.e., where schools or parks are located). To do so, we retrieved
shapefiles for various geographic features (see Table 1) from Open
Hamilton. Figure 3 shows where these spaces are located within SoBi
Hamilton's core service area. Next, we extracted these features from the
PBSP core service area and used pycnophylactic interpolation to
disaggregate and reallocate population from dissemination areas to
smaller polygons. These micro-level polygons are 50 by 50 metres.

\begin{table}

\caption{\label{tab:data-features}\label{tab:landscape-features}Landscape features}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{>{}l|>{\raggedright\arraybackslash}p{30em}}
\toprule
Feature & Description\\
\midrule
\cellcolor{gray!6}{\textbf{Hamilton Bike Share Stations}} & \cellcolor{gray!6}{The location of stations and the number of racks available at each station.}\\
\textbf{Golf Courses} & The location of City and privately owned golf courses.\\
\cellcolor{gray!6}{\textbf{Parks}} & \cellcolor{gray!6}{The location of parks and other green spaces.}\\
\textbf{Designated Large Employment Areas} & The location of large business parks and industrial lands.\\
\cellcolor{gray!6}{\textbf{Municipal Parking Lots}} & \cellcolor{gray!6}{The location of municipal car parks.}\\
\addlinespace
\textbf{Cemeteries} & The location of cemeteries.\\
\cellcolor{gray!6}{\textbf{Environmentally Sensitive Areas}} & \cellcolor{gray!6}{The location of either land or water areas containing natural features or significant ecological functions.}\\
\textbf{Streets} & The street network in Hamilton, including road classification for highways.\\
\cellcolor{gray!6}{\textbf{Educational Institutions}} & \cellcolor{gray!6}{The location of all educational institutions and schools.}\\
\textbf{Places of Worship} & The location of buildings used for religious congregations.\\
\addlinespace
\cellcolor{gray!6}{\textbf{Municipal Service Centres}} & \cellcolor{gray!6}{The location of all municipal service centres, including City Hall.}\\
\textbf{Recreation and Community Centres} & The location of all recreation and community centres.\\
\cellcolor{gray!6}{\textbf{Arenas}} & \cellcolor{gray!6}{The location of all indoor arenas.}\\
\textbf{Emergency Stations} & The location of all Emergency Management Services (EMS) Ambulance stations.\\
\cellcolor{gray!6}{\textbf{Fire Stations}} & \cellcolor{gray!6}{The location of all fire stations.}\\
\addlinespace
\textbf{Police Stations} & The location of all police stations.\\
\cellcolor{gray!6}{\textbf{Railways}} & \cellcolor{gray!6}{The railway network in Hamilton.}\\
\textbf{Hospitals} & The location of all hospitals.\\
\bottomrule
\end{tabular}}
\end{table}

\hypertarget{travel-time-matrix}{%
\subsection{Travel Time Matrix}\label{travel-time-matrix}}

\href{https://download.bbbike.org/osm/bbbike/}{BBBike} is an online
cycle route planner that interfaces with OpenStreetMap. We extracted
OpenStreetMap data for SoBi Hamilton's service area to calculate walking
times from each population cell to nearby bike share stations, using a
walking distance of 10km and a walking time of 30 minutes as a
threshold. A travel time matrix was created with the origins as the
coordinates of the population cells and the destinations as the
coordinates of the bike share stations within the maximum threshold.

\hypertarget{data}{%
\subsection{Data}\label{data}}

All data for this research were accessed from Open Hamilton\footnote{https://open.hamilton.ca/},
an online repository of data curated by the City of Hamilton.

\hypertarget{results}{%
\section{Results}\label{results}}

\begin{figure}

{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-2-1}

}

\caption{Population in all dissemination areas in the SoBi Hamilton core service area and within 30 minutes of walking to a bike share station.}\label{fig:figure-2}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-4-1}

}

\caption{Interpolated population cells in the SoBi Hamilton core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue). The downtown area is outlined in dark green. Geographic features have been extracted. Dissemination area polygons are outlined in grey.}\label{fig:figure-4}
\end{figure}

\hypertarget{accessibility-by-distance-thresholds}{%
\subsection{Accessibility by Distance
Thresholds}\label{accessibility-by-distance-thresholds}}

Consensus regarding the distance that individuals are willing to walk to
access a PBSP is lacking, therefore the literature on walking behaviour
was consulted to determine the thresholds for the sensitivity analysis.
Previous studies have found that living within 250 metres
({\textbf{???}}) and 300 metres ({\textbf{???}}) is correlated with bike
share use, while other research has found that walking trips are less
than 600 metres and rarely more than 1200 metres ({\textbf{???}}) or a
median distance of 650 metres ({\textbf{???}}). As seen in Table
\ref{tab:accessibility-sensibility}, we find that accessibility
increases with a threshold between two and four minutes, but is then
maximized at 5 minutes. At some stations, Hamilton Bike Share will
depict a map to show the user the locations of the other nearest
stations within a five minute walk, which suggests that this is an
average distance that people are anticipated to walk. Accessibility
decreases significantly after eight minutes, which is intuitive given
that demand on a limited supply increases as more people can reach each
bike share station.

For this reason, we experiment with various walking thresholds by
conducting a sensitivity analysis to calculate accessibility at
different walking times: 3 minutes, 5 minutes, 10 minutes, and 15
minutes. We categorize these thresholds as minimum, average, maximum,
and extreme, respectively. At each threshold, we compare accessibility
between the current system and the original system to examine the
contribution of the additional equity stations.

\hypertarget{three-minutes}{%
\subsubsection{Three Minutes}\label{three-minutes}}

At this threshold, we find that there are 25.2 bicycle racks per person
in the original system. With the addition of equity stations, there are
now 25.4 bicycles per person. Figure \ref{fig:figure-6} presents a
comparison of accessibility between the systems. Accessibility is fairly
uniform, with the exception of two small areas near the university and
the waterfront park where accessibility is slightly higher. Equity
stations are shown in green, which helps to demonstrate how they
increase accessibility between the systems.

\begin{figure}

{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-6-1}

}

\caption{Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations.}\label{fig:figure-6}
\end{figure}

\hypertarget{five-minutes}{%
\subsubsection{Five Minutes}\label{five-minutes}}

At this threshold, we find that there are 68.6 bicycle racks per person
in the original system. With the addition of equity stations, there are
now 68.8 bicycles per person. Figure \ref{fig:figure-7} presents a
comparison of accessibility between the systems. Again, accessibility is
fairly uniform, with the exception of one area near the waterfront park
where accessibility is higher.

\begin{figure}

{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-7-1}

}

\caption{Accessibility at 5 minutes walk (average threshold) compared between current system with equity stations and the original system without equity stations.}\label{fig:figure-7}
\end{figure}

\hypertarget{ten-minutes}{%
\subsubsection{Ten Minutes}\label{ten-minutes}}

At this threshold, we find that there are 3.61 bicycle racks per person
in the original system. With the addition of equity stations, there are
now 3.74 bicycles per person. Figure \ref{fig:figure-8} presents a
comparison of accessibility between the systems. We begin to see
differences in accessibility across the service area, with users near
the university and its adjacent neighbourhoods, as well as
neighbourhoods north of the downtown area have slightly higher
accessibility. While the differences are modest, they are more apparent
at this threshold than at shorter walking distances.

\begin{figure}

{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-8-1}

}

\caption{Accessibility at 10 minutes walk (maximum threshold) compared between current system with equity stations and the original system without equity stations.}\label{fig:figure-8}
\end{figure}

\hypertarget{fifteen-minutes}{%
\subsubsection{Fifteen Minutes}\label{fifteen-minutes}}

At this threshold, we find that there are 2.44 bicycle racks per person
in the original system. With the addition of equity stations, there are
now 2.55 bicycles per person. Figure \ref{fig:figure-9} presents a
comparison of accessibility between the systems. Users near the
university and the neighbourhoods north of the downtown area (outlined
in green) have the highest accessibility, followed by those who live in
the city's downtown area. Accessibility in the east end of the core
service area remains low.

\begin{figure}

{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-9-1}

}

\caption{Accessibility at 15 minutes walk (extreme threshold) compared between current system with equity stations and the original system without equity stations.}\label{fig:figure-9}
\end{figure}

\hypertarget{accessibility-by-median-total-income}{%
\subsection{Accessibility by Median Total
Income}\label{accessibility-by-median-total-income}}

Approximately 138,000 people live within Hamilton Bike Share's core
service area or within a 10 minute walk to the service area. We find
that over 118,000 people can access a bicycle share station within a 15
minute walk, which represents roughly 85\% of the total population in
Hamilton Bike Share's core service area {[}see Table
\ref{tab:table-2}{]}. While previous research found that neighbourhoods
with more disadvantage are better serviced by Hamilton Bike Share
({\textbf{???}}), the authors used the Pampalon Deprivation Index from
2011 as a measure of socioeconomic status. Instead, we use median total
income for each dissemination area to examine whether disparities in
accessibility exist within the core service area. We divide income by
quantiles: bottom 20\%, second 20\%, third 20\%, fourth 20\%, and top
20\%. One of the unique properties of the balanced floating catchment
area method is that data can be reaggregated from small to larger
polygons while preserving the total population and number of bicycle
racks at each station. This avoids demand and supply inflation, and also
enables us to present findings in a way that is easier to interpret
(i.e., at the dissemination area level).

Panels 1 and 2 depict bivariate choropleth maps with the combined
spatial distribution of accessibility and median total income per
household for the different thresholds. Our analysis demonstrates that
the stations added to Hamilton's public bicycle share program to
increase equity for disadvantaged neighbourhoods achieved their goal by
increasing accessibility, albeit only modestly. By implementing equity
stations in more areas with lower median total income in Hamilton, the
PBSP has achieved greater horizontal equity by extending the spatial
distribution of bicycles across the city. This is particularly evident
at shorter walking distances of three and five minutes.

However, we found that dissemination areas with lower total household
income do not necessarily have greater access to the program {[}see
Table \ref{tab:table-3}{]}. Income disparities still persist, however
only at certain thresholds. With and without equity stations, people in
the top 20\% of income have the highest level of accessibility at a
threshold of ten and fifteen minutes. At distance thresholds of three
and five minutes, people in the second 20\% have the highest level of
accessibility by a significant amount. This is due to the concentration
of stations near or within the city's downtown area and the ``lower
city'' which historically has been more disadvantaged compared to the
surrounding suburban neighbourhoods. This is an artifact of the city's
steel industry which led to more blue collar and working class people
living in the city's downtown area and ``lower city''. However, the
bottom 20\%, who may benefit the most from Hamilton Bike Share's equity
initiatives, have the lowest accessibility at three minutes threshold
and the second lowest accessibility at all other thresholds.

On the whole, we find that areas that have less than their equitable
share of the level of service are some of the most socioeconomically
disadvantaged areas in the city. This finding aligns with other studies
that have found disparities in station or bike location along income
lines in New York City ({\textbf{???}}), Tampa ({\textbf{???}}), and
Seattle ({\textbf{???}}).

\begin{figure}
\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): without equity stations (top panel) and with equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-3}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): without equity stations (top panel) and with equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-5}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): without equity stations (top panel) and with equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-10}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): without equity stations (top panel) and equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-15}
\end{figure}

\begin{table}

\caption{\label{tab:accessibility-income}\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{lcccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Without Equity Stations} & \multicolumn{2}{c}{With Equity Stations} & \multicolumn{2}{c}{Difference} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
Income Quintile & Accessibility & Population & Accessibility & Population & Accessibility & Population\\
\midrule
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Threshold - 3 minutes}}\\
\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{2.377} & \cellcolor{gray!6}{22359} & \cellcolor{gray!6}{2.424} & \cellcolor{gray!6}{22798} & \cellcolor{gray!6}{0.047} & \cellcolor{gray!6}{439}\\
\hspace{1em}Second 20\% & 12.203 & 9347 & 12.281 & 12420 & 0.078 & 3073\\
\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{3.093} & \cellcolor{gray!6}{7745} & \cellcolor{gray!6}{3.156} & \cellcolor{gray!6}{9455} & \cellcolor{gray!6}{0.063} & \cellcolor{gray!6}{1710}\\
\hspace{1em}Fourth 20\% & 4.119 & 1673 & 4.119 & 1673 & 0.000 & 0\\
\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{3.757} & \cellcolor{gray!6}{2151} & \cellcolor{gray!6}{3.784} & \cellcolor{gray!6}{2416} & \cellcolor{gray!6}{0.027} & \cellcolor{gray!6}{265}\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Threshold - 5 minutes}}\\
\hspace{1em}Bottom 20\% & 1.302 & 35477 & 1.357 & 35803 & 0.055 & 326\\
\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{56.048} & \cellcolor{gray!6}{17513} & \cellcolor{gray!6}{56.137} & \cellcolor{gray!6}{22908} & \cellcolor{gray!6}{0.089} & \cellcolor{gray!6}{5395}\\
\hspace{1em}Third 20\% & 4.259 & 15117 & 4.291 & 18309 & 0.032 & 3192\\
\hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{1.094} & \cellcolor{gray!6}{2867} & \cellcolor{gray!6}{1.095} & \cellcolor{gray!6}{3116} & \cellcolor{gray!6}{0.001} & \cellcolor{gray!6}{249}\\
\hspace{1em}Top 20\% & 6.256 & 4074 & 6.264 & 4540 & 0.008 & 466\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Threshold - 10 minutes}}\\
\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{0.604} & \cellcolor{gray!6}{41824} & \cellcolor{gray!6}{0.622} & \cellcolor{gray!6}{41981} & \cellcolor{gray!6}{0.018} & \cellcolor{gray!6}{157}\\
\hspace{1em}Second 20\% & 0.862 & 27546 & 0.929 & 30503 & 0.067 & 2957\\
\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{0.776} & \cellcolor{gray!6}{22394} & \cellcolor{gray!6}{0.802} & \cellcolor{gray!6}{25128} & \cellcolor{gray!6}{0.026} & \cellcolor{gray!6}{2734}\\
\hspace{1em}Fourth 20\% & 0.225 & 4544 & 0.227 & 4989 & 0.002 & 445\\
\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{1.346} & \cellcolor{gray!6}{7989} & \cellcolor{gray!6}{1.348} & \cellcolor{gray!6}{9078} & \cellcolor{gray!6}{0.002} & \cellcolor{gray!6}{1089}\\
\addlinespace[0.3em]
\multicolumn{7}{l}{\textbf{Threshold - 15 minutes}}\\
\hspace{1em}Bottom 20\% & 0.536 & 42208 & 0.557 & 42327 & 0.021 & 119\\
\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{0.555} & \cellcolor{gray!6}{30507} & \cellcolor{gray!6}{0.614} & \cellcolor{gray!6}{31069} & \cellcolor{gray!6}{0.059} & \cellcolor{gray!6}{562}\\
\hspace{1em}Third 20\% & 0.554 & 26108 & 0.581 & 26660 & 0.027 & 552\\
\hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{0.093} & \cellcolor{gray!6}{6312} & \cellcolor{gray!6}{0.096} & \cellcolor{gray!6}{7435} & \cellcolor{gray!6}{0.003} & \cellcolor{gray!6}{1123}\\
\hspace{1em}Top 20\% & 0.808 & 10209 & 0.811 & 11089 & 0.003 & 880\\
\bottomrule
\multicolumn{7}{l}{\rule{0pt}{1em}\textit{Note: }}\\
\multicolumn{7}{l}{\rule{0pt}{1em} }\\
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} With equity stations = Hamilton Bike Share current system (118 conventional stations, 12 equity stations)}\\
\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{b} Without equity stations = Hamilton Bike Share original system (118 conventional stations, no equity stations)}\\
\end{tabular}}
\end{table}

\hypertarget{study-limitations}{%
\section{Study Limitations}\label{study-limitations}}

This paper did not examine or compare ridership data between
conventional and equity stations. Therefore, further research is needed
to determine whether this encouraged more cycling for low-income
individuals living near equity stations. Other studies have specifically
looked at differences in trip type, frequency, or length among users
from disadvantaged neighbourhoods ({\textbf{???}}; {\textbf{???}};
{\textbf{???}}), but our analysis is limited by the lack of accessible
data to conduct similar analyses for Hamilton Bike Share.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

The addition of specific equity stations to the public bicycle share
program in Hamilton, Ontario had the net effect of increasing
accessibility and reducing both horizontal and vertical inequities. In
particular, accessibility improved the most for the second 20\% income
groups at all thresholds, but the gains were only modest. Congestion
effects were observed at higher thresholds, with accessibility
decreasing significantly once the catchment area is increased to 10
minutes walking. However, We identified specific areas that have both
low accessibility and low income which would benefit from an increased
supply of bicycles. These are ideal candidates for new equity stations.
In this way, our paper has made contributions in a positive way by
applying an intuitive and useful approach to measure accessibility to a
PBSP, and in a normative way by serving to inform future investments in
cycle infrastructure for Hamilton Bike Share.

Wang and Lindsey ({\textbf{???}}) have noted that there is a lack of
research that examines how bike share users' behaviour changes as a
result of program changes to station locations or improvements in
accessibility. As such, a logical next step to this research is to
examine whether Hamilton Bike Share's equity stations increased
ridership or resulted in new memberships in areas that were previously
under-served. An examination of the types of trips undertaken by
residents in these areas would also be informative, such as the study
undertaken by Caspi and Norland ({\textbf{???}}) after bike share
stations were implemented in low-income Philadelphia neighbourhoods. The
bulk of cycling facilities that have been built in Hamilton to date
located in the core service area near the conventional stations. It
would be worthwhile to explore the route choice of bike share trips
departing or ending at the equity stations and to identify factors that
specifically influence trips from these stations, which would extend
existing studies conducted by Scott and colleagues ({\textbf{???}};
{\textbf{???}}; {\textbf{???}}). This paper, combined with additional
studies such as those conceptualized above, would serve as a valuable
case study for Hamilton and other cities with PBSPs who wish to evaluate
and address spatial inequities in accessibility and transportation
options in urban areas.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\end{document}

### 2021-03-19 sobiEquity/raw-data-processing/00-Data-Processing-for-sobiEquity.Rmd
Revised in hub in data package and worked with figures and tables for style

@@
 sobi_hubs <- sobi_hubs %>%
-  dplyr::mutate(hub_status = case_when(OBJECTID == 114 ~ "Deactivated",
-                                       OBJECTID == 115 ~ "Deactivated",
+  dplyr::mutate(hub_status = case_when(OBJECTID == 115 ~ "Deactivated",
                                        OBJECTID == 116 ~ "Deactivated",

### 2021-03-24 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Manuscript draft completed and sent to co-authors for review

@@
 Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021].
-The core service area spans 35 km^2 of the city and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure 1].
+The core service area spans 35 km^2 of the city and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure \ref{fig:hamilton-and-sobi-service-area}].
 This represents roughly one fifth of the population in Hamilton, according to the 2016 Canadian Census.
@@
 A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019].
-As of June 2020, the bike share program has 900 bikes and 130 stations, and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
+As of June 2020, the bike share program has 900 bikes and 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
 
@@
 
-```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The core service area of Hamilton Bike Share is outlined in blue.
-Dissemination areas in the Hamilton Census Metropolitan Area (CMA) are outlined in grey."}
+```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The core service area of Hamilton Bike Share is outlined in black.
+Hamilton Census Metropolitan Area (CMA) is shown in grey."}
 ggplot() + 
@@
 ```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bike share stations in Hamilton, Ontario.
-The service area of Hamilton Bike Share is outlined in blue and the city's downtown core is outlined in light blue.
-Dissemination areas within the Hamilton CMA that touch the bounding box of the core service area are outlined in grey."}
+The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in dark green.
+Hamilton CMA is shown in grey."}
 ggplot() + 
@@
 
-We first plotted the population in all dissemination areas within SoBi Hamilton's core service area.
-Since pycnophylactic interpolation occurs at such a micro scale, we had to ensure that population numbers were not interpolated or allocated to areas where people do not live in Hamilton (i.e., where schools or parks are located).
-To do so, we retrieved shapefiles for various geographic features (see Table 1) from Open Hamilton.
-Figure 3 shows where these spaces are located within SoBi Hamilton's core service area.
-Next, we extracted these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population from dissemination areas to smaller polygons.
-These micro-level polygons are 50 by 50 metres.
+Following the method first presented by Tobler [-@toblerSmoothPycnophylacticInterpolation1979], we used pycnophylactic interpolation to disaggregate population from each dissemination area to smaller polygons that are 50 by 50 metres.
+This method involves smoothing out the population from each dissemination area while preserving total volume [see Figure \ref{fig:da-population}].
+Since pycnophylactic interpolation occurs at such a micro scale, we had to ensure that population numbers were not reallocated or smoothed out to areas where people do not live in Hamilton (i.e., where schools or parks are located).
+To do so, we retrieved shapefiles for various geographic features [see Table \ref{tab:data-features}] from Open Hamilton.
+Next, we extracted these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population [see Figure \ref{fig:interpolated-population}].
 
@@
       booktabs = TRUE,
-      caption = "\\label{tab:landscape-features}Landscape features") %>%
+      caption = "\\label{tab:landscape-features}Landscape features extracted from the Hamilton Bike Share core service area before population is interpolated.") %>%
   kable_paper(full_width = F) %>%
@@
 
-## Travel Time Matrix
-
-[BBBike](https://download.bbbike.org/osm/bbbike/) is an online cycle route planner that interfaces with OpenStreetMap.
-We extracted OpenStreetMap data for SoBi Hamilton's service area to calculate walking times from each population cell to nearby bike share stations, using a walking distance of 10km and a walking time of 30 minutes as a threshold.
-A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bike share stations within the maximum threshold.
-
-## Data
-
-All data for this research were accessed from Open Hamilton^[https://open.hamilton.ca/], an online repository of data curated by the City of Hamilton.
-
-Results
-===================
-
-```{r figure-2, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Population in all dissemination areas in the SoBi Hamilton core service area and within 30 minutes of walking to a bike share station."}
+```{r da-population, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Population in all dissemination areas (outlined in white) in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the service area."}
 ggplot() +
@@
 
-```{r figure-4, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Interpolated population cells in the SoBi Hamilton core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
+```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
 The downtown area is outlined in dark green.
@@
 
-<!-- Suggest to remove this image: all ithe information it gives is already found in other figures: the location of the SoBi stations and the interpolated popualtion
-
-```{r figure-5, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Interpolated population cells in the SoBi Hamilton core service area and within 30 minutes of walking to the core service area.
-Green circles represent the location of the equity stations that were added in 2018.
-Blue circles represent the conventional stations.
-The service area of the PBSP is outlined in black, and dissemination areas from the 2016 Canadian Census are outlined in light gray."}
-ggplot() + 
-  geom_sf(data = cropped_da, 
-          color = NA, 
-          fill = "gray") + 
-  geom_sf(data = population_50x50, 
-          aes(color = population)) +
-  geom_sf(data = sobi_hubs %>% 
-            dplyr::filter(hub_type == "Conventional" &
-                            hub_status == "Active"), 
-          size = 1.5, 
-          shape = 21, 
-          fill = "blue") +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "ERI"), 
-          size = 1.5, shape = 21, 
-          fill = "green") +
-  geom_sf(data = sobi_service, 
-          fill = NA, 
-          colour = "black") +
-  scale_color_distiller(palette = "YlOrRd", 
-                        direction = 1) +
-  geom_sf(data = community_downtown %>%
-            dplyr::filter(NAME == "Hamilton"),
-          fill = NA,
-          color = "dark green") +
-  facet_grid(rows = vars(hub_type)) +
-  theme(panel.background = element_rect(fill = "white"))
-```
---->
+## Travel Time Matrix
+
+[BBBike](https://download.bbbike.org/osm/bbbike/) is an online cycle route planner that interfaces with OpenStreetMap.
+We extracted OpenStreetMap data for SoBi Hamilton's service area to calculate walking times from each population cell to nearby bike share stations, using a walking distance of 10km and a walking time of 30 minutes as a threshold.
+A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bike share stations within the maximum threshold.
+
+## Data
+
+All data for this research were accessed from Open Hamilton^[https://open.hamilton.ca/], an online repository of data curated by the City of Hamilton.
+
+Results
+===================
 
@@
 Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010].
-As seen in Table \ref{tab:accessibility-sensibility}, we find that accessibility increases with a threshold between two and four minutes, but is then maximized at 5 minutes.
+We found that accessibility increases with a threshold between two and four minutes, but is then maximized at 5 minutes.
 At some stations, Hamilton Bike Share will depict a map to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are anticipated to walk.
@@
 
-```{r accessibility-sensibility, eval=FALSE, include = FALSE}
+```{r join-population-to-travel-time-data, include=FALSE}
+# current system with equity stations
+ttm_walk <- left_join(ttm_walk, 
+                      population_50x50 %>%
+                        st_drop_geometry(),
+                      by = "UID")
+```
+
+```{r join-sobi-station-to-travel-time-data, include=FALSE}
+# current system with equity stations
+ttm_walk <- ttm_walk %>% 
+  left_join(sobi_hubs %>%
+              st_drop_geometry() %>%
+              dplyr::select(OBJECTID, 
+                            RACKS_AMOU),
+            by = c("OBJECTID" = "OBJECTID"))
+```
+
+```{r, include=FALSE}
+#data frames for current and original systems
+names(ttm_walk) <- c("UID", 
+                     "hub", 
+                     "travel_time",
+                     "hub_type",
+                     "hub_status",
+                     "population",
+                     "racks")
+```
+
+```{r accessibility-sensibility, eval=FALSE, include=FALSE}
 # Testing different thresholds to check how accessibility changes
@@
 
-```{r join-population-to-travel-time-data, include=FALSE}
-# current system with equity stations
-ttm_walk <- left_join(ttm_walk, 
-                      population_50x50 %>%
-                        st_drop_geometry(),
-                      by = "UID")
-```
-
-```{r join-sobi-station-to-travel-time-data, include=FALSE}
-# current system with equity stations
-ttm_walk <- ttm_walk %>% 
-  left_join(sobi_hubs %>%
-              st_drop_geometry() %>%
-              dplyr::select(OBJECTID, 
-                            RACKS_AMOU),
-            by = c("OBJECTID" = "OBJECTID"))
-```
-
-```{r, include=FALSE}
-#data frames for current and original systems
-names(ttm_walk) <- c("UID", 
-                     "hub", 
-                     "travel_time",
-                     "hub_type",
-                     "hub_status",
-                     "population",
-                     "racks")
-```
-
 <!-- Calculate accessibility for different thresholds -->
@@
 
-At this threshold, we find that there are 25.2 bicycle racks per person in the original system.
+At the minimum threshold, we find that there are 25.2 bicycle racks per person in the original system.
 With the addition of equity stations, there are now 25.4 bicycles per person.
@@
 Accessibility is fairly uniform, with the exception of two small areas near the university and the waterfront park where accessibility is slightly higher.
-Equity stations are shown in green, which helps to demonstrate how they increase accessibility between the systems.
 
@@
 
-At this threshold, we find that there are 68.6 bicycle racks per person in the original system.
+At the average threshold, we find that there are 68.6 bicycle racks per person in the original system.
 With the addition of equity stations, there are now 68.8 bicycles per person.
 Figure \ref{fig:figure-7} presents a comparison of accessibility between the systems.
-Again, accessibility is fairly uniform, with the exception of one area near the waterfront park where accessibility is higher.
+Again, accessibility is fairly uniform, with the exception of one very small area near the waterfront park where accessibility is higher.
 
@@
 
-At this threshold, we find that there are 3.61 bicycle racks per person in the original system.
+At the maximum threshold, we find that there are 3.61 bicycle racks per person in the original system.
 With the addition of equity stations, there are now 3.74 bicycles per person.
 Figure \ref{fig:figure-8} presents a comparison of accessibility between the systems.
-We begin to see differences in accessibility across the service area, with users near the university and its adjacent neighbourhoods, as well as neighbourhoods north of the downtown area have slightly higher accessibility.
+We begin to see differences in accessibility across the service area, with users near the university and its adjacent neighbourhoods, as well as neighbourhoods north of the downtown area, have slightly higher accessibility.
 While the differences are modest, they are more apparent at this threshold than at shorter walking distances.
@@
 
-At this threshold, we find that there are 2.44 bicycle racks per person in the original system.
+At the extreme threshold, we find that there are 2.44 bicycle racks per person in the original system.
 With the addition of equity stations, there are now 2.55 bicycles per person.
@@
 
-Approximately 138,000 people live within Hamilton Bike Share's core service area or within a 10 minute walk to the service area.
-We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in Hamilton Bike Share's core service area [see Table \ref{tab:table-2}].
-While previous research found that neighbourhoods with more disadvantage are better serviced by Hamilton Bike Share [@hosfordWhoArePublic2018], the authors used the Pampalon Deprivation Index from 2011 as a measure of socioeconomic status.
-Instead, we use median total income for each dissemination area to examine whether disparities in accessibility exist within the core service area.
+Approximately 138,000 people live within Hamilton Bike Share's core service area or within a 30 minute walk to the service area.
+We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}].
+While previous research found that neighbourhoods with more disadvantage are better serviced by Hamilton Bike Share [@hosfordWhoArePublic2018], the authors used the Pampalon Deprivation Index from 2011 as a measure of socioeconomic status for dissemination areas across the city not just within the core service area.
+Instead, we use median total income for each dissemination area in the core service area to examine whether disparities in accessibility exist between income groups for those who live in or near the core service area.
 We divide income by quantiles: bottom 20%, second 20%, third 20%, fourth 20%, and top 20%.
-One of the unique properties of the balanced floating catchment area method is that data can be reaggregated from small to larger polygons while preserving the total population and number of bicycle racks at each station.
-This avoids demand and supply inflation, and also enables us to present findings in a way that is easier to interpret (i.e., at the dissemination area level).
-
-Panels 1 and 2 depict bivariate choropleth maps with the combined spatial distribution of accessibility and median total income per household for the different thresholds.
-Our analysis demonstrates that the stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing accessibility, albeit only modestly.
+One of the unique properties of the balanced floating catchment area method is that data can be reaggregated from small to larger polygons while preserving the total population and supply at each station.
+This avoids demand and supply inflation, and also enables us to present findings in a way that is easier to interpret (i.e., across dissemination areas instead of 50 by 50 metre population cells).
+
+Panels 1[\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combined spatial distribution of accessibility and median total income per household for the different thresholds.
+Our analysis demonstrates that the extreme threshold of fifteen minutes allows the most people to access a station.
+We find that stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing accessibility, albeit only modestly [see Table \ref{tab:accessibility-income}].
 By implementing equity stations in more areas with lower median total income in Hamilton, the PBSP has achieved greater horizontal equity by extending the spatial distribution of bicycles across the city.
-This is particularly evident at shorter walking distances of three and five minutes.
-
-However, we found that dissemination areas with lower total household income do not necessarily have greater access to the program [see Table \ref{tab:table-3}].
+This is particularly evident at the minimum and average thresholds of three and five minutes, respectively.
+The largest gains were made for dissemination areas in the second 20%, where an additional 3,073 and 5,395 people could reach a bicycle share station within three and five minutes walk, respectively, after the addition of the equity station.
+
+However, we found that dissemination areas with the lowest total household income do not have much greater access to the program [see Table \ref{tab:table-3}].
 Income disparities still persist, however only at certain thresholds.
 With and without equity stations, people in the top 20% of income have the highest level of accessibility at a threshold of ten and fifteen minutes.
-At distance thresholds of three and five minutes, people in the second 20% have the highest level of accessibility by a significant amount.
-This is due to the concentration of stations near or within the city's downtown area and the "lower city" which historically has been more disadvantaged compared to the surrounding suburban neighbourhoods.
-This is an artifact of the city's steel industry which led to more blue collar and working class people living in the city's downtown area and "lower city".
-However, the bottom 20%, who may benefit the most from Hamilton Bike Share's equity initiatives, have the lowest accessibility at three minutes threshold and the second lowest accessibility at all other thresholds.
+Although dissemination areas in the second 20% have the highest level of accessibility by a significant amount at lower thresholds, the bottom 20%, who may benefit the most from Hamilton Bike Share's equity initiatives, have the lowest accessibility at three minutes threshold and the second lowest accessibility at all other thresholds.
+Panels 1[\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] reveal specific areas where accessibility is low or zero, which are good candidates for additional equity stations.
 
 On the whole, we find that areas that have less than their equitable share of the level of service are some of the most socioeconomically disadvantaged areas in the city.
-This finding aligns with other studies that have found disparities in station or bike location along income lines in New York City [@babagoliExploringHealthSpatial2019], Tampa [@chenExploringEquityPerformance2019], and Seattle [@mooneyFreedomStationSpatial2019].
+While the addition of equity stations increases accessibility for all income groups at all thresholds, they did not increase accessibility significantly for any single income group.
+This finding aligns with other studies that have found disparities in station or bike location along income lines in New York City [@babagoliExploringHealthSpatial2019], Tampa [@chenExploringEquityPerformance2019], and Seattle [@mooneyFreedomStationSpatial2019], where people who have greater advantage have better access to PBSPs.
 
@@
       align = c("l", "c", "c", "c", "c", "c", "c"),
-      caption = "\\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile") %>%
+      caption = "\\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile and between systems (with and without equity stations).") %>%
   kable_styling(latex_options = c("striped", "scale_down")) %>%
@@
 This paper did not examine or compare ridership data between conventional and equity stations.
-Therefore, further research is needed to determine whether this encouraged more cycling for low-income individuals living near equity stations.
-Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighbourhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of accessible data to conduct similar analyses for Hamilton Bike Share.
+Therefore, further research is needed to determine whether the addition of equity stations encouraged more cycling for low-income individuals living near them.
+Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighbourhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of accessible route or user data to conduct similar analyses for Hamilton Bike Share.
 
@@
 The addition of specific equity stations to the public bicycle share program in Hamilton, Ontario had the net effect of increasing accessibility and reducing both horizontal and vertical inequities.
-In particular, accessibility improved the most for the second 20% income groups at all thresholds, but the gains were only modest.
+In particular, accessibility improved the most for the second 20% median total income groups at all thresholds, but the gains were only modest for all income groups.
+Dissemination areas with the bottom 20% had the lowest accessibility at three minutes, and second lowest level of accessibility at five, ten, and fifteen minutes.
 Congestion effects were observed at higher thresholds, with accessibility decreasing significantly once the catchment area is increased to 10 minutes walking.
-However, We identified specific areas that have both low accessibility and low income which would benefit from an increased supply of bicycles.
+Based on our analysis, we identified specific areas that have both low accessibility and low income which would benefit from an increased supply of bicycles.
 These are ideal candidates for new equity stations.

### 2021-03-24 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Manuscript draft completed and sent to co-authors for review

@@
 cities worldwide and a great deal has been learned about their typical
-users ({\textbf{???}}). In many cities, males use bike share more than
-females ({\textbf{???}}; {\textbf{???}}; {\textbf{???}}; {\textbf{???}})
-as do younger age cohorts ({\textbf{???}}; {\textbf{???}};
-{\textbf{???}}). However, one study found that bike share users in
-Washington, DC were more likely to be female ({\textbf{???}}), which
-suggests that the gender gap among cyclists who use bike share is less
-disparate than the gap for personal bicycle use ({\textbf{???}}). There
-is some evidence that bike share users are less likely to own a car
-({\textbf{???}}; {\textbf{???}}). However, the relationship between
-income or education and bike share use is less clear. Stations in
-disadvantaged communities in Chicago generated most of the average
-annual trips ({\textbf{???}}) and individuals from minority or lower
-socioeconomic status neighbourhoods in Minneapolis-St.~Paul used the
-city's PBSP more ({\textbf{???}}). Being university educated was also a
-significant correlate of bike share use in Montreal, Canada
-({\textbf{???}}). On the other hand, members who reside in
+users (Fishman, 2016). In many cities, males use bike share more than
+females (Brey et al., 2017; Nickkar et al., 2019; Reilly, S. M. Wang, et
+al., 2020; Winters et al., 2019) as do younger age cohorts (Brey et al.,
+2017; Buck et al., 2013; Fuller et al., 2011). However, one study found
+that bike share users in Washington, DC were more likely to be female
+(Buck et al., 2013), which suggests that the gender gap among cyclists
+who use bike share is less disparate than the gap for personal bicycle
+use (Fishman, 2016). There is some evidence that bike share users are
+less likely to own a car (Buck et al., 2013; Reilly, Noyes, et al.,
+2020). However, the relationship between income or education and bike
+share use is less clear. Stations in disadvantaged communities in
+Chicago generated most of the average annual trips (Qian and Jaller,
+2020) and individuals from minority or lower socioeconomic status
+neighbourhoods in Minneapolis-St.~Paul used the city's PBSP more (J.
+Wang and Lindsey, 2019a). Being university educated was also a
+significant correlate of bike share use in Montreal, Canada (Fuller et
+al., 2011). On the other hand, members who reside in
 minority-concentrated and lower socioeconomic status neighbourhoods use
-the Minneapolis-St Paul bike share more frequently ({\textbf{???}}).
-Financial savings have been found to motivate those on a low income to
-use bike share ({\textbf{???}}). Many studies have found that proximity
-to a bike share station, either living or working near one, is an
-important determinant of use or having a membership ({\textbf{???}};
-{\textbf{???}}). This makes sense given that individuals are more likely
-to use services or programs that they can easily access. Finally, the
-potential of bike share programs to increase cycling levels has also
-been explored recently ({\textbf{???}}; {\textbf{???}}), but more
+the Minneapolis-St Paul bike share more frequently (J. Wang and Lindsey,
+2019a). Financial savings have been found to motivate those on a low
+income to use bike share (Fishman, 2016). Many studies have found that
+proximity to a bike share station, either living or working near one, is
+an important determinant of use or having a membership (Fishman, 2016;
+Fuller et al., 2011). This makes sense given that individuals are more
+likely to use services or programs that they can easily access. Finally,
+the potential of bike share programs to increase cycling levels has also
+been explored recently (Hosford et al., 2018, 2019), but more
 longitudinal evidence is needed to determine their impact on encouraging
@@
 use beyond the ``typical'' users, which has been the focus of recent
-research ({\textbf{???}}; {\textbf{???}}; {\textbf{???}}), and to offer
-more people the option of using sustainable and active transportation.
+research (Auchincloss et al., 2020; Hull Grasso et al., 2020; MacArthur
+et al., 2020), and to offer more people the option of using sustainable
+and active transportation.
 
@@
 greater access for vulnerable or disadvantaged populations (e.g.,
-vertical equity) ({\textbf{???}}). Both are of interest to researchers
-and transport planners since they are often linked in that advantage, or
-conversely disadvantage, has spatial patterns. Using a negative binomial
-regression model, Qian and Jaller ({\textbf{???}}) estimated ridership
-in Chicago's PBSP among disadvantaged communities and found some
-disparities. While annual members in disadvantaged communities have a
-significantly lower share of trips compared to other areas in the city,
-they make longer trips. This suggests that they may be using PBSP for
-work commuting, which points to the importance of ensuring equitable
-access ({\textbf{???}}). Similar results were found in Philadelphia,
-where lower income areas generate fewer trips and efforts to increase
-equity within the program have not been as successful as intended
-({\textbf{???}}). In the case of Seattle, all neighbourhoods had some
-level of access to dockless bikes but those with higher incomes and more
-residents of higher education had more bikes ({\textbf{???}}). Babagoli
-et al.~({\textbf{???}}) found that neighbourhoods in New York City with
-higher affluence had the greater proportion of Citi Bike stations.
-Overall, these findings suggest that horizontal equity can be achieved
-while vertical equity is harder to attain. Exploring transport equity by
-investigating where bicycle share stations are located, often using
-neighbourhoods or census tracts as the spatial unit of analysis, can
-ignore or miss the benefits that may be derived from adjacent zones.
-Meaning that, stations may be lacking in certain neighbourhoods but
-there may be stations accessible within a reasonable walking cutoff
-time. This is where accessibility becomes an important consideration.
+vertical equity) (Chen et al., 2019). Both are of interest to
+researchers and transport planners since they are often linked in that
+advantage, or conversely disadvantage, has spatial patterns. Using a
+negative binomial regression model, Qian and Jaller (2020) estimated
+ridership in Chicago's PBSP among disadvantaged communities and found
+some disparities. While annual members in disadvantaged communities have
+a significantly lower share of trips compared to other areas in the
+city, they make longer trips. This suggests that they may be using PBSP
+for work commuting, which points to the importance of ensuring equitable
+access (Qian and Jaller, 2020). Similar results were found in
+Philadelphia, where lower income areas generate fewer trips and efforts
+to increase equity within the program have not been as successful as
+intended (Caspi and Noland, 2019). In the case of Seattle, all
+neighbourhoods had some level of access to dockless bikes but those with
+higher incomes and more residents of higher education had more bikes
+(Mooney et al., 2019). Babagoli et al.~(2019) found that neighbourhoods
+in New York City with higher affluence had the greater proportion of
+Citi Bike stations. Overall, these findings suggest that horizontal
+equity can be achieved while vertical equity is harder to attain.
+Exploring transport equity by investigating where bicycle share stations
+are located, often using neighbourhoods or census tracts as the spatial
+unit of analysis, can ignore or miss the benefits that may be derived
+from adjacent zones. Meaning that, stations may be lacking in certain
+neighbourhoods but there may be stations accessible within a reasonable
+walking cutoff time. This is where accessibility becomes an important
+consideration.
 
 Accessibility has been applied in both a positive and normative way to
-inform transportation planning ({\textbf{???}}), but its utility to this
-field has evolved since its conceptualization and become linked with
-recent interest in prioritizing local proximity and modes that are
-suitable for local trips like walking and cycling ({\textbf{???}}). As a
+inform transportation planning (Páez et al., 2012), but its utility to
+this field has evolved since its conceptualization and become linked
+with recent interest in prioritizing local proximity and modes that are
+suitable for local trips like walking and cycling (Levine, 2020). As a
 measure of the ease of reaching potential destinations spread spatially
@@
 transportation network by measuring person-based indicators and
-exploring differences in use between population groups ({\textbf{???}}).
+exploring differences in use between population groups (Páez et al.,
+2012).
 
 In the context of PBSPs, accessibility refers to the distance an
-individual must travel to reach a bicycle share station ({\textbf{???}};
-{\textbf{???}}). Since the time or distance that members of the
-population need to reach a bicycle share station decreases the potential
-of accessing the program and ultimately use of the program, the location
-and size (e.g., maximum number of bicycles available) of stations
-matters. Indeed, distance to bicycle share stations is associated with
-use ({\textbf{???}}; {\textbf{???}}) and can be a barrier to using PBSPs
-({\textbf{???}}). Kabra et al.~({\textbf{???}}) found that the majority
-of bike share usage in Paris comes from areas within 300m of stations.
-Furthermore, unlike other public amenities with greater ability to adapt
-to crowding, like health care services, the utility of stations is
-limited by the number of bicycles that they can hold, which makes
-crowding effects critical. The program may not necessarily be improved
-if stations are easy to reach but offer only a small number of bicycles.
-Likewise, more people may not opt to use the program if the supply of
-bicycles available at the nearest station is insufficient to meet
-demand. The location and size of stations is important to increase the
-utility of this public transportation option for more people, thus
-achieving vertical equity. Accessibility analyses for PBSPs constitute a
-positive and evaluation-based approach that also has the potential to
-inform equity efforts. For instance, Wang and Lindsey ({\textbf{???}})
-investigated whether new or relocated bicycle share stations increased
-accessibility and use, which offered important insights to improve the
-performance of the program.
+individual must travel to reach a bicycle share station (Kabra et al.,
+2020; J. Wang and Lindsey, 2019b). Since the time or distance that
+members of the population need to reach a bicycle share station
+decreases the potential of accessing the program and ultimately use of
+the program, the location and size (e.g., maximum number of bicycles
+available) of stations matters. Indeed, distance to bicycle share
+stations is associated with use (Fuller et al., 2011; J. Wang and
+Lindsey, 2019b) and can be a barrier to using PBSPs (Fishman et al.,
+2014). Kabra et al.~(2020) found that the majority of bike share usage
+in Paris comes from areas within 300m of stations. Furthermore, unlike
+other public amenities with greater ability to adapt to crowding, like
+health care services, the utility of stations is limited by the number
+of bicycles that they can hold, which makes crowding effects critical.
+The program may not necessarily be improved if stations are easy to
+reach but offer only a small number of bicycles. Likewise, more people
+may not opt to use the program if the supply of bicycles available at
+the nearest station is insufficient to meet demand. The location and
+size of stations is important to increase the utility of this public
+transportation option for more people, thus achieving vertical equity.
+Accessibility analyses for PBSPs constitute a positive and
+evaluation-based approach that also has the potential to inform equity
+efforts. For instance, Wang and Lindsey (2019b) investigated whether new
+or relocated bicycle share stations increased accessibility and use,
+which offered important insights to improve the performance of the
+program.
 
 Several approaches have been commonly used for measuring accessibility:
-cumulative opportunities, gravity, and utility-based ({\textbf{???}}).
-Paez et al ({\textbf{???}}) provide a recent overview of various
+cumulative opportunities, gravity, and utility-based (Handy and
+Niemeier, 1997). Paez et al (2012) provide a recent overview of various
 formulations and applications of accessibility in transportation
@@
 opportunities, for example the quantity of bicycle share stations, as
-measured by a function of time, using a negative exponential form
-({\textbf{???}}). It takes both demand and congestion into account
-({\textbf{???}}). Floating catchment area (FCA) methods are a type of
+measured by a function of time, using a negative exponential form (Handy
+and Niemeier, 1997). It takes both demand and congestion into account
+(Páez et al., 2012). Floating catchment area (FCA) methods are a type of
 gravity measure that have been used in healthcare accessibility research
@@
 involves producing flexible catchment areas for populations, recognizing
-that people may be willing to travel to access particular services
-({\textbf{???}}). Thus, this model does a good job of considering
-potential crowding of services using a binary distance function
-({\textbf{???}}). A recent improvement to this approach was achieved
-through a simple and intuitive adjustment to the allocation of supply
-and demand, which addressed the effects of demand and service inflation
-from the conventional approach (see {\textbf{???}}).
+that people may be willing to travel to access particular services (Páez
+et al., 2012). Thus, this model does a good job of considering potential
+crowding of services using a binary distance function (Páez et al.,
+2012). A recent improvement to this approach was achieved through a
+simple and intuitive adjustment to the allocation of supply and demand,
+which addressed the effects of demand and service inflation from the
+conventional approach (see Paez et al., 2019).
 
@@
 bike share equity has typically used a macro-level approach with
-aggregate data from entire neighbourhoods or census tracts
-({\textbf{???}}; {\textbf{???}}; {\textbf{???}}; {\textbf{???}}),
-although there are recent exceptions ({\textbf{???}}; {\textbf{???}}).
-This is also true for studies examining correlates of bike share demand
-({\textbf{???}}). Handy and Niemeier ({\textbf{???}}) note that using
-disaggregated data in accessibility analyses provides a more accurate
-estimate for individuals, which is useful for addressing vertical
-inequities in PBSP usage.
+aggregate data from entire neighbourhoods or census tracts (Babagoli et
+al., 2019; Mooney et al., 2019; Qian and Jaller, 2020; J. Wang and
+Lindsey, 2019a), although there are recent exceptions (Chen et al.,
+2019; Chen and Li, 2021). This is also true for studies examining
+correlates of bike share demand (J. Wang and Lindsey, 2019b). Handy and
+Niemeier (1997) note that using disaggregated data in accessibility
+analyses provides a more accurate estimate for individuals, which is
+useful for addressing vertical inequities in PBSP usage.
 
@@
 case study. The city launched a public bicycle share program, Hamilton
-Bike Share, in March 2015 with 115 stations and 750 bicycles
-({\textbf{???}}). Before June 2020, the program was known as SoBi
-Hamilton. Stations are spaced between 300 and 600 metres apart
-({\textbf{???}}). The core service area spans 35 km\^{}2 of the city and
-roughly 138,000 can reach a bike share hub within 30 minutes of walking
-{[}see Figure 1{]}. This represents roughly one fifth of the population
-in Hamilton, according to the 2016 Canadian Census. The program was
-enthusiastically welcomed in the city - within three weeks of launching,
-10,000 trips had been made ({\textbf{???}}). In 2017, Hamilton Bike
-Share Inc., the non-profit organization that operates the program,
-initiated an equity program, Everyone Rides Initiative (ERI), to remove
-barriers that may prevent individuals from accessing bike share in
-Hamilton. An additional 75 bicycles and 15 stations were added to the
-program, which expanded it to more disadvantaged areas in the core
-service area {[}see Figure 2{]}. The program also offers subsidized
-memberships to individuals who identify as low income. A comparable
-program can be found in Philadelphia (see {\textbf{???}}). As of June
-2020, the bike share program has 900 bikes and 130 stations, and over
-26,000 active memberships ({\textbf{???}}).
+Bike Share, in March 2015 with 115 stations and 750 bicycles (Hamilton,
+2015). Before June 2020, the program was known as SoBi Hamilton.
+Stations are spaced between 300 and 600 metres apart (Scott et al.,
+2021). The core service area spans 35 km\^{}2 of the city and roughly
+138,000 can reach a bike share hub within 30 minutes of walking {[}see
+Figure \ref{fig:hamilton-and-sobi-service-area}{]}. This represents
+roughly one fifth of the population in Hamilton, according to the 2016
+Canadian Census. The program was enthusiastically welcomed in the city -
+within three weeks of launching, 10,000 trips had been made (Hamilton,
+2015). In 2017, Hamilton Bike Share Inc., the non-profit organization
+that operates the program, initiated an equity program, Everyone Rides
+Initiative (ERI), to remove barriers that may prevent individuals from
+accessing bike share in Hamilton. An additional 75 bicycles and 15
+stations were added to the program, which expanded it to more
+disadvantaged areas in the core service area {[}see Figure 2{]}. The
+program also offers subsidized memberships to individuals who identify
+as low income. A comparable program can be found in Philadelphia (see
+Caspi and Noland, 2019). As of June 2020, the bike share program has 900
+bikes and 130 stations {[}see Figure
+\ref{fig:sobi-stations-in-hamilton}{]}, and over 26,000 active
+memberships (Hamilton, 2015).
 
 Hamilton Bike Share has conducted one membership survey to date in 2018
-({\textbf{???}}), and the findings from 420 members are broadly in line
-with the trends that were discussed above (see {\textbf{???}} for a
+(Civicplan, 2017), and the findings from 420 members are broadly in line
+with the trends that were discussed above (see Fishman, 2016 for a
 recent review of the literature). The majority of respondents live
@@
 travelled are longer than the shortest distance route from origin to
-destination ({\textbf{???}}; {\textbf{???}}). Ridership is influenced by
-weather conditions, temporal factors such as university terms,
-employment, and proximity to important destinations like the local
-university, but is not influenced by population and most transportation
-infrastructure variables ({\textbf{???}}). Bike share users prefer to
-deter from the shortest distance route to travel on streets with cycling
-facilities or lower volumes of traffic ({\textbf{???}}) and avoid steep
-slopes and busy roads ({\textbf{???}}). It is important to note that
-these studies used daily ridership or GPS data from trips taken before
-the equity stations were implemented, so these findings are informative
-with respect to the conventional stations.
+destination (Lu et al., 2018; Scott et al., 2021). Ridership is
+influenced by weather conditions, temporal factors such as university
+terms, employment, and proximity to important destinations like the
+local university, but is not influenced by population and most
+transportation infrastructure variables (Scott and Ciuro, 2019). Bike
+share users prefer to deter from the shortest distance route to travel
+on streets with cycling facilities or lower volumes of traffic (Lu et
+al., 2018) and avoid steep slopes and busy roads (Scott et al., 2021).
+It is important to note that these studies used daily ridership or GPS
+data from trips taken before the equity stations were implemented, so
+these findings are informative with respect to the conventional
+stations.
 
-Our analysis builds upon a previous and recent study ({\textbf{???}}),
-which found that disadvantaged areas in Hamilton are better served by
-the city's PBSP compared to other Canadian cities {[}i.e., Toronto,
-Vancouver, Montreal, and Ottawa-Gatineau{]} with PBSPs where advantaged
-areas have greater access. Hosford and Winters ({\textbf{???}})
+Our analysis builds upon a previous and recent study (Hosford and
+Winters, 2018), which found that disadvantaged areas in Hamilton are
+better served by the city's PBSP compared to other Canadian cities
+{[}i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau{]} with PBSPs
+where advantaged areas have greater access. Hosford and Winters (2018)
 acknowledge that ``Hamilton stands out in that the lower income
@@
 of the local university and increasing gentrification in the downtown
-area. In their analysis, Hosford and Winters ({\textbf{???}}) did not
+area. In their analysis, Hosford and Winters (2018) did not
 differentiate between the conventional stations and those added to
@@
 
-\caption{The core service area of Hamilton Bike Share is outlined in blue.
-Dissemination areas in the Hamilton Census Metropolitan Area (CMA) are outlined in grey.}\label{fig:hamilton-and-sobi-service-area}
+\caption{The core service area of Hamilton Bike Share is outlined in black. Hamilton Census Metropolitan Area (CMA) is shown in grey.}\label{fig:hamilton-and-sobi-service-area}
 \end{figure}
@@
 \caption{The spatial distribution of bike share stations in Hamilton, Ontario.
-The service area of Hamilton Bike Share is outlined in blue and the city's downtown core is outlined in light blue.
-Dissemination areas within the Hamilton CMA that touch the bounding box of the core service area are outlined in grey.}\label{fig:sobi-stations-in-hamilton}
+The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in dark green.
+Hamilton CMA is shown in grey.}\label{fig:sobi-stations-in-hamilton}
 \end{figure}
@@
 the number of bicycle racks at a station) by the population who have
-access to the service ({\textbf{???}}). In particular, the Two-Step
+access to the service (Paez et al., 2019). In particular, the Two-Step
 Floating Catchment Area method produces flexible catchment areas for
@@
 account individual behaviour and doesn't assume that people stay within
-the pre-defined boundaries ({\textbf{???}}). This is an important
+the pre-defined boundaries (Paez et al., 2019). This is an important
 property that supports our rationale for applying this method to measure
@@
 result of the overlapping catchment areas produced by the Two-Step
-Floating Catchment Area ({\textbf{???}}). This overlap inflates the
+Floating Catchment Area (Paez et al., 2019). This overlap inflates the
 demand and generates inaccurate or misleading accessibility estimates
-({\textbf{???}}). By adjusting the impedance weights, both supply and
+(Paez et al., 2019). By adjusting the impedance weights, both supply and
 demand are proportionally allocated which eliminates the double-counting
-of population that leads to demand and supply inflation
-({\textbf{???}}). Other benefits of this adjusted method include
-consideration of competition which can occur when catchment areas
-overlap, as well as the preservation of system-wide population and level
-of service. Balanced floating catchment area methods have been used to
-explore accessibility to health care providers ({\textbf{???}}) and
-COVID-19 health care services ({\textbf{???}}), but have not yet been
-used in the cycling literature to explore issues of accessibility.
+of population that leads to demand and supply inflation (Paez et al.,
+2019). Other benefits of this adjusted method include consideration of
+competition which can occur when catchment areas overlap, as well as the
+preservation of system-wide population and level of service. Balanced
+floating catchment area methods have been used to explore accessibility
+to health care providers (Paez et al., 2019) and COVID-19 health care
+services (Pereira et al., 2021), but have not yet been used in the
+cycling literature to explore issues of accessibility.
 
-In their review of accessibility measures, Geurs and van Wee
-({\textbf{???}}) highlight the need for greater inclusion of individual
-spatio-temporal constraints but acknowledge the challenges of acquiring
-and analyzing person-based data. This comes after Kwan's
-({\textbf{???}}) work to show that space-time measures are more capable
-of capturing interpersonal differences, especially the effect of
-space-time constraints on individual behaviour, and are more helpful for
-unraveling gender/ethnic differences. Applying the balanced floating
-catchment area approach to examine accessibility to Hamilton Bike Share
-constitutes a location-based measure with an individual component by
-stratifying according to median household income ({\textbf{???}}), but
+In their review of accessibility measures, Geurs and van Wee (2004)
+highlight the need for greater inclusion of individual spatio-temporal
+constraints but acknowledge the challenges of acquiring and analyzing
+person-based data. This comes after Kwan's (1998) work to show that
+space-time measures are more capable of capturing interpersonal
+differences, especially the effect of space-time constraints on
+individual behaviour, and are more helpful for unraveling gender/ethnic
+differences. Applying the balanced floating catchment area approach to
+examine accessibility to Hamilton Bike Share constitutes a
+location-based measure with an individual component by stratifying
+according to median household income (Geurs and van Wee, 2004), but
 conducting a further sensitivity analysis by adjusting distance
@@
 
-We first plotted the population in all dissemination areas within SoBi
-Hamilton's core service area. Since pycnophylactic interpolation occurs
-at such a micro scale, we had to ensure that population numbers were not
-interpolated or allocated to areas where people do not live in Hamilton
-(i.e., where schools or parks are located). To do so, we retrieved
-shapefiles for various geographic features (see Table 1) from Open
-Hamilton. Figure 3 shows where these spaces are located within SoBi
-Hamilton's core service area. Next, we extracted these features from the
-PBSP core service area and used pycnophylactic interpolation to
-disaggregate and reallocate population from dissemination areas to
-smaller polygons. These micro-level polygons are 50 by 50 metres.
+Following the method first presented by Tobler (1979), we used
+pycnophylactic interpolation to disaggregate population from each
+dissemination area to smaller polygons that are 50 by 50 metres. This
+method involves smoothing out the population from each dissemination
+area while preserving total volume {[}see Figure
+\ref{fig:da-population}{]}. Since pycnophylactic interpolation occurs at
+such a micro scale, we had to ensure that population numbers were not
+reallocated or smoothed out to areas where people do not live in
+Hamilton (i.e., where schools or parks are located). To do so, we
+retrieved shapefiles for various geographic features {[}see Table
+\ref{tab:data-features}{]} from Open Hamilton. Next, we extracted these
+features from the PBSP core service area and used pycnophylactic
+interpolation to disaggregate and reallocate population {[}see Figure
+\ref{fig:interpolated-population}{]}.
 
@@
 
-\caption{\label{tab:data-features}\label{tab:landscape-features}Landscape features}
+\caption{\label{tab:data-features}\label{tab:landscape-features}Landscape features extracted from the Hamilton Bike Share core service area before population is interpolated.}
 \centering
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/da-population-1}
+
+}
+
+\caption{Population in all dissemination areas (outlined in white) in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the service area.}\label{fig:da-population}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/interpolated-population-1}
+
+}
+
+\caption{Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
+The downtown area is outlined in dark green.
+Geographic features have been extracted.
+Dissemination area polygons are outlined in grey.}\label{fig:interpolated-population}
+\end{figure}
+
 \hypertarget{travel-time-matrix}{%
@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-2-1}
-
-}
-
-\caption{Population in all dissemination areas in the SoBi Hamilton core service area and within 30 minutes of walking to a bike share station.}\label{fig:figure-2}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-4-1}
-
-}
-
-\caption{Interpolated population cells in the SoBi Hamilton core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
-The downtown area is outlined in dark green.
-Geographic features have been extracted.
-Dissemination area polygons are outlined in grey.}\label{fig:figure-4}
-\end{figure}
-
 \hypertarget{accessibility-by-distance-thresholds}{%
@@
 was consulted to determine the thresholds for the sensitivity analysis.
-Previous studies have found that living within 250 metres
-({\textbf{???}}) and 300 metres ({\textbf{???}}) is correlated with bike
+Previous studies have found that living within 250 metres (Fuller et
+al., 2011) and 300 metres (Kabra et al., 2020) is correlated with bike
 share use, while other research has found that walking trips are less
-than 600 metres and rarely more than 1200 metres ({\textbf{???}}) or a
-median distance of 650 metres ({\textbf{???}}). As seen in Table
-\ref{tab:accessibility-sensibility}, we find that accessibility
-increases with a threshold between two and four minutes, but is then
-maximized at 5 minutes. At some stations, Hamilton Bike Share will
-depict a map to show the user the locations of the other nearest
-stations within a five minute walk, which suggests that this is an
-average distance that people are anticipated to walk. Accessibility
+than 600 metres and rarely more than 1200 metres (Millward et al., 2013)
+or a median distance of 650 metres (Larsen et al., 2010). We found that
+accessibility increases with a threshold between two and four minutes,
+but is then maximized at 5 minutes. At some stations, Hamilton Bike
+Share will depict a map to show the user the locations of the other
+nearest stations within a five minute walk, which suggests that this is
+an average distance that people are anticipated to walk. Accessibility
 decreases significantly after eight minutes, which is intuitive given
@@
 
-At this threshold, we find that there are 25.2 bicycle racks per person
-in the original system. With the addition of equity stations, there are
-now 25.4 bicycles per person. Figure \ref{fig:figure-6} presents a
-comparison of accessibility between the systems. Accessibility is fairly
-uniform, with the exception of two small areas near the university and
-the waterfront park where accessibility is slightly higher. Equity
-stations are shown in green, which helps to demonstrate how they
-increase accessibility between the systems.
+At the minimum threshold, we find that there are 25.2 bicycle racks per
+person in the original system. With the addition of equity stations,
+there are now 25.4 bicycles per person. Figure \ref{fig:figure-6}
+presents a comparison of accessibility between the systems.
+Accessibility is fairly uniform, with the exception of two small areas
+near the university and the waterfront park where accessibility is
+slightly higher.
 
@@
 
-At this threshold, we find that there are 68.6 bicycle racks per person
-in the original system. With the addition of equity stations, there are
-now 68.8 bicycles per person. Figure \ref{fig:figure-7} presents a
-comparison of accessibility between the systems. Again, accessibility is
-fairly uniform, with the exception of one area near the waterfront park
-where accessibility is higher.
+At the average threshold, we find that there are 68.6 bicycle racks per
+person in the original system. With the addition of equity stations,
+there are now 68.8 bicycles per person. Figure \ref{fig:figure-7}
+presents a comparison of accessibility between the systems. Again,
+accessibility is fairly uniform, with the exception of one very small
+area near the waterfront park where accessibility is higher.
 
@@
 
-At this threshold, we find that there are 3.61 bicycle racks per person
-in the original system. With the addition of equity stations, there are
-now 3.74 bicycles per person. Figure \ref{fig:figure-8} presents a
-comparison of accessibility between the systems. We begin to see
-differences in accessibility across the service area, with users near
-the university and its adjacent neighbourhoods, as well as
-neighbourhoods north of the downtown area have slightly higher
+At the maximum threshold, we find that there are 3.61 bicycle racks per
+person in the original system. With the addition of equity stations,
+there are now 3.74 bicycles per person. Figure \ref{fig:figure-8}
+presents a comparison of accessibility between the systems. We begin to
+see differences in accessibility across the service area, with users
+near the university and its adjacent neighbourhoods, as well as
+neighbourhoods north of the downtown area, have slightly higher
 accessibility. While the differences are modest, they are more apparent
@@
 
-At this threshold, we find that there are 2.44 bicycle racks per person
-in the original system. With the addition of equity stations, there are
-now 2.55 bicycles per person. Figure \ref{fig:figure-9} presents a
-comparison of accessibility between the systems. Users near the
-university and the neighbourhoods north of the downtown area (outlined
-in green) have the highest accessibility, followed by those who live in
-the city's downtown area. Accessibility in the east end of the core
-service area remains low.
+At the extreme threshold, we find that there are 2.44 bicycle racks per
+person in the original system. With the addition of equity stations,
+there are now 2.55 bicycles per person. Figure \ref{fig:figure-9}
+presents a comparison of accessibility between the systems. Users near
+the university and the neighbourhoods north of the downtown area
+(outlined in green) have the highest accessibility, followed by those
+who live in the city's downtown area. Accessibility in the east end of
+the core service area remains low.
 
@@
 Approximately 138,000 people live within Hamilton Bike Share's core
-service area or within a 10 minute walk to the service area. We find
+service area or within a 30 minute walk to the service area. We find
 that over 118,000 people can access a bicycle share station within a 15
 minute walk, which represents roughly 85\% of the total population in
-Hamilton Bike Share's core service area {[}see Table
-\ref{tab:table-2}{]}. While previous research found that neighbourhoods
-with more disadvantage are better serviced by Hamilton Bike Share
-({\textbf{???}}), the authors used the Pampalon Deprivation Index from
-2011 as a measure of socioeconomic status. Instead, we use median total
-income for each dissemination area to examine whether disparities in
-accessibility exist within the core service area. We divide income by
-quantiles: bottom 20\%, second 20\%, third 20\%, fourth 20\%, and top
-20\%. One of the unique properties of the balanced floating catchment
-area method is that data can be reaggregated from small to larger
-polygons while preserving the total population and number of bicycle
-racks at each station. This avoids demand and supply inflation, and also
-enables us to present findings in a way that is easier to interpret
-(i.e., at the dissemination area level).
+the core service area {[}see Table \ref{tab:accessibility-income}{]}.
+While previous research found that neighbourhoods with more disadvantage
+are better serviced by Hamilton Bike Share (Hosford and Winters, 2018),
+the authors used the Pampalon Deprivation Index from 2011 as a measure
+of socioeconomic status for dissemination areas across the city not just
+within the core service area. Instead, we use median total income for
+each dissemination area in the core service area to examine whether
+disparities in accessibility exist between income groups for those who
+live in or near the core service area. We divide income by quantiles:
+bottom 20\%, second 20\%, third 20\%, fourth 20\%, and top 20\%. One of
+the unique properties of the balanced floating catchment area method is
+that data can be reaggregated from small to larger polygons while
+preserving the total population and supply at each station. This avoids
+demand and supply inflation, and also enables us to present findings in
+a way that is easier to interpret (i.e., across dissemination areas
+instead of 50 by 50 metre population cells).
 
-Panels 1 and 2 depict bivariate choropleth maps with the combined
-spatial distribution of accessibility and median total income per
-household for the different thresholds. Our analysis demonstrates that
-the stations added to Hamilton's public bicycle share program to
-increase equity for disadvantaged neighbourhoods achieved their goal by
-increasing accessibility, albeit only modestly. By implementing equity
-stations in more areas with lower median total income in Hamilton, the
-PBSP has achieved greater horizontal equity by extending the spatial
-distribution of bicycles across the city. This is particularly evident
-at shorter walking distances of three and five minutes.
+Panels 1{[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
+{[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
+{[}\ref{fig:figure-bi-map-threshold-10}{]}, and 4
+{[}\ref{fig:figure-bi-map-threshold-15}{]} depict bivariate choropleth
+maps with the combined spatial distribution of accessibility and median
+total income per household for the different thresholds. Our analysis
+demonstrates that the extreme threshold of fifteen minutes allows the
+most people to access a station. We find that stations added to
+Hamilton's public bicycle share program to increase equity for
+disadvantaged neighbourhoods achieved their goal by increasing
+accessibility, albeit only modestly {[}see Table
+\ref{tab:accessibility-income}{]}. By implementing equity stations in
+more areas with lower median total income in Hamilton, the PBSP has
+achieved greater horizontal equity by extending the spatial distribution
+of bicycles across the city. This is particularly evident at the minimum
+and average thresholds of three and five minutes, respectively. The
+largest gains were made for dissemination areas in the second 20\%,
+where an additional 3,073 and 5,395 people could reach a bicycle share
+station within three and five minutes walk, respectively, after the
+addition of the equity station.
 
-However, we found that dissemination areas with lower total household
-income do not necessarily have greater access to the program {[}see
+However, we found that dissemination areas with the lowest total
+household income do not have much greater access to the program {[}see
 Table \ref{tab:table-3}{]}. Income disparities still persist, however
@@
 the top 20\% of income have the highest level of accessibility at a
-threshold of ten and fifteen minutes. At distance thresholds of three
-and five minutes, people in the second 20\% have the highest level of
-accessibility by a significant amount. This is due to the concentration
-of stations near or within the city's downtown area and the ``lower
-city'' which historically has been more disadvantaged compared to the
-surrounding suburban neighbourhoods. This is an artifact of the city's
-steel industry which led to more blue collar and working class people
-living in the city's downtown area and ``lower city''. However, the
-bottom 20\%, who may benefit the most from Hamilton Bike Share's equity
-initiatives, have the lowest accessibility at three minutes threshold
-and the second lowest accessibility at all other thresholds.
+threshold of ten and fifteen minutes. Although dissemination areas in
+the second 20\% have the highest level of accessibility by a significant
+amount at lower thresholds, the bottom 20\%, who may benefit the most
+from Hamilton Bike Share's equity initiatives, have the lowest
+accessibility at three minutes threshold and the second lowest
+accessibility at all other thresholds. Panels
+1{[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
+{[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
+{[}\ref{fig:figure-bi-map-threshold-10}{]}, and 4
+{[}\ref{fig:figure-bi-map-threshold-15}{]} reveal specific areas where
+accessibility is low or zero, which are good candidates for additional
+equity stations.
 
@@
 share of the level of service are some of the most socioeconomically
-disadvantaged areas in the city. This finding aligns with other studies
-that have found disparities in station or bike location along income
-lines in New York City ({\textbf{???}}), Tampa ({\textbf{???}}), and
-Seattle ({\textbf{???}}).
+disadvantaged areas in the city. While the addition of equity stations
+increases accessibility for all income groups at all thresholds, they
+did not increase accessibility significantly for any single income
+group. This finding aligns with other studies that have found
+disparities in station or bike location along income lines in New York
+City (Babagoli et al., 2019), Tampa (Chen et al., 2019), and Seattle
+(Mooney et al., 2019), where people who have greater advantage have
+better access to PBSPs.
 
@@
 
-\caption{\label{tab:accessibility-income}\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile}
+\caption{\label{tab:accessibility-income}\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile and between systems (with and without equity stations).}
 \centering
@@
 \hspace{1em}Fourth 20\% & 4.119 & 1673 & 4.119 & 1673 & 0.000 & 0\\
-\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{3.757} & \cellcolor{gray!6}{2151} & \cellcolor{gray!6}{3.784} & \cellcolor{gray!6}{2416} & \cellcolor{gray!6}{0.027} & \cellcolor{gray!6}{265}\\
+\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{3.408} & \cellcolor{gray!6}{2114} & \cellcolor{gray!6}{3.434} & \cellcolor{gray!6}{2379} & \cellcolor{gray!6}{0.026} & \cellcolor{gray!6}{265}\\
 \addlinespace[0.3em]
@@
 \hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{56.048} & \cellcolor{gray!6}{17513} & \cellcolor{gray!6}{56.137} & \cellcolor{gray!6}{22908} & \cellcolor{gray!6}{0.089} & \cellcolor{gray!6}{5395}\\
-\hspace{1em}Third 20\% & 4.259 & 15117 & 4.291 & 18309 & 0.032 & 3192\\
+\hspace{1em}Third 20\% & 4.258 & 15117 & 4.291 & 18309 & 0.033 & 3192\\
 \hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{1.094} & \cellcolor{gray!6}{2867} & \cellcolor{gray!6}{1.095} & \cellcolor{gray!6}{3116} & \cellcolor{gray!6}{0.001} & \cellcolor{gray!6}{249}\\
-\hspace{1em}Top 20\% & 6.256 & 4074 & 6.264 & 4540 & 0.008 & 466\\
+\hspace{1em}Top 20\% & 5.925 & 4052 & 5.933 & 4518 & 0.008 & 466\\
 \addlinespace[0.3em]
 \multicolumn{7}{l}{\textbf{Threshold - 10 minutes}}\\
-\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{0.604} & \cellcolor{gray!6}{41824} & \cellcolor{gray!6}{0.622} & \cellcolor{gray!6}{41981} & \cellcolor{gray!6}{0.018} & \cellcolor{gray!6}{157}\\
-\hspace{1em}Second 20\% & 0.862 & 27546 & 0.929 & 30503 & 0.067 & 2957\\
-\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{0.776} & \cellcolor{gray!6}{22394} & \cellcolor{gray!6}{0.802} & \cellcolor{gray!6}{25128} & \cellcolor{gray!6}{0.026} & \cellcolor{gray!6}{2734}\\
+\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{0.603} & \cellcolor{gray!6}{41824} & \cellcolor{gray!6}{0.621} & \cellcolor{gray!6}{41981} & \cellcolor{gray!6}{0.018} & \cellcolor{gray!6}{157}\\
+\hspace{1em}Second 20\% & 0.860 & 27546 & 0.927 & 30503 & 0.067 & 2957\\
+\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{0.772} & \cellcolor{gray!6}{22394} & \cellcolor{gray!6}{0.799} & \cellcolor{gray!6}{25128} & \cellcolor{gray!6}{0.027} & \cellcolor{gray!6}{2734}\\
 \hspace{1em}Fourth 20\% & 0.225 & 4544 & 0.227 & 4989 & 0.002 & 445\\
-\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{1.346} & \cellcolor{gray!6}{7989} & \cellcolor{gray!6}{1.348} & \cellcolor{gray!6}{9078} & \cellcolor{gray!6}{0.002} & \cellcolor{gray!6}{1089}\\
+\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{1.160} & \cellcolor{gray!6}{7989} & \cellcolor{gray!6}{1.162} & \cellcolor{gray!6}{9078} & \cellcolor{gray!6}{0.002} & \cellcolor{gray!6}{1089}\\
 \addlinespace[0.3em]
 \multicolumn{7}{l}{\textbf{Threshold - 15 minutes}}\\
-\hspace{1em}Bottom 20\% & 0.536 & 42208 & 0.557 & 42327 & 0.021 & 119\\
-\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{0.555} & \cellcolor{gray!6}{30507} & \cellcolor{gray!6}{0.614} & \cellcolor{gray!6}{31069} & \cellcolor{gray!6}{0.059} & \cellcolor{gray!6}{562}\\
-\hspace{1em}Third 20\% & 0.554 & 26108 & 0.581 & 26660 & 0.027 & 552\\
+\hspace{1em}Bottom 20\% & 0.534 & 42208 & 0.554 & 42327 & 0.020 & 119\\
+\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{0.551} & \cellcolor{gray!6}{30507} & \cellcolor{gray!6}{0.611} & \cellcolor{gray!6}{31069} & \cellcolor{gray!6}{0.060} & \cellcolor{gray!6}{562}\\
+\hspace{1em}Third 20\% & 0.546 & 26108 & 0.573 & 26660 & 0.027 & 552\\
 \hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{0.093} & \cellcolor{gray!6}{6312} & \cellcolor{gray!6}{0.096} & \cellcolor{gray!6}{7435} & \cellcolor{gray!6}{0.003} & \cellcolor{gray!6}{1123}\\
-\hspace{1em}Top 20\% & 0.808 & 10209 & 0.811 & 11089 & 0.003 & 880\\
+\hspace{1em}Top 20\% & 0.712 & 10209 & 0.714 & 11089 & 0.002 & 880\\
 \bottomrule
@@
 conventional and equity stations. Therefore, further research is needed
-to determine whether this encouraged more cycling for low-income
-individuals living near equity stations. Other studies have specifically
-looked at differences in trip type, frequency, or length among users
-from disadvantaged neighbourhoods ({\textbf{???}}; {\textbf{???}};
-{\textbf{???}}), but our analysis is limited by the lack of accessible
-data to conduct similar analyses for Hamilton Bike Share.
+to determine whether the addition of equity stations encouraged more
+cycling for low-income individuals living near them. Other studies have
+specifically looked at differences in trip type, frequency, or length
+among users from disadvantaged neighbourhoods (Caspi and Noland, 2019;
+Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), but our analysis is
+limited by the lack of accessible route or user data to conduct similar
+analyses for Hamilton Bike Share.
 
@@
 accessibility and reducing both horizontal and vertical inequities. In
-particular, accessibility improved the most for the second 20\% income
-groups at all thresholds, but the gains were only modest. Congestion
-effects were observed at higher thresholds, with accessibility
-decreasing significantly once the catchment area is increased to 10
-minutes walking. However, We identified specific areas that have both
-low accessibility and low income which would benefit from an increased
-supply of bicycles. These are ideal candidates for new equity stations.
-In this way, our paper has made contributions in a positive way by
-applying an intuitive and useful approach to measure accessibility to a
-PBSP, and in a normative way by serving to inform future investments in
-cycle infrastructure for Hamilton Bike Share.
+particular, accessibility improved the most for the second 20\% median
+total income groups at all thresholds, but the gains were only modest
+for all income groups. Dissemination areas with the bottom 20\% had the
+lowest accessibility at three minutes, and second lowest level of
+accessibility at five, ten, and fifteen minutes. Congestion effects were
+observed at higher thresholds, with accessibility decreasing
+significantly once the catchment area is increased to 10 minutes
+walking. Based on our analysis, we identified specific areas that have
+both low accessibility and low income which would benefit from an
+increased supply of bicycles. These are ideal candidates for new equity
+stations. In this way, our paper has made contributions in a positive
+way by applying an intuitive and useful approach to measure
+accessibility to a PBSP, and in a normative way by serving to inform
+future investments in cycle infrastructure for Hamilton Bike Share.
 
-Wang and Lindsey ({\textbf{???}}) have noted that there is a lack of
-research that examines how bike share users' behaviour changes as a
-result of program changes to station locations or improvements in
-accessibility. As such, a logical next step to this research is to
-examine whether Hamilton Bike Share's equity stations increased
-ridership or resulted in new memberships in areas that were previously
-under-served. An examination of the types of trips undertaken by
-residents in these areas would also be informative, such as the study
-undertaken by Caspi and Norland ({\textbf{???}}) after bike share
-stations were implemented in low-income Philadelphia neighbourhoods. The
-bulk of cycling facilities that have been built in Hamilton to date
-located in the core service area near the conventional stations. It
-would be worthwhile to explore the route choice of bike share trips
-departing or ending at the equity stations and to identify factors that
-specifically influence trips from these stations, which would extend
-existing studies conducted by Scott and colleagues ({\textbf{???}};
-{\textbf{???}}; {\textbf{???}}). This paper, combined with additional
-studies such as those conceptualized above, would serve as a valuable
-case study for Hamilton and other cities with PBSPs who wish to evaluate
-and address spatial inequities in accessibility and transportation
-options in urban areas.
+Wang and Lindsey (2019b) have noted that there is a lack of research
+that examines how bike share users' behaviour changes as a result of
+program changes to station locations or improvements in accessibility.
+As such, a logical next step to this research is to examine whether
+Hamilton Bike Share's equity stations increased ridership or resulted in
+new memberships in areas that were previously under-served. An
+examination of the types of trips undertaken by residents in these areas
+would also be informative, such as the study undertaken by Caspi and
+Norland (2019) after bike share stations were implemented in low-income
+Philadelphia neighbourhoods. The bulk of cycling facilities that have
+been built in Hamilton to date located in the core service area near the
+conventional stations. It would be worthwhile to explore the route
+choice of bike share trips departing or ending at the equity stations
+and to identify factors that specifically influence trips from these
+stations, which would extend existing studies conducted by Scott and
+colleagues (Lu et al., 2018; Scott and Ciuro, 2019; Scott et al., 2021).
+This paper, combined with additional studies such as those
+conceptualized above, would serve as a valuable case study for Hamilton
+and other cities with PBSPs who wish to evaluate and address spatial
+inequities in accessibility and transportation options in urban areas.
 
@@
 
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-auchinclossDesignBaselineDescription2020}{}%
+Auchincloss, A.H., Michael, Y.L., Fuller, D., Li, S., Niamatullah, S.,
+Fillmore, C.E., Setubal, C., Bettigole, C., 2020. Design and baseline
+description of a cohort of bikeshare users in the city of Philadelphia.
+Journal of Transport \& Health 16, 100836.
+doi:\href{https://doi.org/10.1016/j.jth.2020.100836}{10.1016/j.jth.2020.100836}
+
+\leavevmode\hypertarget{ref-babagoliExploringHealthSpatial2019}{}%
+Babagoli, M.A., Kaufman, T.K., Noyes, P., Sheffield, P.E., 2019.
+Exploring the health and spatial equity implications of the New York
+City Bike share system. Journal of Transport \& Health 13, 200--209.
+doi:\href{https://doi.org/10.1016/j.jth.2019.04.003}{10.1016/j.jth.2019.04.003}
+
+\leavevmode\hypertarget{ref-breyWantRideMy2017}{}%
+Brey, R., Castillo-Manzano, J.I., Castro-Nuño, M., 2017. ``I want to
+ride my bicycle'': Delimiting cyclist typologies. Applied Economics
+Letters 24, 549--552.
+doi:\href{https://doi.org/10.1080/13504851.2016.1210760}{10.1080/13504851.2016.1210760}
+
+\leavevmode\hypertarget{ref-buckAreBikeshareUsers2013}{}%
+Buck, D., Buehler, R., Happ, P., Rawls, B., Chung, P., Borecki, N.,
+2013. Are Bikeshare Users Different from Regular Cyclists?: A First Look
+at Short-Term Users, Annual Members, and Area Cyclists in the
+Washington, D.C., Region. Transportation Research Record 2387, 112--119.
+doi:\href{https://doi.org/10.3141/2387-13}{10.3141/2387-13}
+
+\leavevmode\hypertarget{ref-caspiBikesharingPhiladelphiaLowerincome2019}{}%
+Caspi, O., Noland, R.B., 2019. Bikesharing in Philadelphia: Do
+lower-income areas generate trips? Travel Behaviour and Society 16,
+143--152.
+doi:\href{https://doi.org/10.1016/j.tbs.2019.05.004}{10.1016/j.tbs.2019.05.004}
+
+\leavevmode\hypertarget{ref-chenExploringEquityPerformance2019}{}%
+Chen, Z., Guo, Y., Stuart, A.L., Zhang, Y., Li, X., 2019. Exploring the
+equity performance of bike-sharing systems with disaggregated data: A
+story of southern Tampa. Transportation Research Part A: Policy and
+Practice 130, 529--545.
+doi:\href{https://doi.org/10.1016/j.tra.2019.09.048}{10.1016/j.tra.2019.09.048}
+
+\leavevmode\hypertarget{ref-chenUnobservedHeterogeneityTransportation2021}{}%
+Chen, Z., Li, X., 2021. Unobserved heterogeneity in transportation
+equity analysis: Evidence from a bike-sharing system in southern Tampa.
+Journal of Transport Geography 91, 102956.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2021.102956}{10.1016/j.jtrangeo.2021.102956}
+
+\leavevmode\hypertarget{ref-civicplanSoBiHamiltonMembership2017}{}%
+Civicplan, 2017. SoBi Hamilton Membership Survey. Civicplan \textbar{}
+Planning Engagement Research.
+
+\leavevmode\hypertarget{ref-fishmanBikeshareReviewRecent2016}{}%
+Fishman, E., 2016. Bikeshare: A Review of Recent Literature. Transport
+Reviews 36, 92--113.
+doi:\href{https://doi.org/10.1080/01441647.2015.1033036}{10.1080/01441647.2015.1033036}
+
+\leavevmode\hypertarget{ref-fishmanBarriersBikesharingAnalysis2014}{}%
+Fishman, E., Washington, S., Haworth, N., Mazzei, A., 2014. Barriers to
+bikesharing: An analysis from Melbourne and Brisbane. Journal of
+Transport Geography 41, 325--337.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2014.08.005}{10.1016/j.jtrangeo.2014.08.005}
+
+\leavevmode\hypertarget{ref-fullerUseNewPublic2011}{}%
+Fuller, D., Gauvin, L., Kestens, Y., Daniel, M., Fournier, M., Morency,
+P., Drouin, L., 2011. Use of a New Public Bicycle Share Program in
+Montreal, Canada. American Journal of Preventive Medicine 41, 80--83.
+doi:\href{https://doi.org/10.1016/j.amepre.2011.03.002}{10.1016/j.amepre.2011.03.002}
+
+\leavevmode\hypertarget{ref-geursAccessibilityEvaluationLanduse2004}{}%
+Geurs, K.T., van Wee, B., 2004. Accessibility evaluation of land-use and
+transport strategies: Review and research directions. Journal of
+Transport Geography 12, 127--140.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2003.10.005}{10.1016/j.jtrangeo.2003.10.005}
+
+\leavevmode\hypertarget{ref-hamiltonHamiltonBikeShare2015}{}%
+Hamilton, C. of, 2015. Hamilton Bike Share.
+
+\leavevmode\hypertarget{ref-handyMeasuringAccessibilityExploration1997}{}%
+Handy, S.L., Niemeier, D.A., 1997. Measuring Accessibility: An
+Exploration of Issues and Alternatives. Environment and Planning A:
+Economy and Space 29, 1175--1194.
+doi:\href{https://doi.org/10.1068/a291175}{10.1068/a291175}
+
+\leavevmode\hypertarget{ref-hosfordEvaluationImpactPublic2018}{}%
+Hosford, K., Fuller, D., Lear, S.A., Teschke, K., Gauvin, L., Brauer,
+M., Winters, M., 2018. Evaluation of the impact of a public bicycle
+share program on population bicycling in Vancouver, BC. Preventive
+Medicine Reports 12, 176--181.
+doi:\href{https://doi.org/10.1016/j.pmedr.2018.09.014}{10.1016/j.pmedr.2018.09.014}
+
+\leavevmode\hypertarget{ref-hosfordWhoArePublic2018}{}%
+Hosford, K., Winters, M., 2018. Who Are Public Bicycle Share Programs
+Serving? An Evaluation of the Equity of Spatial Access to Bicycle Share
+Service Areas in Canadian Cities. Transportation Research Record 2672,
+42--50.
+doi:\href{https://doi.org/10.1177/0361198118783107}{10.1177/0361198118783107}
+
+\leavevmode\hypertarget{ref-hosfordEvaluatingImpactImplementing2019}{}%
+Hosford, K., Winters, M., Gauvin, L., Camden, A., Dubé, A.-S., Friedman,
+S.M., Fuller, D., 2019. Evaluating the impact of implementing public
+bicycle share programs on cycling: The International Bikeshare Impacts
+on Cycling and Collisions Study (IBICCS). International Journal of
+Behavioral Nutrition and Physical Activity 16, 107.
+doi:\href{https://doi.org/10.1186/s12966-019-0871-9}{10.1186/s12966-019-0871-9}
+
+\leavevmode\hypertarget{ref-hullgrassoBikeShareEquity2020}{}%
+Hull Grasso, S., Barnes, P., Chavis, C., 2020. Bike Share Equity for
+Underrepresented Groups: Analyzing Barriers to System Usage in
+Baltimore, Maryland. Sustainability 12, 7600.
+doi:\href{https://doi.org/10.3390/su12187600}{10.3390/su12187600}
+
+\leavevmode\hypertarget{ref-kabraBikeShareSystemsAccessibility2020}{}%
+Kabra, A., Belavina, E., Girotra, K., 2020. Bike-Share Systems:
+Accessibility and Availability. Management Science 66, 3803--3824.
+doi:\href{https://doi.org/10.1287/mnsc.2019.3407}{10.1287/mnsc.2019.3407}
+
+\leavevmode\hypertarget{ref-kwanSpaceTimeIntegral1998}{}%
+Kwan, M.-P., 1998. Space-Time and Integral Measures of Individual
+Accessibility: A Comparative Analysis Using a Point-based Framework.
+Geographical Analysis 30, 191--216.
+doi:\href{https://doi.org/10.1111/j.1538-4632.1998.tb00396.x}{10.1111/j.1538-4632.1998.tb00396.x}
+
+\leavevmode\hypertarget{ref-larsenQuarterMileReexamining2010}{}%
+Larsen, J., El-Geneidy, A., Yasmin, F., 2010. Beyond the Quarter Mile:
+Re-examining Travel Distances by Active Transportation. Canadian Journal
+of Urban Research 19, 70--88.
+
+\leavevmode\hypertarget{ref-levineCenturyEvolutionAccessibility2020}{}%
+Levine, J., 2020. A century of evolution of the accessibility concept.
+Transportation Research Part D: Transport and Environment 83, 102309.
+doi:\href{https://doi.org/10.1016/j.trd.2020.102309}{10.1016/j.trd.2020.102309}
+
+\leavevmode\hypertarget{ref-luUnderstandingBikeShare2018}{}%
+Lu, W., Scott, D.M., Dalumpines, R., 2018. Understanding bike share
+cyclist route choice using GPS data: Comparing dominant routes and
+shortest paths. Journal of Transport Geography 71, 172--181.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2018.07.012}{10.1016/j.jtrangeo.2018.07.012}
+
+\leavevmode\hypertarget{ref-macarthurAdaptiveBikeShare2020}{}%
+MacArthur, J., McNeil, N., Cummings, A., Broach, J., 2020. Adaptive Bike
+Share: Expanding Bike Share to People with Disabilities and Older
+Adults. Transportation Research Record 2674, 556--565.
+doi:\href{https://doi.org/10.1177/0361198120925079}{10.1177/0361198120925079}
+
+\leavevmode\hypertarget{ref-millwardActivetransportWalkingBehavior2013}{}%
+Millward, H., Spinney, J., Scott, D., 2013. Active-transport walking
+behavior: Destinations, durations, distances. Journal of Transport
+Geography 28, 101--110.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2012.11.012}{10.1016/j.jtrangeo.2012.11.012}
+
+\leavevmode\hypertarget{ref-mooneyFreedomStationSpatial2019}{}%
+Mooney, S.J., Hosford, K., Howe, B., Yan, A., Winters, M., Bassok, A.,
+Hirsch, J.A., 2019. Freedom from the station: Spatial equity in access
+to dockless bike share. Journal of Transport Geography 74, 91--96.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2018.11.009}{10.1016/j.jtrangeo.2018.11.009}
+
+\leavevmode\hypertarget{ref-nickkarSpatialtemporalGenderLand2019}{}%
+Nickkar, A., Banerjee, S., Chavis, C., Bhuyan, I.A., Barnes, P., 2019. A
+spatial-temporal gender and land use analysis of bikeshare ridership:
+The case study of Baltimore City. City, Culture and Society 18, 100291.
+doi:\href{https://doi.org/10.1016/j.ccs.2019.100291}{10.1016/j.ccs.2019.100291}
+
+\leavevmode\hypertarget{ref-paezDemandLevelService2019}{}%
+Paez, A., Higgins, C.D., Vivona, S.F., 2019. Demand and level of service
+inflation in Floating Catchment Area (FCA) methods. PLoS ONE 14.
+doi:\href{https://doi.org/10.1371/journal.pone.0218773}{10.1371/journal.pone.0218773}
+
+\leavevmode\hypertarget{ref-paezMeasuringAccessibilityPositive2012}{}%
+Páez, A., Scott, D.M., Morency, C., 2012. Measuring accessibility:
+Positive and normative implementations of various accessibility
+indicators. Journal of Transport Geography, Special Section on
+Accessibility and Socio-Economic Activities: Methodological and
+Empirical Aspects 25, 141--153.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2012.03.016}{10.1016/j.jtrangeo.2012.03.016}
+
+\leavevmode\hypertarget{ref-pereiraGeographicAccessCOVID192021}{}%
+Pereira, R.H.M., Braga, C.K.V., Servo, L.M., Serra, B., Amaral, P.,
+Gouveia, N., Paez, A., 2021. Geographic access to COVID-19 healthcare in
+Brazil using a balanced float catchment area approach. Social Science \&
+Medicine 273, 113773.
+doi:\href{https://doi.org/10.1016/j.socscimed.2021.113773}{10.1016/j.socscimed.2021.113773}
+
+\leavevmode\hypertarget{ref-qianBikesharingEquityDisadvantaged2020}{}%
+Qian, X., Jaller, M., 2020. Bikesharing, equity, and disadvantaged
+communities: A case study in Chicago. Transportation Research Part A:
+Policy and Practice 140, 354--371.
+doi:\href{https://doi.org/10.1016/j.tra.2020.07.004}{10.1016/j.tra.2020.07.004}
+
+\leavevmode\hypertarget{ref-reillyNoncyclistsFrequentCyclists2020}{}%
+Reilly, K.H., Noyes, P., Crossa, A., 2020. From non-cyclists to frequent
+cyclists: Factors associated with frequent bike share use in New York
+City. Journal of Transport \& Health 16, 100790.
+doi:\href{https://doi.org/10.1016/j.jth.2019.100790}{10.1016/j.jth.2019.100790}
+
+\leavevmode\hypertarget{ref-reillyGenderDisparitiesNew2020}{}%
+Reilly, K.H., Wang, S.M., Crossa, A., 2020. Gender disparities in New
+York City bike share usage. International Journal of Sustainable
+Transportation 0, 1--9.
+doi:\href{https://doi.org/10.1080/15568318.2020.1861393}{10.1080/15568318.2020.1861393}
+
+\leavevmode\hypertarget{ref-scottWhatFactorsInfluence2019}{}%
+Scott, D.M., Ciuro, C., 2019. What factors influence bike share
+ridership? An investigation of Hamilton, Ontario's bike share hubs.
+Travel Behaviour and Society 16, 50--58.
+doi:\href{https://doi.org/10.1016/j.tbs.2019.04.003}{10.1016/j.tbs.2019.04.003}
+
+\leavevmode\hypertarget{ref-scottRouteChoiceBike2021}{}%
+Scott, D.M., Lu, W., Brown, M.J., 2021. Route choice of bike share
+users: Leveraging GPS data to derive choice sets. Journal of Transport
+Geography 90, 102903.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2020.102903}{10.1016/j.jtrangeo.2020.102903}
+
+\leavevmode\hypertarget{ref-toblerSmoothPycnophylacticInterpolation1979}{}%
+Tobler, W.R., 1979. Smooth Pycnophylactic Interpolation for Geographical
+Regions. Journal of the American Statistical Association 74, 519--530.
+doi:\href{https://doi.org/10.1080/01621459.1979.10481647}{10.1080/01621459.1979.10481647}
+
+\leavevmode\hypertarget{ref-wangNeighborhoodSociodemographicCharacteristics2019}{}%
+Wang, J., Lindsey, G., 2019a. Neighborhood socio-demographic
+characteristics and bike share member patterns of use. Journal of
+Transport Geography 79, 102475.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2019.102475}{10.1016/j.jtrangeo.2019.102475}
+
+\leavevmode\hypertarget{ref-wangNewBikeShare2019}{}%
+Wang, J., Lindsey, G., 2019b. Do new bike share stations increase member
+use: A quasi-experimental study. Transportation Research Part A: Policy
+and Practice 121, 1--11.
+doi:\href{https://doi.org/10.1016/j.tra.2019.01.004}{10.1016/j.tra.2019.01.004}
+
+\leavevmode\hypertarget{ref-wintersWhoAreSuperusers2019}{}%
+Winters, M., Hosford, K., Javaheri, S., 2019. Who are the
+``super-users'' of public bike share? An analysis of public bike share
+members in Vancouver, BC. Preventive Medicine Reports 15, 100946.
+doi:\href{https://doi.org/10.1016/j.pmedr.2019.100946}{10.1016/j.pmedr.2019.100946}
+
 \end{document}

### 2021-03-25 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
CDH edits
text and code edits

@@
 Financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016].
-Many studies have found that proximity to a bike share station, either living or working near one, is an important determinant of use or having a membership [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011].
+Many studies have found that geographic proximity to a bike share station, either living or working near one, is an important determinant of use or having a membership [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011].
 This makes sense given that individuals are more likely to use services or programs that they can easily access.
@@
 
-Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved since its conceptualization and become linked with recent interest in prioritizing local proximity and modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020].
+Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved over the past century and has increasingly become linked with recent interest in prioritizing local proximity and modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020].
 As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequities in usage, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options.
@@
 
-In the context of PBSPs, accessibility refers to the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019].
-Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program and ultimately use of the program, the location and size (e.g., maximum number of bicycles available) of stations matters.
+Beyond the utility derived from using shared bikes to destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019].
+Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., maximum number of bicycles available) of stations matters.
 Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
@@
 
-Several approaches have been commonly used for measuring accessibility: cumulative opportunities, gravity, and utility-based [@handyMeasuringAccessibilityExploration1997].
-Paez et al [-@paezMeasuringAccessibilityPositive2012] provide a recent overview of various formulations and applications of accessibility in transportation research.
-Briefly, the gravity-based approach involves weighting opportunities, for example the quantity of bicycle share stations, as measured by a function of time, using a negative exponential form [@handyMeasuringAccessibilityExploration1997].
-It takes both demand and congestion into account [@paezMeasuringAccessibilityPositive2012].
-Floating catchment area (FCA) methods are a type of gravity measure that have been used in healthcare accessibility research but that are applicable to transportation case studies.
-This approach involves producing flexible catchment areas for populations, recognizing that people may be willing to travel to access particular services [@paezMeasuringAccessibilityPositive2012].
-Thus, this model does a good job of considering potential crowding of services using a binary distance function [@paezMeasuringAccessibilityPositive2012].
-A recent improvement to this approach was achieved through a simple and intuitive adjustment to the allocation of supply and demand, which addressed the effects of demand and service inflation from the conventional approach [see @paezDemandLevelService2019].
+Several approaches have been commonly used for measuring place-based accessibility, including cumulative opportunities, gravity, and utility-based measures [@handyMeasuringAccessibilityExploration1997].
+Geurs and Van Wee [-@geursAccessibilityEvaluationLanduse2004] and Paez et al.
+[-@paezMeasuringAccessibilityPositive2012] provide recent overviews of various formulations and applications of accessibility in transportation research.
+The common gravity-based approach for example involves weighting destination opportunities, such as the quantity of bicycle share stations, by the time required to reach them from an origin using an impedance function [@handyMeasuringAccessibilityExploration1997; @kwanSpaceTimeIntegral1998].
+However, while such measures are suitable for capturing the potential for reaching destinations from a given location, they do not take demand or congestion effects into account.
+
+In contrast, floating catchment area (FCA) methods incorporate information on capacity and demand in calculating accessibility.
+FCA measures have been widely employed in healthcare accessibility research and are typically calculated across two steps.
+In the first, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hosptital weighted by the distance involved in reaching the facility.
+Next, these service level ratios are allocated back to the population centres and summarized as a measure of congested accessibility.
+Thus, this model does a good job of considering potential crowding or compentition for services.
+While there have been many methodological innovations in FCA methods **(e.g.
+[Radke & Mu](https://www.tandfonline.com/doi/abs/10.1080/10824000009480538), [Luo & Wang, 2003](https://journals.sagepub.com/doi/10.1068/b29120), [Luo & Qi, 2009](https://www.sciencedirect.com/science/article/pii/S1353829209000574?via%3Dihub), [Wan et al., 2012](https://www.tandfonline.com/doi/full/10.1080/13658816.2011.624987), [Delamater, 2013](https://www.sciencedirect.com/science/article/pii/S1353829213001020?via%3Dihub))**, a recent improvement to this approach was achieved through a simple and intuitive balanced FCA approach that addressed the effects of demand and service inflation from previous FCA approaches [see @paezDemandLevelService2019].
 
@@
 
-Using balanced floating catchment area methods, a novel approach that has not been used yet in cycling research, we examine accessibility to a public bicycle share program in Hamilton, Ontario.
-Using disaggregate population-level data, we (1) conduct a sensitivity analysis by measuring accessibility and level of service to bike share stations at different thresholds of walking to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median household income of dissemination areas within the core service area.
+Using the balanced floating catchment area method [@paezDemandLevelService2019], a novel approach that has not been used yet in cycling research, with disaggregate population-level data, we examine accessibility to a public bicycle share program in Hamilton, Ontario.
+We (1) conduct a sensitivity analysis by measuring accessibility and level of service to bike share stations at different thresholds of walking to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median household income of dissemination areas within the core service area.
 
@@
 The city launched a public bicycle share program, Hamilton Bike Share, in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015].
-Before June 2020, the program was known as SoBi Hamilton.
+Before June 2020, the program was known as Social Bicycles or SoBi Hamilton.
 Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021].
 The core service area spans 35 km^2 of the city and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure \ref{fig:hamilton-and-sobi-service-area}].
-This represents roughly one fifth of the population in Hamilton, according to the 2016 Canadian Census.
+This represents roughly one fifth of the population in Hamilton according to the 2016 Canadian Census.
 The program was enthusiastically welcomed in the city - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015].
-In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton.
-An additional 75 bicycles and 15 stations were added to the program, which expanded it to more disadvantaged areas in the core service area [see Figure 2].
+In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, the Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton.
+An additional 75 bicycles and 15 stations were added to the program which expanded it to more disadvantaged areas in the core service area [see Figure 2].
 The program also offers subsidized memberships to individuals who identify as low income.
 A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019].
-As of June 2020, the bike share program has 900 bikes and 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
-
-Hamilton Bike Share has conducted one membership survey to date in 2018 [@civicplanSoBiHamiltonMembership2017], and the findings from 420 members are broadly in line with the trends that were discussed above [see @fishmanBikeshareReviewRecent2016 for a recent review of the literature].
+As of June 2020, the bike share program has 900 bikes, 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
+
+Hamilton Bike Share has conducted one membership survey in 2018 [@civicplanSoBiHamiltonMembership2017] and the findings from 420 members are broadly in line with the trends that were discussed above [see @fishmanBikeshareReviewRecent2016 for a recent review of the literature].
 The majority of respondents live within the core service area and the gender split is typical: 57% of respondents are male and 41% are female.
 The majority of respondents, both male and female, are between 25 and 34 years of age, but the percentage of male respondents is higher in the subsequent age groups.
-Respondents use bike share for commuting (40% of trips) or errands and meetings (24% of trips), and nearly 50% of trips have an average length of 11 to 20 minutes.
+Respondents use bike share for commuting (40% of trips) or errands and meetings (24% of trips) and nearly 50% of trips have an average length of 11 to 20 minutes.
 As a result of having a bike share membership, 49% of respondents report that they use their private vehicle less often or much less often but 48% report that their private vehicle use has remained about the same.
@@
 This approach is more appropriate and informative than calculating the provider-to-population ratio (PPR) which simply divides the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019].
-In particular, the Two-Step Floating Catchment Area method produces flexible catchment areas for populations accessing a service rather than using rigid boundaries like PPR, which provides more useful information because it takes into account individual behaviour and doesn't assume that people stay within the pre-defined boundaries [@paezDemandLevelService2019].
+In particular, the Two-Step Floating Catchment Area (2SFCA) method ([Radke & Mu](https://www.tandfonline.com/doi/abs/10.1080/10824000009480538), [Luo & Wang, 2003](https://journals.sagepub.com/doi/10.1068/b29120)) produces flexible catchment areas for populations accessing a service rather than using rigid boundaries like PPR, which provides more useful information because it doesn't assume that people stay within the pre-defined boundaries [@paezDemandLevelService2019].
 This is an important property that supports our rationale for applying this method to measure accessibility to Hamilton Bike Share.
@@
 
-More recently, the _balanced_ floating catchment area approach was developed to address issues with demand and supply inflation that are a result of the overlapping catchment areas produced by the Two-Step Floating Catchment Area [@paezDemandLevelService2019].
-This overlap inflates the demand and generates inaccurate or misleading accessibility estimates [@paezDemandLevelService2019].
-By adjusting the impedance weights, both supply and demand are proportionally allocated which eliminates the double-counting of population that leads to demand and supply inflation [@paezDemandLevelService2019].
+More recently, the _balanced_ floating catchment area approach was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by past FCA methods [@paezDemandLevelService2019].
+Briefly, the overlapping of catchment areas leads to inflation of population totals and deflation of service levels across a study area and generates inaccurate or misleading accessibility estimates.
+In contrast, Paez et al.
+[-@paezDemandLevelService2019] adjust the impedance weights so that both supply and demand are proportionally allocated which eliminates the over-counting of population that leads to distortions in demand and supply.
 Other benefits of this adjusted method include consideration of competition which can occur when catchment areas overlap, as well as the preservation of system-wide population and level of service.
@@
 
+The first step in the balanced floating catchment area method is to allocate the population to be serviced by each Hamilton Bike Share station:
+$$
+P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}
+$$
+
+Next, the level of service at each station (i.e., the maximum number of bicycle racks) is divided by its estimated service population within the established catchment area:
+$$
+L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
+$$
+
+Finally, the accessibility of population cell $i$ is calculated as the weighted sum of the level of service of all stations that can be reached from there according to normalized weights:
+$$
+A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
+$$
+
+The approach uses instead a set of suitably normalized weights as follows:
+$$
+{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
+$$
+
+and
+$$
+{w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
+$$
+
+These weights satisfy the following properties:
+$$
+\sum_{j = 1}^{J} {w^i_{ji}} = 1
+$$
+
+and
+$$
+\sum_{i = 1}^{n} {w^j_{ji}} = 1
+$$
+
+Finally, accessibility is calculated without risk of demand or supply inflation:
+$$
+A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}
+$$
+
+For this paper, we employ a hybrid place-based and person-based approach to calculating accessibility using disaggregate population data.
 In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data.
 This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are more capable of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences.
-Applying the balanced floating catchment area approach to examine accessibility to Hamilton Bike Share constitutes a location-based measure with an individual component by stratifying according to median household income [@geursAccessibilityEvaluationLanduse2004], but conducting a further sensitivity analysis by adjusting distance thresholds would introduce additional spatio-temporal constraints to evaluate equitable accessibility.
-
-The first step in the balanced floating catchment area method is to allocate the population to be serviced by each Hamilton Bike Share station:
-$$
-P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}
-$$
-
-Next, the level of service at each station (i.e., the maximum number of bicycle racks) is divided by its estimated service population within the established catchment area:
-$$
-L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
-$$
-
-Finally, the accessibility of population cell $i$ is calculated as the weighted sum of the level of service of all stations that can be reached from there according to normalized weights:
-$$
-A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
-$$
-
-The approach uses instead a set of suitably normalized weights as follows:
-$$
-{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
-$$
-
-and
-$$
-{w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
-$$
-
-These weights satisfy the following properties:
-$$
-\sum_{j = 1}^{J} {w^i_{ji}} = 1
-$$
-
-and
-$$
-\sum_{i = 1}^{n} {w^j_{ji}} = 1
-$$
-
-Finally, accessibility is calculated without risk of demand or supply inflation:
-$$
-A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}
-$$
+Applying the balanced floating catchment area approach to examine accessibility to Hamilton Bike Share constitutes a location or place-based measure with an individual component by stratifying according to median household income [@geursAccessibilityEvaluationLanduse2004], but conducting a further sensitivity analysis by adjusting distance thresholds would introduce additional spatio-temporal constraints to evaluate equitable accessibility.
 
@@
 To do so, we retrieved shapefiles for various geographic features [see Table \ref{tab:data-features}] from Open Hamilton.
-Next, we extracted these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population [see Figure \ref{fig:interpolated-population}].
+Next, we removed these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population within the remaining area [see Figure \ref{fig:interpolated-population}].
 
@@
 Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010].
-We found that accessibility increases with a threshold between two and four minutes, but is then maximized at 5 minutes.
-At some stations, Hamilton Bike Share will depict a map to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are anticipated to walk.
+In Hamilton, Hamilton Bike Share will depict a map at some stations to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are anticipated to walk.
+In the present case, we found that congested accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at 5 minutes.
 Accessibility decreases significantly after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each bike share station.
@@
 At the minimum threshold, we find that there are 25.2 bicycle racks per person in the original system.
-With the addition of equity stations, there are now 25.4 bicycles per person.
-Figure \ref{fig:figure-6} presents a comparison of accessibility between the systems.
+The addition of equity stations increases this ratio slightly to 25.4 bicycles per person.
+Figure \ref{fig:figure-6} presents a comparison of accessibility between the systems for the population cells.
 Accessibility is fairly uniform, with the exception of two small areas near the university and the waterfront park where accessibility is slightly higher.
@@
 Instead, we use median total income for each dissemination area in the core service area to examine whether disparities in accessibility exist between income groups for those who live in or near the core service area.
-We divide income by quantiles: bottom 20%, second 20%, third 20%, fourth 20%, and top 20%.
-One of the unique properties of the balanced floating catchment area method is that data can be reaggregated from small to larger polygons while preserving the total population and supply at each station.
+We divide income by quintiles: bottom 20%, second 20%, third 20%, fourth 20%, and top 20%.
+One of the unique properties of the balanced floating catchment area method is that data can be re-aggregated from small to larger polygons while preserving the total population and supply at each station.
 This avoids demand and supply inflation, and also enables us to present findings in a way that is easier to interpret (i.e., across dissemination areas instead of 50 by 50 metre population cells).
 
-Panels 1[\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combined spatial distribution of accessibility and median total income per household for the different thresholds.
+Panels 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combined spatial distribution of accessibility and median total income per household for the different thresholds.
 Our analysis demonstrates that the extreme threshold of fifteen minutes allows the most people to access a station.

### 2021-03-26 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Revised changes to the manuscript, with reorganizing the introduction and discussion.

@@
 
-Background
+Introduction
 ===================
 
+The potential of public bicycle share programs (PBSPs) to increase cycling levels [see @hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019] is one of many reasons for implementing such programs in urban areas.
+As a healthy, inexpensive, and convenient form of public transportation, shared bicycles can encourage individuals to take up cycling for short local trips or first and last mile trips to other public transportation instead of using personal vehicles.
+Although these programs are available to the general public and ought to be accessible for all individuals who wish to use them, research from PBSPs located in North American and European cities indicates that there are inequities in usage and access that need to be addressed.
+
+## Literature Review
+
 Public bicycle share programs (PBSP) have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users [@fishmanBikeshareReviewRecent2016].
-In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @reillyGenderDisparitiesNew2020; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011].
+In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @ogilvieInequalitiesUsagePublic2012; @reillyGenderDisparitiesNew2020; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011].
 However, one study found that bike share users in Washington, DC were more likely to be female [@buckAreBikeshareUsers2013], which suggests that the gender gap among cyclists who use bike share is less disparate than the gap for personal bicycle use [@fishmanBikeshareReviewRecent2016].
 There is some evidence that bike share users are less likely to own a car [@buckAreBikeshareUsers2013; @reillyNoncyclistsFrequentCyclists2020].
-However, the relationship between income or education and bike share use is less clear.
-Stations in disadvantaged communities in Chicago generated most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighbourhoods in Minneapolis-St.
+However, the relationship between income or education and bike share use is more nuanced.
+Stations in disadvantaged communities in Chicago were found to generate most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighbourhoods in Minneapolis-St.
 Paul used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019].
-Being university educated was also a significant correlate of bike share use in Montreal, Canada [@fullerUseNewPublic2011].
-On the other hand, members who reside in minority-concentrated and lower socioeconomic status neighbourhoods use the Minneapolis-St Paul bike share more frequently [@wangNeighborhoodSociodemographicCharacteristics2019].
-Financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016].
-Many studies have found that geographic proximity to a bike share station, either living or working near one, is an important determinant of use or having a membership [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011].
+Similar findings were reported in London [@ogilvieInequalitiesUsagePublic2012].
+Being university educated was a significant correlate of bike share use in Montreal, Canada [@fullerUseNewPublic2011].
+Finally, financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016].
+
+Many studies have found that geographic proximity to a bicycle share station, either living or working near one, is an important determinant of use or having a membership [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011].
 This makes sense given that individuals are more likely to use services or programs that they can easily access.
-Finally, the potential of bike share programs to increase cycling levels has also been explored recently [@hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019], but more longitudinal evidence is needed to determine their impact on encouraging more cycling and offering more opportunities for physical activity.
-On the whole, these findings highlight the need for bike share systems to be more highly accessible for diverse populations in order to increase use beyond the "typical" users, which has been the focus of recent research [@auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020], and to offer more people the option of using sustainable and active transportation.
-
-Several studies have recently explored the equity of PBSP in North American cities by primarily examining who has access to bike share (e.g., differences by demographics or socioeconomic status) and where stations are located.
-Equity can be achieved in two different ways: equal spatial distribution across a region (e.g., horizontal equity) or greater access for vulnerable or disadvantaged populations (e.g., vertical equity) [@chenExploringEquityPerformance2019].
+Several studies have recently explored the equity of PBSP by primarily examining who has access to bike share (e.g., differences by demographics or socioeconomic status) and where stations are located.
+It is important to note that equity can be achieved in two different ways: equal spatial distribution across a region (e.g., horizontal equity) or greater access for vulnerable or disadvantaged populations (e.g., vertical equity) [@chenExploringEquityPerformance2019].
 Both are of interest to researchers and transport planners since they are often linked in that advantage, or conversely disadvantage, has spatial patterns.
@@
 This suggests that they may be using PBSP for work commuting, which points to the importance of ensuring equitable access [@qianBikesharingEquityDisadvantaged2020].
-Similar results were found in Philadelphia, where lower income areas generate fewer trips and efforts to increase equity within the program have not been as successful as intended [@caspiBikesharingPhiladelphiaLowerincome2019].
+Similar results were found in Philadelphia, where lower income areas generated fewer trips and it was suggested that efforts to increase equity within the program have not been as successful as intended [@caspiBikesharingPhiladelphiaLowerincome2019].
 In the case of Seattle, all neighbourhoods had some level of access to dockless bikes but those with higher incomes and more residents of higher education had more bikes [@mooneyFreedomStationSpatial2019].
 Babagoli et al.
-[-@babagoliExploringHealthSpatial2019] found that neighbourhoods in New York City with higher affluence had the greater proportion of Citi Bike stations.
-Overall, these findings suggest that horizontal equity can be achieved while vertical equity is harder to attain.
-Exploring transport equity by investigating where bicycle share stations are located, often using neighbourhoods or census tracts as the spatial unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
+[-@babagoliExploringHealthSpatial2019] also found that neighbourhoods in New York City with higher affluence had the greatest proportion of Citi Bike stations.
+Researchers have recommended targeted expansion of stations to areas that use bike share but have inequitable access [@ogilvieInequalitiesUsagePublic2012].
+
+On the whole, existing studies highlight the need for bicycle share systems to be more highly accessible for diverse populations in order to increase use beyond the "typical" users.
+This has been the focus of recent research [among others, see @auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020].
+Offering more people the option of using sustainable and active transportation, particularly those who have lower socioeconomic status and might benefit the most, is a worthy policy goal for cities with PBSPs.
+Several cities in North America have launched specific programs to address barriers to bike share or to expand service to more deprived areas: London, United Kingdom [@ogilvieInequalitiesUsagePublic2012]; Philadelphia, Pennsylvania [@caspiBikesharingPhiladelphiaLowerincome2019], and Hamilton, Ontario [@hosfordWhoArePublic2018].
+Findings to date suggest that horizontal equity can be easier to achieve than vertical equity.
+However, exploring transportation equity by investigating where bicycle share stations are located, often using neighbourhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
 Meaning that, stations may be lacking in certain neighbourhoods but there may be stations accessible within a reasonable walking cutoff time.
@@
 Beyond the utility derived from using shared bikes to destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019].
-Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., maximum number of bicycles available) of stations matters.
+Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., maximum number of bicycle racks available) of stations matters.
 Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
@@
 FCA measures have been widely employed in healthcare accessibility research and are typically calculated across two steps.
-In the first, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hosptital weighted by the distance involved in reaching the facility.
+In the first, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hospital, weighted by the distance involved in reaching the facility.
 Next, these service level ratios are allocated back to the population centres and summarized as a measure of congested accessibility.
-Thus, this model does a good job of considering potential crowding or compentition for services.
-While there have been many methodological innovations in FCA methods **(e.g.
-[Radke & Mu](https://www.tandfonline.com/doi/abs/10.1080/10824000009480538), [Luo & Wang, 2003](https://journals.sagepub.com/doi/10.1068/b29120), [Luo & Qi, 2009](https://www.sciencedirect.com/science/article/pii/S1353829209000574?via%3Dihub), [Wan et al., 2012](https://www.tandfonline.com/doi/full/10.1080/13658816.2011.624987), [Delamater, 2013](https://www.sciencedirect.com/science/article/pii/S1353829213001020?via%3Dihub))**, a recent improvement to this approach was achieved through a simple and intuitive balanced FCA approach that addressed the effects of demand and service inflation from previous FCA approaches [see @paezDemandLevelService2019].
-
-Researchers have also taken different approaches when it comes to aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones.
+Thus, this model does a good job of considering potential crowding or competition for services.
+While there have been many methodological innovations in FCA methods [for example, @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balanced FCA approach that addressed the effects of demand and service inflation from previous FCA approaches [see @paezDemandLevelService2019].
+
+When measuring accessibility, researchers have taken different approaches when it comes to aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones.
 Previous research on bike share equity has typically used a macro-level approach with aggregate data from entire neighbourhoods or census tracts  [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021].
@@
 
+## Original System
+
 This paper uses the city of Hamilton, located in Ontario, Canada, as a case study.
 The city launched a public bicycle share program, Hamilton Bike Share, in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015].
-Before June 2020, the program was known as Social Bicycles or SoBi Hamilton.
+Before June 2020, the program was known as Social Bicycles or SoBi Hamilton but is now called Hamilton Bike Share.
 Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021].
-The core service area spans 35 km^2 of the city and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure \ref{fig:hamilton-and-sobi-service-area}].
+The core service area spans 40 square kilometres of the city although it was planned to be 20 square kilometres [@hamiltonsobi2015], and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure \ref{fig:hamilton-and-sobi-service-area}].
 This represents roughly one fifth of the population in Hamilton according to the 2016 Canadian Census.
-The program was enthusiastically welcomed in the city - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015].
-In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, the Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton.
-An additional 75 bicycles and 15 stations were added to the program which expanded it to more disadvantaged areas in the core service area [see Figure 2].
-The program also offers subsidized memberships to individuals who identify as low income.
+The City of Hamilton undertook a large public engagement campaign to validate the locations of bicycle share stations that had been selected and to crowdsource potential locations for additional locations [@hamiltonsobi2014].
+Most of the suggested locations were in the east end of the core service area that lack stations or in neighbourhoods not serviced by the PBSP.
+Inadequate coverage given the size of the Hamilton Bike Share's service area was identified as a problem early on, with transportation planners noting that the small size (i.e., supply of bicycle racks) and low quantity of stations would lead to challenges in balancing the system [@hamiltonsobi2015].
+Based on public feedback, 13 stations were added to the original system.
+The program was enthusiastically welcomed in the city in 2015 - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015].
+
+## Equity Initiative
+
+In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton.
+Additional bicycles and stations were added to the program which expanded it to more disadvantaged areas in the core service area [see Figure 2].
+The program also offers subsidized memberships to individuals who identify as low income as well as cycle skills education.
 A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019].
-As of June 2020, the bike share program has 900 bikes, 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
-
-Hamilton Bike Share has conducted one membership survey in 2018 [@civicplanSoBiHamiltonMembership2017] and the findings from 420 members are broadly in line with the trends that were discussed above [see @fishmanBikeshareReviewRecent2016 for a recent review of the literature].
-The majority of respondents live within the core service area and the gender split is typical: 57% of respondents are male and 41% are female.
+
+## Current System
+
+As of June 2020, Hamilton Bike Share has 900 bikes, 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
+The core service area remains 40 square kilometres.
+The PBSP has 13 stations that were added as part of the ERI initiative; we refer to these as "equity stations" throughout the paper while all other stations are "conventional stations".
+In total, 117 stations are "conventional" and 13 are "equity".
+
+## Membership
+
+Hamilton Bike Share conducted one membership survey in 2018 [@civicplanSoBiHamiltonMembership2017] and the findings from 420 members are broadly in line with the trends that were discussed above [see @fishmanBikeshareReviewRecent2016 for a recent review of the literature].
+The majority of respondents live within the core service area and the gender split is expected: 57% of respondents are male and 41% are female.
 The majority of respondents, both male and female, are between 25 and 34 years of age, but the percentage of male respondents is higher in the subsequent age groups.
@@
 
-Further research has provided insights about the behaviour and route choice of bike share users in Hamilton.
-The routes most frequently travelled are longer than the shortest distance route from origin to destination [@luUnderstandingBikeShare2018; @scottRouteChoiceBike2021].
-Ridership is influenced by weather conditions, temporal factors such as university terms, employment, and proximity to important destinations like the local university, but is not influenced by population and most transportation infrastructure variables [@scottWhatFactorsInfluence2019].
-Bike share users prefer to deter from the shortest distance route to travel on streets with cycling facilities or lower volumes of traffic [@luUnderstandingBikeShare2018] and avoid steep slopes and busy roads [@scottRouteChoiceBike2021].
-It is important to note that these studies used daily ridership or GPS data from trips taken before the equity stations were implemented, so these findings are informative with respect to the conventional stations.
-
-Our analysis builds upon a previous and recent study [@hosfordWhoArePublic2018], which found that disadvantaged areas in Hamilton are better served by the city's PBSP compared to other Canadian cities [i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau] with PBSPs where advantaged areas have greater access.
+## Relevant Research
+
+Our analysis builds upon a previous and recent study [@hosfordWhoArePublic2018], which found that areas in Hamilton with less advantage are better served by the city's PBSP compared to other Canadian cities [i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau] where areas that are less deprived have greater access.
 Hosford and Winters [-@hosfordWhoArePublic2018] acknowledge that "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas".
-Therefore, the core service area for the PBSP in Hamilton is by default in more disadvantaged areas compared to other Canadian cities, but there is also a great deal of variation in income in the city center as well because of the local university and increasing gentrification in the downtown area.
-In their analysis, Hosford and Winters [-@hosfordWhoArePublic2018] did not differentiate between the conventional stations and those added to increase equity, which we refer to as equity stations in this paper.
-Therefore, this paper continues their work and makes an important contribution to the growing body of literature that examines equity issues in PBSPs.
+Therefore, the core service area for the PBSP in Hamilton is by default in more disadvantaged areas compared to other parts of the city.
+However, there is also a great deal of variation in income in the city center because of the local university and increasing gentrification.
+Hosford and Winters [-@hosfordWhoArePublic2018] took a macro-level and system-wide approach in their analysis by using dissemination areas _across_ the city as the unit of analysis.
+They did not focus specifically on the core service area and did not differentiate between conventional and equity stations, which is worth exploring further.
+Therefore, this paper continues their analysis and makes an important contribution to the growing body of literature that examines equity issues in PBSPs.
 
@@
 This approach is more appropriate and informative than calculating the provider-to-population ratio (PPR) which simply divides the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019].
-In particular, the Two-Step Floating Catchment Area (2SFCA) method ([Radke & Mu](https://www.tandfonline.com/doi/abs/10.1080/10824000009480538), [Luo & Wang, 2003](https://journals.sagepub.com/doi/10.1068/b29120)) produces flexible catchment areas for populations accessing a service rather than using rigid boundaries like PPR, which provides more useful information because it doesn't assume that people stay within the pre-defined boundaries [@paezDemandLevelService2019].
+In particular, the Two-Step Floating Catchment Area (2SFCA) method [@luoMeasuresSpatialAccessibility2003; @radkeSpatialDecompositionsModeling2000] produces flexible catchment areas for populations accessing a service rather than using rigid boundaries like PPR, which provides more useful information because it doesn't assume that people stay within the pre-defined boundaries [@paezDemandLevelService2019].
 This is an important property that supports our rationale for applying this method to measure accessibility to Hamilton Bike Share.
-The City of Hamilton has positioned stations between 300 and 600 metres apart, which would lead to boundaries around these buffers, but it would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.
-
-More recently, the _balanced_ floating catchment area approach was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by past FCA methods [@paezDemandLevelService2019].
+The City of Hamilton has positioned stations between 300 and 600 metres apart, but anticipates that hubs will service those living within a 250 metre buffer from the station [@hamiltonsobi2015].
+The latter constitutes a normative statement: people ought to be able to access a station in less than 600 metres if they live in the core service area, with usage coming from 250 metres around.
+However, it is not known how far people are actually willing to travel to reach a station.
+It would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.
+
+More recently, the _balanced_ floating catchment area (FCA) approach was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by past FCA methods [@paezDemandLevelService2019].
 Briefly, the overlapping of catchment areas leads to inflation of population totals and deflation of service levels across a study area and generates inaccurate or misleading accessibility estimates.
@@
 
-For this paper, we employ a hybrid place-based and person-based approach to calculating accessibility using disaggregate population data.
+For this paper, we employ a hybrid location-based and person-based approach to calculating accessibility using disaggregate population data.
 In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data.
 This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are more capable of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences.
-Applying the balanced floating catchment area approach to examine accessibility to Hamilton Bike Share constitutes a location or place-based measure with an individual component by stratifying according to median household income [@geursAccessibilityEvaluationLanduse2004], but conducting a further sensitivity analysis by adjusting distance thresholds would introduce additional spatio-temporal constraints to evaluate equitable accessibility.
+Applying the balanced floating catchment area approach to allows us to examine accessibility by stratifying according to median household income, which would constitute the individual component of the accessibility measure [@geursAccessibilityEvaluationLanduse2004].
+However, conducting a further sensitivity analysis to measure accessibility at different distance thresholds would help to consider potential spatio-temporal constraints.
+Different people may be willing to travel different distances to access a public bicycle share station.
 
@@
 A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bike share stations within the maximum threshold.
+This provides a more realistic measure of the temporal component by taking infrastructure into account of travel times, rather than using the Euclidean distance from population cell to station.
 
@@
 
-All data for this research were accessed from Open Hamilton^[https://open.hamilton.ca/], an online repository of data curated by the City of Hamilton.
+All data for this research were accessed from Open Hamilton^[https://open.hamilton.ca/], an online and public repository of data curated by the City of Hamilton.
 
@@
 
-Consensus regarding the distance that individuals are willing to walk to access a PBSP is lacking, therefore the literature on walking behaviour was consulted to determine the thresholds for the sensitivity analysis.
+Consensus regarding the distance that individuals are willing to walk to access a PBSP is lacking, but the literature on walking behaviour was consulted to determine the thresholds for the sensitivity analysis.
 Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010].
-In Hamilton, Hamilton Bike Share will depict a map at some stations to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are anticipated to walk.
+In Hamilton, Hamilton Bike Share will depict a map at some stations to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are expected to walk.
+
 In the present case, we found that congested accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at 5 minutes.
@@
 
-For this reason, we experiment with various walking thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
+For this reason, we experiment with various walking thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times from population cell to bicycle share station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
 We categorize these thresholds as minimum, average, maximum, and extreme, respectively.
@@
 
-### Three Minutes
-
-At the minimum threshold, we find that there are 25.2 bicycle racks per person in the original system.
+### Minimum Threshold
+
+With a walking distance of three minutes, we find that there are 25.2 bicycle racks per person in the original system.
 The addition of equity stations increases this ratio slightly to 25.4 bicycles per person.
 Figure \ref{fig:figure-6} presents a comparison of accessibility between the systems for the population cells.
-Accessibility is fairly uniform, with the exception of two small areas near the university and the waterfront park where accessibility is slightly higher.
+Accessibility is fairly uniform overall, with the exception of two small areas where accessibility is slightly higher.
 
@@
 
-### Five Minutes
-
-At the average threshold, we find that there are 68.6 bicycle racks per person in the original system.
+### Average Threshold
+
+With a walking distance of five minutes, we find that there are 68.6 bicycle racks per person in the original system.
 With the addition of equity stations, there are now 68.8 bicycles per person.
 Figure \ref{fig:figure-7} presents a comparison of accessibility between the systems.
-Again, accessibility is fairly uniform, with the exception of one very small area near the waterfront park where accessibility is higher.
+Again, accessibility is fairly uniform, with the exception of one very small area.
+At this threshold, there are more bicycle racks per person than at the minimum threshold.
+This occurs because fewer stations can be reached when the travel time is 3 minutes or less.
 
@@
 
-### Ten Minutes
-
-At the maximum threshold, we find that there are 3.61 bicycle racks per person in the original system.
+### Maximum Threshold
+
+With a walking distance of ten minutes, we find that there are 3.61 bicycle racks per person in the original system.
 With the addition of equity stations, there are now 3.74 bicycles per person.
 Figure \ref{fig:figure-8} presents a comparison of accessibility between the systems.
-We begin to see differences in accessibility across the service area, with users near the university and its adjacent neighbourhoods, as well as neighbourhoods north of the downtown area, have slightly higher accessibility.
+We begin to see differences in accessibility across the service area, with users near the university and its adjacent neighbourhoods, as well as neighbourhoods north of the downtown area (outlined in green), having slightly higher accessibility.
 While the differences are modest, they are more apparent at this threshold than at shorter walking distances.
@@
 
-### Fifteen Minutes
-
-At the extreme threshold, we find that there are 2.44 bicycle racks per person in the original system.
+### Extreme Threshold
+
+With a walking distance of fifteen minutes, we find that there are 2.44 bicycle racks per person in the original system.
 With the addition of equity stations, there are now 2.55 bicycles per person.
@@
 
-## Accessibility by Median Total Income
-
-Approximately 138,000 people live within Hamilton Bike Share's core service area or within a 30 minute walk to the service area.
-We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}].
-While previous research found that neighbourhoods with more disadvantage are better serviced by Hamilton Bike Share [@hosfordWhoArePublic2018], the authors used the Pampalon Deprivation Index from 2011 as a measure of socioeconomic status for dissemination areas across the city not just within the core service area.
-Instead, we use median total income for each dissemination area in the core service area to examine whether disparities in accessibility exist between income groups for those who live in or near the core service area.
+## Accessibility by Median Household Income
+
+One of the unique properties of the balanced floating catchment area method is that data can be reaggregated while preserving the total population in the service area and the supply at each station.
+This avoids demand and supply inflation, and also enables us to present findings in a way that is easier and more intuitive to interpret.
+Therefore, we reaggregate population and accessibility from the smallest 50 by 50 metres population cells to the median household income for each dissemination area.
 We divide income by quintiles: bottom 20%, second 20%, third 20%, fourth 20%, and top 20%.
-One of the unique properties of the balanced floating catchment area method is that data can be re-aggregated from small to larger polygons while preserving the total population and supply at each station.
-This avoids demand and supply inflation, and also enables us to present findings in a way that is easier to interpret (i.e., across dissemination areas instead of 50 by 50 metre population cells).
-
-Panels 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combined spatial distribution of accessibility and median total income per household for the different thresholds.
+
+Panels 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combined spatial distribution of accessibility and median household income at different thresholds.
 Our analysis demonstrates that the extreme threshold of fifteen minutes allows the most people to access a station.
-We find that stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing accessibility, albeit only modestly [see Table \ref{tab:accessibility-income}].
+We find that stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing access, albeit only modestly [see Table \ref{tab:accessibility-income}].
 By implementing equity stations in more areas with lower median total income in Hamilton, the PBSP has achieved greater horizontal equity by extending the spatial distribution of bicycles across the city.
-This is particularly evident at the minimum and average thresholds of three and five minutes, respectively.
+This is particularly evident at the minimum and average thresholds of three and five minutes, respectively, where the equity stations fill a number of gaps in program coverage.
 The largest gains were made for dissemination areas in the second 20%, where an additional 3,073 and 5,395 people could reach a bicycle share station within three and five minutes walk, respectively, after the addition of the equity station.
 
-However, we found that dissemination areas with the lowest total household income do not have much greater access to the program [see Table \ref{tab:table-3}].
+However, we found that dissemination areas with the lowest median household income do not have much greater access to the program [see Table \ref{tab:table-3}].
 Income disparities still persist, however only at certain thresholds.
-With and without equity stations, people in the top 20% of income have the highest level of accessibility at a threshold of ten and fifteen minutes.
-Although dissemination areas in the second 20% have the highest level of accessibility by a significant amount at lower thresholds, the bottom 20%, who may benefit the most from Hamilton Bike Share's equity initiatives, have the lowest accessibility at three minutes threshold and the second lowest accessibility at all other thresholds.
-Panels 1[\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] reveal specific areas where accessibility is low or zero, which are good candidates for additional equity stations.
-
-On the whole, we find that areas that have less than their equitable share of the level of service are some of the most socioeconomically disadvantaged areas in the city.
-While the addition of equity stations increases accessibility for all income groups at all thresholds, they did not increase accessibility significantly for any single income group.
-This finding aligns with other studies that have found disparities in station or bike location along income lines in New York City [@babagoliExploringHealthSpatial2019], Tampa [@chenExploringEquityPerformance2019], and Seattle [@mooneyFreedomStationSpatial2019], where people who have greater advantage have better access to PBSPs.
+With and without equity stations, people in the top 20% of income have the highest level of access at a threshold of ten and fifteen minutes.
+Although dissemination areas in the second 20% have the highest level of access by a significant amount at lower distance thresholds, the bottom 20%, who may benefit the most from Hamilton Bike Share's equity initiatives, have the lowest access at three minutes threshold and the second lowest access at all other thresholds.
 
@@
 
+Discussion
+===================
+
+Using disaggregate data, we examined spatial equity and accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
+Our balanced floating catchment area approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro sale.
+This differentiates our analysis from similar papers exploring equity in PBSPs that use larger geographical units of analysis or that focus on station location instead of level of service.
+In this way, our paper has made contributions in a positive way by applying an intuitive and useful approach to measure accessibility to a PBSP, and in a normative way by serving to inform future investments in cycle infrastructure for Hamilton Bike Share.
+
+The sensitivity analysis revealed that accessibility to Hamilton Bike Share stations is maximized at five minutes and decreases significantly by eight minutes [see Table \ref{tab:accessibility-income}].
+This reflects the normative guide, advertised on some bicycle share stations in Hamilton, that people can access other stations within a five minute walk if the station of origin has no bicycles.
+We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}].
+At a minimum threshold of three minutes, too few stations can be accessed which leads to lower levels of accessibility.
+However, accessibility is at its lowest after eight minutes whereby congestion occurs due to increased potential demand.
+The City of Hamilton has recognized from the launch of the PBSP that significantly more stations and bicycles are needed to service the area [@hamiltonsobi2015].
+With a service area of 40 square kilometres, Hamilton should have between 380 and 440 stations, instead of 130, and 1,500 bicycles instead of 900 [@hamiltonsobi2015].
+Reduced capacity within the system leads to gaps in coverage in some areas of the city "with some areas not having the recommended station density of 300m between stations or 10 stations per square km" [@hamiltonsobi2015].
+This study illustrates the consequences of an imbalanced system whereby levels of accessibility are not equitable across income groups and become very low when congestion occurs.
+
+We found that equity stations achieved their goal of increasing horizontal equity in Hamilton's core service area.
+Panels 1[\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] demonstrate how gaps in service were filled by these additional stations.
+In this respect, there are some commonalities between the expansions of Hamilton Bike Share and that of Citi Bike in New York.
+Babgoli et al.
+[-@babagoliExploringHealthSpatial2019] found a slight but not statistically significant increase in the proportion of neighbourhoods with the highest levels of poverty with stations after the Citi Bike expansion in 2015.
+Although the Citi Bike expansion was not specifically driven by a desire to reduce inequities in access, 16% of neighbourhoods with the most poverty had stations compared to 12% before.
+With the addition of equity stations, there were large gains in accessibility for the second 20% at the average threshold, over 5,000 more people, but much smaller gains for the bottom 20% with only 326 more people.
+
+Vertical inequities, however, continue to persist despite the addition of equity stations.
+While the addition of equity stations modestly increases accessibility for all income groups at all thresholds, they did not increase accessibility significantly for any single income group.
+Most importantly, individuals in the bottom 20% median household income quintile have the second lowest level of access to Hamilton Bike Share at most thresholds (average, maximum, and extreme).
+At the minimum threshold, the bottom 20% have the lowest level of access.
+While previous research found that neighbourhoods with more disadvantage are better serviced by Hamilton Bike Share, the authors used the Pampalon Deprivation Index to determine the level of disadvantage for dissemination areas _across_ the city not just within the core service area [@hosfordWhoArePublic2018].
+Instead, we use median household income for each dissemination area _within_ the core service area.
+We conclude that Hamilton's PBSP, while by default located in areas with more deprivation compared to other cities, has disparities in accessibility between income groups.
+With equity stations, many areas with low median household income in the east end of the service area continue to have low accessibility.
+At higher thresholds of 10 and 15 minutes, the top 20% have the highest level of access to Hamilton Bike Share.
+This aligns with other studies from Tampa [@chenExploringEquityPerformance2019] and Seattle [@mooneyFreedomStationSpatial2019], which have found disparities in station location or access to bicycles between levels of income and education.
+
+Based on our analysis, we identified specific areas that have both low accessibility and low median household income which would benefit from an increased supply of public bicycles.
+These empirical findings provide support to the City of Hamilton's efforts to increase equity and balance the system, and confirm that additional stations and bicycles are needed to improve access not only for the bottom 20% but for all income groups.
+Panels 1[\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] highlight potential locations for new equity stations.
+
 Study Limitations
@@
 Therefore, further research is needed to determine whether the addition of equity stations encouraged more cycling for low-income individuals living near them.
-Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighbourhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of accessible route or user data to conduct similar analyses for Hamilton Bike Share.
+Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighbourhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of publicly available route or user data to conduct similar analyses for Hamilton Bike Share.
 
@@
 The addition of specific equity stations to the public bicycle share program in Hamilton, Ontario had the net effect of increasing accessibility and reducing both horizontal and vertical inequities.
-In particular, accessibility improved the most for the second 20% median total income groups at all thresholds, but the gains were only modest for all income groups.
-Dissemination areas with the bottom 20% had the lowest accessibility at three minutes, and second lowest level of accessibility at five, ten, and fifteen minutes.
+In particular, accessibility improved the most for those in the second 20% median household income at all thresholds, but the gains were only modest for all income groups.
+Dissemination areas with the bottom 20% had the lowest accessibility at three minutes, and second lowest levels of accessibility at five, ten, and fifteen minutes.
 Congestion effects were observed at higher thresholds, with accessibility decreasing significantly once the catchment area is increased to 10 minutes walking.
-Based on our analysis, we identified specific areas that have both low accessibility and low income which would benefit from an increased supply of bicycles.
-These are ideal candidates for new equity stations.
-In this way, our paper has made contributions in a positive way by applying an intuitive and useful approach to measure accessibility to a PBSP, and in a normative way by serving to inform future investments in cycle infrastructure for Hamilton Bike Share.
 

### 2021-03-26 Bike-share-spatial-equity/Cover-Letter.Rmd
Cover letter for submission is written for journal

---
title: "Cover Letter"
output:
  pdf_document: default
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

To the Editor:

Please find attached our paper submitted to the Virtual Special Issue in Transportation Research Part D: Transport and Environment (SI: Transportation Justice). The paper is titled "Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach". We use the city of Hamilton, Ontario in Canada as a case study. The city's public bicycle share program, Hamilton Bike Share, launched in 2015 and started an equity initiative in 2017 which included the addition of 13 stations to the program.

We believe that this paper makes the following contributions to the literature which makes it an excellent inclusion to the special issue on transportation justice:

1.	Our research contributes to the literature on the topic of equity in public bicycle share programs (PBSPs) which has received a lot of attention in North America recently. Other papers include Caspi & Noland, 2019, Hosford & Winters, 2018, Mooney et al., 2019, Qian et al., 2019, among others. The city of Hamilton is the only city in Canada with a PBSP that has an explicit equity program. This makes it an interesting place to examine how specific bike share stations added for the purpose of increasing equity have contributed to addressing horizontal and vertical inequities (see Chen et al., 2019). This case study also builds upon the findings from a recent paper by Hosford & Winters (2018) that explored who is served by PBSPs in Canada.

2. We employ a balanced floating catchment area approach using disaggregate population-level data to measure accessibility at different thresholds: 3 minutes, 5 minutes, 10 minutes, and 15 minutes. This method has not been previously used in the cycling literature to measure accessibility. Unlike other location- and person-based measures of accessibility, floating catchment area methods incorporate information on capacity and demand in calculating accessibility. These measures have been widely employed in healthcare accessibility research and are typically calculated across two steps. In the first, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hospital, weighted by the distance involved in reaching the facility. Next, these service level ratios are allocated back to the population centres and summarized as a measure of congested accessibility. Thus, this model does a good job of considering potential crowding or competition for services, which is an important consideration for PBSPs. Specifically, we use a recent improvement to this approach that was achieved through a simple and intuitive balancing that addressed the effects of demand and service inflation from previous FCA approaches. We also compare accessibility in the system with and without the equity stations at the different thresholds.

3. We found that accessibility is maximized at 5 minutes and gradually decreases until 8 minutes where it plateaus. The analysis indicates that the equity stations increase accessibility, albeit only modestly, for all income quintiles. The second 20% median household income group has the highest level of accessibility at 3 and 5 minutes, while the top 20% has the highest level of accessibility at 10 and 15 minutes. The bottom 20%, those who would arguably benefit the most from inexpensive and healthy transportation options like shared bicycles, have the lowest level of accessibility at 3 minutes and the second lowest level at all other thresholds. This indicates that individuals with the lowest socioeconomic status continue to have poorer access to Hamilton Bike Share despite the addition of equity stations. Based on our analysis, we identify locations in the PBSP core service area where individuals have low median household income and low accessibility, which are ideal candidates for new stations to achieve greater vertical equity in Hamilton Bike Share.

We trust that you will find the paper is worthy of inclusion in the special issue on transportation justice.

Sincerely,

The Authors

### 2021-03-26 Bike-share-spatial-equity/Highlights.Rmd
Highlights for journal

---
title: "Highlights"
output:
  pdf_document: default
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

* Balanced floating catchment area methods are applied to measure accessibility to PBSP

* Contribution of equity stations to accessibility assessed at different thresholds

* Equity stations increase accessibility modestly for all income groups

* Lowest income group has lowest levels of accessibility, even with equity stations

* Locations for new stations are identified to increase equitable access to PBSP

### 2021-03-27 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
change geom_sf to geom_tile

@@
 
-Beyond the utility derived from using shared bikes to destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019].
+Beyond the utility derived from using shared bikes to reach destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019].
 Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., maximum number of bicycle racks available) of stations matters.
@@
 Thus, this model does a good job of considering potential crowding or competition for services.
-While there have been many methodological innovations in FCA methods [for example, @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balanced FCA approach that addressed the effects of demand and service inflation from previous FCA approaches [see @paezDemandLevelService2019].
+While there have been many methodological innovations in FCA methods [for example, @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balanced FCA method that addressed the effects of demand and service inflation from previous FCA approaches [see @paezDemandLevelService2019].
 
@@
 
+```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
+The downtown area is outlined in dark green.
+Geographic features have been extracted.
+Dissemination area polygons are outlined in grey."}
+ggplot() + 
+  geom_sf(data = cropped_da,
+          fill = "gray",
+          color = NA) + 
+  #geom_sf(data = population_50x50,
+  #        aes(color = population)) +
+   #scale_fill_gradient(colours = "YlOrRd") +
+  geom_tile(data = cbind(st_drop_geometry(population_50x50), st_coordinates(population_50x50)),
+            aes(x = X, y = Y, fill = population)) +
+  scale_fill_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA,
+          colour = "black") +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"))
+```
+
 ## Travel Time Matrix
@@
 ```{r figure-6, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
+accessibility_3min <- accessibility %>% dplyr::filter(threshold == "3 min")
 ggplot() +
@@
           color = "gray") +
-  geom_sf(data = accessibility %>%
-            dplyr::filter(threshold == "3 min"),
-          aes(color = accessibility)) +
+  #geom_sf(data = accessibility %>%
+  #          dplyr::filter(threshold == "3 min"),
+  #        aes(color = accessibility)) +
+  #scale_color_distiller(palette = "PuRd", 
+  #                      direction = 1) +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_3min), st_coordinates(accessibility_3min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_distiller(palette = "PuRd", direction = 1) +
   geom_sf(data = sobi_service,
           fill = NA) +
-  scale_color_distiller(palette = "PuRd", 
-                        direction = 1) +
   facet_grid(system ~ threshold)  +
@@
   #         fill = "green") +
-  theme(panel.background = element_rect(fill = "white"))
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank())
 ```
@@
 #accessibility plot for current system with equity stations at 5 minutes walk
+accessibility_5min <- accessibility %>% dplyr::filter(threshold == "5 min")
+
 ggplot() +
@@
           color = "gray") +
-  geom_sf(data = accessibility %>%
-            dplyr::filter(threshold == "5 min"),
-          aes(color = accessibility)) +
+  #geom_sf(data = accessibility %>%
+  #          dplyr::filter(threshold == "5 min"),
+  #        aes(color = accessibility)) +
+  #scale_color_distiller(palette = "PuRd", 
+  #                      direction = 1) +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_5min), st_coordinates(accessibility_5min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_distiller(palette = "PuRd", direction = 1) +
   geom_sf(data = sobi_service,
           fill = NA) +
-  scale_color_distiller(palette = "PuRd", 
-                        direction = 1) +
   facet_grid(system ~ threshold)  +
@@
 #accessibility plot for current system with equity stations at 10 minutes walk
+accessibility_10min <- accessibility %>% dplyr::filter(threshold == "10 min")
 ggplot() +
@@
           color = "lightgray") +
-  geom_sf(data = accessibility %>%
-            dplyr::filter(threshold == "10 min"),
-          aes(color = accessibility)) +
+  #geom_sf(data = accessibility %>%
+  #          dplyr::filter(threshold == "10 min"),
+  #        aes(color = accessibility)) +
+  #scale_color_distiller(palette = "PuRd", 
+  #                      direction = 1) +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_10min), st_coordinates(accessibility_10min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_distiller(palette = "PuRd", direction = 1) +
   geom_sf(data = sobi_service,
           fill = NA) +
-  scale_color_distiller(palette = "PuRd", 
-                        direction = 1) +
   facet_grid(system ~ threshold)  +
@@
 #accessibility plot for current system with equity stations at 15 minutes walk
+accessibility_15min <- accessibility %>% dplyr::filter(threshold == "15 min")
 ggplot() +
@@
           color = "lightgray") +
-  geom_sf(data = accessibility %>%
-            dplyr::filter(threshold == "15 min"),
-          aes(color = accessibility)) +
+  #geom_sf(data = accessibility %>%
+  #          dplyr::filter(threshold == "15 min"),
+  #        aes(color = accessibility)) +
+  #scale_color_distiller(palette = "PuRd", 
+  #                      direction = 1) +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_15min), st_coordinates(accessibility_15min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_distiller(palette = "PuRd", direction = 1) +
   geom_sf(data = sobi_service,
           fill = NA) +
-  scale_color_distiller(palette = "PuRd", 
-                        direction = 1) +
+
   facet_grid(system ~ threshold)  +

### 2021-03-28 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Proof-read and edited first half of paper in preparation for submission

--- 2021-03-26 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
@@
 ---
-title: Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach 
+title: Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach 
 author:
@@
 
-  Public bicycle share programs are implemented to promote cycling as a convenient and sustainable mode of transportation.
+  Public bicycle share programs (PBSPs) are implemented to promote cycling as a convenient and sustainable mode of transportation.
 These systems can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to lower income or disadvantaged populations.
-Recent studies assessing equity in bike sharing have found differences in ridership or membership based on income, education, and trip purpose.
 In Ontario, the City of Hamilton launched a public bicycle share program in 2015 that currently has over 900 operational bicycles and 130 docking stations.
-In 2018, the non-profit organization that operates the program launched an equity initiative to provide subsidized memberships and to expand service by adding twelve docking stations.
-Previous research found that Hamilton’s public bicycle share program targeted well disadvantaged areas in the city compared to programs in other Canadian cities.
-Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program, the location of stations matters.
-Unlike other public amenities with greater ability to adapt to crowding, the utility of stations is limited by the number of bicycles that they can hold, which makes crowding effects critical; the system may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles.
-In this research, we revisit the case of Hamilton and investigate equity differentials in accessibility to bicycle share stations.
-We compare accessibility in the program with and without the equity stations to assess the effect of the initiative.
-Previous research on cycling equity or accessibility has used a macro-level approach with entire neighbourhoods or census tracts as the unit of analysis, although there are recent exceptions.
-In contrast, we implement our analysis parting from microscopic zones to better reflect travel to a docking station, which often happens at the sub-neighborhood level, and conduct a sensitivity analysis at several walking thresholds.
-We then reaggregate the estimated accessibility to dissemination areas for further analysis using census data.
-Equity analysis of this type is possible thanks to a newly developed approach for accessibility using balanced floating catchment areas.
-The addition of equity stations achieved its goal to increase accessibility for lower income populations, although the increase was relatively modest (2.10 bicycles/person with equity stations vs 1.97 bicycles/person without equity stations).
-Further research is needed to determine whether this encouraged more cycling.
+This system was expanded in 2018 by an equity initiative that added twelve docking stations with the explicit objective of increasing spatial equity in access.
+Since the cost of reaching a bicycle share station decreases the potential of accessing the program, the location of stations matters, and previous research found that Hamilton’s public bicycle share program targeted well disadvantaged areas in the city.
+In this research, we revisit the case of Hamilton and investigate differentials in accessibility to bicycle share stations using a balanced floating catchment area (BFCA) accessibility approach.
+We compare accessibility  with and without the equity stations to assess the effect of the initiative.
+We implement our analysis parting from micro zones to better reflect walking to a docking station, and conduct a sensitivity analysis at several walking time thresholds.
+We then reaggregate the estimated accessibility for further analysis using census data.
+The addition of equity stations increased the serviced population at every threshold examined, and although accessibility increased for the whole population, this increase was relatively modest especially for population in the bottom 20% of median household income.
 
@@
 date: "`r Sys.Date()`"
-bibliography: references.bib
+bibliography: [packages.bib, references.bib]
 #linenumbers: true
@@
 
-```{r, include=FALSE}
-rm(list = ls())
+```{r install-data-package, include = FALSE}
+# Run only once if needed to install data package `sobiEquity`
+if (!require("sobiEquity", character.only = TRUE)) {
+      devtools::install_github("https://github.com/paezha/Accessibility-Sobi-Hamilton", subdir = "sobiEquity")
+  }
 ```
@@
 library(units)
-#library(disk.frame)
-#library(pycno)
-#library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
+library(disk.frame)
+library(pycno)
+library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
+```
+
+```{r generate-package-bibliography, include=FALSE}
+knitr::write_bib(.packages(), "packages.bib")
 ```
@@
 ```{r invoke-data, include=FALSE}
+# The data is part of data package `sobiEquity` 
 data("community_downtown")
@@
 
+\newpage
+
 Introduction
@@
 
-The potential of public bicycle share programs (PBSPs) to increase cycling levels [see @hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019] is one of many reasons for implementing such programs in urban areas.
+The potential of public bicycle share programs (PBSPs) to increase cycling levels is but one of many reasons for implementing such programs in urban areas [@hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019].
 As a healthy, inexpensive, and convenient form of public transportation, shared bicycles can encourage individuals to take up cycling for short local trips or first and last mile trips to other public transportation instead of using personal vehicles.
-Although these programs are available to the general public and ought to be accessible for all individuals who wish to use them, research from PBSPs located in North American and European cities indicates that there are inequities in usage and access that need to be addressed.
-
-## Literature Review
-
-Public bicycle share programs (PBSP) have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users [@fishmanBikeshareReviewRecent2016].
+These systems can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to lower income or disadvantaged populations.
+As these systems become increasingly common, their introduction has been accompanied by a flurry of research that investigates them from the perspective of spatial and transportation equity [e.g., @smith2015exploring; @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @hullgrassoBikeShareEquity2020; @qianBikesharingEquityDisadvantaged2020; @qianEnhancingEquitableService2020].
+Indeed, although these programs are available to the general public and ought to be accessible to any individuals who wish to use them, research on PBSPs located in North American and European cities indicates that there are inequities in spatial accessibility in need of redress.
+
+In this paper we investigate the case of the PBSP in the City of Hamilton, in Ontario, Canada.
+Hamilton Bike Share (previously also known as Social Bicycles, or SoBi) was launched in 2015 and currently has over 900 operational bicycles and 130 docking stations.
+An interesting feature of this system is that it was expanded by an equity initiative that introduced twelve docking stations in 2018 with the explicit objective of increasing spatial accessibility.
+Hamilton Bike Share was already studied by @hosfordWhoArePublic2018 as part of a selection of PBSPs in Canada.
+These authors found that most of the cities in their research could benefit of greater efforts to expand service to areas with lower socio-economic populations - but Hamilton fared somewhat better than other cities in this respect.
+What distinguishes our research is the use of a balanced floating catchment area (BFCA) accessibility approach that accounts for the supply of stations and the potential demand from the population serviced.
+This requires a more disaggregated approach than the use of census geographies [q.v., @hosfordWhoArePublic2018], since many trips to a bike share station likely happen at a level lower than even the smallest census geography.
+For this reason, we implement our analysis parting from micro population zones to better reflect the friction of walking to a docking station.
+Further, we conduct a sensitivity analysis at several walking time thresholds.
+In terms of spatial equity, we compare accessibility with and without the equity stations to assess the effect of the initiative, before reaggregating the data for further analysis using median household income information from the census.
+We find that the addition of equity stations does increase the serviced population at every threshold examined.
+However, although accessibility increased for the whole population, this increase was relatively modest, especially for population in the bottom 20% of median household income.
+The analysis also identifies areas under-serviced within the core service area of the system.
+
+This paper is an example of open and reproducible research that uses only open software for transportation and statistical analysis [@bivand2020progress; @lovelace2021open].
+All data were obtained from publicly available sources and organized in the form a data package.
+Following best practices in spatial data science [@brunsdon2020opening], the code and data needed to reproduce (or modify/extend) the analysis are available for download.^[https://drive.google.com/drive/folders/1ZGRpSN2LxS2Fq2tLVJgfm1sI6I26vzZh?usp=sharing]
+
+Literature Review
+===================
+
+Public bicycle share programs have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users [@fishmanBikeshareReviewRecent2016].
 In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @ogilvieInequalitiesUsagePublic2012; @reillyGenderDisparitiesNew2020; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011].
@@
 There is some evidence that bike share users are less likely to own a car [@buckAreBikeshareUsers2013; @reillyNoncyclistsFrequentCyclists2020].
-However, the relationship between income or education and bike share use is more nuanced.
-Stations in disadvantaged communities in Chicago were found to generate most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighbourhoods in Minneapolis-St.
+However, the relationship between income or education and bike share use is less clear-cut.
+Stations in disadvantaged communities in Chicago have been found to generate most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighbourhoods in Minneapolis-St.
 Paul used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019].
@@
 Being university educated was a significant correlate of bike share use in Montreal, Canada [@fullerUseNewPublic2011].
-Finally, financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016].
-
-Many studies have found that geographic proximity to a bicycle share station, either living or working near one, is an important determinant of use or having a membership [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011].
-This makes sense given that individuals are more likely to use services or programs that they can easily access.
-Several studies have recently explored the equity of PBSP by primarily examining who has access to bike share (e.g., differences by demographics or socioeconomic status) and where stations are located.
+Perhaps not coincidentally, financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016].
+
+Many studies have found that geographic proximity to a bicycle share station is an important determinant of membership and use [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011].
+This makes sense given that individuals are more likely to use services or programs that they can easily reach.
+Several studies have recently explored the equity of PBSPs by primarily examining who has access to bike share (e.g., differences by demographics or socioeconomic status) and where stations are located.
 It is important to note that equity can be achieved in two different ways: equal spatial distribution across a region (e.g., horizontal equity) or greater access for vulnerable or disadvantaged populations (e.g., vertical equity) [@chenExploringEquityPerformance2019].
@@
 While annual members in disadvantaged communities have a significantly lower share of trips compared to other areas in the city, they make longer trips.
-This suggests that they may be using PBSP for work commuting, which points to the importance of ensuring equitable access [@qianBikesharingEquityDisadvantaged2020].
+This suggests that they may be using PBSPs for utilitarian trips (e.g., commuting), which points to the importance of ensuring equitable access [@qianBikesharingEquityDisadvantaged2020].
 Similar results were found in Philadelphia, where lower income areas generated fewer trips and it was suggested that efforts to increase equity within the program have not been as successful as intended [@caspiBikesharingPhiladelphiaLowerincome2019].
@@
 On the whole, existing studies highlight the need for bicycle share systems to be more highly accessible for diverse populations in order to increase use beyond the "typical" users.
-This has been the focus of recent research [among others, see @auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020].
+This has been the focus of recent research [see, among others, @auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020].
 Offering more people the option of using sustainable and active transportation, particularly those who have lower socioeconomic status and might benefit the most, is a worthy policy goal for cities with PBSPs.
-Several cities in North America have launched specific programs to address barriers to bike share or to expand service to more deprived areas: London, United Kingdom [@ogilvieInequalitiesUsagePublic2012]; Philadelphia, Pennsylvania [@caspiBikesharingPhiladelphiaLowerincome2019], and Hamilton, Ontario [@hosfordWhoArePublic2018].
+Several cities have launched specific programs to address barriers to bike share or to expand service to more deprived areas: London, United Kingdom [@ogilvieInequalitiesUsagePublic2012]; Philadelphia, Pennsylvania [@caspiBikesharingPhiladelphiaLowerincome2019], and Hamilton, Ontario [@hosfordWhoArePublic2018].
 Findings to date suggest that horizontal equity can be easier to achieve than vertical equity.
 However, exploring transportation equity by investigating where bicycle share stations are located, often using neighbourhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
-Meaning that, stations may be lacking in certain neighbourhoods but there may be stations accessible within a reasonable walking cutoff time.
-This is where accessibility becomes an important consideration.
+Meaning that, stations may be lacking in certain neighbourhoods but there may be stations accessible within a reasonable walking time.
+This is where geographical accessibility becomes an important consideration.
 
 Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved over the past century and has increasingly become linked with recent interest in prioritizing local proximity and modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020].
-As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequities in usage, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options.
+As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequities in the provision of infrastructure, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options.
 It also addresses some of the challenges of other performance measures such as level of service within a transportation network by measuring person-based indicators and exploring differences in use between population groups [@paezMeasuringAccessibilityPositive2012].
@@
 Beyond the utility derived from using shared bikes to destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019].
-Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., maximum number of bicycle racks available) of stations matters.
+Since the time or distance needed to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., maximum number of bicycle racks available) of stations matters.
 Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
 Kabra et al.
-[-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300m of stations.
-Furthermore, unlike other public amenities with greater ability to adapt to crowding, like health care services, the utility of stations is limited by the number of bicycles that they can hold, which makes crowding effects critical.
+[-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300 metres of stations, which amounts to 2-4 minutes walking by an able-bodied adult.
+Furthermore, similar to other public amenities affected by crowding, like health care services [e.g., @pereiraGeographicAccessCOVID192021], the utility of stations is limited by the number of bicycles that they can hold.
 The program may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles.
@@
 Thus, this model does a good job of considering potential crowding or competition for services.
-While there have been many methodological innovations in FCA methods [for example, @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balanced FCA approach that addressed the effects of demand and service inflation from previous FCA approaches [see @paezDemandLevelService2019].
+While there have been many methodological innovations in FCA methods [for example, @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balancing of the impedance that addressed the effects of demand and service inflation found in previous FCA approaches [see @paezDemandLevelService2019].
 
 When measuring accessibility, researchers have taken different approaches when it comes to aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones.
-Previous research on bike share equity has typically used a macro-level approach with aggregate data from entire neighbourhoods or census tracts  [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021].
+Previous research on bike share equity has typically used a meso- or macro-level approach with aggregate data from entire neighbourhoods or census tracts [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021].
 This is also true for studies examining correlates of bike share demand [@wangNewBikeShare2019].
@@
 
-Using the balanced floating catchment area method [@paezDemandLevelService2019], a novel approach that has not been used yet in cycling research, with disaggregate population-level data, we examine accessibility to a public bicycle share program in Hamilton, Ontario.
-We (1) conduct a sensitivity analysis by measuring accessibility and level of service to bike share stations at different thresholds of walking to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median household income of dissemination areas within the core service area.
+Using the balanced floating catchment area method [@paezDemandLevelService2019], a novel approach that has not been used yet in cycling research, with disaggregate population-level data, in this paper we examine accessibility to a public bicycle share program in Hamilton, Ontario.
+We (1) conduct a sensitivity analysis by measuring accessibility and level of service to bike share stations at different walking time thresholds to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median household income of dissemination areas within the core service area.
 
@@
 
-This paper uses the city of Hamilton, located in Ontario, Canada, as a case study.
-The city launched a public bicycle share program, Hamilton Bike Share, in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015].
+The focus of this study is the city of Hamilton, located in Ontario, Canada.
+The city launched a public bicycle share program in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015].
 Before June 2020, the program was known as Social Bicycles or SoBi Hamilton but is now called Hamilton Bike Share.
 Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021].
-The core service area spans 40 square kilometres of the city although it was planned to be 20 square kilometres [@hamiltonsobi2015], and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure \ref{fig:hamilton-and-sobi-service-area}].
-This represents roughly one fifth of the population in Hamilton according to the 2016 Canadian Census.
+The core service area spans 40 sq.km of the city although it was planned to be 20 sq.km [@hamiltonsobi2015], and roughly 138,000 people are within 30 minutes walking of a bike share station [see Figure \ref{fig:hamilton-and-sobi-service-area}].
+This represents roughly one fifth of the total population of the Hamilton Census Metropolitan Area according to the 2016 Canadian Census.
 The City of Hamilton undertook a large public engagement campaign to validate the locations of bicycle share stations that had been selected and to crowdsource potential locations for additional locations [@hamiltonsobi2014].
 Most of the suggested locations were in the east end of the core service area that lack stations or in neighbourhoods not serviced by the PBSP.
-Inadequate coverage given the size of the Hamilton Bike Share's service area was identified as a problem early on, with transportation planners noting that the small size (i.e., supply of bicycle racks) and low quantity of stations would lead to challenges in balancing the system [@hamiltonsobi2015].
+The program was enthusiastically welcomed in the city in 2015 - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015], however inadequate coverage given the size of the Hamilton Bike Share's service area was identified as a problem early on, and transportation planners noted that the small size (i.e., supply of bicycle racks) and low quantity of stations would lead to challenges in balancing the system [@hamiltonsobi2015].
 Based on public feedback, 13 stations were added to the original system.
-The program was enthusiastically welcomed in the city in 2015 - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015].
 
@@
 
-In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton.
+In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), with the objective of reducing barriers that may prevent individuals from accessing bike share in Hamilton.
 Additional bicycles and stations were added to the program which expanded it to more disadvantaged areas in the core service area [see Figure 2].
-The program also offers subsidized memberships to individuals who identify as low income as well as cycle skills education.
+The program also offers subsidized memberships to individuals who identify as low income, and complements these initiatives with cycle skills education.
 A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019].
@@
 As of June 2020, Hamilton Bike Share has 900 bikes, 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
-The core service area remains 40 square kilometres.
+The core service area remains 40 sq.km.
 The PBSP has 13 stations that were added as part of the ERI initiative; we refer to these as "equity stations" throughout the paper while all other stations are "conventional stations".
@@
 As a result of having a bike share membership, 49% of respondents report that they use their private vehicle less often or much less often but 48% report that their private vehicle use has remained about the same.
-This suggests that SoBi Hamilton has been useful for certain kinds of trips but not all, meaning that some trips continue to require a private vehicle.
+This suggests that Hamilton Bike Share has been useful for certain kinds of trips but not all, meaning that some trips continue to require a private vehicle.
 
@@
 Hosford and Winters [-@hosfordWhoArePublic2018] acknowledge that "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas".
-Therefore, the core service area for the PBSP in Hamilton is by default in more disadvantaged areas compared to other parts of the city.
+Therefore, the core service area for the PBSP in Hamilton by default covers some of the less advantaged areas in the city.
 However, there is also a great deal of variation in income in the city center because of the local university and increasing gentrification.
 Hosford and Winters [-@hosfordWhoArePublic2018] took a macro-level and system-wide approach in their analysis by using dissemination areas _across_ the city as the unit of analysis.
-They did not focus specifically on the core service area and did not differentiate between conventional and equity stations, which is worth exploring further.
-Therefore, this paper continues their analysis and makes an important contribution to the growing body of literature that examines equity issues in PBSPs.
-
-```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "The core service area of Hamilton Bike Share is outlined in black.
+They did not focus specifically on the core service area and did not differentiate between conventional and equity stations.
+
+```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "The core service area of Hamilton Bike Share is outlined in black.
 Hamilton Census Metropolitan Area (CMA) is shown in grey."}
@@
 Floating catchment area (FCA) methods are an approach commonly used in the healthcare accessibility literature.
-This approach is more appropriate and informative than calculating the provider-to-population ratio (PPR) which simply divides the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019].
-In particular, the Two-Step Floating Catchment Area (2SFCA) method [@luoMeasuresSpatialAccessibility2003; @radkeSpatialDecompositionsModeling2000] produces flexible catchment areas for populations accessing a service rather than using rigid boundaries like PPR, which provides more useful information because it doesn't assume that people stay within the pre-defined boundaries [@paezDemandLevelService2019].
+This approach is more appropriate and informative than calculating provider-to-population ratios (PPR) that simply divide the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019].
+In particular, the Two-Step Floating Catchment Area (2SFCA) method [@luoMeasuresSpatialAccessibility2003; @radkeSpatialDecompositionsModeling2000] produces flexible catchment areas instead of using rigid boundaries like PPR.
+This provides more useful information because it does not assume that people are limited to service within pre-defined boundaries [@paezDemandLevelService2019].
 This is an important property that supports our rationale for applying this method to measure accessibility to Hamilton Bike Share.
@@
 
-More recently, the _balanced_ floating catchment area (FCA) approach was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by past FCA methods [@paezDemandLevelService2019].
-Briefly, the overlapping of catchment areas leads to inflation of population totals and deflation of service levels across a study area and generates inaccurate or misleading accessibility estimates.
+More recently, the _balanced_ floating catchment area (FCA) approach was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by earlier FCA methods [@paezDemandLevelService2019].
+Briefly, overlapping catchment areas lead to inflation of population totals and deflation of service levels across a study area and generates inaccurate or misleading accessibility estimates.
 In contrast, Paez et al.
-[-@paezDemandLevelService2019] adjust the impedance weights so that both supply and demand are proportionally allocated which eliminates the over-counting of population that leads to distortions in demand and supply.
-Other benefits of this adjusted method include consideration of competition which can occur when catchment areas overlap, as well as the preservation of system-wide population and level of service.
+[-@paezDemandLevelService2019] adjust the impedance weights so that both supply and demand are proportionally allocated.
+The results is a FCA method that balances the population and level of service by eliminating the over-counting of population and level of service that leads to distortions in demand and supply.
+Other benefits of this adjusted method include consideration of competition which can occur when catchment areas overlap, as well as the preservation of population and level of service.
 Balanced floating catchment area methods have been used to explore accessibility to health care providers [@paezDemandLevelService2019] and COVID-19 health care services [@pereiraGeographicAccessCOVID192021], but have not yet been used in the cycling literature to explore issues of accessibility.
 
-The first step in the balanced floating catchment area method is to allocate the population to be serviced by each Hamilton Bike Share station:
+The first step in the floating catchment area method is to allocate the population to be serviced by each Hamilton Bike Share station:
 $$
@@
 $$
-
-Next, the level of service at each station (i.e., the maximum number of bicycle racks) is divided by its estimated service population within the established catchment area:
+As seen in the equation above, the population allocated to station $j$ is the weighted sum of the population in the region; a spatial weight $w_{ij}$ represents the friction that the population at $i$ faces when reaching station $j$, and is usually given by a distance-decay function, so that each station is assumed to service only a segment of the population within a limited geographical range.
+
+Next, the supply at each station (i.e., the maximum number of bicycle racks) is divided by its estimated service population within the established catchment area; this gives the level of service of station $j$ in bicycle racks per person:
 $$
@@
 
-Finally, the accessibility of population cell $i$ is calculated as the weighted sum of the level of service of all stations that can be reached from there according to normalized weights:
+Finally, the accessibility of population cell $i$ is calculated as the weighted sum of the level of service of all stations that can be reached from there according to the spatial weights:
 $$
@@
 
-The approach uses instead a set of suitably normalized weights as follows:
+The balanced approach of @paezDemandLevelService2019 replaces the spatial weights with normalized versions as follows:
 $$
@@
 
-and
+\noindent and:
 $$
@@
 
-Finally, accessibility is calculated without risk of demand or supply inflation:
+With these weights accessibility can be calculated without risk of demand or supply inflation:
 $$
@@
 $$
-
+By allocating the population and level of service proportionally, this method preserves the values of the population and level of service and provides a regional provider-to-population ratio since:
+$$
+\sum_{i=1}^n A_i = \sum_{j=1}^J L_j = \frac{\sum_{j=1}^JS_j}{\sum_{i=1}^n P_i}
+$$
+
+In fact, since the proportional allocation procedure means that any proportion of the population allocated to a station is never allocated to other stations, and conversely any level of service allocated to a population is never re-allocated elsewhere, this property is replicated for any level of aggregation.
 For this paper, we employ a hybrid location-based and person-based approach to calculating accessibility using disaggregate population data.
 In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data.
-This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are more capable of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences.
-Applying the balanced floating catchment area approach to allows us to examine accessibility by stratifying according to median household income, which would constitute the individual component of the accessibility measure [@geursAccessibilityEvaluationLanduse2004].
+This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are  better able of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences.
+Applying the balanced floating catchment area approach allows us to examine accessibility by stratifying according to median household income, which would constitute the individual component of the accessibility measure [@geursAccessibilityEvaluationLanduse2004].
 However, conducting a further sensitivity analysis to measure accessibility at different distance thresholds would help to consider potential spatio-temporal constraints.
@@
 
-Following the method first presented by Tobler [-@toblerSmoothPycnophylacticInterpolation1979], we used pycnophylactic interpolation to disaggregate population from each dissemination area to smaller polygons that are 50 by 50 metres.
-This method involves smoothing out the population from each dissemination area while preserving total volume [see Figure \ref{fig:da-population}].
-Since pycnophylactic interpolation occurs at such a micro scale, we had to ensure that population numbers were not reallocated or smoothed out to areas where people do not live in Hamilton (i.e., where schools or parks are located).
-To do so, we retrieved shapefiles for various geographic features [see Table \ref{tab:data-features}] from Open Hamilton.
+To obtain population at sub-census geography levels (at the micro scale) we use pycnophylactic interpolation [@toblerSmoothPycnophylacticInterpolation1979].
+We obtained population data from the 2016 Census of Canada for dissemination areas (DAs, the smallest publicly available census geography in Canada).
+This zonal values of the population were interpolated to smaller polygons 50-by-50 m in size.
+Pycnophilactic interpolation involves smoothing out the population from each dissemination area while preserving total volume [see Figure \ref{fig:da-population}].
+When interpolating the population at this high level of resolution it is important to ensure that population numbers were not allocated to areas where people do not live in Hamilton (e.g., to parks, large institutional buildings, etc.) To do so, we retrieved shapefiles for various geographic features [see Table \ref{tab:data-features}] from Open Hamilton.
 Next, we removed these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population within the remaining area [see Figure \ref{fig:interpolated-population}].
@@
 
-```{r da-population, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Population in all dissemination areas (outlined in white) in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the service area."}
+```{r da-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Population in all dissemination areas (outlined in white) that are inside or touch the bounding box of the Hamilton Bike Share core service area (outlined in black)."}
 ggplot() +
@@
 
-```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
-The downtown area is outlined in dark green.
-Geographic features have been extracted.
-Dissemination area polygons are outlined in grey."}
+```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
+Each micro population cell is 50-by-50 m in size.
+The downtown area is outlined in dark green."}
 ggplot() + 
@@
 
-[BBBike](https://download.bbbike.org/osm/bbbike/) is an online cycle route planner that interfaces with OpenStreetMap.
-We extracted OpenStreetMap data for SoBi Hamilton's service area to calculate walking times from each population cell to nearby bike share stations, using a walking distance of 10km and a walking time of 30 minutes as a threshold.
+To calculate walking times from the centroid of our micro population cells we extracted OpenStreetMap data for Hamilton Bike Share's service area from [BBBike](https://download.bbbike.org/osm/bbbike/), an online cycle route planner that interfaces with OpenStreetMap.
+OpenStreetMap data provides the networks for calculate walking times from each population cell to nearby bike share stations, using a maximum walking distance of 10 km and walking time of 30 minutes as thresholds.
 A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bike share stations within the maximum threshold.
-This provides a more realistic measure of the temporal component by taking infrastructure into account of travel times, rather than using the Euclidean distance from population cell to station.
+This process provides a more realistic measure of the friction of reaching bike share stations by taking infrastructure into account of travel times, rather than using the Euclidean distance from population cell to station.
+Routing and travel time calculations were completed using the `R` package `r5r`, used for rapid realistic routing operations [@Pereira2021r5r].
 
@@
 
-All data for this research were accessed from Open Hamilton^[https://open.hamilton.ca/], an online and public repository of data curated by the City of Hamilton.
+All data for this research were accessed from publicly available Census of Canada sources, from OpenStreetMaps, and from Open Hamilton^[https://open.hamilton.ca/], a public online repository of data curated by the City of Hamilton.
 
@@
 
-Consensus regarding the distance that individuals are willing to walk to access a PBSP is lacking, but the literature on walking behaviour was consulted to determine the thresholds for the sensitivity analysis.
+Consensus regarding the distance that individuals are willing to walk to access a PBSP station is lacking, but the literature on walking behaviour can provide some guidelines to determine the thresholds for sensitivity analysis.
 Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010].
@@
 In the present case, we found that congested accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at 5 minutes.
-Accessibility decreases significantly after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each bike share station.
+Accessibility decreases substantially after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each bike share station.
 
@@
 At each threshold, we compare accessibility between the current system and the original system to examine the contribution of the additional equity stations.
+When considering the results reported below, it is important to remember that accessibility is technically a form of smoothing [@okelly2003aggregate, pp.
+7-8]: smaller thresholds produce less smoothing (which can result in "spiky" accessibility landscapes), and larger thresholds produce more smoothing, and fewer spikes.
 
@@
 
-With a walking distance of three minutes, we find that there are 25.2 bicycle racks per person in the original system.
+With a walking distance of three minutes, we find that there are 25.2 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
 The addition of equity stations increases this ratio slightly to 25.4 bicycles per person.
@@
 Accessibility is fairly uniform overall, with the exception of two small areas where accessibility is slightly higher.
-
- ```{r figure-6, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
+This high level of system-wide accessibility occurs because the population that can reach bike share stations when travel time is 3 minutes or less is very limited, and accessibility is strongly shaped by a few locations that concentrate population and stations.
+
+ ```{r figure-6, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
 ggplot() +
@@
 With a walking distance of five minutes, we find that there are 68.6 bicycle racks per person in the original system.
-With the addition of equity stations, there are now 68.8 bicycles per person.
+With the addition of equity stations, there are now 68.8 bicycle racks per person.
+At this threshold, there are more bicycle racks per person than at the minimum threshold.
+System-wide accessibility has in fact increased: the population that can reach the stations has grown, but not to the the point that congestion effects begin to take place.
 Figure \ref{fig:figure-7} presents a comparison of accessibility between the systems.
 Again, accessibility is fairly uniform, with the exception of one very small area.
-At this threshold, there are more bicycle racks per person than at the minimum threshold.
-This occurs because fewer stations can be reached when the travel time is 3 minutes or less.
-
-```{r figure-7, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 5 minutes walk (average threshold) compared between current system with equity stations and the original system without equity stations."}
+
+```{r figure-7, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 5 minutes walk (average threshold) compared between current system with equity stations and the original system without equity stations."}
 #accessibility plot for current system with equity stations at 5 minutes walk
@@
 Figure \ref{fig:figure-8} presents a comparison of accessibility between the systems.
-We begin to see differences in accessibility across the service area, with users near the university and its adjacent neighbourhoods, as well as neighbourhoods north of the downtown area (outlined in green), having slightly higher accessibility.
+Differences in accessibility across the service area are now apparent, with users near the university and its adjacent neighbourhoods, as well as neighbourhoods north of the downtown area (outlined in green), having slightly higher accessibility.
 While the differences are modest, they are more apparent at this threshold than at shorter walking distances.
 
-```{r figure-8, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) compared between current system with equity stations and the original system without equity stations."}
+```{r figure-8, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) compared between current system with equity stations and the original system without equity stations."}
 #accessibility plot for current system with equity stations at 10 minutes walk
@@
 
-```{r figure-9, echo=FALSE, fig.align = 'center', out.width = "65%", fig.cap = "Accessibility at 15 minutes walk (extreme threshold) compared between current system with equity stations and the original system without equity stations."}
+```{r figure-9, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 15 minutes walk (extreme threshold) compared between current system with equity stations and the original system without equity stations."}
 #accessibility plot for current system with equity stations at 15 minutes walk
@@
 One of the unique properties of the balanced floating catchment area method is that data can be reaggregated while preserving the total population in the service area and the supply at each station.
-This avoids demand and supply inflation, and also enables us to present findings in a way that is easier and more intuitive to interpret.
-Therefore, we reaggregate population and accessibility from the smallest 50 by 50 metres population cells to the median household income for each dissemination area.
-We divide income by quintiles: bottom 20%, second 20%, third 20%, fourth 20%, and top 20%.
-
-Panels 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combined spatial distribution of accessibility and median household income at different thresholds.
-Our analysis demonstrates that the extreme threshold of fifteen minutes allows the most people to access a station.
-We find that stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing access, albeit only modestly [see Table \ref{tab:accessibility-income}].
+This avoids demand and supply inflation, and also enables us to present findings in a way that is perhaps more intuitive to interpret.
+Therefore, we reaggregate population and accessibility from the 50-by-50 m population cells to the dissemination area.
+We then can compare accessibility to median household income statistics from the census.
+
+Panels 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combine the spatial distribution of accessibility and median household income, using tertiles for the coloring scheme.
+As expected, the analysis indicates that the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service areas of stations.
+We find that stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing access
+
+, albeit only modestly [see Table \ref{tab:accessibility-income}].
 By implementing equity stations in more areas with lower median total income in Hamilton, the PBSP has achieved greater horizontal equity by extending the spatial distribution of bicycles across the city.
@@
 
-```{r figure-10, echo=FALSE, fig.align = 'center', out.width = "65%"}
+```{r figure-10, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-11, echo=FALSE, fig.align = 'center', out.width = "65%"}
+```{r figure-11, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-12, echo=FALSE, fig.align = 'center', out.width = "65%"}
+```{r figure-12, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-13, echo=FALSE, fig.align = 'center', out.width = "65%"}
+```{r figure-13, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-14, echo=FALSE, fig.align = 'center', out.width = "65%"}
+```{r figure-14, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-15, echo=FALSE, fig.align = 'center', out.width = "65%"}
+```{r figure-15, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-16, echo=FALSE, fig.align = 'center', out.width = "65%"}
+```{r figure-16, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-17, echo=FALSE, fig.align = 'center', out.width = "65%"}
+```{r figure-17, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 The City of Hamilton has recognized from the launch of the PBSP that significantly more stations and bicycles are needed to service the area [@hamiltonsobi2015].
-With a service area of 40 square kilometres, Hamilton should have between 380 and 440 stations, instead of 130, and 1,500 bicycles instead of 900 [@hamiltonsobi2015].
+With a service area of 40 sq.km, Hamilton should have between 380 and 440 stations, instead of 130, and 1,500 bicycles instead of 900 [@hamiltonsobi2015].
 Reduced capacity within the system leads to gaps in coverage in some areas of the city "with some areas not having the recommended station density of 300m between stations or 10 stations per square km" [@hamiltonsobi2015].

### 2021-03-28 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Proof-read and edited first half of paper in preparation for submission

@@
             pdfauthor={},
-            pdftitle={Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach},
+            pdftitle={Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach},
             colorlinks=false,
@@
 
-  \title{Examining spatial equity and accessibility to a public bicycle share
+  \title{Examining spatial equity in accessibility to a public bicycle share
 program using a balanced floating catchment area approach}
@@
   \begin{abstract}
-  Public bicycle share programs are implemented to promote cycling as a
-  convenient and sustainable mode of transportation. These systems can
-  also play a role in addressing transportation needs and advancing
+  Public bicycle share programs (PBSPs) are implemented to promote cycling
+  as a convenient and sustainable mode of transportation. These systems
+  can also play a role in addressing transportation needs and advancing
   transportation equity if they make cycling more accessible to lower
-  income or disadvantaged populations. Recent studies assessing equity in
-  bike sharing have found differences in ridership or membership based on
-  income, education, and trip purpose. In Ontario, the City of Hamilton
+  income or disadvantaged populations. In Ontario, the City of Hamilton
   launched a public bicycle share program in 2015 that currently has over
-  900 operational bicycles and 130 docking stations. In 2018, the
-  non-profit organization that operates the program launched an equity
-  initiative to provide subsidized memberships and to expand service by
-  adding twelve docking stations. Previous research found that Hamilton's
-  public bicycle share program targeted well disadvantaged areas in the
-  city compared to programs in other Canadian cities. Since the time or
-  distance that members of the population need to reach a bicycle share
-  station decreases the potential of accessing the program, the location
-  of stations matters. Unlike other public amenities with greater ability
-  to adapt to crowding, the utility of stations is limited by the number
-  of bicycles that they can hold, which makes crowding effects critical;
-  the system may not necessarily be improved if stations are easy to reach
-  but offer only a small number of bicycles. In this research, we revisit
-  the case of Hamilton and investigate equity differentials in
-  accessibility to bicycle share stations. We compare accessibility in the
-  program with and without the equity stations to assess the effect of the
-  initiative. Previous research on cycling equity or accessibility has
-  used a macro-level approach with entire neighbourhoods or census tracts
-  as the unit of analysis, although there are recent exceptions. In
-  contrast, we implement our analysis parting from microscopic zones to
-  better reflect travel to a docking station, which often happens at the
-  sub-neighborhood level, and conduct a sensitivity analysis at several
-  walking thresholds. We then reaggregate the estimated accessibility to
-  dissemination areas for further analysis using census data. Equity
-  analysis of this type is possible thanks to a newly developed approach
-  for accessibility using balanced floating catchment areas. The addition
-  of equity stations achieved its goal to increase accessibility for lower
-  income populations, although the increase was relatively modest (2.10
-  bicycles/person with equity stations vs 1.97 bicycles/person without
-  equity stations). Further research is needed to determine whether this
-  encouraged more cycling.
+  900 operational bicycles and 130 docking stations. This system was
+  expanded in 2018 by an equity initiative that added twelve docking
+  stations with the explicit objective of increasing spatial equity in
+  access. Since the cost of reaching a bicycle share station decreases the
+  potential of accessing the program, the location of stations matters,
+  and previous research found that Hamilton's public bicycle share program
+  targeted well disadvantaged areas in the city. In this research, we
+  revisit the case of Hamilton and investigate differentials in
+  accessibility to bicycle share stations using a balanced floating
+  catchment area (BFCA) accessibility approach. We compare accessibility
+  with and without the equity stations to assess the effect of the
+  initiative. We implement our analysis parting from micro zones to better
+  reflect walking to a docking station, and conduct a sensitivity analysis
+  at several walking time thresholds. We then reaggregate the estimated
+  accessibility for further analysis using census data. The addition of
+  equity stations increased the serviced population at every threshold
+  examined, and although accessibility increased for the whole population,
+  this increase was relatively modest especially for population in the
+  bottom 20\% of median household income.
   \end{abstract}
@@
 
-\hypertarget{background}{%
-\section{Background}\label{background}}
-
-Public bicycle share programs (PBSP) have been implemented in over 800
-cities worldwide and a great deal has been learned about their typical
-users (Fishman, 2016). In many cities, males use bike share more than
-females (Brey et al., 2017; Nickkar et al., 2019; Reilly, S. M. Wang, et
-al., 2020; Winters et al., 2019) as do younger age cohorts (Brey et al.,
-2017; Buck et al., 2013; Fuller et al., 2011). However, one study found
-that bike share users in Washington, DC were more likely to be female
-(Buck et al., 2013), which suggests that the gender gap among cyclists
-who use bike share is less disparate than the gap for personal bicycle
-use (Fishman, 2016). There is some evidence that bike share users are
-less likely to own a car (Buck et al., 2013; Reilly, Noyes, et al.,
-2020). However, the relationship between income or education and bike
-share use is less clear. Stations in disadvantaged communities in
-Chicago generated most of the average annual trips (Qian and Jaller,
-2020) and individuals from minority or lower socioeconomic status
-neighbourhoods in Minneapolis-St.~Paul used the city's PBSP more (J.
-Wang and Lindsey, 2019a). Being university educated was also a
-significant correlate of bike share use in Montreal, Canada (Fuller et
-al., 2011). On the other hand, members who reside in
-minority-concentrated and lower socioeconomic status neighbourhoods use
-the Minneapolis-St Paul bike share more frequently (J. Wang and Lindsey,
-2019a). Financial savings have been found to motivate those on a low
-income to use bike share (Fishman, 2016). Many studies have found that
-proximity to a bike share station, either living or working near one, is
-an important determinant of use or having a membership (Fishman, 2016;
-Fuller et al., 2011). This makes sense given that individuals are more
-likely to use services or programs that they can easily access. Finally,
-the potential of bike share programs to increase cycling levels has also
-been explored recently (Hosford et al., 2018, 2019), but more
-longitudinal evidence is needed to determine their impact on encouraging
-more cycling and offering more opportunities for physical activity. On
-the whole, these findings highlight the need for bike share systems to
-be more highly accessible for diverse populations in order to increase
-use beyond the ``typical'' users, which has been the focus of recent
-research (Auchincloss et al., 2020; Hull Grasso et al., 2020; MacArthur
-et al., 2020), and to offer more people the option of using sustainable
-and active transportation.
-
-Several studies have recently explored the equity of PBSP in North
-American cities by primarily examining who has access to bike share
-(e.g., differences by demographics or socioeconomic status) and where
-stations are located. Equity can be achieved in two different ways:
+\newpage
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+The potential of public bicycle share programs (PBSPs) to increase
+cycling levels is but one of many reasons for implementing such programs
+in urban areas (Hosford et al., 2018, 2019). As a healthy, inexpensive,
+and convenient form of public transportation, shared bicycles can
+encourage individuals to take up cycling for short local trips or first
+and last mile trips to other public transportation instead of using
+personal vehicles. These systems can also play a role in addressing
+transportation needs and advancing transportation equity if they make
+cycling more accessible to lower income or disadvantaged populations. As
+these systems become increasingly common, their introduction has been
+accompanied by a flurry of research that investigates them from the
+perspective of spatial and transportation equity (e.g., Smith et al.,
+2015; Hosford and Winters, 2018; Hull Grasso et al., 2020; Mooney et
+al., 2019; Qian and Jaller, 2020; Qian et al., 2020). Indeed, although
+these programs are available to the general public and ought to be
+accessible to any individuals who wish to use them, research on PBSPs
+located in North American and European cities indicates that there are
+inequities in spatial accessibility in need of redress.
+
+In this paper we investigate the case of the PBSP in the City of
+Hamilton, in Ontario, Canada. Hamilton Bike Share (previously also known
+as Social Bicycles, or SoBi) was launched in 2015 and currently has over
+900 operational bicycles and 130 docking stations. An interesting
+feature of this system is that it was expanded by an equity initiative
+that introduced twelve docking stations in 2018 with the explicit
+objective of increasing spatial accessibility. Hamilton Bike Share was
+already studied by Hosford and Winters (2018) as part of a selection of
+PBSPs in Canada. These authors found that most of the cities in their
+research could benefit of greater efforts to expand service to areas
+with lower socio-economic populations - but Hamilton fared somewhat
+better than other cities in this respect. What distinguishes our
+research is the use of a balanced floating catchment area (BFCA)
+accessibility approach that accounts for the supply of stations and the
+potential demand from the population serviced. This requires a more
+disaggregated approach than the use of census geographies (q.v., Hosford
+and Winters, 2018), since many trips to a bike share station likely
+happen at a level lower than even the smallest census geography. For
+this reason, we implement our analysis parting from micro population
+zones to better reflect the friction of walking to a docking station.
+Further, we conduct a sensitivity analysis at several walking time
+thresholds. In terms of spatial equity, we compare accessibility with
+and without the equity stations to assess the effect of the initiative,
+before reaggregating the data for further analysis using median
+household income information from the census. We find that the addition
+of equity stations does increase the serviced population at every
+threshold examined. However, although accessibility increased for the
+whole population, this increase was relatively modest, especially for
+population in the bottom 20\% of median household income. The analysis
+also identifies areas under-serviced within the core service area of the
+system.
+
+This paper is an example of open and reproducible research that uses
+only open software for transportation and statistical analysis (Bivand,
+2020; Lovelace, 2021). All data were obtained from publicly available
+sources and organized in the form a data package. Following best
+practices in spatial data science (Brunsdon and Comber, 2020), the code
+and data needed to reproduce (or modify/extend) the analysis are
+available for download.\footnote{https://drive.google.com/drive/folders/1ZGRpSN2LxS2Fq2tLVJgfm1sI6I26vzZh?usp=sharing}
+
+\hypertarget{literature-review}{%
+\section{Literature Review}\label{literature-review}}
+
+Public bicycle share programs have been implemented in over 800 cities
+worldwide and a great deal has been learned about their typical users
+(Fishman, 2016). In many cities, males use bike share more than females
+({\textbf{???}}; Brey et al., 2017; Nickkar et al., 2019; Reilly, S. M.
+Wang, et al., 2020; Winters et al., 2019) as do younger age cohorts
+(Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011). However,
+one study found that bike share users in Washington, DC were more likely
+to be female (Buck et al., 2013), which suggests that the gender gap
+among cyclists who use bike share is less disparate than the gap for
+personal bicycle use (Fishman, 2016). There is some evidence that bike
+share users are less likely to own a car (Buck et al., 2013; Reilly,
+Noyes, et al., 2020). However, the relationship between income or
+education and bike share use is less clear-cut. Stations in
+disadvantaged communities in Chicago have been found to generate most of
+the average annual trips (Qian and Jaller, 2020) and individuals from
+minority or lower socioeconomic status neighbourhoods in
+Minneapolis-St.~Paul used the city's PBSP more (J. Wang and Lindsey,
+2019a). Similar findings were reported in London ({\textbf{???}}). Being
+university educated was a significant correlate of bike share use in
+Montreal, Canada (Fuller et al., 2011). Perhaps not coincidentally,
+financial savings have been found to motivate those on a low income to
+use bike share (Fishman, 2016).
+
+Many studies have found that geographic proximity to a bicycle share
+station is an important determinant of membership and use (Fishman,
+2016; Fuller et al., 2011). This makes sense given that individuals are
+more likely to use services or programs that they can easily reach.
+Several studies have recently explored the equity of PBSPs by primarily
+examining who has access to bike share (e.g., differences by
+demographics or socioeconomic status) and where stations are located. It
+is important to note that equity can be achieved in two different ways:
 equal spatial distribution across a region (e.g., horizontal equity) or
@@
 a significantly lower share of trips compared to other areas in the
-city, they make longer trips. This suggests that they may be using PBSP
-for work commuting, which points to the importance of ensuring equitable
-access (Qian and Jaller, 2020). Similar results were found in
-Philadelphia, where lower income areas generate fewer trips and efforts
-to increase equity within the program have not been as successful as
-intended (Caspi and Noland, 2019). In the case of Seattle, all
-neighbourhoods had some level of access to dockless bikes but those with
-higher incomes and more residents of higher education had more bikes
-(Mooney et al., 2019). Babagoli et al.~(2019) found that neighbourhoods
-in New York City with higher affluence had the greater proportion of
-Citi Bike stations. Overall, these findings suggest that horizontal
-equity can be achieved while vertical equity is harder to attain.
-Exploring transport equity by investigating where bicycle share stations
-are located, often using neighbourhoods or census tracts as the spatial
-unit of analysis, can ignore or miss the benefits that may be derived
-from adjacent zones. Meaning that, stations may be lacking in certain
-neighbourhoods but there may be stations accessible within a reasonable
-walking cutoff time. This is where accessibility becomes an important
-consideration.
+city, they make longer trips. This suggests that they may be using PBSPs
+for utilitarian trips (e.g., commuting), which points to the importance
+of ensuring equitable access (Qian and Jaller, 2020). Similar results
+were found in Philadelphia, where lower income areas generated fewer
+trips and it was suggested that efforts to increase equity within the
+program have not been as successful as intended (Caspi and Noland,
+2019). In the case of Seattle, all neighbourhoods had some level of
+access to dockless bikes but those with higher incomes and more
+residents of higher education had more bikes (Mooney et al., 2019).
+Babagoli et al.~(2019) also found that neighbourhoods in New York City
+with higher affluence had the greatest proportion of Citi Bike stations.
+Researchers have recommended targeted expansion of stations to areas
+that use bike share but have inequitable access ({\textbf{???}}).
+
+On the whole, existing studies highlight the need for bicycle share
+systems to be more highly accessible for diverse populations in order to
+increase use beyond the ``typical'' users. This has been the focus of
+recent research (see, among others, Auchincloss et al., 2020; Hull
+Grasso et al., 2020; MacArthur et al., 2020). Offering more people the
+option of using sustainable and active transportation, particularly
+those who have lower socioeconomic status and might benefit the most, is
+a worthy policy goal for cities with PBSPs. Several cities have launched
+specific programs to address barriers to bike share or to expand service
+to more deprived areas: London, United Kingdom ({\textbf{???}});
+Philadelphia, Pennsylvania (Caspi and Noland, 2019), and Hamilton,
+Ontario (Hosford and Winters, 2018). Findings to date suggest that
+horizontal equity can be easier to achieve than vertical equity.
+However, exploring transportation equity by investigating where bicycle
+share stations are located, often using neighbourhoods or census tracts
+as the geographical unit of analysis, can ignore or miss the benefits
+that may be derived from adjacent zones. Meaning that, stations may be
+lacking in certain neighbourhoods but there may be stations accessible
+within a reasonable walking time. This is where geographical
+accessibility becomes an important consideration.
 
@@
 inform transportation planning (Páez et al., 2012), but its utility to
-this field has evolved since its conceptualization and become linked
-with recent interest in prioritizing local proximity and modes that are
-suitable for local trips like walking and cycling (Levine, 2020). As a
-measure of the ease of reaching potential destinations spread spatially
-in a given area, accessibility is relevant to PBSPs because it can
-identify current inequities in usage, as well as guide interventions
-that increase access for groups that are under-serviced or address gaps
-in transportation options. It also addresses some of the challenges of
-other performance measures such as level of service within a
-transportation network by measuring person-based indicators and
-exploring differences in use between population groups (Páez et al.,
-2012).
-
-In the context of PBSPs, accessibility refers to the distance an
-individual must travel to reach a bicycle share station (Kabra et al.,
-2020; J. Wang and Lindsey, 2019b). Since the time or distance that
-members of the population need to reach a bicycle share station
-decreases the potential of accessing the program and ultimately use of
-the program, the location and size (e.g., maximum number of bicycles
-available) of stations matters. Indeed, distance to bicycle share
-stations is associated with use (Fuller et al., 2011; J. Wang and
-Lindsey, 2019b) and can be a barrier to using PBSPs (Fishman et al.,
-2014). Kabra et al.~(2020) found that the majority of bike share usage
-in Paris comes from areas within 300m of stations. Furthermore, unlike
-other public amenities with greater ability to adapt to crowding, like
-health care services, the utility of stations is limited by the number
-of bicycles that they can hold, which makes crowding effects critical.
-The program may not necessarily be improved if stations are easy to
-reach but offer only a small number of bicycles. Likewise, more people
-may not opt to use the program if the supply of bicycles available at
-the nearest station is insufficient to meet demand. The location and
-size of stations is important to increase the utility of this public
+this field has evolved over the past century and has increasingly become
+linked with recent interest in prioritizing local proximity and modes
+that are suitable for local trips like walking and cycling (Levine,
+2020). As a measure of the ease of reaching potential destinations
+spread spatially in a given area, accessibility is relevant to PBSPs
+because it can identify current inequities in the provision of
+infrastructure, as well as guide interventions that increase access for
+groups that are under-serviced or address gaps in transportation
+options. It also addresses some of the challenges of other performance
+measures such as level of service within a transportation network by
+measuring person-based indicators and exploring differences in use
+between population groups (Páez et al., 2012).
+
+Beyond the utility derived from using shared bikes to destinations of
+value, an important aspect of accessibility associated with PBSPs is the
+distance an individual must travel to reach a bicycle share station
+(Kabra et al., 2020; J. Wang and Lindsey, 2019b). Since the time or
+distance needed to reach a bicycle share station decreases the potential
+of accessing the program and ultimately their use of the program, the
+location and size (e.g., maximum number of bicycle racks available) of
+stations matters. Indeed, distance to bicycle share stations is
+associated with use (Fuller et al., 2011; J. Wang and Lindsey, 2019b)
+and can be a barrier to using PBSPs (Fishman et al., 2014). Kabra et
+al.~(2020) found that the majority of bike share usage in Paris comes
+from areas within 300 metres of stations, which amounts to 2-4 minutes
+walking by an able-bodied adult. Furthermore, similar to other public
+amenities affected by crowding, like health care services (e.g.,
+Pereira, Carlos Kauê Vieira Braga, et al., 2021), the utility of
+stations is limited by the number of bicycles that they can hold. The
+program may not necessarily be improved if stations are easy to reach
+but offer only a small number of bicycles. Likewise, more people may not
+opt to use the program if the supply of bicycles available at the
+nearest station is insufficient to meet demand. The location and size of
+stations is important to increase the utility of this public
 transportation option for more people, thus achieving vertical equity.
@@
 
-Several approaches have been commonly used for measuring accessibility:
-cumulative opportunities, gravity, and utility-based (Handy and
-Niemeier, 1997). Paez et al (2012) provide a recent overview of various
+Several approaches have been commonly used for measuring place-based
+accessibility, including cumulative opportunities, gravity, and
+utility-based measures (Handy and Niemeier, 1997). Geurs and Van Wee
+(2004) and Paez et al.~(2012) provide recent overviews of various
 formulations and applications of accessibility in transportation
-research. Briefly, the gravity-based approach involves weighting
-opportunities, for example the quantity of bicycle share stations, as
-measured by a function of time, using a negative exponential form (Handy
-and Niemeier, 1997). It takes both demand and congestion into account
-(Páez et al., 2012). Floating catchment area (FCA) methods are a type of
-gravity measure that have been used in healthcare accessibility research
-but that are applicable to transportation case studies. This approach
-involves producing flexible catchment areas for populations, recognizing
-that people may be willing to travel to access particular services (Páez
-et al., 2012). Thus, this model does a good job of considering potential
-crowding of services using a binary distance function (Páez et al.,
-2012). A recent improvement to this approach was achieved through a
-simple and intuitive adjustment to the allocation of supply and demand,
-which addressed the effects of demand and service inflation from the
-conventional approach (see Paez et al., 2019).
-
-Researchers have also taken different approaches when it comes to
-aggregation of data, either by using the individual or household as the
-smallest unit of analysis or larger spatial zones. Previous research on
-bike share equity has typically used a macro-level approach with
-aggregate data from entire neighbourhoods or census tracts (Babagoli et
-al., 2019; Mooney et al., 2019; Qian and Jaller, 2020; J. Wang and
-Lindsey, 2019a), although there are recent exceptions (Chen et al.,
-2019; Chen and Li, 2021). This is also true for studies examining
-correlates of bike share demand (J. Wang and Lindsey, 2019b). Handy and
-Niemeier (1997) note that using disaggregated data in accessibility
-analyses provides a more accurate estimate for individuals, which is
-useful for addressing vertical inequities in PBSP usage.
-
-Using balanced floating catchment area methods, a novel approach that
-has not been used yet in cycling research, we examine accessibility to a
-public bicycle share program in Hamilton, Ontario. Using disaggregate
-population-level data, we (1) conduct a sensitivity analysis by
-measuring accessibility and level of service to bike share stations at
-different thresholds of walking to reach a station: 3 minutes, 5
-minutes, 10 minutes, and 15 minutes; (2) explore the contribution of
-specific stations that were added to Hamilton's PBSP to reducing both
-horizontal and vertical inequities; and (3) examine whether disparities
-in accessibility exist according to median household income of
-dissemination areas within the core service area.
+research. The common gravity-based approach for example involves
+weighting destination opportunities, such as the quantity of bicycle
+share stations, by the time required to reach them from an origin using
+an impedance function (Handy and Niemeier, 1997; Kwan, 1998). However,
+while such measures are suitable for capturing the potential for
+reaching destinations from a given location, they do not take demand or
+congestion effects into account.
+
+In contrast, floating catchment area (FCA) methods incorporate
+information on capacity and demand in calculating accessibility. FCA
+measures have been widely employed in healthcare accessibility research
+and are typically calculated across two steps. In the first, a ratio of
+supply to demand at service locations is calculated, such as the number
+of beds at a hospital divided by the number of people within the
+catchment area of the hospital, weighted by the distance involved in
+reaching the facility. Next, these service level ratios are allocated
+back to the population centres and summarized as a measure of congested
+accessibility. Thus, this model does a good job of considering potential
+crowding or competition for services. While there have been many
+methodological innovations in FCA methods (for example, {\textbf{???}};
+{\textbf{???}}; {\textbf{???}}; {\textbf{???}}; {\textbf{???}}), a
+recent improvement to this approach was achieved through a simple and
+intuitive balancing of the impedance that addressed the effects of
+demand and service inflation found in previous FCA approaches (see Paez
+et al., 2019).
+
+When measuring accessibility, researchers have taken different
+approaches when it comes to aggregation of data, either by using the
+individual or household as the smallest unit of analysis or larger
+spatial zones. Previous research on bike share equity has typically used
+a meso- or macro-level approach with aggregate data from entire
+neighbourhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
+2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
+are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
+also true for studies examining correlates of bike share demand (J. Wang
+and Lindsey, 2019b). Handy and Niemeier (1997) note that using
+disaggregated data in accessibility analyses provides a more accurate
+estimate for individuals, which is useful for addressing vertical
+inequities in PBSP usage.
+
+Using the balanced floating catchment area method (Paez et al., 2019), a
+novel approach that has not been used yet in cycling research, with
+disaggregate population-level data, in this paper we examine
+accessibility to a public bicycle share program in Hamilton, Ontario. We
+(1) conduct a sensitivity analysis by measuring accessibility and level
+of service to bike share stations at different walking time thresholds
+to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes;
+(2) explore the contribution of specific stations that were added to
+Hamilton's PBSP to reducing both horizontal and vertical inequities; and
+(3) examine whether disparities in accessibility exist according to
+median household income of dissemination areas within the core service
+area.
 
@@
 
-This paper uses the city of Hamilton, located in Ontario, Canada, as a
-case study. The city launched a public bicycle share program, Hamilton
-Bike Share, in March 2015 with 115 stations and 750 bicycles (Hamilton,
-2015). Before June 2020, the program was known as SoBi Hamilton.
-Stations are spaced between 300 and 600 metres apart (Scott et al.,
-2021). The core service area spans 35 km\^{}2 of the city and roughly
-138,000 can reach a bike share hub within 30 minutes of walking {[}see
-Figure \ref{fig:hamilton-and-sobi-service-area}{]}. This represents
-roughly one fifth of the population in Hamilton, according to the 2016
-Canadian Census. The program was enthusiastically welcomed in the city -
-within three weeks of launching, 10,000 trips had been made (Hamilton,
-2015). In 2017, Hamilton Bike Share Inc., the non-profit organization
-that operates the program, initiated an equity program, Everyone Rides
-Initiative (ERI), to remove barriers that may prevent individuals from
-accessing bike share in Hamilton. An additional 75 bicycles and 15
-stations were added to the program, which expanded it to more
-disadvantaged areas in the core service area {[}see Figure 2{]}. The
-program also offers subsidized memberships to individuals who identify
-as low income. A comparable program can be found in Philadelphia (see
-Caspi and Noland, 2019). As of June 2020, the bike share program has 900
-bikes and 130 stations {[}see Figure
-\ref{fig:sobi-stations-in-hamilton}{]}, and over 26,000 active
-memberships (Hamilton, 2015).
-
-Hamilton Bike Share has conducted one membership survey to date in 2018
-(Civicplan, 2017), and the findings from 420 members are broadly in line
-with the trends that were discussed above (see Fishman, 2016 for a
-recent review of the literature). The majority of respondents live
-within the core service area and the gender split is typical: 57\% of
-respondents are male and 41\% are female. The majority of respondents,
-both male and female, are between 25 and 34 years of age, but the
-percentage of male respondents is higher in the subsequent age groups.
-Respondents use bike share for commuting (40\% of trips) or errands and
-meetings (24\% of trips), and nearly 50\% of trips have an average
-length of 11 to 20 minutes. As a result of having a bike share
-membership, 49\% of respondents report that they use their private
-vehicle less often or much less often but 48\% report that their private
-vehicle use has remained about the same. This suggests that SoBi
-Hamilton has been useful for certain kinds of trips but not all, meaning
-that some trips continue to require a private vehicle.
-
-Further research has provided insights about the behaviour and route
-choice of bike share users in Hamilton. The routes most frequently
-travelled are longer than the shortest distance route from origin to
-destination (Lu et al., 2018; Scott et al., 2021). Ridership is
-influenced by weather conditions, temporal factors such as university
-terms, employment, and proximity to important destinations like the
-local university, but is not influenced by population and most
-transportation infrastructure variables (Scott and Ciuro, 2019). Bike
-share users prefer to deter from the shortest distance route to travel
-on streets with cycling facilities or lower volumes of traffic (Lu et
-al., 2018) and avoid steep slopes and busy roads (Scott et al., 2021).
-It is important to note that these studies used daily ridership or GPS
-data from trips taken before the equity stations were implemented, so
-these findings are informative with respect to the conventional
-stations.
+\hypertarget{original-system}{%
+\subsection{Original System}\label{original-system}}
+
+The focus of this study is the city of Hamilton, located in Ontario,
+Canada. The city launched a public bicycle share program in March 2015
+with 115 stations and 750 bicycles (Hamilton, 2015). Before June 2020,
+the program was known as Social Bicycles or SoBi Hamilton but is now
+called Hamilton Bike Share. Stations are spaced between 300 and 600
+metres apart (Scott et al., 2021). The core service area spans 40 sq.km
+of the city although it was planned to be 20 sq.km ({\textbf{???}}), and
+roughly 138,000 people are within 30 minutes walking of a bike share
+station {[}see Figure \ref{fig:hamilton-and-sobi-service-area}{]}. This
+represents roughly one fifth of the total population of the Hamilton
+Census Metropolitan Area according to the 2016 Canadian Census. The City
+of Hamilton undertook a large public engagement campaign to validate the
+locations of bicycle share stations that had been selected and to
+crowdsource potential locations for additional locations
+({\textbf{???}}). Most of the suggested locations were in the east end
+of the core service area that lack stations or in neighbourhoods not
+serviced by the PBSP. The program was enthusiastically welcomed in the
+city in 2015 - within three weeks of launching, 10,000 trips had been
+made (Hamilton, 2015), however inadequate coverage given the size of the
+Hamilton Bike Share's service area was identified as a problem early on,
+and transportation planners noted that the small size (i.e., supply of
+bicycle racks) and low quantity of stations would lead to challenges in
+balancing the system ({\textbf{???}}). Based on public feedback, 13
+stations were added to the original system.
+
+\hypertarget{equity-initiative}{%
+\subsection{Equity Initiative}\label{equity-initiative}}
+
+In 2017, Hamilton Bike Share Inc., the non-profit organization that
+operates the program, initiated an equity program, Everyone Rides
+Initiative (ERI), with the objective of reducing barriers that may
+prevent individuals from accessing bike share in Hamilton. Additional
+bicycles and stations were added to the program which expanded it to
+more disadvantaged areas in the core service area {[}see Figure 2{]}.
+The program also offers subsidized memberships to individuals who
+identify as low income, and complements these initiatives with cycle
+skills education. A comparable program can be found in Philadelphia (see
+Caspi and Noland, 2019).
+
+\hypertarget{current-system}{%
+\subsection{Current System}\label{current-system}}
+
+As of June 2020, Hamilton Bike Share has 900 bikes, 130 stations {[}see
+Figure \ref{fig:sobi-stations-in-hamilton}{]}, and over 26,000 active
+memberships (Hamilton, 2015). The core service area remains 40 sq.km.
+The PBSP has 13 stations that were added as part of the ERI initiative;
+we refer to these as ``equity stations'' throughout the paper while all
+other stations are ``conventional stations''. In total, 117 stations are
+``conventional'' and 13 are ``equity''.
+
+\hypertarget{membership}{%
+\subsection{Membership}\label{membership}}
+
+Hamilton Bike Share conducted one membership survey in 2018 (Civicplan,
+2017) and the findings from 420 members are broadly in line with the
+trends that were discussed above (see Fishman, 2016 for a recent review
+of the literature). The majority of respondents live within the core
+service area and the gender split is expected: 57\% of respondents are
+male and 41\% are female. The majority of respondents, both male and
+female, are between 25 and 34 years of age, but the percentage of male
+respondents is higher in the subsequent age groups. Respondents use bike
+share for commuting (40\% of trips) or errands and meetings (24\% of
+trips) and nearly 50\% of trips have an average length of 11 to 20
+minutes. As a result of having a bike share membership, 49\% of
+respondents report that they use their private vehicle less often or
+much less often but 48\% report that their private vehicle use has
+remained about the same. This suggests that Hamilton Bike Share has been
+useful for certain kinds of trips but not all, meaning that some trips
+continue to require a private vehicle.
+
+\hypertarget{relevant-research}{%
+\subsection{Relevant Research}\label{relevant-research}}
 
 Our analysis builds upon a previous and recent study (Hosford and
-Winters, 2018), which found that disadvantaged areas in Hamilton are
-better served by the city's PBSP compared to other Canadian cities
-{[}i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau{]} with PBSPs
-where advantaged areas have greater access. Hosford and Winters (2018)
-acknowledge that ``Hamilton stands out in that the lower income
+Winters, 2018), which found that areas in Hamilton with less advantage
+are better served by the city's PBSP compared to other Canadian cities
+{[}i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau{]} where
+areas that are less deprived have greater access. Hosford and Winters
+(2018) acknowledge that ``Hamilton stands out in that the lower income
 neighborhoods are located near the city center and wealthier
 neighborhoods are in the surrounding suburban areas''. Therefore, the
-core service area for the PBSP in Hamilton is by default in more
-disadvantaged areas compared to other Canadian cities, but there is also
-a great deal of variation in income in the city center as well because
-of the local university and increasing gentrification in the downtown
-area. In their analysis, Hosford and Winters (2018) did not
-differentiate between the conventional stations and those added to
-increase equity, which we refer to as equity stations in this paper.
-Therefore, this paper continues their work and makes an important
-contribution to the growing body of literature that examines equity
-issues in PBSPs.
+core service area for the PBSP in Hamilton by default covers some of the
+less advantaged areas in the city. However, there is also a great deal
+of variation in income in the city center because of the local
+university and increasing gentrification. Hosford and Winters (2018)
+took a macro-level and system-wide approach in their analysis by using
+dissemination areas \emph{across} the city as the unit of analysis. They
+did not focus specifically on the core service area and did not
+differentiate between conventional and equity stations.
 
@@
 
-{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/hamilton-and-sobi-service-area-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/hamilton-and-sobi-service-area-1}
 
@@
 the healthcare accessibility literature. This approach is more
-appropriate and informative than calculating the provider-to-population
-ratio (PPR) which simply divides the level of supply of a service (e.g.,
+appropriate and informative than calculating provider-to-population
+ratios (PPR) that simply divide the level of supply of a service (e.g.,
 the number of bicycle racks at a station) by the population who have
 access to the service (Paez et al., 2019). In particular, the Two-Step
-Floating Catchment Area method produces flexible catchment areas for
-populations accessing a service rather than using rigid boundaries like
-PPR, which provides more useful information because it takes into
-account individual behaviour and doesn't assume that people stay within
-the pre-defined boundaries (Paez et al., 2019). This is an important
-property that supports our rationale for applying this method to measure
-accessibility to Hamilton Bike Share. The City of Hamilton has
-positioned stations between 300 and 600 metres apart, which would lead
-to boundaries around these buffers, but it would be reasonable to assume
-that people are willing to walk beyond this threshold to access other
-stations if the ones nearest them have no supply of bicycles.
-
-More recently, the \emph{balanced} floating catchment area approach was
-developed to address issues with demand and supply inflation that are a
-result of the overlapping catchment areas produced by the Two-Step
-Floating Catchment Area (Paez et al., 2019). This overlap inflates the
-demand and generates inaccurate or misleading accessibility estimates
-(Paez et al., 2019). By adjusting the impedance weights, both supply and
-demand are proportionally allocated which eliminates the double-counting
-of population that leads to demand and supply inflation (Paez et al.,
-2019). Other benefits of this adjusted method include consideration of
-competition which can occur when catchment areas overlap, as well as the
-preservation of system-wide population and level of service. Balanced
-floating catchment area methods have been used to explore accessibility
-to health care providers (Paez et al., 2019) and COVID-19 health care
-services (Pereira et al., 2021), but have not yet been used in the
-cycling literature to explore issues of accessibility.
-
-In their review of accessibility measures, Geurs and van Wee (2004)
-highlight the need for greater inclusion of individual spatio-temporal
-constraints but acknowledge the challenges of acquiring and analyzing
-person-based data. This comes after Kwan's (1998) work to show that
-space-time measures are more capable of capturing interpersonal
-differences, especially the effect of space-time constraints on
-individual behaviour, and are more helpful for unraveling gender/ethnic
-differences. Applying the balanced floating catchment area approach to
-examine accessibility to Hamilton Bike Share constitutes a
-location-based measure with an individual component by stratifying
-according to median household income (Geurs and van Wee, 2004), but
-conducting a further sensitivity analysis by adjusting distance
-thresholds would introduce additional spatio-temporal constraints to
-evaluate equitable accessibility.
-
-The first step in the balanced floating catchment area method is to
-allocate the population to be serviced by each Hamilton Bike Share
-station: \[
+Floating Catchment Area (2SFCA) method ({\textbf{???}}; {\textbf{???}})
+produces flexible catchment areas instead of using rigid boundaries like
+PPR. This provides more useful information because it does not assume
+that people are limited to service within pre-defined boundaries (Paez
+et al., 2019). This is an important property that supports our rationale
+for applying this method to measure accessibility to Hamilton Bike
+Share. The City of Hamilton has positioned stations between 300 and 600
+metres apart, but anticipates that hubs will service those living within
+a 250 metre buffer from the station ({\textbf{???}}). The latter
+constitutes a normative statement: people ought to be able to access a
+station in less than 600 metres if they live in the core service area,
+with usage coming from 250 metres around. However, it is not known how
+far people are actually willing to travel to reach a station. It would
+be reasonable to assume that people are willing to walk beyond this
+threshold to access other stations if the ones nearest them have no
+supply of bicycles.
+
+More recently, the \emph{balanced} floating catchment area (FCA)
+approach was developed to address issues with demand and supply
+inflation that result from the overlapping catchment areas produced by
+earlier FCA methods (Paez et al., 2019). Briefly, overlapping catchment
+areas lead to inflation of population totals and deflation of service
+levels across a study area and generates inaccurate or misleading
+accessibility estimates. In contrast, Paez et al.~(2019) adjust the
+impedance weights so that both supply and demand are proportionally
+allocated. The results is a FCA method that balances the population and
+level of service by eliminating the over-counting of population and
+level of service that leads to distortions in demand and supply. Other
+benefits of this adjusted method include consideration of competition
+which can occur when catchment areas overlap, as well as the
+preservation of population and level of service. Balanced floating
+catchment area methods have been used to explore accessibility to health
+care providers (Paez et al., 2019) and COVID-19 health care services
+(Pereira, Carlos Kauê Vieira Braga, et al., 2021), but have not yet been
+used in the cycling literature to explore issues of accessibility.
+
+The first step in the floating catchment area method is to allocate the
+population to be serviced by each Hamilton Bike Share station: \[
 P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}
-\]
-
-Next, the level of service at each station (i.e., the maximum number of
-bicycle racks) is divided by its estimated service population within the
-established catchment area: \[
+\] As seen in the equation above, the population allocated to station
+\(j\) is the weighted sum of the population in the region; a spatial
+weight \(w_{ij}\) represents the friction that the population at \(i\)
+faces when reaching station \(j\), and is usually given by a
+distance-decay function, so that each station is assumed to service only
+a segment of the population within a limited geographical range.
+
+Next, the supply at each station (i.e., the maximum number of bicycle
+racks) is divided by its estimated service population within the
+established catchment area; this gives the level of service of station
+\(j\) in bicycle racks per person: \[
 L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
@@
 weighted sum of the level of service of all stations that can be reached
-from there according to normalized weights: \[
+from there according to the spatial weights: \[
 A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
@@
 
-The approach uses instead a set of suitably normalized weights as
-follows: \[
+The balanced approach of Paez et al. (2019) replaces the spatial weights
+with normalized versions as follows: \[
 {w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
@@
 
-and \[
+\noindent and: \[
 {w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
@@
 
-and \[
+\noindent and: \[
 \sum_{i = 1}^{n} {w^j_{ji}} = 1
@@
 
-Finally, accessibility is calculated without risk of demand or supply
-inflation: \[
+With these weights accessibility can be calculated without risk of
+demand or supply inflation: \[
 A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}
+\] By allocating the population and level of service proportionally,
+this method preserves the values of the population and level of service
+and provides a regional provider-to-population ratio since: \[
+\sum_{i=1}^n A_i = \sum_{j=1}^J L_j = \frac{\sum_{j=1}^JS_j}{\sum_{i=1}^n P_i}
 \]
+
+In fact, since the proportional allocation procedure means that any
+proportion of the population allocated to a station is never allocated
+to other stations, and conversely any level of service allocated to a
+population is never re-allocated elsewhere, this property is replicated
+for any level of aggregation. For this paper, we employ a hybrid
+location-based and person-based approach to calculating accessibility
+using disaggregate population data. In their review of accessibility
+measures, Geurs and van Wee (2004) highlight the need for greater
+inclusion of individual spatio-temporal constraints but acknowledge the
+challenges of acquiring and analyzing person-based data. This comes
+after Kwan's (1998) work to show that space-time measures are better
+able of capturing interpersonal differences, especially the effect of
+space-time constraints on individual behaviour, and are more helpful for
+unraveling gender/ethnic differences. Applying the balanced floating
+catchment area approach allows us to examine accessibility by
+stratifying according to median household income, which would constitute
+the individual component of the accessibility measure (Geurs and van
+Wee, 2004). However, conducting a further sensitivity analysis to
+measure accessibility at different distance thresholds would help to
+consider potential spatio-temporal constraints. Different people may be
+willing to travel different distances to access a public bicycle share
+station.
 
@@
 
-Following the method first presented by Tobler (1979), we used
-pycnophylactic interpolation to disaggregate population from each
-dissemination area to smaller polygons that are 50 by 50 metres. This
-method involves smoothing out the population from each dissemination
-area while preserving total volume {[}see Figure
-\ref{fig:da-population}{]}. Since pycnophylactic interpolation occurs at
-such a micro scale, we had to ensure that population numbers were not
-reallocated or smoothed out to areas where people do not live in
-Hamilton (i.e., where schools or parks are located). To do so, we
-retrieved shapefiles for various geographic features {[}see Table
-\ref{tab:data-features}{]} from Open Hamilton. Next, we extracted these
-features from the PBSP core service area and used pycnophylactic
-interpolation to disaggregate and reallocate population {[}see Figure
-\ref{fig:interpolated-population}{]}.
+To obtain population at sub-census geography levels (at the micro scale)
+we use pycnophylactic interpolation (Tobler, 1979). We obtained
+population data from the 2016 Census of Canada for dissemination areas
+(DAs, the smallest publicly available census geography in Canada). This
+zonal values of the population were interpolated to smaller polygons 50
+by 50 metres in size. Pycnophilactic interpolation involves smoothing
+out the population from each dissemination area while preserving total
+volume {[}see Figure \ref{fig:da-population}{]}. When interpolating the
+population at this high level of resolution it is important to ensure
+that population numbers were not allocated to areas where people do not
+live in Hamilton (e.g., to parks, large institutional buildings, etc.)
+To do so, we retrieved shapefiles for various geographic features {[}see
+Table \ref{tab:data-features}{]} from Open Hamilton. Next, we removed
+these features from the PBSP core service area and used pycnophylactic
+interpolation to disaggregate and reallocate population within the
+remaining area {[}see Figure \ref{fig:interpolated-population}{]}.
 
@@
 
-{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/da-population-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/da-population-1}
 
@@
 
-\caption{Population in all dissemination areas (outlined in white) in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the service area.}\label{fig:da-population}
+\caption{Population in all dissemination areas (outlined in white) that are inside or touch the bounding box of the Hamilton Bike Share core service area (outlined in black).}\label{fig:da-population}
 \end{figure}
@@
 
-{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/interpolated-population-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/interpolated-population-1}
 
@@
 \caption{Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
-The downtown area is outlined in dark green.
-Geographic features have been extracted.
-Dissemination area polygons are outlined in grey.}\label{fig:interpolated-population}
+Each micro population cell is 50-by-50 m in size.
+The downtown area is outlined in dark green.}\label{fig:interpolated-population}
 \end{figure}
@@
 
-\href{https://download.bbbike.org/osm/bbbike/}{BBBike} is an online
-cycle route planner that interfaces with OpenStreetMap. We extracted
-OpenStreetMap data for SoBi Hamilton's service area to calculate walking
-times from each population cell to nearby bike share stations, using a
-walking distance of 10km and a walking time of 30 minutes as a
-threshold. A travel time matrix was created with the origins as the
-coordinates of the population cells and the destinations as the
-coordinates of the bike share stations within the maximum threshold.
+To calculate walking times from the centroid of our micro population
+cells we extracted OpenStreetMap data for Hamilton Bike Share's service
+area from \href{https://download.bbbike.org/osm/bbbike/}{BBBike}, an
+online cycle route planner that interfaces with OpenStreetMap.
+OpenStreetMap data provides the networks for calculate walking times
+from each population cell to nearby bike share stations, using a maximum
+walking distance of 10 km and walking time of 30 minutes as thresholds.
+A travel time matrix was created with the origins as the coordinates of
+the population cells and the destinations as the coordinates of the bike
+share stations within the maximum threshold. This process provides a
+more realistic measure of the friction of reaching bike share stations
+by taking infrastructure into account of travel times, rather than using
+the Euclidean distance from population cell to station. Routing and
+travel time calculations were completed using the \texttt{R} package
+\texttt{r5r}, used for rapid realistic routing operations (Pereira,
+Saraiva, et al., 2021).
 
@@
 
-All data for this research were accessed from Open Hamilton\footnote{https://open.hamilton.ca/},
-an online repository of data curated by the City of Hamilton.
+All data for this research were accessed from publicly available Census
+of Canada sources, from OpenStreetMaps, and from Open Hamilton\footnote{https://open.hamilton.ca/},
+a public online repository of data curated by the City of Hamilton.
 
@@
 Consensus regarding the distance that individuals are willing to walk to
-access a PBSP is lacking, therefore the literature on walking behaviour
-was consulted to determine the thresholds for the sensitivity analysis.
-Previous studies have found that living within 250 metres (Fuller et
-al., 2011) and 300 metres (Kabra et al., 2020) is correlated with bike
-share use, while other research has found that walking trips are less
-than 600 metres and rarely more than 1200 metres (Millward et al., 2013)
-or a median distance of 650 metres (Larsen et al., 2010). We found that
-accessibility increases with a threshold between two and four minutes,
-but is then maximized at 5 minutes. At some stations, Hamilton Bike
-Share will depict a map to show the user the locations of the other
-nearest stations within a five minute walk, which suggests that this is
-an average distance that people are anticipated to walk. Accessibility
-decreases significantly after eight minutes, which is intuitive given
+access a PBSP station is lacking, but the literature on walking
+behaviour was consulted to determine the thresholds for the sensitivity
+analysis. Previous studies have found that living within 250 metres
+(Fuller et al., 2011) and 300 metres (Kabra et al., 2020) is correlated
+with bike share use, while other research has found that walking trips
+are less than 600 metres and rarely more than 1200 metres (Millward et
+al., 2013) or a median distance of 650 metres (Larsen et al., 2010). In
+Hamilton, Hamilton Bike Share will depict a map at some stations to show
+the user the locations of the other nearest stations within a five
+minute walk, which suggests that this is an average distance that people
+are expected to walk.
+
+In the present case, we found that congested accessibility calculated
+using the balanced FCA approach increases with a threshold between two
+and four minutes, but is then maximized at 5 minutes. Accessibility
+decreases substantially after eight minutes, which is intuitive given
 that demand on a limited supply increases as more people can reach each
@@
 conducting a sensitivity analysis to calculate accessibility at
-different walking times: 3 minutes, 5 minutes, 10 minutes, and 15
-minutes. We categorize these thresholds as minimum, average, maximum,
-and extreme, respectively. At each threshold, we compare accessibility
-between the current system and the original system to examine the
-contribution of the additional equity stations.
-
-\hypertarget{three-minutes}{%
-\subsubsection{Three Minutes}\label{three-minutes}}
-
-At the minimum threshold, we find that there are 25.2 bicycle racks per
-person in the original system. With the addition of equity stations,
-there are now 25.4 bicycles per person. Figure \ref{fig:figure-6}
-presents a comparison of accessibility between the systems.
-Accessibility is fairly uniform, with the exception of two small areas
-near the university and the waterfront park where accessibility is
+different walking times from population cell to bicycle share station: 3
+minutes, 5 minutes, 10 minutes, and 15 minutes. We categorize these
+thresholds as minimum, average, maximum, and extreme, respectively. At
+each threshold, we compare accessibility between the current system and
+the original system to examine the contribution of the additional equity
+stations.
+
+\hypertarget{minimum-threshold}{%
+\subsubsection{Minimum Threshold}\label{minimum-threshold}}
+
+With a walking distance of three minutes, we find that there are 25.2
+bicycle racks per person in the original system. The addition of equity
+stations increases this ratio slightly to 25.4 bicycles per person.
+Figure \ref{fig:figure-6} presents a comparison of accessibility between
+the systems for the population cells. Accessibility is fairly uniform
+overall, with the exception of two small areas where accessibility is
 slightly higher.
@@
 
-{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-6-1}
+ {\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-6-1}
 
@@
 
-\hypertarget{five-minutes}{%
-\subsubsection{Five Minutes}\label{five-minutes}}
-
-At the average threshold, we find that there are 68.6 bicycle racks per
-person in the original system. With the addition of equity stations,
-there are now 68.8 bicycles per person. Figure \ref{fig:figure-7}
-presents a comparison of accessibility between the systems. Again,
-accessibility is fairly uniform, with the exception of one very small
-area near the waterfront park where accessibility is higher.
+\hypertarget{average-threshold}{%
+\subsubsection{Average Threshold}\label{average-threshold}}
+
+With a walking distance of five minutes, we find that there are 68.6
+bicycle racks per person in the original system. With the addition of
+equity stations, there are now 68.8 bicycles per person. Figure
+\ref{fig:figure-7} presents a comparison of accessibility between the
+systems. Again, accessibility is fairly uniform, with the exception of
+one very small area. At this threshold, there are more bicycle racks per
+person than at the minimum threshold. This occurs because fewer stations
+can be reached when the travel time is 3 minutes or less.
 
@@
 
-{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-7-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-7-1}
 
@@
 
-\hypertarget{ten-minutes}{%
-\subsubsection{Ten Minutes}\label{ten-minutes}}
-
-At the maximum threshold, we find that there are 3.61 bicycle racks per
-person in the original system. With the addition of equity stations,
-there are now 3.74 bicycles per person. Figure \ref{fig:figure-8}
-presents a comparison of accessibility between the systems. We begin to
-see differences in accessibility across the service area, with users
-near the university and its adjacent neighbourhoods, as well as
-neighbourhoods north of the downtown area, have slightly higher
-accessibility. While the differences are modest, they are more apparent
-at this threshold than at shorter walking distances.
+\hypertarget{maximum-threshold}{%
+\subsubsection{Maximum Threshold}\label{maximum-threshold}}
+
+With a walking distance of ten minutes, we find that there are 3.61
+bicycle racks per person in the original system. With the addition of
+equity stations, there are now 3.74 bicycles per person. Figure
+\ref{fig:figure-8} presents a comparison of accessibility between the
+systems. We begin to see differences in accessibility across the service
+area, with users near the university and its adjacent neighbourhoods, as
+well as neighbourhoods north of the downtown area (outlined in green),
+having slightly higher accessibility. While the differences are modest,
+they are more apparent at this threshold than at shorter walking
+distances.
 
@@
 
-{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-8-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-8-1}
 
@@
 
-\hypertarget{fifteen-minutes}{%
-\subsubsection{Fifteen Minutes}\label{fifteen-minutes}}
-
-At the extreme threshold, we find that there are 2.44 bicycle racks per
-person in the original system. With the addition of equity stations,
-there are now 2.55 bicycles per person. Figure \ref{fig:figure-9}
-presents a comparison of accessibility between the systems. Users near
-the university and the neighbourhoods north of the downtown area
-(outlined in green) have the highest accessibility, followed by those
-who live in the city's downtown area. Accessibility in the east end of
-the core service area remains low.
+\hypertarget{extreme-threshold}{%
+\subsubsection{Extreme Threshold}\label{extreme-threshold}}
+
+With a walking distance of fifteen minutes, we find that there are 2.44
+bicycle racks per person in the original system. With the addition of
+equity stations, there are now 2.55 bicycles per person. Figure
+\ref{fig:figure-9} presents a comparison of accessibility between the
+systems. Users near the university and the neighbourhoods north of the
+downtown area (outlined in green) have the highest accessibility,
+followed by those who live in the city's downtown area. Accessibility in
+the east end of the core service area remains low.
 
@@
 
-{\centering \includegraphics[width=0.65\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-9-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-9-1}
 
@@
 
-\hypertarget{accessibility-by-median-total-income}{%
-\subsection{Accessibility by Median Total
-Income}\label{accessibility-by-median-total-income}}
-
-Approximately 138,000 people live within Hamilton Bike Share's core
-service area or within a 30 minute walk to the service area. We find
-that over 118,000 people can access a bicycle share station within a 15
-minute walk, which represents roughly 85\% of the total population in
-the core service area {[}see Table \ref{tab:accessibility-income}{]}.
-While previous research found that neighbourhoods with more disadvantage
-are better serviced by Hamilton Bike Share (Hosford and Winters, 2018),
-the authors used the Pampalon Deprivation Index from 2011 as a measure
-of socioeconomic status for dissemination areas across the city not just
-within the core service area. Instead, we use median total income for
-each dissemination area in the core service area to examine whether
-disparities in accessibility exist between income groups for those who
-live in or near the core service area. We divide income by quantiles:
-bottom 20\%, second 20\%, third 20\%, fourth 20\%, and top 20\%. One of
-the unique properties of the balanced floating catchment area method is
-that data can be reaggregated from small to larger polygons while
-preserving the total population and supply at each station. This avoids
-demand and supply inflation, and also enables us to present findings in
-a way that is easier to interpret (i.e., across dissemination areas
-instead of 50 by 50 metre population cells).
-
-Panels 1{[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
+\hypertarget{accessibility-by-median-household-income}{%
+\subsection{Accessibility by Median Household
+Income}\label{accessibility-by-median-household-income}}
+
+One of the unique properties of the balanced floating catchment area
+method is that data can be reaggregated while preserving the total
+population in the service area and the supply at each station. This
+avoids demand and supply inflation, and also enables us to present
+findings in a way that is easier and more intuitive to interpret.
+Therefore, we reaggregate population and accessibility from the smallest
+50 by 50 metres population cells to the median household income for each
+dissemination area. We divide income by quintiles: bottom 20\%, second
+20\%, third 20\%, fourth 20\%, and top 20\%.
+
+Panels 1 {[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
 {[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
@@
 maps with the combined spatial distribution of accessibility and median
-total income per household for the different thresholds. Our analysis
-demonstrates that the extreme threshold of fifteen minutes allows the
-most people to access a station. We find that stations added to
-Hamilton's public bicycle share program to increase equity for
-disadvantaged neighbourhoods achieved their goal by increasing
-accessibility, albeit only modestly {[}see Table
-\ref{tab:accessibility-income}{]}. By implementing equity stations in
-more areas with lower median total income in Hamilton, the PBSP has
-achieved greater horizontal equity by extending the spatial distribution
-of bicycles across the city. This is particularly evident at the minimum
-and average thresholds of three and five minutes, respectively. The
-largest gains were made for dissemination areas in the second 20\%,
-where an additional 3,073 and 5,395 people could reach a bicycle share
-station within three and five minutes walk, respectively, after the
-addition of the equity station.
-
-However, we found that dissemination areas with the lowest total
+household income at different thresholds. Our analysis demonstrates that
+the extreme threshold of fifteen minutes allows the most people to
+access a station. We find that stations added to Hamilton's public
+bicycle share program to increase equity for disadvantaged
+neighbourhoods achieved their goal by increasing access, albeit only
+modestly {[}see Table \ref{tab:accessibility-income}{]}. By implementing
+equity stations in more areas with lower median total income in
+Hamilton, the PBSP has achieved greater horizontal equity by extending
+the spatial distribution of bicycles across the city. This is
+particularly evident at the minimum and average thresholds of three and
+five minutes, respectively, where the equity stations fill a number of
+gaps in program coverage. The largest gains were made for dissemination
+areas in the second 20\%, where an additional 3,073 and 5,395 people
+could reach a bicycle share station within three and five minutes walk,
+respectively, after the addition of the equity station.
+
+However, we found that dissemination areas with the lowest median
 household income do not have much greater access to the program {[}see
@@
 only at certain thresholds. With and without equity stations, people in
-the top 20\% of income have the highest level of accessibility at a
-threshold of ten and fifteen minutes. Although dissemination areas in
-the second 20\% have the highest level of accessibility by a significant
-amount at lower thresholds, the bottom 20\%, who may benefit the most
-from Hamilton Bike Share's equity initiatives, have the lowest
-accessibility at three minutes threshold and the second lowest
-accessibility at all other thresholds. Panels
-1{[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
-{[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
-{[}\ref{fig:figure-bi-map-threshold-10}{]}, and 4
-{[}\ref{fig:figure-bi-map-threshold-15}{]} reveal specific areas where
-accessibility is low or zero, which are good candidates for additional
-equity stations.
-
-On the whole, we find that areas that have less than their equitable
-share of the level of service are some of the most socioeconomically
-disadvantaged areas in the city. While the addition of equity stations
-increases accessibility for all income groups at all thresholds, they
-did not increase accessibility significantly for any single income
-group. This finding aligns with other studies that have found
-disparities in station or bike location along income lines in New York
-City (Babagoli et al., 2019), Tampa (Chen et al., 2019), and Seattle
-(Mooney et al., 2019), where people who have greater advantage have
-better access to PBSPs.
+the top 20\% of income have the highest level of access at a threshold
+of ten and fifteen minutes. Although dissemination areas in the second
+20\% have the highest level of access by a significant amount at lower
+distance thresholds, the bottom 20\%, who may benefit the most from
+Hamilton Bike Share's equity initiatives, have the lowest access at
+three minutes threshold and the second lowest access at all other
+thresholds.
 
@@
 \hspace{1em}Fourth 20\% & 4.119 & 1673 & 4.119 & 1673 & 0.000 & 0\\
-\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{3.408} & \cellcolor{gray!6}{2114} & \cellcolor{gray!6}{3.434} & \cellcolor{gray!6}{2379} & \cellcolor{gray!6}{0.026} & \cellcolor{gray!6}{265}\\
+\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{3.757} & \cellcolor{gray!6}{2151} & \cellcolor{gray!6}{3.784} & \cellcolor{gray!6}{2416} & \cellcolor{gray!6}{0.027} & \cellcolor{gray!6}{265}\\
 \addlinespace[0.3em]
@@
 \hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{56.048} & \cellcolor{gray!6}{17513} & \cellcolor{gray!6}{56.137} & \cellcolor{gray!6}{22908} & \cellcolor{gray!6}{0.089} & \cellcolor{gray!6}{5395}\\
-\hspace{1em}Third 20\% & 4.258 & 15117 & 4.291 & 18309 & 0.033 & 3192\\
+\hspace{1em}Third 20\% & 4.259 & 15117 & 4.291 & 18309 & 0.032 & 3192\\
 \hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{1.094} & \cellcolor{gray!6}{2867} & \cellcolor{gray!6}{1.095} & \cellcolor{gray!6}{3116} & \cellcolor{gray!6}{0.001} & \cellcolor{gray!6}{249}\\
-\hspace{1em}Top 20\% & 5.925 & 4052 & 5.933 & 4518 & 0.008 & 466\\
+\hspace{1em}Top 20\% & 6.256 & 4074 & 6.264 & 4540 & 0.008 & 466\\
 \addlinespace[0.3em]
 \multicolumn{7}{l}{\textbf{Threshold - 10 minutes}}\\
-\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{0.603} & \cellcolor{gray!6}{41824} & \cellcolor{gray!6}{0.621} & \cellcolor{gray!6}{41981} & \cellcolor{gray!6}{0.018} & \cellcolor{gray!6}{157}\\
-\hspace{1em}Second 20\% & 0.860 & 27546 & 0.927 & 30503 & 0.067 & 2957\\
-\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{0.772} & \cellcolor{gray!6}{22394} & \cellcolor{gray!6}{0.799} & \cellcolor{gray!6}{25128} & \cellcolor{gray!6}{0.027} & \cellcolor{gray!6}{2734}\\
+\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{0.604} & \cellcolor{gray!6}{41824} & \cellcolor{gray!6}{0.622} & \cellcolor{gray!6}{41981} & \cellcolor{gray!6}{0.018} & \cellcolor{gray!6}{157}\\
+\hspace{1em}Second 20\% & 0.862 & 27546 & 0.929 & 30503 & 0.067 & 2957\\
+\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{0.776} & \cellcolor{gray!6}{22394} & \cellcolor{gray!6}{0.802} & \cellcolor{gray!6}{25128} & \cellcolor{gray!6}{0.026} & \cellcolor{gray!6}{2734}\\
 \hspace{1em}Fourth 20\% & 0.225 & 4544 & 0.227 & 4989 & 0.002 & 445\\
-\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{1.160} & \cellcolor{gray!6}{7989} & \cellcolor{gray!6}{1.162} & \cellcolor{gray!6}{9078} & \cellcolor{gray!6}{0.002} & \cellcolor{gray!6}{1089}\\
+\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{1.346} & \cellcolor{gray!6}{7989} & \cellcolor{gray!6}{1.348} & \cellcolor{gray!6}{9078} & \cellcolor{gray!6}{0.002} & \cellcolor{gray!6}{1089}\\
 \addlinespace[0.3em]
 \multicolumn{7}{l}{\textbf{Threshold - 15 minutes}}\\
-\hspace{1em}Bottom 20\% & 0.534 & 42208 & 0.554 & 42327 & 0.020 & 119\\
-\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{0.551} & \cellcolor{gray!6}{30507} & \cellcolor{gray!6}{0.611} & \cellcolor{gray!6}{31069} & \cellcolor{gray!6}{0.060} & \cellcolor{gray!6}{562}\\
-\hspace{1em}Third 20\% & 0.546 & 26108 & 0.573 & 26660 & 0.027 & 552\\
+\hspace{1em}Bottom 20\% & 0.536 & 42208 & 0.557 & 42327 & 0.021 & 119\\
+\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{0.555} & \cellcolor{gray!6}{30507} & \cellcolor{gray!6}{0.614} & \cellcolor{gray!6}{31069} & \cellcolor{gray!6}{0.059} & \cellcolor{gray!6}{562}\\
+\hspace{1em}Third 20\% & 0.554 & 26108 & 0.581 & 26660 & 0.027 & 552\\
 \hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{0.093} & \cellcolor{gray!6}{6312} & \cellcolor{gray!6}{0.096} & \cellcolor{gray!6}{7435} & \cellcolor{gray!6}{0.003} & \cellcolor{gray!6}{1123}\\
-\hspace{1em}Top 20\% & 0.712 & 10209 & 0.714 & 11089 & 0.002 & 880\\
+\hspace{1em}Top 20\% & 0.808 & 10209 & 0.811 & 11089 & 0.003 & 880\\
 \bottomrule
@@
 
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+Using disaggregate data, we examined spatial equity and accessibility to
+Hamilton Bike Share, with a particular focus on assessing the
+contribution of the program's equity stations. Our balanced floating
+catchment area approach, combined with pycnophylactic interpolation,
+enabled us to measure accessibility on a micro sale. This differentiates
+our analysis from similar papers exploring equity in PBSPs that use
+larger geographical units of analysis or that focus on station location
+instead of level of service. In this way, our paper has made
+contributions in a positive way by applying an intuitive and useful
+approach to measure accessibility to a PBSP, and in a normative way by
+serving to inform future investments in cycle infrastructure for
+Hamilton Bike Share.
+
+The sensitivity analysis revealed that accessibility to Hamilton Bike
+Share stations is maximized at five minutes and decreases significantly
+by eight minutes {[}see Table \ref{tab:accessibility-income}{]}. This
+reflects the normative guide, advertised on some bicycle share stations
+in Hamilton, that people can access other stations within a five minute
+walk if the station of origin has no bicycles. We find that over 118,000
+people can access a bicycle share station within a 15 minute walk, which
+represents roughly 85\% of the total population in the core service area
+{[}see Table \ref{tab:accessibility-income}{]}. At a minimum threshold
+of three minutes, too few stations can be accessed which leads to lower
+levels of accessibility. However, accessibility is at its lowest after
+eight minutes whereby congestion occurs due to increased potential
+demand. The City of Hamilton has recognized from the launch of the PBSP
+that significantly more stations and bicycles are needed to service the
+area ({\textbf{???}}). With a service area of 40 sq.km, Hamilton should
+have between 380 and 440 stations, instead of 130, and 1,500 bicycles
+instead of 900 ({\textbf{???}}). Reduced capacity within the system
+leads to gaps in coverage in some areas of the city ``with some areas
+not having the recommended station density of 300m between stations or
+10 stations per square km'' ({\textbf{???}}). This study illustrates the
+consequences of an imbalanced system whereby levels of accessibility are
+not equitable across income groups and become very low when congestion
+occurs.
+
+We found that equity stations achieved their goal of increasing
+horizontal equity in Hamilton's core service area. Panels
+1{[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
+{[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
+{[}\ref{fig:figure-bi-map-threshold-10}{]}, and 4
+{[}\ref{fig:figure-bi-map-threshold-15}{]} demonstrate how gaps in
+service were filled by these additional stations. In this respect, there
+are some commonalities between the expansions of Hamilton Bike Share and
+that of Citi Bike in New York. Babgoli et al.~(2019) found a slight but
+not statistically significant increase in the proportion of
+neighbourhoods with the highest levels of poverty with stations after
+the Citi Bike expansion in 2015. Although the Citi Bike expansion was
+not specifically driven by a desire to reduce inequities in access, 16\%
+of neighbourhoods with the most poverty had stations compared to 12\%
+before. With the addition of equity stations, there were large gains in
+accessibility for the second 20\% at the average threshold, over 5,000
+more people, but much smaller gains for the bottom 20\% with only 326
+more people.
+
+Vertical inequities, however, continue to persist despite the addition
+of equity stations. While the addition of equity stations modestly
+increases accessibility for all income groups at all thresholds, they
+did not increase accessibility significantly for any single income
+group. Most importantly, individuals in the bottom 20\% median household
+income quintile have the second lowest level of access to Hamilton Bike
+Share at most thresholds (average, maximum, and extreme). At the minimum
+threshold, the bottom 20\% have the lowest level of access. While
+previous research found that neighbourhoods with more disadvantage are
+better serviced by Hamilton Bike Share, the authors used the Pampalon
+Deprivation Index to determine the level of disadvantage for
+dissemination areas \emph{across} the city not just within the core
+service area (Hosford and Winters, 2018). Instead, we use median
+household income for each dissemination area \emph{within} the core
+service area. We conclude that Hamilton's PBSP, while by default located
+in areas with more deprivation compared to other cities, has disparities
+in accessibility between income groups. With equity stations, many areas
+with low median household income in the east end of the service area
+continue to have low accessibility. At higher thresholds of 10 and 15
+minutes, the top 20\% have the highest level of access to Hamilton Bike
+Share. This aligns with other studies from Tampa (Chen et al., 2019) and
+Seattle (Mooney et al., 2019), which have found disparities in station
+location or access to bicycles between levels of income and education.
+
+Based on our analysis, we identified specific areas that have both low
+accessibility and low median household income which would benefit from
+an increased supply of public bicycles. These empirical findings provide
+support to the City of Hamilton's efforts to increase equity and balance
+the system, and confirm that additional stations and bicycles are needed
+to improve access not only for the bottom 20\% but for all income
+groups. Panels 1{[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
+{[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
+{[}\ref{fig:figure-bi-map-threshold-10}{]}, and 4
+{[}\ref{fig:figure-bi-map-threshold-15}{]} highlight potential locations
+for new equity stations.
+
 \hypertarget{study-limitations}{%
@@
 Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), but our analysis is
-limited by the lack of accessible route or user data to conduct similar
-analyses for Hamilton Bike Share.
+limited by the lack of publicly available route or user data to conduct
+similar analyses for Hamilton Bike Share.
 
@@
 accessibility and reducing both horizontal and vertical inequities. In
-particular, accessibility improved the most for the second 20\% median
-total income groups at all thresholds, but the gains were only modest
-for all income groups. Dissemination areas with the bottom 20\% had the
-lowest accessibility at three minutes, and second lowest level of
-accessibility at five, ten, and fifteen minutes. Congestion effects were
-observed at higher thresholds, with accessibility decreasing
+particular, accessibility improved the most for those in the second 20\%
+median household income at all thresholds, but the gains were only
+modest for all income groups. Dissemination areas with the bottom 20\%
+had the lowest accessibility at three minutes, and second lowest levels
+of accessibility at five, ten, and fifteen minutes. Congestion effects
+were observed at higher thresholds, with accessibility decreasing
 significantly once the catchment area is increased to 10 minutes
-walking. Based on our analysis, we identified specific areas that have
-both low accessibility and low income which would benefit from an
-increased supply of bicycles. These are ideal candidates for new equity
-stations. In this way, our paper has made contributions in a positive
-way by applying an intuitive and useful approach to measure
-accessibility to a PBSP, and in a normative way by serving to inform
-future investments in cycle infrastructure for Hamilton Bike Share.
+walking.
 
@@
 
+\leavevmode\hypertarget{ref-bivand2020progress}{}%
+Bivand, R.S., 2020. Progress in the r ecosystem for representing and
+handling spatial data. Journal of Geographical Systems 1--32.
+doi:\href{https://doi.org/10.1007/s10109-020-00336-0}{10.1007/s10109-020-00336-0}
+
 \leavevmode\hypertarget{ref-breyWantRideMy2017}{}%
@@
 doi:\href{https://doi.org/10.1080/13504851.2016.1210760}{10.1080/13504851.2016.1210760}
+
+\leavevmode\hypertarget{ref-brunsdon2020opening}{}%
+Brunsdon, C., Comber, A., 2020. Opening practice: Supporting
+reproducibility and critical spatial data science. Journal of
+Geographical Systems 1--20.
+doi:\href{https://doi.org/10.1007/s10109-020-00334-2}{10.1007/s10109-020-00334-2}
 
@@
 
+\leavevmode\hypertarget{ref-lovelace2021open}{}%
+Lovelace, R., 2021. Open source tools for geographic analysis in
+transport planning. Journal of Geographical Systems 1--32.
+doi:\href{https://doi.org/doi.org/10.1007/s10109-020-00342-2}{doi.org/10.1007/s10109-020-00342-2}
+
 \leavevmode\hypertarget{ref-luUnderstandingBikeShare2018}{}%
@@
 
+\leavevmode\hypertarget{ref-Pereira2021r5r}{}%
+Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga, C.K.V., Conway,
+M.W., 2021. R5r: Rapid realistic routing on multimodal transport
+networks with r\textsuperscript{5} in r. Findings.
+doi:\href{https://doi.org/10.32866/001c.21262}{10.32866/001c.21262}
+
 \leavevmode\hypertarget{ref-qianBikesharingEquityDisadvantaged2020}{}%
@@
 
+\leavevmode\hypertarget{ref-qianEnhancingEquitableService2020}{}%
+Qian, X., Jaller, M., Niemeier, D., 2020. Enhancing equitable service
+level: Which can address better, dockless or dock-based Bikeshare
+systems? Journal of Transport Geography 86, 102784.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2020.102784}{10.1016/j.jtrangeo.2020.102784}
+
 \leavevmode\hypertarget{ref-reillyNoncyclistsFrequentCyclists2020}{}%
@@
 
+\leavevmode\hypertarget{ref-smith2015exploring}{}%
+Smith, C.S., Oh, J.-S., Lei, C., 2015. Exploring the equity dimensions
+of us bicycle sharing systems. Western Michigan University.
+Transportation Research Center for Livable~\ldots.
+
 \leavevmode\hypertarget{ref-toblerSmoothPycnophylacticInterpolation1979}{}%

### 2021-03-28 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd

@@
     email: desjae@mcmaster.ca
-    affiliation: Some School
+    affiliation: McMaster University
     footnote: 1
@@
     email: cd.higgins@utoronto.ca
-    affiliation: Another University
-    footnote: 2
+    affiliation: University of Toronto Scarborough
+    footnote: 
   - name: Antonio Paez
     email: paezha@mcmaster.ca
-    affiliation: Some School
-    footnote: 2
+    affiliation: McMaster University
+    footnote: 
 address:
@@
     text: "Corresponding Author"
-  - code: 2
-    text: "Equal contribution"
 abstract:
@@
 We then reaggregate the estimated accessibility for further analysis using census data.
-The addition of equity stations increased the serviced population at every threshold examined, and although accessibility increased for the whole population, this increase was relatively modest especially for population in the bottom 20% of median household income.
+Our findings indicate that the addition of equity stations increased the horizontal equity of the system by growing the serviced population at every threshold examined.
+With respect to vertical equity, although accessibility increased for the whole population, this increase was relatively modest especially for population in the bottom 20% of median total household income.
 
@@
 library(data.table)
+library(disk.frame)
 library(dplyr)
@@
 library(kableExtra)
-library(patchwork)
-library(readr)
+library(pycno)
+library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.htmllibrary(readr)
 library(raster)
@@
 library(units)
-library(disk.frame)
-library(pycno)
-library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.html
 ```
@@
 ```{r generate-package-bibliography, include=FALSE}
-knitr::write_bib(.packages(), "packages.bib")
+knitr::write_bib(c(.packages(), "knitr", "rticles"), "packages.bib")
 ```
@@
 As these systems become increasingly common, their introduction has been accompanied by a flurry of research that investigates them from the perspective of spatial and transportation equity [e.g., @smith2015exploring; @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @hullgrassoBikeShareEquity2020; @qianBikesharingEquityDisadvantaged2020; @qianEnhancingEquitableService2020].
-Indeed, although these programs are available to the general public and ought to be accessible to any individuals who wish to use them, research on PBSPs located in North American and European cities indicates that there are inequities in spatial accessibility in need of redress.
+Indeed, although these programs are available to the general public and ought to be accessible to any individuals who wish to use them, research on PBSPs located in North American and European cities indicates that inequity in spatial accessibility needs redressing.
 
@@
 Further, we conduct a sensitivity analysis at several walking time thresholds.
-In terms of spatial equity, we compare accessibility with and without the equity stations to assess the effect of the initiative, before reaggregating the data for further analysis using median household income information from the census.
+In terms of spatial equity, we compare accessibility with and without the equity stations to assess the effect of the initiative, before reaggregating the data for further analysis using median total household income information from the census.
 We find that the addition of equity stations does increase the serviced population at every threshold examined.
-However, although accessibility increased for the whole population, this increase was relatively modest, especially for population in the bottom 20% of median household income.
+However, although accessibility increased for the whole population, this increase was relatively modest, especially for population in the bottom 20% of median total household income.
 The analysis also identifies areas under-serviced within the core service area of the system.
@@
 Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved over the past century and has increasingly become linked with recent interest in prioritizing local proximity and modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020].
-As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequities in the provision of infrastructure, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options.
+As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequaltiy in the provision of infrastructure, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options.
 It also addresses some of the challenges of other performance measures such as level of service within a transportation network by measuring person-based indicators and exploring differences in use between population groups [@paezMeasuringAccessibilityPositive2012].
@@
 Beyond the utility derived from using shared bikes to destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019].
-Since the time or distance needed to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., maximum number of bicycle racks available) of stations matters.
+Since the time or distance needed to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., number of bicycle racks) of stations matters.
 Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
 Kabra et al.
-[-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300 metres of stations, which amounts to 2-4 minutes walking by an able-bodied adult.
+[-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300 m of stations, which amounts to 2-4 minutes walking by an able-bodied adult.
 Furthermore, similar to other public amenities affected by crowding, like health care services [e.g., @pereiraGeographicAccessCOVID192021], the utility of stations is limited by the number of bicycles that they can hold.
 The program may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles.
-Likewise, more people may not opt to use the program if the supply of bicycles available at the nearest station is insufficient to meet demand.
+Likewise, more people may not opt to use the program if the supply of bicycles available at nearby stations is insufficient to meet demand.
 The location and size of stations is important to increase the utility of this public transportation option for more people, thus achieving vertical equity.
@@
 Thus, this model does a good job of considering potential crowding or competition for services.
-While there have been many methodological innovations in FCA methods [for example, @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balancing of the impedance that addressed the effects of demand and service inflation found in previous FCA approaches [see @paezDemandLevelService2019].
+While there have been many methodological innovations in FCA methods [for example, @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balancing of the impedance that addressed the effects of demand and service inflation found in earlier FCA approaches [see @paezDemandLevelService2019].
 
@@
 This is also true for studies examining correlates of bike share demand [@wangNewBikeShare2019].
-Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals, which is useful for addressing vertical inequities in PBSP usage.
+Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals, which is useful for addressing vertical inequity in PBSP usage.
 
 Using the balanced floating catchment area method [@paezDemandLevelService2019], a novel approach that has not been used yet in cycling research, with disaggregate population-level data, in this paper we examine accessibility to a public bicycle share program in Hamilton, Ontario.
-We (1) conduct a sensitivity analysis by measuring accessibility and level of service to bike share stations at different walking time thresholds to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median household income of dissemination areas within the core service area.
+We (1) conduct a sensitivity analysis by measuring accessibility and level of service to bike share stations at different walking time thresholds to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median total household income of dissemination areas within the core service area.
 
@@
 This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are  better able of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences.
-Applying the balanced floating catchment area approach allows us to examine accessibility by stratifying according to median household income, which would constitute the individual component of the accessibility measure [@geursAccessibilityEvaluationLanduse2004].
+Applying the balanced floating catchment area approach allows us to examine accessibility by stratifying according to median total household income, which would constitute the individual component of the accessibility measure [@geursAccessibilityEvaluationLanduse2004].
 However, conducting a further sensitivity analysis to measure accessibility at different distance thresholds would help to consider potential spatio-temporal constraints.
@@
 ```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
-Each micro population cell is 50-by-50 m in size.
+Each population cell is 50-by-50 m in size.
 The downtown area is outlined in dark green."}
@@
           color = NA) + 
-  geom_sf(data = population_50x50,
-          aes(color = population)) +
-  scale_color_distiller(palette = "YlOrRd", 
+  #geom_sf(data = population_50x50,
+  #        aes(color = population)) +
+   #scale_fill_gradient(colours = "YlOrRd") +
+  geom_tile(data = cbind(st_drop_geometry(population_50x50), 
+                         st_coordinates(population_50x50)),
+            aes(x = X, 
+                y = Y, 
+                fill = population)) +
+  scale_fill_distiller(palette = "YlOrRd", 
                         direction = 1) +
@@
 ```{r figure-6, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
+#accessibility plot for current system with equity stations at 3 minutes walk
+accessibility_3min <- accessibility %>% 
+  dplyr::filter(threshold == "3 min")
+
 ggplot() +
@@
 #accessibility plot for current system with equity stations at 5 minutes walk
+accessibility_5min <- accessibility %>% 
+  dplyr::filter(threshold == "5 min")
+
 ggplot() +
@@
 #accessibility plot for current system with equity stations at 10 minutes walk
+accessibility_10min <- accessibility %>% 
+  dplyr::filter(threshold == "10 min")
+
 ggplot() +
@@
 #accessibility plot for current system with equity stations at 15 minutes walk
+accessibility_15min <- accessibility %>% 
+  dplyr::filter(threshold == "15 min")
+
 ggplot() +
@@
   #         fill = "green") +
-  theme(panel.background = element_rect(fill = "white"))
-```
-
-## Accessibility by Median Household Income
-
-One of the unique properties of the balanced floating catchment area method is that data can be reaggregated while preserving the total population in the service area and the supply at each station.
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank())
+```
+
+## Accessibility by Median Total Household Income
+
+A unique property of the balanced floating catchment area method is that data can be reaggregated while preserving the total population in the service area and the supply at each station.
 This avoids demand and supply inflation, and also enables us to present findings in a way that is perhaps more intuitive to interpret.
 Therefore, we reaggregate population and accessibility from the 50-by-50 m population cells to the dissemination area.
-We then can compare accessibility to median household income statistics from the census.
-
-Panels 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combine the spatial distribution of accessibility and median household income, using tertiles for the coloring scheme.
+We then can compare accessibility to median total household income statistics from the census.
+
+Figures 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combine the spatial distribution of accessibility and median total household income, using tertiles for the coloring scheme.
+In each figure the top panel is the system without the equity stations, and the bottom panel is accessibility with the equity stations.
 As expected, the analysis indicates that the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service areas of stations.
-We find that stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing access
-
-, albeit only modestly [see Table \ref{tab:accessibility-income}].
-By implementing equity stations in more areas with lower median total income in Hamilton, the PBSP has achieved greater horizontal equity by extending the spatial distribution of bicycles across the city.
+We find that stations added to Hamilton's public bicycle share program did indeed expand the spatial coverage of the system.
+By implementing equity stations in areas with lower median total household income in Hamilton, the PBSP attained greater horizontal equity by extending the spatial distribution of bicycles across the city.
 This is particularly evident at the minimum and average thresholds of three and five minutes, respectively, where the equity stations fill a number of gaps in program coverage.
-The largest gains were made for dissemination areas in the second 20%, where an additional 3,073 and 5,395 people could reach a bicycle share station within three and five minutes walk, respectively, after the addition of the equity station.
-
-However, we found that dissemination areas with the lowest median household income do not have much greater access to the program [see Table \ref{tab:table-3}].
-Income disparities still persist, however only at certain thresholds.
-With and without equity stations, people in the top 20% of income have the highest level of access at a threshold of ten and fifteen minutes.
-Although dissemination areas in the second 20% have the highest level of access by a significant amount at lower distance thresholds, the bottom 20%, who may benefit the most from Hamilton Bike Share's equity initiatives, have the lowest access at three minutes threshold and the second lowest access at all other thresholds.
 
@@
 # join accessibility to population cells
-test <- test %>% st_join(population_50x50 %>% dplyr::select(-UID)) %>%
+test <- test %>% 
+  st_join(population_50x50 %>% 
+            dplyr::select(-UID)) %>%
   rename(population50x50 = population)
@@
   quantile(c(0, 0.2, 0.4, 0.6, 0.8, 1), na.rm = TRUE)
+```
+
+```{r population-core-service-by-income, include=FALSE}
+# Find the population by income quantile
+population_by_income <- population_50x50 %>%
+  dplyr::select(-population) %>%
+  st_join(hamilton_da_2016) %>%
+  st_drop_geometry() %>%
+  distinct(DA,
+           .keep_all = TRUE) %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE)) %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population)) %>%
+  drop_na()
 ```
@@
   group_by(income_quintile) %>%
-  summarize(accessibility = sum(accessibility),
-            population = sum(population50x50))
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
 ```
@@
 
+To further explore this issue, we aggregate accessibility by median total household income across all geographies (see Table \ref{tab:accessibility-income}).
+Similar to @hosfordWhoArePublic2018, we find that Hamilton's PBSP already serviced a large proportion of the population in the bottom and second quintiles of median total household income.
+This may well be an artifact of the spatial socio-economic and demographic profile of Hamilton, where the most dense parts of the city (where a PBSP is most easily launched) are also those with relatively lower incomes.
+On the other hand, the levels of accessibility to Hamilton's PBSP are generally lower for populations in the bottom 20% of median total household income, compared to populations in the top 20%.
+
+The results indicate that the addition of equity stations increased horizontal equity by growing the population serviced irrespective of the walking threshold.
+The largest gains were made for dissemination areas in the second 20%, where an additional 3,073 and 5,395 people could reach a bicycle share station within three and five minutes walk, respectively, after the addition of the equity station.
+However, we find that there were only small changes in the population in the bottom 20% of median total household income who are serviced by the equity stations, and the accessibility gains are also quite modest and smaller than for populations in the second and third quintiles of median total household income.
+This suggests that income disparities persist, albeit this depends on the walking time thresholds.
+With and without equity stations, people in the top 20% of income have the highest level of access at a threshold of ten and fifteen minutes.
+Although dissemination areas in the second 20% have the highest level of access by a significant amount at lower distance thresholds, the bottom 20%, who may benefit the most from Hamilton Bike Share's equity initiatives, have the lowest access at three minutes threshold and the second lowest access at all other thresholds.
+
 ```{r accessibility-income, echo=FALSE}
@@
 
-colnames(accessibility_income) <- c("iq", "accessibility_conv", "population_conv", "accessibility_all", "population_all")
+colnames(accessibility_income) <- c("iq", "population_conv", "accessibility_conv", "population_all", "accessibility_all")
 
 accessibility_income <- accessibility_income %>%
-  mutate(accessibility_conv = round(accessibility_conv, 3), 
+  transmute(iq,
+            population = rep(population_by_income$population, 4),
             population_conv = round(population_conv),
+            accessibility_conv = round(accessibility_conv, 3), 
+            population_all = round(population_all),
             accessibility_all = round(accessibility_all, 3),
-         population_all = round(population_all),
-         accessibility_diff = accessibility_all - accessibility_conv,
-         population_diff = population_all - population_conv)
+            population_diff = population_all - population_conv, 
+            accessibility_diff = accessibility_all - accessibility_conv)
 
@@
       col.names = c("Income Quintile",
+                    "Total Population",
+                    "Population",
                     "Accessibility",
@@
                     "Population",
-                    "Accessibility",
-                    "Population"),
+                    "Accessibility"),
       align = c("l", "c", "c", "c", "c", "c", "c"),
-      caption = "\\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile and between systems (with and without equity stations).") %>%
+      caption = "\\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile and between systems (with and without equity stations).
+Total population is the population by income quintile in the DAs that have any PBSP service at all.") %>%
   kable_styling(latex_options = c("striped", "scale_down")) %>%
@@
             label_row_css = "background-color: #8f9498; color: #fff;") %>%
-  add_header_above(c(" " = 1, "Without Equity Stations" = 2, "With Equity Stations" = 2, "Difference" = 2)) %>%
+  add_header_above(c(" " = 2, "Without Equity Stations" = 2, "With Equity Stations" = 2, "Difference" = 2)) %>%
   #column_spec(2, color = "white",
@@
 
-Using disaggregate data, we examined spatial equity and accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
-Our balanced floating catchment area approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro sale.
+Using disaggregate data, in this paper we examined spatial equity and accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
+Use of a balanced floating catchment area approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro sale.
 This differentiates our analysis from similar papers exploring equity in PBSPs that use larger geographical units of analysis or that focus on station location instead of level of service.
@@
 The sensitivity analysis revealed that accessibility to Hamilton Bike Share stations is maximized at five minutes and decreases significantly by eight minutes [see Table \ref{tab:accessibility-income}].
-This reflects the normative guide, advertised on some bicycle share stations in Hamilton, that people can access other stations within a five minute walk if the station of origin has no bicycles.
+This reflects the normative guide, advertised on some bicycle share stations in Hamilton showing other stations within a five minute walk.
 We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}].
-At a minimum threshold of three minutes, too few stations can be accessed which leads to lower levels of accessibility.
-However, accessibility is at its lowest after eight minutes whereby congestion occurs due to increased potential demand.
-The City of Hamilton has recognized from the launch of the PBSP that significantly more stations and bicycles are needed to service the area [@hamiltonsobi2015].
-With a service area of 40 sq.km, Hamilton should have between 380 and 440 stations, instead of 130, and 1,500 bicycles instead of 900 [@hamiltonsobi2015].
+At a minimum threshold of three minutes, too few people can reach stations which leads to relatively high levels of service (since there is little crowding).
+However, accessibility is at its lowest after eight minutes whereby congestion effects due to increased potential demand kick in.
+The City of Hamilton has recognized from the launch of the PBSP that substantially more stations and bicycles are needed to service the area [@hamiltonsobi2015].
+With a service area of 40 sq.km, it is estimated that Hamilton should have between 380 and 440 stations, instead of 130, and 1,500 bicycles instead of 900 [@hamiltonsobi2015].
 Reduced capacity within the system leads to gaps in coverage in some areas of the city "with some areas not having the recommended station density of 300m between stations or 10 stations per square km" [@hamiltonsobi2015].
-This study illustrates the consequences of an imbalanced system whereby levels of accessibility are not equitable across income groups and become very low when congestion occurs.
-
-We found that equity stations achieved their goal of increasing horizontal equity in Hamilton's core service area.
-Panels 1[\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] demonstrate how gaps in service were filled by these additional stations.
+This study illustrates the consequences of an imbalanced system whereby levels of accessibility are not equitable across income groups.
+
+We found that equity stations increased horizontal equity in Hamilton's core service area.
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in service were filled by these additional stations.
 In this respect, there are some commonalities between the expansions of Hamilton Bike Share and that of Citi Bike in New York.
@@
 [-@babagoliExploringHealthSpatial2019] found a slight but not statistically significant increase in the proportion of neighbourhoods with the highest levels of poverty with stations after the Citi Bike expansion in 2015.
-Although the Citi Bike expansion was not specifically driven by a desire to reduce inequities in access, 16% of neighbourhoods with the most poverty had stations compared to 12% before.
+Although the Citi Bike expansion was not specifically driven by a desire to reduce inequity in access, 16% of neighbourhoods with the highest levels of poverty had stations compared to 12% before.
 With the addition of equity stations, there were large gains in accessibility for the second 20% at the average threshold, over 5,000 more people, but much smaller gains for the bottom 20% with only 326 more people.
 
-Vertical inequities, however, continue to persist despite the addition of equity stations.
-While the addition of equity stations modestly increases accessibility for all income groups at all thresholds, they did not increase accessibility significantly for any single income group.
-Most importantly, individuals in the bottom 20% median household income quintile have the second lowest level of access to Hamilton Bike Share at most thresholds (average, maximum, and extreme).
+Vertical inequity, however, persisted despite the addition of equity stations.
+While the addition of equity stations seems to modestly increase accessibility for all income groups at all thresholds, they did not increase accessibility substantially for any single income group.
+Most importantly, individuals in the bottom 20% of median total household income have the second lowest level of access to Hamilton Bike Share at most thresholds (average, maximum, and extreme).
 At the minimum threshold, the bottom 20% have the lowest level of access.
 While previous research found that neighbourhoods with more disadvantage are better serviced by Hamilton Bike Share, the authors used the Pampalon Deprivation Index to determine the level of disadvantage for dissemination areas _across_ the city not just within the core service area [@hosfordWhoArePublic2018].
-Instead, we use median household income for each dissemination area _within_ the core service area.
+Instead, we use median total household income for each dissemination area _within_ the core service area.
 We conclude that Hamilton's PBSP, while by default located in areas with more deprivation compared to other cities, has disparities in accessibility between income groups.
@@
 These empirical findings provide support to the City of Hamilton's efforts to increase equity and balance the system, and confirm that additional stations and bicycles are needed to improve access not only for the bottom 20% but for all income groups.
-Panels 1[\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] highlight potential locations for new equity stations.
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} highlight potential locations for new equity stations.
 
@@
 
+An additional limitation is the lack of publicly available information about the number of bicycles at each station.
+Hamilton Bike share works to balance the number of bicycles in stations, but it is reasonable to expect that the number of bicycles will not match exactly the number of racks at every station.
+Ideally, instead of number of bicycle racks as our measure of supply, we would have liked to use the average number of bicycles at stations, perhaps at different times during the day too.
+Should these data become available, it would be worthwhile to revisit the question to examine how well operation of the system (including balancing of bicycles among stations) works to maintain the nominal levels of accessibility examined in this paper.
+
 Conclusion
@@
 
-The addition of specific equity stations to the public bicycle share program in Hamilton, Ontario had the net effect of increasing accessibility and reducing both horizontal and vertical inequities.
+The addition of specific equity stations to the public bicycle share program in Hamilton, Ontario had the net effect of increasing accessibility and reducing to some extent both horizontal and vertical inequities.
 In particular, accessibility improved the most for those in the second 20% median household income at all thresholds, but the gains were only modest for all income groups.
@@
 Wang and Lindsey [-@wangNewBikeShare2019] have noted that there is a lack of research that examines how bike share users' behaviour changes as a result of program changes to station locations or improvements in accessibility.
-As such, a logical next step to this research is to examine whether Hamilton Bike Share's equity stations increased ridership or resulted in new memberships in areas that were previously under-served.
+As such, a logical next step to this research is to examine whether Hamilton Bike Share's equity stations increased ridership or resulted in new memberships in areas that were previously under-serviced.
 An examination of the types of trips undertaken by residents in these areas would also be informative, such as the study undertaken by Caspi and Norland [-@caspiBikesharingPhiladelphiaLowerincome2019] after bike share stations were implemented in low-income Philadelphia neighbourhoods.
@@
 It would be worthwhile to explore the route choice of bike share trips departing or ending at the equity stations and to identify factors that specifically influence trips from these stations, which would extend existing studies conducted by Scott and colleagues [@luUnderstandingBikeShare2018; @scottWhatFactorsInfluence2019; @scottRouteChoiceBike2021].
-This paper, combined with additional studies such as those conceptualized above, would serve as a valuable case study for Hamilton and other cities with PBSPs who wish to evaluate and address spatial inequities in accessibility and transportation options in urban areas.
+This paper, combined with additional studies such as those conceptualized above, would serve as a valuable case study for Hamilton and other cities with PBSPs that wish to evaluate and address spatial inequuality in accessibility and transportation options in urban areas.
+
+Acknowledgments
+===================
+
+This research was completed using open software, and the authors wish to acknowledge the developers of the following `R` packages: `biscale` [@R-biscale], `cowplot` [@R-cowplot], `data.table` [@R-data.table], `disk.frame` [@R-disk.frame], `gdistance` [@R-gdistance], `gridExtra` [@R-gridExtra], `kableExtra` [@R-kableExtra], `knitr` [@R-knitr], `pycno` [@R-pycno], `r5r` [@R-r5r], `raster` [@R-raster], `rgdal` [@R-rgdal], `rticles` [@R-rticles], `sf` [@R-sf], `tidyverse` [@R-tidyverse], `tinytex` [@R-tinytex],
+`units` [@R-units].
 

### 2021-03-28 Bike-share-spatial-equity/Bike-share-spatial-equity.tex

@@
 program using a balanced floating catchment area approach}
-    \author[Some School]{Elise Desjardins\corref{1}}
+    \author[McMaster University]{Elise Desjardins\corref{1}}
    \ead{desjae@mcmaster.ca} 
-    \author[Another University]{Christopher D. Higgins\corref{2}}
+    \author[University of Toronto Scarborough]{Christopher D. Higgins}
    \ead{cd.higgins@utoronto.ca} 
-    \author[Some School]{Antonio Paez\corref{2}}
+    \author[McMaster University]{Antonio Paez}
    \ead{paezha@mcmaster.ca} 
@@
       \cortext[1]{Corresponding Author}
-    \cortext[2]{Equal contribution}
 
@@
   at several walking time thresholds. We then reaggregate the estimated
-  accessibility for further analysis using census data. The addition of
-  equity stations increased the serviced population at every threshold
-  examined, and although accessibility increased for the whole population,
-  this increase was relatively modest especially for population in the
-  bottom 20\% of median household income.
+  accessibility for further analysis using census data. Our findings
+  indicate that the addition of equity stations increased the horizontal
+  equity of the system by growing the serviced population at every
+  threshold examined. With respect to vertical equity, although
+  accessibility increased for the whole population, this increase was
+  relatively modest especially for population in the bottom 20\% of median
+  total household income.
   \end{abstract}
@@
 accessible to any individuals who wish to use them, research on PBSPs
-located in North American and European cities indicates that there are
-inequities in spatial accessibility in need of redress.
+located in North American and European cities indicates that inequity in
+spatial accessibility needs redressing.
 
@@
 and without the equity stations to assess the effect of the initiative,
-before reaggregating the data for further analysis using median
+before reaggregating the data for further analysis using median total
 household income information from the census. We find that the addition
@@
 whole population, this increase was relatively modest, especially for
-population in the bottom 20\% of median household income. The analysis
-also identifies areas under-serviced within the core service area of the
-system.
+population in the bottom 20\% of median total household income. The
+analysis also identifies areas under-serviced within the core service
+area of the system.
 
@@
 spread spatially in a given area, accessibility is relevant to PBSPs
-because it can identify current inequities in the provision of
+because it can identify current inequaltiy in the provision of
 infrastructure, as well as guide interventions that increase access for
@@
 of accessing the program and ultimately their use of the program, the
-location and size (e.g., maximum number of bicycle racks available) of
-stations matters. Indeed, distance to bicycle share stations is
-associated with use (Fuller et al., 2011; J. Wang and Lindsey, 2019b)
-and can be a barrier to using PBSPs (Fishman et al., 2014). Kabra et
-al.~(2020) found that the majority of bike share usage in Paris comes
-from areas within 300 metres of stations, which amounts to 2-4 minutes
-walking by an able-bodied adult. Furthermore, similar to other public
-amenities affected by crowding, like health care services (e.g.,
-Pereira, Carlos Kauê Vieira Braga, et al., 2021), the utility of
-stations is limited by the number of bicycles that they can hold. The
-program may not necessarily be improved if stations are easy to reach
-but offer only a small number of bicycles. Likewise, more people may not
-opt to use the program if the supply of bicycles available at the
-nearest station is insufficient to meet demand. The location and size of
-stations is important to increase the utility of this public
-transportation option for more people, thus achieving vertical equity.
-Accessibility analyses for PBSPs constitute a positive and
-evaluation-based approach that also has the potential to inform equity
-efforts. For instance, Wang and Lindsey (2019b) investigated whether new
-or relocated bicycle share stations increased accessibility and use,
-which offered important insights to improve the performance of the
-program.
+location and size (e.g., number of bicycle racks) of stations matters.
+Indeed, distance to bicycle share stations is associated with use
+(Fuller et al., 2011; J. Wang and Lindsey, 2019b) and can be a barrier
+to using PBSPs (Fishman et al., 2014). Kabra et al.~(2020) found that
+the majority of bike share usage in Paris comes from areas within 300 m
+of stations, which amounts to 2-4 minutes walking by an able-bodied
+adult. Furthermore, similar to other public amenities affected by
+crowding, like health care services (e.g., Pereira, Carlos Kauê Vieira
+Braga, et al., 2021), the utility of stations is limited by the number
+of bicycles that they can hold. The program may not necessarily be
+improved if stations are easy to reach but offer only a small number of
+bicycles. Likewise, more people may not opt to use the program if the
+supply of bicycles available at nearby stations is insufficient to meet
+demand. The location and size of stations is important to increase the
+utility of this public transportation option for more people, thus
+achieving vertical equity. Accessibility analyses for PBSPs constitute a
+positive and evaluation-based approach that also has the potential to
+inform equity efforts. For instance, Wang and Lindsey (2019b)
+investigated whether new or relocated bicycle share stations increased
+accessibility and use, which offered important insights to improve the
+performance of the program.
 
@@
 intuitive balancing of the impedance that addressed the effects of
-demand and service inflation found in previous FCA approaches (see Paez
+demand and service inflation found in earlier FCA approaches (see Paez
 et al., 2019).
@@
 estimate for individuals, which is useful for addressing vertical
-inequities in PBSP usage.
+inequity in PBSP usage.
 
@@
 (3) examine whether disparities in accessibility exist according to
-median household income of dissemination areas within the core service
-area.
+median total household income of dissemination areas within the core
+service area.
 
@@
 catchment area approach allows us to examine accessibility by
-stratifying according to median household income, which would constitute
-the individual component of the accessibility measure (Geurs and van
-Wee, 2004). However, conducting a further sensitivity analysis to
-measure accessibility at different distance thresholds would help to
+stratifying according to median total household income, which would
+constitute the individual component of the accessibility measure (Geurs
+and van Wee, 2004). However, conducting a further sensitivity analysis
+to measure accessibility at different distance thresholds would help to
 consider potential spatio-temporal constraints. Different people may be
@@
 (DAs, the smallest publicly available census geography in Canada). This
-zonal values of the population were interpolated to smaller polygons 50
-by 50 metres in size. Pycnophilactic interpolation involves smoothing
-out the population from each dissemination area while preserving total
+zonal values of the population were interpolated to smaller polygons
+50-by-50 m in size. Pycnophilactic interpolation involves smoothing out
+the population from each dissemination area while preserving total
 volume {[}see Figure \ref{fig:da-population}{]}. When interpolating the
@@
 \caption{Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
-Each micro population cell is 50-by-50 m in size.
+Each population cell is 50-by-50 m in size.
 The downtown area is outlined in dark green.}\label{fig:interpolated-population}
@@
 access a PBSP station is lacking, but the literature on walking
-behaviour was consulted to determine the thresholds for the sensitivity
-analysis. Previous studies have found that living within 250 metres
-(Fuller et al., 2011) and 300 metres (Kabra et al., 2020) is correlated
-with bike share use, while other research has found that walking trips
-are less than 600 metres and rarely more than 1200 metres (Millward et
-al., 2013) or a median distance of 650 metres (Larsen et al., 2010). In
-Hamilton, Hamilton Bike Share will depict a map at some stations to show
-the user the locations of the other nearest stations within a five
-minute walk, which suggests that this is an average distance that people
-are expected to walk.
+behaviour can provide some guidelines to determine the thresholds for
+sensitivity analysis. Previous studies have found that living within 250
+metres (Fuller et al., 2011) and 300 metres (Kabra et al., 2020) is
+correlated with bike share use, while other research has found that
+walking trips are less than 600 metres and rarely more than 1200 metres
+(Millward et al., 2013) or a median distance of 650 metres (Larsen et
+al., 2010). In Hamilton, Hamilton Bike Share will depict a map at some
+stations to show the user the locations of the other nearest stations
+within a five minute walk, which suggests that this is an average
+distance that people are expected to walk.
 
@@
 the original system to examine the contribution of the additional equity
-stations.
+stations. When considering the results reported below, it is important
+to remember that accessibility is technically a form of smoothing
+(O'Kelly and Horner, 2003, pp. 7--8): smaller thresholds produce less
+smoothing (which can result in ``spiky'' accessibility landscapes), and
+larger thresholds produce more smoothing, and fewer spikes.
 
@@
 With a walking distance of three minutes, we find that there are 25.2
-bicycle racks per person in the original system. The addition of equity
+bicycle racks per person system-wide in the original system
+configuration (i.e., without equity stations). The addition of equity
 stations increases this ratio slightly to 25.4 bicycles per person.
@@
 overall, with the exception of two small areas where accessibility is
-slightly higher.
+slightly higher. This high level of system-wide accessibility occurs
+because the population that can reach bike share stations when travel
+time is 3 minutes or less is very limited, and accessibility is strongly
+shaped by a few locations that concentrate population and stations.
 
@@
 bicycle racks per person in the original system. With the addition of
-equity stations, there are now 68.8 bicycles per person. Figure
+equity stations, there are now 68.8 bicycle racks per person. At this
+threshold, there are more bicycle racks per person than at the minimum
+threshold. System-wide accessibility has in fact increased: the
+population that can reach the stations has grown, but not to the the
+point that congestion effects begin to take place. Figure
 \ref{fig:figure-7} presents a comparison of accessibility between the
 systems. Again, accessibility is fairly uniform, with the exception of
-one very small area. At this threshold, there are more bicycle racks per
-person than at the minimum threshold. This occurs because fewer stations
-can be reached when the travel time is 3 minutes or less.
+one very small area.
 
@@
 \ref{fig:figure-8} presents a comparison of accessibility between the
-systems. We begin to see differences in accessibility across the service
-area, with users near the university and its adjacent neighbourhoods, as
-well as neighbourhoods north of the downtown area (outlined in green),
-having slightly higher accessibility. While the differences are modest,
-they are more apparent at this threshold than at shorter walking
-distances.
+systems. Differences in accessibility across the service area are now
+apparent, with users near the university and its adjacent
+neighbourhoods, as well as neighbourhoods north of the downtown area
+(outlined in green), having slightly higher accessibility. While the
+differences are modest, they are more apparent at this threshold than at
+shorter walking distances.
 
@@
 
-\hypertarget{accessibility-by-median-household-income}{%
-\subsection{Accessibility by Median Household
-Income}\label{accessibility-by-median-household-income}}
-
-One of the unique properties of the balanced floating catchment area
-method is that data can be reaggregated while preserving the total
-population in the service area and the supply at each station. This
-avoids demand and supply inflation, and also enables us to present
-findings in a way that is easier and more intuitive to interpret.
-Therefore, we reaggregate population and accessibility from the smallest
-50 by 50 metres population cells to the median household income for each
-dissemination area. We divide income by quintiles: bottom 20\%, second
-20\%, third 20\%, fourth 20\%, and top 20\%.
-
-Panels 1 {[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
+\hypertarget{accessibility-by-median-total-household-income}{%
+\subsection{Accessibility by Median Total Household
+Income}\label{accessibility-by-median-total-household-income}}
+
+A unique property of the balanced floating catchment area method is that
+data can be reaggregated while preserving the total population in the
+service area and the supply at each station. This avoids demand and
+supply inflation, and also enables us to present findings in a way that
+is perhaps more intuitive to interpret. Therefore, we reaggregate
+population and accessibility from the 50-by-50 m population cells to the
+dissemination area. We then can compare accessibility to median total
+household income statistics from the census.
+
+Figures 1 {[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
 {[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
@@
 {[}\ref{fig:figure-bi-map-threshold-15}{]} depict bivariate choropleth
-maps with the combined spatial distribution of accessibility and median
-household income at different thresholds. Our analysis demonstrates that
-the extreme threshold of fifteen minutes allows the most people to
-access a station. We find that stations added to Hamilton's public
-bicycle share program to increase equity for disadvantaged
-neighbourhoods achieved their goal by increasing access, albeit only
-modestly {[}see Table \ref{tab:accessibility-income}{]}. By implementing
-equity stations in more areas with lower median total income in
-Hamilton, the PBSP has achieved greater horizontal equity by extending
-the spatial distribution of bicycles across the city. This is
-particularly evident at the minimum and average thresholds of three and
-five minutes, respectively, where the equity stations fill a number of
-gaps in program coverage. The largest gains were made for dissemination
-areas in the second 20\%, where an additional 3,073 and 5,395 people
-could reach a bicycle share station within three and five minutes walk,
-respectively, after the addition of the equity station.
-
-However, we found that dissemination areas with the lowest median
-household income do not have much greater access to the program {[}see
-Table \ref{tab:table-3}{]}. Income disparities still persist, however
-only at certain thresholds. With and without equity stations, people in
-the top 20\% of income have the highest level of access at a threshold
-of ten and fifteen minutes. Although dissemination areas in the second
-20\% have the highest level of access by a significant amount at lower
-distance thresholds, the bottom 20\%, who may benefit the most from
-Hamilton Bike Share's equity initiatives, have the lowest access at
-three minutes threshold and the second lowest access at all other
-thresholds.
+maps with the combine the spatial distribution of accessibility and
+median total household income, using tertiles for the coloring scheme.
+In each figure the top panel is the system without the equity stations,
+and the bottom panel is accessibility with the equity stations. As
+expected, the analysis indicates that the extreme threshold of fifteen
+minutes is associated with the largest number of people who are within
+the assumed service areas of stations. We find that stations added to
+Hamilton's public bicycle share program did indeed expand the spatial
+coverage of the system. By implementing equity stations in areas with
+lower median total household income in Hamilton, the PBSP attained
+greater horizontal equity by extending the spatial distribution of
+bicycles across the city. This is particularly evident at the minimum
+and average thresholds of three and five minutes, respectively, where
+the equity stations fill a number of gaps in program coverage.
 
@@
 
+To further explore this issue, we aggregate accessibility by median
+total household income across all geographies (see Table
+\ref{tab:accessibility-income}). Similar to Hosford and Winters (2018),
+we find that Hamilton's PBSP already serviced a large proportion of the
+population in the bottom and second quintiles of median total household
+income. This may well be an artifact of the spatial socio-economic and
+demographic profile of Hamilton, where the most dense parts of the city
+(where a PBSP is most easily launched) are also those with relatively
+lower incomes. On the other hand, the levels of accessibility to
+Hamilton's PBSP are generally lower for populations in the bottom 20\%
+of median total household income, compared to populations in the top
+20\%.
+
+The results indicate that the addition of equity stations increased
+horizontal equity by growing the population serviced irrespective of the
+walking threshold. The largest gains were made for dissemination areas
+in the second 20\%, where an additional 3,073 and 5,395 people could
+reach a bicycle share station within three and five minutes walk,
+respectively, after the addition of the equity station. However, we find
+that there were only small changes in the population in the bottom 20\%
+of median total household income who are serviced by the equity
+stations, and the accessibility gains are also quite modest and smaller
+than for populations in the second and third quintiles of median total
+household income. This suggests that income disparities persist, albeit
+this depends on the walking time thresholds. With and without equity
+stations, people in the top 20\% of income have the highest level of
+access at a threshold of ten and fifteen minutes. Although dissemination
+areas in the second 20\% have the highest level of access by a
+significant amount at lower distance thresholds, the bottom 20\%, who
+may benefit the most from Hamilton Bike Share's equity initiatives, have
+the lowest access at three minutes threshold and the second lowest
+access at all other thresholds.
+
 \begin{table}
 
-\caption{\label{tab:accessibility-income}\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile and between systems (with and without equity stations).}
+\caption{\label{tab:accessibility-income}\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile and between systems (with and without equity stations).
+Total population is the population by income quintile in the DAs that have any PBSP service at all.}
 \centering
 \resizebox{\linewidth}{!}{
-\begin{tabular}[t]{lcccccc}
+\begin{tabular}[t]{lccccccl}
 \toprule
-\multicolumn{1}{c}{ } & \multicolumn{2}{c}{Without Equity Stations} & \multicolumn{2}{c}{With Equity Stations} & \multicolumn{2}{c}{Difference} \\
-\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7}
-Income Quintile & Accessibility & Population & Accessibility & Population & Accessibility & Population\\
+\multicolumn{2}{c}{ } & \multicolumn{2}{c}{Without Equity Stations} & \multicolumn{2}{c}{With Equity Stations} & \multicolumn{2}{c}{Difference} \\
+\cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8}
+Income Quintile & Total Population & Population & Accessibility & Population & Accessibility & Population & Accessibility\\
 \midrule
 \addlinespace[0.3em]
-\multicolumn{7}{l}{\textbf{Threshold - 3 minutes}}\\
-\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{2.377} & \cellcolor{gray!6}{22359} & \cellcolor{gray!6}{2.424} & \cellcolor{gray!6}{22798} & \cellcolor{gray!6}{0.047} & \cellcolor{gray!6}{439}\\
-\hspace{1em}Second 20\% & 12.203 & 9347 & 12.281 & 12420 & 0.078 & 3073\\
-\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{3.093} & \cellcolor{gray!6}{7745} & \cellcolor{gray!6}{3.156} & \cellcolor{gray!6}{9455} & \cellcolor{gray!6}{0.063} & \cellcolor{gray!6}{1710}\\
-\hspace{1em}Fourth 20\% & 4.119 & 1673 & 4.119 & 1673 & 0.000 & 0\\
-\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{3.757} & \cellcolor{gray!6}{2151} & \cellcolor{gray!6}{3.784} & \cellcolor{gray!6}{2416} & \cellcolor{gray!6}{0.027} & \cellcolor{gray!6}{265}\\
+\multicolumn{8}{l}{\textbf{Threshold - 3 minutes}}\\
+\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{43441} & \cellcolor{gray!6}{22359} & \cellcolor{gray!6}{2.377} & \cellcolor{gray!6}{22798} & \cellcolor{gray!6}{2.424} & \cellcolor{gray!6}{439} & \cellcolor{gray!6}{0.047}\\
+\hspace{1em}Second 20\% & 33312 & 9347 & 12.203 & 12420 & 12.281 & 3073 & 0.078\\
+\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{30940} & \cellcolor{gray!6}{7745} & \cellcolor{gray!6}{3.093} & \cellcolor{gray!6}{9455} & \cellcolor{gray!6}{3.156} & \cellcolor{gray!6}{1710} & \cellcolor{gray!6}{0.063}\\
+\hspace{1em}Fourth 20\% & 20185 & 1673 & 4.119 & 1673 & 4.119 & 0 & 0.000\\
+\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{27541} & \cellcolor{gray!6}{2151} & \cellcolor{gray!6}{3.757} & \cellcolor{gray!6}{2416} & \cellcolor{gray!6}{3.784} & \cellcolor{gray!6}{265} & \cellcolor{gray!6}{0.027}\\
 \addlinespace[0.3em]
-\multicolumn{7}{l}{\textbf{Threshold - 5 minutes}}\\
-\hspace{1em}Bottom 20\% & 1.302 & 35477 & 1.357 & 35803 & 0.055 & 326\\
-\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{56.048} & \cellcolor{gray!6}{17513} & \cellcolor{gray!6}{56.137} & \cellcolor{gray!6}{22908} & \cellcolor{gray!6}{0.089} & \cellcolor{gray!6}{5395}\\
-\hspace{1em}Third 20\% & 4.259 & 15117 & 4.291 & 18309 & 0.032 & 3192\\
-\hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{1.094} & \cellcolor{gray!6}{2867} & \cellcolor{gray!6}{1.095} & \cellcolor{gray!6}{3116} & \cellcolor{gray!6}{0.001} & \cellcolor{gray!6}{249}\\
-\hspace{1em}Top 20\% & 6.256 & 4074 & 6.264 & 4540 & 0.008 & 466\\
+\multicolumn{8}{l}{\textbf{Threshold - 5 minutes}}\\
+\hspace{1em}Bottom 20\% & 43441 & 35477 & 1.302 & 35803 & 1.357 & 326 & 0.055\\
+\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{33312} & \cellcolor{gray!6}{17513} & \cellcolor{gray!6}{56.048} & \cellcolor{gray!6}{22908} & \cellcolor{gray!6}{56.137} & \cellcolor{gray!6}{5395} & \cellcolor{gray!6}{0.089}\\
+\hspace{1em}Third 20\% & 30940 & 15117 & 4.259 & 18309 & 4.291 & 3192 & 0.032\\
+\hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{20185} & \cellcolor{gray!6}{2867} & \cellcolor{gray!6}{1.094} & \cellcolor{gray!6}{3116} & \cellcolor{gray!6}{1.095} & \cellcolor{gray!6}{249} & \cellcolor{gray!6}{0.001}\\
+\hspace{1em}Top 20\% & 27541 & 4074 & 6.256 & 4540 & 6.264 & 466 & 0.008\\
 \addlinespace[0.3em]
-\multicolumn{7}{l}{\textbf{Threshold - 10 minutes}}\\
-\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{0.604} & \cellcolor{gray!6}{41824} & \cellcolor{gray!6}{0.622} & \cellcolor{gray!6}{41981} & \cellcolor{gray!6}{0.018} & \cellcolor{gray!6}{157}\\
-\hspace{1em}Second 20\% & 0.862 & 27546 & 0.929 & 30503 & 0.067 & 2957\\
-\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{0.776} & \cellcolor{gray!6}{22394} & \cellcolor{gray!6}{0.802} & \cellcolor{gray!6}{25128} & \cellcolor{gray!6}{0.026} & \cellcolor{gray!6}{2734}\\
-\hspace{1em}Fourth 20\% & 0.225 & 4544 & 0.227 & 4989 & 0.002 & 445\\
-\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{1.346} & \cellcolor{gray!6}{7989} & \cellcolor{gray!6}{1.348} & \cellcolor{gray!6}{9078} & \cellcolor{gray!6}{0.002} & \cellcolor{gray!6}{1089}\\
+\multicolumn{8}{l}{\textbf{Threshold - 10 minutes}}\\
+\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{43441} & \cellcolor{gray!6}{41824} & \cellcolor{gray!6}{0.604} & \cellcolor{gray!6}{41981} & \cellcolor{gray!6}{0.622} & \cellcolor{gray!6}{157} & \cellcolor{gray!6}{0.018}\\
+\hspace{1em}Second 20\% & 33312 & 27546 & 0.862 & 30503 & 0.929 & 2957 & 0.067\\
+\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{30940} & \cellcolor{gray!6}{22394} & \cellcolor{gray!6}{0.776} & \cellcolor{gray!6}{25128} & \cellcolor{gray!6}{0.802} & \cellcolor{gray!6}{2734} & \cellcolor{gray!6}{0.026}\\
+\hspace{1em}Fourth 20\% & 20185 & 4544 & 0.225 & 4989 & 0.227 & 445 & 0.002\\
+\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{27541} & \cellcolor{gray!6}{7989} & \cellcolor{gray!6}{1.346} & \cellcolor{gray!6}{9078} & \cellcolor{gray!6}{1.348} & \cellcolor{gray!6}{1089} & \cellcolor{gray!6}{0.002}\\
 \addlinespace[0.3em]
-\multicolumn{7}{l}{\textbf{Threshold - 15 minutes}}\\
-\hspace{1em}Bottom 20\% & 0.536 & 42208 & 0.557 & 42327 & 0.021 & 119\\
-\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{0.555} & \cellcolor{gray!6}{30507} & \cellcolor{gray!6}{0.614} & \cellcolor{gray!6}{31069} & \cellcolor{gray!6}{0.059} & \cellcolor{gray!6}{562}\\
-\hspace{1em}Third 20\% & 0.554 & 26108 & 0.581 & 26660 & 0.027 & 552\\
-\hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{0.093} & \cellcolor{gray!6}{6312} & \cellcolor{gray!6}{0.096} & \cellcolor{gray!6}{7435} & \cellcolor{gray!6}{0.003} & \cellcolor{gray!6}{1123}\\
-\hspace{1em}Top 20\% & 0.808 & 10209 & 0.811 & 11089 & 0.003 & 880\\
+\multicolumn{8}{l}{\textbf{Threshold - 15 minutes}}\\
+\hspace{1em}Bottom 20\% & 43441 & 42208 & 0.536 & 42327 & 0.557 & 119 & 0.021\\
+\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{33312} & \cellcolor{gray!6}{30507} & \cellcolor{gray!6}{0.555} & \cellcolor{gray!6}{31069} & \cellcolor{gray!6}{0.614} & \cellcolor{gray!6}{562} & \cellcolor{gray!6}{0.059}\\
+\hspace{1em}Third 20\% & 30940 & 26108 & 0.554 & 26660 & 0.581 & 552 & 0.027\\
+\hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{20185} & \cellcolor{gray!6}{6312} & \cellcolor{gray!6}{0.093} & \cellcolor{gray!6}{7435} & \cellcolor{gray!6}{0.096} & \cellcolor{gray!6}{1123} & \cellcolor{gray!6}{0.003}\\
+\hspace{1em}Top 20\% & 27541 & 10209 & 0.808 & 11089 & 0.811 & 880 & 0.003\\
 \bottomrule
-\multicolumn{7}{l}{\rule{0pt}{1em}\textit{Note: }}\\
-\multicolumn{7}{l}{\rule{0pt}{1em} }\\
-\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{a} With equity stations = Hamilton Bike Share current system (118 conventional stations, 12 equity stations)}\\
-\multicolumn{7}{l}{\rule{0pt}{1em}\textsuperscript{b} Without equity stations = Hamilton Bike Share original system (118 conventional stations, no equity stations)}\\
+\multicolumn{8}{l}{\rule{0pt}{1em}\textit{Note: }}\\
+\multicolumn{8}{l}{\rule{0pt}{1em} }\\
+\multicolumn{8}{l}{\rule{0pt}{1em}\textsuperscript{a} With equity stations = Hamilton Bike Share current system (118 conventional stations, 12 equity stations)}\\
+\multicolumn{8}{l}{\rule{0pt}{1em}\textsuperscript{b} Without equity stations = Hamilton Bike Share original system (118 conventional stations, no equity stations)}\\
 \end{tabular}}
@@
 
-Using disaggregate data, we examined spatial equity and accessibility to
-Hamilton Bike Share, with a particular focus on assessing the
-contribution of the program's equity stations. Our balanced floating
-catchment area approach, combined with pycnophylactic interpolation,
-enabled us to measure accessibility on a micro sale. This differentiates
-our analysis from similar papers exploring equity in PBSPs that use
-larger geographical units of analysis or that focus on station location
-instead of level of service. In this way, our paper has made
-contributions in a positive way by applying an intuitive and useful
+Using disaggregate data, in this paper we examined spatial equity and
+accessibility to Hamilton Bike Share, with a particular focus on
+assessing the contribution of the program's equity stations. Use of a
+balanced floating catchment area approach, combined with pycnophylactic
+interpolation, enabled us to measure accessibility on a micro sale. This
+differentiates our analysis from similar papers exploring equity in
+PBSPs that use larger geographical units of analysis or that focus on
+station location instead of level of service. In this way, our paper has
+made contributions in a positive way by applying an intuitive and useful
 approach to measure accessibility to a PBSP, and in a normative way by
@@
 reflects the normative guide, advertised on some bicycle share stations
-in Hamilton, that people can access other stations within a five minute
-walk if the station of origin has no bicycles. We find that over 118,000
-people can access a bicycle share station within a 15 minute walk, which
-represents roughly 85\% of the total population in the core service area
-{[}see Table \ref{tab:accessibility-income}{]}. At a minimum threshold
-of three minutes, too few stations can be accessed which leads to lower
-levels of accessibility. However, accessibility is at its lowest after
-eight minutes whereby congestion occurs due to increased potential
-demand. The City of Hamilton has recognized from the launch of the PBSP
-that significantly more stations and bicycles are needed to service the
-area ({\textbf{???}}). With a service area of 40 sq.km, Hamilton should
-have between 380 and 440 stations, instead of 130, and 1,500 bicycles
-instead of 900 ({\textbf{???}}). Reduced capacity within the system
-leads to gaps in coverage in some areas of the city ``with some areas
-not having the recommended station density of 300m between stations or
-10 stations per square km'' ({\textbf{???}}). This study illustrates the
-consequences of an imbalanced system whereby levels of accessibility are
-not equitable across income groups and become very low when congestion
-occurs.
-
-We found that equity stations achieved their goal of increasing
-horizontal equity in Hamilton's core service area. Panels
-1{[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
-{[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
-{[}\ref{fig:figure-bi-map-threshold-10}{]}, and 4
-{[}\ref{fig:figure-bi-map-threshold-15}{]} demonstrate how gaps in
-service were filled by these additional stations. In this respect, there
-are some commonalities between the expansions of Hamilton Bike Share and
+in Hamilton showing other stations within a five minute walk. We find
+that over 118,000 people can access a bicycle share station within a 15
+minute walk, which represents roughly 85\% of the total population in
+the core service area {[}see Table \ref{tab:accessibility-income}{]}. At
+a minimum threshold of three minutes, too few people can reach stations
+which leads to relatively high levels of service (since there is little
+crowding). However, accessibility is at its lowest after eight minutes
+whereby congestion effects due to increased potential demand kick in.
+The City of Hamilton has recognized from the launch of the PBSP that
+substantially more stations and bicycles are needed to service the area
+({\textbf{???}}). With a service area of 40 sq.km, it is estimated that
+Hamilton should have between 380 and 440 stations, instead of 130, and
+1,500 bicycles instead of 900 ({\textbf{???}}). Reduced capacity within
+the system leads to gaps in coverage in some areas of the city ``with
+some areas not having the recommended station density of 300m between
+stations or 10 stations per square km'' ({\textbf{???}}). This study
+illustrates the consequences of an imbalanced system whereby levels of
+accessibility are not equitable across income groups.
+
+We found that equity stations increased horizontal equity in Hamilton's
+core service area. Figures \ref{fig:figure-bi-map-threshold-3},
+\ref{fig:figure-bi-map-threshold-5},
+\ref{fig:figure-bi-map-threshold-10}, and
+\ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in service
+were filled by these additional stations. In this respect, there are
+some commonalities between the expansions of Hamilton Bike Share and
 that of Citi Bike in New York. Babgoli et al.~(2019) found a slight but
@@
 the Citi Bike expansion in 2015. Although the Citi Bike expansion was
-not specifically driven by a desire to reduce inequities in access, 16\%
-of neighbourhoods with the most poverty had stations compared to 12\%
-before. With the addition of equity stations, there were large gains in
-accessibility for the second 20\% at the average threshold, over 5,000
-more people, but much smaller gains for the bottom 20\% with only 326
-more people.
-
-Vertical inequities, however, continue to persist despite the addition
-of equity stations. While the addition of equity stations modestly
-increases accessibility for all income groups at all thresholds, they
-did not increase accessibility significantly for any single income
-group. Most importantly, individuals in the bottom 20\% median household
-income quintile have the second lowest level of access to Hamilton Bike
+not specifically driven by a desire to reduce inequity in access, 16\%
+of neighbourhoods with the highest levels of poverty had stations
+compared to 12\% before. With the addition of equity stations, there
+were large gains in accessibility for the second 20\% at the average
+threshold, over 5,000 more people, but much smaller gains for the bottom
+20\% with only 326 more people.
+
+Vertical inequity, however, persisted despite the addition of equity
+stations. While the addition of equity stations seems to modestly
+increase accessibility for all income groups at all thresholds, they did
+not increase accessibility substantially for any single income group.
+Most importantly, individuals in the bottom 20\% of median total
+household income have the second lowest level of access to Hamilton Bike
 Share at most thresholds (average, maximum, and extreme). At the minimum
@@
 dissemination areas \emph{across} the city not just within the core
-service area (Hosford and Winters, 2018). Instead, we use median
+service area (Hosford and Winters, 2018). Instead, we use median total
 household income for each dissemination area \emph{within} the core
@@
 to improve access not only for the bottom 20\% but for all income
-groups. Panels 1{[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
-{[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
-{[}\ref{fig:figure-bi-map-threshold-10}{]}, and 4
-{[}\ref{fig:figure-bi-map-threshold-15}{]} highlight potential locations
-for new equity stations.
+groups. Figures \ref{fig:figure-bi-map-threshold-3},
+\ref{fig:figure-bi-map-threshold-5},
+\ref{fig:figure-bi-map-threshold-10}, and
+\ref{fig:figure-bi-map-threshold-15} highlight potential locations for
+new equity stations.
 
@@
 
+An additional limitation is the lack of publicly available information
+about the number of bicycles at each station. Hamilton Bike share works
+to balance the number of bicycles in stations, but it is reasonable to
+expect that the number of bicycles will not match exactly the number of
+racks at every station. Ideally, instead of number of bicycle racks as
+our measure of supply, we would have liked to use the average number of
+bicycles at stations, perhaps at different times during the day too.
+Should these data become available, it would be worthwhile to revisit
+the question to examine how well operation of the system (including
+balancing of bicycles among stations) works to maintain the nominal
+levels of accessibility examined in this paper.
+
 \hypertarget{conclusion}{%
@@
 program in Hamilton, Ontario had the net effect of increasing
-accessibility and reducing both horizontal and vertical inequities. In
-particular, accessibility improved the most for those in the second 20\%
-median household income at all thresholds, but the gains were only
-modest for all income groups. Dissemination areas with the bottom 20\%
-had the lowest accessibility at three minutes, and second lowest levels
-of accessibility at five, ten, and fifteen minutes. Congestion effects
-were observed at higher thresholds, with accessibility decreasing
-significantly once the catchment area is increased to 10 minutes
-walking.
+accessibility and reducing to some extent both horizontal and vertical
+inequities. In particular, accessibility improved the most for those in
+the second 20\% median household income at all thresholds, but the gains
+were only modest for all income groups. Dissemination areas with the
+bottom 20\% had the lowest accessibility at three minutes, and second
+lowest levels of accessibility at five, ten, and fifteen minutes.
+Congestion effects were observed at higher thresholds, with
+accessibility decreasing significantly once the catchment area is
+increased to 10 minutes walking.
 
@@
 Hamilton Bike Share's equity stations increased ridership or resulted in
-new memberships in areas that were previously under-served. An
+new memberships in areas that were previously under-serviced. An
 examination of the types of trips undertaken by residents in these areas
@@
 conceptualized above, would serve as a valuable case study for Hamilton
-and other cities with PBSPs who wish to evaluate and address spatial
-inequities in accessibility and transportation options in urban areas.
+and other cities with PBSPs that wish to evaluate and address spatial
+inequuality in accessibility and transportation options in urban areas.
+
+\hypertarget{acknowledgments}{%
+\section{Acknowledgments}\label{acknowledgments}}
+
+This research was completed using open software, and the authors wish to
+acknowledge the developers of the following \texttt{R} packages:
+\texttt{biscale} (Prener et al., 2020), \texttt{cowplot} (Wilke, 2020),
+\texttt{data.table} (Dowle and Srinivasan, 2020), \texttt{disk.frame}
+(ZJ, 2020), \texttt{gdistance} (van Etten, 2020), \texttt{gridExtra}
+(Auguie, 2017), \texttt{kableExtra} (Zhu, 2020), \texttt{knitr} (Xie,
+2021a), \texttt{pycno} (Brunsdon, 2014), \texttt{r5r} (Saraiva et al.,
+2021), \texttt{raster} (Hijmans, 2020), \texttt{rgdal} (Bivand et al.,
+2021), \texttt{rticles} (Allaire et al., 2021), \texttt{sf} (Pebesma,
+2021), \texttt{tidyverse} (Wickham, 2019), \texttt{tinytex} (Xie,
+2021b), \texttt{units} (Pebesma et al., 2020).
 
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-R-rticles}{}%
+Allaire, J., Xie, Y., R Foundation, Wickham, H., Journal of Statistical
+Software, Vaidyanathan, R., Association for Computing Machinery,
+Boettiger, C., Elsevier, Broman, K., Mueller, K., Quast, B., Pruim, R.,
+Marwick, B., Wickham, C., Keyes, O., Yu, M., Emaasit, D., Onkelinx, T.,
+Gasparini, A., Desautels, M.-A., Leutnant, D., MDPI, Taylor and Francis,
+Ögreden, O., Hance, D., Nüst, D., Uvesten, P., Campitelli, E.,
+Muschelli, J., Hayes, A., Kamvar, Z.N., Ross, N., Cannoodt, R., Luguern,
+D., Kaplan, D.M., Kreutzer, S., Wang, S., Hesselberth, J., Dervieux, C.,
+2021. Rticles: Article formats for r markdown.
+
 \leavevmode\hypertarget{ref-auchinclossDesignBaselineDescription2020}{}%
@@
 
+\leavevmode\hypertarget{ref-R-gridExtra}{}%
+Auguie, B., 2017. GridExtra: Miscellaneous functions for "grid"
+graphics.
+
 \leavevmode\hypertarget{ref-babagoliExploringHealthSpatial2019}{}%
@@
 
+\leavevmode\hypertarget{ref-R-rgdal}{}%
+Bivand, R., Keitt, T., Rowlingson, B., 2021. Rgdal: Bindings for the
+geospatial data abstraction library.
+
 \leavevmode\hypertarget{ref-bivand2020progress}{}%
@@
 doi:\href{https://doi.org/10.1080/13504851.2016.1210760}{10.1080/13504851.2016.1210760}
+
+\leavevmode\hypertarget{ref-R-pycno}{}%
+Brunsdon, C., 2014. Pycno: Pycnophylactic interpolation.
 
@@
 
+\leavevmode\hypertarget{ref-R-data.table}{}%
+Dowle, M., Srinivasan, A., 2020. Data.table: Extension of `data.frame`.
+
 \leavevmode\hypertarget{ref-fishmanBikeshareReviewRecent2016}{}%
@@
 doi:\href{https://doi.org/10.1068/a291175}{10.1068/a291175}
+
+\leavevmode\hypertarget{ref-R-raster}{}%
+Hijmans, R.J., 2020. Raster: Geographic data analysis and modeling.
 
@@
 
+\leavevmode\hypertarget{ref-okelly2003aggregate}{}%
+O'Kelly, M.E., Horner, M.W., 2003. Aggregate accessibility to population
+at the county level: U.S. 1940-2000. Journal of Geographical Systems 5,
+5--23.
+
 \leavevmode\hypertarget{ref-paezDemandLevelService2019}{}%
@@
 
+\leavevmode\hypertarget{ref-R-sf}{}%
+Pebesma, E., 2021. Sf: Simple features for r.
+
+\leavevmode\hypertarget{ref-R-units}{}%
+Pebesma, E., Mailund, T., Kalinowski, T., 2020. Units: Measurement units
+for r vectors.
+
 \leavevmode\hypertarget{ref-pereiraGeographicAccessCOVID192021}{}%
@@
 
+\leavevmode\hypertarget{ref-R-biscale}{}%
+Prener, C., Grossenbacher, T., Zehr, A., 2020. Biscale: Tools and
+palettes for bivariate thematic mapping.
+
 \leavevmode\hypertarget{ref-qianBikesharingEquityDisadvantaged2020}{}%
@@
 
+\leavevmode\hypertarget{ref-R-r5r}{}%
+Saraiva, M., Pereira, R.H.M., Herszenhut, D., Braga, C.K.V., 2021. R5r:
+Rapid realistic routing with r5.
+
 \leavevmode\hypertarget{ref-scottWhatFactorsInfluence2019}{}%
@@
 
+\leavevmode\hypertarget{ref-R-gdistance}{}%
+van Etten, J., 2020. Gdistance: Distances and routes on geographical
+grids.
+
 \leavevmode\hypertarget{ref-wangNeighborhoodSociodemographicCharacteristics2019}{}%
@@
 
+\leavevmode\hypertarget{ref-R-tidyverse}{}%
+Wickham, H., 2019. Tidyverse: Easily install and load the tidyverse.
+
+\leavevmode\hypertarget{ref-R-cowplot}{}%
+Wilke, C.O., 2020. Cowplot: Streamlined plot theme and plot annotations
+for ggplot2.
+
 \leavevmode\hypertarget{ref-wintersWhoAreSuperusers2019}{}%
@@
 
+\leavevmode\hypertarget{ref-R-knitr}{}%
+Xie, Y., 2021a. Knitr: A general-purpose package for dynamic report
+generation in r.
+
+\leavevmode\hypertarget{ref-R-tinytex}{}%
+Xie, Y., 2021b. Tinytex: Helper functions to install and maintain tex
+live, and compile latex documents.
+
+\leavevmode\hypertarget{ref-R-kableExtra}{}%
+Zhu, H., 2020. KableExtra: Construct complex table with kable and pipe
+syntax.
+
+\leavevmode\hypertarget{ref-R-disk.frame}{}%
+ZJ, D., 2020. Disk.frame: Larger-than-ram disk-based data manipulation
+framework.
+
 \end{document}

### 2021-03-28 Bike-share-spatial-equity/Cover-Letter.Rmd

@@
 Please find attached our paper submitted to the Virtual Special Issue in Transportation Research Part D: Transport and Environment (SI: Transportation Justice).
-The paper is titled "Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach".
+The paper is titled "Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach".
 We use the city of Hamilton, Ontario in Canada as a case study.
@@
 1.	Our research contributes to the literature on the topic of equity in public bicycle share programs (PBSPs) which has received a lot of attention in North America recently.
-Other papers include Caspi & Noland, 2019, Hosford & Winters, 2018, Mooney et al., 2019, Qian et al., 2019, among others.
-The city of Hamilton is the only city in Canada with a PBSP that has an explicit equity program.
+Recent papers on this topic include Caspi & Noland, 2019, Hosford & Winters, 2018, Mooney et al., 2019, Qian et al., 2019, among others.
+To the best of our knowledge, the city of Hamilton is the only city in Canada with a PBSP that has an explicit equity program.
 This makes it an interesting place to examine how specific bike share stations added for the purpose of increasing equity have contributed to addressing horizontal and vertical inequities (see Chen et al., 2019).

### 2021-03-28 Bike-share-spatial-equity/Highlights.Rmd

@@
 
-* Contribution of equity stations to accessibility assessed at different thresholds
+* Contribution of equity stations to accessibility assessed at different thresholds for walking time to stations
 
-* Equity stations increase accessibility modestly for all income groups
+* Equity stations increased both horizontal and vertical accessibility, albeit only modestly
 

### 2021-03-29 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Final proof-read and edit before submission

@@
             pdfauthor={},
-            pdftitle={Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach},
+            pdftitle={Examining horizontal and vertical equity to a public bicycle share program using a balanced floating catchment area accessibility approach},
             colorlinks=false,
@@
 
-  \title{Examining spatial equity in accessibility to a public bicycle share
-program using a balanced floating catchment area approach}
+  \title{Examining horizontal and vertical equity to a public bicycle share
+program using a balanced floating catchment area accessibility approach}
     \author[McMaster University]{Elise Desjardins\corref{1}}
@@
    \ead{paezha@mcmaster.ca} 
-      \address[McMaster University]{School of Earth, Environment \& Society, 1280 Main Street West,
-Hamilton, ON L8S4L8}
-    \address[University of Toronto Scarborough]{Department of Geography \& Planning, 1265 Military Trail, Toronto, ON
-M1C1A4}
+      \address[McMaster University]{School of Earth, Environment \& Society, McMaster University, 1280 Main
+Street West, Hamilton, ON L8S4L8}
+    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
+1265 Military Trail, Toronto, ON M1C1A4}
       \cortext[1]{Corresponding Author}
@@
   \begin{abstract}
-  Public bicycle share programs (PBSPs) are implemented to promote cycling
-  as a convenient and sustainable mode of transportation. These systems
-  can also play a role in addressing transportation needs and advancing
-  transportation equity if they make cycling more accessible to lower
-  income or disadvantaged populations. In Ontario, the City of Hamilton
-  launched a public bicycle share program in 2015 that currently has over
-  900 operational bicycles and 130 docking stations. This system was
-  expanded in 2018 by an equity initiative that added twelve docking
-  stations with the explicit objective of increasing spatial equity in
-  access. Since the cost of reaching a bicycle share station decreases the
-  potential of accessing the program, the location of stations matters,
-  and previous research found that Hamilton's public bicycle share program
-  targeted well disadvantaged areas in the city. In this research, we
-  revisit the case of Hamilton and investigate differentials in
-  accessibility to bicycle share stations using a balanced floating
-  catchment area (BFCA) accessibility approach. We compare accessibility
-  with and without the equity stations to assess the effect of the
-  initiative. We implement our analysis parting from micro zones to better
-  reflect walking to a docking station, and conduct a sensitivity analysis
-  at several walking time thresholds. We then reaggregate the estimated
+  Public bicycle share programs (PBSPs) can play a role in advancing
+  transportation equity if they make cycling more accessible to
+  disadvantaged populations. In Ontario, the City of Hamilton expanded
+  their PBSP in 2018 by adding twelve docking stations with the explicit
+  objective of increasing spatial equity in access. In this case study, we
+  investigate differentials in accessibility to stations using a balanced
+  floating catchment area accessibility approach and compare accessibility
+  with and without the equity stations. We analyze micro zones to better
+  reflect walking to a station and conduct a sensitivity analysis at
+  several walking time thresholds. We then reaggregate the estimated
   accessibility for further analysis using census data. Our findings
-  indicate that the addition of equity stations increased the horizontal
-  equity of the system by growing the serviced population at every
-  threshold examined. With respect to vertical equity, although
-  accessibility increased for the whole population, this increase was
-  relatively modest especially for population in the bottom 20\% of median
-  total household income.
+  indicate that equity stations increased the serviced population at every
+  threshold examined. Although accessibility increased for the whole
+  population, this increase was relatively modest especially for
+  population in the bottom 20\% of median total household income.
   \end{abstract}
@@
 \hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{1. Introduction}\label{introduction}}
 
@@
 and last mile trips to other public transportation instead of using
-personal vehicles. These systems can also play a role in addressing
+personal vehicles. These programs can also play a role in addressing
 transportation needs and advancing transportation equity if they make
-cycling more accessible to lower income or disadvantaged populations. As
-these systems become increasingly common, their introduction has been
-accompanied by a flurry of research that investigates them from the
-perspective of spatial and transportation equity (e.g., Smith et al.,
-2015; Hosford and Winters, 2018; Hull Grasso et al., 2020; Mooney et
-al., 2019; Qian and Jaller, 2020; Qian et al., 2020). Indeed, although
-these programs are available to the general public and ought to be
-accessible to any individuals who wish to use them, research on PBSPs
-located in North American and European cities indicates that inequity in
-spatial accessibility needs redressing.
-
-In this paper we investigate the case of the PBSP in the City of
-Hamilton, in Ontario, Canada. Hamilton Bike Share (previously also known
-as Social Bicycles, or SoBi) was launched in 2015 and currently has over
-900 operational bicycles and 130 docking stations. An interesting
-feature of this system is that it was expanded by an equity initiative
-that introduced twelve docking stations in 2018 with the explicit
-objective of increasing spatial accessibility. Hamilton Bike Share was
-already studied by Hosford and Winters (2018) as part of a selection of
-PBSPs in Canada. These authors found that most of the cities in their
-research could benefit of greater efforts to expand service to areas
-with lower socio-economic populations - but Hamilton fared somewhat
+cycling more accessible to individuals and groups with lower
+socioeconomic status. As these programs become increasingly common,
+their introduction has been accompanied by a flurry of research that
+investigates them from the perspective of spatial and transportation
+equity (e.g., Hosford and Winters, 2018; Hull Grasso et al., 2020;
+Mooney et al., 2019; Qian and Jaller, 2021, 2020; Smith et al., 2015).
+Indeed, although these programs are available to the general public and
+ought to be accessible to any individual who wishes to use them,
+research on PBSPs located in North American and European cities
+indicates that inequity in spatial accessibility needs redressing.
+
+In this paper, we investigate the case of the PBSP in the city of
+Hamilton, in Ontario, Canada. Hamilton Bike Share was launched in 2015
+and currently has over 900 operational bicycles and 130 docking
+stations. An interesting feature of this program is that it was expanded
+by an equity initiative that introduced twelve stations in 2018 with the
+explicit objective of increasing spatial accessibility. Hamilton Bike
+Share was already studied by Hosford and Winters (2018) as part of a
+selection of PBSPs in Canada. The authors found that most of the cities
+in their research could benefit of greater efforts to expand service to
+areas with lower socioeconomic groups - but Hamilton fared somewhat
 better than other cities in this respect. What distinguishes our
-research is the use of a balanced floating catchment area (BFCA)
-accessibility approach that accounts for the supply of stations and the
-potential demand from the population serviced. This requires a more
-disaggregated approach than the use of census geographies (q.v., Hosford
-and Winters, 2018), since many trips to a bike share station likely
-happen at a level lower than even the smallest census geography. For
-this reason, we implement our analysis parting from micro population
-zones to better reflect the friction of walking to a docking station.
-Further, we conduct a sensitivity analysis at several walking time
-thresholds. In terms of spatial equity, we compare accessibility with
-and without the equity stations to assess the effect of the initiative,
-before reaggregating the data for further analysis using median total
-household income information from the census. We find that the addition
-of equity stations does increase the serviced population at every
-threshold examined. However, although accessibility increased for the
-whole population, this increase was relatively modest, especially for
-population in the bottom 20\% of median total household income. The
-analysis also identifies areas under-serviced within the core service
-area of the system.
+research from existing literature is the use of a balanced floating
+catchment area (BFCA) accessibility approach that accounts for the
+supply of stations and the potential demand from the population
+serviced. This requires a more disaggregated approach than the use of
+census geographies (q.v., Hosford and Winters, 2018), since walking
+trips to a bicycle share station likely happen at a level lower than
+even the smallest census geography. For this reason, we implement our
+analysis parting from micro population zones to better reflect the
+friction of walking to a docking station, which is an important
+component of a bike share trip (Chen et al., 2019). Further, we conduct
+a sensitivity analysis at several walking time thresholds. In terms of
+spatial equity, we compare accessibility with and without the equity
+stations to assess the effect of the initiative, before reaggregating
+the data for further analysis using median total household income
+information from the census.
 
@@
 2020; Lovelace, 2021). All data were obtained from publicly available
-sources and organized in the form a data package. Following best
+sources and organized in the form of a data package. Following best
 practices in spatial data science (Brunsdon and Comber, 2020), the code
-and data needed to reproduce (or modify/extend) the analysis are
-available for download.\footnote{https://drive.google.com/drive/folders/1ZGRpSN2LxS2Fq2tLVJgfm1sI6I26vzZh?usp=sharing}
+and data needed to reproduce, modify or extend the analysis are
+available for download.\footnote{https://github.com/paezha/Accessibility-Sobi-Hamilton}
 
 \hypertarget{literature-review}{%
-\section{Literature Review}\label{literature-review}}
+\section{2. Literature Review}\label{literature-review}}
 
@@
 (Fishman, 2016). In many cities, males use bike share more than females
-({\textbf{???}}; Brey et al., 2017; Nickkar et al., 2019; Reilly, S. M.
-Wang, et al., 2020; Winters et al., 2019) as do younger age cohorts
-(Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011). However,
-one study found that bike share users in Washington, DC were more likely
-to be female (Buck et al., 2013), which suggests that the gender gap
-among cyclists who use bike share is less disparate than the gap for
-personal bicycle use (Fishman, 2016). There is some evidence that bike
-share users are less likely to own a car (Buck et al., 2013; Reilly,
-Noyes, et al., 2020). However, the relationship between income or
-education and bike share use is less clear-cut. Stations in
+(Brey et al., 2017; Nickkar et al., 2019; Ogilvie and Goodman, 2012;
+Reilly, S. M. Wang, et al., 2020; Winters et al., 2019) as do younger
+age cohorts (Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011).
+However, one study found that bike share users in Washington, DC were
+more likely to be female (Buck et al., 2013), which suggests that the
+gender gap among cyclists who use bike share is less disparate than the
+gap for personal bicycle use (Fishman, 2016). There is some evidence
+that bike share users are less likely to own a car (Buck et al., 2013;
+Reilly, Noyes, et al., 2020). However, the relationship between income
+or education and bike share use is less clear-cut. Stations in
 disadvantaged communities in Chicago have been found to generate most of
 the average annual trips (Qian and Jaller, 2020) and individuals from
-minority or lower socioeconomic status neighbourhoods in
+minority or lower socioeconomic status neighborhoods in
 Minneapolis-St.~Paul used the city's PBSP more (J. Wang and Lindsey,
-2019a). Similar findings were reported in London ({\textbf{???}}). Being
-university educated was a significant correlate of bike share use in
-Montreal, Canada (Fuller et al., 2011). Perhaps not coincidentally,
-financial savings have been found to motivate those on a low income to
-use bike share (Fishman, 2016).
+2019a). Similar findings were reported in London (Ogilvie and Goodman,
+2012). Being university educated was a significant correlate of bike
+share use in Montreal, Canada (Fuller et al., 2011). Perhaps not
+coincidentally, financial savings have been found to motivate those on a
+low income to use bike share (Fishman, 2016).
 
@@
 more likely to use services or programs that they can easily reach.
-Several studies have recently explored the equity of PBSPs by primarily
-examining who has access to bike share (e.g., differences by
-demographics or socioeconomic status) and where stations are located. It
-is important to note that equity can be achieved in two different ways:
-equal spatial distribution across a region (e.g., horizontal equity) or
-greater access for vulnerable or disadvantaged populations (e.g.,
-vertical equity) (Chen et al., 2019). Both are of interest to
-researchers and transport planners since they are often linked in that
-advantage, or conversely disadvantage, has spatial patterns. Using a
-negative binomial regression model, Qian and Jaller (2020) estimated
-ridership in Chicago's PBSP among disadvantaged communities and found
-some disparities. While annual members in disadvantaged communities have
-a significantly lower share of trips compared to other areas in the
-city, they make longer trips. This suggests that they may be using PBSPs
-for utilitarian trips (e.g., commuting), which points to the importance
-of ensuring equitable access (Qian and Jaller, 2020). Similar results
-were found in Philadelphia, where lower income areas generated fewer
-trips and it was suggested that efforts to increase equity within the
-program have not been as successful as intended (Caspi and Noland,
-2019). In the case of Seattle, all neighbourhoods had some level of
-access to dockless bikes but those with higher incomes and more
-residents of higher education had more bikes (Mooney et al., 2019).
-Babagoli et al.~(2019) also found that neighbourhoods in New York City
-with higher affluence had the greatest proportion of Citi Bike stations.
-Researchers have recommended targeted expansion of stations to areas
-that use bike share but have inequitable access ({\textbf{???}}).
-
-On the whole, existing studies highlight the need for bicycle share
-systems to be more highly accessible for diverse populations in order to
-increase use beyond the ``typical'' users. This has been the focus of
-recent research (see, among others, Auchincloss et al., 2020; Hull
-Grasso et al., 2020; MacArthur et al., 2020). Offering more people the
-option of using sustainable and active transportation, particularly
-those who have lower socioeconomic status and might benefit the most, is
-a worthy policy goal for cities with PBSPs. Several cities have launched
+Several studies have recently explored the equity of PBSPs in North
+America by primarily examining who has access to bike share (e.g.,
+differences by demographics or socioeconomic status) and where stations
+are located. It is important to note that equity can be achieved in two
+different ways: balanced or equal spatial distribution across a region
+for all similar groups (e.g., horizontal equity) or greater and targeted
+access to accommodate vulnerable or disadvantaged populations (e.g.,
+vertical equity) (see Chen et al., 2019). Both are of interest to
+researchers and transportation planners since they are often linked in
+that advantage, or conversely disadvantage, has spatial patterns. Using
+a negative binomial regression model, Qian and Jaller (2020) estimated
+ridership in Chicago's PBSP and found some disparities. A minority of
+bike share stations were found to be located in disadvantaged
+communities, while annual members from disadvantaged communities have a
+lower share of trips compared to other areas in the city. This suggests
+that they may be using PBSPs for utilitarian trips (e.g., commuting),
+which points to the importance of ensuring equitable access (Qian and
+Jaller, 2020). Similar results were found in Philadelphia. Despite
+efforts to increase equity within the city's PBSP, census block groups
+with lower median income generated fewer trips which suggested that
+efforts have not been as successful as intended (Caspi and Noland,
+2019). Trips from stations in such areas are utilitarian (e.g.,
+commuting to work), which points to the importance of ensuring equitable
+access (Caspi and Noland, 2019). In the case of Seattle, all
+neighborhoods had some level of access to dockless bicycles but those
+with higher incomes and more residents of higher education had more
+bikes (Mooney et al., 2019). Babagoli et al.~(2019) also found that
+neighborhoods in New York City with higher affluence had the greatest
+proportion of Citi Bike stations.
+
+On the whole, existing studies highlight the need for PBSPs to be more
+highly accessible for diverse populations in order to increase use
+beyond the ``typical'' users. This has been the focus of recent research
+(see, among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
+MacArthur et al., 2020). Offering more people the option of using
+sustainable and active transportation, particularly those who have lower
+socioeconomic status and might benefit the most, is a worthy policy goal
+for cities with PBSPs. A few North American cities have launched
 specific programs to address barriers to bike share or to expand service
-to more deprived areas: London, United Kingdom ({\textbf{???}});
-Philadelphia, Pennsylvania (Caspi and Noland, 2019), and Hamilton,
-Ontario (Hosford and Winters, 2018). Findings to date suggest that
-horizontal equity can be easier to achieve than vertical equity.
-However, exploring transportation equity by investigating where bicycle
-share stations are located, often using neighbourhoods or census tracts
-as the geographical unit of analysis, can ignore or miss the benefits
-that may be derived from adjacent zones. Meaning that, stations may be
-lacking in certain neighbourhoods but there may be stations accessible
-within a reasonable walking time. This is where geographical
-accessibility becomes an important consideration.
+to more deprived areas: Philadelphia, Pennsylvania (Caspi and Noland,
+2019), and Hamilton, Ontario (Hosford and Winters, 2018). However,
+exploring transportation equity by investigating where bicycle share
+stations are located, often using neighborhoods or census tracts as the
+geographical unit of analysis, can ignore or miss the benefits that may
+be derived from adjacent zones. Meaning that, stations may be lacking in
+certain neighborhoods but there may be stations accessible within a
+reasonable walking time. This is where geographical accessibility
+becomes an important consideration.
 
@@
 this field has evolved over the past century and has increasingly become
-linked with recent interest in prioritizing local proximity and modes
-that are suitable for local trips like walking and cycling (Levine,
-2020). As a measure of the ease of reaching potential destinations
-spread spatially in a given area, accessibility is relevant to PBSPs
-because it can identify current inequaltiy in the provision of
-infrastructure, as well as guide interventions that increase access for
-groups that are under-serviced or address gaps in transportation
-options. It also addresses some of the challenges of other performance
-measures such as level of service within a transportation network by
-measuring person-based indicators and exploring differences in use
-between population groups (Páez et al., 2012).
+linked with recent planning interests in prioritizing modes that are
+suitable for local trips like walking and cycling (Levine, 2020). As a
+measure of the ease of reaching potential destinations spread spatially
+in a given area, accessibility is relevant to PBSPs because it can
+identify current inequality in the provision of infrastructure, as well
+as guide interventions that increase access for groups that are
+under-serviced or address gaps in transportation options. It also
+addresses some of the challenges of other performance measures such as
+level of service within a transportation network by measuring
+person-based indicators and exploring differences in use between
+population groups (Páez et al., 2012).
 
@@
 crowding, like health care services (e.g., Pereira, Carlos Kauê Vieira
-Braga, et al., 2021), the utility of stations is limited by the number
-of bicycles that they can hold. The program may not necessarily be
-improved if stations are easy to reach but offer only a small number of
-bicycles. Likewise, more people may not opt to use the program if the
-supply of bicycles available at nearby stations is insufficient to meet
-demand. The location and size of stations is important to increase the
-utility of this public transportation option for more people, thus
-achieving vertical equity. Accessibility analyses for PBSPs constitute a
-positive and evaluation-based approach that also has the potential to
-inform equity efforts. For instance, Wang and Lindsey (2019b)
-investigated whether new or relocated bicycle share stations increased
-accessibility and use, which offered important insights to improve the
-performance of the program.
+Braga, et al., 2021), the utility of stations is limited by the
+\emph{maximum} number of bicycles that they can hold. The program may
+not necessarily be improved if stations are easy to reach but offer only
+a small number of bicycles. Likewise, more people may not opt to use the
+program if the supply of bicycles available at nearby stations is
+insufficient to meet demand. The location and size of stations is
+important to increase the utility of this public transportation option
+for more people, thus achieving vertical equity. Accessibility analyses
+for PBSPs constitute a positive and evaluation-based approach that also
+has the potential to inform equity efforts. For instance, Wang and
+Lindsey (2019b) investigated whether new or relocated bicycle share
+stations increased accessibility and use, which offered important
+insights to improve the performance of the program.
 
@@
 reaching the facility. Next, these service level ratios are allocated
-back to the population centres and summarized as a measure of congested
+back to the population centers and summarized as a measure of congested
 accessibility. Thus, this model does a good job of considering potential
 crowding or competition for services. While there have been many
-methodological innovations in FCA methods (for example, {\textbf{???}};
-{\textbf{???}}; {\textbf{???}}; {\textbf{???}}; {\textbf{???}}), a
-recent improvement to this approach was achieved through a simple and
-intuitive balancing of the impedance that addressed the effects of
-demand and service inflation found in earlier FCA approaches (see Paez
-et al., 2019).
+methodological innovations in FCA methods (for example, Delamater, 2013;
+Luo and Qi, 2009; Luo and Wang, 2003; Radke and Mu, 2000; Wan et al.,
+2012), a recent improvement to this approach was achieved through a
+simple and intuitive balancing of the impedance that addressed the
+effects of demand and service inflation found in earlier FCA approaches
+(see Paez et al., 2019).
 
@@
 spatial zones. Previous research on bike share equity has typically used
-a meso- or macro-level approach with aggregate data from entire
-neighbourhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
+a meso- or macro-level approach with aggregated data from entire
+neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
 2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
@@
 estimate for individuals, which is useful for addressing vertical
-inequity in PBSP usage.
+inequity in PBSP usage. Chen et al.~(2019, p. 530) are in favour of
+using disaggregated data, which they did in their recent analysis of
+Tampa's PBSP, because they note ``the use of aggregated data might
+hinder our understanding of the equity impacts since individual
+disparities are absorbed after aggregation''.
 
@@
 novel approach that has not been used yet in cycling research, with
-disaggregate population-level data, in this paper we examine
-accessibility to a public bicycle share program in Hamilton, Ontario. We
-(1) conduct a sensitivity analysis by measuring accessibility and level
-of service to bike share stations at different walking time thresholds
+disaggregated population data, we examine accessibility to a public
+bicycle share program in Hamilton, Ontario. In this paper, we (1)
+conduct a sensitivity analysis by measuring accessibility and level of
+service to bicycle share stations at different walking time thresholds
 to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes;
@@
 \hypertarget{sec:study}{%
-\section{Case Study}\label{sec:study}}
+\section{3. Case Study}\label{sec:study}}
 
 \hypertarget{original-system}{%
-\subsection{Original System}\label{original-system}}
+\subsection{3.1. Original System}\label{original-system}}
 
@@
 Canada. The city launched a public bicycle share program in March 2015
-with 115 stations and 750 bicycles (Hamilton, 2015). Before June 2020,
-the program was known as Social Bicycles or SoBi Hamilton but is now
-called Hamilton Bike Share. Stations are spaced between 300 and 600
-metres apart (Scott et al., 2021). The core service area spans 40 sq.km
-of the city although it was planned to be 20 sq.km ({\textbf{???}}), and
+with 115 stations and 750 bicycles (Hamilton, 2015a). Before June 2020,
+the program was known as Social Bicycles or SoBi Hamilton, but is now
+called Hamilton Bike Share. Stations are spaced between 300 and 600 m
+apart (Scott et al., 2021). The core service area spans 40 sq.km of the
+city although it was planned to be 20 sq.km (Hamilton, 2015b), and
 roughly 138,000 people are within 30 minutes walking of a bike share
@@
 locations of bicycle share stations that had been selected and to
-crowdsource potential locations for additional locations
-({\textbf{???}}). Most of the suggested locations were in the east end
-of the core service area that lack stations or in neighbourhoods not
-serviced by the PBSP. The program was enthusiastically welcomed in the
-city in 2015 - within three weeks of launching, 10,000 trips had been
-made (Hamilton, 2015), however inadequate coverage given the size of the
-Hamilton Bike Share's service area was identified as a problem early on,
-and transportation planners noted that the small size (i.e., supply of
+crowdsource potential locations for additional locations (Hamilton,
+2014). Most of the suggested locations were in the east end of the core
+service area that lack stations or in neighborhoods not serviced by the
+PBSP. The program was enthusiastically welcomed in the city in 2015 -
+within three weeks of launching, 10,000 trips had been made (Hamilton,
+2015a), however inadequate coverage given the size of Hamilton Bike
+Share's service area was identified as a problem early on, and
+transportation planners noted that the small size (i.e., supply of
 bicycle racks) and low quantity of stations would lead to challenges in
-balancing the system ({\textbf{???}}). Based on public feedback, 13
-stations were added to the original system.
+balancing the system (Hamilton, 2015b). Based on public feedback,
+thirteen stations were added to the original system.
 
 \hypertarget{equity-initiative}{%
-\subsection{Equity Initiative}\label{equity-initiative}}
+\subsection{3.2. Equity Initiative}\label{equity-initiative}}
 
@@
 bicycles and stations were added to the program which expanded it to
-more disadvantaged areas in the core service area {[}see Figure 2{]}.
-The program also offers subsidized memberships to individuals who
-identify as low income, and complements these initiatives with cycle
-skills education. A comparable program can be found in Philadelphia (see
-Caspi and Noland, 2019).
+more disadvantaged areas in the core service area. The program also
+offers subsidized memberships to individuals who identify as low income,
+and complements this service with cycle skills education. A comparable
+program can be found in Philadelphia (see Caspi and Noland, 2019).
 
 \hypertarget{current-system}{%
-\subsection{Current System}\label{current-system}}
+\subsection{3.3. Current System}\label{current-system}}
 
@@
 Figure \ref{fig:sobi-stations-in-hamilton}{]}, and over 26,000 active
-memberships (Hamilton, 2015). The core service area remains 40 sq.km.
-The PBSP has 13 stations that were added as part of the ERI initiative;
-we refer to these as ``equity stations'' throughout the paper while all
-other stations are ``conventional stations''. In total, 117 stations are
-``conventional'' and 13 are ``equity''.
+memberships (Hamilton, 2015a). The core service area remains 40 sq.km.
+The PBSP has twelve stations that were added as part of the ERI
+initiative; we refer to these as ``equity stations'' throughout the
+paper while all other stations are ``conventional stations''. In total,
+118 stations are ``conventional'' and 12 are ``equity''.
 
 \hypertarget{membership}{%
-\subsection{Membership}\label{membership}}
+\subsection{3.4. Membership}\label{membership}}
 
@@
 useful for certain kinds of trips but not all, meaning that some trips
-continue to require a private vehicle.
+continue to require a personal vehicle.
 
 \hypertarget{relevant-research}{%
-\subsection{Relevant Research}\label{relevant-research}}
+\subsection{3.5. Relevant Research}\label{relevant-research}}
 
@@
 
-\caption{The core service area of Hamilton Bike Share is outlined in black. Hamilton Census Metropolitan Area (CMA) is shown in grey.}\label{fig:hamilton-and-sobi-service-area}
+\caption{The core service area of Hamilton Bike Share is outlined in black. Hamilton Census Metropolitan Area is shown in grey.}\label{fig:hamilton-and-sobi-service-area}
 \end{figure}
@@
 
-\caption{The spatial distribution of bike share stations in Hamilton, Ontario.
-The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in dark green.
-Hamilton CMA is shown in grey.}\label{fig:sobi-stations-in-hamilton}
+\caption{The spatial distribution of bicycle share stations in Hamilton, Ontario.
+The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in pink.
+Hamilton Census Metropolitan Area is shown in grey.}\label{fig:sobi-stations-in-hamilton}
 \end{figure}
@@
 \hypertarget{sec:methods}{%
-\section{Methods}\label{sec:methods}}
+\section{4. Methods and Data}\label{sec:methods}}
 
 \hypertarget{floating-catchment-area}{%
-\subsection{Floating Catchment Area}\label{floating-catchment-area}}
+\subsection{4.1. Floating Catchment
+Area}\label{floating-catchment-area}}
 
@@
 access to the service (Paez et al., 2019). In particular, the Two-Step
-Floating Catchment Area (2SFCA) method ({\textbf{???}}; {\textbf{???}})
-produces flexible catchment areas instead of using rigid boundaries like
-PPR. This provides more useful information because it does not assume
-that people are limited to service within pre-defined boundaries (Paez
-et al., 2019). This is an important property that supports our rationale
-for applying this method to measure accessibility to Hamilton Bike
-Share. The City of Hamilton has positioned stations between 300 and 600
-metres apart, but anticipates that hubs will service those living within
-a 250 metre buffer from the station ({\textbf{???}}). The latter
-constitutes a normative statement: people ought to be able to access a
-station in less than 600 metres if they live in the core service area,
-with usage coming from 250 metres around. However, it is not known how
+Floating Catchment Area (2SFCA) method (Luo and Wang, 2003; Radke and
+Mu, 2000) produces flexible catchment areas instead of using rigid
+boundaries like PPR. This provides more useful information because it
+does not assume that people are limited to service within pre-defined
+boundaries (Paez et al., 2019). This is an important property that
+supports our rationale for applying this method to measure accessibility
+to Hamilton Bike Share. The City of Hamilton has positioned stations
+between 300 and 600 m apart, but anticipates that hubs will service
+those living within a 250 m buffer from the station (Hamilton, 2015b).
+The latter constitutes a normative statement: people ought to be able to
+access a station in less than 600 m if they live in the core service
+area, with usage coming from 250 m around. However, it is not known how
 far people are actually willing to travel to reach a station. It would
@@
 
-More recently, the \emph{balanced} floating catchment area (FCA)
-approach was developed to address issues with demand and supply
-inflation that result from the overlapping catchment areas produced by
-earlier FCA methods (Paez et al., 2019). Briefly, overlapping catchment
-areas lead to inflation of population totals and deflation of service
-levels across a study area and generates inaccurate or misleading
-accessibility estimates. In contrast, Paez et al.~(2019) adjust the
-impedance weights so that both supply and demand are proportionally
-allocated. The results is a FCA method that balances the population and
-level of service by eliminating the over-counting of population and
-level of service that leads to distortions in demand and supply. Other
-benefits of this adjusted method include consideration of competition
-which can occur when catchment areas overlap, as well as the
-preservation of population and level of service. Balanced floating
-catchment area methods have been used to explore accessibility to health
-care providers (Paez et al., 2019) and COVID-19 health care services
-(Pereira, Carlos Kauê Vieira Braga, et al., 2021), but have not yet been
-used in the cycling literature to explore issues of accessibility.
+More recently, the \emph{balanced} floating catchment area approach was
+developed to address issues with demand and supply inflation that result
+from the overlapping catchment areas produced by earlier FCA methods
+(Paez et al., 2019). Briefly, overlapping catchment areas lead to
+inflation of population totals and deflation of service levels across a
+study area and generates inaccurate or misleading accessibility
+estimates. In contrast, Paez et al.~(2019) adjusted the impedance
+weights so that both supply and demand are proportionally allocated. The
+result is a FCA method that balances the population and level of service
+by eliminating the over-counting of population and level of service that
+leads to distortions in demand and supply. Other benefits of this
+adjusted method include consideration of competition which can occur
+when catchment areas overlap, as well as the preservation of population
+and level of service. Balanced floating catchment area methods have been
+used to explore accessibility to health care providers (Paez et al.,
+2019) and COVID-19 health care services (Pereira, Carlos Kauê Vieira
+Braga, et al., 2021), but have not yet been used in the cycling
+literature to explore issues of accessibility.
 
@@
 P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}
-\] As seen in the equation above, the population allocated to station
-\(j\) is the weighted sum of the population in the region; a spatial
-weight \(w_{ij}\) represents the friction that the population at \(i\)
-faces when reaching station \(j\), and is usually given by a
-distance-decay function, so that each station is assumed to service only
-a segment of the population within a limited geographical range.
+\]
+
+As seen in the equation above, the population allocated to station \(j\)
+is the weighted sum of the population in the region; a spatial weight
+\(w_{ij}\) represents the friction that the population at \(i\) faces
+when reaching station \(j\), and is usually given by a distance-decay
+function, so that each station is assumed to service only a segment of
+the population within a limited geographical range.
 
@@
 
-The balanced approach of Paez et al. (2019) replaces the spatial weights
+The balanced approach of Paez et al (2019) replaces the spatial weights
 with normalized versions as follows: \[
@@
 
-With these weights accessibility can be calculated without risk of
+With these weights, accessibility can be calculated without risk of
 demand or supply inflation: \[
 A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}
-\] By allocating the population and level of service proportionally,
-this method preserves the values of the population and level of service
-and provides a regional provider-to-population ratio since: \[
+\]
+
+By allocating the population and level of service proportionally, this
+method preserves the values of the population and level of service and
+provides a regional provider-to-population ratio since: \[
 \sum_{i=1}^n A_i = \sum_{j=1}^J L_j = \frac{\sum_{j=1}^JS_j}{\sum_{i=1}^n P_i}
@@
 location-based and person-based approach to calculating accessibility
-using disaggregate population data. In their review of accessibility
+using disaggregated population data. In their review of accessibility
 measures, Geurs and van Wee (2004) highlight the need for greater
@@
 catchment area approach allows us to examine accessibility by
-stratifying according to median total household income, which would
-constitute the individual component of the accessibility measure (Geurs
+stratifying according to median total household income, which
+constitutes the individual component of the accessibility measure (Geurs
 and van Wee, 2004). However, conducting a further sensitivity analysis
-to measure accessibility at different distance thresholds would help to
-consider potential spatio-temporal constraints. Different people may be
-willing to travel different distances to access a public bicycle share
-station.
+to measure accessibility at different walking time thresholds would help
+to consider potential spatio-temporal constraints. Different people may
+be willing to travel different distances to access a public bicycle
+share station.
 
 \hypertarget{pycnophylactic-interpolation}{%
-\subsection{Pycnophylactic
+\subsection{4.2. Pycnophylactic
 Interpolation}\label{pycnophylactic-interpolation}}
 
-To obtain population at sub-census geography levels (at the micro scale)
-we use pycnophylactic interpolation (Tobler, 1979). We obtained
+To obtain population at sub-census geography levels (at the micro
+scale), we use pycnophylactic interpolation (Tobler, 1979). We obtained
 population data from the 2016 Census of Canada for dissemination areas
-(DAs, the smallest publicly available census geography in Canada). This
-zonal values of the population were interpolated to smaller polygons
-50-by-50 m in size. Pycnophilactic interpolation involves smoothing out
-the population from each dissemination area while preserving total
-volume {[}see Figure \ref{fig:da-population}{]}. When interpolating the
-population at this high level of resolution it is important to ensure
-that population numbers were not allocated to areas where people do not
-live in Hamilton (e.g., to parks, large institutional buildings, etc.)
-To do so, we retrieved shapefiles for various geographic features {[}see
-Table \ref{tab:data-features}{]} from Open Hamilton. Next, we removed
-these features from the PBSP core service area and used pycnophylactic
-interpolation to disaggregate and reallocate population within the
-remaining area {[}see Figure \ref{fig:interpolated-population}{]}.
+(DA), which is the smallest publicly available census geography in
+Canada. These zonal values of the population were interpolated to
+smaller polygons 50-by-50 m in size. Pycnophylactic interpolation
+involves smoothing out the population from each dissemination area while
+preserving total volume {[}see Figure \ref{fig:da-population}{]}. When
+interpolating the population at this high level of resolution, it is
+important to ensure that population numbers were not allocated to areas
+where people do not live in Hamilton (for example, to parks, large
+institutional buildings, etc.). To do so, we retrieved shapefiles for
+various geographic features {[}see Table \ref{tab:data-features}{]} from
+Open Hamilton. Next, we removed these features from the PBSP core
+service area and used pycnophylactic interpolation to disaggregate and
+reallocate population within the remaining area {[}see Figure
+\ref{fig:interpolated-population}{]}.
 
@@
 \addlinespace
-\cellcolor{gray!6}{\textbf{Municipal Service Centres}} & \cellcolor{gray!6}{The location of all municipal service centres, including City Hall.}\\
-\textbf{Recreation and Community Centres} & The location of all recreation and community centres.\\
+\cellcolor{gray!6}{\textbf{Municipal Service Centers}} & \cellcolor{gray!6}{The location of all municipal service centers, including City Hall.}\\
+\textbf{Recreation and Community Centers} & The location of all recreation and community centers.\\
 \cellcolor{gray!6}{\textbf{Arenas}} & \cellcolor{gray!6}{The location of all indoor arenas.}\\
@@
 
-\caption{Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
+\caption{Interpolated population in the Hamilton Bike Share's core service area (outlined in black) and within 30 minutes of walking to the core service area.
 Each population cell is 50-by-50 m in size.
@@
 \hypertarget{travel-time-matrix}{%
-\subsection{Travel Time Matrix}\label{travel-time-matrix}}
+\subsection{4.3. Travel Time Matrix}\label{travel-time-matrix}}
 
 To calculate walking times from the centroid of our micro population
-cells we extracted OpenStreetMap data for Hamilton Bike Share's service
+cells, we extracted OpenStreetMap data for Hamilton Bike Share's service
 area from \href{https://download.bbbike.org/osm/bbbike/}{BBBike}, an
 online cycle route planner that interfaces with OpenStreetMap.
-OpenStreetMap data provides the networks for calculate walking times
-from each population cell to nearby bike share stations, using a maximum
-walking distance of 10 km and walking time of 30 minutes as thresholds.
-A travel time matrix was created with the origins as the coordinates of
-the population cells and the destinations as the coordinates of the bike
-share stations within the maximum threshold. This process provides a
-more realistic measure of the friction of reaching bike share stations
-by taking infrastructure into account of travel times, rather than using
-the Euclidean distance from population cell to station. Routing and
-travel time calculations were completed using the \texttt{R} package
-\texttt{r5r}, used for rapid realistic routing operations (Pereira,
-Saraiva, et al., 2021).
+OpenStreetMap data provides the networks for calculating walking times
+from each population cell to nearby bicycle share stations, using a
+maximum walking distance of 10 km and walking time of 30 minutes as
+thresholds. A travel time matrix was created with the origins as the
+coordinates of the population cells and the destinations as the
+coordinates of the bicycle share stations within the maximum threshold.
+This process provides a more realistic measure of the friction of
+reaching stations by taking infrastructure into account of travel times,
+rather than using the Euclidean distance from population cell to
+station. Routing and travel time calculations were completed using the
+\texttt{R} package \texttt{r5r}, used for rapid realistic routing
+operations (Pereira, Saraiva, et al., 2021).
 
 \hypertarget{data}{%
-\subsection{Data}\label{data}}
+\subsection{4.4. Data}\label{data}}
 
@@
 \hypertarget{results}{%
-\section{Results}\label{results}}
+\section{5. Results}\label{results}}
 
 \hypertarget{accessibility-by-distance-thresholds}{%
-\subsection{Accessibility by Distance
+\subsection{5.1. Accessibility by Distance
 Thresholds}\label{accessibility-by-distance-thresholds}}
@@
 Consensus regarding the distance that individuals are willing to walk to
-access a PBSP station is lacking, but the literature on walking
+access a bicycle share station is lacking, but the literature on walking
 behaviour can provide some guidelines to determine the thresholds for
 sensitivity analysis. Previous studies have found that living within 250
-metres (Fuller et al., 2011) and 300 metres (Kabra et al., 2020) is
-correlated with bike share use, while other research has found that
-walking trips are less than 600 metres and rarely more than 1200 metres
-(Millward et al., 2013) or a median distance of 650 metres (Larsen et
-al., 2010). In Hamilton, Hamilton Bike Share will depict a map at some
-stations to show the user the locations of the other nearest stations
-within a five minute walk, which suggests that this is an average
-distance that people are expected to walk.
+m (Fuller et al., 2011) and 300 m (Kabra et al., 2020) is correlated
+with bike share use, while other research has found that walking trips
+are less than 600 m and rarely more than 1200 m (Millward et al., 2013)
+or a median distance of 650 m (Larsen et al., 2010). In Hamilton,
+Hamilton Bike Share will depict a map at some stations to show the user
+the locations of the other nearest stations within a five minute walk,
+which suggests that this is an average distance that people are expected
+to walk. National Association of City Transportation Officials (NACTO)
+has a similar normative guide (City Transportation Officials, 2015).
 
@@
 that demand on a limited supply increases as more people can reach each
-bike share station.
-
-For this reason, we experiment with various walking thresholds by
-conducting a sensitivity analysis to calculate accessibility at
-different walking times from population cell to bicycle share station: 3
-minutes, 5 minutes, 10 minutes, and 15 minutes. We categorize these
-thresholds as minimum, average, maximum, and extreme, respectively. At
-each threshold, we compare accessibility between the current system and
-the original system to examine the contribution of the additional equity
-stations. When considering the results reported below, it is important
-to remember that accessibility is technically a form of smoothing
-(O'Kelly and Horner, 2003, pp. 7--8): smaller thresholds produce less
-smoothing (which can result in ``spiky'' accessibility landscapes), and
-larger thresholds produce more smoothing, and fewer spikes.
+station.
+
+For this reason, we experiment with various thresholds by conducting a
+sensitivity analysis to calculate accessibility at different walking
+times from population cell to bicycle share station: 3 minutes, 5
+minutes, 10 minutes, and 15 minutes. We categorize these thresholds as
+minimum, average, maximum, and extreme, respectively. At each threshold,
+we compare accessibility between the current system and the original
+system to examine the contribution of the additional equity stations.
+When considering the results reported below, it is important to remember
+that accessibility is technically a form of smoothing (O'Kelly and
+Horner, 2003, pp. 7--8): smaller thresholds produce less smoothing
+(which can result in ``spiky'' accessibility landscapes), while larger
+thresholds produce more smoothing and fewer spikes.
 
 \hypertarget{minimum-threshold}{%
-\subsubsection{Minimum Threshold}\label{minimum-threshold}}
+\subsubsection{5.1.1. Minimum Threshold}\label{minimum-threshold}}
 
@@
 slightly higher. This high level of system-wide accessibility occurs
-because the population that can reach bike share stations when travel
+because the population that can reach bicycle share stations when travel
 time is 3 minutes or less is very limited, and accessibility is strongly
@@
 \hypertarget{average-threshold}{%
-\subsubsection{Average Threshold}\label{average-threshold}}
+\subsubsection{5.1.2. Average Threshold}\label{average-threshold}}
 
 With a walking distance of five minutes, we find that there are 68.6
-bicycle racks per person in the original system. With the addition of
+bicycle racks per person system-wide in the original system
+configuration (i.e., without equity stations). With the addition of
 equity stations, there are now 68.8 bicycle racks per person. At this
@@
 \hypertarget{maximum-threshold}{%
-\subsubsection{Maximum Threshold}\label{maximum-threshold}}
+\subsubsection{5.1.3. Maximum Threshold}\label{maximum-threshold}}
 
 With a walking distance of ten minutes, we find that there are 3.61
-bicycle racks per person in the original system. With the addition of
+bicycle racks per person system-wide in the original system
+configuration (i.e., without equity stations). With the addition of
 equity stations, there are now 3.74 bicycles per person. Figure
@@
 systems. Differences in accessibility across the service area are now
-apparent, with users near the university and its adjacent
-neighbourhoods, as well as neighbourhoods north of the downtown area
-(outlined in green), having slightly higher accessibility. While the
+apparent, with users near the university and its adjacent neighborhoods,
+as well as neighborhoods north of the downtown area (the latter is
+outlined in green), having slightly higher accessibility. While the
 differences are modest, they are more apparent at this threshold than at
@@
 \hypertarget{extreme-threshold}{%
-\subsubsection{Extreme Threshold}\label{extreme-threshold}}
+\subsubsection{5.1.4. Extreme Threshold}\label{extreme-threshold}}
 
 With a walking distance of fifteen minutes, we find that there are 2.44
-bicycle racks per person in the original system. With the addition of
+bicycle racks per person system-wide in the original system
+configuration (i.e., without equity stations). With the addition of
 equity stations, there are now 2.55 bicycles per person. Figure
 \ref{fig:figure-9} presents a comparison of accessibility between the
-systems. Users near the university and the neighbourhoods north of the
-downtown area (outlined in green) have the highest accessibility,
-followed by those who live in the city's downtown area. Accessibility in
-the east end of the core service area remains low.
+systems. Users near the university and the neighborhoods north of the
+downtown area (the latter is outlined in green) have the highest
+accessibility, followed by those who live in the city's downtown area.
+Accessibility in the east end, where equity stations were implemented,
+of the core service area remains low.
 
@@
 \hypertarget{accessibility-by-median-total-household-income}{%
-\subsection{Accessibility by Median Total Household
+\subsection{5.2. Accessibility by Median Total Household
 Income}\label{accessibility-by-median-total-household-income}}
@@
 dissemination area. We then can compare accessibility to median total
-household income statistics from the census.
+household income statistics from the Canadian census results from 2016.
 
@@
 {[}\ref{fig:figure-bi-map-threshold-15}{]} depict bivariate choropleth
-maps with the combine the spatial distribution of accessibility and
-median total household income, using tertiles for the coloring scheme.
-In each figure the top panel is the system without the equity stations,
-and the bottom panel is accessibility with the equity stations. As
-expected, the analysis indicates that the extreme threshold of fifteen
-minutes is associated with the largest number of people who are within
-the assumed service areas of stations. We find that stations added to
-Hamilton's public bicycle share program did indeed expand the spatial
-coverage of the system. By implementing equity stations in areas with
-lower median total household income in Hamilton, the PBSP attained
-greater horizontal equity by extending the spatial distribution of
-bicycles across the city. This is particularly evident at the minimum
-and average thresholds of three and five minutes, respectively, where
-the equity stations fill a number of gaps in program coverage.
+maps that combine the spatial distribution of accessibility and median
+total household income, using tertiles for the coloring scheme. In each
+figure, the top panel is the system without the equity stations (i.e.,
+the original system configuration), and the bottom panel is
+accessibility in the current system with equity stations. As expected,
+the analysis indicates that the extreme threshold of fifteen minutes is
+associated with the largest number of people who are within the assumed
+service area of stations. We find that stations added to Hamilton's
+public bicycle share program did indeed expand the spatial coverage of
+the system, leading to more balance across the service area. By
+implementing equity stations in areas with lower median total household
+income in Hamilton, the PBSP attained greater horizontal equity by
+extending the spatial distribution of bicycles across the city. This is
+particularly evident at the minimum and average thresholds of three and
+five minutes, respectively, where the equity stations fill a number of
+gaps in program coverage.
 
@@
 
-To further explore this issue, we aggregate accessibility by median
-total household income across all geographies (see Table
-\ref{tab:accessibility-income}). Similar to Hosford and Winters (2018),
-we find that Hamilton's PBSP already serviced a large proportion of the
-population in the bottom and second quintiles of median total household
-income. This may well be an artifact of the spatial socio-economic and
-demographic profile of Hamilton, where the most dense parts of the city
-(where a PBSP is most easily launched) are also those with relatively
-lower incomes. On the other hand, the levels of accessibility to
-Hamilton's PBSP are generally lower for populations in the bottom 20\%
-of median total household income, compared to populations in the top
-20\%.
-
-The results indicate that the addition of equity stations increased
-horizontal equity by growing the population serviced irrespective of the
-walking threshold. The largest gains were made for dissemination areas
-in the second 20\%, where an additional 3,073 and 5,395 people could
-reach a bicycle share station within three and five minutes walk,
-respectively, after the addition of the equity station. However, we find
-that there were only small changes in the population in the bottom 20\%
-of median total household income who are serviced by the equity
-stations, and the accessibility gains are also quite modest and smaller
-than for populations in the second and third quintiles of median total
-household income. This suggests that income disparities persist, albeit
-this depends on the walking time thresholds. With and without equity
-stations, people in the top 20\% of income have the highest level of
-access at a threshold of ten and fifteen minutes. Although dissemination
-areas in the second 20\% have the highest level of access by a
-significant amount at lower distance thresholds, the bottom 20\%, who
-may benefit the most from Hamilton Bike Share's equity initiatives, have
-the lowest access at three minutes threshold and the second lowest
-access at all other thresholds.
+To further explore this issue and examine vertical equity, we aggregate
+accessibility by median total household income across all geographies
+(see Table \ref{tab:accessibility-income}). Similar to Hosford and
+Winters (2018), we find that Hamilton's PBSP already serviced a large
+proportion of the population in the bottom and second quintiles of
+median total household income. This may well be an artifact of the
+spatial socioeconomic and demographic profile of Hamilton, where the
+most dense parts of the city (where a PBSP is most easily launched) are
+also those with relatively lower incomes. On the other hand, the levels
+of accessibility to Hamilton's PBSP are generally lower for populations
+in the bottom 20\% of median total household income, compared to
+populations in the top 20\%.
+
+The spatial mapping reveals that the addition of equity stations
+increased horizontal equity by growing the population serviced
+irrespective of the walking threshold. The largest gains were made for
+dissemination areas in the second 20\%, where an additional 3,073 and
+5,395 people could reach a bicycle share station within three and five
+minutes walk, respectively, after the addition of equity stations.
+However, we find that there were only small increases in the population
+in the bottom 20\% of median total household income who are serviced by
+the equity stations, and the accessibility gains are also quite modest
+and smaller than for populations in the second and third quintiles of
+median total household income. This suggests that income disparities
+persist, albeit this depends on the walking time thresholds. With and
+without equity stations, people in the top 20\% of income have the
+highest level of access at a threshold of ten and fifteen minutes.
+Although dissemination areas in the second 20\% have the highest level
+of access by a significant amount at lower distance thresholds, the
+bottom 20\%, who may benefit the most from Hamilton Bike Share's equity
+initiatives, have the lowest access at three minutes threshold and the
+second lowest access at all other thresholds.
 
@@
 \hspace{1em}Fourth 20\% & 20185 & 1673 & 4.119 & 1673 & 4.119 & 0 & 0.000\\
-\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{27541} & \cellcolor{gray!6}{2151} & \cellcolor{gray!6}{3.757} & \cellcolor{gray!6}{2416} & \cellcolor{gray!6}{3.784} & \cellcolor{gray!6}{265} & \cellcolor{gray!6}{0.027}\\
+\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{27541} & \cellcolor{gray!6}{2114} & \cellcolor{gray!6}{3.408} & \cellcolor{gray!6}{2379} & \cellcolor{gray!6}{3.434} & \cellcolor{gray!6}{265} & \cellcolor{gray!6}{0.026}\\
 \addlinespace[0.3em]
@@
 \hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{33312} & \cellcolor{gray!6}{17513} & \cellcolor{gray!6}{56.048} & \cellcolor{gray!6}{22908} & \cellcolor{gray!6}{56.137} & \cellcolor{gray!6}{5395} & \cellcolor{gray!6}{0.089}\\
-\hspace{1em}Third 20\% & 30940 & 15117 & 4.259 & 18309 & 4.291 & 3192 & 0.032\\
+\hspace{1em}Third 20\% & 30940 & 15117 & 4.258 & 18309 & 4.291 & 3192 & 0.033\\
 \hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{20185} & \cellcolor{gray!6}{2867} & \cellcolor{gray!6}{1.094} & \cellcolor{gray!6}{3116} & \cellcolor{gray!6}{1.095} & \cellcolor{gray!6}{249} & \cellcolor{gray!6}{0.001}\\
-\hspace{1em}Top 20\% & 27541 & 4074 & 6.256 & 4540 & 6.264 & 466 & 0.008\\
+\hspace{1em}Top 20\% & 27541 & 4052 & 5.925 & 4518 & 5.933 & 466 & 0.008\\
 \addlinespace[0.3em]
 \multicolumn{8}{l}{\textbf{Threshold - 10 minutes}}\\
-\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{43441} & \cellcolor{gray!6}{41824} & \cellcolor{gray!6}{0.604} & \cellcolor{gray!6}{41981} & \cellcolor{gray!6}{0.622} & \cellcolor{gray!6}{157} & \cellcolor{gray!6}{0.018}\\
-\hspace{1em}Second 20\% & 33312 & 27546 & 0.862 & 30503 & 0.929 & 2957 & 0.067\\
-\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{30940} & \cellcolor{gray!6}{22394} & \cellcolor{gray!6}{0.776} & \cellcolor{gray!6}{25128} & \cellcolor{gray!6}{0.802} & \cellcolor{gray!6}{2734} & \cellcolor{gray!6}{0.026}\\
+\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{43441} & \cellcolor{gray!6}{41824} & \cellcolor{gray!6}{0.603} & \cellcolor{gray!6}{41981} & \cellcolor{gray!6}{0.621} & \cellcolor{gray!6}{157} & \cellcolor{gray!6}{0.018}\\
+\hspace{1em}Second 20\% & 33312 & 27546 & 0.860 & 30503 & 0.927 & 2957 & 0.067\\
+\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{30940} & \cellcolor{gray!6}{22394} & \cellcolor{gray!6}{0.772} & \cellcolor{gray!6}{25128} & \cellcolor{gray!6}{0.799} & \cellcolor{gray!6}{2734} & \cellcolor{gray!6}{0.027}\\
 \hspace{1em}Fourth 20\% & 20185 & 4544 & 0.225 & 4989 & 0.227 & 445 & 0.002\\
-\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{27541} & \cellcolor{gray!6}{7989} & \cellcolor{gray!6}{1.346} & \cellcolor{gray!6}{9078} & \cellcolor{gray!6}{1.348} & \cellcolor{gray!6}{1089} & \cellcolor{gray!6}{0.002}\\
+\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{27541} & \cellcolor{gray!6}{7989} & \cellcolor{gray!6}{1.160} & \cellcolor{gray!6}{9078} & \cellcolor{gray!6}{1.162} & \cellcolor{gray!6}{1089} & \cellcolor{gray!6}{0.002}\\
 \addlinespace[0.3em]
 \multicolumn{8}{l}{\textbf{Threshold - 15 minutes}}\\
-\hspace{1em}Bottom 20\% & 43441 & 42208 & 0.536 & 42327 & 0.557 & 119 & 0.021\\
-\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{33312} & \cellcolor{gray!6}{30507} & \cellcolor{gray!6}{0.555} & \cellcolor{gray!6}{31069} & \cellcolor{gray!6}{0.614} & \cellcolor{gray!6}{562} & \cellcolor{gray!6}{0.059}\\
-\hspace{1em}Third 20\% & 30940 & 26108 & 0.554 & 26660 & 0.581 & 552 & 0.027\\
+\hspace{1em}Bottom 20\% & 43441 & 42208 & 0.534 & 42327 & 0.554 & 119 & 0.020\\
+\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{33312} & \cellcolor{gray!6}{30507} & \cellcolor{gray!6}{0.551} & \cellcolor{gray!6}{31069} & \cellcolor{gray!6}{0.611} & \cellcolor{gray!6}{562} & \cellcolor{gray!6}{0.060}\\
+\hspace{1em}Third 20\% & 30940 & 26108 & 0.546 & 26660 & 0.573 & 552 & 0.027\\
 \hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{20185} & \cellcolor{gray!6}{6312} & \cellcolor{gray!6}{0.093} & \cellcolor{gray!6}{7435} & \cellcolor{gray!6}{0.096} & \cellcolor{gray!6}{1123} & \cellcolor{gray!6}{0.003}\\
-\hspace{1em}Top 20\% & 27541 & 10209 & 0.808 & 11089 & 0.811 & 880 & 0.003\\
+\hspace{1em}Top 20\% & 27541 & 10209 & 0.712 & 11089 & 0.714 & 880 & 0.002\\
 \bottomrule
@@
 \hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-Using disaggregate data, in this paper we examined spatial equity and
+\section{6. Discussion}\label{discussion}}
+
+Using disaggregated data in this paper, we examined spatial equity and
 accessibility to Hamilton Bike Share, with a particular focus on
@@
 balanced floating catchment area approach, combined with pycnophylactic
-interpolation, enabled us to measure accessibility on a micro sale. This
-differentiates our analysis from similar papers exploring equity in
-PBSPs that use larger geographical units of analysis or that focus on
-station location instead of level of service. In this way, our paper has
-made contributions in a positive way by applying an intuitive and useful
-approach to measure accessibility to a PBSP, and in a normative way by
-serving to inform future investments in cycle infrastructure for
-Hamilton Bike Share.
+interpolation, enabled us to measure accessibility on a micro scale
+which is a sensible approach to avoid the ``absorption of disparities'',
+as articulated by Chen et al.~(2019). This differentiates our analysis
+from similar papers exploring equity in PBSPs that use larger
+geographical units of analysis or that focus on station location instead
+of level of service. In this way, our paper has made contributions in a
+positive way by applying an intuitive and useful approach to measure
+accessibility to a PBSP, and in a normative way by serving to inform
+future investments in cycle infrastructure for Hamilton Bike Share.
 
@@
 by eight minutes {[}see Table \ref{tab:accessibility-income}{]}. This
-reflects the normative guide, advertised on some bicycle share stations
-in Hamilton showing other stations within a five minute walk. We find
+reflects the normative guide advertised on some bicycle share stations
+in Hamilton showing other stations within a five minute walk, as well as
+the directive of NACTO (City Transportation Officials, 2015). We find
 that over 118,000 people can access a bicycle share station within a 15
@@
 a minimum threshold of three minutes, too few people can reach stations
-which leads to relatively high levels of service (since there is little
-crowding). However, accessibility is at its lowest after eight minutes
+which leads to relatively high levels of service since there is little
+crowding. However, accessibility is at its lowest after eight minutes
 whereby congestion effects due to increased potential demand kick in.
@@
 substantially more stations and bicycles are needed to service the area
-({\textbf{???}}). With a service area of 40 sq.km, it is estimated that
-Hamilton should have between 380 and 440 stations, instead of 130, and
-1,500 bicycles instead of 900 ({\textbf{???}}). Reduced capacity within
+(Hamilton, 2015b). With a service area of 40 sq.km, it is estimated that
+Hamilton should have between 380 and 440 stations instead of 130, and
+1,500 bicycles instead of 900 (Hamilton, 2015b). Reduced capacity within
 the system leads to gaps in coverage in some areas of the city ``with
 some areas not having the recommended station density of 300m between
-stations or 10 stations per square km'' ({\textbf{???}}). This study
+stations or 10 stations per square km'' (Hamilton, 2015b). This study
 illustrates the consequences of an imbalanced system whereby levels of
@@
 \ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in service
-were filled by these additional stations. In this respect, there are
-some commonalities between the expansions of Hamilton Bike Share and
+were filled by these additional stations. This indicates that more
+individuals at all income quintiles can access Hamilton Bike Share,
+which leads to a more balanced or equal program. In this respect, there
+are some commonalities between the expansions of Hamilton Bike Share and
 that of Citi Bike in New York. Babgoli et al.~(2019) found a slight but
 not statistically significant increase in the proportion of
-neighbourhoods with the highest levels of poverty with stations after
-the Citi Bike expansion in 2015. Although the Citi Bike expansion was
-not specifically driven by a desire to reduce inequity in access, 16\%
-of neighbourhoods with the highest levels of poverty had stations
-compared to 12\% before. With the addition of equity stations, there
+neighborhoods with the highest levels of poverty with stations after the
+Citi Bike expansion in 2015. Although the Citi Bike expansion was not
+specifically driven by a desire to reduce inequity in access, 16\% of
+neighborhoods with the highest levels of poverty had stations compared
+to 12\% before. Similarly, with the addition of equity stations, there
 were large gains in accessibility for the second 20\% at the average
@@
 
-Vertical inequity, however, persisted despite the addition of equity
-stations. While the addition of equity stations seems to modestly
-increase accessibility for all income groups at all thresholds, they did
-not increase accessibility substantially for any single income group.
-Most importantly, individuals in the bottom 20\% of median total
-household income have the second lowest level of access to Hamilton Bike
-Share at most thresholds (average, maximum, and extreme). At the minimum
-threshold, the bottom 20\% have the lowest level of access. While
-previous research found that neighbourhoods with more disadvantage are
-better serviced by Hamilton Bike Share, the authors used the Pampalon
-Deprivation Index to determine the level of disadvantage for
-dissemination areas \emph{across} the city not just within the core
-service area (Hosford and Winters, 2018). Instead, we use median total
-household income for each dissemination area \emph{within} the core
-service area. We conclude that Hamilton's PBSP, while by default located
-in areas with more deprivation compared to other cities, has disparities
-in accessibility between income groups. With equity stations, many areas
-with low median household income in the east end of the service area
-continue to have low accessibility. At higher thresholds of 10 and 15
-minutes, the top 20\% have the highest level of access to Hamilton Bike
-Share. This aligns with other studies from Tampa (Chen et al., 2019) and
-Seattle (Mooney et al., 2019), which have found disparities in station
-location or access to bicycles between levels of income and education.
+Vertical inequity, however, persisted as evidenced by differences in
+accessibility according to income quintile. While the addition of equity
+stations seems to modestly increase accessibility for all income groups
+at all thresholds, they did not increase accessibility substantially for
+any single income group. Most importantly, individuals in the bottom
+20\% of median total household income have the second lowest level of
+access to Hamilton Bike Share at most thresholds (average, maximum, and
+extreme). At the minimum threshold, the bottom 20\% have the lowest
+level of access. While previous research found that neighborhoods with
+more disadvantage are better serviced by Hamilton Bike Share, the
+authors used the Pampalon Deprivation Index to determine the level of
+disadvantage for dissemination areas \emph{across} the city not just
+within the core service area (Hosford and Winters, 2018). Instead, we
+use median total household income for each dissemination area
+\emph{within} the core service area. We conclude that Hamilton's PBSP,
+while by default located in areas with more deprivation compared to
+other cities, has disparities in accessibility between income groups.
+With equity stations, many areas with low median total household income
+in the east end of the service area continue to have low accessibility.
+At the maximum and extreme thresholds, the top 20\% have the highest
+level of access to Hamilton Bike Share. These findings aligns with other
+studies from Tampa (Chen et al., 2019) and Seattle (Mooney et al.,
+2019), which have found disparities in station location or access to
+bicycles between levels of income and education.
 
 Based on our analysis, we identified specific areas that have both low
-accessibility and low median household income which would benefit from
-an increased supply of public bicycles. These empirical findings provide
-support to the City of Hamilton's efforts to increase equity and balance
-the system, and confirm that additional stations and bicycles are needed
-to improve access not only for the bottom 20\% but for all income
-groups. Figures \ref{fig:figure-bi-map-threshold-3},
+accessibility and low median total household income which would benefit
+from an increased supply of public bicycles. These empirical findings
+provide support to the City of Hamilton's efforts to increase equity and
+balance the system, and confirm that additional stations and bicycles
+are needed to improve access not only for the bottom 20\% but for all
+income groups. Figures \ref{fig:figure-bi-map-threshold-3},
 \ref{fig:figure-bi-map-threshold-5},
@@
 \ref{fig:figure-bi-map-threshold-15} highlight potential locations for
-new equity stations.
+new equity stations to better accommodate groups with lower
+socioeconomic status.
 
 \hypertarget{study-limitations}{%
-\section{Study Limitations}\label{study-limitations}}
+\section{7. Study Limitations}\label{study-limitations}}
 
@@
 specifically looked at differences in trip type, frequency, or length
-among users from disadvantaged neighbourhoods (Caspi and Noland, 2019;
+among users from disadvantaged neighborhoods (Caspi and Noland, 2019;
 Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), but our analysis is
-limited by the lack of publicly available route or user data to conduct
-similar analyses for Hamilton Bike Share.
+limited by the lack of publicly available route and individual user data
+to conduct similar analyses for Hamilton Bike Share.
 
@@
 about the number of bicycles at each station. Hamilton Bike share works
-to balance the number of bicycles in stations, but it is reasonable to
-expect that the number of bicycles will not match exactly the number of
-racks at every station. Ideally, instead of number of bicycle racks as
-our measure of supply, we would have liked to use the average number of
-bicycles at stations, perhaps at different times during the day too.
-Should these data become available, it would be worthwhile to revisit
-the question to examine how well operation of the system (including
-balancing of bicycles among stations) works to maintain the nominal
-levels of accessibility examined in this paper.
+to balance the number of bicycles across stations, but it is reasonable
+to expect that the number of bicycles will not match exactly the number
+of racks at every station. Ideally, instead of number of bicycle racks
+as our measure of supply, we would have liked to use the average number
+of bicycles at stations, perhaps at different times during the day or
+different seasons. Should this data become available, it would be
+worthwhile to revisit the question to examine how well operation of the
+system (including balancing of bicycles across stations) works to
+maintain the nominal levels of accessibility examined in this paper.
 
 \hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+\section{8. Conclusion}\label{conclusion}}
 
@@
 inequities. In particular, accessibility improved the most for those in
-the second 20\% median household income at all thresholds, but the gains
-were only modest for all income groups. Dissemination areas with the
-bottom 20\% had the lowest accessibility at three minutes, and second
-lowest levels of accessibility at five, ten, and fifteen minutes.
+the second 20\% median total household income at all thresholds, but the
+gains were only modest for all income groups. Dissemination areas with
+the bottom 20\% had the lowest accessibility at three minutes, and
+second lowest levels of accessibility at five, ten, and fifteen minutes.
 Congestion effects were observed at higher thresholds, with
 accessibility decreasing significantly once the catchment area is
-increased to 10 minutes walking.
+increased to ten minutes walking.
 
@@
 Norland (2019) after bike share stations were implemented in low-income
-Philadelphia neighbourhoods. The bulk of cycling facilities that have
-been built in Hamilton to date located in the core service area near the
-conventional stations. It would be worthwhile to explore the route
+Philadelphia neighborhoods. The bulk of cycling facilities that have
+been built in Hamilton to date are located in the core service area near
+the conventional stations. It would be worthwhile to explore the route
 choice of bike share trips departing or ending at the equity stations
@@
 and other cities with PBSPs that wish to evaluate and address spatial
-inequuality in accessibility and transportation options in urban areas.
+inequality in accessibility and transportation options in urban areas.
 
@@
 (Auguie, 2017), \texttt{kableExtra} (Zhu, 2020), \texttt{knitr} (Xie,
-2021a), \texttt{pycno} (Brunsdon, 2014), \texttt{r5r} (Saraiva et al.,
-2021), \texttt{raster} (Hijmans, 2020), \texttt{rgdal} (Bivand et al.,
-2021), \texttt{rticles} (Allaire et al., 2021), \texttt{sf} (Pebesma,
-2021), \texttt{tidyverse} (Wickham, 2019), \texttt{tinytex} (Xie,
-2021b), \texttt{units} (Pebesma et al., 2020).
+2020), \texttt{pycno} (Brunsdon, 2014), \texttt{r5r} (Saraiva et al.,
+2020), \texttt{raster} (Hijmans, 2020), \texttt{rgdal} (Bivand et al.,
+2020), \texttt{rticles} (Allaire et al., 2021), \texttt{sf} (Pebesma,
+2020), \texttt{tidyverse} (Wickham, 2019), \texttt{tinytex} (Xie, 2021),
+\texttt{units} (Pebesma et al., 2020).
 
@@
 \leavevmode\hypertarget{ref-R-rgdal}{}%
-Bivand, R., Keitt, T., Rowlingson, B., 2021. Rgdal: Bindings for the
+Bivand, R., Keitt, T., Rowlingson, B., 2020. Rgdal: Bindings for the
 geospatial data abstraction library.
@@
 
+\leavevmode\hypertarget{ref-nactowalkingstation2015}{}%
+City Transportation Officials, N.A. of, 2015. Walkable station spacing
+is key to successful, equitable bike share {[}WWW Document{]}. URL
+\url{https://nacto.org/wp-content/uploads/2015/09/NACTO_Walkable-Station-Spacing-Is-Key-For-Bike-Share_Sc.pdf}
+
 \leavevmode\hypertarget{ref-civicplanSoBiHamiltonMembership2017}{}%
@@
 Planning Engagement Research.
+
+\leavevmode\hypertarget{ref-delamaterSpatialAccessibilitySuboptimally2013}{}%
+Delamater, P.L., 2013. Spatial accessibility in suboptimally configured
+health care systems: A modified two-step floating catchment area
+(M2SFCA) metric. Health \& Place 24, 30--43.
+doi:\href{https://doi.org/10.1016/j.healthplace.2013.07.012}{10.1016/j.healthplace.2013.07.012}
 
@@
 
+\leavevmode\hypertarget{ref-hamiltonsobi2014}{}%
+Hamilton, C. of, 2014. Hamilton bike share public engagement report
+{[}WWW Document{]}. URL
+\url{https://hamilton.socialbicycles.com/assets/pdf/Social_Cyclist_Bike_Share_Report.pdf}
+
 \leavevmode\hypertarget{ref-hamiltonHamiltonBikeShare2015}{}%
-Hamilton, C. of, 2015. Hamilton Bike Share.
+Hamilton, C. of, 2015a. Hamilton Bike Share.
+
+\leavevmode\hypertarget{ref-hamiltonsobi2015}{}%
+Hamilton, C. of, 2015b. Public bike share transit system implementation
+plan (pw13015c) - (city wide) {[}WWW Document{]}. URL
+\url{https://pub-hamilton.escribemeetings.com/filestream.ashx?DocumentId=118356}
 
@@
 
+\leavevmode\hypertarget{ref-luoEnhancedTwostepFloating2009}{}%
+Luo, W., Qi, Y., 2009. An enhanced two-step floating catchment area
+(E2SFCA) method for measuring spatial accessibility to primary care
+physicians. Health \& Place 15, 1100--1107.
+doi:\href{https://doi.org/10.1016/j.healthplace.2009.06.002}{10.1016/j.healthplace.2009.06.002}
+
+\leavevmode\hypertarget{ref-luoMeasuresSpatialAccessibility2003}{}%
+Luo, W., Wang, F., 2003. Measures of Spatial Accessibility to Health
+Care in a GIS Environment: Synthesis and a Case Study in the Chicago
+Region. Environment and Planning B: Planning and Design 30, 865--884.
+doi:\href{https://doi.org/10.1068/b29120}{10.1068/b29120}
+
 \leavevmode\hypertarget{ref-macarthurAdaptiveBikeShare2020}{}%
@@
 doi:\href{https://doi.org/10.1016/j.ccs.2019.100291}{10.1016/j.ccs.2019.100291}
+
+\leavevmode\hypertarget{ref-ogilvieInequalitiesUsagePublic2012}{}%
+Ogilvie, F., Goodman, A., 2012. Inequalities in usage of a public
+bicycle sharing scheme: Socio-demographic predictors of uptake and usage
+of the London (UK) cycle hire scheme. Preventive Medicine 55, 40--45.
+doi:\href{https://doi.org/10.1016/j.ypmed.2012.05.002}{10.1016/j.ypmed.2012.05.002}
 
@@
 \leavevmode\hypertarget{ref-R-sf}{}%
-Pebesma, E., 2021. Sf: Simple features for r.
+Pebesma, E., 2020. Sf: Simple features for r.
 
@@
 
-\leavevmode\hypertarget{ref-qianEnhancingEquitableService2020}{}%
-Qian, X., Jaller, M., Niemeier, D., 2020. Enhancing equitable service
-level: Which can address better, dockless or dock-based Bikeshare
-systems? Journal of Transport Geography 86, 102784.
-doi:\href{https://doi.org/10.1016/j.jtrangeo.2020.102784}{10.1016/j.jtrangeo.2020.102784}
+\leavevmode\hypertarget{ref-qianBikeshareDestinationChoices2021}{}%
+Qian, X., Jaller, M., 2021. Bikeshare destination choices and
+accessibility among disadvantaged communities. Transportation Research
+Part D: Transport and Environment 91, 102686.
+doi:\href{https://doi.org/10.1016/j.trd.2020.102686}{10.1016/j.trd.2020.102686}
+
+\leavevmode\hypertarget{ref-radkeSpatialDecompositionsModeling2000}{}%
+Radke, J., Mu, L., 2000. Spatial Decompositions, Modeling and Mapping
+Service Regions to Predict Access to Social Programs. Geographic
+Information Sciences 6, 105--112.
+doi:\href{https://doi.org/10.1080/10824000009480538}{10.1080/10824000009480538}
 
@@
 \leavevmode\hypertarget{ref-R-r5r}{}%
-Saraiva, M., Pereira, R.H.M., Herszenhut, D., Braga, C.K.V., 2021. R5r:
+Saraiva, M., Pereira, R.H.M., Herszenhut, D., Braga, C.K.V., 2020. R5r:
 Rapid realistic routing with r5.
@@
 
+\leavevmode\hypertarget{ref-wanThreestepFloatingCatchment2012}{}%
+Wan, N., Zou, B., Sternberg, T., 2012. A three-step floating catchment
+area method for analyzing spatial access to health services.
+International Journal of Geographical Information Science 26,
+1073--1089.
+doi:\href{https://doi.org/10.1080/13658816.2011.624987}{10.1080/13658816.2011.624987}
+
 \leavevmode\hypertarget{ref-wangNeighborhoodSociodemographicCharacteristics2019}{}%
@@
 \leavevmode\hypertarget{ref-R-knitr}{}%
-Xie, Y., 2021a. Knitr: A general-purpose package for dynamic report
+Xie, Y., 2020. Knitr: A general-purpose package for dynamic report
 generation in r.
@@
 \leavevmode\hypertarget{ref-R-tinytex}{}%
-Xie, Y., 2021b. Tinytex: Helper functions to install and maintain tex
+Xie, Y., 2021. Tinytex: Helper functions to install and maintain tex
 live, and compile latex documents.

### 2021-03-29 sobiEquity/R/sobiEquity.R
Update data package for submission

@@ -1,7 +1,8 @@
 #' sobiEquity
 #'
-#' sobiEquity is a companion package for the paper "Examining spatial equity and accessibility
-#'  to a public bicycle share program using a balanced floating catchment area approach".
+#' sobiEquity is a companion package for the paper "Examining horizontal and vertical 
+#' equity to a public bicycle share program using a balanced floating catchment area 
+#' accessibility approach".
 #'   The package includes the data used in the analysis.
 #'
 #' @author Elise Desjardins, \email{desjae@@mcmaster.ca}
@@ -22,3 +23,76 @@ NULL
 #' @source Statistics Canada
 "hamilton_cma"
 
+#' Hamilton Community Downtown.
+#'
+#' Boundary of the downtown center of Hamilton, in Ontario, Canada.
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords transportation
+#' @name community_downtown
+#' @usage data(community_downtown)
+#' @source Open Hamilton
+"community_downtown"
+
+#' Hamilton Dissemination Areas.
+#'
+#' Boundary of the Hamilton DAs, in Ontario, Canada, with median total household income and
+#' population levels within each DA.
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords transportation
+#' @name hamilton_da_2016
+#' @usage data(hamilton_da_2016)
+#' @source Statistics Canada
+"hamilton_da_2016"
+
+#' Hamilton Interpolated Population.
+#'
+#' Interpolated population from Hamilton DAs, in 50-by-50 m cells.
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords transportation
+#' @name population_50x50
+#' @usage data(population_50x50)
+#' @source Statistics Canada
+"population_50x50"
+
+#' Hamilton Bike Share Service Area
+#'
+#' Boundary of the Hamilton Bike Share service area of Hamilton, in Ontario, Canada.
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords transportation
+#' @name sobi_service
+#' @usage data(sobi_service)
+#' @source Open Hamilton
+"sobi_service"
+
+#' Hamilton Bike Share Stations
+#'
+#' Location of the Hamilton Bike Share stations of Hamilton, in Ontario, Canada.
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords transportation
+#' @name sobi_hubs
[…]

### 2021-03-29 README.Rmd
Created README.Rmd file

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Accessibility-Sobi-Hamilton

<!-- badges: start -->
<!-- badges: end -->

The goal of Accessibility-Sobi-Hamilton is to ...

What is special about using `README.Rmd` instead of just `README.md`? You can include R chunks like so:

```{r cars}
summary(cars)
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/master/examples>.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub.

### 2021-03-29 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd

@@
 ---
-title: Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach 
+title: Examining horizontal and vertical equity to a public bicycle share program using a balanced floating catchment area accessibility approach 
 author:
@@
   - code: McMaster University
-    address: School of Earth, Environment & Society, 1280 Main Street West, Hamilton, ON L8S4L8
+    address: School of Earth, Environment & Society, McMaster University, 1280 Main Street West, Hamilton, ON L8S4L8
   - code: University of Toronto Scarborough
-    address: Department of Geography & Planning, 1265 Military Trail, Toronto, ON M1C1A4
+    address: Department of Geography & Planning, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON M1C1A4
 footnote:
@@
 
-  Public bicycle share programs (PBSPs) are implemented to promote cycling as a convenient and sustainable mode of transportation.
-These systems can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to lower income or disadvantaged populations.
-In Ontario, the City of Hamilton launched a public bicycle share program in 2015 that currently has over 900 operational bicycles and 130 docking stations.
-This system was expanded in 2018 by an equity initiative that added twelve docking stations with the explicit objective of increasing spatial equity in access.
-Since the cost of reaching a bicycle share station decreases the potential of accessing the program, the location of stations matters, and previous research found that Hamilton’s public bicycle share program targeted well disadvantaged areas in the city.
-In this research, we revisit the case of Hamilton and investigate differentials in accessibility to bicycle share stations using a balanced floating catchment area (BFCA) accessibility approach.
-We compare accessibility  with and without the equity stations to assess the effect of the initiative.
-We implement our analysis parting from micro zones to better reflect walking to a docking station, and conduct a sensitivity analysis at several walking time thresholds.
+  Public bicycle share programs (PBSPs) can play a role in advancing transportation equity if they make cycling more accessible to disadvantaged populations.
+In Ontario, the City of Hamilton expanded their PBSP in 2018 by adding twelve docking stations with the explicit objective of increasing spatial equity in access.
+In this case study, we investigate differentials in accessibility to stations using a balanced floating catchment area accessibility approach and compare accessibility with and without the equity stations.
+We analyze micro zones to better reflect walking to a station and conduct a sensitivity analysis at several walking time thresholds.
 We then reaggregate the estimated accessibility for further analysis using census data.
-Our findings indicate that the addition of equity stations increased the horizontal equity of the system by growing the serviced population at every threshold examined.
-With respect to vertical equity, although accessibility increased for the whole population, this increase was relatively modest especially for population in the bottom 20% of median total household income.
+Our findings indicate that equity stations increased the serviced population at every threshold examined.
+Although accessibility increased for the whole population, this increase was relatively modest especially for population in the bottom 20% of median total household income.
 
@@
 
-Introduction
+1. Introduction
 ===================
@@
 As a healthy, inexpensive, and convenient form of public transportation, shared bicycles can encourage individuals to take up cycling for short local trips or first and last mile trips to other public transportation instead of using personal vehicles.
-These systems can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to lower income or disadvantaged populations.
-As these systems become increasingly common, their introduction has been accompanied by a flurry of research that investigates them from the perspective of spatial and transportation equity [e.g., @smith2015exploring; @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @hullgrassoBikeShareEquity2020; @qianBikesharingEquityDisadvantaged2020; @qianEnhancingEquitableService2020].
-Indeed, although these programs are available to the general public and ought to be accessible to any individuals who wish to use them, research on PBSPs located in North American and European cities indicates that inequity in spatial accessibility needs redressing.
-
-In this paper we investigate the case of the PBSP in the City of Hamilton, in Ontario, Canada.
-Hamilton Bike Share (previously also known as Social Bicycles, or SoBi) was launched in 2015 and currently has over 900 operational bicycles and 130 docking stations.
-An interesting feature of this system is that it was expanded by an equity initiative that introduced twelve docking stations in 2018 with the explicit objective of increasing spatial accessibility.
-Hamilton Bike Share was already studied by @hosfordWhoArePublic2018 as part of a selection of PBSPs in Canada.
-These authors found that most of the cities in their research could benefit of greater efforts to expand service to areas with lower socio-economic populations - but Hamilton fared somewhat better than other cities in this respect.
-What distinguishes our research is the use of a balanced floating catchment area (BFCA) accessibility approach that accounts for the supply of stations and the potential demand from the population serviced.
-This requires a more disaggregated approach than the use of census geographies [q.v., @hosfordWhoArePublic2018], since many trips to a bike share station likely happen at a level lower than even the smallest census geography.
-For this reason, we implement our analysis parting from micro population zones to better reflect the friction of walking to a docking station.
+These programs can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to individuals and groups with lower socioeconomic status.
+As these programs become increasingly common, their introduction has been accompanied by a flurry of research that investigates them from the perspective of spatial and transportation equity [e.g., @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @hullgrassoBikeShareEquity2020; @qianBikesharingEquityDisadvantaged2020; @qianBikeshareDestinationChoices2021; @smith2015exploring].
+Indeed, although these programs are available to the general public and ought to be accessible to any individual who wishes to use them, research on PBSPs located in North American and European cities indicates that inequity in spatial accessibility needs redressing.
+
+In this paper, we investigate the case of the PBSP in the city of Hamilton, in Ontario, Canada.
+Hamilton Bike Share was launched in 2015 and currently has over 900 operational bicycles and 130 docking stations.
+An interesting feature of this program is that it was expanded by an equity initiative that introduced twelve stations in 2018 with the explicit objective of increasing spatial accessibility.
+Hamilton Bike Share was already studied by Hosford and Winters [-@hosfordWhoArePublic2018] as part of a selection of PBSPs in Canada.
+The authors found that most of the cities in their research could benefit of greater efforts to expand service to areas with lower socioeconomic groups - but Hamilton fared somewhat better than other cities in this respect.
+What distinguishes our research from existing literature is the use of a balanced floating catchment area (BFCA) accessibility approach that accounts for the supply of stations and the potential demand from the population serviced.
+This requires a more disaggregated approach than the use of census geographies [q.v., @hosfordWhoArePublic2018], since walking trips to a bicycle share station likely happen at a level lower than even the smallest census geography.
+For this reason, we implement our analysis parting from micro population zones to better reflect the friction of walking to a docking station, which is an important component of a bike share trip [@chenExploringEquityPerformance2019].
 Further, we conduct a sensitivity analysis at several walking time thresholds.
 In terms of spatial equity, we compare accessibility with and without the equity stations to assess the effect of the initiative, before reaggregating the data for further analysis using median total household income information from the census.
-We find that the addition of equity stations does increase the serviced population at every threshold examined.
-However, although accessibility increased for the whole population, this increase was relatively modest, especially for population in the bottom 20% of median total household income.
-The analysis also identifies areas under-serviced within the core service area of the system.
 
 This paper is an example of open and reproducible research that uses only open software for transportation and statistical analysis [@bivand2020progress; @lovelace2021open].
-All data were obtained from publicly available sources and organized in the form a data package.
-Following best practices in spatial data science [@brunsdon2020opening], the code and data needed to reproduce (or modify/extend) the analysis are available for download.^[https://drive.google.com/drive/folders/1ZGRpSN2LxS2Fq2tLVJgfm1sI6I26vzZh?usp=sharing]
-
-Literature Review
+All data were obtained from publicly available sources and organized in the form of a data package.
+Following best practices in spatial data science [@brunsdon2020opening], the code and data needed to reproduce, modify or extend the analysis are available for download.^[https://github.com/paezha/Accessibility-Sobi-Hamilton]
+
+2. Literature Review
 ===================
@@
 However, the relationship between income or education and bike share use is less clear-cut.
-Stations in disadvantaged communities in Chicago have been found to generate most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighbourhoods in Minneapolis-St.
+Stations in disadvantaged communities in Chicago have been found to generate most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighborhoods in Minneapolis-St.
 Paul used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019].
@@
 This makes sense given that individuals are more likely to use services or programs that they can easily reach.
-Several studies have recently explored the equity of PBSPs by primarily examining who has access to bike share (e.g., differences by demographics or socioeconomic status) and where stations are located.
-It is important to note that equity can be achieved in two different ways: equal spatial distribution across a region (e.g., horizontal equity) or greater access for vulnerable or disadvantaged populations (e.g., vertical equity) [@chenExploringEquityPerformance2019].
-Both are of interest to researchers and transport planners since they are often linked in that advantage, or conversely disadvantage, has spatial patterns.
-Using a negative binomial regression model, Qian and Jaller [-@qianBikesharingEquityDisadvantaged2020] estimated ridership in Chicago's PBSP among disadvantaged communities and found some disparities.
-While annual members in disadvantaged communities have a significantly lower share of trips compared to other areas in the city, they make longer trips.
+Several studies have recently explored the equity of PBSPs in North America by primarily examining who has access to bike share (e.g., differences by demographics or socioeconomic status) and where stations are located.
+It is important to note that equity can be achieved in two different ways: balanced or equal spatial distribution across a region for all similar groups (e.g., horizontal equity) or greater and targeted access to accommodate vulnerable or disadvantaged populations (e.g., vertical equity) [see @chenExploringEquityPerformance2019].
+Both are of interest to researchers and transportation planners since they are often linked in that advantage, or conversely disadvantage, has spatial patterns.
+Using a negative binomial regression model, Qian and Jaller [-@qianBikesharingEquityDisadvantaged2020] estimated ridership in Chicago's PBSP and found some disparities.
+A minority of bike share stations were found to be located in disadvantaged communities, while annual members from disadvantaged communities have a lower share of trips compared to other areas in the city.
 This suggests that they may be using PBSPs for utilitarian trips (e.g., commuting), which points to the importance of ensuring equitable access [@qianBikesharingEquityDisadvantaged2020].
-Similar results were found in Philadelphia, where lower income areas generated fewer trips and it was suggested that efforts to increase equity within the program have not been as successful as intended [@caspiBikesharingPhiladelphiaLowerincome2019].
-In the case of Seattle, all neighbourhoods had some level of access to dockless bikes but those with higher incomes and more residents of higher education had more bikes [@mooneyFreedomStationSpatial2019].
+Similar results were found in Philadelphia.
+Despite efforts to increase equity within the city's PBSP, census block groups with lower median income generated fewer trips which suggested that efforts have not been as successful as intended [@caspiBikesharingPhiladelphiaLowerincome2019].
+Trips from stations in such areas are utilitarian (e.g., commuting to work), which points to the importance of ensuring equitable access [@caspiBikesharingPhiladelphiaLowerincome2019].
+In the case of Seattle, all neighborhoods had some level of access to dockless bicycles but those with higher incomes and more residents of higher education had more bikes [@mooneyFreedomStationSpatial2019].
 Babagoli et al.
-[-@babagoliExploringHealthSpatial2019] also found that neighbourhoods in New York City with higher affluence had the greatest proportion of Citi Bike stations.
-Researchers have recommended targeted expansion of stations to areas that use bike share but have inequitable access [@ogilvieInequalitiesUsagePublic2012].
-
-On the whole, existing studies highlight the need for bicycle share systems to be more highly accessible for diverse populations in order to increase use beyond the "typical" users.
+[-@babagoliExploringHealthSpatial2019] also found that neighborhoods in New York City with higher affluence had the greatest proportion of Citi Bike stations.
+
+On the whole, existing studies highlight the need for PBSPs to be more highly accessible for diverse populations in order to increase use beyond the "typical" users.
 This has been the focus of recent research [see, among others, @auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020].
 Offering more people the option of using sustainable and active transportation, particularly those who have lower socioeconomic status and might benefit the most, is a worthy policy goal for cities with PBSPs.
-Several cities have launched specific programs to address barriers to bike share or to expand service to more deprived areas: London, United Kingdom [@ogilvieInequalitiesUsagePublic2012]; Philadelphia, Pennsylvania [@caspiBikesharingPhiladelphiaLowerincome2019], and Hamilton, Ontario [@hosfordWhoArePublic2018].
-Findings to date suggest that horizontal equity can be easier to achieve than vertical equity.
-However, exploring transportation equity by investigating where bicycle share stations are located, often using neighbourhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
-Meaning that, stations may be lacking in certain neighbourhoods but there may be stations accessible within a reasonable walking time.
+A few North American cities have launched specific programs to address barriers to bike share or to expand service to more deprived areas: Philadelphia, Pennsylvania [@caspiBikesharingPhiladelphiaLowerincome2019], and Hamilton, Ontario [@hosfordWhoArePublic2018].
+However, exploring transportation equity by investigating where bicycle share stations are located, often using neighborhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
+Meaning that, stations may be lacking in certain neighborhoods but there may be stations accessible within a reasonable walking time.
 This is where geographical accessibility becomes an important consideration.
 
-Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved over the past century and has increasingly become linked with recent interest in prioritizing local proximity and modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020].
-As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequaltiy in the provision of infrastructure, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options.
+Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved over the past century and has increasingly become linked with recent planning interests in prioritizing modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020].
+As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequality in the provision of infrastructure, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options.
 It also addresses some of the challenges of other performance measures such as level of service within a transportation network by measuring person-based indicators and exploring differences in use between population groups [@paezMeasuringAccessibilityPositive2012].
@@
 [-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300 m of stations, which amounts to 2-4 minutes walking by an able-bodied adult.
-Furthermore, similar to other public amenities affected by crowding, like health care services [e.g., @pereiraGeographicAccessCOVID192021], the utility of stations is limited by the number of bicycles that they can hold.
+Furthermore, similar to other public amenities affected by crowding, like health care services [e.g., @pereiraGeographicAccessCOVID192021], the utility of stations is limited by the _maximum_ number of bicycles that they can hold.
 The program may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles.
@@
 In the first, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hospital, weighted by the distance involved in reaching the facility.
-Next, these service level ratios are allocated back to the population centres and summarized as a measure of congested accessibility.
+Next, these service level ratios are allocated back to the population centers and summarized as a measure of congested accessibility.
 Thus, this model does a good job of considering potential crowding or competition for services.
@@
 When measuring accessibility, researchers have taken different approaches when it comes to aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones.
-Previous research on bike share equity has typically used a meso- or macro-level approach with aggregate data from entire neighbourhoods or census tracts [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021].
+Previous research on bike share equity has typically used a meso- or macro-level approach with aggregated data from entire neighborhoods or census tracts [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021].
 This is also true for studies examining correlates of bike share demand [@wangNewBikeShare2019].
 Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals, which is useful for addressing vertical inequity in PBSP usage.
-
-Using the balanced floating catchment area method [@paezDemandLevelService2019], a novel approach that has not been used yet in cycling research, with disaggregate population-level data, in this paper we examine accessibility to a public bicycle share program in Hamilton, Ontario.
-We (1) conduct a sensitivity analysis by measuring accessibility and level of service to bike share stations at different walking time thresholds to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median total household income of dissemination areas within the core service area.
-
-Case Study {#sec:study}
+Chen et al.
+[-@chenExploringEquityPerformance2019, pp.
+530] are in favour of using disaggregated data, which they did in their recent analysis of Tampa's PBSP, because they note "the use of aggregated data might hinder our understanding of the equity impacts since individual disparities are absorbed after aggregation".
+
+Using the balanced floating catchment area method [@paezDemandLevelService2019], a novel approach that has not been used yet in cycling research, with disaggregated population data, we examine accessibility to a public bicycle share program in Hamilton, Ontario.
+In this paper, we (1) conduct a sensitivity analysis by measuring accessibility and level of service to bicycle share stations at different walking time thresholds to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median total household income of dissemination areas within the core service area.
+
+3. Case Study {#sec:study}
 ======================
 
-## Original System
+## 3.1. Original System
 
@@
 The city launched a public bicycle share program in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015].
-Before June 2020, the program was known as Social Bicycles or SoBi Hamilton but is now called Hamilton Bike Share.
-Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021].
+Before June 2020, the program was known as Social Bicycles or SoBi Hamilton, but is now called Hamilton Bike Share.
+Stations are spaced between 300 and 600 m apart [@scottRouteChoiceBike2021].
 The core service area spans 40 sq.km of the city although it was planned to be 20 sq.km [@hamiltonsobi2015], and roughly 138,000 people are within 30 minutes walking of a bike share station [see Figure \ref{fig:hamilton-and-sobi-service-area}].
@@
 The City of Hamilton undertook a large public engagement campaign to validate the locations of bicycle share stations that had been selected and to crowdsource potential locations for additional locations [@hamiltonsobi2014].
-Most of the suggested locations were in the east end of the core service area that lack stations or in neighbourhoods not serviced by the PBSP.
-The program was enthusiastically welcomed in the city in 2015 - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015], however inadequate coverage given the size of the Hamilton Bike Share's service area was identified as a problem early on, and transportation planners noted that the small size (i.e., supply of bicycle racks) and low quantity of stations would lead to challenges in balancing the system [@hamiltonsobi2015].
-Based on public feedback, 13 stations were added to the original system.
-
-## Equity Initiative
+Most of the suggested locations were in the east end of the core service area that lack stations or in neighborhoods not serviced by the PBSP.
+The program was enthusiastically welcomed in the city in 2015 - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015], however inadequate coverage given the size of Hamilton Bike Share's service area was identified as a problem early on, and transportation planners noted that the small size (i.e., supply of bicycle racks) and low quantity of stations would lead to challenges in balancing the system [@hamiltonsobi2015].
+Based on public feedback, thirteen stations were added to the original system.
+
+## 3.2. Equity Initiative
 
 In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), with the objective of reducing barriers that may prevent individuals from accessing bike share in Hamilton.
-Additional bicycles and stations were added to the program which expanded it to more disadvantaged areas in the core service area [see Figure 2].
-The program also offers subsidized memberships to individuals who identify as low income, and complements these initiatives with cycle skills education.
+Additional bicycles and stations were added to the program which expanded it to more disadvantaged areas in the core service area.
+The program also offers subsidized memberships to individuals who identify as low income, and complements this service with cycle skills education.
 A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019].
 
-## Current System
+## 3.3. Current System
 
@@
 The core service area remains 40 sq.km.
-The PBSP has 13 stations that were added as part of the ERI initiative; we refer to these as "equity stations" throughout the paper while all other stations are "conventional stations".
-In total, 117 stations are "conventional" and 13 are "equity".
-
-## Membership
+The PBSP has twelve stations that were added as part of the ERI initiative; we refer to these as "equity stations" throughout the paper while all other stations are "conventional stations".
+In total, 118 stations are "conventional" and 12 are "equity".
+
+## 3.4. Membership
 
@@
 As a result of having a bike share membership, 49% of respondents report that they use their private vehicle less often or much less often but 48% report that their private vehicle use has remained about the same.
-This suggests that Hamilton Bike Share has been useful for certain kinds of trips but not all, meaning that some trips continue to require a private vehicle.
-
-## Relevant Research
+This suggests that Hamilton Bike Share has been useful for certain kinds of trips but not all, meaning that some trips continue to require a personal vehicle.
+
+## 3.5. Relevant Research
 
@@
 ```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "The core service area of Hamilton Bike Share is outlined in black.
-Hamilton Census Metropolitan Area (CMA) is shown in grey."}
+Hamilton Census Metropolitan Area is shown in grey."}
 ggplot() + 
@@
 
-```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bike share stations in Hamilton, Ontario.
-The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in dark green.
-Hamilton CMA is shown in grey."}
+```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bicycle share stations in Hamilton, Ontario.
+The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in pink.
+Hamilton Census Metropolitan Area is shown in grey."}
 ggplot() + 
@@
           size = 1,
-          color = "dark green") +
+          color = "pink") +
   geom_sf(data = sobi_hubs %>% 
@@
 
-Methods {#sec:methods}
+4. Methods and Data {#sec:methods}
 ======================
 
-## Floating Catchment Area
+## 4.1. Floating Catchment Area
 
@@
 This is an important property that supports our rationale for applying this method to measure accessibility to Hamilton Bike Share.
-The City of Hamilton has positioned stations between 300 and 600 metres apart, but anticipates that hubs will service those living within a 250 metre buffer from the station [@hamiltonsobi2015].
-The latter constitutes a normative statement: people ought to be able to access a station in less than 600 metres if they live in the core service area, with usage coming from 250 metres around.
+The City of Hamilton has positioned stations between 300 and 600 m apart, but anticipates that hubs will service those living within a 250 m buffer from the station [@hamiltonsobi2015].
+The latter constitutes a normative statement: people ought to be able to access a station in less than 600 m if they live in the core service area, with usage coming from 250 m around.
 However, it is not known how far people are actually willing to travel to reach a station.
@@
 
-More recently, the _balanced_ floating catchment area (FCA) approach was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by earlier FCA methods [@paezDemandLevelService2019].
+More recently, the _balanced_ floating catchment area approach was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by earlier FCA methods [@paezDemandLevelService2019].
 Briefly, overlapping catchment areas lead to inflation of population totals and deflation of service levels across a study area and generates inaccurate or misleading accessibility estimates.
 In contrast, Paez et al.
-[-@paezDemandLevelService2019] adjust the impedance weights so that both supply and demand are proportionally allocated.
-The results is a FCA method that balances the population and level of service by eliminating the over-counting of population and level of service that leads to distortions in demand and supply.
+[-@paezDemandLevelService2019] adjusted the impedance weights so that both supply and demand are proportionally allocated.
+The result is a FCA method that balances the population and level of service by eliminating the over-counting of population and level of service that leads to distortions in demand and supply.
 Other benefits of this adjusted method include consideration of competition which can occur when catchment areas overlap, as well as the preservation of population and level of service.
@@
 $$
+
 As seen in the equation above, the population allocated to station $j$ is the weighted sum of the population in the region; a spatial weight $w_{ij}$ represents the friction that the population at $i$ faces when reaching station $j$, and is usually given by a distance-decay function, so that each station is assumed to service only a segment of the population within a limited geographical range.
@@
 
-The balanced approach of @paezDemandLevelService2019 replaces the spatial weights with normalized versions as follows:
+The balanced approach of Paez et al [-@paezDemandLevelService2019] replaces the spatial weights with normalized versions as follows:
 $$
@@
 
-With these weights accessibility can be calculated without risk of demand or supply inflation:
+With these weights, accessibility can be calculated without risk of demand or supply inflation:
 $$
@@
 $$
+
 By allocating the population and level of service proportionally, this method preserves the values of the population and level of service and provides a regional provider-to-population ratio since:
@@
 In fact, since the proportional allocation procedure means that any proportion of the population allocated to a station is never allocated to other stations, and conversely any level of service allocated to a population is never re-allocated elsewhere, this property is replicated for any level of aggregation.
-For this paper, we employ a hybrid location-based and person-based approach to calculating accessibility using disaggregate population data.
+For this paper, we employ a hybrid location-based and person-based approach to calculating accessibility using disaggregated population data.
 In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data.
 This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are better able of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences.
-Applying the balanced floating catchment area approach allows us to examine accessibility by stratifying according to median total household income, which would constitute the individual component of the accessibility measure [@geursAccessibilityEvaluationLanduse2004].
-However, conducting a further sensitivity analysis to measure accessibility at different distance thresholds would help to consider potential spatio-temporal constraints.
+Applying the balanced floating catchment area approach allows us to examine accessibility by stratifying according to median total household income, which constitutes the individual component of the accessibility measure [@geursAccessibilityEvaluationLanduse2004].
+However, conducting a further sensitivity analysis to measure accessibility at different walking time thresholds would help to consider potential spatio-temporal constraints.
 Different people may be willing to travel different distances to access a public bicycle share station.
 
-## Pycnophylactic Interpolation
-
-To obtain population at sub-census geography levels (at the micro scale) we use pycnophylactic interpolation [@toblerSmoothPycnophylacticInterpolation1979].
-We obtained population data from the 2016 Census of Canada for dissemination areas (DAs, the smallest publicly available census geography in Canada).
-This zonal values of the population were interpolated to smaller polygons 50-by-50 m in size.
-Pycnophilactic interpolation involves smoothing out the population from each dissemination area while preserving total volume [see Figure \ref{fig:da-population}].
-When interpolating the population at this high level of resolution it is important to ensure that population numbers were not allocated to areas where people do not live in Hamilton (e.g., to parks, large institutional buildings, etc.) To do so, we retrieved shapefiles for various geographic features [see Table \ref{tab:data-features}] from Open Hamilton.
+## 4.2. Pycnophylactic Interpolation
+
+To obtain population at sub-census geography levels (at the micro scale), we use pycnophylactic interpolation [@toblerSmoothPycnophylacticInterpolation1979].
+We obtained population data from the 2016 Census of Canada for dissemination areas (DA), which is the smallest publicly available census geography in Canada.
+These zonal values of the population were interpolated to smaller polygons 50-by-50 m in size.
+Pycnophylactic interpolation involves smoothing out the population from each dissemination area while preserving total volume [see Figure \ref{fig:da-population}].
+When interpolating the population at this high level of resolution, it is important to ensure that population numbers were not allocated to areas where people do not live in Hamilton (for example, to parks, large institutional buildings, etc.).
+To do so, we retrieved shapefiles for various geographic features [see Table \ref{tab:data-features}] from Open Hamilton.
 Next, we removed these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population within the remaining area [see Figure \ref{fig:interpolated-population}].
@@
 data_tbl <- data.frame(
-  Feature = c("Hamilton Bike Share Stations", "Golf Courses", "Parks", "Designated Large Employment Areas", "Municipal Parking Lots", "Cemeteries", "Environmentally Sensitive Areas", "Streets", "Educational Institutions", "Places of Worship", "Municipal Service Centres", "Recreation and Community Centres", "Arenas", "Emergency Stations", "Fire Stations", "Police Stations", "Railways", "Hospitals"),
+  Feature = c("Hamilton Bike Share Stations", "Golf Courses", "Parks", "Designated Large Employment Areas", "Municipal Parking Lots", "Cemeteries", "Environmentally Sensitive Areas", "Streets", "Educational Institutions", "Places of Worship", "Municipal Service Centers", "Recreation and Community Centers", "Arenas", "Emergency Stations", "Fire Stations", "Police Stations", "Railways", "Hospitals"),
   Description = c(
@@
     "The location of buildings used for religious congregations.",
-    "The location of all municipal service centres, including City Hall.",
-    "The location of all recreation and community centres.",
+    "The location of all municipal service centers, including City Hall.",
+    "The location of all recreation and community centers.",
     "The location of all indoor arenas.",
@@
 
-```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Interpolated population in the Hamilton Bike Share core service area (outlined in black) and within 30 minutes of walking to the core service area (outlined in blue).
+```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Interpolated population in the Hamilton Bike Share's core service area (outlined in black) and within 30 minutes of walking to the core service area.
 Each population cell is 50-by-50 m in size.
@@
 
-## Travel Time Matrix
-
-To calculate walking times from the centroid of our micro population cells we extracted OpenStreetMap data for Hamilton Bike Share's service area from [BBBike](https://download.bbbike.org/osm/bbbike/), an online cycle route planner that interfaces with OpenStreetMap.
-OpenStreetMap data provides the networks for calculate walking times from each population cell to nearby bike share stations, using a maximum walking distance of 10 km and walking time of 30 minutes as thresholds.
-A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bike share stations within the maximum threshold.
-This process provides a more realistic measure of the friction of reaching bike share stations by taking infrastructure into account of travel times, rather than using the Euclidean distance from population cell to station.
+## 4.3. Travel Time Matrix
+
+To calculate walking times from the centroid of our micro population cells, we extracted OpenStreetMap data for Hamilton Bike Share's service area from [BBBike](https://download.bbbike.org/osm/bbbike/), an online cycle route planner that interfaces with OpenStreetMap.
+OpenStreetMap data provides the networks for calculating walking times from each population cell to nearby bicycle share stations, using a maximum walking distance of 10 km and walking time of 30 minutes as thresholds.
+A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bicycle share stations within the maximum threshold.
+This process provides a more realistic measure of the friction of reaching stations by taking infrastructure into account of travel times, rather than using the Euclidean distance from population cell to station.
 Routing and travel time calculations were completed using the `R` package `r5r`, used for rapid realistic routing operations [@Pereira2021r5r].
 
-## Data
+## 4.4. Data
 
@@
 
-Results
+5. Results
 ===================
 
-## Accessibility by Distance Thresholds
-
-Consensus regarding the distance that individuals are willing to walk to access a PBSP station is lacking, but the literature on walking behaviour can provide some guidelines to determine the thresholds for sensitivity analysis.
-Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010].
+## 5.1. Accessibility by Distance Thresholds
+
+Consensus regarding the distance that individuals are willing to walk to access a bicycle share station is lacking, but the literature on walking behaviour can provide some guidelines to determine the thresholds for sensitivity analysis.
+Previous studies have found that living within 250 m [@fullerUseNewPublic2011] and 300 m [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 m and rarely more than 1200 m [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 m [@larsenQuarterMileReexamining2010].
 In Hamilton, Hamilton Bike Share will depict a map at some stations to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are expected to walk.
+National Association of City Transportation Officials (NACTO) has a similar normative guide [@nactowalkingstation2015].
 
 In the present case, we found that congested accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at 5 minutes.
-Accessibility decreases substantially after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each bike share station.
-
-For this reason, we experiment with various walking thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times from population cell to bicycle share station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
+Accessibility decreases substantially after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each station.
+
+For this reason, we experiment with various thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times from population cell to bicycle share station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
 We categorize these thresholds as minimum, average, maximum, and extreme, respectively.
@@
 When considering the results reported below, it is important to remember that accessibility is technically a form of smoothing [@okelly2003aggregate, pp.
-7-8]: smaller thresholds produce less smoothing (which can result in "spiky" accessibility landscapes), and larger thresholds produce more smoothing, and fewer spikes.
+7-8]: smaller thresholds produce less smoothing (which can result in "spiky" accessibility landscapes), while larger thresholds produce more smoothing and fewer spikes.
 
@@
 
-### Minimum Threshold
+### 5.1.1. Minimum Threshold
 
@@
 Accessibility is fairly uniform overall, with the exception of two small areas where accessibility is slightly higher.
-This high level of system-wide accessibility occurs because the population that can reach bike share stations when travel time is 3 minutes or less is very limited, and accessibility is strongly shaped by a few locations that concentrate population and stations.
+This high level of system-wide accessibility occurs because the population that can reach bicycle share stations when travel time is 3 minutes or less is very limited, and accessibility is strongly shaped by a few locations that concentrate population and stations.
 
@@
 
-### Average Threshold
-
-With a walking distance of five minutes, we find that there are 68.6 bicycle racks per person in the original system.
+### 5.1.2. Average Threshold
+
+With a walking distance of five minutes, we find that there are 68.6 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
 With the addition of equity stations, there are now 68.8 bicycle racks per person.
@@
 
-### Maximum Threshold
-
-With a walking distance of ten minutes, we find that there are 3.61 bicycle racks per person in the original system.
+### 5.1.3. Maximum Threshold
+
+With a walking distance of ten minutes, we find that there are 3.61 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
 With the addition of equity stations, there are now 3.74 bicycles per person.
 Figure \ref{fig:figure-8} presents a comparison of accessibility between the systems.
-Differences in accessibility across the service area are now apparent, with users near the university and its adjacent neighbourhoods, as well as neighbourhoods north of the downtown area (outlined in green), having slightly higher accessibility.
+Differences in accessibility across the service area are now apparent, with users near the university and its adjacent neighborhoods, as well as neighborhoods north of the downtown area (the latter is outlined in green), having slightly higher accessibility.
 While the differences are modest, they are more apparent at this threshold than at shorter walking distances.
@@
 
-### Extreme Threshold
-
-With a walking distance of fifteen minutes, we find that there are 2.44 bicycle racks per person in the original system.
+### 5.1.4. Extreme Threshold
+
+With a walking distance of fifteen minutes, we find that there are 2.44 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
 With the addition of equity stations, there are now 2.55 bicycles per person.
 Figure \ref{fig:figure-9} presents a comparison of accessibility between the systems.
-Users near the university and the neighbourhoods north of the downtown area (outlined in green) have the highest accessibility, followed by those who live in the city's downtown area.
-Accessibility in the east end of the core service area remains low.
+Users near the university and the neighborhoods north of the downtown area (the latter is outlined in green) have the highest accessibility, followed by those who live in the city's downtown area.
+Accessibility in the east end, where equity stations were implemented, of the core service area remains low.
 
@@
 
-## Accessibility by Median Total Household Income
+## 5.2. Accessibility by Median Total Household Income
 
@@
 Therefore, we reaggregate population and accessibility from the 50-by-50 m population cells to the dissemination area.
-We then can compare accessibility to median total household income statistics from the census.
-
-Figures 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combine the spatial distribution of accessibility and median total household income, using tertiles for the coloring scheme.
-In each figure the top panel is the system without the equity stations, and the bottom panel is accessibility with the equity stations.
-As expected, the analysis indicates that the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service areas of stations.
-We find that stations added to Hamilton's public bicycle share program did indeed expand the spatial coverage of the system.
+We then can compare accessibility to median total household income statistics from the Canadian census results from 2016.
+
+Figures 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps that combine the spatial distribution of accessibility and median total household income, using tertiles for the coloring scheme.
+In each figure, the top panel is the system without the equity stations (i.e., the original system configuration), and the bottom panel is accessibility in the current system with equity stations.
+As expected, the analysis indicates that the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service area of stations.
+We find that stations added to Hamilton's public bicycle share program did indeed expand the spatial coverage of the system, leading to more balance across the service area.
 By implementing equity stations in areas with lower median total household income in Hamilton, the PBSP attained greater horizontal equity by extending the spatial distribution of bicycles across the city.
@@
 
-To further explore this issue, we aggregate accessibility by median total household income across all geographies (see Table \ref{tab:accessibility-income}).
-Similar to @hosfordWhoArePublic2018, we find that Hamilton's PBSP already serviced a large proportion of the population in the bottom and second quintiles of median total household income.
-This may well be an artifact of the spatial socio-economic and demographic profile of Hamilton, where the most dense parts of the city (where a PBSP is most easily launched) are also those with relatively lower incomes.
+To further explore this issue and examine vertical equity, we aggregate accessibility by median total household income across all geographies (see Table \ref{tab:accessibility-income}).
+Similar to Hosford and Winters  [-@hosfordWhoArePublic2018], we find that Hamilton's PBSP already serviced a large proportion of the population in the bottom and second quintiles of median total household income.
+This may well be an artifact of the spatial socioeconomic and demographic profile of Hamilton, where the most dense parts of the city (where a PBSP is most easily launched) are also those with relatively lower incomes.
 On the other hand, the levels of accessibility to Hamilton's PBSP are generally lower for populations in the bottom 20% of median total household income, compared to populations in the top 20%.
 
-The results indicate that the addition of equity stations increased horizontal equity by growing the population serviced irrespective of the walking threshold.
-The largest gains were made for dissemination areas in the second 20%, where an additional 3,073 and 5,395 people could reach a bicycle share station within three and five minutes walk, respectively, after the addition of the equity station.
-However, we find that there were only small changes in the population in the bottom 20% of median total household income who are serviced by the equity stations, and the accessibility gains are also quite modest and smaller than for populations in the second and third quintiles of median total household income.
+The spatial mapping reveals that the addition of equity stations increased horizontal equity by growing the population serviced irrespective of the walking threshold.
+The largest gains were made for dissemination areas in the second 20%, where an additional 3,073 and 5,395 people could reach a bicycle share station within three and five minutes walk, respectively, after the addition of equity stations.
+However, we find that there were only small increases in the population in the bottom 20% of median total household income who are serviced by the equity stations, and the accessibility gains are also quite modest and smaller than for populations in the second and third quintiles of median total household income.
 This suggests that income disparities persist, albeit this depends on the walking time thresholds.
@@
 
-Discussion
+6. Discussion
 ===================
 
-Using disaggregate data, in this paper we examined spatial equity and accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
-Use of a balanced floating catchment area approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro sale.
+Using disaggregated data in this paper, we examined spatial equity and accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
+Use of a balanced floating catchment area approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro scale which is a sensible approach to avoid the "absorption of disparities", as articulated by Chen et al.
+[-@chenExploringEquityPerformance2019].
 This differentiates our analysis from similar papers exploring equity in PBSPs that use larger geographical units of analysis or that focus on station location instead of level of service.
@@
 The sensitivity analysis revealed that accessibility to Hamilton Bike Share stations is maximized at five minutes and decreases significantly by eight minutes [see Table \ref{tab:accessibility-income}].
-This reflects the normative guide, advertised on some bicycle share stations in Hamilton showing other stations within a five minute walk.
+This reflects the normative guide advertised on some bicycle share stations in Hamilton showing other stations within a five minute walk, as well as the directive of NACTO [@nactowalkingstation2015].
 We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}].
-At a minimum threshold of three minutes, too few people can reach stations which leads to relatively high levels of service (since there is little crowding).
+At a minimum threshold of three minutes, too few people can reach stations which leads to relatively high levels of service since there is little crowding.
 However, accessibility is at its lowest after eight minutes whereby congestion effects due to increased potential demand kick in.
 The City of Hamilton has recognized from the launch of the PBSP that substantially more stations and bicycles are needed to service the area [@hamiltonsobi2015].
-With a service area of 40 sq.km, it is estimated that Hamilton should have between 380 and 440 stations, instead of 130, and 1,500 bicycles instead of 900 [@hamiltonsobi2015].
+With a service area of 40 sq.km, it is estimated that Hamilton should have between 380 and 440 stations instead of 130, and 1,500 bicycles instead of 900 [@hamiltonsobi2015].
 Reduced capacity within the system leads to gaps in coverage in some areas of the city "with some areas not having the recommended station density of 300m between stations or 10 stations per square km" [@hamiltonsobi2015].
@@
 Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in service were filled by these additional stations.
+This indicates that more individuals at all income quintiles can access Hamilton Bike Share, which leads to a more balanced or equal program.
 In this respect, there are some commonalities between the expansions of Hamilton Bike Share and that of Citi Bike in New York.
 Babgoli et al.
-[-@babagoliExploringHealthSpatial2019] found a slight but not statistically significant increase in the proportion of neighbourhoods with the highest levels of poverty with stations after the Citi Bike expansion in 2015.
-Although the Citi Bike expansion was not specifically driven by a desire to reduce inequity in access, 16% of neighbourhoods with the highest levels of poverty had stations compared to 12% before.
-With the addition of equity stations, there were large gains in accessibility for the second 20% at the average threshold, over 5,000 more people, but much smaller gains for the bottom 20% with only 326 more people.
-
-Vertical inequity, however, persisted despite the addition of equity stations.
+[-@babagoliExploringHealthSpatial2019] found a slight but not statistically significant increase in the proportion of neighborhoods with the highest levels of poverty with stations after the Citi Bike expansion in 2015.
+Although the Citi Bike expansion was not specifically driven by a desire to reduce inequity in access, 16% of neighborhoods with the highest levels of poverty had stations compared to 12% before.
+Similarly, with the addition of equity stations, there were large gains in accessibility for the second 20% at the average threshold, over 5,000 more people, but much smaller gains for the bottom 20% with only 326 more people.
+
+Vertical inequity, however, persisted as evidenced by differences in accessibility according to income quintile.
 While the addition of equity stations seems to modestly increase accessibility for all income groups at all thresholds, they did not increase accessibility substantially for any single income group.
@@
 At the minimum threshold, the bottom 20% have the lowest level of access.
-While previous research found that neighbourhoods with more disadvantage are better serviced by Hamilton Bike Share, the authors used the Pampalon Deprivation Index to determine the level of disadvantage for dissemination areas _across_ the city not just within the core service area [@hosfordWhoArePublic2018].
+While previous research found that neighborhoods with more disadvantage are better serviced by Hamilton Bike Share, the authors used the Pampalon Deprivation Index to determine the level of disadvantage for dissemination areas _across_ the city not just within the core service area [@hosfordWhoArePublic2018].
 Instead, we use median total household income for each dissemination area _within_ the core service area.
 We conclude that Hamilton's PBSP, while by default located in areas with more deprivation compared to other cities, has disparities in accessibility between income groups.
-With equity stations, many areas with low median household income in the east end of the service area continue to have low accessibility.
-At higher thresholds of 10 and 15 minutes, the top 20% have the highest level of access to Hamilton Bike Share.
-This aligns with other studies from Tampa [@chenExploringEquityPerformance2019] and Seattle [@mooneyFreedomStationSpatial2019], which have found disparities in station location or access to bicycles between levels of income and education.
-
-Based on our analysis, we identified specific areas that have both low accessibility and low median household income which would benefit from an increased supply of public bicycles.
+With equity stations, many areas with low median total household income in the east end of the service area continue to have low accessibility.
+At the maximum and extreme thresholds, the top 20% have the highest level of access to Hamilton Bike Share.
+These findings aligns with other studies from Tampa [@chenExploringEquityPerformance2019] and Seattle [@mooneyFreedomStationSpatial2019], which have found disparities in station location or access to bicycles between levels of income and education.
+
+Based on our analysis, we identified specific areas that have both low accessibility and low median total household income which would benefit from an increased supply of public bicycles.
 These empirical findings provide support to the City of Hamilton's efforts to increase equity and balance the system, and confirm that additional stations and bicycles are needed to improve access not only for the bottom 20% but for all income groups.
-Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} highlight potential locations for new equity stations.
-
-Study Limitations
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} highlight potential locations for new equity stations to better accommodate groups with lower socioeconomic status.
+
+7. Study Limitations
 ===================
@@
 Therefore, further research is needed to determine whether the addition of equity stations encouraged more cycling for low-income individuals living near them.
-Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighbourhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of publicly available route or user data to conduct similar analyses for Hamilton Bike Share.
+Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighborhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of publicly available route and individual user data to conduct similar analyses for Hamilton Bike Share.
 
 An additional limitation is the lack of publicly available information about the number of bicycles at each station.
-Hamilton Bike share works to balance the number of bicycles in stations, but it is reasonable to expect that the number of bicycles will not match exactly the number of racks at every station.
-Ideally, instead of number of bicycle racks as our measure of supply, we would have liked to use the average number of bicycles at stations, perhaps at different times during the day too.
-Should these data become available, it would be worthwhile to revisit the question to examine how well operation of the system (including balancing of bicycles among stations) works to maintain the nominal levels of accessibility examined in this paper.
-
-Conclusion
+Hamilton Bike share works to balance the number of bicycles across stations, but it is reasonable to expect that the number of bicycles will not match exactly the number of racks at every station.
+Ideally, instead of number of bicycle racks as our measure of supply, we would have liked to use the average number of bicycles at stations, perhaps at different times during the day or different seasons.
+Should this data become available, it would be worthwhile to revisit the question to examine how well operation of the system (including balancing of bicycles across stations) works to maintain the nominal levels of accessibility examined in this paper.
+
+8. Conclusion
 ===================
@@
 The addition of specific equity stations to the public bicycle share program in Hamilton, Ontario had the net effect of increasing accessibility and reducing to some extent both horizontal and vertical inequities.
-In particular, accessibility improved the most for those in the second 20% median household income at all thresholds, but the gains were only modest for all income groups.
+In particular, accessibility improved the most for those in the second 20% median total household income at all thresholds, but the gains were only modest for all income groups.
 Dissemination areas with the bottom 20% had the lowest accessibility at three minutes, and second lowest levels of accessibility at five, ten, and fifteen minutes.
-Congestion effects were observed at higher thresholds, with accessibility decreasing significantly once the catchment area is increased to 10 minutes walking.
+Congestion effects were observed at higher thresholds, with accessibility decreasing significantly once the catchment area is increased to ten minutes walking.
 
@@
 As such, a logical next step to this research is to examine whether Hamilton Bike Share's equity stations increased ridership or resulted in new memberships in areas that were previously under-serviced.
-An examination of the types of trips undertaken by residents in these areas would also be informative, such as the study undertaken by Caspi and Norland [-@caspiBikesharingPhiladelphiaLowerincome2019] after bike share stations were implemented in low-income Philadelphia neighbourhoods.
-The bulk of cycling facilities that have been built in Hamilton to date located in the core service area near the conventional stations.
+An examination of the types of trips undertaken by residents in these areas would also be informative, such as the study undertaken by Caspi and Norland [-@caspiBikesharingPhiladelphiaLowerincome2019] after bike share stations were implemented in low-income Philadelphia neighborhoods.
+The bulk of cycling facilities that have been built in Hamilton to date are located in the core service area near the conventional stations.
 It would be worthwhile to explore the route choice of bike share trips departing or ending at the equity stations and to identify factors that specifically influence trips from these stations, which would extend existing studies conducted by Scott and colleagues [@luUnderstandingBikeShare2018; @scottWhatFactorsInfluence2019; @scottRouteChoiceBike2021].
-This paper, combined with additional studies such as those conceptualized above, would serve as a valuable case study for Hamilton and other cities with PBSPs that wish to evaluate and address spatial inequuality in accessibility and transportation options in urban areas.
+This paper, combined with additional studies such as those conceptualized above, would serve as a valuable case study for Hamilton and other cities with PBSPs that wish to evaluate and address spatial inequality in accessibility and transportation options in urban areas.
 

### 2021-03-29 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
First version of README.Rmd before submission

@@
 library(sf)
-library(sobiEquity)
+library(sobiEquity) # Data package
 library(tidyverse)
@@
 ```{r join-population-to-travel-time-data, include=FALSE}
-# current system with equity stations
 ttm_walk <- left_join(ttm_walk, 
@@
 ```{r join-sobi-station-to-travel-time-data, include=FALSE}
-# current system with equity stations
 ttm_walk <- ttm_walk %>% 
@@
 ```{r, include=FALSE}
-#data frames for current and original systems
 names(ttm_walk) <- c("UID", 

### 2021-03-29 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
First version of README.Rmd before submission

@@
 \hspace{1em}Fourth 20\% & 20185 & 1673 & 4.119 & 1673 & 4.119 & 0 & 0.000\\
-\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{27541} & \cellcolor{gray!6}{2114} & \cellcolor{gray!6}{3.408} & \cellcolor{gray!6}{2379} & \cellcolor{gray!6}{3.434} & \cellcolor{gray!6}{265} & \cellcolor{gray!6}{0.026}\\
+\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{27541} & \cellcolor{gray!6}{2151} & \cellcolor{gray!6}{3.757} & \cellcolor{gray!6}{2416} & \cellcolor{gray!6}{3.784} & \cellcolor{gray!6}{265} & \cellcolor{gray!6}{0.027}\\
 \addlinespace[0.3em]
@@
 \hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{33312} & \cellcolor{gray!6}{17513} & \cellcolor{gray!6}{56.048} & \cellcolor{gray!6}{22908} & \cellcolor{gray!6}{56.137} & \cellcolor{gray!6}{5395} & \cellcolor{gray!6}{0.089}\\
-\hspace{1em}Third 20\% & 30940 & 15117 & 4.258 & 18309 & 4.291 & 3192 & 0.033\\
+\hspace{1em}Third 20\% & 30940 & 15117 & 4.259 & 18309 & 4.291 & 3192 & 0.032\\
 \hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{20185} & \cellcolor{gray!6}{2867} & \cellcolor{gray!6}{1.094} & \cellcolor{gray!6}{3116} & \cellcolor{gray!6}{1.095} & \cellcolor{gray!6}{249} & \cellcolor{gray!6}{0.001}\\
-\hspace{1em}Top 20\% & 27541 & 4052 & 5.925 & 4518 & 5.933 & 466 & 0.008\\
+\hspace{1em}Top 20\% & 27541 & 4074 & 6.256 & 4540 & 6.264 & 466 & 0.008\\
 \addlinespace[0.3em]
 \multicolumn{8}{l}{\textbf{Threshold - 10 minutes}}\\
-\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{43441} & \cellcolor{gray!6}{41824} & \cellcolor{gray!6}{0.603} & \cellcolor{gray!6}{41981} & \cellcolor{gray!6}{0.621} & \cellcolor{gray!6}{157} & \cellcolor{gray!6}{0.018}\\
-\hspace{1em}Second 20\% & 33312 & 27546 & 0.860 & 30503 & 0.927 & 2957 & 0.067\\
-\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{30940} & \cellcolor{gray!6}{22394} & \cellcolor{gray!6}{0.772} & \cellcolor{gray!6}{25128} & \cellcolor{gray!6}{0.799} & \cellcolor{gray!6}{2734} & \cellcolor{gray!6}{0.027}\\
+\hspace{1em}\cellcolor{gray!6}{Bottom 20\%} & \cellcolor{gray!6}{43441} & \cellcolor{gray!6}{41824} & \cellcolor{gray!6}{0.604} & \cellcolor{gray!6}{41981} & \cellcolor{gray!6}{0.622} & \cellcolor{gray!6}{157} & \cellcolor{gray!6}{0.018}\\
+\hspace{1em}Second 20\% & 33312 & 27546 & 0.862 & 30503 & 0.929 & 2957 & 0.067\\
+\hspace{1em}\cellcolor{gray!6}{Third 20\%} & \cellcolor{gray!6}{30940} & \cellcolor{gray!6}{22394} & \cellcolor{gray!6}{0.776} & \cellcolor{gray!6}{25128} & \cellcolor{gray!6}{0.802} & \cellcolor{gray!6}{2734} & \cellcolor{gray!6}{0.026}\\
 \hspace{1em}Fourth 20\% & 20185 & 4544 & 0.225 & 4989 & 0.227 & 445 & 0.002\\
-\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{27541} & \cellcolor{gray!6}{7989} & \cellcolor{gray!6}{1.160} & \cellcolor{gray!6}{9078} & \cellcolor{gray!6}{1.162} & \cellcolor{gray!6}{1089} & \cellcolor{gray!6}{0.002}\\
+\hspace{1em}\cellcolor{gray!6}{Top 20\%} & \cellcolor{gray!6}{27541} & \cellcolor{gray!6}{7989} & \cellcolor{gray!6}{1.346} & \cellcolor{gray!6}{9078} & \cellcolor{gray!6}{1.348} & \cellcolor{gray!6}{1089} & \cellcolor{gray!6}{0.002}\\
 \addlinespace[0.3em]
 \multicolumn{8}{l}{\textbf{Threshold - 15 minutes}}\\
-\hspace{1em}Bottom 20\% & 43441 & 42208 & 0.534 & 42327 & 0.554 & 119 & 0.020\\
-\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{33312} & \cellcolor{gray!6}{30507} & \cellcolor{gray!6}{0.551} & \cellcolor{gray!6}{31069} & \cellcolor{gray!6}{0.611} & \cellcolor{gray!6}{562} & \cellcolor{gray!6}{0.060}\\
-\hspace{1em}Third 20\% & 30940 & 26108 & 0.546 & 26660 & 0.573 & 552 & 0.027\\
+\hspace{1em}Bottom 20\% & 43441 & 42208 & 0.536 & 42327 & 0.557 & 119 & 0.021\\
+\hspace{1em}\cellcolor{gray!6}{Second 20\%} & \cellcolor{gray!6}{33312} & \cellcolor{gray!6}{30507} & \cellcolor{gray!6}{0.555} & \cellcolor{gray!6}{31069} & \cellcolor{gray!6}{0.614} & \cellcolor{gray!6}{562} & \cellcolor{gray!6}{0.059}\\
+\hspace{1em}Third 20\% & 30940 & 26108 & 0.554 & 26660 & 0.581 & 552 & 0.027\\
 \hspace{1em}\cellcolor{gray!6}{Fourth 20\%} & \cellcolor{gray!6}{20185} & \cellcolor{gray!6}{6312} & \cellcolor{gray!6}{0.093} & \cellcolor{gray!6}{7435} & \cellcolor{gray!6}{0.096} & \cellcolor{gray!6}{1123} & \cellcolor{gray!6}{0.003}\\
-\hspace{1em}Top 20\% & 27541 & 10209 & 0.712 & 11089 & 0.714 & 880 & 0.002\\
+\hspace{1em}Top 20\% & 27541 & 10209 & 0.808 & 11089 & 0.811 & 880 & 0.003\\
 \bottomrule
@@
 (Auguie, 2017), \texttt{kableExtra} (Zhu, 2020), \texttt{knitr} (Xie,
-2020), \texttt{pycno} (Brunsdon, 2014), \texttt{r5r} (Saraiva et al.,
-2020), \texttt{raster} (Hijmans, 2020), \texttt{rgdal} (Bivand et al.,
-2020), \texttt{rticles} (Allaire et al., 2021), \texttt{sf} (Pebesma,
-2020), \texttt{tidyverse} (Wickham, 2019), \texttt{tinytex} (Xie, 2021),
-\texttt{units} (Pebesma et al., 2020).
+2021a), \texttt{pycno} (Brunsdon, 2014), \texttt{r5r} (Saraiva et al.,
+2021), \texttt{raster} (Hijmans, 2020), \texttt{rgdal} (Bivand et al.,
+2021), \texttt{rticles} (Allaire et al., 2021), \texttt{sf} (Pebesma,
+2021), \texttt{tidyverse} (Wickham, 2019), \texttt{tinytex} (Xie,
+2021b), \texttt{units} (Pebesma et al., 2020).
 
@@
 \leavevmode\hypertarget{ref-R-rgdal}{}%
-Bivand, R., Keitt, T., Rowlingson, B., 2020. Rgdal: Bindings for the
+Bivand, R., Keitt, T., Rowlingson, B., 2021. Rgdal: Bindings for the
 geospatial data abstraction library.
@@
 \leavevmode\hypertarget{ref-R-sf}{}%
-Pebesma, E., 2020. Sf: Simple features for r.
+Pebesma, E., 2021. Sf: Simple features for r.
 
@@
 \leavevmode\hypertarget{ref-R-r5r}{}%
-Saraiva, M., Pereira, R.H.M., Herszenhut, D., Braga, C.K.V., 2020. R5r:
+Saraiva, M., Pereira, R.H.M., Herszenhut, D., Braga, C.K.V., 2021. R5r:
 Rapid realistic routing with r5.
@@
 \leavevmode\hypertarget{ref-R-knitr}{}%
-Xie, Y., 2020. Knitr: A general-purpose package for dynamic report
+Xie, Y., 2021a. Knitr: A general-purpose package for dynamic report
 generation in r.
@@
 \leavevmode\hypertarget{ref-R-tinytex}{}%
-Xie, Y., 2021. Tinytex: Helper functions to install and maintain tex
+Xie, Y., 2021b. Tinytex: Helper functions to install and maintain tex
 live, and compile latex documents.

### 2021-03-29 README.Rmd
First version of README.Rmd before submission

@@
 ---
-output: github_document
+output:
+  github_document:
+    pandoc_args: --webtex
+always_allow_html: true
+bibliography: [Bike-share-spatial-equity/packages.bib, Bike-share-spatial-equity/references.bib]
 ---
@@
 
-The goal of Accessibility-Sobi-Hamilton is to ...
+Elise Desjardins (McMaster University)  
+Christopher D. Higgins (University of Toronto Scarborough)  
+Antonio Páez (McMaster University)
 
-What is special about using `README.Rmd` instead of just `README.md`? You can include R chunks like so:
+Paper submitted to *Transportation Research Part D: Transport and Environment*
 
-```{r cars}
-summary(cars)
+## Abstract
+
+Public bicycle share programs (PBSPs) can play a role in advancing transportation equity if they make cycling more accessible to disadvantaged populations.
+In Ontario, the City of Hamilton expanded their PBSP in 2018 by adding twelve docking stations with the explicit objective of increasing spatial equity in access.
+In this case study, we investigate differentials in accessibility to stations using a balanced floating catchment area accessibility approach and compare accessibility with and without the equity stations.
+We analyze micro zones to better reflect walking to a station and conduct a sensitivity analysis at several walking time thresholds.
+We then reaggregate the estimated accessibility for further analysis using census data.
+Our findings indicate that equity stations increased the serviced population at every threshold examined.
+Although accessibility increased for the whole population, this increase was relatively modest especially for population in the bottom 20% of median total household income.
+
+## Keywords
+
+- Public Bike Sharing Programs  
+- Transport equity  
+- Horizontal equity  
+- Vertical equity  
+- Accessibility  
+- Balanced floating catchment areas  
+- Hamilton  
+- Ontario
+- Reproducible research
+
+```{r load-packages, include=FALSE}
+library(biscale)
+library(cowplot)
+library(data.table)
+library(disk.frame)
+library(dplyr)
+library(gdistance)
+library(gridExtra)
+library(kableExtra)
+library(pycno)
+library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.htmllibrary(readr)
+library(raster)
+library(rgdal)
+library(sf)
+library(sobiEquity) # Data package
+library(tidyverse)
+library(tinytex)
+library(units)
 ```
 
-You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date.
-`devtools::build_readme()` is handy for this.
-You could also use GitHub Actions to re-render `README.Rmd` every time you push.
-An example workflow can be found here: <https://github.com/r-lib/actions/tree/master/examples>.
-
-You can also embed plots, for example:
-
-```{r pressure, echo = FALSE}
-plot(pressure)
+```{r generate-package-bibliography, include=FALSE}
+knitr::write_bib(c(.packages(), "knitr", "rticles"), "packages.bib")
 ```
 
-In that case, don't forget to commit and push the resulting figure files, so they display on GitHub.
+```{r invoke-data, include=FALSE}
+# The data is part of data package `sobiEquity` 
+data("community_downtown")
+data("hamilton_cma")
+data("hamilton_da_2016")
+data("population_50x50")
+data("sobi_service")
+data("sobi_hubs")
+data("ttm_walk")
+```
+
+```{r crop-das, include=FALSE}
+# Crop DAs using bounding box of service area
+sobi_service_bbox <- sobi_service %>% 
+  st_bbox() %>%
+  st_as_sfc()
+
+cropped_da <- hamilton_da_2016 %>%
+  st_crop(sobi_service_bbox) %>%
+  dplyr::select(DA, geometry)
+```
+
+\newpage
+
+## Introduction
+
+The potential of public bicycle share programs (PBSPs) to increase cycling levels is but one of many reasons for implementing such programs in urban areas [@hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019].
+As a healthy, inexpensive, and convenient form of public transportation, shared bicycles can encourage individuals to take up cycling for short local trips or first and last mile trips to other public transportation instead of using personal vehicles.
+These programs can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to individuals and groups with lower socioeconomic status.
+As these programs become increasingly common, their introduction has been accompanied by a flurry of research that investigates them from the perspective of spatial and transportation equity [e.g., @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @hullgrassoBikeShareEquity2020; @qianBikesharingEquityDisadvantaged2020; @qianBikeshareDestinationChoices2021; @smith2015exploring].
+Indeed, although these programs are available to the general public and ought to be accessible to any individual who wishes to use them, research on PBSPs located in North American and European cities indicates that inequity in spatial accessibility needs redressing.
+
+In this paper, we investigate the case of the PBSP in the city of Hamilton, in Ontario, Canada.
+Hamilton Bike Share was launched in 2015 and currently has over 900 operational bicycles and 130 docking stations.
+An interesting feature of this program is that it was expanded by an equity initiative that introduced twelve stations in 2018 with the explicit objective of increasing spatial accessibility.
+Hamilton Bike Share was already studied by Hosford and Winters [-@hosfordWhoArePublic2018] as part of a selection of PBSPs in Canada.
+The authors found that most of the cities in their research could benefit of greater efforts to expand service to areas with lower socioeconomic groups - but Hamilton fared somewhat better than other cities in this respect.
+What distinguishes our research from existing literature is the use of a balanced floating catchment area (BFCA) accessibility approach that accounts for the supply of stations and the potential demand from the population serviced.
+This requires a more disaggregated approach than the use of census geographies [q.v., @hosfordWhoArePublic2018], since walking trips to a bicycle share station likely happen at a level lower than even the smallest census geography.
+For this reason, we implement our analysis parting from micro population zones to better reflect the friction of walking to a docking station, which is an important component of a bike share trip [@chenExploringEquityPerformance2019].
+Further, we conduct a sensitivity analysis at several walking time thresholds.
+In terms of spatial equity, we compare accessibility with and without the equity stations to assess the effect of the initiative, before reaggregating the data for further analysis using median total household income information from the census.
+
+This paper is an example of open and reproducible research that uses only open software for transportation and statistical analysis [@bivand2020progress; @lovelace2021open].
+All data were obtained from publicly available sources and organized in the form of a data package.
+Following best practices in spatial data science [@brunsdon2020opening], the code and data needed to reproduce, modify or extend the analysis are available for download.^[https://github.com/paezha/Accessibility-Sobi-Hamilton]
+
+## Literature Review
+
+Public bicycle share programs have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users [@fishmanBikeshareReviewRecent2016].
+In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @ogilvieInequalitiesUsagePublic2012; @reillyGenderDisparitiesNew2020; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011].
+However, one study found that bike share users in Washington, DC were more likely to be female [@buckAreBikeshareUsers2013], which suggests that the gender gap among cyclists who use bike share is less disparate than the gap for personal bicycle use [@fishmanBikeshareReviewRecent2016].
+There is some evidence that bike share users are less likely to own a car [@buckAreBikeshareUsers2013; @reillyNoncyclistsFrequentCyclists2020].
+However, the relationship between income or education and bike share use is less clear-cut.
+Stations in disadvantaged communities in Chicago have been found to generate most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighborhoods in Minneapolis-St.
+Paul used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019].
+Similar findings were reported in London [@ogilvieInequalitiesUsagePublic2012].
+Being university educated was a significant correlate of bike share use in Montreal, Canada [@fullerUseNewPublic2011].
+Perhaps not coincidentally, financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016].
+
+Many studies have found that geographic proximity to a bicycle share station is an important determinant of membership and use [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011].
+This makes sense given that individuals are more likely to use services or programs that they can easily reach.
+Several studies have recently explored the equity of PBSPs in North America by primarily examining who has access to bike share (e.g., differences by demographics or socioeconomic status) and where stations are located.
+It is important to note that equity can be achieved in two different ways: balanced or equal spatial distribution across a region for all similar groups (e.g., horizontal equity) or greater and targeted access to accommodate vulnerable or disadvantaged populations (e.g., vertical equity) [see @chenExploringEquityPerformance2019].
+Both are of interest to researchers and transportation planners since they are often linked in that advantage, or conversely disadvantage, has spatial patterns.
+Using a negative binomial regression model, Qian and Jaller [-@qianBikesharingEquityDisadvantaged2020] estimated ridership in Chicago's PBSP and found some disparities.
+A minority of bike share stations were found to be located in disadvantaged communities, while annual members from disadvantaged communities have a lower share of trips compared to other areas in the city.
+This suggests that they may be using PBSPs for utilitarian trips (e.g., commuting), which points to the importance of ensuring equitable access [@qianBikesharingEquityDisadvantaged2020].
+Similar results were found in Philadelphia.
+Despite efforts to increase equity within the city's PBSP, census block groups with lower median income generated fewer trips which suggested that efforts have not been as successful as intended [@caspiBikesharingPhiladelphiaLowerincome2019].
+Trips from stations in such areas are utilitarian (e.g., commuting to work), which points to the importance of ensuring equitable access [@caspiBikesharingPhiladelphiaLowerincome2019].
+In the case of Seattle, all neighborhoods had some level of access to dockless bicycles but those with higher incomes and more residents of higher education had more bikes [@mooneyFreedomStationSpatial2019].
+Babagoli et al.
+[-@babagoliExploringHealthSpatial2019] also found that neighborhoods in New York City with higher affluence had the greatest proportion of Citi Bike stations.
+
+On the whole, existing studies highlight the need for PBSPs to be more highly accessible for diverse populations in order to increase use beyond the "typical" users.
+This has been the focus of recent research [see, among others, @auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020].
+Offering more people the option of using sustainable and active transportation, particularly those who have lower socioeconomic status and might benefit the most, is a worthy policy goal for cities with PBSPs.
+A few North American cities have launched specific programs to address barriers to bike share or to expand service to more deprived areas: Philadelphia, Pennsylvania [@caspiBikesharingPhiladelphiaLowerincome2019], and Hamilton, Ontario [@hosfordWhoArePublic2018].
+However, exploring transportation equity by investigating where bicycle share stations are located, often using neighborhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
+Meaning that, stations may be lacking in certain neighborhoods but there may be stations accessible within a reasonable walking time.
+This is where geographical accessibility becomes an important consideration.
+
+Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved over the past century and has increasingly become linked with recent planning interests in prioritizing modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020].
+As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequality in the provision of infrastructure, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options.
+It also addresses some of the challenges of other performance measures such as level of service within a transportation network by measuring person-based indicators and exploring differences in use between population groups [@paezMeasuringAccessibilityPositive2012].
+
+Beyond the utility derived from using shared bikes to destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019].
+Since the time or distance needed to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., number of bicycle racks) of stations matters.
+Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
+Kabra et al.
+[-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300 m of stations, which amounts to 2-4 minutes walking by an able-bodied adult.
+Furthermore, similar to other public amenities affected by crowding, like health care services [e.g., @pereiraGeographicAccessCOVID192021], the utility of stations is limited by the _maximum_ number of bicycles that they can hold.
+The program may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles.
+Likewise, more people may not opt to use the program if the supply of bicycles available at nearby stations is insufficient to meet demand.
+The location and size of stations is important to increase the utility of this public transportation option for more people, thus achieving vertical equity.
+Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts.
+For instance, Wang and Lindsey [-@wangNewBikeShare2019] investigated whether new or relocated bicycle share stations increased accessibility and use, which offered important insights to improve the performance of the program.
+
+Several approaches have been commonly used for measuring place-based accessibility, including cumulative opportunities, gravity, and utility-based measures [@handyMeasuringAccessibilityExploration1997].
+Geurs and Van Wee [-@geursAccessibilityEvaluationLanduse2004] and Paez et al.
+[-@paezMeasuringAccessibilityPositive2012] provide recent overviews of various formulations and applications of accessibility in transportation research.
+The common gravity-based approach for example involves weighting destination opportunities, such as the quantity of bicycle share stations, by the time required to reach them from an origin using an impedance function [@handyMeasuringAccessibilityExploration1997; @kwanSpaceTimeIntegral1998].
+However, while such measures are suitable for capturing the potential for reaching destinations from a given location, they do not take demand or congestion effects into account.
+
+In contrast, floating catchment area (FCA) methods incorporate information on capacity and demand in calculating accessibility.
+FCA measures have been widely employed in healthcare accessibility research and are typically calculated across two steps.
+In the first, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hospital, weighted by the distance involved in reaching the facility.
+Next, these service level ratios are allocated back to the population centers and summarized as a measure of congested accessibility.
+Thus, this model does a good job of considering potential crowding or competition for services.
+While there have been many methodological innovations in FCA methods [for example, @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balancing of the impedance that addressed the effects of demand and service inflation found in earlier FCA approaches [see @paezDemandLevelService2019].
+
+When measuring accessibility, researchers have taken different approaches when it comes to aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones.
+Previous research on bike share equity has typically used a meso- or macro-level approach with aggregated data from entire neighborhoods or census tracts [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021].
+This is also true for studies examining correlates of bike share demand [@wangNewBikeShare2019].
+Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals, which is useful for addressing vertical inequity in PBSP usage.
+Chen et al.
+[-@chenExploringEquityPerformance2019, pp.
+530] are in favour of using disaggregated data, which they did in their recent analysis of Tampa's PBSP, because they note "the use of aggregated data might hinder our understanding of the equity impacts since individual disparities are absorbed after aggregation".
+
+Using the balanced floating catchment area method [@paezDemandLevelService2019], a novel approach that has not been used yet in cycling research, with disaggregated population data, we examine accessibility to a public bicycle share program in Hamilton, Ontario.
+In this paper, we (1) conduct a sensitivity analysis by measuring accessibility and level of service to bicycle share stations at different walking time thresholds to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median total household income of dissemination areas within the core service area.
+
+## Case Study {#sec:study}
+
+### Original System
+
+The focus of this study is the city of Hamilton, located in Ontario, Canada.
+The city launched a public bicycle share program in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015].
+Before June 2020, the program was known as Social Bicycles or SoBi Hamilton, but is now called Hamilton Bike Share.
+Stations are spaced between 300 and 600 m apart [@scottRouteChoiceBike2021].
+The core service area spans 40 sq.km of the city although it was planned to be 20 sq.km [@hamiltonsobi2015], and roughly 138,000 people are within 30 minutes walking of a bike share station [see Figure \ref{fig:hamilton-and-sobi-service-area}].
+This represents roughly one fifth of the total population of the Hamilton Census Metropolitan Area according to the 2016 Canadian Census.
+The City of Hamilton undertook a large public engagement campaign to validate the locations of bicycle share stations that had been selected and to crowdsource potential locations for additional locations [@hamiltonsobi2014].
+Most of the suggested locations were in the east end of the core service area that lack stations or in neighborhoods not serviced by the PBSP.
+The program was enthusiastically welcomed in the city in 2015 - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015], however inadequate coverage given the size of Hamilton Bike Share's service area was identified as a problem early on, and transportation planners noted that the small size (i.e., supply of bicycle racks) and low quantity of stations would lead to challenges in balancing the system [@hamiltonsobi2015].
+Based on public feedback, thirteen stations were added to the original system.
+
+### Equity Initiative
+
+In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), with the objective of reducing barriers that may prevent individuals from accessing bike share in Hamilton.
+Additional bicycles and stations were added to the program which expanded it to more disadvantaged areas in the core service area.
+The program also offers subsidized memberships to individuals who identify as low income, and complements this service with cycle skills education.
+A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019].
+
+### Current System
+
+As of June 2020, Hamilton Bike Share has 900 bikes, 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
+The core service area remains 40 sq.km.
+The PBSP has twelve stations that were added as part of the ERI initiative; we refer to these as "equity stations" throughout the paper while all other stations are "conventional stations".
+In total, 118 stations are "conventional" and 12 are "equity".
+
+### Membership
+
+Hamilton Bike Share conducted one membership survey in 2018 [@civicplanSoBiHamiltonMembership2017] and the findings from 420 members are broadly in line with the trends that were discussed above [see @fishmanBikeshareReviewRecent2016 for a recent review of the literature].
+The majority of respondents live within the core service area and the gender split is expected: 57% of respondents are male and 41% are female.
+The majority of respondents, both male and female, are between 25 and 34 years of age, but the percentage of male respondents is higher in the subsequent age groups.
+Respondents use bike share for commuting (40% of trips) or errands and meetings (24% of trips) and nearly 50% of trips have an average length of 11 to 20 minutes.
+As a result of having a bike share membership, 49% of respondents report that they use their private vehicle less often or much less often but 48% report that their private vehicle use has remained about the same.
+This suggests that Hamilton Bike Share has been useful for certain kinds of trips but not all, meaning that some trips continue to require a personal vehicle.
+
+### Relevant Research
+
+Our analysis builds upon a previous and recent study [@hosfordWhoArePublic2018], which found that areas in Hamilton with less advantage are better served by the city's PBSP compared to other Canadian cities [i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau] where areas that are less deprived have greater access.
+Hosford and Winters [-@hosfordWhoArePublic2018] acknowledge that "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas".
+Therefore, the core service area for the PBSP in Hamilton by default covers some of the less advantaged areas in the city.
+However, there is also a great deal of variation in income in the city center because of the local university and increasing gentrification.
+Hosford and Winters [-@hosfordWhoArePublic2018] took a macro-level and system-wide approach in their analysis by using dissemination areas _across_ the city as the unit of analysis.
+They did not focus specifically on the core service area and did not differentiate between conventional and equity stations.
+
+```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "The core service area of Hamilton Bike Share is outlined in black.
+Hamilton Census Metropolitan Area is shown in grey."}
+ggplot() + 
+  geom_sf(data = hamilton_da_2016,
+          color = NA,
+          fill = "gray") + 
+  geom_sf(data = sobi_service,
+          fill = NA,
+          colour = "black") + 
+  theme(panel.background = element_rect(fill = "white"))
+```
+
+```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bicycle share stations in Hamilton, Ontario.
+The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in pink.
+Hamilton Census Metropolitan Area is shown in grey."}
+ggplot() + 
+  geom_sf(data = cropped_da,
+          color = NA,
+          fill = "gray") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "black") +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          size = 1,
+          color = "pink") +
+  geom_sf(data = sobi_hubs %>% 
+            dplyr::filter(hub_type == "Conventional" & 
+                            hub_status == "Active"), # Remove deactivated hubs
+          aes(size = RACKS_AMOU),
+          shape = 21,
+          fill = "blue") +
+  geom_sf(data =  sobi_hubs %>% 
+            dplyr::filter(hub_type == "ERI"), 
+          aes(size = RACKS_AMOU), 
+          shape = 21,
+          fill = "green") +
+  scale_size(name = "Number of racks", range = c(0.5, 4)) +
+  theme(panel.background = element_rect(fill = "white"))
+```
+
+## Methods and Data {#sec:methods}
+
+### Floating Catchment Area
+
+Floating catchment area (FCA) methods are an approach commonly used in the healthcare accessibility literature.
+This approach is more appropriate and informative than calculating provider-to-population ratios (PPR) that simply divide the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019].
+In particular, the Two-Step Floating Catchment Area (2SFCA) method [@luoMeasuresSpatialAccessibility2003; @radkeSpatialDecompositionsModeling2000] produces flexible catchment areas instead of using rigid boundaries like PPR.
+This provides more useful information because it does not assume that people are limited to service within pre-defined boundaries [@paezDemandLevelService2019].
+This is an important property that supports our rationale for applying this method to measure accessibility to Hamilton Bike Share.
+The City of Hamilton has positioned stations between 300 and 600 m apart, but anticipates that hubs will service those living within a 250 m buffer from the station [@hamiltonsobi2015].
+The latter constitutes a normative statement: people ought to be able to access a station in less than 600 m if they live in the core service area, with usage coming from 250 m around.
+However, it is not known how far people are actually willing to travel to reach a station.
+It would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.
+
+More recently, the _balanced_ floating catchment area approach was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by earlier FCA methods [@paezDemandLevelService2019].
+Briefly, overlapping catchment areas lead to inflation of population totals and deflation of service levels across a study area and generates inaccurate or misleading accessibility estimates.
+In contrast, Paez et al.
+[-@paezDemandLevelService2019] adjusted the impedance weights so that both supply and demand are proportionally allocated.
+The result is a FCA method that balances the population and level of service by eliminating the over-counting of population and level of service that leads to distortions in demand and supply.
+Other benefits of this adjusted method include consideration of competition which can occur when catchment areas overlap, as well as the preservation of population and level of service.
+Balanced floating catchment area methods have been used to explore accessibility to health care providers [@paezDemandLevelService2019] and COVID-19 health care services [@pereiraGeographicAccessCOVID192021], but have not yet been used in the cycling literature to explore issues of accessibility.
+
+The first step in the floating catchment area method is to allocate the population to be serviced by each Hamilton Bike Share station:
+$$
+P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}
+$$
+
+As seen in the equation above, the population allocated to station $j$ is the weighted sum of the population in the region; a spatial weight $w_{ij}$ represents the friction that the population at $i$ faces when reaching station $j$, and is usually given by a distance-decay function, so that each station is assumed to service only a segment of the population within a limited geographical range.
+
+Next, the supply at each station (i.e., the maximum number of bicycle racks) is divided by its estimated service population within the established catchment area; this gives the level of service of station $j$ in bicycle racks per person:
+$$
+L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
+$$
+
+Finally, the accessibility of population cell $i$ is calculated as the weighted sum of the level of service of all stations that can be reached from there according to the spatial weights:
+$$
+A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
+$$
+
+The balanced approach of Paez et al [-@paezDemandLevelService2019] replaces the spatial weights with normalized versions as follows:
+$$
+{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
+$$
+
+\noindent and:
+$$
+{w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
+$$
+
+These weights satisfy the following properties:
+$$
+\sum_{j = 1}^{J} {w^i_{ji}} = 1
+$$
+
+\noindent and:
+$$
+\sum_{i = 1}^{n} {w^j_{ji}} = 1
+$$
+
+With these weights, accessibility can be calculated without risk of demand or supply inflation:
+$$
+A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}
+$$
+
+By allocating the population and level of service proportionally, this method preserves the values of the population and level of service and provides a regional provider-to-population ratio since:
+$$
+\sum_{i=1}^n A_i = \sum_{j=1}^J L_j = \frac{\sum_{j=1}^JS_j}{\sum_{i=1}^n P_i}
+$$
+
+In fact, since the proportional allocation procedure means that any proportion of the population allocated to a station is never allocated to other stations, and conversely any level of service allocated to a population is never re-allocated elsewhere, this property is replicated for any level of aggregation.
+For this paper, we employ a hybrid location-based and person-based approach to calculating accessibility using disaggregated population data.
+In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data.
+This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are better able of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences.
+Applying the balanced floating catchment area approach allows us to examine accessibility by stratifying according to median total household income, which constitutes the individual component of the accessibility measure [@geursAccessibilityEvaluationLanduse2004].
+However, conducting a further sensitivity analysis to measure accessibility at different walking time thresholds would help to consider potential spatio-temporal constraints.
+Different people may be willing to travel different distances to access a public bicycle share station.
+
+### Pycnophylactic Interpolation
+
+To obtain population at sub-census geography levels (at the micro scale), we use pycnophylactic interpolation [@toblerSmoothPycnophylacticInterpolation1979].
+We obtained population data from the 2016 Census of Canada for dissemination areas (DA), which is the smallest publicly available census geography in Canada.
+These zonal values of the population were interpolated to smaller polygons 50-by-50 m in size.
+Pycnophylactic interpolation involves smoothing out the population from each dissemination area while preserving total volume [see Figure \ref{fig:da-population}].
+When interpolating the population at this high level of resolution, it is important to ensure that population numbers were not allocated to areas where people do not live in Hamilton (for example, to parks, large institutional buildings, etc.).
+To do so, we retrieved shapefiles for various geographic features [see Table \ref{tab:data-features}] from Open Hamilton.
+Next, we removed these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population within the remaining area [see Figure \ref{fig:interpolated-population}].
+
+```{r data-features, echo=FALSE}
+data_tbl <- data.frame(
+  Feature = c("Hamilton Bike Share Stations", "Golf Courses", "Parks", "Designated Large Employment Areas", "Municipal Parking Lots", "Cemeteries", "Environmentally Sensitive Areas", "Streets", "Educational Institutions", "Places of Worship", "Municipal Service Centers", "Recreation and Community Centers", "Arenas", "Emergency Stations", "Fire Stations", "Police Stations", "Railways", "Hospitals"),
+  Description = c(
+    "The location of stations and the number of racks available at each station.",
+    "The location of City and privately owned golf courses.", 
+    "The location of parks and other green spaces.",
+    "The location of large business parks and industrial lands.",
+    "The location of municipal car parks.",
+    "The location of cemeteries.",
+    "The location of either land or water areas containing natural features or significant ecological functions.",
+    "The street network in Hamilton, including road classification for highways.",
+    "The location of all educational institutions and schools.",
+    "The location of buildings used for religious congregations.",
+    "The location of all municipal service centers, including City Hall.",
+    "The location of all recreation and community centers.",
+    "The location of all indoor arenas.",
+    "The location of all Emergency Management Services (EMS) Ambulance stations.",
+    "The location of all fire stations.",
+    "The location of all police stations.",
+    "The railway network in Hamilton.",
+    "The location of all hospitals."
+  )
+)
+
+kable(data_tbl,
+      "html",
+      #"latex",
+      booktabs = TRUE,
+      caption = "\\label{tab:landscape-features}Landscape features extracted from the Hamilton Bike Share core service area before population is interpolated.") %>%
+  kable_paper(full_width = F) %>%
+  column_spec(1, bold = T, border_right = T) %>%
+  column_spec(2, width = "30em") %>%
+  kable_styling(latex_options = c("striped", "scale_down"))
+```
+
+```{r da-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Population in all dissemination areas (outlined in white) that are inside or touch the bounding box of the Hamilton Bike Share core service area (outlined in black)."}
+ggplot() +
+  geom_sf(data = hamilton_da_2016 %>%
+            dplyr::filter(sobi_service == "Yes"),
+          aes(fill = population),
+          color = "white") +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  scale_fill_distiller(palette = "YlOrRd", 
+                       direction = 1) +
+  theme(panel.background = element_rect(fill = "white"))
+```
+
+```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Interpolated population in the Hamilton Bike Share's core service area (outlined in black) and within 30 minutes of walking to the core service area.
+Each population cell is 50-by-50 m in size.
+The downtown area is outlined in dark green."}
+ggplot() + 
+  geom_sf(data = cropped_da,
+          fill = "gray",
+          color = NA) + 
+  #geom_sf(data = population_50x50,
+  #        aes(color = population)) +
+   #scale_fill_gradient(colours = "YlOrRd") +
+  geom_tile(data = cbind(st_drop_geometry(population_50x50), 
+                         st_coordinates(population_50x50)),
+            aes(x = X, 
+                y = Y, 
+                fill = population)) +
+  scale_fill_distiller(palette = "YlOrRd", 
+                        direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA,
+          colour = "black") +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"))
+```
+
+### Travel Time Matrix
+
+To calculate walking times from the centroid of our micro population cells, we extracted OpenStreetMap data for Hamilton Bike Share's service area from [BBBike](https://download.bbbike.org/osm/bbbike/), an online cycle route planner that interfaces with OpenStreetMap.
+OpenStreetMap data provides the networks for calculating walking times from each population cell to nearby bicycle share stations, using a maximum walking distance of 10 km and walking time of 30 minutes as thresholds.
+A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bicycle share stations within the maximum threshold.
+This process provides a more realistic measure of the friction of reaching stations by taking infrastructure into account of travel times, rather than using the Euclidean distance from population cell to station.
+Routing and travel time calculations were completed using the `R` package `r5r`, used for rapid realistic routing operations [@Pereira2021r5r].
+
+### Data
+
+All data for this research were accessed from publicly available Census of Canada sources, from OpenStreetMaps, and from Open Hamilton^[https://open.hamilton.ca/], a public online repository of data curated by the City of Hamilton.
+
+## Results
+
+### Accessibility by Distance Thresholds
+
+Consensus regarding the distance that individuals are willing to walk to access a bicycle share station is lacking, but the literature on walking behaviour can provide some guidelines to determine the thresholds for sensitivity analysis.
+Previous studies have found that living within 250 m [@fullerUseNewPublic2011] and 300 m [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 m and rarely more than 1200 m [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 m [@larsenQuarterMileReexamining2010].
+In Hamilton, Hamilton Bike Share will depict a map at some stations to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are expected to walk.
+National Association of City Transportation Officials (NACTO) has a similar normative guide [@nactowalkingstation2015].
+
+In the present case, we found that congested accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at 5 minutes.
+Accessibility decreases substantially after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each station.
+
+For this reason, we experiment with various thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times from population cell to bicycle share station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
+We categorize these thresholds as minimum, average, maximum, and extreme, respectively.
+At each threshold, we compare accessibility between the current system and the original system to examine the contribution of the additional equity stations.
+When considering the results reported below, it is important to remember that accessibility is technically a form of smoothing [@okelly2003aggregate, pp.
+7-8]: smaller thresholds produce less smoothing (which can result in "spiky" accessibility landscapes), while larger thresholds produce more smoothing and fewer spikes.
+
+```{r join-population-to-travel-time-data, include=FALSE}
+ttm_walk <- left_join(ttm_walk, 
+                      population_50x50 %>%
+                        st_drop_geometry(),
+                      by = "UID")
+```
+
+```{r join-sobi-station-to-travel-time-data, include=FALSE}
+ttm_walk <- ttm_walk %>% 
+  left_join(sobi_hubs %>%
+              st_drop_geometry() %>%
+              dplyr::select(OBJECTID, 
+                            RACKS_AMOU),
+            by = c("OBJECTID" = "OBJECTID"))
+```
+
+```{r, include=FALSE}
+names(ttm_walk) <- c("UID", 
+                     "hub", 
+                     "travel_time",
+                     "hub_type",
+                     "hub_status",
+                     "population",
+                     "racks")
+```
+
+```{r accessibility-sensibility, eval=FALSE, include=FALSE}
+# Testing different thresholds to check how accessibility changes
+accessibility_test <- data.frame(threshold = c(3:15), 
+                                 accessibility = numeric(13))
+for(i in 1:13){
+  results_test <- ttm_walk %>%
+    dplyr::filter(hub_type == "Conventional" &
+                    hub_status == "Active") %>%
+    b2sfca(threshold = i +2 )
+  accessibility_test$accessibility[i] <- sum(results_test$accessibility$accessibility)
+}
+
+ggplot(accessibility_test, 
+       aes(x = threshold, y = accessibility)) +
+  geom_line()
+```
+
+<!-- Calculate accessibility for different thresholds -->
+
+```{r 3-min-conventional, include=FALSE}
+results_3min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 3)
+```
+
+```{r 3-min-all, include=FALSE}
+results_3min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 3)
+```
+
+```{r 3-min-sum-conventional, include=FALSE}
+sum(results_3min_conventional$los$los)
+sum(results_3min_conventional$accessibility$accessibility)
+```
+
+```{r 3-min-sum-all, include=FALSE}
+sum(results_3min_all$los$los)
+sum(results_3min_all$accessibility$accessibility)
+```
+```{r 5-min-conventional, include=FALSE}
+results_5min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 5)
+```
+
+```{r 5-min-all, include=FALSE}
+results_5min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 5)
+```
+
+```{r, 5-min-sum-conventional, include=FALSE}
+sum(results_5min_conventional$los$los)
+sum(results_5min_conventional$accessibility$accessibility)
+```
+
+```{r 5-min-sum-all, include=FALSE}
+sum(results_5min_all$los$los)
+sum(results_5min_all$accessibility$accessibility)
+```
+
+```{r 10-min-conventional, include=FALSE}
+results_10min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 10)
+```
+
+```{r 10-min-all, include=FALSE}
+results_10min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 10)
+```
+
+```{r 10-min-sum-conventional, include=FALSE}
+sum(results_10min_conventional$los$los)
+sum(results_10min_conventional$accessibility$accessibility)
+```
+
+```{r, 10-min-sum-all, include=FALSE}
+sum(results_10min_all$los$los)
+sum(results_10min_all$accessibility$accessibility)
+```
+
+```{r 15-min-conventional, include=FALSE}
+results_15min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 15)
+```
+
+```{r 15-min-all, include=FALSE}
+results_15min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 15)
+```
+
+```{r 15-min-sum-conventional, include=FALSE}
+sum(results_15min_conventional$los$los)
+sum(results_15min_conventional$accessibility$accessibility)
+```
+
+```{r 15-min-sum-all, include=FALSE}
+sum(results_15min_all$los$los)
+sum(results_15min_all$accessibility$accessibility)
+```
+
+```{r, bind-all-results, include=FALSE}
+los <- rbind(data.frame(results_3min_all$los, system = "All stations", threshold = "3 min"),
+             data.frame(results_3min_conventional$los, system = "Conventional", threshold = "3 min"),
+             data.frame(results_5min_all$los, system = "All stations", threshold = "5 min"),
+             data.frame(results_5min_conventional$los, system = "Conventional", threshold = "5 min"),
+             data.frame(results_10min_all$los, system = "All stations", threshold = "10 min"),
+             data.frame(results_10min_conventional$los, system = "Conventional", threshold = "10 min"),
+             data.frame(results_15min_all$los, system = "All stations", threshold = "15 min"),
+             data.frame(results_15min_conventional$los, system = "Conventional", threshold = "15 min"))
+
+accessibility <- rbind(data.frame(results_3min_all$accessibility, system = "All stations", threshold = "3 min"),
+                       data.frame(results_3min_conventional$accessibility, system = "Conventional", threshold = "3 min"),
+                       data.frame(results_5min_all$accessibility, system = "All stations", threshold = "5 min"),
+                       data.frame(results_5min_conventional$accessibility, system = "Conventional", threshold = "5 min"),
+                       data.frame(results_10min_all$accessibility, system = "All stations", threshold = "10 min"),
+                       data.frame(results_10min_conventional$accessibility, system = "Conventional", threshold = "10 min"),
+                       data.frame(results_15min_all$accessibility, system = "All stations", threshold = "15 min"),
+                       data.frame(results_15min_conventional$accessibility, system = "Conventional", threshold = "15 min"))
+```
+
+```{r, add-geometry-to-results, include=FALSE}
+los <- los %>%
+  left_join(sobi_hubs %>%
+              dplyr::select(OBJECTID),
+            by = c("hub" = "OBJECTID")) %>%
+  st_as_sf()
+
+accessibility <- accessibility %>%
+  left_join(population_50x50 %>%
+              dplyr::select(UID),
+            by = c("UID" = "UID")) %>%
+  st_as_sf()
+```
+
+#### Minimum Threshold
+
+With a walking distance of three minutes, we find that there are 25.2 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+The addition of equity stations increases this ratio slightly to 25.4 bicycles per person.
+Figure \ref{fig:figure-6} presents a comparison of accessibility between the systems for the population cells.
+Accessibility is fairly uniform overall, with the exception of two small areas where accessibility is slightly higher.
+This high level of system-wide accessibility occurs because the population that can reach bicycle share stations when travel time is 3 minutes or less is very limited, and accessibility is strongly shaped by a few locations that concentrate population and stations.
+
+```{r figure-6, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
+#accessibility plot for current system with equity stations at 3 minutes walk
+accessibility_3min <- accessibility %>% 
+  dplyr::filter(threshold == "3 min")
+
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "gray") +
+  #geom_sf(data = accessibility %>%
+  #          dplyr::filter(threshold == "3 min"),
+  #        aes(color = accessibility)) +
+  #scale_color_distiller(palette = "PuRd", 
+  #                      direction = 1) +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_3min), st_coordinates(accessibility_3min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_distiller(palette = "PuRd", direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ threshold)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  # geom_sf(data =  sobi_hubs %>% 
+  #           dplyr::filter(hub_type == "ERI"), 
+  #         size = 1.5, shape = 21, 
+  #         fill = "green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank())
+```
+
+#### Average Threshold
+
+With a walking distance of five minutes, we find that there are 68.6 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+With the addition of equity stations, there are now 68.8 bicycle racks per person.
+At this threshold, there are more bicycle racks per person than at the minimum threshold.
+System-wide accessibility has in fact increased: the population that can reach the stations has grown, but not to the the point that congestion effects begin to take place.
+Figure \ref{fig:figure-7} presents a comparison of accessibility between the systems.
+Again, accessibility is fairly uniform, with the exception of one very small area.
+
+```{r figure-7, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 5 minutes walk (average threshold) compared between current system with equity stations and the original system without equity stations."}
+#accessibility plot for current system with equity stations at 5 minutes walk
+accessibility_5min <- accessibility %>% 
+  dplyr::filter(threshold == "5 min")
+
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "gray") +
+  #geom_sf(data = accessibility %>%
+  #          dplyr::filter(threshold == "5 min"),
+  #        aes(color = accessibility)) +
+  #scale_color_distiller(palette = "PuRd", 
+  #                      direction = 1) +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_5min), st_coordinates(accessibility_5min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_distiller(palette = "PuRd", direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ threshold)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  # geom_sf(data =  sobi_hubs %>% 
+  #           dplyr::filter(hub_type == "ERI"), 
+  #         size = 1.5, shape = 21, 
+  #         fill = "green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank())
+```
+
+#### Maximum Threshold
+
+With a walking distance of ten minutes, we find that there are 3.61 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+With the addition of equity stations, there are now 3.74 bicycles per person.
+Figure \ref{fig:figure-8} presents a comparison of accessibility between the systems.
+Differences in accessibility across the service area are now apparent, with users near the university and its adjacent neighborhoods, as well as neighborhoods north of the downtown area (the latter is outlined in green), having slightly higher accessibility.
+While the differences are modest, they are more apparent at this threshold than at shorter walking distances.
+
+```{r figure-8, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) compared between current system with equity stations and the original system without equity stations."}
+#accessibility plot for current system with equity stations at 10 minutes walk
+accessibility_10min <- accessibility %>% 
+  dplyr::filter(threshold == "10 min")
+
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  #geom_sf(data = accessibility %>%
+  #          dplyr::filter(threshold == "10 min"),
+  #        aes(color = accessibility)) +
+  #scale_color_distiller(palette = "PuRd", 
+  #                      direction = 1) +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_10min), st_coordinates(accessibility_10min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_distiller(palette = "PuRd", direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ threshold)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  # geom_sf(data =  sobi_hubs %>% 
+  #           dplyr::filter(hub_type == "ERI"), 
+  #         size = 1.5, shape = 21, 
+  #         fill = "green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank())
+```
+
+#### Extreme Threshold
+
+With a walking distance of fifteen minutes, we find that there are 2.44 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+With the addition of equity stations, there are now 2.55 bicycles per person.
+Figure \ref{fig:figure-9} presents a comparison of accessibility between the systems.
+Users near the university and the neighborhoods north of the downtown area (the latter is outlined in green) have the highest accessibility, followed by those who live in the city's downtown area.
+Accessibility in the east end, where equity stations were implemented, of the core service area remains low.
+
+```{r figure-9, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 15 minutes walk (extreme threshold) compared between current system with equity stations and the original system without equity stations."}
+#accessibility plot for current system with equity stations at 15 minutes walk
+accessibility_15min <- accessibility %>% 
+  dplyr::filter(threshold == "15 min")
+
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  #geom_sf(data = accessibility %>%
+  #          dplyr::filter(threshold == "15 min"),
+  #        aes(color = accessibility)) +
+  #scale_color_distiller(palette = "PuRd", 
+  #                      direction = 1) +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_15min), st_coordinates(accessibility_15min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_distiller(palette = "PuRd", direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+
+  facet_grid(system ~ threshold)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  # geom_sf(data =  sobi_hubs %>% 
+  #           dplyr::filter(hub_type == "ERI"), 
+  #         size = 1.5, shape = 21, 
+  #         fill = "green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank())
+```
+
+### Accessibility by Median Total Household Income
+
+A unique property of the balanced floating catchment area method is that data can be reaggregated while preserving the total population in the service area and the supply at each station.
+This avoids demand and supply inflation, and also enables us to present findings in a way that is perhaps more intuitive to interpret.
+Therefore, we reaggregate population and accessibility from the 50-by-50 m population cells to the dissemination area.
+We then can compare accessibility to median total household income statistics from the Canadian census results from 2016.
+
+Figures 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps that combine the spatial distribution of accessibility and median total household income, using tertiles for the coloring scheme.
+In each figure, the top panel is the system without the equity stations (i.e., the original system configuration), and the bottom panel is accessibility in the current system with equity stations.
+As expected, the analysis indicates that the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service area of stations.
+We find that stations added to Hamilton's public bicycle share program did indeed expand the spatial coverage of the system, leading to more balance across the service area.
+By implementing equity stations in areas with lower median total household income in Hamilton, the PBSP attained greater horizontal equity by extending the spatial distribution of bicycles across the city.
+This is particularly evident at the minimum and average thresholds of three and five minutes, respectively, where the equity stations fill a number of gaps in program coverage.
+
+```{r, include=FALSE}
+# drop geometry
+test <- accessibility
+
+# join accessibility to population cells
+test <- test %>% 
+  st_join(population_50x50 %>% 
+            dplyr::select(-UID)) %>%
+  rename(population50x50 = population)
+```
+
+```{r, include=FALSE}
+test <- test %>%
+  st_join(hamilton_da_2016)
+```
+
+```{r, include=FALSE}
+agg_accessibility_da <- test %>%
+  group_by(DA, 
+           system,
+           threshold) %>%
+  summarize(accessibility = sum(accessibility),
+            median_total_income = first(median_total_income), 
+            .groups = "drop")
+```
+
+```{r figure-10, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "3 min", 
+                system == "All stations") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_3_all <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+
+legend <- bi_legend(pal = "DkCyan",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6) + 
+  theme(plot.background = element_rect(color = "lightgrey"))
+```
+
+```{r figure-11, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "3 min", 
+                system == "Conventional") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_3_conv <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+```
+
+```{r figure-bi-map-threshold-3, include = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): without equity stations (top panel) and with equity stations (bottom panel)"}
+ggdraw() + draw_plot(bi_map_3_conv, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_3_all, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
+```
+
+```{r, out.width="0.3\\linewidth", echo=FALSE, fig.align="center", fig.cap=c("Bivariate map of accessibility and income (threshold: 3 min): without equity stations (top panel) and with equity stations (bottom panel)"), echo=FALSE}
+knitr::include_graphics("./Bike-share-spatial-equity/Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1.pdf")
+```
+
+```{r figure-12, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "5 min", 
+                system == "All stations") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_5_all <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+```
+
+```{r figure-13, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "5 min", 
+                system == "Conventional") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_5_conv <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+```
+
+```{r figure-bi-map-threshold-5, include = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): without equity stations (top panel) and with equity stations (bottom panel)"}
+ggdraw() + draw_plot(bi_map_5_conv, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_5_all, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
+```
+
+```{r, echo=FALSE, fig.align="center", fig.cap=c("Bivariate map of accessibility and income (threshold: 5 min): without equity stations (top panel) and with equity stations (bottom panel)"), echo=FALSE}
+knitr::include_graphics("./Bike-share-spatial-equity/Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1.pdf")
+```
+
+```{r figure-14, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "10 min", 
+                system == "All stations") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_10_all <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+```
+
+```{r figure-15, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "10 min", 
+                system == "Conventional") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_10_conv <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+```
+
+```{r figure-bi-map-threshold-10, include = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): without equity stations (top panel) and with equity stations (bottom panel)"}
+ggdraw() + 
+  draw_plot(bi_map_10_conv, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_10_all, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
+```
+
+```{r, echo=FALSE, fig.align="center", fig.cap=c("Bivariate map of accessibility and income (threshold: 10 min): without equity stations (top panel) and with equity stations (bottom panel)"), echo=FALSE}
+knitr::include_graphics("./Bike-share-spatial-equity/Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-10-1.pdf")
+```
+
+```{r figure-16, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "15 min", 
+                system == "All stations") %>%
+  drop_na() %>% 
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_15_all <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+```
+
+```{r figure-17, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "15 min", 
+                system == "Conventional") %>%
+  drop_na() %>% 
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_15_conv <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  bi_scale_fill(pal = "DkCyan", 
+                dim = 3) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+```
+
+```{r figure-bi-map-threshold-15, include = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): without equity stations (top panel) and equity stations (bottom panel)"}
+ggdraw() + 
+  draw_plot(bi_map_15_conv, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_15_all, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
+```
+
+```{r, echo=FALSE, fig.align="center", fig.cap=c("Bivariate map of accessibility and income (threshold: 15 min): without equity stations (top panel) and with equity stations (bottom panel)"), echo=FALSE}
+knitr::include_graphics("./Bike-share-spatial-equity/Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-15-1.pdf")
+```
+
+```{r, include=FALSE}
+# Find the income quantiles of the population served by SoBi
+iq <- hamilton_da_2016 %>% 
+  dplyr::filter(sobi_service == "Yes") %>%
+  pull(median_total_income) %>%
+  quantile(c(0, 0.2, 0.4, 0.6, 0.8, 1), na.rm = TRUE)
+```
+
+```{r population-core-service-by-income, include=FALSE}
+# Find the population by income quantile
+population_by_income <- population_50x50 %>%
+  dplyr::select(-population) %>%
+  st_join(hamilton_da_2016) %>%
+  st_drop_geometry() %>%
+  distinct(DA,
+           .keep_all = TRUE) %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE)) %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population)) %>%
+  drop_na()
+```
+
+```{r income-all-3, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_all_3 <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "3 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_all_3 <- agg_accessibility_all_3 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-con-3, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_con_3 <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "3 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_con_3 <- agg_accessibility_con_3 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-all-5, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_all_5 <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "5 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_all_5 <- agg_accessibility_all_5 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-con-5, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_con_5 <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "5 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_con_5 <- agg_accessibility_con_5 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-all-10, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_all_10 <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_all_10 <- agg_accessibility_all_10 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-con-10, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_con_10 <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_con_10 <- agg_accessibility_con_10 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-all-15, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_all_15 <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "15 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_all_15 <- agg_accessibility_all_15 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-con-15, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_con_15 <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "15 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_con_15 <- agg_accessibility_con_15 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-all-hubs, include=FALSE}
+all_hubs <- rbind(inc_stats_all_3,
+                  inc_stats_all_5,
+                  inc_stats_all_10,
+                  inc_stats_all_15) %>%
+  drop_na() # there is a very small number of population cells without income data
+
+all_hubs
+```
+
+```{r income-conventional-hubs, include=FALSE}
+conventional_hubs <- rbind(inc_stats_con_3,
+                           inc_stats_con_5,
+                           inc_stats_con_10,
+                           inc_stats_con_15) %>%
+  drop_na() # there is a very small number of population cells without income data
+
+conventional_hubs
+```
+
+To further explore this issue and examine vertical equity, we aggregate accessibility by median total household income across all geographies (see Table \ref{tab:accessibility-income}).
+Similar to Hosford and Winters  [-@hosfordWhoArePublic2018], we find that Hamilton's PBSP already serviced a large proportion of the population in the bottom and second quintiles of median total household income.
+This may well be an artifact of the spatial socioeconomic and demographic profile of Hamilton, where the most dense parts of the city (where a PBSP is most easily launched) are also those with relatively lower incomes.
+On the other hand, the levels of accessibility to Hamilton's PBSP are generally lower for populations in the bottom 20% of median total household income, compared to populations in the top 20%.
+
+The spatial mapping reveals that the addition of equity stations increased horizontal equity by growing the population serviced irrespective of the walking threshold.
+The largest gains were made for dissemination areas in the second 20%, where an additional 3,073 and 5,395 people could reach a bicycle share station within three and five minutes walk, respectively, after the addition of equity stations.
+However, we find that there were only small increases in the population in the bottom 20% of median total household income who are serviced by the equity stations, and the accessibility gains are also quite modest and smaller than for populations in the second and third quintiles of median total household income.
+This suggests that income disparities persist, albeit this depends on the walking time thresholds.
+With and without equity stations, people in the top 20% of income have the highest level of access at a threshold of ten and fifteen minutes.
+Although dissemination areas in the second 20% have the highest level of access by a significant amount at lower distance thresholds, the bottom 20%, who may benefit the most from Hamilton Bike Share's equity initiatives, have the lowest access at three minutes threshold and the second lowest access at all other thresholds.
+
+```{r accessibility-income, echo=FALSE}
+accessibility_income <- cbind(
+  conventional_hubs,
+  all_hubs[, 2:3])
+
+colnames(accessibility_income) <- c("iq", "population_conv", "accessibility_conv", "population_all", "accessibility_all")
+
+accessibility_income <- accessibility_income %>%
+  transmute(iq,
+            population = rep(population_by_income$population, 4),
+            population_conv = round(population_conv),
+            accessibility_conv = round(accessibility_conv, 3), 
+            population_all = round(population_all),
+            accessibility_all = round(accessibility_all, 3),
+            population_diff = population_all - population_conv, 
+            accessibility_diff = accessibility_all - accessibility_conv)
+
+kable(accessibility_income,
+      "html",
+      #"latex",
+      booktabs = TRUE,
+      col.names = c("Income Quintile",
+                    "Total Population",
+                    "Population",
+                    "Accessibility",
+                    "Population",
+                    "Accessibility",
+                    "Population",
+                    "Accessibility"),
+      align = c("l", "c", "c", "c", "c", "c", "c"),
+      caption = "\\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile and between systems (with and without equity stations).
+Total population is the population by income quintile in the DAs that have any PBSP service at all.") %>%
+  kable_styling(latex_options = c("striped", "scale_down")) %>%
+  pack_rows("Threshold - 3 minutes", 1, 5, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Threshold - 5 minutes", 6, 10, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Threshold - 10 minutes", 11, 15, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Threshold - 15 minutes", 16, 20, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  add_header_above(c(" " = 2, "Without Equity Stations" = 2, "With Equity Stations" = 2, "Difference" = 2)) %>%
+  #column_spec(2, color = "white",
+  #background = spec_color(accessibility_income$accessibility[1:40], end = 1)) %>%
+  footnote(general = " ",
+           alphabet = c("With equity stations = Hamilton Bike Share current system (118 conventional stations, 12 equity stations)", "Without equity stations = Hamilton Bike Share original system (118 conventional stations, no equity stations)"))
+```
+
+## Discussion
+
+Using disaggregated data in this paper, we examined spatial equity and accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
+Use of a balanced floating catchment area approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro scale which is a sensible approach to avoid the "absorption of disparities", as articulated by Chen et al.
+[-@chenExploringEquityPerformance2019].
+This differentiates our analysis from similar papers exploring equity in PBSPs that use larger geographical units of analysis or that focus on station location instead of level of service.
+In this way, our paper has made contributions in a positive way by applying an intuitive and useful approach to measure accessibility to a PBSP, and in a normative way by serving to inform future investments in cycle infrastructure for Hamilton Bike Share.
+
+The sensitivity analysis revealed that accessibility to Hamilton Bike Share stations is maximized at five minutes and decreases significantly by eight minutes [see Table \ref{tab:accessibility-income}].
+This reflects the normative guide advertised on some bicycle share stations in Hamilton showing other stations within a five minute walk, as well as the directive of NACTO [@nactowalkingstation2015].
+We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}].
+At a minimum threshold of three minutes, too few people can reach stations which leads to relatively high levels of service since there is little crowding.
+However, accessibility is at its lowest after eight minutes whereby congestion effects due to increased potential demand kick in.
+The City of Hamilton has recognized from the launch of the PBSP that substantially more stations and bicycles are needed to service the area [@hamiltonsobi2015].
+With a service area of 40 sq.km, it is estimated that Hamilton should have between 380 and 440 stations instead of 130, and 1,500 bicycles instead of 900 [@hamiltonsobi2015].
+Reduced capacity within the system leads to gaps in coverage in some areas of the city "with some areas not having the recommended station density of 300m between stations or 10 stations per square km" [@hamiltonsobi2015].
+This study illustrates the consequences of an imbalanced system whereby levels of accessibility are not equitable across income groups.
+
+We found that equity stations increased horizontal equity in Hamilton's core service area.
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in service were filled by these additional stations.
+This indicates that more individuals at all income quintiles can access Hamilton Bike Share, which leads to a more balanced or equal program.
+In this respect, there are some commonalities between the expansions of Hamilton Bike Share and that of Citi Bike in New York.
+Babgoli et al.
+[-@babagoliExploringHealthSpatial2019] found a slight but not statistically significant increase in the proportion of neighborhoods with the highest levels of poverty with stations after the Citi Bike expansion in 2015.
+Although the Citi Bike expansion was not specifically driven by a desire to reduce inequity in access, 16% of neighborhoods with the highest levels of poverty had stations compared to 12% before.
+Similarly, with the addition of equity stations, there were large gains in accessibility for the second 20% at the average threshold, over 5,000 more people, but much smaller gains for the bottom 20% with only 326 more people.
+
+Vertical inequity, however, persisted as evidenced by differences in accessibility according to income quintile.
+While the addition of equity stations seems to modestly increase accessibility for all income groups at all thresholds, they did not increase accessibility substantially for any single income group.
+Most importantly, individuals in the bottom 20% of median total household income have the second lowest level of access to Hamilton Bike Share at most thresholds (average, maximum, and extreme).
+At the minimum threshold, the bottom 20% have the lowest level of access.
+While previous research found that neighborhoods with more disadvantage are better serviced by Hamilton Bike Share, the authors used the Pampalon Deprivation Index to determine the level of disadvantage for dissemination areas _across_ the city not just within the core service area [@hosfordWhoArePublic2018].
+Instead, we use median total household income for each dissemination area _within_ the core service area.
+We conclude that Hamilton's PBSP, while by default located in areas with more deprivation compared to other cities, has disparities in accessibility between income groups.
+With equity stations, many areas with low median total household income in the east end of the service area continue to have low accessibility.
+At the maximum and extreme thresholds, the top 20% have the highest level of access to Hamilton Bike Share.
+These findings aligns with other studies from Tampa [@chenExploringEquityPerformance2019] and Seattle [@mooneyFreedomStationSpatial2019], which have found disparities in station location or access to bicycles between levels of income and education.
+
+Based on our analysis, we identified specific areas that have both low accessibility and low median total household income which would benefit from an increased supply of public bicycles.
+These empirical findings provide support to the City of Hamilton's efforts to increase equity and balance the system, and confirm that additional stations and bicycles are needed to improve access not only for the bottom 20% but for all income groups.
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} highlight potential locations for new equity stations to better accommodate groups with lower socioeconomic status.
+
+## Study Limitations
+
+This paper did not examine or compare ridership data between conventional and equity stations.
+Therefore, further research is needed to determine whether the addition of equity stations encouraged more cycling for low-income individuals living near them.
+Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighborhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of publicly available route and individual user data to conduct similar analyses for Hamilton Bike Share.
+
+An additional limitation is the lack of publicly available information about the number of bicycles at each station.
+Hamilton Bike share works to balance the number of bicycles across stations, but it is reasonable to expect that the number of bicycles will not match exactly the number of racks at every station.
+Ideally, instead of number of bicycle racks as our measure of supply, we would have liked to use the average number of bicycles at stations, perhaps at different times during the day or different seasons.
+Should this data become available, it would be worthwhile to revisit the question to examine how well operation of the system (including balancing of bicycles across stations) works to maintain the nominal levels of accessibility examined in this paper.
+
+## Conclusion
+
+The addition of specific equity stations to the public bicycle share program in Hamilton, Ontario had the net effect of increasing accessibility and reducing to some extent both horizontal and vertical inequities.
+In particular, accessibility improved the most for those in the second 20% median total household income at all thresholds, but the gains were only modest for all income groups.
+Dissemination areas with the bottom 20% had the lowest accessibility at three minutes, and second lowest levels of accessibility at five, ten, and fifteen minutes.
+Congestion effects were observed at higher thresholds, with accessibility decreasing significantly once the catchment area is increased to ten minutes walking.
+
+Wang and Lindsey [-@wangNewBikeShare2019] have noted that there is a lack of research that examines how bike share users' behaviour changes as a result of program changes to station locations or improvements in accessibility.
+As such, a logical next step to this research is to examine whether Hamilton Bike Share's equity stations increased ridership or resulted in new memberships in areas that were previously under-serviced.
+An examination of the types of trips undertaken by residents in these areas would also be informative, such as the study undertaken by Caspi and Norland [-@caspiBikesharingPhiladelphiaLowerincome2019] after bike share stations were implemented in low-income Philadelphia neighborhoods.
+The bulk of cycling facilities that have been built in Hamilton to date are located in the core service area near the conventional stations.
+It would be worthwhile to explore the route choice of bike share trips departing or ending at the equity stations and to identify factors that specifically influence trips from these stations, which would extend existing studies conducted by Scott and colleagues [@luUnderstandingBikeShare2018; @scottWhatFactorsInfluence2019; @scottRouteChoiceBike2021].
+This paper, combined with additional studies such as those conceptualized above, would serve as a valuable case study for Hamilton and other cities with PBSPs that wish to evaluate and address spatial inequality in accessibility and transportation options in urban areas.
+
+## Acknowledgments
+
+This research was completed using open software, and the authors wish to acknowledge the developers of the following `R` packages: `biscale` [@R-biscale], `cowplot` [@R-cowplot], `data.table` [@R-data.table], `disk.frame` [@R-disk.frame], `gdistance` [@R-gdistance], `gridExtra` [@R-gridExtra], `kableExtra` [@R-kableExtra], `knitr` [@R-knitr], `pycno` [@R-pycno], `r5r` [@R-r5r], `raster` [@R-raster], `rgdal` [@R-rgdal], `rticles` [@R-rticles], `sf` [@R-sf], `tidyverse` [@R-tidyverse], `tinytex` [@R-tinytex],
+`units` [@R-units].
+
+References {#references .unnumbered}
+===================

### 2021-03-29 sobiEquity.Rcheck/00_pkg_src/sobiEquity/R/sobiEquity.R

@@ -1,7 +1,8 @@
 #' sobiEquity
 #'
-#' sobiEquity is a companion package for the paper "Examining spatial equity and accessibility
-#'  to a public bicycle share program using a balanced floating catchment area approach".
+#' sobiEquity is a companion package for the paper "Examining horizontal and vertical 
+#' equity to a public bicycle share program using a balanced floating catchment area 
+#' accessibility approach".
 #'   The package includes the data used in the analysis.
 #'
 #' @author Elise Desjardins, \email{desjae@@mcmaster.ca}
@@ -21,3 +22,77 @@ NULL
 #' @usage data(hamilton_cma)
 #' @source Statistics Canada
 "hamilton_cma"
+
+#' Hamilton Community Downtown.
+#'
+#' Boundary of the downtown center of Hamilton, in Ontario, Canada.
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords transportation
+#' @name community_downtown
+#' @usage data(community_downtown)
+#' @source Open Hamilton
+"community_downtown"
+
+#' Hamilton Dissemination Areas.
+#'
+#' Boundary of the Hamilton DAs, in Ontario, Canada, with median total household income and
+#' population levels within each DA.
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords transportation
+#' @name hamilton_da_2016
+#' @usage data(hamilton_da_2016)
+#' @source Statistics Canada
+"hamilton_da_2016"
+
+#' Hamilton Interpolated Population.
+#'
+#' Interpolated population from Hamilton DAs, in 50-by-50 m cells.
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords transportation
+#' @name population_50x50
+#' @usage data(population_50x50)
+#' @source Statistics Canada
+"population_50x50"
+
+#' Hamilton Bike Share Service Area
+#'
+#' Boundary of the Hamilton Bike Share service area of Hamilton, in Ontario, Canada.
+#'
+#' @docType data
+#' @keywords datasets
+#' @keywords transportation
+#' @name sobi_service
+#' @usage data(sobi_service)
+#' @source Open Hamilton
+"sobi_service"
+
+#' Hamilton Bike Share Stations
+#'
+#' Location of the Hamilton Bike Share stations of Hamilton, in Ontario, Canada.
+#'
+#' @docType data
+#' @keywords datasets
[…]

### 2021-03-29 sobiEquity.Rcheck/sobiEquity-Ex.R

@@ -1,0 +1,52 @@
+pkgname <- "sobiEquity"
+source(file.path(R.home("share"), "R", "examples-header.R"))
+options(warn = 1)
+options(pager = "console")
+base::assign(".ExTimings", "sobiEquity-Ex.timings", pos = 'CheckExEnv')
+base::cat("name\tuser\tsystem\telapsed\n", file=base::get(".ExTimings", pos = 'CheckExEnv'))
+base::assign(".format_ptime",
+function(x) {
+  if(!is.na(x[4L])) x[1L] <- x[1L] + x[4L]
+  if(!is.na(x[5L])) x[2L] <- x[2L] + x[5L]
+  options(OutDec = '.')
+  format(x[1L:3L], digits = 7L)
+},
+pos = 'CheckExEnv')
+
+### * </HEADER>
+library('sobiEquity')
+
+base::assign(".oldSearch", base::search(), pos = 'CheckExEnv')
+base::assign(".old_wd", base::getwd(), pos = 'CheckExEnv')
+cleanEx()
+nameEx("b2sfca")
+### * b2sfca
+
+flush(stderr()); flush(stdout())
+
+base::assign(".ptime", proc.time(), pos = "CheckExEnv")
+### Name: b2sfca
+### Title: Compute balanced floating catchment area accessibility
+### Aliases: b2sfca
+
+### ** Examples
+
+#add(1, 1)
+#add(10, 1)
+
+
+
+base::assign(".dptime", (proc.time() - get(".ptime", pos = "CheckExEnv")), pos = "CheckExEnv")
+base::cat("b2sfca", base::get(".format_ptime", pos = 'CheckExEnv')(get(".dptime", pos = "CheckExEnv")), "\n", file=base::get(".ExTimings", pos = 'CheckExEnv'), append=TRUE, sep="\t")
+### * <FOOTER>
+###
+cleanEx()
+options(digits = 7L)
+base::cat("Time elapsed: ", proc.time() - base::get("ptime", pos = 'CheckExEnv'),"\n")
+grDevices::dev.off()
+###
+### Local variables: ***
+### mode: outline-minor ***
+### outline-regexp: "\\(> \\)?### [*]+" ***
+### End: ***
+quit('no')

### 2021-03-31 Bike-share-spatial-equity/Cover-Letter.Rmd
Final draft of Cover Letter for Submission
Title of paper is revised to fix grammatical error; changes made at the request of the editor are described

@@
 Please find attached our paper submitted to the Virtual Special Issue in Transportation Research Part D: Transport and Environment (SI: Transportation Justice).
-The paper is titled "Examining spatial equity in accessibility to a public bicycle share program using a balanced floating catchment area approach".
+The paper is titled "Examining spatial equity of bike share: a balanced floating catchment area approach".
 We use the city of Hamilton, Ontario in Canada as a case study.
@@
 Specifically, we use a recent improvement to this approach that was achieved through a simple and intuitive balancing that addressed the effects of demand and service inflation from previous FCA approaches.
-We also compare accessibility in the system with and without the equity stations at the different thresholds.
+In terms of spatial equity, we compare accessibility with and without the equity stations to assess the effect of the initiative, before reaggregating the data for further analysis using median total household income information from the census.
 
@@
 The analysis indicates that the equity stations increase accessibility, albeit only modestly, for all income quintiles.
-The second 20% median household income group has the highest level of accessibility at 3 and 5 minutes, while the top 20% has the highest level of accessibility at 10 and 15 minutes.
+The second 20% median total household income group has the highest level of accessibility at 3 and 5 minutes, while the top 20% has the highest level of accessibility at 10 and 15 minutes.
 The bottom 20%, those who would arguably benefit the most from inexpensive and healthy transportation options like shared bicycles, have the lowest level of accessibility at 3 minutes and the second lowest level at all other thresholds.
@@
 We trust that you will find the paper is worthy of inclusion in the special issue on transportation justice.
+The following corrections have been made: i) the title has been corrected to fix a grammatical error; and ii) the submission is double-spaced.
 

### 2021-03-31 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Final submission

@@
 ---
-title: Examining horizontal and vertical equity to a public bicycle share program using a balanced floating catchment area accessibility approach 
+title: Examining spatial equity of bike share - a balanced floating catchment area approach 
 author:
@@
 output: rticles::elsevier_article
+header-includes:
+    - \usepackage{setspace}\doublespacing
 ---
@@
 Offering more people the option of using sustainable and active transportation, particularly those who have lower socioeconomic status and might benefit the most, is a worthy policy goal for cities with PBSPs.
-A few North American cities have launched specific programs to address barriers to bike share or to expand service to more deprived areas: Philadelphia, Pennsylvania [@caspiBikesharingPhiladelphiaLowerincome2019], and Hamilton, Ontario [@hosfordWhoArePublic2018].
+A few North American cities have launched specific programs to address barriers to accessing bike share or to expand service to more deprived areas: Philadelphia, Pennsylvania [@caspiBikesharingPhiladelphiaLowerincome2019], and Hamilton, Ontario [@hosfordWhoArePublic2018].
 However, exploring transportation equity by investigating where bicycle share stations are located, often using neighborhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
@@
 ```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bicycle share stations in Hamilton, Ontario.
+Conventional stations are in blue and equity stations are in green.
 The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in pink.
@@
 
-Figures 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps that combine the spatial distribution of accessibility and median total household income, using tertiles for the coloring scheme.
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} depict bivariate choropleth maps that combine the spatial distribution of accessibility and median total household income, using tertiles for the coloring scheme.
 In each figure, the top panel is the system without the equity stations (i.e., the original system configuration), and the bottom panel is accessibility in the current system with equity stations.

### 2021-03-31 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Final submission

--- 2021-03-29 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
@@
             pdfauthor={},
-            pdftitle={Examining horizontal and vertical equity to a public bicycle share program using a balanced floating catchment area accessibility approach},
+            pdftitle={Examining spatial equity of bike share - a balanced floating catchment area approach},
             colorlinks=false,
@@
 % Pandoc header
+\usepackage{setspace}\doublespacing
 \usepackage{booktabs}
@@
 
-  \title{Examining horizontal and vertical equity to a public bicycle share
-program using a balanced floating catchment area accessibility approach}
+  \title{Examining spatial equity of bike share - a balanced floating catchment
+area approach}
     \author[McMaster University]{Elise Desjardins\corref{1}}
@@
 for cities with PBSPs. A few North American cities have launched
-specific programs to address barriers to bike share or to expand service
-to more deprived areas: Philadelphia, Pennsylvania (Caspi and Noland,
-2019), and Hamilton, Ontario (Hosford and Winters, 2018). However,
-exploring transportation equity by investigating where bicycle share
-stations are located, often using neighborhoods or census tracts as the
-geographical unit of analysis, can ignore or miss the benefits that may
-be derived from adjacent zones. Meaning that, stations may be lacking in
-certain neighborhoods but there may be stations accessible within a
-reasonable walking time. This is where geographical accessibility
-becomes an important consideration.
+specific programs to address barriers to accessing bike share or to
+expand service to more deprived areas: Philadelphia, Pennsylvania (Caspi
+and Noland, 2019), and Hamilton, Ontario (Hosford and Winters, 2018).
+However, exploring transportation equity by investigating where bicycle
+share stations are located, often using neighborhoods or census tracts
+as the geographical unit of analysis, can ignore or miss the benefits
+that may be derived from adjacent zones. Meaning that, stations may be
+lacking in certain neighborhoods but there may be stations accessible
+within a reasonable walking time. This is where geographical
+accessibility becomes an important consideration.
 
@@
 \caption{The spatial distribution of bicycle share stations in Hamilton, Ontario.
+Conventional stations are in blue and equity stations are in green.
 The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in pink.
@@
 
-Figures 1 {[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
-{[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
-{[}\ref{fig:figure-bi-map-threshold-10}{]}, and 4
-{[}\ref{fig:figure-bi-map-threshold-15}{]} depict bivariate choropleth
-maps that combine the spatial distribution of accessibility and median
-total household income, using tertiles for the coloring scheme. In each
+Figures \ref{fig:figure-bi-map-threshold-3},
+\ref{fig:figure-bi-map-threshold-5},
+\ref{fig:figure-bi-map-threshold-10}, and
+\ref{fig:figure-bi-map-threshold-15} depict bivariate choropleth maps
+that combine the spatial distribution of accessibility and median total
+household income, using tertiles for the coloring scheme. In each
 figure, the top panel is the system without the equity stations (i.e.,

### 2021-03-31 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Updated title in README.Rmd

@@
 ---
-title: Examining spatial equity of bike share - a balanced floating catchment area approach 
+title: "Examining spatial equity of bike share: A balanced floating catchment area approach" 
 author:

### 2021-06-15 Bike-share-spatial-equity/Comparative-Analysis.Rmd
Added a function to calculate accessibility using the conventional 2SFCA approach (no balancing to allocate population and los proportionally). New notebook has an example of inflation/deflation effects.

---
title: "R Notebook"
output: html_notebook
---

```{r load-packages, include=FALSE}
library(biscale)
library(cowplot)
library(data.table)
library(disk.frame)
library(dplyr)
library(gdistance)
library(gridExtra)
library(kableExtra)
library(pycno)
library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.htmllibrary(readr)
library(raster)
library(rgdal)
library(sf)
library(sobiEquity) # Data package
library(tidyverse)
library(tinytex)
library(units)
```

```{r invoke-data, include=FALSE}
# The data is part of data package `sobiEquity` 
data("community_downtown")
data("hamilton_cma")
data("hamilton_da_2016")
data("population_50x50")
data("sobi_service")
data("sobi_hubs")
data("ttm_walk")
```

```{r crop-das, include=FALSE}
# Crop DAs using bounding box of service area
sobi_service_bbox <- sobi_service %>% 
  st_bbox() %>%
  st_as_sfc()

cropped_da <- hamilton_da_2016 %>%
  st_crop(sobi_service_bbox) %>%
  dplyr::select(DA, geometry)
```

```{r join-population-to-travel-time-data, include=FALSE}
ttm_walk <- left_join(ttm_walk, 
                      population_50x50 %>%
                        st_drop_geometry(),
                      by = "UID")
```

```{r join-sobi-station-to-travel-time-data, include=FALSE}
ttm_walk <- ttm_walk %>% 
  left_join(sobi_hubs %>%
              st_drop_geometry() %>%
              dplyr::select(OBJECTID, 
                            RACKS_AMOU),
            by = c("OBJECTID" = "OBJECTID"))
```

```{r, include=FALSE}
names(ttm_walk) <- c("UID", 
                     "hub", 
                     "travel_time",
                     "hub_type",
                     "hub_status",
                     "population",
                     "racks")
```

# Conventional 2-stage floating catchment areas

In this approach the impedance is not normalized to allocate population and level of service proportionally. It does not have the property of preserving these quantities. The function to calculate this is `c2sfca`.

As an example, calculate the 2SFCA using a threshold of 3 minutes and the non-equity stations only:
```{r conventional-3-min-non-equity, include=FALSE}
conventional_3min_non_equity <- ttm_walk %>%
  dplyr::filter(hub_type == "Conventional" &
                  hub_status == "Active") %>%
  c2sfca(threshold = 3)
```

Still a threshold of 3 minutes, but with all stations:
```{r conventional-3-min-all, include=FALSE}
conventional_3min_all <- ttm_walk %>%
  dplyr::filter(hub_status == "Active") %>%
  c2sfca(threshold = 3)
```

Plot the results (non-equity stations as an example). First add the geometry:
```{r, add-geometry-to-results, include=FALSE}
los <- conventional_3min_non_equity$los %>%
  left_join(sobi_hubs %>%
              dplyr::select(OBJECTID),
            by = c("hub" = "OBJECTID")) %>%
  st_as_sf()

accessibility <- conventional_3min_non_equity$accessibility %>%
  left_join(population_50x50 %>%
              dplyr::select(UID),
            by = c("UID" = "UID")) %>%
  st_as_sf()
```

Then plot:
```{r}
ggplot() +
  geom_sf(data = population_50x50,
          color = "gray") +
  geom_sf(data = los,
            aes(size = los, color = los)) +
  scale_color_distiller(palette = "PuRd", direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  theme(panel.background = element_rect(fill = "white"),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

```{r}
ggplot() +
  geom_sf(data = population_50x50,
          color = "gray") +
  geom_tile(data = cbind(st_drop_geometry(accessibility), st_coordinates(accessibility)),
            aes(x = X, y = Y, fill = accessibility)) +
  scale_fill_distiller(palette = "PuRd", direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  theme(panel.background = element_rect(fill = "white"),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

We can verify that the los and accessibility are not identical, a consequence of the non-proportional allocation of population and los:
```{r conventional-3-min-sum-non-equity}
sum(conventional_3min_non_equity$los$los)
sum(conventional_3min_non_equity$accessibility$accessibility)
```

```{r conventional-3-min-sum-all}
sum(conventional_3min_non_equity$los$los)
sum(conventional_3min_non_equity$accessibility$accessibility)
```

# Balanced 2 stage floating catchment areas

The balanced 2-stage floating catchment areas approach 
```{r balanced-3-min-non-equity, include=FALSE}
balanced_3min_non_equity <- ttm_walk %>%
  dplyr::filter(hub_type == "Conventional" &
                  hub_status == "Active") %>%
  b2sfca(threshold = 3)
```

```{r balanced-3-min-all, include=FALSE}
balanced_3min_all <- ttm_walk %>%
  dplyr::filter(hub_status == "Active") %>%
  b2sfca(threshold = 3)
```

```{r balanced-3-min-sum-non-equity}
sum(balanced_3min_non_equity$los$los)
sum(balanced_3min_non_equity$accessibility$accessibility)
```

```{r balanced-3-min-sum-all}
sum(conventional_3min_all$los$los)
sum(conventional_3min_all$accessibility$accessibility)
```

Plot the results (non-equity stations as an example). First add the geometry:
```{r, balanced-add-geometry-to-results, include=FALSE}
los <- balanced_3min_non_equity$los %>%
  left_join(sobi_hubs %>%
              dplyr::select(OBJECTID),
            by = c("hub" = "OBJECTID")) %>%
  st_as_sf()

accessibility <- balanced_3min_non_equity$accessibility %>%
  left_join(population_50x50 %>%
              dplyr::select(UID),
            by = c("UID" = "UID")) %>%
  st_as_sf()
```

Then plot:
```{r}
ggplot() +
  geom_sf(data = population_50x50,
          color = "gray") +
  geom_sf(data = los,
            aes(size = los, color = los)) +
  scale_color_distiller(palette = "PuRd", direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  theme(panel.background = element_rect(fill = "white"),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

```{r}
ggplot() +
  geom_sf(data = population_50x50,
          color = "gray") +
  geom_tile(data = cbind(st_drop_geometry(accessibility), st_coordinates(accessibility)),
            aes(x = X, y = Y, fill = accessibility)) +
  scale_fill_distiller(palette = "PuRd", direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  theme(panel.background = element_rect(fill = "white"),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

# Compare results

There is an inflation/deflection effect in the conventional 2-stage floating catchment area method. By how much is it inflated? Compare:
```{r}
los_inflation <- data.frame(hub = conventional_3min_non_equity$los$hub, 
                       los_inflation = conventional_3min_non_equity$los$los/balanced_3min_non_equity$los$los) %>% # Ratio of conventional los to balanced los
  left_join(sobi_hubs %>%
              dplyr::select(OBJECTID),
            by = c("hub" = "OBJECTID")) %>%
  st_as_sf()

accessibility_inflation <- data.frame(UID = conventional_3min_non_equity$accessibility$UID,
                            accessibility_inflation = conventional_3min_non_equity$accessibility$accessibility/balanced_3min_non_equity$accessibility$accessibility) %>%
  left_join(population_50x50 %>%
              dplyr::select(UID),
            by = c("UID" = "UID")) %>%
  st_as_sf()
```

Plot inflation and deflation:
```{r}
ggplot() +
  geom_sf(data = population_50x50,
          color = "gray") +
  geom_sf(data = los_inflation,
            aes(size = los_inflation, color = los_inflation)) +
  scale_color_distiller(palette = "PuRd", direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  theme(panel.background = element_rect(fill = "white"),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

```{r}
ggplot() +
  geom_sf(data = population_50x50,
          color = "gray") +
  geom_tile(data = cbind(st_drop_geometry(accessibility_inflation), st_coordinates(accessibility_inflation)),
            aes(x = X, y = Y, fill = accessibility_inflation)) +
  scale_fill_distiller(palette = "PuRd", direction = 1) +
  geom_sf(data = sobi_service,
          fill = NA) +
  geom_sf(data = community_downtown %>%
            dplyr::filter(NAME == "Hamilton"),
          fill = NA,
          color = "dark green") +
  theme(panel.background = element_rect(fill = "white"),
        axis.title.x=element_blank(),
        axis.title.y=element_blank())
```

### 2021-06-15 sobiEquity.Rcheck/00_pkg_src/sobiEquity/R/c2sfca.R

@@ -1,0 +1,61 @@
+#' Compute conventional floating catchment area accessibility
+#'
+#' @param ttm A travel time matrix with a column of origin IDs, a column of destination IDs, travel times, type of destinations, population at the origin, and supply at the destination.
+#' @param threshold A threshold for the binary impedance function
+#' @return The level of service \code{los} and accessibility \code{accessibility}.
+#' @export
+#' @examples
+#' #add(1, 1)
+#' #add(10, 1)
+
+c2sfca <- function(ttm = ttm, threshold = threshold){
+  # calculate the impedance
+  ttm <- ttm %>%
+    dplyr::mutate(impedance_binary = ifelse(travel_time <= threshold, 1, 0))
+
+  # Calculate the sum of the impedance by population for the balancing factors. In the binary case, this should be the same as the number of SoBi hubs that each interpolated population unit can reach
+  #sum_b1 <- ttm %>%
+  #  dplyr::group_by(UID) %>%
+  #  dplyr::summarize(sum_b1 = sum(impedance_binary),
+  #            .groups = "drop")
+
+  # Calculate the sum of the impedance by SoBi hubs for the balancing factors. In the binary case, this should be the same as the number of interpolated population cells that each SoBi hub serves
+  #sum_b2 <- ttm %>%
+  #  dplyr::group_by(hub) %>%
+  #  dplyr::summarize(sum_b2 = sum(impedance_binary))
+
+  # Join the sum of the impedance to the table to calculate the balanced impedance values:
+  #ttm <- ttm %>%
+  #  dplyr::left_join(sum_b1, by = "UID") %>%
+  #  dplyr::left_join(sum_b2, by = "hub")
+
+  # **Important:** notice that some values of `sum_b1` are zeros! This means that some population units cannot reach a SoBi hub in at most 24 min. Since the balancing impedance is calculated by diving the impedance by `sum_b1`, this will lead to NaNs (divisions by zero). To avoid issues we will remove any population units that reach zero hubs:
+  #ttm <- ttm %>%
+  #  dplyr::filter(sum_b1 > 0,
+  #                sum_b2 > 0)
+
+  # Calculate balanced impedance values:
+  ttm <- ttm %>%
+    dplyr::mutate(impedance_1 = impedance_binary,
+                  impedance_2 = impedance_binary)
+
+  # The level of service of each hub is the number of bicycle racks at the hub, divided by the population that they serve:
+  los <- ttm %>%
[…]

### 2021-06-15 sobiEquity.Rcheck/00_pkg_src/sobiEquity/R/sobiEquity.R

@@ -1,9 +1,14 @@
 #' sobiEquity
 #'
-#' sobiEquity is a companion package for the paper "Examining horizontal and vertical 
-#' equity to a public bicycle share program using a balanced floating catchment area 
+#' sobiEquity is a companion package for the paper "Examining horizontal and vertical
+#' equity to a public bicycle share program using a balanced floating catchment area
 #' accessibility approach".
-#'   The package includes the data used in the analysis.
+#'
+#' The package includes the data used in the analysis.
+#'
+#' Contains information licensed under the Open Government Licence – Canada (https://open.canada.ca/en/open-government-licence-canada).
+#'
+#' Contains public sector Data made available under the City of Hamilton’s Open Data Licence (https://www.hamilton.ca/city-initiatives/strategies-actions/open-data-licence-terms-and-conditions).
 #'
 #' @author Elise Desjardins, \email{desjae@@mcmaster.ca}
 #' @author Antonio Paez, \email{paezha@@mcmaster.ca}
@@ -32,7 +37,7 @@ NULL
 #' @keywords transportation
 #' @name community_downtown
 #' @usage data(community_downtown)
-#' @source Open Hamilton
+#' @source Open Hamilton (https://open.hamilton.ca/)
 "community_downtown"
 
 #' Hamilton Dissemination Areas.
@@ -69,7 +74,7 @@ NULL
 #' @keywords transportation
 #' @name sobi_service
 #' @usage data(sobi_service)
-#' @source Open Hamilton
+#' @source Open Hamilton (https://open.hamilton.ca/)
 "sobi_service"
 
 #' Hamilton Bike Share Stations
@@ -81,7 +86,7 @@ NULL
 #' @keywords transportation
 #' @name sobi_hubs
 #' @usage data(sobi_hubs)
-#' @source Open Hamilton
+#' @source Open Hamilton (https://open.hamilton.ca/)
 "sobi_hubs"
 
 #' Travel Time Matrix
@@ -94,5 +99,4 @@ NULL
 #' @keywords transportation
 #' @name ttm_walk
 #' @usage data(ttm_walk)
-#' @source BBBike
 "ttm_walk"

### 2021-06-15 sobiEquity.Rcheck/sobiEquity-Ex.R

@@ -38,6 +38,26 @@ base::assign(".ptime", proc.time(), pos = "CheckExEnv")
 
 base::assign(".dptime", (proc.time() - get(".ptime", pos = "CheckExEnv")), pos = "CheckExEnv")
 base::cat("b2sfca", base::get(".format_ptime", pos = 'CheckExEnv')(get(".dptime", pos = "CheckExEnv")), "\n", file=base::get(".ExTimings", pos = 'CheckExEnv'), append=TRUE, sep="\t")
+cleanEx()
+nameEx("c2sfca")
+### * c2sfca
+
+flush(stderr()); flush(stdout())
+
+base::assign(".ptime", proc.time(), pos = "CheckExEnv")
+### Name: c2sfca
+### Title: Compute conventional floating catchment area accessibility
+### Aliases: c2sfca
+
+### ** Examples
+
+#add(1, 1)
+#add(10, 1)
+
+
+
+base::assign(".dptime", (proc.time() - get(".ptime", pos = "CheckExEnv")), pos = "CheckExEnv")
+base::cat("c2sfca", base::get(".format_ptime", pos = 'CheckExEnv')(get(".dptime", pos = "CheckExEnv")), "\n", file=base::get(".ExTimings", pos = 'CheckExEnv'), append=TRUE, sep="\t")
 ### * <FOOTER>
 ###
 cleanEx()

### 2021-06-15 sobiEquity/R/c2sfca.R

@@ -1,0 +1,60 @@
+#' Compute conventional floating catchment area accessibility
+#'
+#' @param ttm A travel time matrix with a column of origin IDs, a column of destination IDs, travel times, type of destinations, population at the origin, and supply at the destination.
+#' @param threshold A threshold for the binary impedance function
+#' @return The level of service \code{los} and accessibility \code{accessibility}.
+#' @export
+#' @examples
+#' #add(1, 1)
+#' #add(10, 1)
+
+c2sfca <- function(ttm = ttm, threshold = threshold){
+  # calculate the impedance
+  ttm <- ttm %>%
+    dplyr::mutate(impedance_binary = ifelse(travel_time <= threshold, 1, 0))
+
+  # Calculate the sum of the impedance by population for the balancing factors. In the binary case, this should be the same as the number of SoBi hubs that each interpolated population unit can reach
+  #sum_b1 <- ttm %>%
+  #  dplyr::group_by(UID) %>%
+  #  dplyr::summarize(sum_b1 = sum(impedance_binary),
+  #            .groups = "drop")
+
+  # Calculate the sum of the impedance by SoBi hubs for the balancing factors. In the binary case, this should be the same as the number of interpolated population cells that each SoBi hub serves
+  #sum_b2 <- ttm %>%
+  #  dplyr::group_by(hub) %>%
+  #  dplyr::summarize(sum_b2 = sum(impedance_binary))
+
+  # Join the sum of the impedance to the table to calculate the balanced impedance values:
+  #ttm <- ttm %>%
+  #  dplyr::left_join(sum_b1, by = "UID") %>%
+  #  dplyr::left_join(sum_b2, by = "hub")
+
+  # **Important:** notice that some values of `impedance_binary` are zeros! This means that some population units cannot reach a SoBi hub in at most the threshold. To avoid issues we will remove any population units that reach zero hubs:
+  ttm <- ttm %>%
+    dplyr::filter(impedance_binary > 0)
+
+  # Calculate balanced impedance values:
+  ttm <- ttm %>%
+    dplyr::mutate(impedance_1 = impedance_binary,
+                  impedance_2 = impedance_binary)
+
+  # The level of service of each hub is the number of bicycle racks at the hub, divided by the population that they serve:
+  los <- ttm %>%
+    dplyr::group_by(hub) %>%
+    dplyr::summarize(los = dplyr::first(racks) / sum((population * impedance_1)),
+              .groups = "drop")
+
[…]

### 2021-06-16 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Revisions to manuscript

--- 2021-03-31 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
@@
 ---
-title: Examining spatial equity of bike share - a balanced floating catchment area approach 
+title: Examining equity in accessibility to bike share - a balanced floating catchment area approach 
 author:
@@
 
-  Public bicycle share programs (PBSPs) can play a role in advancing transportation equity if they make cycling more accessible to disadvantaged populations.
-In Ontario, the City of Hamilton expanded their PBSP in 2018 by adding twelve docking stations with the explicit objective of increasing spatial equity in access.
-In this case study, we investigate differentials in accessibility to stations using a balanced floating catchment area accessibility approach and compare accessibility with and without the equity stations.
+  Public bicycle share programs (PBSPs) can play a role in advancing transportation equity if they make bicycling more accessible to disadvantaged populations.
+In Ontario, the City of Hamilton expanded their PBSP in 2018 by adding twelve "equity" stations with the explicit objective of increasing access for under-serviced neighbourhoods.
+In this case study, we investigate differentials in accessibility to stations using a balanced floating catchment area approach and compare accessibility with and without the equity stations.
 We analyze micro zones to better reflect walking to a station and conduct a sensitivity analysis at several walking time thresholds.
 We then reaggregate the estimated accessibility for further analysis using census data.
-Our findings indicate that equity stations increased the serviced population at every threshold examined.
-Although accessibility increased for the whole population, this increase was relatively modest especially for population in the bottom 20% of median total household income.
+Our findings indicate that equity stations increased accessibility for the serviced population at every threshold examined, but the increase was relatively modest especially for population in the bottom 20% of median total household income.
 
@@
 ---
+
+```{r}
+rm(list = ls())
+```
 
@@
 
-The potential of public bicycle share programs (PBSPs) to increase cycling levels is but one of many reasons for implementing such programs in urban areas [@hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019].
-As a healthy, inexpensive, and convenient form of public transportation, shared bicycles can encourage individuals to take up cycling for short local trips or first and last mile trips to other public transportation instead of using personal vehicles.
-These programs can also play a role in addressing transportation needs and advancing transportation equity if they make cycling more accessible to individuals and groups with lower socioeconomic status.
-As these programs become increasingly common, their introduction has been accompanied by a flurry of research that investigates them from the perspective of spatial and transportation equity [e.g., @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @hullgrassoBikeShareEquity2020; @qianBikesharingEquityDisadvantaged2020; @qianBikeshareDestinationChoices2021; @smith2015exploring].
-Indeed, although these programs are available to the general public and ought to be accessible to any individual who wishes to use them, research on PBSPs located in North American and European cities indicates that inequity in spatial accessibility needs redressing.
-
-In this paper, we investigate the case of the PBSP in the city of Hamilton, in Ontario, Canada.
-Hamilton Bike Share was launched in 2015 and currently has over 900 operational bicycles and 130 docking stations.
-An interesting feature of this program is that it was expanded by an equity initiative that introduced twelve stations in 2018 with the explicit objective of increasing spatial accessibility.
-Hamilton Bike Share was already studied by Hosford and Winters [-@hosfordWhoArePublic2018] as part of a selection of PBSPs in Canada.
-The authors found that most of the cities in their research could benefit of greater efforts to expand service to areas with lower socioeconomic groups - but Hamilton fared somewhat better than other cities in this respect.
-What distinguishes our research from existing literature is the use of a balanced floating catchment area (BFCA) accessibility approach that accounts for the supply of stations and the potential demand from the population serviced.
-This requires a more disaggregated approach than the use of census geographies [q.v., @hosfordWhoArePublic2018], since walking trips to a bicycle share station likely happen at a level lower than even the smallest census geography.
-For this reason, we implement our analysis parting from micro population zones to better reflect the friction of walking to a docking station, which is an important component of a bike share trip [@chenExploringEquityPerformance2019].
-Further, we conduct a sensitivity analysis at several walking time thresholds.
-In terms of spatial equity, we compare accessibility with and without the equity stations to assess the effect of the initiative, before reaggregating the data for further analysis using median total household income information from the census.
+The potential of public bicycle share programs (PBSPs) to increase bicycling levels is but one of many reasons for implementing such programs in urban areas [see @hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019].
+As a healthy, inexpensive, and convenient form of public transportation, shared bicycles can encourage individuals to take up bicycling for short local trips or first and last mile trips to other public transportation instead of using personal vehicles.
+These programs can also play a role in advancing transportation equity if they make bicycling more accessible to disadvantaged populations.
+Although PBSPs are available to the general public in over 800 cities worldwide [@fishmanBikeshareReviewRecent2016], and ought to be accessible to any individual who wishes to use them, research on PBSPs indicates that inequities persist with respect to who can use and access them.
+
+Many PBSPs in North America now offer specific programs to address equity that primarily focus on removing cost barriers and increasing access for groups that are under-represented among existing users [@trec2019].
+Hamilton Bike Share (HBS), located in Hamilton, Ontario, was the only Canadian PBSP included in a North American scan of bike share equity programs [see @trec2019].
+HBS was launched in 2015 and currently has over 900 operational bicycles and 130 docking stations.
+An equity program, Everyone Rides Initiative, was implemented in 2018 which expanded the program by introducing twelve "equity" stations which expanded HBS to more disadvantaged neighbourhoods in the core service area.
+
+This paper examines accessibility to Hamilton Bike Share using a balanced floating catchment area (BFCA) approach.
+We also conduct a comparative analysis to the conventional two-step floating catchment area approach to highlight the benefits of this method which, to our knowledge, has not yet been used in the cycling literature.
+The paper also assesses the contribution of HBS's equity stations to reducing inequities in accessibility for different groups according to median total household income, and provides policy recommendations to further improve equity.
 
@@
 
-2. Literature Review
+2.1. Literature Review
 ===================
+
+## 1.1. Public Bicycle Share Programs
 
@@
 In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @ogilvieInequalitiesUsagePublic2012; @reillyGenderDisparitiesNew2020; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011].
-However, one study found that bike share users in Washington, DC were more likely to be female [@buckAreBikeshareUsers2013], which suggests that the gender gap among cyclists who use bike share is less disparate than the gap for personal bicycle use [@fishmanBikeshareReviewRecent2016].
+However, one study found that bike share users in Washington, DC were more likely to be female [@buckAreBikeshareUsers2013], which suggests that the gender gap among bicyclists who use PBSPs is less disparate than the gap for personal bicycle use [@fishmanBikeshareReviewRecent2016].
 There is some evidence that bike share users are less likely to own a car [@buckAreBikeshareUsers2013; @reillyNoncyclistsFrequentCyclists2020].
@@
 Stations in disadvantaged communities in Chicago have been found to generate most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighborhoods in Minneapolis-St.
-Paul used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019].
+Paul also used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019].
 Similar findings were reported in London [@ogilvieInequalitiesUsagePublic2012].
 Being university educated was a significant correlate of bike share use in Montreal, Canada [@fullerUseNewPublic2011].
-Perhaps not coincidentally, financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016].
-
-Many studies have found that geographic proximity to a bicycle share station is an important determinant of membership and use [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011].
-This makes sense given that individuals are more likely to use services or programs that they can easily reach.
-Several studies have recently explored the equity of PBSPs in North America by primarily examining who has access to bike share (e.g., differences by demographics or socioeconomic status) and where stations are located.
-It is important to note that equity can be achieved in two different ways: balanced or equal spatial distribution across a region for all similar groups (e.g., horizontal equity) or greater and targeted access to accommodate vulnerable or disadvantaged populations (e.g., vertical equity) [see @chenExploringEquityPerformance2019].
-Both are of interest to researchers and transportation planners since they are often linked in that advantage, or conversely disadvantage, has spatial patterns.
-Using a negative binomial regression model, Qian and Jaller [-@qianBikesharingEquityDisadvantaged2020] estimated ridership in Chicago's PBSP and found some disparities.
-A minority of bike share stations were found to be located in disadvantaged communities, while annual members from disadvantaged communities have a lower share of trips compared to other areas in the city.
-This suggests that they may be using PBSPs for utilitarian trips (e.g., commuting), which points to the importance of ensuring equitable access [@qianBikesharingEquityDisadvantaged2020].
+Not coincidentally, financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016].
+
+## 1.2. Equity of PBSPs
+
+The introduction of PBSPs has been accompanied by a flurry of research focusing on benefits from them.
+Many studies have examined differences in demographics or socioeconomic status between those who use or have access to PBSPs and those who don't, while other research has explored spatial inequities in where stations are located [e.g., @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @hullgrassoBikeShareEquity2020; @qianBikesharingEquityDisadvantaged2020; @qianBikeshareDestinationChoices2021; @smith2015exploring].
+In Chicago, Qian and Jaller [-@qianBikesharingEquityDisadvantaged2020] estimated ridership in the city's PBSP and found that a minority of bike share stations were located in disadvantaged communities, while annual members from such areas had a lower share of trips compared to other areas in the city.
 Similar results were found in Philadelphia.
-Despite efforts to increase equity within the city's PBSP, census block groups with lower median income generated fewer trips which suggested that efforts have not been as successful as intended [@caspiBikesharingPhiladelphiaLowerincome2019].
-Trips from stations in such areas are utilitarian (e.g., commuting to work), which points to the importance of ensuring equitable access [@caspiBikesharingPhiladelphiaLowerincome2019].
+Despite efforts to increase equity within the city's PBSP, census block groups with lower median income generated fewer trips [@caspiBikesharingPhiladelphiaLowerincome2019].
+Trips from stations in such areas were utilitarian (e.g., commuting to work), which points to the importance of ensuring equitable access [@caspiBikesharingPhiladelphiaLowerincome2019].
 In the case of Seattle, all neighborhoods had some level of access to dockless bicycles but those with higher incomes and more residents of higher education had more bikes [@mooneyFreedomStationSpatial2019].
@@
 
-On the whole, existing studies highlight the need for PBSPs to be more highly accessible for diverse populations in order to increase use beyond the "typical" users.
+Chen et al.
+[-@chenExploringEquityPerformance2019] distinguish two types of equity related to bike share, horizontal and vertical, based on the work of Delbosc and Currie [@delboscUsingLorenzCurves2011].
+Horizontal equity leads to balanced or equal distribution of accessibility and costs for all similar groups, while vertical equity would involve greater and targeted access for under-represented or disadvantaged populations [see @chenExploringEquityPerformance2019].
+Both are of interest to researchers and transportation planners since they are often linked in that advantage, or conversely disadvantage, has spatial patterns.
+Bike share equity programs, as identified by McNeil et al.
+[-@trec2019], can be considered efforts to improve vertical equity since they favour groups that have benefited less from PBSPs.
+
+On the whole, existing studies highlight the need for PBSPs to be more accessible for diverse populations in order to increase use beyond the "typical" users.
 This has been the focus of recent research [see, among others, @auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020].
 Offering more people the option of using sustainable and active transportation, particularly those who have lower socioeconomic status and might benefit the most, is a worthy policy goal for cities with PBSPs.
-A few North American cities have launched specific programs to address barriers to accessing bike share or to expand service to more deprived areas: Philadelphia, Pennsylvania [@caspiBikesharingPhiladelphiaLowerincome2019], and Hamilton, Ontario [@hosfordWhoArePublic2018].
 However, exploring transportation equity by investigating where bicycle share stations are located, often using neighborhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
@@
 
+## 2.3. Accessibility Approaches
+
 Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved over the past century and has increasingly become linked with recent planning interests in prioritizing modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020].
-As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequality in the provision of infrastructure, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options.
-It also addresses some of the challenges of other performance measures such as level of service within a transportation network by measuring person-based indicators and exploring differences in use between population groups [@paezMeasuringAccessibilityPositive2012].
-
-Beyond the utility derived from using shared bikes to destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019].
-Since the time or distance needed to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., number of bicycle racks) of stations matters.
-Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
+Beyond the utility derived from using shared bicycles to access destinations of value, docking stations themselves are amenities because they offer a transportation service.
+Therefore, the ease of reaching these stations, which are spread spatially in a given area, can affect use of the PBSP.
+
+The location and size of docking stations are important to increase the utility of PBSPs for more people.
+These two factors are relevant to accessibility.
+Since the time or distance needed to reach a docking station decreases the potential of accessing the program, the location of stations matters.
+Indeed, distance to stations is associated with bike share use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
 Kabra et al.
-[-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300 m of stations, which amounts to 2-4 minutes walking by an able-bodied adult.
-Furthermore, similar to other public amenities affected by crowding, like health care services [e.g., @pereiraGeographicAccessCOVID192021], the utility of stations is limited by the _maximum_ number of bicycles that they can hold.
-The program may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles.
+[-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300 m of stations, which amounts to 2-4 minutes walking by an adult who does not have a disability.
+Similar to other public amenities affected by crowding, the utility of stations is also limited by the _maximum_ number of bicycles that they can hold (e.g., their size).
+The program may not necessarily be accessed if stations are easy to reach but offer only a small number of bicycles.
 Likewise, more people may not opt to use the program if the supply of bicycles available at nearby stations is insufficient to meet demand.
-The location and size of stations is important to increase the utility of this public transportation option for more people, thus achieving vertical equity.
 Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts.
@@
 Several approaches have been commonly used for measuring place-based accessibility, including cumulative opportunities, gravity, and utility-based measures [@handyMeasuringAccessibilityExploration1997].
-Geurs and Van Wee [-@geursAccessibilityEvaluationLanduse2004] and Paez et al.
-[-@paezMeasuringAccessibilityPositive2012] provide recent overviews of various formulations and applications of accessibility in transportation research.
-The common gravity-based approach for example involves weighting destination opportunities, such as the quantity of bicycle share stations, by the time required to reach them from an origin using an impedance function [@handyMeasuringAccessibilityExploration1997; @kwanSpaceTimeIntegral1998].
-However, while such measures are suitable for capturing the potential for reaching destinations from a given location, they do not take demand or congestion effects into account.
-
-In contrast, floating catchment area (FCA) methods incorporate information on capacity and demand in calculating accessibility.
-FCA measures have been widely employed in healthcare accessibility research and are typically calculated across two steps.
-In the first, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hospital, weighted by the distance involved in reaching the facility.
+The gravity-based approach involves weighting destination opportunities, such as the quantity of bicycle share stations, by the time required to reach them from an origin using an impedance function [@handyMeasuringAccessibilityExploration1997; @kwanSpaceTimeIntegral1998].
+While such measures are suitable for capturing the potential for reaching destinations from a given location, they do not take demand or congestion effects into account which is an important consideration when calculating accessibility for amenities such as bike share stations.
+
+In contrast, floating catchment area (FCA) methods have been widely employed in health care accessibility research.
+The benefit of this approach is that the method incorporates information on capacity, demand, and congestion in calculating accessibility.
+This approach is more appropriate and informative than calculating provider-to-population ratios (PPR) that simply divide the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019].
+In particular, the Two-Step Floating Catchment Area (2SFCA) method [@luoMeasuresSpatialAccessibility2003; @radkeSpatialDecompositionsModeling2000] produces flexible catchment areas instead of using rigid boundaries like PPR.
+In the first step, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hospital, weighted by the distance involved in reaching the facility.
 Next, these service level ratios are allocated back to the population centers and summarized as a measure of congested accessibility.
 Thus, this model does a good job of considering potential crowding or competition for services.
-While there have been many methodological innovations in FCA methods [for example, @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balancing of the impedance that addressed the effects of demand and service inflation found in earlier FCA approaches [see @paezDemandLevelService2019].
-
-When measuring accessibility, researchers have taken different approaches when it comes to aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones.
-Previous research on bike share equity has typically used a meso- or macro-level approach with aggregated data from entire neighborhoods or census tracts [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021].
-This is also true for studies examining correlates of bike share demand [@wangNewBikeShare2019].
-Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals, which is useful for addressing vertical inequity in PBSP usage.
+However, overlapping catchment areas from the conventional FCA approach lead to inflation of population totals and deflation of service levels across a study area which generates inaccurate or misleading accessibility estimates [@paezDemandLevelService2019].
+While there have been many methodological innovations in FCA methods [e.g., @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balancing of the impedance that addressed the effects of demand and service inflation found in earlier FCA approaches [see @paezDemandLevelService2019].
+This provides more useful information because it does not assume that people are limited to service within pre-defined boundaries [@paezDemandLevelService2019].
+
+When measuring accessibility, researchers have also taken different approaches when it comes to the aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones.
+Previous research on bicycle share equity has typically used a meso- or macro-level approach with aggregated data from entire neighborhoods or census tracts [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021].
+This is also true for studies examining correlates of bicycle share demand [@wangNewBikeShare2019].
+Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals.
 Chen et al.
@@
 
-Using the balanced floating catchment area method [@paezDemandLevelService2019], a novel approach that has not been used yet in cycling research, with disaggregated population data, we examine accessibility to a public bicycle share program in Hamilton, Ontario.
-In this paper, we (1) conduct a sensitivity analysis by measuring accessibility and level of service to bicycle share stations at different walking time thresholds to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median total household income of dissemination areas within the core service area.
+## 2.4 Previous Research
+
+Our analysis builds upon a previous study [@hosfordWhoArePublic2018], which found that areas in Hamilton with less advantage are better served by the city's PBSP compared to other Canadian cities (i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau) where areas that are less deprived have greater access.
+Hosford and Winters [-@hosfordWhoArePublic2018, pp.
+] acknowledge that "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas".
+Therefore, the core service area for the PBSP in Hamilton by default covers more of the less advantaged areas in the city.
+However, there is also a great deal of variation in income in the city center because of the local university and increasing gentrification.
+Hosford and Winters [-@hosfordWhoArePublic2018] took a macro-level approach in their analysis by using dissemination areas _across_ the city as the unit of analysis.
+They did not focus specifically on the core service area and did not differentiate between conventional and equity stations.
+
+Therefore, this paper contributes to the cycling literature by expanding the analysis of Hosford and Winters [-@hosfordWhoArePublic2018] to assess equity in accessibility to HBS and explore the contribution of the "equity" stations.
 
@@
 
-The focus of this study is the city of Hamilton, located in Ontario, Canada.
-The city launched a public bicycle share program in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015].
-Before June 2020, the program was known as Social Bicycles or SoBi Hamilton, but is now called Hamilton Bike Share.
-Stations are spaced between 300 and 600 m apart [@scottRouteChoiceBike2021].
-The core service area spans 40 sq.km of the city although it was planned to be 20 sq.km [@hamiltonsobi2015], and roughly 138,000 people are within 30 minutes walking of a bike share station [see Figure \ref{fig:hamilton-and-sobi-service-area}].
+The case study of this paper is the city of Hamilton, Ontario, Canada.
+Before June 2020, the program was known as Social Bicycles or SoBi Hamilton, but is now called Hamilton Bike Share (HBS).
+The core service area spans 40 sq.km of the city, although it was planned to be 20 sq.km [@hamiltonsobi2015], and roughly 138,000 people are within 30 minutes walking of a station [see Figure \ref{fig:hamilton-and-sobi-service-area}].
 This represents roughly one fifth of the total population of the Hamilton Census Metropolitan Area according to the 2016 Canadian Census.
-The City of Hamilton undertook a large public engagement campaign to validate the locations of bicycle share stations that had been selected and to crowdsource potential locations for additional locations [@hamiltonsobi2014].
+The City of Hamilton undertook a large public engagement campaign to validate the locations of stations that had been selected and to crowdsource potential locations for additional locations [@hamiltonsobi2014].
 Most of the suggested locations were in the east end of the core service area that lack stations or in neighborhoods not serviced by the PBSP.
 The program was enthusiastically welcomed in the city in 2015 - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015], however inadequate coverage given the size of Hamilton Bike Share's service area was identified as a problem early on, and transportation planners noted that the small size (i.e., supply of bicycle racks) and low quantity of stations would lead to challenges in balancing the system [@hamiltonsobi2015].
-Based on public feedback, thirteen stations were added to the original system.
 
@@
 
-In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), with the objective of reducing barriers that may prevent individuals from accessing bike share in Hamilton.
-Additional bicycles and stations were added to the program which expanded it to more disadvantaged areas in the core service area.
-The program also offers subsidized memberships to individuals who identify as low income, and complements this service with cycle skills education.
+Hamilton Bike Share's equity program, Everyone Rides Initiative (ERI), was launched in 2018 with the objective of reducing barriers that may prevent individuals from accessing the program.
+Additional bicycles and twelve "equity" stations were added to the core service area.
+The equity program also offers subsidized memberships to individuals who identify as low income, and complements this service with cycle skills education.
 A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019].
@@
 
-As of June 2020, Hamilton Bike Share has 900 bikes, 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
+As of June 2020, HBS has 900 bikes, 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
 The core service area remains 40 sq.km.
-The PBSP has twelve stations that were added as part of the ERI initiative; we refer to these as "equity stations" throughout the paper while all other stations are "conventional stations".
-In total, 118 stations are "conventional" and 12 are "equity".
-
-## 3.4. Membership
-
-Hamilton Bike Share conducted one membership survey in 2018 [@civicplanSoBiHamiltonMembership2017] and the findings from 420 members are broadly in line with the trends that were discussed above [see @fishmanBikeshareReviewRecent2016 for a recent review of the literature].
-The majority of respondents live within the core service area and the gender split is expected: 57% of respondents are male and 41% are female.
-The majority of respondents, both male and female, are between 25 and 34 years of age, but the percentage of male respondents is higher in the subsequent age groups.
-Respondents use bike share for commuting (40% of trips) or errands and meetings (24% of trips) and nearly 50% of trips have an average length of 11 to 20 minutes.
-As a result of having a bike share membership, 49% of respondents report that they use their private vehicle less often or much less often but 48% report that their private vehicle use has remained about the same.
-This suggests that Hamilton Bike Share has been useful for certain kinds of trips but not all, meaning that some trips continue to require a personal vehicle.
-
-## 3.5. Relevant Research
-
-Our analysis builds upon a previous and recent study [@hosfordWhoArePublic2018], which found that areas in Hamilton with less advantage are better served by the city's PBSP compared to other Canadian cities [i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau] where areas that are less deprived have greater access.
-Hosford and Winters [-@hosfordWhoArePublic2018] acknowledge that "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas".
-Therefore, the core service area for the PBSP in Hamilton by default covers some of the less advantaged areas in the city.
-However, there is also a great deal of variation in income in the city center because of the local university and increasing gentrification.
-Hosford and Winters [-@hosfordWhoArePublic2018] took a macro-level and system-wide approach in their analysis by using dissemination areas _across_ the city as the unit of analysis.
-They did not focus specifically on the core service area and did not differentiate between conventional and equity stations.
+The program has twelve "equity" stations and 118 "conventional" stations (i.e., that were not explicitly added to address inequities in the program).
+The City of Hamilton has positioned stations between 300 and 600 m apart [@scottRouteChoiceBike2021], but anticipates that they will service those living within a 250 m buffer [@hamiltonsobi2015].
+The latter constitutes a normative statement: people ought to be able to access a station in less than 600 m if they live in the core service area, with most usage coming from 250 m around.
+However, it is not known how far people are actually willing to travel to reach a station.
+It would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.
 
@@
 
-## 4.1. Floating Catchment Area
-
-Floating catchment area (FCA) methods are an approach commonly used in the healthcare accessibility literature.
-This approach is more appropriate and informative than calculating provider-to-population ratios (PPR) that simply divide the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019].
-In particular, the Two-Step Floating Catchment Area (2SFCA) method [@luoMeasuresSpatialAccessibility2003; @radkeSpatialDecompositionsModeling2000] produces flexible catchment areas instead of using rigid boundaries like PPR.
-This provides more useful information because it does not assume that people are limited to service within pre-defined boundaries [@paezDemandLevelService2019].
-This is an important property that supports our rationale for applying this method to measure accessibility to Hamilton Bike Share.
-The City of Hamilton has positioned stations between 300 and 600 m apart, but anticipates that hubs will service those living within a 250 m buffer from the station [@hamiltonsobi2015].
-The latter constitutes a normative statement: people ought to be able to access a station in less than 600 m if they live in the core service area, with usage coming from 250 m around.
-However, it is not known how far people are actually willing to travel to reach a station.
-It would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.
-
-More recently, the _balanced_ floating catchment area approach was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by earlier FCA methods [@paezDemandLevelService2019].
-Briefly, overlapping catchment areas lead to inflation of population totals and deflation of service levels across a study area and generates inaccurate or misleading accessibility estimates.
-In contrast, Paez et al.
+## 4.1. Balanced Floating Catchment Area
+
+The BFCA method was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by earlier FCA methods [@paezDemandLevelService2019].
+Paez et al.
 [-@paezDemandLevelService2019] adjusted the impedance weights so that both supply and demand are proportionally allocated.
 The result is a FCA method that balances the population and level of service by eliminating the over-counting of population and level of service that leads to distortions in demand and supply.
-Other benefits of this adjusted method include consideration of competition which can occur when catchment areas overlap, as well as the preservation of population and level of service.
-Balanced floating catchment area methods have been used to explore accessibility to health care providers [@paezDemandLevelService2019] and COVID-19 health care services [@pereiraGeographicAccessCOVID192021], but have not yet been used in the cycling literature to explore issues of accessibility.
-
-The first step in the floating catchment area method is to allocate the population to be serviced by each Hamilton Bike Share station:
-$$
+
+The first step in the FCA method is to allocate the population to be serviced by each docking station:
+\begin{equation}
+\label{eq:population-allocation}
 P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}
-$$
+\end{equation}
 
@@
 Next, the supply at each station (i.e., the maximum number of bicycle racks) is divided by its estimated service population within the established catchment area; this gives the level of service of station $j$ in bicycle racks per person:
-$$
+\begin{equation}
+\label{eq:level-of-service}
 L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
-$$
+\end{equation}
 
 Finally, the accessibility of population cell $i$ is calculated as the weighted sum of the level of service of all stations that can be reached from there according to the spatial weights:
-$$
+\begin{equation}
+\label{eq:FCA-accessibility}
 A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
-$$
+\end{equation}
 
 The balanced approach of Paez et al [-@paezDemandLevelService2019] replaces the spatial weights with normalized versions as follows:
-$$
+\begin{equation}
+\label{eq:spatial-weights}
 {w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
-$$
+\end{equation}
 
 \noindent and:
-$$
+\begin{equation}
+\label{eq:spatial-weights-2}
 {w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
-$$
+\end{equation}
 
 These weights satisfy the following properties:
-$$
+\begin{equation}
+\label{eq:weights}
 \sum_{j = 1}^{J} {w^i_{ji}} = 1
-$$
+\end{equation}
 
 \noindent and:
-$$
+\begin{equation}
+\label{eq:weights-2}
 \sum_{i = 1}^{n} {w^j_{ji}} = 1
-$$
+\end{equation}
 
 With these weights, accessibility can be calculated without risk of demand or supply inflation:
-$$
+\begin{equation}
+\label{eq:balanced-accessibility}
 A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}
-$$
+\end{equation}
 
 By allocating the population and level of service proportionally, this method preserves the values of the population and level of service and provides a regional provider-to-population ratio since:
-$$
+\begin{equation}
+\label{eq:proportionality}
 \sum_{i=1}^n A_i = \sum_{j=1}^J L_j = \frac{\sum_{j=1}^JS_j}{\sum_{i=1}^n P_i}
-$$
+\end{equation}
 
 In fact, since the proportional allocation procedure means that any proportion of the population allocated to a station is never allocated to other stations, and conversely any level of service allocated to a population is never re-allocated elsewhere, this property is replicated for any level of aggregation.
-For this paper, we employ a hybrid location-based and person-based approach to calculating accessibility using disaggregated population data.
-In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data.
-This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are better able of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences.
-Applying the balanced floating catchment area approach allows us to examine accessibility by stratifying according to median total household income, which constitutes the individual component of the accessibility measure [@geursAccessibilityEvaluationLanduse2004].
-However, conducting a further sensitivity analysis to measure accessibility at different walking time thresholds would help to consider potential spatio-temporal constraints.
-Different people may be willing to travel different distances to access a public bicycle share station.
 
@@
 
-To obtain population at sub-census geography levels (at the micro scale), we use pycnophylactic interpolation [@toblerSmoothPycnophylacticInterpolation1979].
+Since walking trips to a bicycle share station likely happen at a level lower than even the smallest census geography, this requires a more disaggregated approach than the use of census geographies.
+For this reason, we implemented our analysis parting from micro population zones to better reflect the friction of walking to a docking station, which is an important component of a bike share trip [@chenExploringEquityPerformance2019].
+
+To obtain population at sub-census geography levels, we used pycnophylactic interpolation [@toblerSmoothPycnophylacticInterpolation1979].
 We obtained population data from the 2016 Census of Canada for dissemination areas (DA), which is the smallest publicly available census geography in Canada.
-These zonal values of the population were interpolated to smaller polygons 50-by-50 m in size.
-Pycnophylactic interpolation involves smoothing out the population from each dissemination area while preserving total volume [see Figure \ref{fig:da-population}].
-When interpolating the population at this high level of resolution, it is important to ensure that population numbers were not allocated to areas where people do not live in Hamilton (for example, to parks, large institutional buildings, etc.).
+These zonal values of the population were interpolated to smaller polygons that are 50-by-50 m in size.
+Pycnophylactic interpolation involves smoothing out the population from each DA while preserving total volume [see Figure \ref{fig:da-population}].
+When interpolating the population at this high level of resolution, it is important to ensure that population numbers were not allocated to areas where people do not live in Hamilton (for example, to parks or large institutional buildings, etc.).
 To do so, we retrieved shapefiles for various geographic features [see Table \ref{tab:data-features}] from Open Hamilton.
 Next, we removed these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population within the remaining area [see Figure \ref{fig:interpolated-population}].
-
-```{r data-features, echo=FALSE}
-data_tbl <- data.frame(
-  Feature = c("Hamilton Bike Share Stations", "Golf Courses", "Parks", "Designated Large Employment Areas", "Municipal Parking Lots", "Cemeteries", "Environmentally Sensitive Areas", "Streets", "Educational Institutions", "Places of Worship", "Municipal Service Centers", "Recreation and Community Centers", "Arenas", "Emergency Stations", "Fire Stations", "Police Stations", "Railways", "Hospitals"),
-  Description = c(
-    "The location of stations and the number of racks available at each station.",
-    "The location of City and privately owned golf courses.", 
-    "The location of parks and other green spaces.",
-    "The location of large business parks and industrial lands.",
-    "The location of municipal car parks.",
-    "The location of cemeteries.",
-    "The location of either land or water areas containing natural features or significant ecological functions.",
-    "The street network in Hamilton, including road classification for highways.",
-    "The location of all educational institutions and schools.",
-    "The location of buildings used for religious congregations.",
-    "The location of all municipal service centers, including City Hall.",
-    "The location of all recreation and community centers.",
-    "The location of all indoor arenas.",
-    "The location of all Emergency Management Services (EMS) Ambulance stations.",
-    "The location of all fire stations.",
-    "The location of all police stations.",
-    "The railway network in Hamilton.",
-    "The location of all hospitals."
-  )
-)
-
-kable(data_tbl,
-      "latex",
-      booktabs = TRUE,
-      caption = "\\label{tab:landscape-features}Landscape features extracted from the Hamilton Bike Share core service area before population is interpolated.") %>%
-  kable_paper(full_width = F) %>%
-  column_spec(1, bold = T, border_right = T) %>%
-  column_spec(2, width = "30em") %>%
-  kable_styling(latex_options = c("striped", "scale_down"))
-```
 
@@
 
-To calculate walking times from the centroid of our micro population cells, we extracted OpenStreetMap data for Hamilton Bike Share's service area from [BBBike](https://download.bbbike.org/osm/bbbike/), an online cycle route planner that interfaces with OpenStreetMap.
+To calculate walking times from the centroid of the micro population cells to docking stations, we extracted OpenStreetMap data for Hamilton Bike Share's service area from [BBBike](https://download.bbbike.org/osm/bbbike/), an online cycle route planner that interfaces with OpenStreetMap.
 OpenStreetMap data provides the networks for calculating walking times from each population cell to nearby bicycle share stations, using a maximum walking distance of 10 km and walking time of 30 minutes as thresholds.
 A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bicycle share stations within the maximum threshold.
-This process provides a more realistic measure of the friction of reaching stations by taking infrastructure into account of travel times, rather than using the Euclidean distance from population cell to station.
+This process provides a more realistic measure of the friction of reaching stations by taking infrastructure into account network travel times, rather than using the Euclidean distance from population cell to station.
 Routing and travel time calculations were completed using the `R` package `r5r`, used for rapid realistic routing operations [@Pereira2021r5r].
@@
 
-Consensus regarding the distance that individuals are willing to walk to access a bicycle share station is lacking, but the literature on walking behaviour can provide some guidelines to determine the thresholds for sensitivity analysis.
-Previous studies have found that living within 250 m [@fullerUseNewPublic2011] and 300 m [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 m and rarely more than 1200 m [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 m [@larsenQuarterMileReexamining2010].
-In Hamilton, Hamilton Bike Share will depict a map at some stations to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are expected to walk.
+Consensus regarding the distance that individuals are willing to walk to access a bike share station is lacking, but the literature on walking behaviour can provide some guidelines to determine the thresholds for our sensitivity analysis.
+Previous studies have found that living within 250 m [@fullerUseNewPublic2011] and 300 m [@kabraBikeShareSystemsAccessibility2020] of a docking station is correlated with bicycle share use, while other research has found that walking trips are less than 600 m and rarely more than 1200 m [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 m [@larsenQuarterMileReexamining2010].
+Hamilton Bike Share will often depict a map at some docking stations to show the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are expected to walk.
 National Association of City Transportation Officials (NACTO) has a similar normative guide [@nactowalkingstation2015].
 
-In the present case, we found that congested accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at 5 minutes.
+In the present case, we found that accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at 5 minutes.
 Accessibility decreases substantially after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each station.
 
-For this reason, we experiment with various thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times from population cell to bicycle share station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
-We categorize these thresholds as minimum, average, maximum, and extreme, respectively.
-At each threshold, we compare accessibility between the current system and the original system to examine the contribution of the additional equity stations.
+For this reason, we experimented with various thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times from population cell to bike share stations: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
+We categorized these thresholds as minimum, average, maximum, and extreme, respectively.
+At each threshold, we compared accessibility between the current system and the original system to examine the contribution of the twelve equity stations.
 When considering the results reported below, it is important to remember that accessibility is technically a form of smoothing [@okelly2003aggregate, pp.
@@
 ```
+
 ```{r 5-min-conventional, include=FALSE}
@@
 
-With a walking distance of three minutes, we find that there are 25.2 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+With a walking distance of three minutes, we found that there are 25.2 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
 The addition of equity stations increases this ratio slightly to 25.4 bicycles per person.
@@
 
-With a walking distance of five minutes, we find that there are 68.6 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+With a walking distance of five minutes, we found that there are 68.6 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
 With the addition of equity stations, there are now 68.8 bicycle racks per person.
@@
 
-With a walking distance of ten minutes, we find that there are 3.61 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+With a walking distance of ten minutes, we found that there are 3.61 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
 With the addition of equity stations, there are now 3.74 bicycles per person.
@@
 
-With a walking distance of fifteen minutes, we find that there are 2.44 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+With a walking distance of fifteen minutes, we found that there are 2.44 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
 With the addition of equity stations, there are now 2.55 bicycles per person.
@@
 
-## 5.2. Accessibility by Median Total Household Income
-
-A unique property of the balanced floating catchment area method is that data can be reaggregated while preserving the total population in the service area and the supply at each station.
+## 5.2. Comparative Analysis
+
+To highlight the benefit of using the BFCA approach, we conducted a comparative analysis with the 2SFCA method.
+The latter method does not normalize the impedance to allocate population and level of service proportionally.
+
+```{r conventional-3-min-non-equity, include=FALSE}
+conventional_3min_non_equity <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  c2sfca(threshold = 3)
+```
+
+```{r conventional-3-min-all, include=FALSE}
+conventional_3min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  c2sfca(threshold = 3)
+```
+
+```{r conventional-3-min-sum-non-equity}
+sum(conventional_3min_non_equity$los$los)
+sum(conventional_3min_non_equity$accessibility$accessibility)
+```
+
+```{r conventional-3-min-sum-all}
+sum(conventional_3min_non_equity$los$los)
+sum(conventional_3min_non_equity$accessibility$accessibility)
+```
+
+```{r conventional-5-min-non-equity, include=FALSE}
+conventional_5min_non_equity <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  c2sfca(threshold = 5)
+```
+
+```{r conventional-5-min-all, include=FALSE}
+conventional_5min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  c2sfca(threshold = 5)
+```
+
+```{r conventional-5-min-sum-non-equity}
+sum(conventional_5min_non_equity$los$los)
+sum(conventional_5min_non_equity$accessibility$accessibility)
+```
+
+```{r conventional-5-min-sum-all}
+sum(conventional_5min_non_equity$los$los)
+sum(conventional_5min_non_equity$accessibility$accessibility)
+```
+
+```{r conventional-10-min-non-equity, include=FALSE}
+conventional_10min_non_equity <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  c2sfca(threshold = 10)
+```
+
+```{r conventional-10-min-all, include=FALSE}
+conventional_10min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  c2sfca(threshold = 10)
+```
+
+```{r conventional-10-min-sum-non-equity}
+sum(conventional_10min_non_equity$los$los)
+sum(conventional_10min_non_equity$accessibility$accessibility)
+```
+
+```{r conventional-10-min-sum-all}
+sum(conventional_10min_non_equity$los$los)
+sum(conventional_10min_non_equity$accessibility$accessibility)
+```
+
+```{r conventional-15-min-non-equity, include=FALSE}
+conventional_15min_non_equity <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  c2sfca(threshold = 15)
+```
+
+```{r conventional-15-min-all, include=FALSE}
+conventional_15min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  c2sfca(threshold = 15)
+```
+
+```{r conventional-15-min-sum-non-equity}
+sum(conventional_15min_non_equity$los$los)
+sum(conventional_15min_non_equity$accessibility$accessibility)
+```
+
+```{r conventional-15-min-sum-all}
+sum(conventional_15min_non_equity$los$los)
+sum(conventional_15min_non_equity$accessibility$accessibility)
+```
+
+## 5.3. Accessibility by Median Total Household Income
+
+A unique property of the balanced FCA method is that data can be reaggregated while preserving the total population in the core service area and the supply at each docking station.
 This avoids demand and supply inflation, and also enables us to present findings in a way that is perhaps more intuitive to interpret.
-Therefore, we reaggregate population and accessibility from the 50-by-50 m population cells to the dissemination area.
-We then can compare accessibility to median total household income statistics from the Canadian census results from 2016.
+Therefore, we reaggregated population and accessibility from the 50-by-50 m population cells to the dissemination area.
+We then can compared accessibility to median total household income statistics from the 2016 Canadian Census.
 
@@
 As expected, the analysis indicates that the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service area of stations.
-We find that stations added to Hamilton's public bicycle share program did indeed expand the spatial coverage of the system, leading to more balance across the service area.
+We found that stations added to HBS did indeed expand the spatial coverage of the system, leading to more balance across the core service area.
 By implementing equity stations in areas with lower median total household income in Hamilton, the PBSP attained greater horizontal equity by extending the spatial distribution of bicycles across the city.
@@
 
-To further explore this issue and examine vertical equity, we aggregate accessibility by median total household income across all geographies (see Table \ref{tab:accessibility-income}).
-Similar to Hosford and Winters  [-@hosfordWhoArePublic2018], we find that Hamilton's PBSP already serviced a large proportion of the population in the bottom and second quintiles of median total household income.
+To further explore this issue and examine vertical equity, we aggregated accessibility by median total household income across all geographies (see Table \ref{tab:accessibility-income}).
+Similar to Hosford and Winters  [-@hosfordWhoArePublic2018], we found that Hamilton's PBSP already serviced a large proportion of the population in the bottom and second quintiles of median total household income.
 This may well be an artifact of the spatial socioeconomic and demographic profile of Hamilton, where the most dense parts of the city (where a PBSP is most easily launched) are also those with relatively lower incomes.
@@
 
-Using disaggregated data in this paper, we examined spatial equity and accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
+It does not have the property of preserving these quantities. This produces a There is an inflation/deflection effect
+
+Using disaggregated data, we examined spatial equity and accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
 Use of a balanced floating catchment area approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro scale which is a sensible approach to avoid the "absorption of disparities", as articulated by Chen et al.
@@
 This paper did not examine or compare ridership data between conventional and equity stations.
-Therefore, further research is needed to determine whether the addition of equity stations encouraged more cycling for low-income individuals living near them.
-Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighborhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of publicly available route and individual user data to conduct similar analyses for Hamilton Bike Share.
+Therefore, further research is needed to determine whether the addition of equity stations encouraged more bicycling for disadvantaged groups living near them.
+Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighborhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of publicly available route and individual user data to conduct similar analyses for HBS.
 
 An additional limitation is the lack of publicly available information about the number of bicycles at each station.
-Hamilton Bike share works to balance the number of bicycles across stations, but it is reasonable to expect that the number of bicycles will not match exactly the number of racks at every station.
+HBS works to balance the number of bicycles across stations, but it is reasonable to expect that the number of bicycles will not match exactly the number of racks at every station.
 Ideally, instead of number of bicycle racks as our measure of supply, we would have liked to use the average number of bicycles at stations, perhaps at different times during the day or different seasons.
@@
 
-The addition of specific equity stations to the public bicycle share program in Hamilton, Ontario had the net effect of increasing accessibility and reducing to some extent both horizontal and vertical inequities.
+The addition of specific stations to improve equity within the PBSP in Hamilton had the net effect of increasing accessibility and reducing to some extent both horizontal and vertical inequities.
 In particular, accessibility improved the most for those in the second 20% median total household income at all thresholds, but the gains were only modest for all income groups.

### 2021-06-16 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Revisions with Elise

@@
 library(gdistance)
+library(ggmap)
+library(ggspatial)
 library(gridExtra)
@@
 It would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.
-
-```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "The core service area of Hamilton Bike Share is outlined in black.
+```{r}
+
+```
+
+```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "The core service area of Hamilton Bike Share is outlined in blue.
 Hamilton Census Metropolitan Area is shown in grey."}
-ggplot() + 
+geo_context <- ggplot() + 
   geom_sf(data = hamilton_da_2016,
@@
           fill = NA,
-          colour = "black") + 
-  theme(panel.background = element_rect(fill = "white"))
+          colour = "blue") + 
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank()) +
+  #annotate("text", x = -80, y = 43.5, label = "Lake Ontario") +
+  annotation_scale(location = "bl", width_hint = 0.4) +
+  annotation_north_arrow(location = "tl", which_north = "true", 
+        pad_x = unit(0.25, "in"), pad_y = unit(0.25, "in"),
+        style = north_arrow_fancy_orienteering)
+
+# Inset map
+geo_context_inset <- get_stamenmap(bbox = c(left = -79,
+                                    bottom = 43,
+                                    right = -81,
+                                    top = 44),
+          maptype = "toner-lite", 
+          crop = FALSE,
+          zoom = 8)
+# plot map
+geo_context_inset <- ggmap(geo_context_inset)  +
+    coord_sf(crs = st_crs(4326))  +
+    #xlab("Longitude") + ylab("Latitude") +
+  annotation_scale(location = "bl", width_hint = 0.4) +
+  theme_minimal() +
+  theme(axis.text = element_blank(),
+        panel.border = element_rect(color = "black",
+                                    fill = NA),
+        axis.title = element_blank())
+
+#geo_context + geo_context_inset
+ggdraw() + 
+  draw_plot(geo_context, 0.0, 0.0, 0.8, 1) + 
+  draw_plot(geo_context_inset, 0.5, 0.4, 0.6, 0.6) +
+  draw_plot_label(x = c(.25,.5), y = c(0.45,.6), 
+                  label =  c("Hamilton Bikeshare \nCore Service Area",  "Lake Ontario"),
+                  hjust = 0, 
+                  size = 8) +
+  draw_line(x = c(0.34, 0.4),
+    y = c(0.43, 0.5),
+    color = "blue", 
+    size = 1,
+    lineend = "round",
+    linejoin = "mitre",
+    arrow = arrow(length = unit(0.1, "inches")))
 ```
@@
   geom_sf(data = sobi_hubs %>% 
-            dplyr::filter(hub_type == "Conventional" & 
-                            hub_status == "Active"), # Remove deactivated hubs
-          aes(size = RACKS_AMOU),
-          shape = 21,
-          fill = "blue") +
-  geom_sf(data =  sobi_hubs %>% 
-            dplyr::filter(hub_type == "ERI"), 
-          aes(size = RACKS_AMOU), 
-          shape = 21,
-          fill = "green") +
-  scale_size(name = "Number of racks", range = c(0.5, 4)) +
-  theme(panel.background = element_rect(fill = "white"))
+            dplyr::filter(hub_status == "Active"), # Remove deactivated hubs
+          aes(size = RACKS_AMOU, fill = hub_type),
+          shape = 21) +
+  scale_size(name = "Number of racks", 
+             range = c(0.5, 3.5)) +
+  scale_fill_manual(name = "Hub Type",
+                    values = c("Conventional" = "blue", "ERI" = "green")) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        legend.position = "bottom",
+        axis.ticks = element_blank()) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "tl", which_north = "true", 
+        pad_x = unit(0.25, "in"), pad_y = unit(0.20, "in"),
+        style = north_arrow_fancy_orienteering)
 ```
@@
 The first step in the FCA method is to allocate the population to be serviced by each docking station:
+
 \begin{equation}
@@
 Next, we removed these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population within the remaining area [see Figure \ref{fig:interpolated-population}].
-
-```{r da-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Population in all dissemination areas (outlined in white) that are inside or touch the bounding box of the Hamilton Bike Share core service area (outlined in black)."}
-ggplot() +
-  geom_sf(data = hamilton_da_2016 %>%
-            dplyr::filter(sobi_service == "Yes"),
-          aes(fill = population),
-          color = "white") +
-  geom_sf(data = sobi_service,
-          fill = NA) +
-  scale_fill_distiller(palette = "YlOrRd", 
-                       direction = 1) +
-  theme(panel.background = element_rect(fill = "white"))
-```
 
@@
           color = "dark green") +
-  theme(panel.background = element_rect(fill = "white"))
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.title = element_blank(),
+        axis.ticks = element_blank()) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "tl", which_north = "true", 
+        pad_x = unit(0.25, "in"), pad_y = unit(0.20, "in"),
+        style = north_arrow_fancy_orienteering)
 ```
@@
 The addition of equity stations increases this ratio slightly to 25.4 bicycles per person.
-Figure \ref{fig:figure-6} presents a comparison of accessibility between the systems for the population cells.
 Accessibility is fairly uniform overall, with the exception of two small areas where accessibility is slightly higher.
 This high level of system-wide accessibility occurs because the population that can reach bicycle share stations when travel time is 3 minutes or less is very limited, and accessibility is strongly shaped by a few locations that concentrate population and stations.
-
-```{r figure-6, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
+For this reason, accessibility is highly concentrated in small areas around the stations.
+The map is not shown but can be recreated using the source code and data.
+
+```{r figure-6, eval = FALSE, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
 #accessibility plot for current system with equity stations at 3 minutes walk
@@
         axis.title.x=element_blank(),
-        axis.title.y=element_blank())
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
+
 ```
@@
             aes(x = X, y = Y, fill = accessibility)) +
-  scale_fill_distiller(palette = "PuRd", direction = 1) +
+  scale_fill_distiller(palette = "PuRd", 
+                       direction = 1,
+                       trans = "log",
+                       breaks = c(0.001, 0.02, 0.4, 8)) +
   geom_sf(data = sobi_service,
@@
         axis.title.x=element_blank(),
-        axis.title.y=element_blank())
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
 ```
@@
             aes(x = X, y = Y, fill = accessibility)) +
-  scale_fill_distiller(palette = "PuRd", direction = 1) +
+  scale_fill_distiller(palette = "PuRd",
+                       direction = 1,
+                       trans = "log",
+                       breaks = c(0.00004, 0.0003, 0.002)) +
   geom_sf(data = sobi_service,
@@
             aes(x = X, y = Y, fill = accessibility)) +
-  scale_fill_distiller(palette = "PuRd", direction = 1) +
+  scale_fill_distiller(palette = "PuRd", 
+                       direction = 1,
+                       trans = "log",
+                       breaks = c(0.000006, 0.00005, 0.0003)) +
   geom_sf(data = sobi_service,
@@
 To highlight the benefit of using the BFCA approach, we conducted a comparative analysis with the 2SFCA method.
-The latter method does not normalize the impedance to allocate population and level of service proportionally.
+The latter method does not allocate population and level of service proportionally, which means that there are often substantial amounts of multiple counting each.
+Figure X shows for the 10-minute threshold the level of inflation/deflation that results when the balanced and the conventional approaches are compared.
+The inflation/deflation is defined as the ratio of the level of service and accessibility calculated using the conventional approach to the balanced approach.
+This effect has been documented in PAEZ-HIGGINS-VIVONA and Chen?.
+Inspection of Fig.
+REFERENCE TO FIGURE, shows that in some locations, the level of service is deflated by up to 40\% (or whatever) and accessibility is inflated by as much as 40\% (or whatever).
 
@@
 
-```{r conventional-5-min-sum-non-equity}
-sum(conventional_5min_non_equity$los$los)
-sum(conventional_5min_non_equity$accessibility$accessibility)
-```
-
-```{r conventional-5-min-sum-all}
-sum(conventional_5min_non_equity$los$los)
-sum(conventional_5min_non_equity$accessibility$accessibility)
-```
-
 ```{r conventional-10-min-non-equity, include=FALSE}
@@
 
-A unique property of the balanced FCA method is that data can be reaggregated while preserving the total population in the core service area and the supply at each docking station.
-This avoids demand and supply inflation, and also enables us to present findings in a way that is perhaps more intuitive to interpret.
-Therefore, we reaggregated population and accessibility from the 50-by-50 m population cells to the dissemination area.
-We then can compared accessibility to median total household income statistics from the 2016 Canadian Census.
+Next, we wish to examine accessibility differences by income levels.
+To this end we take our micro-zones and impute the median household income of their corresponding DA.
+Use of bivariate maps helps us to visualize the level of accessibility by level of income.
+Median total household income statistics were drawn from the 2016 Canadian Census.
 
@@
 
-To further explore this issue and examine vertical equity, we aggregated accessibility by median total household income across all geographies (see Table \ref{tab:accessibility-income}).
+To examine vertical equity we need to estimate what level of accessibility accrues to how many people in different income strata.
+A unique property of the balanced FCA method (which does not hold for the conventional approach because of the inflation/deflation issues discussed above) is that data can be re-aggregated accessibility and population using income as an aggregator criterion, while preserving the total population and the level of service.
+This avoids demand and supply inflation, and also enables us to present findings in a way that is perhaps more intuitive to interpret.
+Therefore, we aggregate the accessibility and population by the median household income as imputed from the DAs (see Table \ref{tab:accessibility-income}).
 Similar to Hosford and Winters  [-@hosfordWhoArePublic2018], we found that Hamilton's PBSP already serviced a large proportion of the population in the bottom and second quintiles of median total household income.
@@
 
-It does not have the property of preserving these quantities. This produces a There is an inflation/deflection effect
-
 Using disaggregated data, we examined spatial equity and accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.

### 2021-06-18 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Major revisions of manuscript

@@
 Many PBSPs in North America now offer specific programs to address equity that primarily focus on removing cost barriers and increasing access for groups that are under-represented among existing users [@trec2019].
+The locations of docking stations is a common consideration for increasing equity [@howlandCurrentEffortsMake2017].
 Hamilton Bike Share (HBS), located in Hamilton, Ontario, was the only Canadian PBSP included in a North American scan of bike share equity programs [see @trec2019].
@@
 
-The location and size of docking stations are important to increase the utility of PBSPs for more people.
-These two factors are relevant to accessibility.
+The location and size of docking stations are two factors that are relevant to accessibility.
 Since the time or distance needed to reach a docking station decreases the potential of accessing the program, the location of stations matters.
-Indeed, distance to stations is associated with bike share use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
 Kabra et al.
@@
 Similar to other public amenities affected by crowding, the utility of stations is also limited by the _maximum_ number of bicycles that they can hold (e.g., their size).
-The program may not necessarily be accessed if stations are easy to reach but offer only a small number of bicycles.
-Likewise, more people may not opt to use the program if the supply of bicycles available at nearby stations is insufficient to meet demand.
 Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts.
@@
 
-Therefore, this paper contributes to the cycling literature by expanding the analysis of Hosford and Winters [-@hosfordWhoArePublic2018] to assess equity in accessibility to HBS and explore the contribution of the "equity" stations.
+Therefore, this paper contributes to the cycling literature by expanding the analysis of Hosford and Winters [-@hosfordWhoArePublic2018] to assess equity in accessibility to HBS and explore the contribution of the "equity" stations added to Hamilton's PBSP.
 
@@
 ```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bicycle share stations in Hamilton, Ontario.
-Conventional stations are in blue and equity stations are in green.
-The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in pink.
+Conventional stations are in blue and equity (ERI) stations are in orange.
+The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in dark green.
 Hamilton Census Metropolitan Area is shown in grey."}
@@
           size = 1,
-          color = "pink") +
+          color = "darkgreen") +
   geom_sf(data = sobi_hubs %>% 
@@
   scale_fill_manual(name = "Hub Type",
-                    values = c("Conventional" = "blue", "ERI" = "green")) +
+                    values = c("Conventional" = "blue", "ERI" = "orange")) +
   theme(panel.background = element_rect(fill = "white"),
@@
 
-In the present case, we found that accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at 5 minutes.
+In the present case, we found that accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at five minutes.
 Accessibility decreases substantially after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each station.
 
-For this reason, we experimented with various thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times from population cell to bike share stations: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
+For this reason, we experimented with various thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times from population cell to bike share stations: three minutes, five minutes, ten minutes, and fifteen minutes.
 We categorized these thresholds as minimum, average, maximum, and extreme, respectively.
@@
 Accessibility is fairly uniform overall, with the exception of two small areas where accessibility is slightly higher.
-This high level of system-wide accessibility occurs because the population that can reach bicycle share stations when travel time is 3 minutes or less is very limited, and accessibility is strongly shaped by a few locations that concentrate population and stations.
-For this reason, accessibility is highly concentrated in small areas around the stations.
-The map is not shown but can be recreated using the source code and data.
+This high level of system-wide accessibility occurs because the population that can reach a docking station when travel time is three minutes or less is very limited, and accessibility is strongly shaped by a few locations that concentrate population and stations.
+For this reason, accessibility is highly concentrated in small areas around those stations.
+The map is not shown, since it is less informative, but can be recreated using the source code and data.
 
@@
 The latter method does not allocate population and level of service proportionally, which means that there are often substantial amounts of multiple counting each.
-Figure X shows for the 10-minute threshold the level of inflation/deflation that results when the balanced and the conventional approaches are compared.
-The inflation/deflation is defined as the ratio of the level of service and accessibility calculated using the conventional approach to the balanced approach.
-This effect has been documented in PAEZ-HIGGINS-VIVONA and Chen?.
-Inspection of Fig.
-REFERENCE TO FIGURE, shows that in some locations, the level of service is deflated by up to 40\% (or whatever) and accessibility is inflated by as much as 40\% (or whatever).
-
-```{r conventional-3-min-non-equity, include=FALSE}
-conventional_3min_non_equity <- ttm_walk %>%
-  dplyr::filter(hub_type == "Conventional" &
-                  hub_status == "Active") %>%
-  c2sfca(threshold = 3)
-```
-
-```{r conventional-3-min-all, include=FALSE}
-conventional_3min_all <- ttm_walk %>%
-  dplyr::filter(hub_status == "Active") %>%
-  c2sfca(threshold = 3)
-```
-
-```{r conventional-3-min-sum-non-equity}
-sum(conventional_3min_non_equity$los$los)
-sum(conventional_3min_non_equity$accessibility$accessibility)
-```
-
-```{r conventional-3-min-sum-all}
-sum(conventional_3min_non_equity$los$los)
-sum(conventional_3min_non_equity$accessibility$accessibility)
-```
+Figure \ref{fig:figure-10} shows the level of inflation and deflation at five minutes that results when the balanced and the conventional approaches are compared.
+The inflation and deflation is defined as the ratio of the level of service and accessibility calculated using the conventional approach to the balanced approach.
+This effect has been documented in other studies [@chenMeasuringAccessibilityHealth2021; @paezDemandLevelService2019].
+Figure \ref{fig:figure-10} shows that in some locations the level of service is deflated by up to 80%.
+This is evident for the equity stations in the east end of the core service area.
+Noticeably, the level of service is least deflated in the downtown area, likely due to the density of stations.
+Figure \ref{fig:figure-11} shows that accessibility is inflated by as much as 50% in many areas across the city.
 
@@
 
-```{r conventional-10-min-non-equity, include=FALSE}
-conventional_10min_non_equity <- ttm_walk %>%
+```{r conventional-5-min-sum-non-equity, include=FALSE}
+sum(conventional_5min_non_equity$los$los)
+sum(conventional_5min_non_equity$accessibility$accessibility)
+```
+
+```{r conventional-5-min-sum-all, include=FALSE}
+sum(conventional_5min_non_equity$los$los)
+sum(conventional_5min_non_equity$accessibility$accessibility)
+```
+
+```{r balanced-5-min-non-equity, include=FALSE}
+balanced_5min_non_equity <- ttm_walk %>%
   dplyr::filter(hub_type == "Conventional" &
                   hub_status == "Active") %>%
-  c2sfca(threshold = 10)
-```
-
-```{r conventional-10-min-all, include=FALSE}
-conventional_10min_all <- ttm_walk %>%
+  b2sfca(threshold = 5)
+```
+
+```{r balanced-5-min-all, include=FALSE}
+balanced_5min_all <- ttm_walk %>%
   dplyr::filter(hub_status == "Active") %>%
-  c2sfca(threshold = 10)
-```
-
-```{r conventional-10-min-sum-non-equity}
-sum(conventional_10min_non_equity$los$los)
-sum(conventional_10min_non_equity$accessibility$accessibility)
-```
-
-```{r conventional-10-min-sum-all}
-sum(conventional_10min_non_equity$los$los)
-sum(conventional_10min_non_equity$accessibility$accessibility)
-```
-
-```{r conventional-15-min-non-equity, include=FALSE}
-conventional_15min_non_equity <- ttm_walk %>%
-  dplyr::filter(hub_type == "Conventional" &
-                  hub_status == "Active") %>%
-  c2sfca(threshold = 15)
-```
-
-```{r conventional-15-min-all, include=FALSE}
-conventional_15min_all <- ttm_walk %>%
-  dplyr::filter(hub_status == "Active") %>%
-  c2sfca(threshold = 15)
-```
-
-```{r conventional-15-min-sum-non-equity}
-sum(conventional_15min_non_equity$los$los)
-sum(conventional_15min_non_equity$accessibility$accessibility)
-```
-
-```{r conventional-15-min-sum-all}
-sum(conventional_15min_non_equity$los$los)
-sum(conventional_15min_non_equity$accessibility$accessibility)
+  b2sfca(threshold = 5)
+```
+
+```{r balanced-5-min-sum-non-equity, include=FALSE}
+sum(balanced_5min_non_equity$los$los)
+sum(balanced_5min_non_equity$accessibility$accessibility)
+```
+
+```{r balanced-5-min-sum-all, include=FALSE}
+sum(conventional_5min_all$los$los)
+sum(conventional_5min_all$accessibility$accessibility)
+```
+
+```{r los-inflation-5-min, include=FALSE}
+los_inflation <- data.frame(hub = conventional_5min_non_equity$los$hub, 
+                       los_inflation = conventional_5min_non_equity$los$los/balanced_5min_non_equity$los$los) %>% # Ratio of conventional los to balanced los
+  left_join(sobi_hubs %>%
+              dplyr::select(OBJECTID),
+            by = c("hub" = "OBJECTID")) %>%
+  st_as_sf()
+
+accessibility_inflation <- data.frame(UID = conventional_5min_non_equity$accessibility$UID,
+                            accessibility_inflation = conventional_5min_non_equity$accessibility$accessibility/balanced_5min_non_equity$accessibility$accessibility) %>%
+  left_join(population_50x50 %>%
+              dplyr::select(UID),
+            by = c("UID" = "UID")) %>%
+  st_as_sf()
+```
+
+```{r figure-10, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Deflation of level of service at 5 minutes walk (average threshold) in the current system with equity stations."}
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "gray") +
+  geom_sf(data = los_inflation,
+            aes(size = los_inflation, color = los_inflation)) +
+  scale_color_distiller(palette = "PuRd", direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
+```
+
+```{r figure-11, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Inflation of accessibility at 5 minutes walk (average threshold) in the current system with equity stations."}
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "gray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_inflation), st_coordinates(accessibility_inflation)),
+            aes(x = X, y = Y, fill = accessibility_inflation)) +
+  scale_fill_distiller(palette = "PuRd", direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
 ```
@@
 
-Next, we wish to examine accessibility differences by income levels.
-To this end we take our micro-zones and impute the median household income of their corresponding DA.
+Next, we wish to examine accessibility differences by income level.
+To this end, we took our micro-zones and imputed the median household income of their corresponding DA.
 Use of bivariate maps helps us to visualize the level of accessibility by level of income.
@@
 
-```{r figure-10, echo=FALSE, fig.align = 'center', out.width = "90%"}
+```{r figure-12, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
           color = "lightgrey") +
-  bi_scale_fill(pal = "DkCyan", 
+  bi_scale_fill(pal = "GrPink", 
                 dim = 3) +
@@
 
-legend <- bi_legend(pal = "DkCyan",
+legend <- bi_legend(pal = "GrPink",
                     dim = 3,
@@
 
-```{r figure-11, echo=FALSE, fig.align = 'center', out.width = "90%"}
+```{r figure-13, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-bi-map-threshold-3, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): without equity stations (top panel) and with equity stations (bottom panel)"}
-ggdraw() + draw_plot(bi_map_3_conv, 0.00, 0.5, 1, 0.5) + 
-  draw_plot(bi_map_3_all, 0.00, 0.0, 1, 0.5) + 
-  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
-```
-
-```{r figure-12, echo=FALSE, fig.align = 'center', out.width = "90%"}
+```{r figure-bi-map-threshold-3, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): with equity stations (top panel) and without equity stations (bottom panel)"}
+ggdraw() + draw_plot(bi_map_3_all, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_3_conv, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2) +
+
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "t", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
+```
+
+```{r figure-14, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-13, echo=FALSE, fig.align = 'center', out.width = "90%"}
+```{r figure-15, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-bi-map-threshold-5, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): without equity stations (top panel) and with equity stations (bottom panel)"}
-ggdraw() + draw_plot(bi_map_5_conv, 0.00, 0.5, 1, 0.5) + 
-  draw_plot(bi_map_5_all, 0.00, 0.0, 1, 0.5) + 
-  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
-```
-
-```{r figure-14, echo=FALSE, fig.align = 'center', out.width = "90%"}
+```{r figure-bi-map-threshold-5, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): with equity stations (top panel) and without equity stations (bottom panel)"}
+ggdraw() + draw_plot(bi_map_5_all, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_5_conv, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2) +
+
+annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "t", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
+```
+
+```{r figure-16, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
           fill = NA,
-          color = "lightgrey") +  bi_scale_fill(pal = "DkCyan", 
+          color = "lightgrey") +  bi_scale_fill(pal = "GrPink", 
                 dim = 3) +
@@
 
-```{r figure-15, echo=FALSE, fig.align = 'center', out.width = "90%"}
+```{r figure-17, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-bi-map-threshold-10, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): without equity stations (top panel) and with equity stations (bottom panel)"}
-ggdraw() + draw_plot(bi_map_10_conv, 0.00, 0.5, 1, 0.5) + 
-  draw_plot(bi_map_10_all, 0.00, 0.0, 1, 0.5) + 
-  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
-```
-
-```{r figure-16, echo=FALSE, fig.align = 'center', out.width = "90%"}
+```{r figure-bi-map-threshold-10, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): with equity stations (top panel) and without equity stations (bottom panel)"}
+ggdraw() + draw_plot(bi_map_10_all, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_10_conv, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2) +
+
+annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "t", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
+```
+
+```{r figure-18, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-17, echo=FALSE, fig.align = 'center', out.width = "90%"}
+```{r figure-19, echo=FALSE, fig.align = 'center', out.width = "90%"}
 data_map <- agg_accessibility_da %>%
@@
 
-```{r figure-bi-map-threshold-15, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): without equity stations (top panel) and equity stations (bottom panel)"}
-ggdraw() + draw_plot(bi_map_15_conv, 0.00, 0.5, 1, 0.5) + 
-  draw_plot(bi_map_15_all, 0.00, 0.0, 1, 0.5) + 
-  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
+```{r figure-bi-map-threshold-15, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): with equity stations (top panel) and without equity stations (bottom panel)"}
+ggdraw() + draw_plot(bi_map_15_all, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_15_conv, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2) +
+
+annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "t", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
 ```
@@
 
-To examine vertical equity we need to estimate what level of accessibility accrues to how many people in different income strata.
-A unique property of the balanced FCA method (which does not hold for the conventional approach because of the inflation/deflation issues discussed above) is that data can be re-aggregated accessibility and population using income as an aggregator criterion, while preserving the total population and the level of service.
+To examine whether equity was increased for groups with lower income, we need to estimate what level of accessibility accrues to how many people in different income strata.
+A unique property of the BFCA method, which does not hold for the conventional 2SFCA approach because of the inflation/deflation issues discussed above, is that accessibility and population data can be re-aggregated using income as an aggregator criterion, while preserving the total population and the level of service.
 This avoids demand and supply inflation, and also enables us to present findings in a way that is perhaps more intuitive to interpret.
-Therefore, we aggregate the accessibility and population by the median household income as imputed from the DAs (see Table \ref{tab:accessibility-income}).
+
+Therefore, we aggregated the accessibility and population by the median household income as imputed from the DAs (see Table \ref{tab:accessibility-income}).
 Similar to Hosford and Winters  [-@hosfordWhoArePublic2018], we found that Hamilton's PBSP already serviced a large proportion of the population in the bottom and second quintiles of median total household income.
 This may well be an artifact of the spatial socioeconomic and demographic profile of Hamilton, where the most dense parts of the city (where a PBSP is most easily launched) are also those with relatively lower incomes.
-On the other hand, the levels of accessibility to Hamilton's PBSP are generally lower for populations in the bottom 20% of median total household income, compared to populations in the top 20%.
-
-The spatial mapping reveals that the addition of equity stations increased horizontal equity by growing the population serviced irrespective of the walking threshold.
+On the other hand, we found the levels of accessibility to Hamilton's PBSP are generally lower for populations in the bottom 20% of median total household income, compared to populations in the top 20%.
+
+The spatial mapping revealed that the addition of equity stations increased accessibility to HBS by growing the population serviced irrespective of the walking threshold.
 The largest gains were made for dissemination areas in the second 20%, where an additional 3,073 and 5,395 people could reach a bicycle share station within three and five minutes walk, respectively, after the addition of equity stations.
 However, we find that there were only small increases in the population in the bottom 20% of median total household income who are serviced by the equity stations, and the accessibility gains are also quite modest and smaller than for populations in the second and third quintiles of median total household income.
-This suggests that income disparities persist, albeit this depends on the walking time thresholds.
+This suggests that inequities in accessibility to HBS persist according to income, albeit this depends on the walking time thresholds.
 With and without equity stations, people in the top 20% of income have the highest level of access at a threshold of ten and fifteen minutes.
-Although dissemination areas in the second 20% have the highest level of access by a significant amount at lower distance thresholds, the bottom 20%, who may benefit the most from Hamilton Bike Share's equity initiatives, have the lowest access at three minutes threshold and the second lowest access at all other thresholds.
+Although dissemination areas in the second 20% have the highest level of access by a significant amount at lower distance thresholds, the bottom 20%, who may benefit the most from the equity stations, have the lowest access at three minutes threshold and the second lowest access at all other thresholds.
 
@@
 
-Using disaggregated data, we examined spatial equity and accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
-Use of a balanced floating catchment area approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro scale which is a sensible approach to avoid the "absorption of disparities", as articulated by Chen et al.
-[-@chenExploringEquityPerformance2019].
+Using disaggregated population data, we examined equity in accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
+Using a BFCA approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro scale which better reflects the friction [*is this the right word?*] of walking thresholds and avoids the "absorption of disparities", as articulated by Chen et al.
+[-@chenExploringEquityPerformance2019] and elsewhere in the transportation literature [@rowangouldIdentifyingEnvironmentalJustice2016].
+The BFCA method also considers demand and supply in the calculation of accessibility, which is important for services like PBSPs but not accounted for in other common accessibility models.
+Unlike the 2SFCA method, there was no inflation or deflation in our accessibility estimates because population and level of service were preserved (see Figure \ref{fig:figure-10} and Figure \ref{fig:figure-11}).
 This differentiates our analysis from similar papers exploring equity in PBSPs that use larger geographical units of analysis or that focus on station location instead of level of service.
-In this way, our paper has made contributions in a positive way by applying an intuitive and useful approach to measure accessibility to a PBSP, and in a normative way by serving to inform future investments in cycle infrastructure for Hamilton Bike Share.
-
-The sensitivity analysis revealed that accessibility to Hamilton Bike Share stations is maximized at five minutes and decreases significantly by eight minutes [see Table \ref{tab:accessibility-income}].
-This reflects the normative guide advertised on some bicycle share stations in Hamilton showing other stations within a five minute walk, as well as the directive of NACTO [@nactowalkingstation2015].
+In this way, our paper has made an important contribution by applying an intuitive and useful approach to measure accessibility to a PBSP.
+This open and reproducible research could be replicated for additional transportation and statistical analysis using other individual-level data in Hamilton or revised for PBSPs in other North American cities.
+
+The sensitivity analysis revealed that accessibility to docking stations is maximized at five minutes and decreases significantly by eight minutes [see Table \ref{tab:accessibility-income}].
+This reflects the normative guide advertised on some docking stations in Hamilton showing other stations within a five minute walk, as well as the directive of NACTO [@nactowalkingstation2015].
 We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}].
-At a minimum threshold of three minutes, too few people can reach stations which leads to relatively high levels of service since there is little crowding.
+At a minimum threshold of three minutes, too few people can reach stations which leads to relatively high levels of service since there is little crowding for the stations.
 However, accessibility is at its lowest after eight minutes whereby congestion effects due to increased potential demand kick in.
+Chen et al.
+[-@chenExploringEquityPerformance2019] also found that accessibility decreases as larger geographical units are used in the estimation, since more people can access the service.
+
 The City of Hamilton has recognized from the launch of the PBSP that substantially more stations and bicycles are needed to service the area [@hamiltonsobi2015].
 With a service area of 40 sq.km, it is estimated that Hamilton should have between 380 and 440 stations instead of 130, and 1,500 bicycles instead of 900 [@hamiltonsobi2015].
-Reduced capacity within the system leads to gaps in coverage in some areas of the city "with some areas not having the recommended station density of 300m between stations or 10 stations per square km" [@hamiltonsobi2015].
-This study illustrates the consequences of an imbalanced system whereby levels of accessibility are not equitable across income groups.
-
-We found that equity stations increased horizontal equity in Hamilton's core service area.
-Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in service were filled by these additional stations.
-This indicates that more individuals at all income quintiles can access Hamilton Bike Share, which leads to a more balanced or equal program.
-In this respect, there are some commonalities between the expansions of Hamilton Bike Share and that of Citi Bike in New York.
-Babgoli et al.
-[-@babagoliExploringHealthSpatial2019] found a slight but not statistically significant increase in the proportion of neighborhoods with the highest levels of poverty with stations after the Citi Bike expansion in 2015.
-Although the Citi Bike expansion was not specifically driven by a desire to reduce inequity in access, 16% of neighborhoods with the highest levels of poverty had stations compared to 12% before.
-Similarly, with the addition of equity stations, there were large gains in accessibility for the second 20% at the average threshold, over 5,000 more people, but much smaller gains for the bottom 20% with only 326 more people.
-
-Vertical inequity, however, persisted as evidenced by differences in accessibility according to income quintile.
+Reduced capacity within the system led to gaps in coverage in some areas of the city "with some areas not having the recommended station density of 300m between stations or 10 stations per square km" [@hamiltonsobi2015].
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in service were filled by the additional equity stations.
+The improvement in accessibility is most noticeable at the average and maximum thresholds in the east end of the city, which corresponds to the neighbourhoods where stations were added.
+However, the contribution of equity stations to reducing inequities in accessibility was modest.
+
+The equity stations were added to HBS as a targeted initiative to increase access for disadvantaged communities.
+However, we found that inequities persisted as evidenced by differences in accessibility according to income quintile (Table \ref{tab:accessibility-income}).
 While the addition of equity stations seems to modestly increase accessibility for all income groups at all thresholds, they did not increase accessibility substantially for any single income group.
@@
 At the minimum threshold, the bottom 20% have the lowest level of access.
-While previous research found that neighborhoods with more disadvantage are better serviced by Hamilton Bike Share, the authors used the Pampalon Deprivation Index to determine the level of disadvantage for dissemination areas _across_ the city not just within the core service area [@hosfordWhoArePublic2018].
+This suggests that the equity stations may not have sufficiently addressed vertical equity.
+While previous research found that neighborhoods with more disadvantage are better serviced by HBS, the authors used the Pampalon Deprivation Index to determine the level of disadvantage for dissemination areas _across_ the city not just within the core service area [@hosfordWhoArePublic2018].
 Instead, we use median total household income for each dissemination area _within_ the core service area.
@@
 At the maximum and extreme thresholds, the top 20% have the highest level of access to Hamilton Bike Share.
-These findings aligns with other studies from Tampa [@chenExploringEquityPerformance2019] and Seattle [@mooneyFreedomStationSpatial2019], which have found disparities in station location or access to bicycles between levels of income and education.
-
-Based on our analysis, we identified specific areas that have both low accessibility and low median total household income which would benefit from an increased supply of public bicycles.
-These empirical findings provide support to the City of Hamilton's efforts to increase equity and balance the system, and confirm that additional stations and bicycles are needed to improve access not only for the bottom 20% but for all income groups.
+These findings aligns with other studies from Tampa [@chenExploringEquityPerformance2019], Philadelphia [@caspiBikesharingPhiladelphiaLowerincome2019] and Seattle [@mooneyFreedomStationSpatial2019], which have found disparities in station location, annual trips, or access to bicycles, respectively, between levels of income and education.
+With the addition of equity stations, there were large gains in accessibility for the second 20% at the average threshold, amounting to over 5,000 more people, but much smaller gains for the bottom 20% with only 326 more people who are able to access a HBS station.
+
+This paper highlights the importance of both the location and size of docking stations for increasing equity in accessibility to PBSPs.
+Several studies have examined who can access PBSPs based on the location of stations [@babagoliExploringHealthSpatial2019; @qianBikesharingEquityDisadvantaged2020; @qianEnhancingEquitableService2020; @wangNewBikeShare2019], but the potential supply of bicycles at stations in disadvantaged communities is not considered in the literature.
+The size of stations (i.e., potential supply of bicycles) was also not identified by PBSP owners and operators as a factor that needed to be considered in equity efforts [@howlandCurrentEffortsMake2017].
+PBSPs may not necessarily reduce inequities in access if stations are easy to reach for disadvantaged groups but offer only a small number of bicycles.
+Likewise, more people may not opt to use the program if the supply of bicycles available at nearby stations is insufficient to meet demand.
+Babgoli et al.
+[-@babagoliExploringHealthSpatial2019] found a slight but not statistically significant increase in the proportion of neighborhoods with the highest levels of poverty with stations after the Citi Bike expansion in 2015.
+Although the Citi Bike expansion was not specifically driven by a desire to reduce inequity in access, 16% of neighborhoods with the highest levels of poverty had stations compared to 12% before.
+This leaves open the question as to whether the stations held enough bicycles to meet demand, and whether congestion effects in neighbourhoods with more poverty are reducing the potential of bike share trips.
+
+The following policy and planning implications can inform efforts for increasing equity in Hamilton Bike Share, and potentially other PBSPs in North America.
+Our analysis has identified specific areas that have both low accessibility and low median total household income.
 Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} highlight potential locations for new equity stations to better accommodate groups with lower socioeconomic status.
+Additional stations are needed to further reduce inequities in accessibility.
+This information will be useful to transportation planners in Hamilton and HBS operators.
+Our findings also suggest that PBSPs should consider increasing the size of stations in disadvantaged communities.
+Docking stations added to target vertical equity should have an adequate supply of public bicycles, particularly more in areas that have a strong interest in adopting bike share.
+Finally, distance to stations is associated with bike share use [@fullerUseNewPublic2011; @wangNewBikeShare2019], especially when the trip starts at the residence [@fishmanBikeShareSynthesis2013], and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
+To address this issue, transportation planners and PBSP operators should consult with groups and communities who are under-represented among bike share users to determine how far they are willing to walk to reach a docking station and to get their input on the location of new docking stations.
 
@@
 
-An additional limitation is the lack of publicly available information about the number of bicycles at each station.
+An additional limitation is the lack of publicly available information about the average number of bicycles at each station on a daily basis.
 HBS works to balance the number of bicycles across stations, but it is reasonable to expect that the number of bicycles will not match exactly the number of racks at every station.
-Ideally, instead of number of bicycle racks as our measure of supply, we would have liked to use the average number of bicycles at stations, perhaps at different times during the day or different seasons.
+Ideally, instead of number of bicycle racks as our measure of supply, we would have liked to use the average number of bicycles at stations, perhaps at different times during the day across many seasons.
 Should this data become available, it would be worthwhile to revisit the question to examine how well operation of the system (including balancing of bicycles across stations) works to maintain the nominal levels of accessibility examined in this paper.
@@
 
-The addition of specific stations to improve equity within the PBSP in Hamilton had the net effect of increasing accessibility and reducing to some extent both horizontal and vertical inequities.
+The addition of specific stations to improve equity within the PBSP in Hamilton had the net effect of increasing accessibility and reducing to some extent vertical inequities.
 In particular, accessibility improved the most for those in the second 20% median total household income at all thresholds, but the gains were only modest for all income groups.
@@
 It would be worthwhile to explore the route choice of bike share trips departing or ending at the equity stations and to identify factors that specifically influence trips from these stations, which would extend existing studies conducted by Scott and colleagues [@luUnderstandingBikeShare2018; @scottWhatFactorsInfluence2019; @scottRouteChoiceBike2021].
-This paper, combined with additional studies such as those conceptualized above, would serve as a valuable case study for Hamilton and other cities with PBSPs that wish to evaluate and address spatial inequality in accessibility and transportation options in urban areas.
+This paper, combined with additional studies such as those conceptualized above, would serve as a valuable case study for Hamilton and other cities with PBSPs that wish to evaluate and address inequities in accessibility and transportation options in urban areas.
 

### 2021-06-18 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Major revisions of manuscript

--- 2021-03-29 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
@@
 \bibliographystyle{elsarticle-harv}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\BuiltInTok}[1]{#1}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
+\newcommand{\ExtensionTok}[1]{#1}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ImportTok}[1]{#1}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
+\newcommand{\NormalTok}[1]{#1}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
+\newcommand{\RegionMarkerTok}[1]{#1}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \ifxetex
@@
             pdfauthor={},
-            pdftitle={Examining horizontal and vertical equity to a public bicycle share program using a balanced floating catchment area accessibility approach},
+            pdftitle={Examining equity in accessibility to bike share - a balanced floating catchment area approach},
             colorlinks=false,
@@
 
-  \title{Examining horizontal and vertical equity to a public bicycle share
-program using a balanced floating catchment area accessibility approach}
+  \title{Examining equity in accessibility to bike share - a balanced floating
+catchment area approach}
     \author[McMaster University]{Elise Desjardins\corref{1}}
@@
   Public bicycle share programs (PBSPs) can play a role in advancing
-  transportation equity if they make cycling more accessible to
+  transportation equity if they make bicycling more accessible to
   disadvantaged populations. In Ontario, the City of Hamilton expanded
-  their PBSP in 2018 by adding twelve docking stations with the explicit
-  objective of increasing spatial equity in access. In this case study, we
-  investigate differentials in accessibility to stations using a balanced
-  floating catchment area accessibility approach and compare accessibility
-  with and without the equity stations. We analyze micro zones to better
-  reflect walking to a station and conduct a sensitivity analysis at
-  several walking time thresholds. We then reaggregate the estimated
-  accessibility for further analysis using census data. Our findings
-  indicate that equity stations increased the serviced population at every
-  threshold examined. Although accessibility increased for the whole
-  population, this increase was relatively modest especially for
-  population in the bottom 20\% of median total household income.
+  their PBSP in 2018 by adding twelve ``equity'' stations with the
+  explicit objective of increasing access for under-serviced
+  neighbourhoods. In this case study, we investigate differentials in
+  accessibility to stations using a balanced floating catchment area
+  approach and compare accessibility with and without the equity stations.
+  We analyze micro zones to better reflect walking to a station and
+  conduct a sensitivity analysis at several walking time thresholds. We
+  then reaggregate the estimated accessibility for further analysis using
+  census data. Our findings indicate that equity stations increased
+  accessibility for the serviced population at every threshold examined,
+  but the increase was relatively modest especially for population in the
+  bottom 20\% of median total household income.
   \end{abstract}
@@
  \end{frontmatter}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{rm}\NormalTok{(}\DataTypeTok{list =} \KeywordTok{ls}\NormalTok{())}
+\end{Highlighting}
+\end{Shaded}
 
@@
 The potential of public bicycle share programs (PBSPs) to increase
-cycling levels is but one of many reasons for implementing such programs
-in urban areas (Hosford et al., 2018, 2019). As a healthy, inexpensive,
-and convenient form of public transportation, shared bicycles can
-encourage individuals to take up cycling for short local trips or first
-and last mile trips to other public transportation instead of using
-personal vehicles. These programs can also play a role in addressing
-transportation needs and advancing transportation equity if they make
-cycling more accessible to individuals and groups with lower
-socioeconomic status. As these programs become increasingly common,
-their introduction has been accompanied by a flurry of research that
-investigates them from the perspective of spatial and transportation
-equity (e.g., Hosford and Winters, 2018; Hull Grasso et al., 2020;
-Mooney et al., 2019; Qian and Jaller, 2021, 2020; Smith et al., 2015).
-Indeed, although these programs are available to the general public and
+bicycling levels is but one of many reasons for implementing such
+programs in urban areas (see Hosford et al., 2018, 2019). As a healthy,
+inexpensive, and convenient form of public transportation, shared
+bicycles can encourage individuals to take up bicycling for short local
+trips or first and last mile trips to other public transportation
+instead of using personal vehicles. These programs can also play a role
+in advancing transportation equity if they make bicycling more
+accessible to disadvantaged populations. Although PBSPs are available to
+the general public in over 800 cities worldwide (Fishman, 2016), and
 ought to be accessible to any individual who wishes to use them,
-research on PBSPs located in North American and European cities
-indicates that inequity in spatial accessibility needs redressing.
-
-In this paper, we investigate the case of the PBSP in the city of
-Hamilton, in Ontario, Canada. Hamilton Bike Share was launched in 2015
-and currently has over 900 operational bicycles and 130 docking
-stations. An interesting feature of this program is that it was expanded
-by an equity initiative that introduced twelve stations in 2018 with the
-explicit objective of increasing spatial accessibility. Hamilton Bike
-Share was already studied by Hosford and Winters (2018) as part of a
-selection of PBSPs in Canada. The authors found that most of the cities
-in their research could benefit of greater efforts to expand service to
-areas with lower socioeconomic groups - but Hamilton fared somewhat
-better than other cities in this respect. What distinguishes our
-research from existing literature is the use of a balanced floating
-catchment area (BFCA) accessibility approach that accounts for the
-supply of stations and the potential demand from the population
-serviced. This requires a more disaggregated approach than the use of
-census geographies (q.v., Hosford and Winters, 2018), since walking
-trips to a bicycle share station likely happen at a level lower than
-even the smallest census geography. For this reason, we implement our
-analysis parting from micro population zones to better reflect the
-friction of walking to a docking station, which is an important
-component of a bike share trip (Chen et al., 2019). Further, we conduct
-a sensitivity analysis at several walking time thresholds. In terms of
-spatial equity, we compare accessibility with and without the equity
-stations to assess the effect of the initiative, before reaggregating
-the data for further analysis using median total household income
-information from the census.
+research on PBSPs indicates that inequities persist with respect to who
+can use and access them.
+
+Many PBSPs in North America now offer specific programs to address
+equity that primarily focus on removing cost barriers and increasing
+access for groups that are under-represented among existing users
+(McNeil et al., 2019). The locations of docking stations is a common
+consideration for increasing equity (Howland et al., 2017). Hamilton
+Bike Share (HBS), located in Hamilton, Ontario, was the only Canadian
+PBSP included in a North American scan of bike share equity programs
+(see McNeil et al., 2019). HBS was launched in 2015 and currently has
+over 900 operational bicycles and 130 docking stations. An equity
+program, Everyone Rides Initiative, was implemented in 2018 which
+expanded the program by introducing twelve ``equity'' stations which
+expanded HBS to more disadvantaged neighbourhoods in the core service
+area.
+
+This paper examines accessibility to Hamilton Bike Share using a
+balanced floating catchment area (BFCA) approach. We also conduct a
+comparative analysis to the conventional two-step floating catchment
+area approach to highlight the benefits of this method which, to our
+knowledge, has not yet been used in the cycling literature. The paper
+also assesses the contribution of HBS's equity stations to reducing
+inequities in accessibility for different groups according to median
+total household income, and provides policy recommendations to further
+improve equity.
 
@@
 \hypertarget{literature-review}{%
-\section{2. Literature Review}\label{literature-review}}
+\section{2.1. Literature Review}\label{literature-review}}
+
+\hypertarget{public-bicycle-share-programs}{%
+\subsection{1.1. Public Bicycle Share
+Programs}\label{public-bicycle-share-programs}}
 
@@
 more likely to be female (Buck et al., 2013), which suggests that the
-gender gap among cyclists who use bike share is less disparate than the
-gap for personal bicycle use (Fishman, 2016). There is some evidence
-that bike share users are less likely to own a car (Buck et al., 2013;
+gender gap among bicyclists who use PBSPs is less disparate than the gap
+for personal bicycle use (Fishman, 2016). There is some evidence that
+bike share users are less likely to own a car (Buck et al., 2013;
 Reilly, Noyes, et al., 2020). However, the relationship between income
@@
 minority or lower socioeconomic status neighborhoods in
-Minneapolis-St.~Paul used the city's PBSP more (J. Wang and Lindsey,
-2019a). Similar findings were reported in London (Ogilvie and Goodman,
-2012). Being university educated was a significant correlate of bike
-share use in Montreal, Canada (Fuller et al., 2011). Perhaps not
+Minneapolis-St.~Paul also used the city's PBSP more (J. Wang and
+Lindsey, 2019a). Similar findings were reported in London (Ogilvie and
+Goodman, 2012). Being university educated was a significant correlate of
+bike share use in Montreal, Canada (Fuller et al., 2011). Not
 coincidentally, financial savings have been found to motivate those on a
@@
 
-Many studies have found that geographic proximity to a bicycle share
-station is an important determinant of membership and use (Fishman,
-2016; Fuller et al., 2011). This makes sense given that individuals are
-more likely to use services or programs that they can easily reach.
-Several studies have recently explored the equity of PBSPs in North
-America by primarily examining who has access to bike share (e.g.,
-differences by demographics or socioeconomic status) and where stations
-are located. It is important to note that equity can be achieved in two
-different ways: balanced or equal spatial distribution across a region
-for all similar groups (e.g., horizontal equity) or greater and targeted
-access to accommodate vulnerable or disadvantaged populations (e.g.,
-vertical equity) (see Chen et al., 2019). Both are of interest to
-researchers and transportation planners since they are often linked in
-that advantage, or conversely disadvantage, has spatial patterns. Using
-a negative binomial regression model, Qian and Jaller (2020) estimated
-ridership in Chicago's PBSP and found some disparities. A minority of
-bike share stations were found to be located in disadvantaged
-communities, while annual members from disadvantaged communities have a
-lower share of trips compared to other areas in the city. This suggests
-that they may be using PBSPs for utilitarian trips (e.g., commuting),
-which points to the importance of ensuring equitable access (Qian and
-Jaller, 2020). Similar results were found in Philadelphia. Despite
-efforts to increase equity within the city's PBSP, census block groups
-with lower median income generated fewer trips which suggested that
-efforts have not been as successful as intended (Caspi and Noland,
-2019). Trips from stations in such areas are utilitarian (e.g.,
+\hypertarget{equity-of-pbsps}{%
+\subsection{1.2. Equity of PBSPs}\label{equity-of-pbsps}}
+
+The introduction of PBSPs has been accompanied by a flurry of research
+focusing on benefits from them. Many studies have examined differences
+in demographics or socioeconomic status between those who use or have
+access to PBSPs and those who don't, while other research has explored
+spatial inequities in where stations are located (e.g., Hosford and
+Winters, 2018; Hull Grasso et al., 2020; Mooney et al., 2019; Qian and
+Jaller, 2021, 2020; Smith et al., 2015). In Chicago, Qian and Jaller
+(2020) estimated ridership in the city's PBSP and found that a minority
+of bike share stations were located in disadvantaged communities, while
+annual members from such areas had a lower share of trips compared to
+other areas in the city. Similar results were found in Philadelphia.
+Despite efforts to increase equity within the city's PBSP, census block
+groups with lower median income generated fewer trips (Caspi and Noland,
+2019). Trips from stations in such areas were utilitarian (e.g.,
 commuting to work), which points to the importance of ensuring equitable
@@
 
+Chen et al.~(2019) distinguish two types of equity related to bike
+share, horizontal and vertical, based on the work of Delbosc and Currie
+(Delbosc and Currie, 2011). Horizontal equity leads to balanced or equal
+distribution of accessibility and costs for all similar groups, while
+vertical equity would involve greater and targeted access for
+under-represented or disadvantaged populations (see Chen et al., 2019).
+Both are of interest to researchers and transportation planners since
+they are often linked in that advantage, or conversely disadvantage, has
+spatial patterns. Bike share equity programs, as identified by McNeil et
+al.~(2019), can be considered efforts to improve vertical equity since
+they favour groups that have benefited less from PBSPs.
+
 On the whole, existing studies highlight the need for PBSPs to be more
-highly accessible for diverse populations in order to increase use
-beyond the ``typical'' users. This has been the focus of recent research
-(see, among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
+accessible for diverse populations in order to increase use beyond the
+``typical'' users. This has been the focus of recent research (see,
+among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
 MacArthur et al., 2020). Offering more people the option of using
@@
 socioeconomic status and might benefit the most, is a worthy policy goal
-for cities with PBSPs. A few North American cities have launched
-specific programs to address barriers to bike share or to expand service
-to more deprived areas: Philadelphia, Pennsylvania (Caspi and Noland,
-2019), and Hamilton, Ontario (Hosford and Winters, 2018). However,
-exploring transportation equity by investigating where bicycle share
-stations are located, often using neighborhoods or census tracts as the
-geographical unit of analysis, can ignore or miss the benefits that may
-be derived from adjacent zones. Meaning that, stations may be lacking in
-certain neighborhoods but there may be stations accessible within a
-reasonable walking time. This is where geographical accessibility
-becomes an important consideration.
+for cities with PBSPs. However, exploring transportation equity by
+investigating where bicycle share stations are located, often using
+neighborhoods or census tracts as the geographical unit of analysis, can
+ignore or miss the benefits that may be derived from adjacent zones.
+Meaning that, stations may be lacking in certain neighborhoods but there
+may be stations accessible within a reasonable walking time. This is
+where geographical accessibility becomes an important consideration.
+
+\hypertarget{accessibility-approaches}{%
+\subsection{2.3. Accessibility
+Approaches}\label{accessibility-approaches}}
 
@@
 linked with recent planning interests in prioritizing modes that are
-suitable for local trips like walking and cycling (Levine, 2020). As a
-measure of the ease of reaching potential destinations spread spatially
-in a given area, accessibility is relevant to PBSPs because it can
-identify current inequality in the provision of infrastructure, as well
-as guide interventions that increase access for groups that are
-under-serviced or address gaps in transportation options. It also
-addresses some of the challenges of other performance measures such as
-level of service within a transportation network by measuring
-person-based indicators and exploring differences in use between
-population groups (Páez et al., 2012).
-
-Beyond the utility derived from using shared bikes to destinations of
-value, an important aspect of accessibility associated with PBSPs is the
-distance an individual must travel to reach a bicycle share station
-(Kabra et al., 2020; J. Wang and Lindsey, 2019b). Since the time or
-distance needed to reach a bicycle share station decreases the potential
-of accessing the program and ultimately their use of the program, the
-location and size (e.g., number of bicycle racks) of stations matters.
-Indeed, distance to bicycle share stations is associated with use
-(Fuller et al., 2011; J. Wang and Lindsey, 2019b) and can be a barrier
-to using PBSPs (Fishman et al., 2014). Kabra et al.~(2020) found that
-the majority of bike share usage in Paris comes from areas within 300 m
-of stations, which amounts to 2-4 minutes walking by an able-bodied
-adult. Furthermore, similar to other public amenities affected by
-crowding, like health care services (e.g., Pereira, Carlos Kauê Vieira
-Braga, et al., 2021), the utility of stations is limited by the
-\emph{maximum} number of bicycles that they can hold. The program may
-not necessarily be improved if stations are easy to reach but offer only
-a small number of bicycles. Likewise, more people may not opt to use the
-program if the supply of bicycles available at nearby stations is
-insufficient to meet demand. The location and size of stations is
-important to increase the utility of this public transportation option
-for more people, thus achieving vertical equity. Accessibility analyses
-for PBSPs constitute a positive and evaluation-based approach that also
-has the potential to inform equity efforts. For instance, Wang and
-Lindsey (2019b) investigated whether new or relocated bicycle share
+suitable for local trips like walking and cycling (Levine, 2020). Beyond
+the utility derived from using shared bicycles to access destinations of
+value, docking stations themselves are amenities because they offer a
+transportation service. Therefore, the ease of reaching these stations,
+which are spread spatially in a given area, can affect use of the PBSP.
+
+The location and size of docking stations are two factors that are
+relevant to accessibility. Since the time or distance needed to reach a
+docking station decreases the potential of accessing the program, the
+location of stations matters. Kabra et al.~(2020) found that the
+majority of bike share usage in Paris comes from areas within 300 m of
+stations, which amounts to 2-4 minutes walking by an adult who does not
+have a disability. Similar to other public amenities affected by
+crowding, the utility of stations is also limited by the \emph{maximum}
+number of bicycles that they can hold (e.g., their size). Accessibility
+analyses for PBSPs constitute a positive and evaluation-based approach
+that also has the potential to inform equity efforts. For instance, Wang
+and Lindsey (2019b) investigated whether new or relocated bicycle share
 stations increased accessibility and use, which offered important
@@
 accessibility, including cumulative opportunities, gravity, and
-utility-based measures (Handy and Niemeier, 1997). Geurs and Van Wee
-(2004) and Paez et al.~(2012) provide recent overviews of various
-formulations and applications of accessibility in transportation
-research. The common gravity-based approach for example involves
-weighting destination opportunities, such as the quantity of bicycle
-share stations, by the time required to reach them from an origin using
-an impedance function (Handy and Niemeier, 1997; Kwan, 1998). However,
-while such measures are suitable for capturing the potential for
-reaching destinations from a given location, they do not take demand or
-congestion effects into account.
-
-In contrast, floating catchment area (FCA) methods incorporate
-information on capacity and demand in calculating accessibility. FCA
-measures have been widely employed in healthcare accessibility research
-and are typically calculated across two steps. In the first, a ratio of
-supply to demand at service locations is calculated, such as the number
-of beds at a hospital divided by the number of people within the
-catchment area of the hospital, weighted by the distance involved in
-reaching the facility. Next, these service level ratios are allocated
-back to the population centers and summarized as a measure of congested
-accessibility. Thus, this model does a good job of considering potential
-crowding or competition for services. While there have been many
-methodological innovations in FCA methods (for example, Delamater, 2013;
-Luo and Qi, 2009; Luo and Wang, 2003; Radke and Mu, 2000; Wan et al.,
-2012), a recent improvement to this approach was achieved through a
-simple and intuitive balancing of the impedance that addressed the
-effects of demand and service inflation found in earlier FCA approaches
-(see Paez et al., 2019).
-
-When measuring accessibility, researchers have taken different
-approaches when it comes to aggregation of data, either by using the
-individual or household as the smallest unit of analysis or larger
-spatial zones. Previous research on bike share equity has typically used
-a meso- or macro-level approach with aggregated data from entire
-neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
-2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
-are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
-also true for studies examining correlates of bike share demand (J. Wang
-and Lindsey, 2019b). Handy and Niemeier (1997) note that using
-disaggregated data in accessibility analyses provides a more accurate
-estimate for individuals, which is useful for addressing vertical
-inequity in PBSP usage. Chen et al.~(2019, p. 530) are in favour of
-using disaggregated data, which they did in their recent analysis of
-Tampa's PBSP, because they note ``the use of aggregated data might
-hinder our understanding of the equity impacts since individual
-disparities are absorbed after aggregation''.
-
-Using the balanced floating catchment area method (Paez et al., 2019), a
-novel approach that has not been used yet in cycling research, with
-disaggregated population data, we examine accessibility to a public
-bicycle share program in Hamilton, Ontario. In this paper, we (1)
-conduct a sensitivity analysis by measuring accessibility and level of
-service to bicycle share stations at different walking time thresholds
-to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes;
-(2) explore the contribution of specific stations that were added to
-Hamilton's PBSP to reducing both horizontal and vertical inequities; and
-(3) examine whether disparities in accessibility exist according to
-median total household income of dissemination areas within the core
-service area.
-
-\hypertarget{sec:study}{%
-\section{3. Case Study}\label{sec:study}}
-
-\hypertarget{original-system}{%
-\subsection{3.1. Original System}\label{original-system}}
-
-The focus of this study is the city of Hamilton, located in Ontario,
-Canada. The city launched a public bicycle share program in March 2015
-with 115 stations and 750 bicycles (Hamilton, 2015a). Before June 2020,
-the program was known as Social Bicycles or SoBi Hamilton, but is now
-called Hamilton Bike Share. Stations are spaced between 300 and 600 m
-apart (Scott et al., 2021). The core service area spans 40 sq.km of the
-city although it was planned to be 20 sq.km (Hamilton, 2015b), and
-roughly 138,000 people are within 30 minutes walking of a bike share
-station {[}see Figure \ref{fig:hamilton-and-sobi-service-area}{]}. This
-represents roughly one fifth of the total population of the Hamilton
-Census Metropolitan Area according to the 2016 Canadian Census. The City
-of Hamilton undertook a large public engagement campaign to validate the
-locations of bicycle share stations that had been selected and to
-crowdsource potential locations for additional locations (Hamilton,
-2014). Most of the suggested locations were in the east end of the core
-service area that lack stations or in neighborhoods not serviced by the
-PBSP. The program was enthusiastically welcomed in the city in 2015 -
-within three weeks of launching, 10,000 trips had been made (Hamilton,
-2015a), however inadequate coverage given the size of Hamilton Bike
-Share's service area was identified as a problem early on, and
-transportation planners noted that the small size (i.e., supply of
-bicycle racks) and low quantity of stations would lead to challenges in
-balancing the system (Hamilton, 2015b). Based on public feedback,
-thirteen stations were added to the original system.
-
-\hypertarget{equity-initiative}{%
-\subsection{3.2. Equity Initiative}\label{equity-initiative}}
-
-In 2017, Hamilton Bike Share Inc., the non-profit organization that
-operates the program, initiated an equity program, Everyone Rides
-Initiative (ERI), with the objective of reducing barriers that may
-prevent individuals from accessing bike share in Hamilton. Additional
-bicycles and stations were added to the program which expanded it to
-more disadvantaged areas in the core service area. The program also
-offers subsidized memberships to individuals who identify as low income,
-and complements this service with cycle skills education. A comparable
-program can be found in Philadelphia (see Caspi and Noland, 2019).
-
-\hypertarget{current-system}{%
-\subsection{3.3. Current System}\label{current-system}}
-
-As of June 2020, Hamilton Bike Share has 900 bikes, 130 stations {[}see
-Figure \ref{fig:sobi-stations-in-hamilton}{]}, and over 26,000 active
-memberships (Hamilton, 2015a). The core service area remains 40 sq.km.
-The PBSP has twelve stations that were added as part of the ERI
-initiative; we refer to these as ``equity stations'' throughout the
-paper while all other stations are ``conventional stations''. In total,
-118 stations are ``conventional'' and 12 are ``equity''.
-
-\hypertarget{membership}{%
-\subsection{3.4. Membership}\label{membership}}
-
-Hamilton Bike Share conducted one membership survey in 2018 (Civicplan,
-2017) and the findings from 420 members are broadly in line with the
-trends that were discussed above (see Fishman, 2016 for a recent review
-of the literature). The majority of respondents live within the core
-service area and the gender split is expected: 57\% of respondents are
-male and 41\% are female. The majority of respondents, both male and
-female, are between 25 and 34 years of age, but the percentage of male
-respondents is higher in the subsequent age groups. Respondents use bike
-share for commuting (40\% of trips) or errands and meetings (24\% of
-trips) and nearly 50\% of trips have an average length of 11 to 20
-minutes. As a result of having a bike share membership, 49\% of
-respondents report that they use their private vehicle less often or
-much less often but 48\% report that their private vehicle use has
-remained about the same. This suggests that Hamilton Bike Share has been
-useful for certain kinds of trips but not all, meaning that some trips
-continue to require a personal vehicle.
-
-\hypertarget{relevant-research}{%
-\subsection{3.5. Relevant Research}\label{relevant-research}}
-
-Our analysis builds upon a previous and recent study (Hosford and
-Winters, 2018), which found that areas in Hamilton with less advantage
-are better served by the city's PBSP compared to other Canadian cities
-{[}i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau{]} where
-areas that are less deprived have greater access. Hosford and Winters
-(2018) acknowledge that ``Hamilton stands out in that the lower income
-neighborhoods are located near the city center and wealthier
-neighborhoods are in the surrounding suburban areas''. Therefore, the
-core service area for the PBSP in Hamilton by default covers some of the
-less advantaged areas in the city. However, there is also a great deal
-of variation in income in the city center because of the local
-university and increasing gentrification. Hosford and Winters (2018)
-took a macro-level and system-wide approach in their analysis by using
-dissemination areas \emph{across} the city as the unit of analysis. They
-did not focus specifically on the core service area and did not
-differentiate between conventional and equity stations.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/hamilton-and-sobi-service-area-1}
-
-}
-
-\caption{The core service area of Hamilton Bike Share is outlined in black. Hamilton Census Metropolitan Area is shown in grey.}\label{fig:hamilton-and-sobi-service-area}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/sobi-stations-in-hamilton-1}
-
-}
-
-\caption{The spatial distribution of bicycle share stations in Hamilton, Ontario.
-The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in pink.
-Hamilton Census Metropolitan Area is shown in grey.}\label{fig:sobi-stations-in-hamilton}
-\end{figure}
-
-\hypertarget{sec:methods}{%
-\section{4. Methods and Data}\label{sec:methods}}
-
-\hypertarget{floating-catchment-area}{%
-\subsection{4.1. Floating Catchment
-Area}\label{floating-catchment-area}}
-
-Floating catchment area (FCA) methods are an approach commonly used in
-the healthcare accessibility literature. This approach is more
-appropriate and informative than calculating provider-to-population
+utility-based measures (Handy and Niemeier, 1997). The gravity-based
+approach involves weighting destination opportunities, such as the
+quantity of bicycle share stations, by the time required to reach them
+from an origin using an impedance function (Handy and Niemeier, 1997;
+Kwan, 1998). While such measures are suitable for capturing the
+potential for reaching destinations from a given location, they do not
+take demand or congestion effects into account which is an important
+consideration when calculating accessibility for amenities such as bike
+share stations.
+
+In contrast, floating catchment area (FCA) methods have been widely
+employed in health care accessibility research. The benefit of this
+approach is that the method incorporates information on capacity,
+demand, and congestion in calculating accessibility. This approach is
+more appropriate and informative than calculating provider-to-population
 ratios (PPR) that simply divide the level of supply of a service (e.g.,
@@
 Mu, 2000) produces flexible catchment areas instead of using rigid
-boundaries like PPR. This provides more useful information because it
-does not assume that people are limited to service within pre-defined
-boundaries (Paez et al., 2019). This is an important property that
-supports our rationale for applying this method to measure accessibility
-to Hamilton Bike Share. The City of Hamilton has positioned stations
-between 300 and 600 m apart, but anticipates that hubs will service
-those living within a 250 m buffer from the station (Hamilton, 2015b).
-The latter constitutes a normative statement: people ought to be able to
-access a station in less than 600 m if they live in the core service
-area, with usage coming from 250 m around. However, it is not known how
-far people are actually willing to travel to reach a station. It would
-be reasonable to assume that people are willing to walk beyond this
+boundaries like PPR. In the first step, a ratio of supply to demand at
+service locations is calculated, such as the number of beds at a
+hospital divided by the number of people within the catchment area of
+the hospital, weighted by the distance involved in reaching the
+facility. Next, these service level ratios are allocated back to the
+population centers and summarized as a measure of congested
+accessibility. Thus, this model does a good job of considering potential
+crowding or competition for services. However, overlapping catchment
+areas from the conventional FCA approach lead to inflation of population
+totals and deflation of service levels across a study area which
+generates inaccurate or misleading accessibility estimates (Paez et al.,
+2019). While there have been many methodological innovations in FCA
+methods (e.g., Delamater, 2013; Luo and Qi, 2009; Luo and Wang, 2003;
+Radke and Mu, 2000; Wan et al., 2012), a recent improvement to this
+approach was achieved through a simple and intuitive balancing of the
+impedance that addressed the effects of demand and service inflation
+found in earlier FCA approaches (see Paez et al., 2019). This provides
+more useful information because it does not assume that people are
+limited to service within pre-defined boundaries (Paez et al., 2019).
+
+When measuring accessibility, researchers have also taken different
+approaches when it comes to the aggregation of data, either by using the
+individual or household as the smallest unit of analysis or larger
+spatial zones. Previous research on bicycle share equity has typically
+used a meso- or macro-level approach with aggregated data from entire
+neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
+2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
+are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
+also true for studies examining correlates of bicycle share demand (J.
+Wang and Lindsey, 2019b). Handy and Niemeier (1997) note that using
+disaggregated data in accessibility analyses provides a more accurate
+estimate for individuals. Chen et al.~(2019, p. 530) are in favour of
+using disaggregated data, which they did in their recent analysis of
+Tampa's PBSP, because they note ``the use of aggregated data might
+hinder our understanding of the equity impacts since individual
+disparities are absorbed after aggregation''.
+
+\hypertarget{previous-research}{%
+\subsection{2.4 Previous Research}\label{previous-research}}
+
+Our analysis builds upon a previous study (Hosford and Winters, 2018),
+which found that areas in Hamilton with less advantage are better served
+by the city's PBSP compared to other Canadian cities (i.e., Toronto,
+Vancouver, Montreal, and Ottawa-Gatineau) where areas that are less
+deprived have greater access. Hosford and Winters (2018, pp.~)
+acknowledge that ``Hamilton stands out in that the lower income
+neighborhoods are located near the city center and wealthier
+neighborhoods are in the surrounding suburban areas''. Therefore, the
+core service area for the PBSP in Hamilton by default covers more of the
+less advantaged areas in the city. However, there is also a great deal
+of variation in income in the city center because of the local
+university and increasing gentrification. Hosford and Winters (2018)
+took a macro-level approach in their analysis by using dissemination
+areas \emph{across} the city as the unit of analysis. They did not focus
+specifically on the core service area and did not differentiate between
+conventional and equity stations.
+
+Therefore, this paper contributes to the cycling literature by expanding
+the analysis of Hosford and Winters (2018) to assess equity in
+accessibility to HBS and explore the contribution of the ``equity''
+stations added to Hamilton's PBSP.
+
+\hypertarget{sec:study}{%
+\section{3. Case Study}\label{sec:study}}
+
+\hypertarget{original-system}{%
+\subsection{3.1. Original System}\label{original-system}}
+
+The case study of this paper is the city of Hamilton, Ontario, Canada.
+Before June 2020, the program was known as Social Bicycles or SoBi
+Hamilton, but is now called Hamilton Bike Share (HBS). The core service
+area spans 40 sq.km of the city, although it was planned to be 20 sq.km
+(Hamilton, 2015a), and roughly 138,000 people are within 30 minutes
+walking of a station {[}see Figure
+\ref{fig:hamilton-and-sobi-service-area}{]}. This represents roughly one
+fifth of the total population of the Hamilton Census Metropolitan Area
+according to the 2016 Canadian Census. The City of Hamilton undertook a
+large public engagement campaign to validate the locations of stations
+that had been selected and to crowdsource potential locations for
+additional locations (Hamilton, 2014). Most of the suggested locations
+were in the east end of the core service area that lack stations or in
+neighborhoods not serviced by the PBSP. The program was enthusiastically
+welcomed in the city in 2015 - within three weeks of launching, 10,000
+trips had been made (Hamilton, 2015b), however inadequate coverage given
+the size of Hamilton Bike Share's service area was identified as a
+problem early on, and transportation planners noted that the small size
+(i.e., supply of bicycle racks) and low quantity of stations would lead
+to challenges in balancing the system (Hamilton, 2015a).
+
+\hypertarget{equity-initiative}{%
+\subsection{3.2. Equity Initiative}\label{equity-initiative}}
+
+Hamilton Bike Share's equity program, Everyone Rides Initiative (ERI),
+was launched in 2018 with the objective of reducing barriers that may
+prevent individuals from accessing the program. Additional bicycles and
+twelve ``equity'' stations were added to the core service area. The
+equity program also offers subsidized memberships to individuals who
+identify as low income, and complements this service with cycle skills
+education. A comparable program can be found in Philadelphia (see Caspi
+and Noland, 2019).
+
+\hypertarget{current-system}{%
+\subsection{3.3. Current System}\label{current-system}}
+
+As of June 2020, HBS has 900 bikes, 130 stations {[}see Figure
+\ref{fig:sobi-stations-in-hamilton}{]}, and over 26,000 active
+memberships (Hamilton, 2015b). The core service area remains 40 sq.km.
+The program has twelve ``equity'' stations and 118 ``conventional''
+stations (i.e., that were not explicitly added to address inequities in
+the program). The City of Hamilton has positioned stations between 300
+and 600 m apart (Scott et al., 2021), but anticipates that they will
+service those living within a 250 m buffer (Hamilton, 2015a). The latter
+constitutes a normative statement: people ought to be able to access a
+station in less than 600 m if they live in the core service area, with
+most usage coming from 250 m around. However, it is not known how far
+people are actually willing to travel to reach a station. It would be
+reasonable to assume that people are willing to walk beyond this
 threshold to access other stations if the ones nearest them have no
@@
 
-More recently, the \emph{balanced} floating catchment area approach was
-developed to address issues with demand and supply inflation that result
-from the overlapping catchment areas produced by earlier FCA methods
-(Paez et al., 2019). Briefly, overlapping catchment areas lead to
-inflation of population totals and deflation of service levels across a
-study area and generates inaccurate or misleading accessibility
-estimates. In contrast, Paez et al.~(2019) adjusted the impedance
-weights so that both supply and demand are proportionally allocated. The
-result is a FCA method that balances the population and level of service
-by eliminating the over-counting of population and level of service that
-leads to distortions in demand and supply. Other benefits of this
-adjusted method include consideration of competition which can occur
-when catchment areas overlap, as well as the preservation of population
-and level of service. Balanced floating catchment area methods have been
-used to explore accessibility to health care providers (Paez et al.,
-2019) and COVID-19 health care services (Pereira, Carlos Kauê Vieira
-Braga, et al., 2021), but have not yet been used in the cycling
-literature to explore issues of accessibility.
-
-The first step in the floating catchment area method is to allocate the
-population to be serviced by each Hamilton Bike Share station: \[
+\begin{verbatim}
+## Source : http://tile.stamen.com/toner-lite/8/70/93.png
+\end{verbatim}
+
+\begin{verbatim}
+## Source : http://tile.stamen.com/toner-lite/8/71/93.png
+\end{verbatim}
+
+\begin{verbatim}
+## Source : http://tile.stamen.com/toner-lite/8/70/94.png
+\end{verbatim}
+
+\begin{verbatim}
+## Source : http://tile.stamen.com/toner-lite/8/71/94.png
+\end{verbatim}
+
+\begin{verbatim}
+## Coordinate system already present. Adding new coordinate system, which will replace the existing one.
+\end{verbatim}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/hamilton-and-sobi-service-area-1}
+
+}
+
+\caption{The core service area of Hamilton Bike Share is outlined in blue. Hamilton Census Metropolitan Area is shown in grey.}\label{fig:hamilton-and-sobi-service-area}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/sobi-stations-in-hamilton-1}
+
+}
+
+\caption{The spatial distribution of bicycle share stations in Hamilton, Ontario.
+Conventional stations are in blue and equity (ERI) stations are in orange.
+The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in dark green.
+Hamilton Census Metropolitan Area is shown in grey.}\label{fig:sobi-stations-in-hamilton}
+\end{figure}
+
+\hypertarget{sec:methods}{%
+\section{4. Methods and Data}\label{sec:methods}}
+
+\hypertarget{balanced-floating-catchment-area}{%
+\subsection{4.1. Balanced Floating Catchment
+Area}\label{balanced-floating-catchment-area}}
+
+The BFCA method was developed to address issues with demand and supply
+inflation that result from the overlapping catchment areas produced by
+earlier FCA methods (Paez et al., 2019). Paez et al.~(2019) adjusted the
+impedance weights so that both supply and demand are proportionally
+allocated. The result is a FCA method that balances the population and
+level of service by eliminating the over-counting of population and
+level of service that leads to distortions in demand and supply.
+
+The first step in the FCA method is to allocate the population to be
+serviced by each docking station:
+
+\begin{equation}
+\label{eq:population-allocation}
 P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}
-\]
+\end{equation}
 
@@
 established catchment area; this gives the level of service of station
-\(j\) in bicycle racks per person: \[
+\(j\) in bicycle racks per person: \begin{equation}
+\label{eq:level-of-service}
 L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
-\]
+\end{equation}
 
@@
 weighted sum of the level of service of all stations that can be reached
-from there according to the spatial weights: \[
+from there according to the spatial weights: \begin{equation}
+\label{eq:FCA-accessibility}
 A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
-\]
+\end{equation}
 
 The balanced approach of Paez et al (2019) replaces the spatial weights
-with normalized versions as follows: \[
+with normalized versions as follows: \begin{equation}
+\label{eq:spatial-weights}
 {w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
-\]
-
-\noindent and: \[
+\end{equation}
+
+\noindent and: \begin{equation}
+\label{eq:spatial-weights-2}
 {w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
-\]
-
-These weights satisfy the following properties: \[
+\end{equation}
+
+These weights satisfy the following properties: \begin{equation}
+\label{eq:weights}
 \sum_{j = 1}^{J} {w^i_{ji}} = 1
-\]
-
-\noindent and: \[
+\end{equation}
+
+\noindent and: \begin{equation}
+\label{eq:weights-2}
 \sum_{i = 1}^{n} {w^j_{ji}} = 1
-\]
+\end{equation}
 
 With these weights, accessibility can be calculated without risk of
-demand or supply inflation: \[
+demand or supply inflation: \begin{equation}
+\label{eq:balanced-accessibility}
 A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}
-\]
+\end{equation}
 
@@
 method preserves the values of the population and level of service and
-provides a regional provider-to-population ratio since: \[
+provides a regional provider-to-population ratio since: \begin{equation}
+\label{eq:proportionality}
 \sum_{i=1}^n A_i = \sum_{j=1}^J L_j = \frac{\sum_{j=1}^JS_j}{\sum_{i=1}^n P_i}
-\]
+\end{equation}
 
@@
 population is never re-allocated elsewhere, this property is replicated
-for any level of aggregation. For this paper, we employ a hybrid
-location-based and person-based approach to calculating accessibility
-using disaggregated population data. In their review of accessibility
-measures, Geurs and van Wee (2004) highlight the need for greater
-inclusion of individual spatio-temporal constraints but acknowledge the
-challenges of acquiring and analyzing person-based data. This comes
-after Kwan's (1998) work to show that space-time measures are better
-able of capturing interpersonal differences, especially the effect of
-space-time constraints on individual behaviour, and are more helpful for
-unraveling gender/ethnic differences. Applying the balanced floating
-catchment area approach allows us to examine accessibility by
-stratifying according to median total household income, which
-constitutes the individual component of the accessibility measure (Geurs
-and van Wee, 2004). However, conducting a further sensitivity analysis
-to measure accessibility at different walking time thresholds would help
-to consider potential spatio-temporal constraints. Different people may
-be willing to travel different distances to access a public bicycle
-share station.
+for any level of aggregation.
 
@@
 
-To obtain population at sub-census geography levels (at the micro
-scale), we use pycnophylactic interpolation (Tobler, 1979). We obtained
-population data from the 2016 Census of Canada for dissemination areas
-(DA), which is the smallest publicly available census geography in
-Canada. These zonal values of the population were interpolated to
-smaller polygons 50-by-50 m in size. Pycnophylactic interpolation
-involves smoothing out the population from each dissemination area while
-preserving total volume {[}see Figure \ref{fig:da-population}{]}. When
-interpolating the population at this high level of resolution, it is
-important to ensure that population numbers were not allocated to areas
-where people do not live in Hamilton (for example, to parks, large
-institutional buildings, etc.). To do so, we retrieved shapefiles for
-various geographic features {[}see Table \ref{tab:data-features}{]} from
-Open Hamilton. Next, we removed these features from the PBSP core
-service area and used pycnophylactic interpolation to disaggregate and
-reallocate population within the remaining area {[}see Figure
-\ref{fig:interpolated-population}{]}.
-
-\begin{table}
-
-\caption{\label{tab:data-features}\label{tab:landscape-features}Landscape features extracted from the Hamilton Bike Share core service area before population is interpolated.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{>{}l|>{\raggedright\arraybackslash}p{30em}}
-\toprule
-Feature & Description\\
-\midrule
-\cellcolor{gray!6}{\textbf{Hamilton Bike Share Stations}} & \cellcolor{gray!6}{The location of stations and the number of racks available at each station.}\\
-\textbf{Golf Courses} & The location of City and privately owned golf courses.\\
-\cellcolor{gray!6}{\textbf{Parks}} & \cellcolor{gray!6}{The location of parks and other green spaces.}\\
-\textbf{Designated Large Employment Areas} & The location of large business parks and industrial lands.\\
-\cellcolor{gray!6}{\textbf{Municipal Parking Lots}} & \cellcolor{gray!6}{The location of municipal car parks.}\\
-\addlinespace
-\textbf{Cemeteries} & The location of cemeteries.\\
-\cellcolor{gray!6}{\textbf{Environmentally Sensitive Areas}} & \cellcolor{gray!6}{The location of either land or water areas containing natural features or significant ecological functions.}\\
-\textbf{Streets} & The street network in Hamilton, including road classification for highways.\\
-\cellcolor{gray!6}{\textbf{Educational Institutions}} & \cellcolor{gray!6}{The location of all educational institutions and schools.}\\
-\textbf{Places of Worship} & The location of buildings used for religious congregations.\\
-\addlinespace
-\cellcolor{gray!6}{\textbf{Municipal Service Centers}} & \cellcolor{gray!6}{The location of all municipal service centers, including City Hall.}\\
-\textbf{Recreation and Community Centers} & The location of all recreation and community centers.\\
-\cellcolor{gray!6}{\textbf{Arenas}} & \cellcolor{gray!6}{The location of all indoor arenas.}\\
-\textbf{Emergency Stations} & The location of all Emergency Management Services (EMS) Ambulance stations.\\
-\cellcolor{gray!6}{\textbf{Fire Stations}} & \cellcolor{gray!6}{The location of all fire stations.}\\
-\addlinespace
-\textbf{Police Stations} & The location of all police stations.\\
-\cellcolor{gray!6}{\textbf{Railways}} & \cellcolor{gray!6}{The railway network in Hamilton.}\\
-\textbf{Hospitals} & The location of all hospitals.\\
-\bottomrule
-\end{tabular}}
-\end{table}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/da-population-1}
-
-}
-
-\caption{Population in all dissemination areas (outlined in white) that are inside or touch the bounding box of the Hamilton Bike Share core service area (outlined in black).}\label{fig:da-population}
-\end{figure}
+Since walking trips to a bicycle share station likely happen at a level
+lower than even the smallest census geography, this requires a more
+disaggregated approach than the use of census geographies. For this
+reason, we implemented our analysis parting from micro population zones
+to better reflect the friction of walking to a docking station, which is
+an important component of a bike share trip (Chen et al., 2019).
+
+To obtain population at sub-census geography levels, we used
+pycnophylactic interpolation (Tobler, 1979). We obtained population data
+from the 2016 Census of Canada for dissemination areas (DA), which is
+the smallest publicly available census geography in Canada. These zonal
+values of the population were interpolated to smaller polygons that are
+50-by-50 m in size. Pycnophylactic interpolation involves smoothing out
+the population from each DA while preserving total volume {[}see Figure
+\ref{fig:da-population}{]}. When interpolating the population at this
+high level of resolution, it is important to ensure that population
+numbers were not allocated to areas where people do not live in Hamilton
+(for example, to parks or large institutional buildings, etc.). To do
+so, we retrieved shapefiles for various geographic features {[}see Table
+\ref{tab:data-features}{]} from Open Hamilton. Next, we removed these
+features from the PBSP core service area and used pycnophylactic
+interpolation to disaggregate and reallocate population within the
+remaining area {[}see Figure \ref{fig:interpolated-population}{]}.
 
@@
 
-To calculate walking times from the centroid of our micro population
-cells, we extracted OpenStreetMap data for Hamilton Bike Share's service
-area from \href{https://download.bbbike.org/osm/bbbike/}{BBBike}, an
-online cycle route planner that interfaces with OpenStreetMap.
-OpenStreetMap data provides the networks for calculating walking times
-from each population cell to nearby bicycle share stations, using a
-maximum walking distance of 10 km and walking time of 30 minutes as
-thresholds. A travel time matrix was created with the origins as the
-coordinates of the population cells and the destinations as the
-coordinates of the bicycle share stations within the maximum threshold.
-This process provides a more realistic measure of the friction of
-reaching stations by taking infrastructure into account of travel times,
-rather than using the Euclidean distance from population cell to
-station. Routing and travel time calculations were completed using the
-\texttt{R} package \texttt{r5r}, used for rapid realistic routing
-operations (Pereira, Saraiva, et al., 2021).
+To calculate walking times from the centroid of the micro population
+cells to docking stations, we extracted OpenStreetMap data for Hamilton
+Bike Share's service area from
+\href{https://download.bbbike.org/osm/bbbike/}{BBBike}, an online cycle
+route planner that interfaces with OpenStreetMap. OpenStreetMap data
+provides the networks for calculating walking times from each population
+cell to nearby bicycle share stations, using a maximum walking distance
+of 10 km and walking time of 30 minutes as thresholds. A travel time
+matrix was created with the origins as the coordinates of the population
+cells and the destinations as the coordinates of the bicycle share
+stations within the maximum threshold. This process provides a more
+realistic measure of the friction of reaching stations by taking
+infrastructure into account network travel times, rather than using the
+Euclidean distance from population cell to station. Routing and travel
+time calculations were completed using the \texttt{R} package
+\texttt{r5r}, used for rapid realistic routing operations (Pereira et
+al., 2021).
 
@@
 Consensus regarding the distance that individuals are willing to walk to
-access a bicycle share station is lacking, but the literature on walking
+access a bike share station is lacking, but the literature on walking
 behaviour can provide some guidelines to determine the thresholds for
-sensitivity analysis. Previous studies have found that living within 250
-m (Fuller et al., 2011) and 300 m (Kabra et al., 2020) is correlated
-with bike share use, while other research has found that walking trips
-are less than 600 m and rarely more than 1200 m (Millward et al., 2013)
-or a median distance of 650 m (Larsen et al., 2010). In Hamilton,
-Hamilton Bike Share will depict a map at some stations to show the user
-the locations of the other nearest stations within a five minute walk,
-which suggests that this is an average distance that people are expected
-to walk. National Association of City Transportation Officials (NACTO)
-has a similar normative guide (City Transportation Officials, 2015).
-
-In the present case, we found that congested accessibility calculated
-using the balanced FCA approach increases with a threshold between two
-and four minutes, but is then maximized at 5 minutes. Accessibility
-decreases substantially after eight minutes, which is intuitive given
-that demand on a limited supply increases as more people can reach each
-station.
-
-For this reason, we experiment with various thresholds by conducting a
+our sensitivity analysis. Previous studies have found that living within
+250 m (Fuller et al., 2011) and 300 m (Kabra et al., 2020) of a docking
+station is correlated with bicycle share use, while other research has
+found that walking trips are less than 600 m and rarely more than 1200 m
+(Millward et al., 2013) or a median distance of 650 m (Larsen et al.,
+2010). Hamilton Bike Share will often depict a map at some docking
+stations to show the locations of the other nearest stations within a
+five minute walk, which suggests that this is an average distance that
+people are expected to walk. National Association of City Transportation
+Officials (NACTO) has a similar normative guide (City Transportation
+Officials, 2015).
+
+In the present case, we found that accessibility calculated using the
+balanced FCA approach increases with a threshold between two and four
+minutes, but is then maximized at five minutes. Accessibility decreases
+substantially after eight minutes, which is intuitive given that demand
+on a limited supply increases as more people can reach each station.
+
+For this reason, we experimented with various thresholds by conducting a
 sensitivity analysis to calculate accessibility at different walking
-times from population cell to bicycle share station: 3 minutes, 5
-minutes, 10 minutes, and 15 minutes. We categorize these thresholds as
-minimum, average, maximum, and extreme, respectively. At each threshold,
-we compare accessibility between the current system and the original
-system to examine the contribution of the additional equity stations.
-When considering the results reported below, it is important to remember
-that accessibility is technically a form of smoothing (O'Kelly and
-Horner, 2003, pp. 7--8): smaller thresholds produce less smoothing
-(which can result in ``spiky'' accessibility landscapes), while larger
-thresholds produce more smoothing and fewer spikes.
+times from population cell to bike share stations: three minutes, five
+minutes, ten minutes, and fifteen minutes. We categorized these
+thresholds as minimum, average, maximum, and extreme, respectively. At
+each threshold, we compared accessibility between the current system and
+the original system to examine the contribution of the twelve equity
+stations. When considering the results reported below, it is important
+to remember that accessibility is technically a form of smoothing
+(O'Kelly and Horner, 2003, pp. 7--8): smaller thresholds produce less
+smoothing (which can result in ``spiky'' accessibility landscapes),
+while larger thresholds produce more smoothing and fewer spikes.
 
@@
 
-With a walking distance of three minutes, we find that there are 25.2
+With a walking distance of three minutes, we found that there are 25.2
 bicycle racks per person system-wide in the original system
@@
 stations increases this ratio slightly to 25.4 bicycles per person.
-Figure \ref{fig:figure-6} presents a comparison of accessibility between
-the systems for the population cells. Accessibility is fairly uniform
-overall, with the exception of two small areas where accessibility is
-slightly higher. This high level of system-wide accessibility occurs
-because the population that can reach bicycle share stations when travel
-time is 3 minutes or less is very limited, and accessibility is strongly
-shaped by a few locations that concentrate population and stations.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-6-1}
-
-}
-
-\caption{Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations.}\label{fig:figure-6}
-\end{figure}
+Accessibility is fairly uniform overall, with the exception of two small
+areas where accessibility is slightly higher. This high level of
+system-wide accessibility occurs because the population that can reach a
+docking station when travel time is three minutes or less is very
+limited, and accessibility is strongly shaped by a few locations that
+concentrate population and stations. For this reason, accessibility is
+highly concentrated in small areas around those stations. The map is not
+shown, since it is less informative, but can be recreated using the
+source code and data.
 
@@
 
-With a walking distance of five minutes, we find that there are 68.6
+With a walking distance of five minutes, we found that there are 68.6
 bicycle racks per person system-wide in the original system
@@
 
-With a walking distance of ten minutes, we find that there are 3.61
+With a walking distance of ten minutes, we found that there are 3.61
 bicycle racks per person system-wide in the original system
@@
 
-With a walking distance of fifteen minutes, we find that there are 2.44
+With a walking distance of fifteen minutes, we found that there are 2.44
 bicycle racks per person system-wide in the original system
@@
 
+\hypertarget{comparative-analysis}{%
+\subsection{5.2. Comparative Analysis}\label{comparative-analysis}}
+
+To highlight the benefit of using the BFCA approach, we conducted a
+comparative analysis with the 2SFCA method. The latter method does not
+allocate population and level of service proportionally, which means
+that there are often substantial amounts of multiple counting each.
+Figure \ref{fig:figure-10} shows the level of inflation and deflation at
+five minutes that results when the balanced and the conventional
+approaches are compared. The inflation and deflation is defined as the
+ratio of the level of service and accessibility calculated using the
+conventional approach to the balanced approach. This effect has been
+documented in other studies (Chen et al., 2021; Paez et al., 2019).
+Figure \ref{fig:figure-10} shows that in some locations the level of
+service is deflated by up to 80\%. This is evident for the equity
+stations in the east end of the core service area. Noticeably, the level
+of service is least deflated in the downtown area, likely due to the
+density of stations. Figure \ref{fig:figure-11} shows that accessibility
+is inflated by as much as 50\% in many areas across the city.
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-10-1}
+
+}
+
+\caption{Deflation of level of service at 5 minutes walk (average threshold) in the current system with equity stations.}\label{fig:figure-10}
+\end{figure}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-11-1}
+
+}
+
+\caption{Inflation of accessibility at 5 minutes walk (average threshold) in the current system with equity stations.}\label{fig:figure-11}
+\end{figure}
+
 \hypertarget{accessibility-by-median-total-household-income}{%
-\subsection{5.2. Accessibility by Median Total Household
+\subsection{5.3. Accessibility by Median Total Household
 Income}\label{accessibility-by-median-total-household-income}}
 
-A unique property of the balanced floating catchment area method is that
-data can be reaggregated while preserving the total population in the
-service area and the supply at each station. This avoids demand and
-supply inflation, and also enables us to present findings in a way that
-is perhaps more intuitive to interpret. Therefore, we reaggregate
-population and accessibility from the 50-by-50 m population cells to the
-dissemination area. We then can compare accessibility to median total
-household income statistics from the Canadian census results from 2016.
-
-Figures 1 {[}\ref{fig:figure-bi-map-threshold-3}{]}, 2
-{[}\ref{fig:figure-bi-map-threshold-5}{]}, 3
-{[}\ref{fig:figure-bi-map-threshold-10}{]}, and 4
-{[}\ref{fig:figure-bi-map-threshold-15}{]} depict bivariate choropleth
-maps that combine the spatial distribution of accessibility and median
-total household income, using tertiles for the coloring scheme. In each
+Next, we wish to examine accessibility differences by income level. To
+this end, we took our micro-zones and imputed the median household
+income of their corresponding DA. Use of bivariate maps helps us to
+visualize the level of accessibility by level of income. Median total
+household income statistics were drawn from the 2016 Canadian Census.
+
+Figures \ref{fig:figure-bi-map-threshold-3},
+\ref{fig:figure-bi-map-threshold-5},
+\ref{fig:figure-bi-map-threshold-10}, and
+\ref{fig:figure-bi-map-threshold-15} depict bivariate choropleth maps
+that combine the spatial distribution of accessibility and median total
+household income, using tertiles for the coloring scheme. In each
 figure, the top panel is the system without the equity stations (i.e.,
@@
 associated with the largest number of people who are within the assumed
-service area of stations. We find that stations added to Hamilton's
-public bicycle share program did indeed expand the spatial coverage of
-the system, leading to more balance across the service area. By
-implementing equity stations in areas with lower median total household
-income in Hamilton, the PBSP attained greater horizontal equity by
-extending the spatial distribution of bicycles across the city. This is
-particularly evident at the minimum and average thresholds of three and
-five minutes, respectively, where the equity stations fill a number of
-gaps in program coverage.
+service area of stations. We found that stations added to HBS did indeed
+expand the spatial coverage of the system, leading to more balance
+across the core service area. By implementing equity stations in areas
+with lower median total household income in Hamilton, the PBSP attained
+greater horizontal equity by extending the spatial distribution of
+bicycles across the city. This is particularly evident at the minimum
+and average thresholds of three and five minutes, respectively, where
+the equity stations fill a number of gaps in program coverage.
+
+\begin{verbatim}
+## Using plotunit = 'm'
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in f(...): True north is not meaningful without coord_sf()
+\end{verbatim}
 
 \begin{figure}
-\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): without equity stations (top panel) and with equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-3}
+\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-3}
 \end{figure}
 
+\begin{verbatim}
+## Using plotunit = 'm'
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in f(...): True north is not meaningful without coord_sf()
+\end{verbatim}
+
 \begin{figure}
-\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): without equity stations (top panel) and with equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-5}
+\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-5}
 \end{figure}
 
+\begin{verbatim}
+## Using plotunit = 'm'
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in f(...): True north is not meaningful without coord_sf()
+\end{verbatim}
+
 \begin{figure}
-\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): without equity stations (top panel) and with equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-10}
+\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-10}
 \end{figure}
 
+\begin{verbatim}
+## Using plotunit = 'm'
+\end{verbatim}
+
+\begin{verbatim}
+## Warning in f(...): True north is not meaningful without coord_sf()
+\end{verbatim}
+
 \begin{figure}
-\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): without equity stations (top panel) and equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-15}
+\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-15}
 \end{figure}
 
-To further explore this issue and examine vertical equity, we aggregate
-accessibility by median total household income across all geographies
-(see Table \ref{tab:accessibility-income}). Similar to Hosford and
-Winters (2018), we find that Hamilton's PBSP already serviced a large
-proportion of the population in the bottom and second quintiles of
-median total household income. This may well be an artifact of the
-spatial socioeconomic and demographic profile of Hamilton, where the
-most dense parts of the city (where a PBSP is most easily launched) are
-also those with relatively lower incomes. On the other hand, the levels
-of accessibility to Hamilton's PBSP are generally lower for populations
-in the bottom 20\% of median total household income, compared to
-populations in the top 20\%.
-
-The spatial mapping reveals that the addition of equity stations
-increased horizontal equity by growing the population serviced
+To examine whether equity was increased for groups with lower income, we
+need to estimate what level of accessibility accrues to how many people
+in different income strata. A unique property of the BFCA method, which
+does not hold for the conventional 2SFCA approach because of the
+inflation/deflation issues discussed above, is that accessibility and
+population data can be re-aggregated using income as an aggregator
+criterion, while preserving the total population and the level of
+service. This avoids demand and supply inflation, and also enables us to
+present findings in a way that is perhaps more intuitive to interpret.
+
+Therefore, we aggregated the accessibility and population by the median
+household income as imputed from the DAs (see Table
+\ref{tab:accessibility-income}). Similar to Hosford and Winters (2018),
+we found that Hamilton's PBSP already serviced a large proportion of the
+population in the bottom and second quintiles of median total household
+income. This may well be an artifact of the spatial socioeconomic and
+demographic profile of Hamilton, where the most dense parts of the city
+(where a PBSP is most easily launched) are also those with relatively
+lower incomes. On the other hand, we found the levels of accessibility
+to Hamilton's PBSP are generally lower for populations in the bottom
+20\% of median total household income, compared to populations in the
+top 20\%.
+
+The spatial mapping revealed that the addition of equity stations
+increased accessibility to HBS by growing the population serviced
 irrespective of the walking threshold. The largest gains were made for
@@
 and smaller than for populations in the second and third quintiles of
-median total household income. This suggests that income disparities
-persist, albeit this depends on the walking time thresholds. With and
-without equity stations, people in the top 20\% of income have the
-highest level of access at a threshold of ten and fifteen minutes.
-Although dissemination areas in the second 20\% have the highest level
-of access by a significant amount at lower distance thresholds, the
-bottom 20\%, who may benefit the most from Hamilton Bike Share's equity
-initiatives, have the lowest access at three minutes threshold and the
-second lowest access at all other thresholds.
+median total household income. This suggests that inequities in
+accessibility to HBS persist according to income, albeit this depends on
+the walking time thresholds. With and without equity stations, people in
+the top 20\% of income have the highest level of access at a threshold
+of ten and fifteen minutes. Although dissemination areas in the second
+20\% have the highest level of access by a significant amount at lower
+distance thresholds, the bottom 20\%, who may benefit the most from the
+equity stations, have the lowest access at three minutes threshold and
+the second lowest access at all other thresholds.
 
@@
 
-Using disaggregated data in this paper, we examined spatial equity and
-accessibility to Hamilton Bike Share, with a particular focus on
-assessing the contribution of the program's equity stations. Use of a
-balanced floating catchment area approach, combined with pycnophylactic
-interpolation, enabled us to measure accessibility on a micro scale
-which is a sensible approach to avoid the ``absorption of disparities'',
-as articulated by Chen et al.~(2019). This differentiates our analysis
-from similar papers exploring equity in PBSPs that use larger
-geographical units of analysis or that focus on station location instead
-of level of service. In this way, our paper has made contributions in a
-positive way by applying an intuitive and useful approach to measure
-accessibility to a PBSP, and in a normative way by serving to inform
-future investments in cycle infrastructure for Hamilton Bike Share.
-
-The sensitivity analysis revealed that accessibility to Hamilton Bike
-Share stations is maximized at five minutes and decreases significantly
-by eight minutes {[}see Table \ref{tab:accessibility-income}{]}. This
-reflects the normative guide advertised on some bicycle share stations
-in Hamilton showing other stations within a five minute walk, as well as
-the directive of NACTO (City Transportation Officials, 2015). We find
-that over 118,000 people can access a bicycle share station within a 15
+Using disaggregated population data, we examined equity in accessibility
+to Hamilton Bike Share, with a particular focus on assessing the
+contribution of the program's equity stations. Using a BFCA approach,
+combined with pycnophylactic interpolation, enabled us to measure
+accessibility on a micro scale which better reflects the friction
+{[}\emph{is this the right word?}{]} of walking thresholds and avoids
+the ``absorption of disparities'', as articulated by Chen et al.~(2019)
+and elsewhere in the transportation literature (Rowangould et al.,
+2016). The BFCA method also considers demand and supply in the
+calculation of accessibility, which is important for services like PBSPs
+but not accounted for in other common accessibility models. Unlike the
+2SFCA method, there was no inflation or deflation in our accessibility
+estimates because population and level of service were preserved (see
+Figure \ref{fig:figure-10} and Figure \ref{fig:figure-11}). This
+differentiates our analysis from similar papers exploring equity in
+PBSPs that use larger geographical units of analysis or that focus on
+station location instead of level of service. In this way, our paper has
+made an important contribution by applying an intuitive and useful
+approach to measure accessibility to a PBSP. This open and reproducible
+research could be replicated for additional transportation and
+statistical analysis using other individual-level data in Hamilton or
+revised for PBSPs in other North American cities.
+
+The sensitivity analysis revealed that accessibility to docking stations
+is maximized at five minutes and decreases significantly by eight
+minutes {[}see Table \ref{tab:accessibility-income}{]}. This reflects
+the normative guide advertised on some docking stations in Hamilton
+showing other stations within a five minute walk, as well as the
+directive of NACTO (City Transportation Officials, 2015). We find that
+over 118,000 people can access a bicycle share station within a 15
 minute walk, which represents roughly 85\% of the total population in
@@
 which leads to relatively high levels of service since there is little
-crowding. However, accessibility is at its lowest after eight minutes
-whereby congestion effects due to increased potential demand kick in.
+crowding for the stations. However, accessibility is at its lowest after
+eight minutes whereby congestion effects due to increased potential
+demand kick in. Chen et al.~(2019) also found that accessibility
+decreases as larger geographical units are used in the estimation, since
+more people can access the service.
+
 The City of Hamilton has recognized from the launch of the PBSP that
 substantially more stations and bicycles are needed to service the area
-(Hamilton, 2015b). With a service area of 40 sq.km, it is estimated that
+(Hamilton, 2015a). With a service area of 40 sq.km, it is estimated that
 Hamilton should have between 380 and 440 stations instead of 130, and
-1,500 bicycles instead of 900 (Hamilton, 2015b). Reduced capacity within
-the system leads to gaps in coverage in some areas of the city ``with
-some areas not having the recommended station density of 300m between
-stations or 10 stations per square km'' (Hamilton, 2015b). This study
-illustrates the consequences of an imbalanced system whereby levels of
-accessibility are not equitable across income groups.
-
-We found that equity stations increased horizontal equity in Hamilton's
-core service area. Figures \ref{fig:figure-bi-map-threshold-3},
+1,500 bicycles instead of 900 (Hamilton, 2015a). Reduced capacity within
+the system led to gaps in coverage in some areas of the city ``with some
+areas not having the recommended station density of 300m between
+stations or 10 stations per square km'' (Hamilton, 2015a). Figures
+\ref{fig:figure-bi-map-threshold-3},
 \ref{fig:figure-bi-map-threshold-5},
@@
 \ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in service
-were filled by these additional stations. This indicates that more
-individuals at all income quintiles can access Hamilton Bike Share,
-which leads to a more balanced or equal program. In this respect, there
-are some commonalities between the expansions of Hamilton Bike Share and
-that of Citi Bike in New York. Babgoli et al.~(2019) found a slight but
-not statistically significant increase in the proportion of
-neighborhoods with the highest levels of poverty with stations after the
-Citi Bike expansion in 2015. Although the Citi Bike expansion was not
-specifically driven by a desire to reduce inequity in access, 16\% of
-neighborhoods with the highest levels of poverty had stations compared
-to 12\% before. Similarly, with the addition of equity stations, there
-were large gains in accessibility for the second 20\% at the average
-threshold, over 5,000 more people, but much smaller gains for the bottom
-20\% with only 326 more people.
-
-Vertical inequity, however, persisted as evidenced by differences in
-accessibility according to income quintile. While the addition of equity
-stations seems to modestly increase accessibility for all income groups
-at all thresholds, they did not increase accessibility substantially for
-any single income group. Most importantly, individuals in the bottom
-20\% of median total household income have the second lowest level of
-access to Hamilton Bike Share at most thresholds (average, maximum, and
-extreme). At the minimum threshold, the bottom 20\% have the lowest
-level of access. While previous research found that neighborhoods with
-more disadvantage are better serviced by Hamilton Bike Share, the
-authors used the Pampalon Deprivation Index to determine the level of
-disadvantage for dissemination areas \emph{across} the city not just
-within the core service area (Hosford and Winters, 2018). Instead, we
-use median total household income for each dissemination area
-\emph{within} the core service area. We conclude that Hamilton's PBSP,
-while by default located in areas with more deprivation compared to
-other cities, has disparities in accessibility between income groups.
-With equity stations, many areas with low median total household income
-in the east end of the service area continue to have low accessibility.
-At the maximum and extreme thresholds, the top 20\% have the highest
-level of access to Hamilton Bike Share. These findings aligns with other
-studies from Tampa (Chen et al., 2019) and Seattle (Mooney et al.,
-2019), which have found disparities in station location or access to
-bicycles between levels of income and education.
-
-Based on our analysis, we identified specific areas that have both low
-accessibility and low median total household income which would benefit
-from an increased supply of public bicycles. These empirical findings
-provide support to the City of Hamilton's efforts to increase equity and
-balance the system, and confirm that additional stations and bicycles
-are needed to improve access not only for the bottom 20\% but for all
-income groups. Figures \ref{fig:figure-bi-map-threshold-3},
+were filled by the additional equity stations. The improvement in
+accessibility is most noticeable at the average and maximum thresholds
+in the east end of the city, which corresponds to the neighbourhoods
+where stations were added. However, the contribution of equity stations
+to reducing inequities in accessibility was modest.
+
+The equity stations were added to HBS as a targeted initiative to
+increase access for disadvantaged communities. However, we found that
+inequities persisted as evidenced by differences in accessibility
+according to income quintile (Table \ref{tab:accessibility-income}).
+While the addition of equity stations seems to modestly increase
+accessibility for all income groups at all thresholds, they did not
+increase accessibility substantially for any single income group. Most
+importantly, individuals in the bottom 20\% of median total household
+income have the second lowest level of access to Hamilton Bike Share at
+most thresholds (average, maximum, and extreme). At the minimum
+threshold, the bottom 20\% have the lowest level of access. This
+suggests that the equity stations may not have sufficiently addressed
+vertical equity. While previous research found that neighborhoods with
+more disadvantage are better serviced by HBS, the authors used the
+Pampalon Deprivation Index to determine the level of disadvantage for
+dissemination areas \emph{across} the city not just within the core
+service area (Hosford and Winters, 2018). Instead, we use median total
+household income for each dissemination area \emph{within} the core
+service area. We conclude that Hamilton's PBSP, while by default located
+in areas with more deprivation compared to other cities, has disparities
+in accessibility between income groups. With equity stations, many areas
+with low median total household income in the east end of the service
+area continue to have low accessibility. At the maximum and extreme
+thresholds, the top 20\% have the highest level of access to Hamilton
+Bike Share. These findings aligns with other studies from Tampa (Chen et
+al., 2019), Philadelphia (Caspi and Noland, 2019) and Seattle (Mooney et
+al., 2019), which have found disparities in station location, annual
+trips, or access to bicycles, respectively, between levels of income and
+education. With the addition of equity stations, there were large gains
+in accessibility for the second 20\% at the average threshold, amounting
+to over 5,000 more people, but much smaller gains for the bottom 20\%
+with only 326 more people who are able to access a HBS station.
+
+This paper highlights the importance of both the location and size of
+docking stations for increasing equity in accessibility to PBSPs.
+Several studies have examined who can access PBSPs based on the location
+of stations (Babagoli et al., 2019; Qian and Jaller, 2020; Qian et al.,
+2020; J. Wang and Lindsey, 2019b), but the potential supply of bicycles
+at stations in disadvantaged communities is not considered in the
+literature. The size of stations (i.e., potential supply of bicycles)
+was also not identified by PBSP owners and operators as a factor that
+needed to be considered in equity efforts (Howland et al., 2017). PBSPs
+may not necessarily reduce inequities in access if stations are easy to
+reach for disadvantaged groups but offer only a small number of
+bicycles. Likewise, more people may not opt to use the program if the
+supply of bicycles available at nearby stations is insufficient to meet
+demand. Babgoli et al.~(2019) found a slight but not statistically
+significant increase in the proportion of neighborhoods with the highest
+levels of poverty with stations after the Citi Bike expansion in 2015.
+Although the Citi Bike expansion was not specifically driven by a desire
+to reduce inequity in access, 16\% of neighborhoods with the highest
+levels of poverty had stations compared to 12\% before. This leaves open
+the question as to whether the stations held enough bicycles to meet
+demand, and whether congestion effects in neighbourhoods with more
+poverty are reducing the potential of bike share trips.
+
+The following policy and planning implications can inform efforts for
+increasing equity in Hamilton Bike Share, and potentially other PBSPs in
+North America. Our analysis has identified specific areas that have both
+low accessibility and low median total household income. Figures
+\ref{fig:figure-bi-map-threshold-3},
 \ref{fig:figure-bi-map-threshold-5},
@@
 new equity stations to better accommodate groups with lower
-socioeconomic status.
+socioeconomic status. Additional stations are needed to further reduce
+inequities in accessibility. This information will be useful to
+transportation planners in Hamilton and HBS operators. Our findings also
+suggest that PBSPs should consider increasing the size of stations in
+disadvantaged communities. Docking stations added to target vertical
+equity should have an adequate supply of public bicycles, particularly
+more in areas that have a strong interest in adopting bike share.
+Finally, distance to stations is associated with bike share use (Fuller
+et al., 2011; J. Wang and Lindsey, 2019b), especially when the trip
+starts at the residence (Fishman et al., 2013), and can be a barrier to
+using PBSPs (Fishman et al., 2014). To address this issue,
+transportation planners and PBSP operators should consult with groups
+and communities who are under-represented among bike share users to
+determine how far they are willing to walk to reach a docking station
+and to get their input on the location of new docking stations.
 
@@
 to determine whether the addition of equity stations encouraged more
-cycling for low-income individuals living near them. Other studies have
+bicycling for disadvantaged groups living near them. Other studies have
 specifically looked at differences in trip type, frequency, or length
@@
 limited by the lack of publicly available route and individual user data
-to conduct similar analyses for Hamilton Bike Share.
+to conduct similar analyses for HBS.
 
 An additional limitation is the lack of publicly available information
-about the number of bicycles at each station. Hamilton Bike share works
-to balance the number of bicycles across stations, but it is reasonable
-to expect that the number of bicycles will not match exactly the number
-of racks at every station. Ideally, instead of number of bicycle racks
-as our measure of supply, we would have liked to use the average number
-of bicycles at stations, perhaps at different times during the day or
-different seasons. Should this data become available, it would be
-worthwhile to revisit the question to examine how well operation of the
-system (including balancing of bicycles across stations) works to
-maintain the nominal levels of accessibility examined in this paper.
+about the average number of bicycles at each station on a daily basis.
+HBS works to balance the number of bicycles across stations, but it is
+reasonable to expect that the number of bicycles will not match exactly
+the number of racks at every station. Ideally, instead of number of
+bicycle racks as our measure of supply, we would have liked to use the
+average number of bicycles at stations, perhaps at different times
+during the day across many seasons. Should this data become available,
+it would be worthwhile to revisit the question to examine how well
+operation of the system (including balancing of bicycles across
+stations) works to maintain the nominal levels of accessibility examined
+in this paper.
 
@@
 
-The addition of specific equity stations to the public bicycle share
-program in Hamilton, Ontario had the net effect of increasing
-accessibility and reducing to some extent both horizontal and vertical
-inequities. In particular, accessibility improved the most for those in
-the second 20\% median total household income at all thresholds, but the
-gains were only modest for all income groups. Dissemination areas with
-the bottom 20\% had the lowest accessibility at three minutes, and
-second lowest levels of accessibility at five, ten, and fifteen minutes.
-Congestion effects were observed at higher thresholds, with
-accessibility decreasing significantly once the catchment area is
-increased to ten minutes walking.
+The addition of specific stations to improve equity within the PBSP in
+Hamilton had the net effect of increasing accessibility and reducing to
+some extent vertical inequities. In particular, accessibility improved
+the most for those in the second 20\% median total household income at
+all thresholds, but the gains were only modest for all income groups.
+Dissemination areas with the bottom 20\% had the lowest accessibility at
+three minutes, and second lowest levels of accessibility at five, ten,
+and fifteen minutes. Congestion effects were observed at higher
+thresholds, with accessibility decreasing significantly once the
+catchment area is increased to ten minutes walking.
 
@@
 conceptualized above, would serve as a valuable case study for Hamilton
-and other cities with PBSPs that wish to evaluate and address spatial
-inequality in accessibility and transportation options in urban areas.
+and other cities with PBSPs that wish to evaluate and address inequities
+in accessibility and transportation options in urban areas.
 
@@
 \texttt{biscale} (Prener et al., 2020), \texttt{cowplot} (Wilke, 2020),
-\texttt{data.table} (Dowle and Srinivasan, 2020), \texttt{disk.frame}
-(ZJ, 2020), \texttt{gdistance} (van Etten, 2020), \texttt{gridExtra}
-(Auguie, 2017), \texttt{kableExtra} (Zhu, 2020), \texttt{knitr} (Xie,
+\texttt{data.table} (Dowle and Srinivasan, 2021), \texttt{disk.frame}
+(ZJ, 2021), \texttt{gdistance} (van Etten, 2020), \texttt{gridExtra}
+(Auguie, 2017), \texttt{kableExtra} (Zhu, 2021), \texttt{knitr} (Xie,
 2021a), \texttt{pycno} (Brunsdon, 2014), \texttt{r5r} (Saraiva et al.,
@@
 2021), \texttt{rticles} (Allaire et al., 2021), \texttt{sf} (Pebesma,
-2021), \texttt{tidyverse} (Wickham, 2019), \texttt{tinytex} (Xie,
-2021b), \texttt{units} (Pebesma et al., 2020).
+2021), \texttt{tidyverse} (Wickham, 2021), \texttt{tinytex} (Xie,
+2021b), \texttt{units} (Pebesma et al., 2021).
 
@@
 
+\leavevmode\hypertarget{ref-chenMeasuringAccessibilityHealth2021}{}%
+Chen, W., Cheng, L., Chen, X., Chen, J., Cao, M., 2021. Measuring
+accessibility to health care services for older bus passengers: A finer
+spatial resolution. Journal of Transport Geography 93, 103068.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2021.103068}{10.1016/j.jtrangeo.2021.103068}
+
 \leavevmode\hypertarget{ref-chenExploringEquityPerformance2019}{}%
@@
 
-\leavevmode\hypertarget{ref-civicplanSoBiHamiltonMembership2017}{}%
-Civicplan, 2017. SoBi Hamilton Membership Survey. Civicplan \textbar{}
-Planning Engagement Research.
-
 \leavevmode\hypertarget{ref-delamaterSpatialAccessibilitySuboptimally2013}{}%
@@
 
+\leavevmode\hypertarget{ref-delboscUsingLorenzCurves2011}{}%
+Delbosc, A., Currie, G., 2011. Using Lorenz curves to assess public
+transport equity. Journal of Transport Geography, Special section on
+Alternative Travel futures 19, 1252--1259.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2011.02.008}{10.1016/j.jtrangeo.2011.02.008}
+
 \leavevmode\hypertarget{ref-R-data.table}{}%
-Dowle, M., Srinivasan, A., 2020. Data.table: Extension of `data.frame`.
+Dowle, M., Srinivasan, A., 2021. Data.table: Extension of `data.frame`.
 
@@
 doi:\href{https://doi.org/10.1080/01441647.2015.1033036}{10.1080/01441647.2015.1033036}
+
+\leavevmode\hypertarget{ref-fishmanBikeShareSynthesis2013}{}%
+Fishman, E., Washington, S., Haworth, N., 2013. Bike Share: A Synthesis
+of the Literature. Transport Reviews 33, 148--165.
+doi:\href{https://doi.org/10.1080/01441647.2013.775612}{10.1080/01441647.2013.775612}
 
@@
 
-\leavevmode\hypertarget{ref-geursAccessibilityEvaluationLanduse2004}{}%
-Geurs, K.T., van Wee, B., 2004. Accessibility evaluation of land-use and
-transport strategies: Review and research directions. Journal of
-Transport Geography 12, 127--140.
-doi:\href{https://doi.org/10.1016/j.jtrangeo.2003.10.005}{10.1016/j.jtrangeo.2003.10.005}
-
 \leavevmode\hypertarget{ref-hamiltonsobi2014}{}%
@@
 
-\leavevmode\hypertarget{ref-hamiltonHamiltonBikeShare2015}{}%
-Hamilton, C. of, 2015a. Hamilton Bike Share.
-
 \leavevmode\hypertarget{ref-hamiltonsobi2015}{}%
-Hamilton, C. of, 2015b. Public bike share transit system implementation
+Hamilton, C. of, 2015a. Public bike share transit system implementation
 plan (pw13015c) - (city wide) {[}WWW Document{]}. URL
 \url{https://pub-hamilton.escribemeetings.com/filestream.ashx?DocumentId=118356}
+
+\leavevmode\hypertarget{ref-hamiltonHamiltonBikeShare2015}{}%
+Hamilton, C. of, 2015b. Hamilton Bike Share.
 
@@
 
+\leavevmode\hypertarget{ref-howlandCurrentEffortsMake2017}{}%
+Howland, S., McNeil, N., Broach, J., Rankins, K., MacArthur, J., Dill,
+J., 2017. Current Efforts to Make Bikeshare More Equitable: Survey of
+System Owners and Operators. Transportation Research Record 2662,
+160--167. doi:\href{https://doi.org/10.3141/2662-18}{10.3141/2662-18}
+
 \leavevmode\hypertarget{ref-hullgrassoBikeShareEquity2020}{}%
@@
 doi:\href{https://doi.org/10.1177/0361198120925079}{10.1177/0361198120925079}
+
+\leavevmode\hypertarget{ref-trec2019}{}%
+McNeil, N., MacArthur, J., Broach, J., Cummings, A., Stark, R.-L., 2019.
+National scan of bik national scan of bike share equity pr quity
+programs: Approaches and best practices for promoting equity in bike
+share {[}WWW Document{]}. URL
+\url{https://pdxscholar.library.pdx.edu/cgi/viewcontent.cgi?article=1191\&context=trec_reports}
 
@@
 \leavevmode\hypertarget{ref-R-units}{}%
-Pebesma, E., Mailund, T., Kalinowski, T., 2020. Units: Measurement units
-for r vectors.
-
-\leavevmode\hypertarget{ref-pereiraGeographicAccessCOVID192021}{}%
-Pereira, R.H.M., Braga, C.K.V., Servo, L.M., Serra, B., Amaral, P.,
-Gouveia, N., Paez, A., 2021. Geographic access to COVID-19 healthcare in
-Brazil using a balanced float catchment area approach. Social Science \&
-Medicine 273, 113773.
-doi:\href{https://doi.org/10.1016/j.socscimed.2021.113773}{10.1016/j.socscimed.2021.113773}
+Pebesma, E., Mailund, T., Kalinowski, T., Ucar, I., 2021. Units:
+Measurement units for r vectors.
 
@@
 
+\leavevmode\hypertarget{ref-qianEnhancingEquitableService2020}{}%
+Qian, X., Jaller, M., Niemeier, D., 2020. Enhancing equitable service
+level: Which can address better, dockless or dock-based Bikeshare
+systems? Journal of Transport Geography 86, 102784.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2020.102784}{10.1016/j.jtrangeo.2020.102784}
+
 \leavevmode\hypertarget{ref-radkeSpatialDecompositionsModeling2000}{}%
@@
 
+\leavevmode\hypertarget{ref-rowangouldIdentifyingEnvironmentalJustice2016}{}%
+Rowangould, D., Karner, A., London, J., 2016. Identifying environmental
+justice communities for transportation analysis. Transportation Research
+Part A: Policy and Practice 88, 151--162.
+doi:\href{https://doi.org/10.1016/j.tra.2016.04.002}{10.1016/j.tra.2016.04.002}
+
 \leavevmode\hypertarget{ref-R-r5r}{}%
-Saraiva, M., Pereira, R.H.M., Herszenhut, D., Braga, C.K.V., 2021. R5r:
-Rapid realistic routing with r5.
+Saraiva, M., Pereira, R.H.M., Herszenhut, D., Braga, C.K.V., Conway,
+M.W., 2021. R5r: Rapid realistic routing with r5.
 
@@
 Scott, D.M., Lu, W., Brown, M.J., 2021. Route choice of bike share
-users: Leveraging GPS data to derive choice sets. Journal of Transport
+users: Leveraging gps data to derive choice sets. Journal of Transport
 Geography 90, 102903.
@@
 of us bicycle sharing systems. Western Michigan University.
-Transportation Research Center for Livable~\ldots.
+Transportation Research Center for Livable~.
 
@@
 \leavevmode\hypertarget{ref-R-tidyverse}{}%
-Wickham, H., 2019. Tidyverse: Easily install and load the tidyverse.
+Wickham, H., 2021. Tidyverse: Easily install and load the tidyverse.
 
@@
 \leavevmode\hypertarget{ref-R-kableExtra}{}%
-Zhu, H., 2020. KableExtra: Construct complex table with kable and pipe
+Zhu, H., 2021. KableExtra: Construct complex table with kable and pipe
 syntax.
@@
 \leavevmode\hypertarget{ref-R-disk.frame}{}%
-ZJ, D., 2020. Disk.frame: Larger-than-ram disk-based data manipulation
+ZJ, D., 2021. Disk.frame: Larger-than-ram disk-based data manipulation
 framework.

### 2021-06-21 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Edits after Elise revised the paper

@@
 
-```{r}
+```{r echo=FALSE}
 rm(list = ls())
@@
 
-```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "The core service area of Hamilton Bike Share is outlined in blue.
+```{r hamilton-and-sobi-service-area, echo=FALSE, message=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "The core service area of Hamilton Bike Share is outlined in blue.
 Hamilton Census Metropolitan Area is shown in grey."}
@@
 
-By allocating the population and level of service proportionally, this method preserves the values of the population and level of service and provides a regional provider-to-population ratio since:
+By allocating the population and level of service proportionally, this method preserves the values of the population and level of service since:
 \begin{equation}
 \label{eq:proportionality}
-\sum_{i=1}^n A_i = \sum_{j=1}^J L_j = \frac{\sum_{j=1}^JS_j}{\sum_{i=1}^n P_i}
+\sum_{i=1}^n A_i = \sum_{j=1}^J L_j
 \end{equation}
@@
 
-With a walking distance of three minutes, we found that there are 25.2 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
-The addition of equity stations increases this ratio slightly to 25.4 bicycles per person.
-Accessibility is fairly uniform overall, with the exception of two small areas where accessibility is slightly higher.
+With a walking distance of three minutes, we found that the total level of service is 25.2 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+The addition of equity stations increases this slightly to 25.4 bicycles per person.
+This total level of service is allocated to the population to obtain the levels of accessibility, which turn out to be relatively uniform overall, with the exception of two small areas where accessibility is slightly higher.
 This high level of system-wide accessibility occurs because the population that can reach a docking station when travel time is three minutes or less is very limited, and accessibility is strongly shaped by a few locations that concentrate population and stations.
@@
           color = "lightgrey") +
+  ggtitle("All stations") +
   bi_scale_fill(pal = "GrPink", 
                 dim = 3) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering) +
   theme(panel.background = element_rect(fill = "white"),
@@
         axis.ticks = element_blank(),
-        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
 
@@
           color = "lightgrey") +
+  ggtitle("Conventional") +
   bi_scale_fill(pal = "GrPink", 
                 dim = 3) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering) +
   theme(panel.background = element_rect(fill = "white"),
@@
         axis.ticks = element_blank(),
-        plot.margin = unit(c(0, 0, 0, 0), "cm"))
-```
-
-```{r figure-bi-map-threshold-3, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): with equity stations (top panel) and without equity stations (bottom panel)"}
-ggdraw() + draw_plot(bi_map_3_all, 0.00, 0.5, 1, 0.5) + 
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-bi-map-threshold-3, echo = FALSE, out.width="1.2\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): with equity stations (top panel) and without equity stations (bottom panel)"}
+ggdraw() + 
+  draw_plot(bi_map_3_all, 0.00, 0.5, 1, 0.5) + 
   draw_plot(bi_map_3_conv, 0.00, 0.0, 1, 0.5) + 
-  draw_plot(legend, 0.06, 0.4, 0.2, 0.2) +
-
-  annotation_scale(location = "bl", width_hint = 0.4,
-                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
-  annotation_north_arrow(location = "t", which_north = "true", 
-        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
-        style = north_arrow_fancy_orienteering)
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
 ```
@@
         axis.ticks = element_blank(),
-        plot.margin = unit(c(0, 0, 0, 0), "cm"))
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
 ```
@@
         axis.ticks = element_blank(),
-        plot.margin = unit(c(0, 0, 0, 0), "cm"))
-```
-
-```{r figure-bi-map-threshold-5, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): with equity stations (top panel) and without equity stations (bottom panel)"}
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-bi-map-threshold-5, echo = FALSE, out.width="1.2\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): with equity stations (top panel) and without equity stations (bottom panel)"}
 ggdraw() + draw_plot(bi_map_5_all, 0.00, 0.5, 1, 0.5) + 
   draw_plot(bi_map_5_conv, 0.00, 0.0, 1, 0.5) + 
-  draw_plot(legend, 0.06, 0.4, 0.2, 0.2) +
-
-annotation_scale(location = "bl", width_hint = 0.4,
-                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
-  annotation_north_arrow(location = "t", which_north = "true", 
-        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
-        style = north_arrow_fancy_orienteering)
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
 ```
@@
           fill = NA,
-          color = "lightgrey") +  bi_scale_fill(pal = "GrPink", 
+          color = "lightgrey") +
+  ggtitle("All stations") +
+  bi_scale_fill(pal = "GrPink", 
                 dim = 3) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering) +
   theme(panel.background = element_rect(fill = "white"),
@@
         axis.ticks = element_blank(),
-        plot.margin = unit(c(0, 0, 0, 0), "cm"))
-```
-
-```{r figure-bi-map-threshold-10, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): with equity stations (top panel) and without equity stations (bottom panel)"}
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-bi-map-threshold-10, echo = FALSE, out.width="1.2\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): with equity stations (top panel) and without equity stations (bottom panel)"}
 ggdraw() + draw_plot(bi_map_10_all, 0.00, 0.5, 1, 0.5) + 
   draw_plot(bi_map_10_conv, 0.00, 0.0, 1, 0.5) + 
-  draw_plot(legend, 0.06, 0.4, 0.2, 0.2) +
-
-annotation_scale(location = "bl", width_hint = 0.4,
-                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
-  annotation_north_arrow(location = "t", which_north = "true", 
-        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
-        style = north_arrow_fancy_orienteering)
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
 ```
@@
         axis.ticks = element_blank(),
-        plot.margin = unit(c(0, 0, 0, 0), "cm"))
-```
-
-```{r figure-bi-map-threshold-15, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): with equity stations (top panel) and without equity stations (bottom panel)"}
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-bi-map-threshold-15, echo = FALSE, out.width="1.2\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): with equity stations (top panel) and without equity stations (bottom panel)"}
 ggdraw() + draw_plot(bi_map_15_all, 0.00, 0.5, 1, 0.5) + 
   draw_plot(bi_map_15_conv, 0.00, 0.0, 1, 0.5) + 
-  draw_plot(legend, 0.06, 0.4, 0.2, 0.2) +
-
-annotation_scale(location = "bl", width_hint = 0.4,
-                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
-  annotation_north_arrow(location = "t", which_north = "true", 
-        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
-        style = north_arrow_fancy_orienteering)
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
 ```
@@
 Using disaggregated population data, we examined equity in accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
-Using a BFCA approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro scale which better reflects the friction [*is this the right word?*] of walking thresholds and avoids the "absorption of disparities", as articulated by Chen et al.
+Using a BFCA approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro scale which better scale at which walking takes place and avoids the "absorption of disparities", as articulated by Chen et al.
 [-@chenExploringEquityPerformance2019] and elsewhere in the transportation literature [@rowangouldIdentifyingEnvironmentalJustice2016].

### 2021-06-21 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Edits after Elise revised the paper

@@
 
-  \journal{Transportation Research Part D: Transport and Environment} % Sets Journal name
+  \journal{Transportation Research Part D: Transport and
+Environment} % Sets Journal name
 
@@
 \bibliographystyle{elsarticle-harv}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
 \ifxetex
@@
 % Pandoc citation processing
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc} % for calculating minipage widths
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
-  \title{Examining equity in accessibility to bike share - a balanced floating
-catchment area approach}
+  \title{Examining equity in accessibility to bike share - a balanced
+floating catchment area approach}
     \author[McMaster University]{Elise Desjardins\corref{1}}
@@
    \ead{paezha@mcmaster.ca} 
-      \address[McMaster University]{School of Earth, Environment \& Society, McMaster University, 1280 Main
-Street West, Hamilton, ON L8S4L8}
-    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
-1265 Military Trail, Toronto, ON M1C1A4}
+      \address[McMaster University]{School of Earth, Environment \&
+Society, McMaster University, 1280 Main Street West, Hamilton, ON
+L8S4L8}
+    \address[University of Toronto Scarborough]{Department of Geography
+\& Planning, University of Toronto Scarborough, 1265 Military Trail,
+Toronto, ON M1C1A4}
       \cortext[1]{Corresponding Author}
@@
   accessibility to stations using a balanced floating catchment area
-  approach and compare accessibility with and without the equity stations.
-  We analyze micro zones to better reflect walking to a station and
-  conduct a sensitivity analysis at several walking time thresholds. We
-  then reaggregate the estimated accessibility for further analysis using
-  census data. Our findings indicate that equity stations increased
-  accessibility for the serviced population at every threshold examined,
-  but the increase was relatively modest especially for population in the
-  bottom 20\% of median total household income.
+  approach and compare accessibility with and without the equity
+  stations. We analyze micro zones to better reflect walking to a
+  station and conduct a sensitivity analysis at several walking time
+  thresholds. We then reaggregate the estimated accessibility for
+  further analysis using census data. Our findings indicate that equity
+  stations increased accessibility for the serviced population at every
+  threshold examined, but the increase was relatively modest especially
+  for population in the bottom 20\% of median total household income.
   \end{abstract}
@@
  \end{frontmatter}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{rm}\NormalTok{(}\DataTypeTok{list =} \KeywordTok{ls}\NormalTok{())}
-\end{Highlighting}
-\end{Shaded}
 
@@
 bicycling levels is but one of many reasons for implementing such
-programs in urban areas (see Hosford et al., 2018, 2019). As a healthy,
+programs in urban areas (Hosford et al., 2019, see 2018). As a healthy,
 inexpensive, and convenient form of public transportation, shared
@@
 (Brey et al., 2017; Nickkar et al., 2019; Ogilvie and Goodman, 2012;
-Reilly, S. M. Wang, et al., 2020; Winters et al., 2019) as do younger
-age cohorts (Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011).
-However, one study found that bike share users in Washington, DC were
-more likely to be female (Buck et al., 2013), which suggests that the
-gender gap among bicyclists who use PBSPs is less disparate than the gap
-for personal bicycle use (Fishman, 2016). There is some evidence that
-bike share users are less likely to own a car (Buck et al., 2013;
-Reilly, Noyes, et al., 2020). However, the relationship between income
-or education and bike share use is less clear-cut. Stations in
-disadvantaged communities in Chicago have been found to generate most of
-the average annual trips (Qian and Jaller, 2020) and individuals from
-minority or lower socioeconomic status neighborhoods in
-Minneapolis-St.~Paul also used the city's PBSP more (J. Wang and
-Lindsey, 2019a). Similar findings were reported in London (Ogilvie and
-Goodman, 2012). Being university educated was a significant correlate of
-bike share use in Montreal, Canada (Fuller et al., 2011). Not
-coincidentally, financial savings have been found to motivate those on a
-low income to use bike share (Fishman, 2016).
+Reilly et al., 2020b; Winters et al., 2019) as do younger age cohorts
+(Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011). However,
+one study found that bike share users in Washington, DC were more likely
+to be female (Buck et al., 2013), which suggests that the gender gap
+among bicyclists who use PBSPs is less disparate than the gap for
+personal bicycle use (Fishman, 2016). There is some evidence that bike
+share users are less likely to own a car (Buck et al., 2013; Reilly et
+al., 2020a). However, the relationship between income or education and
+bike share use is less clear-cut. Stations in disadvantaged communities
+in Chicago have been found to generate most of the average annual trips
+(Qian and Jaller, 2020) and individuals from minority or lower
+socioeconomic status neighborhoods in Minneapolis-St.~Paul also used the
+city's PBSP more (Wang and Lindsey, 2019a). Similar findings were
+reported in London (Ogilvie and Goodman, 2012). Being university
+educated was a significant correlate of bike share use in Montreal,
+Canada (Fuller et al., 2011). Not coincidentally, financial savings have
+been found to motivate those on a low income to use bike share (Fishman,
+2016).
 
@@
 with higher incomes and more residents of higher education had more
-bikes (Mooney et al., 2019). Babagoli et al.~(2019) also found that
+bikes (Mooney et al., 2019). Babagoli et al. (2019) also found that
 neighborhoods in New York City with higher affluence had the greatest
@@
 
-Chen et al.~(2019) distinguish two types of equity related to bike
+Chen et al. (2019) distinguish two types of equity related to bike
 share, horizontal and vertical, based on the work of Delbosc and Currie
@@
 spatial patterns. Bike share equity programs, as identified by McNeil et
-al.~(2019), can be considered efforts to improve vertical equity since
+al. (2019), can be considered efforts to improve vertical equity since
 they favour groups that have benefited less from PBSPs.
@@
 docking station decreases the potential of accessing the program, the
-location of stations matters. Kabra et al.~(2020) found that the
+location of stations matters. Kabra et al. (2020) found that the
 majority of bike share usage in Paris comes from areas within 300 m of
@@
 neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
-2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
+2019; Qian and Jaller, 2020; Wang and Lindsey, 2019a), although there
 are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
-also true for studies examining correlates of bicycle share demand (J.
-Wang and Lindsey, 2019b). Handy and Niemeier (1997) note that using
+also true for studies examining correlates of bicycle share demand (Wang
+and Lindsey, 2019b). Handy and Niemeier (1997) note that using
 disaggregated data in accessibility analyses provides a more accurate
-estimate for individuals. Chen et al.~(2019, p. 530) are in favour of
+estimate for individuals. Chen et al. (2019, p. 530) are in favour of
 using disaggregated data, which they did in their recent analysis of
@@
 hinder our understanding of the equity impacts since individual
-disparities are absorbed after aggregation''.
+disparities are absorbed after aggregation.''
 
@@
 neighborhoods are located near the city center and wealthier
-neighborhoods are in the surrounding suburban areas''. Therefore, the
+neighborhoods are in the surrounding suburban areas.'' Therefore, the
 core service area for the PBSP in Hamilton by default covers more of the
@@
 
-\begin{verbatim}
-## Source : http://tile.stamen.com/toner-lite/8/70/93.png
-\end{verbatim}
-
-\begin{verbatim}
-## Source : http://tile.stamen.com/toner-lite/8/71/93.png
-\end{verbatim}
-
-\begin{verbatim}
-## Source : http://tile.stamen.com/toner-lite/8/70/94.png
-\end{verbatim}
-
-\begin{verbatim}
-## Source : http://tile.stamen.com/toner-lite/8/71/94.png
-\end{verbatim}
-
-\begin{verbatim}
-## Coordinate system already present. Adding new coordinate system, which will replace the existing one.
-\end{verbatim}
-
 \begin{figure}
@@
 inflation that result from the overlapping catchment areas produced by
-earlier FCA methods (Paez et al., 2019). Paez et al.~(2019) adjusted the
+earlier FCA methods (Paez et al., 2019). Paez et al. (2019) adjusted the
 impedance weights so that both supply and demand are proportionally
@@
 By allocating the population and level of service proportionally, this
-method preserves the values of the population and level of service and
-provides a regional provider-to-population ratio since: \begin{equation}
+method preserves the values of the population and level of service
+since: \begin{equation}
 \label{eq:proportionality}
-\sum_{i=1}^n A_i = \sum_{j=1}^J L_j = \frac{\sum_{j=1}^JS_j}{\sum_{i=1}^n P_i}
+\sum_{i=1}^n A_i = \sum_{j=1}^J L_j
 \end{equation}
@@
 All data for this research were accessed from publicly available Census
-of Canada sources, from OpenStreetMaps, and from Open Hamilton\footnote{https://open.hamilton.ca/},
-a public online repository of data curated by the City of Hamilton.
+of Canada sources, from OpenStreetMaps, and from Open
+Hamilton\footnote{https://open.hamilton.ca/}, a public online repository
+of data curated by the City of Hamilton.
 
@@
 
-With a walking distance of three minutes, we found that there are 25.2
-bicycle racks per person system-wide in the original system
-configuration (i.e., without equity stations). The addition of equity
-stations increases this ratio slightly to 25.4 bicycles per person.
-Accessibility is fairly uniform overall, with the exception of two small
-areas where accessibility is slightly higher. This high level of
-system-wide accessibility occurs because the population that can reach a
-docking station when travel time is three minutes or less is very
-limited, and accessibility is strongly shaped by a few locations that
-concentrate population and stations. For this reason, accessibility is
-highly concentrated in small areas around those stations. The map is not
-shown, since it is less informative, but can be recreated using the
-source code and data.
+With a walking distance of three minutes, we found that the total level
+of service is 25.2 bicycle racks per person system-wide in the original
+system configuration (i.e., without equity stations). The addition of
+equity stations increases this slightly to 25.4 bicycles per person.
+This total level of service is allocated to the population to obtain the
+levels of accessibility, which turn out to be relatively uniform
+overall, with the exception of two small areas where accessibility is
+slightly higher. This high level of system-wide accessibility occurs
+because the population that can reach a docking station when travel time
+is three minutes or less is very limited, and accessibility is strongly
+shaped by a few locations that concentrate population and stations. For
+this reason, accessibility is highly concentrated in small areas around
+those stations. The map is not shown, since it is less informative, but
+can be recreated using the source code and data.
 
@@
 
-\begin{verbatim}
-## Using plotunit = 'm'
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in f(...): True north is not meaningful without coord_sf()
-\end{verbatim}
-
 \begin{figure}
-\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-3}
+\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-3}
 \end{figure}
 
-\begin{verbatim}
-## Using plotunit = 'm'
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in f(...): True north is not meaningful without coord_sf()
-\end{verbatim}
-
 \begin{figure}
-\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-5}
+\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-5}
 \end{figure}
 
-\begin{verbatim}
-## Using plotunit = 'm'
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in f(...): True north is not meaningful without coord_sf()
-\end{verbatim}
-
 \begin{figure}
-\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-10}
+\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-10}
 \end{figure}
 
-\begin{verbatim}
-## Using plotunit = 'm'
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in f(...): True north is not meaningful without coord_sf()
-\end{verbatim}
-
 \begin{figure}
-\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-15}
+\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-15}
 \end{figure}
@@
 combined with pycnophylactic interpolation, enabled us to measure
-accessibility on a micro scale which better reflects the friction
-{[}\emph{is this the right word?}{]} of walking thresholds and avoids
-the ``absorption of disparities'', as articulated by Chen et al.~(2019)
-and elsewhere in the transportation literature (Rowangould et al.,
-2016). The BFCA method also considers demand and supply in the
-calculation of accessibility, which is important for services like PBSPs
-but not accounted for in other common accessibility models. Unlike the
-2SFCA method, there was no inflation or deflation in our accessibility
-estimates because population and level of service were preserved (see
-Figure \ref{fig:figure-10} and Figure \ref{fig:figure-11}). This
-differentiates our analysis from similar papers exploring equity in
-PBSPs that use larger geographical units of analysis or that focus on
-station location instead of level of service. In this way, our paper has
-made an important contribution by applying an intuitive and useful
-approach to measure accessibility to a PBSP. This open and reproducible
-research could be replicated for additional transportation and
-statistical analysis using other individual-level data in Hamilton or
-revised for PBSPs in other North American cities.
+accessibility on a micro scale which better scale at which walking takes
+place and avoids the ``absorption of disparities,'' as articulated by
+Chen et al. (2019) and elsewhere in the transportation literature
+(Rowangould et al., 2016). The BFCA method also considers demand and
+supply in the calculation of accessibility, which is important for
+services like PBSPs but not accounted for in other common accessibility
+models. Unlike the 2SFCA method, there was no inflation or deflation in
+our accessibility estimates because population and level of service were
+preserved (see Figure \ref{fig:figure-10} and Figure
+\ref{fig:figure-11}). This differentiates our analysis from similar
+papers exploring equity in PBSPs that use larger geographical units of
+analysis or that focus on station location instead of level of service.
+In this way, our paper has made an important contribution by applying an
+intuitive and useful approach to measure accessibility to a PBSP. This
+open and reproducible research could be replicated for additional
+transportation and statistical analysis using other individual-level
+data in Hamilton or revised for PBSPs in other North American cities.
 
@@
 eight minutes whereby congestion effects due to increased potential
-demand kick in. Chen et al.~(2019) also found that accessibility
+demand kick in. Chen et al. (2019) also found that accessibility
 decreases as larger geographical units are used in the estimation, since
@@
 Several studies have examined who can access PBSPs based on the location
-of stations (Babagoli et al., 2019; Qian and Jaller, 2020; Qian et al.,
-2020; J. Wang and Lindsey, 2019b), but the potential supply of bicycles
-at stations in disadvantaged communities is not considered in the
+of stations (Babagoli et al., 2019; Qian et al., 2020; Qian and Jaller,
+2020; Wang and Lindsey, 2019b), but the potential supply of bicycles at
+stations in disadvantaged communities is not considered in the
 literature. The size of stations (i.e., potential supply of bicycles)
@@
 supply of bicycles available at nearby stations is insufficient to meet
-demand. Babgoli et al.~(2019) found a slight but not statistically
+demand. Babgoli et al. (2019) found a slight but not statistically
 significant increase in the proportion of neighborhoods with the highest
@@
 Finally, distance to stations is associated with bike share use (Fuller
-et al., 2011; J. Wang and Lindsey, 2019b), especially when the trip
-starts at the residence (Fishman et al., 2013), and can be a barrier to
-using PBSPs (Fishman et al., 2014). To address this issue,
-transportation planners and PBSP operators should consult with groups
-and communities who are under-represented among bike share users to
-determine how far they are willing to walk to reach a docking station
-and to get their input on the location of new docking stations.
+et al., 2011; Wang and Lindsey, 2019b), especially when the trip starts
+at the residence (Fishman et al., 2013), and can be a barrier to using
+PBSPs (Fishman et al., 2014). To address this issue, transportation
+planners and PBSP operators should consult with groups and communities
+who are under-represented among bike share users to determine how far
+they are willing to walk to reach a docking station and to get their
+input on the location of new docking stations.
 
@@
 among users from disadvantaged neighborhoods (Caspi and Noland, 2019;
-Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), but our analysis is
+Qian and Jaller, 2020; Wang and Lindsey, 2019a), but our analysis is
 limited by the lack of publicly available route and individual user data
@@
 stations, which would extend existing studies conducted by Scott and
-colleagues (Lu et al., 2018; Scott and Ciuro, 2019; Scott et al., 2021).
+colleagues (Lu et al., 2018; Scott et al., 2021; Scott and Ciuro, 2019).
 This paper, combined with additional studies such as those
@@
 2021), \texttt{rticles} (Allaire et al., 2021), \texttt{sf} (Pebesma,
-2021), \texttt{tidyverse} (Wickham, 2021), \texttt{tinytex} (Xie,
+2021), \texttt{tidyverse} (Wickham, 2019), \texttt{tinytex} (Xie,
 2021b), \texttt{units} (Pebesma et al., 2021).
@@
 \hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
 \leavevmode\hypertarget{ref-R-rticles}{}%
@@
 Fillmore, C.E., Setubal, C., Bettigole, C., 2020. Design and baseline
-description of a cohort of bikeshare users in the city of Philadelphia.
-Journal of Transport \& Health 16, 100836.
+description of a cohort of bikeshare users in the city of
+{Philadelphia}. Journal of Transport \& Health 16, 100836.
 doi:\href{https://doi.org/10.1016/j.jth.2020.100836}{10.1016/j.jth.2020.100836}
@@
 \leavevmode\hypertarget{ref-R-gridExtra}{}%
-Auguie, B., 2017. GridExtra: Miscellaneous functions for "grid"
+Auguie, B., 2017. gridExtra: Miscellaneous functions for "grid"
 graphics.
@@
 Babagoli, M.A., Kaufman, T.K., Noyes, P., Sheffield, P.E., 2019.
-Exploring the health and spatial equity implications of the New York
-City Bike share system. Journal of Transport \& Health 13, 200--209.
+Exploring the health and spatial equity implications of the {New York
+City Bike} share system. Journal of Transport \& Health 13, 200--209.
 doi:\href{https://doi.org/10.1016/j.jth.2019.04.003}{10.1016/j.jth.2019.04.003}
@@
 \leavevmode\hypertarget{ref-breyWantRideMy2017}{}%
-Brey, R., Castillo-Manzano, J.I., Castro-Nuño, M., 2017. ``I want to
-ride my bicycle'': Delimiting cyclist typologies. Applied Economics
+Brey, R., Castillo-Manzano, J.I., Castro-Nuño, M., 2017. {`{I} want to
+ride my bicycle'}: Delimiting cyclist typologies. Applied Economics
 Letters 24, 549--552.
@@
 Buck, D., Buehler, R., Happ, P., Rawls, B., Chung, P., Borecki, N.,
-2013. Are Bikeshare Users Different from Regular Cyclists?: A First Look
-at Short-Term Users, Annual Members, and Area Cyclists in the
-Washington, D.C., Region. Transportation Research Record 2387, 112--119.
+2013. Are {Bikeshare Users Different} from {Regular Cyclists}?: {A First
+Look} at {Short}-{Term Users}, {Annual Members}, and {Area Cyclists} in
+the {Washington}, {D}.{C}., {Region}. Transportation Research Record
+2387, 112--119.
 doi:\href{https://doi.org/10.3141/2387-13}{10.3141/2387-13}
@@
 \leavevmode\hypertarget{ref-caspiBikesharingPhiladelphiaLowerincome2019}{}%
-Caspi, O., Noland, R.B., 2019. Bikesharing in Philadelphia: Do
+Caspi, O., Noland, R.B., 2019. Bikesharing in {Philadelphia}: {Do}
 lower-income areas generate trips? Travel Behaviour and Society 16,
@@
 Chen, W., Cheng, L., Chen, X., Chen, J., Cao, M., 2021. Measuring
-accessibility to health care services for older bus passengers: A finer
-spatial resolution. Journal of Transport Geography 93, 103068.
+accessibility to health care services for older bus passengers: {A}
+finer spatial resolution. Journal of Transport Geography 93, 103068.
 doi:\href{https://doi.org/10.1016/j.jtrangeo.2021.103068}{10.1016/j.jtrangeo.2021.103068}
@@
 Chen, Z., Guo, Y., Stuart, A.L., Zhang, Y., Li, X., 2019. Exploring the
-equity performance of bike-sharing systems with disaggregated data: A
-story of southern Tampa. Transportation Research Part A: Policy and
+equity performance of bike-sharing systems with disaggregated data: {A}
+story of southern {Tampa}. Transportation Research Part A: Policy and
 Practice 130, 529--545.
@@
 Chen, Z., Li, X., 2021. Unobserved heterogeneity in transportation
-equity analysis: Evidence from a bike-sharing system in southern Tampa.
-Journal of Transport Geography 91, 102956.
+equity analysis: {Evidence} from a bike-sharing system in southern
+{Tampa}. Journal of Transport Geography 91, 102956.
 doi:\href{https://doi.org/10.1016/j.jtrangeo.2021.102956}{10.1016/j.jtrangeo.2021.102956}
@@
 Delamater, P.L., 2013. Spatial accessibility in suboptimally configured
-health care systems: A modified two-step floating catchment area
-(M2SFCA) metric. Health \& Place 24, 30--43.
+health care systems: {A} modified two-step floating catchment area
+({M2SFCA}) metric. Health \& Place 24, 30--43.
 doi:\href{https://doi.org/10.1016/j.healthplace.2013.07.012}{10.1016/j.healthplace.2013.07.012}
@@
 \leavevmode\hypertarget{ref-delboscUsingLorenzCurves2011}{}%
-Delbosc, A., Currie, G., 2011. Using Lorenz curves to assess public
+Delbosc, A., Currie, G., 2011. Using {Lorenz} curves to assess public
 transport equity. Journal of Transport Geography, Special section on
-Alternative Travel futures 19, 1252--1259.
+{Alternative Travel} futures 19, 1252--1259.
 doi:\href{https://doi.org/10.1016/j.jtrangeo.2011.02.008}{10.1016/j.jtrangeo.2011.02.008}
@@
 \leavevmode\hypertarget{ref-fishmanBikeshareReviewRecent2016}{}%
-Fishman, E., 2016. Bikeshare: A Review of Recent Literature. Transport
-Reviews 36, 92--113.
+Fishman, E., 2016. Bikeshare: {A Review} of {Recent Literature}.
+Transport Reviews 36, 92--113.
 doi:\href{https://doi.org/10.1080/01441647.2015.1033036}{10.1080/01441647.2015.1033036}
@@
 \leavevmode\hypertarget{ref-fishmanBikeShareSynthesis2013}{}%
-Fishman, E., Washington, S., Haworth, N., 2013. Bike Share: A Synthesis
-of the Literature. Transport Reviews 33, 148--165.
+Fishman, E., Washington, S., Haworth, N., 2013. Bike {Share}: {A
+Synthesis} of the {Literature}. Transport Reviews 33, 148--165.
 doi:\href{https://doi.org/10.1080/01441647.2013.775612}{10.1080/01441647.2013.775612}
@@
 Fishman, E., Washington, S., Haworth, N., Mazzei, A., 2014. Barriers to
-bikesharing: An analysis from Melbourne and Brisbane. Journal of
+bikesharing: An analysis from {Melbourne} and {Brisbane}. Journal of
 Transport Geography 41, 325--337.
@@
 Fuller, D., Gauvin, L., Kestens, Y., Daniel, M., Fournier, M., Morency,
-P., Drouin, L., 2011. Use of a New Public Bicycle Share Program in
-Montreal, Canada. American Journal of Preventive Medicine 41, 80--83.
+P., Drouin, L., 2011. Use of a {New Public Bicycle Share Program} in
+{Montreal}, {Canada}. American Journal of Preventive Medicine 41,
+80--83.
 doi:\href{https://doi.org/10.1016/j.amepre.2011.03.002}{10.1016/j.amepre.2011.03.002}
@@
 Hamilton, C. of, 2015a. Public bike share transit system implementation
-plan (pw13015c) - (city wide) {[}WWW Document{]}. URL
+plan (PW13015c) - (city wide) {[}WWW Document{]}. URL
 \url{https://pub-hamilton.escribemeetings.com/filestream.ashx?DocumentId=118356}
@@
 \leavevmode\hypertarget{ref-hamiltonHamiltonBikeShare2015}{}%
-Hamilton, C. of, 2015b. Hamilton Bike Share.
+Hamilton, C. of, 2015b. Hamilton {Bike Share}.
 
 \leavevmode\hypertarget{ref-handyMeasuringAccessibilityExploration1997}{}%
-Handy, S.L., Niemeier, D.A., 1997. Measuring Accessibility: An
-Exploration of Issues and Alternatives. Environment and Planning A:
+Handy, S.L., Niemeier, D.A., 1997. Measuring {Accessibility}: {An
+Exploration} of {Issues} and {Alternatives}. Environment and Planning A:
 Economy and Space 29, 1175--1194.
@@
 M., Winters, M., 2018. Evaluation of the impact of a public bicycle
-share program on population bicycling in Vancouver, BC. Preventive
+share program on population bicycling in {Vancouver}, {BC}. Preventive
 Medicine Reports 12, 176--181.
@@
 \leavevmode\hypertarget{ref-hosfordWhoArePublic2018}{}%
-Hosford, K., Winters, M., 2018. Who Are Public Bicycle Share Programs
-Serving? An Evaluation of the Equity of Spatial Access to Bicycle Share
-Service Areas in Canadian Cities. Transportation Research Record 2672,
-42--50.
+Hosford, K., Winters, M., 2018. Who {Are Public Bicycle Share Programs
+Serving}? {An Evaluation} of the {Equity} of {Spatial Access} to
+{Bicycle Share Service Areas} in {Canadian Cities}. Transportation
+Research Record 2672, 42--50.
 doi:\href{https://doi.org/10.1177/0361198118783107}{10.1177/0361198118783107}
@@
 S.M., Fuller, D., 2019. Evaluating the impact of implementing public
-bicycle share programs on cycling: The International Bikeshare Impacts
-on Cycling and Collisions Study (IBICCS). International Journal of
+bicycle share programs on cycling: The {International Bikeshare Impacts}
+on {Cycling} and {Collisions Study} ({IBICCS}). International Journal of
 Behavioral Nutrition and Physical Activity 16, 107.
@@
 Howland, S., McNeil, N., Broach, J., Rankins, K., MacArthur, J., Dill,
-J., 2017. Current Efforts to Make Bikeshare More Equitable: Survey of
-System Owners and Operators. Transportation Research Record 2662,
+J., 2017. Current {Efforts} to {Make Bikeshare More Equitable}: {Survey}
+of {System Owners} and {Operators}. Transportation Research Record 2662,
 160--167. doi:\href{https://doi.org/10.3141/2662-18}{10.3141/2662-18}
@@
 \leavevmode\hypertarget{ref-hullgrassoBikeShareEquity2020}{}%
-Hull Grasso, S., Barnes, P., Chavis, C., 2020. Bike Share Equity for
-Underrepresented Groups: Analyzing Barriers to System Usage in
-Baltimore, Maryland. Sustainability 12, 7600.
+Hull Grasso, S., Barnes, P., Chavis, C., 2020. Bike {Share Equity} for
+{Underrepresented Groups}: {Analyzing Barriers} to {System Usage} in
+{Baltimore}, {Maryland}. Sustainability 12, 7600.
 doi:\href{https://doi.org/10.3390/su12187600}{10.3390/su12187600}
@@
 \leavevmode\hypertarget{ref-kabraBikeShareSystemsAccessibility2020}{}%
-Kabra, A., Belavina, E., Girotra, K., 2020. Bike-Share Systems:
-Accessibility and Availability. Management Science 66, 3803--3824.
+Kabra, A., Belavina, E., Girotra, K., 2020. Bike-{Share Systems}:
+{Accessibility} and {Availability}. Management Science 66, 3803--3824.
 doi:\href{https://doi.org/10.1287/mnsc.2019.3407}{10.1287/mnsc.2019.3407}
@@
 \leavevmode\hypertarget{ref-kwanSpaceTimeIntegral1998}{}%
-Kwan, M.-P., 1998. Space-Time and Integral Measures of Individual
-Accessibility: A Comparative Analysis Using a Point-based Framework.
-Geographical Analysis 30, 191--216.
+Kwan, M.-P., 1998. Space-{Time} and {Integral Measures} of {Individual
+Accessibility}: {A Comparative Analysis Using} a {Point}-based
+{Framework}. Geographical Analysis 30, 191--216.
 doi:\href{https://doi.org/10.1111/j.1538-4632.1998.tb00396.x}{10.1111/j.1538-4632.1998.tb00396.x}
@@
 \leavevmode\hypertarget{ref-larsenQuarterMileReexamining2010}{}%
-Larsen, J., El-Geneidy, A., Yasmin, F., 2010. Beyond the Quarter Mile:
-Re-examining Travel Distances by Active Transportation. Canadian Journal
-of Urban Research 19, 70--88.
+Larsen, J., El-Geneidy, A., Yasmin, F., 2010. Beyond the {Quarter Mile}:
+{Re}-examining {Travel Distances} by {Active Transportation}. Canadian
+Journal of Urban Research 19, 70--88.
 
@@
 Lu, W., Scott, D.M., Dalumpines, R., 2018. Understanding bike share
-cyclist route choice using GPS data: Comparing dominant routes and
+cyclist route choice using {GPS} data: {Comparing} dominant routes and
 shortest paths. Journal of Transport Geography 71, 172--181.
@@
 Luo, W., Qi, Y., 2009. An enhanced two-step floating catchment area
-(E2SFCA) method for measuring spatial accessibility to primary care
+({E2SFCA}) method for measuring spatial accessibility to primary care
 physicians. Health \& Place 15, 1100--1107.
@@
 \leavevmode\hypertarget{ref-luoMeasuresSpatialAccessibility2003}{}%
-Luo, W., Wang, F., 2003. Measures of Spatial Accessibility to Health
-Care in a GIS Environment: Synthesis and a Case Study in the Chicago
-Region. Environment and Planning B: Planning and Design 30, 865--884.
-doi:\href{https://doi.org/10.1068/b29120}{10.1068/b29120}
+Luo, W., Wang, F., 2003. Measures of {Spatial Accessibility} to {Health
+Care} in a {GIS Environment}: {Synthesis} and a {Case Study} in the
+{Chicago Region}. Environment and Planning B: Planning and Design 30,
+865--884. doi:\href{https://doi.org/10.1068/b29120}{10.1068/b29120}
 
 \leavevmode\hypertarget{ref-macarthurAdaptiveBikeShare2020}{}%
-MacArthur, J., McNeil, N., Cummings, A., Broach, J., 2020. Adaptive Bike
-Share: Expanding Bike Share to People with Disabilities and Older
-Adults. Transportation Research Record 2674, 556--565.
+MacArthur, J., McNeil, N., Cummings, A., Broach, J., 2020. Adaptive
+{Bike Share}: {Expanding Bike Share} to {People} with {Disabilities} and
+{Older Adults}. Transportation Research Record 2674, 556--565.
 doi:\href{https://doi.org/10.1177/0361198120925079}{10.1177/0361198120925079}
@@
 Mooney, S.J., Hosford, K., Howe, B., Yan, A., Winters, M., Bassok, A.,
-Hirsch, J.A., 2019. Freedom from the station: Spatial equity in access
+Hirsch, J.A., 2019. Freedom from the station: {Spatial} equity in access
 to dockless bike share. Journal of Transport Geography 74, 91--96.
@@
 spatial-temporal gender and land use analysis of bikeshare ridership:
-The case study of Baltimore City. City, Culture and Society 18, 100291.
+{The} case study of {Baltimore City}. City, Culture and Society 18,
+100291.
 doi:\href{https://doi.org/10.1016/j.ccs.2019.100291}{10.1016/j.ccs.2019.100291}
+
+\leavevmode\hypertarget{ref-okelly2003aggregate}{}%
+O'Kelly, M.E., Horner, M.W., 2003. Aggregate accessibility to population
+at the county level: U.s. 1940-2000. Journal of Geographical Systems 5,
+5--23.
 
@@
 Ogilvie, F., Goodman, A., 2012. Inequalities in usage of a public
-bicycle sharing scheme: Socio-demographic predictors of uptake and usage
-of the London (UK) cycle hire scheme. Preventive Medicine 55, 40--45.
+bicycle sharing scheme: {Socio}-demographic predictors of uptake and
+usage of the {London} ({UK}) cycle hire scheme. Preventive Medicine 55,
+40--45.
 doi:\href{https://doi.org/10.1016/j.ypmed.2012.05.002}{10.1016/j.ypmed.2012.05.002}
-
-\leavevmode\hypertarget{ref-okelly2003aggregate}{}%
-O'Kelly, M.E., Horner, M.W., 2003. Aggregate accessibility to population
-at the county level: U.S. 1940-2000. Journal of Geographical Systems 5,
-5--23.
 
@@
 Paez, A., Higgins, C.D., Vivona, S.F., 2019. Demand and level of service
-inflation in Floating Catchment Area (FCA) methods. PLoS ONE 14.
+inflation in {Floating Catchment Area} ({FCA}) methods. PLoS ONE 14.
 doi:\href{https://doi.org/10.1371/journal.pone.0218773}{10.1371/journal.pone.0218773}
@@
 Positive and normative implementations of various accessibility
-indicators. Journal of Transport Geography, Special Section on
-Accessibility and Socio-Economic Activities: Methodological and
-Empirical Aspects 25, 141--153.
+indicators. Journal of Transport Geography, Special {section} on
+{Accessibility} and {Socio}-{Economic Activities}: {Methodological} and
+{Empirical Aspects} 25, 141--153.
 doi:\href{https://doi.org/10.1016/j.jtrangeo.2012.03.016}{10.1016/j.jtrangeo.2012.03.016}
@@
 Pereira, R.H.M., Saraiva, M., Herszenhut, D., Braga, C.K.V., Conway,
-M.W., 2021. R5r: Rapid realistic routing on multimodal transport
+M.W., 2021. r5r: Rapid realistic routing on multimodal transport
 networks with r\textsuperscript{5} in r. Findings.
@@
 Qian, X., Jaller, M., 2020. Bikesharing, equity, and disadvantaged
-communities: A case study in Chicago. Transportation Research Part A:
-Policy and Practice 140, 354--371.
+communities: {A} case study in {Chicago}. Transportation Research Part
+A: Policy and Practice 140, 354--371.
 doi:\href{https://doi.org/10.1016/j.tra.2020.07.004}{10.1016/j.tra.2020.07.004}
@@
 Qian, X., Jaller, M., Niemeier, D., 2020. Enhancing equitable service
-level: Which can address better, dockless or dock-based Bikeshare
+level: {Which} can address better, dockless or dock-based {Bikeshare}
 systems? Journal of Transport Geography 86, 102784.
@@
 \leavevmode\hypertarget{ref-radkeSpatialDecompositionsModeling2000}{}%
-Radke, J., Mu, L., 2000. Spatial Decompositions, Modeling and Mapping
-Service Regions to Predict Access to Social Programs. Geographic
-Information Sciences 6, 105--112.
+Radke, J., Mu, L., 2000. Spatial {Decompositions}, {Modeling} and
+{Mapping Service Regions} to {Predict Access} to {Social Programs}.
+Geographic Information Sciences 6, 105--112.
 doi:\href{https://doi.org/10.1080/10824000009480538}{10.1080/10824000009480538}
@@
 \leavevmode\hypertarget{ref-reillyNoncyclistsFrequentCyclists2020}{}%
-Reilly, K.H., Noyes, P., Crossa, A., 2020. From non-cyclists to frequent
-cyclists: Factors associated with frequent bike share use in New York
-City. Journal of Transport \& Health 16, 100790.
+Reilly, K.H., Noyes, P., Crossa, A., 2020a. From non-cyclists to
+frequent cyclists: {Factors} associated with frequent bike share use in
+{New York City}. Journal of Transport \& Health 16, 100790.
 doi:\href{https://doi.org/10.1016/j.jth.2019.100790}{10.1016/j.jth.2019.100790}
@@
 \leavevmode\hypertarget{ref-reillyGenderDisparitiesNew2020}{}%
-Reilly, K.H., Wang, S.M., Crossa, A., 2020. Gender disparities in New
-York City bike share usage. International Journal of Sustainable
+Reilly, K.H., Wang, S.M., Crossa, A., 2020b. Gender disparities in {New
+York City} bike share usage. International Journal of Sustainable
 Transportation 0, 1--9.
@@
 Saraiva, M., Pereira, R.H.M., Herszenhut, D., Braga, C.K.V., Conway,
-M.W., 2021. R5r: Rapid realistic routing with r5.
+M.W., 2021. r5r: Rapid realistic routing with R5.
 
@@
 Scott, D.M., Ciuro, C., 2019. What factors influence bike share
-ridership? An investigation of Hamilton, Ontario's bike share hubs.
-Travel Behaviour and Society 16, 50--58.
+ridership? {An} investigation of {Hamilton}, {Ontario}'s bike share
+hubs. Travel Behaviour and Society 16, 50--58.
 doi:\href{https://doi.org/10.1016/j.tbs.2019.04.003}{10.1016/j.tbs.2019.04.003}
@@
 Scott, D.M., Lu, W., Brown, M.J., 2021. Route choice of bike share
-users: Leveraging gps data to derive choice sets. Journal of Transport
+users: Leveraging GPS data to derive choice sets. Journal of Transport
 Geography 90, 102903.
@@
 Smith, C.S., Oh, J.-S., Lei, C., 2015. Exploring the equity dimensions
-of us bicycle sharing systems. Western Michigan University.
+of US bicycle sharing systems. Western Michigan University.
 Transportation Research Center for Livable~.
@@
 \leavevmode\hypertarget{ref-toblerSmoothPycnophylacticInterpolation1979}{}%
-Tobler, W.R., 1979. Smooth Pycnophylactic Interpolation for Geographical
-Regions. Journal of the American Statistical Association 74, 519--530.
+Tobler, W.R., 1979. Smooth {Pycnophylactic Interpolation} for
+{Geographical Regions}. Journal of the American Statistical Association
+74, 519--530.
 doi:\href{https://doi.org/10.1080/01621459.1979.10481647}{10.1080/01621459.1979.10481647}
@@
 Wang, J., Lindsey, G., 2019b. Do new bike share stations increase member
-use: A quasi-experimental study. Transportation Research Part A: Policy
-and Practice 121, 1--11.
+use: {A} quasi-experimental study. Transportation Research Part A:
+Policy and Practice 121, 1--11.
 doi:\href{https://doi.org/10.1016/j.tra.2019.01.004}{10.1016/j.tra.2019.01.004}
@@
 \leavevmode\hypertarget{ref-R-tidyverse}{}%
-Wickham, H., 2021. Tidyverse: Easily install and load the tidyverse.
+Wickham, H., 2019. Tidyverse: Easily install and load the tidyverse.
 
@@
 Winters, M., Hosford, K., Javaheri, S., 2019. Who are the
-``super-users'' of public bike share? An analysis of public bike share
-members in Vancouver, BC. Preventive Medicine Reports 15, 100946.
+{`super-users'} of public bike share? {An} analysis of public bike share
+members in {Vancouver}, {BC}. Preventive Medicine Reports 15, 100946.
 doi:\href{https://doi.org/10.1016/j.pmedr.2019.100946}{10.1016/j.pmedr.2019.100946}
@@
 \leavevmode\hypertarget{ref-R-tinytex}{}%
-Xie, Y., 2021b. Tinytex: Helper functions to install and maintain tex
-live, and compile latex documents.
+Xie, Y., 2021b. Tinytex: Helper functions to install and maintain TeX
+live, and compile LaTeX documents.
 
 \leavevmode\hypertarget{ref-R-kableExtra}{}%
-Zhu, H., 2021. KableExtra: Construct complex table with kable and pipe
+Zhu, H., 2021. kableExtra: Construct complex table with kable and pipe
 syntax.
@@
 \leavevmode\hypertarget{ref-R-disk.frame}{}%
-ZJ, D., 2021. Disk.frame: Larger-than-ram disk-based data manipulation
+ZJ, D., 2021. Disk.frame: Larger-than-RAM disk-based data manipulation
 framework.
 
+\end{CSLReferences}
+
 \end{document}

### 2021-08-23 Bike-share-spatial-equity/Cover-Letter.Rmd
Revised cover letter - June 2021

@@
 
-Please find attached our paper submitted to the Virtual Special Issue in Transportation Research Part D: Transport and Environment (SI: Transportation Justice).
-The paper is titled "Examining spatial equity of bike share: a balanced floating catchment area approach".
+Please find attached our revised paper submitted to the Virtual Special Issue in Transportation Research Part D: Transport and Environment (SI: Transportation Justice).
+The paper is titled "Examining equity in accessibility to bike share: a balanced floating catchment area approach".
 We use the city of Hamilton, Ontario in Canada as a case study.
@@
 
-We believe that this paper makes the following contributions to the literature which makes it an excellent inclusion to the special issue on transportation justice:
+We would like to thank the anonymous reviewers for their constructive feedback.
+We believe that the paper is stronger as a result and makes the following contributions to the literature which makes it an excellent inclusion to the special issue on transportation justice:
 
@@
 To the best of our knowledge, the city of Hamilton is the only city in Canada with a PBSP that has an explicit equity program.
-This makes it an interesting place to examine how specific bike share stations added for the purpose of increasing equity have contributed to addressing horizontal and vertical inequities (see Chen et al., 2019).
+This makes it an interesting place to examine how specific docking stations added for the purpose of increasing equity have contributed to addressing vertical inequities (see Chen et al., 2019).
 This case study also builds upon the findings from a recent paper by Hosford & Winters (2018) that explored who is served by PBSPs in Canada.
@@
 2.
-We employ a balanced floating catchment area approach using disaggregate population-level data to measure accessibility at different thresholds: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
+We employ a balanced floating catchment area (BFCA) approach using disaggregate population-level data to measure accessibility at different thresholds: 3 minutes, 5 minutes, 10 minutes, and 15 minutes.
 This method has not been previously used in the cycling literature to measure accessibility.
-Unlike other location- and person-based measures of accessibility, floating catchment area methods incorporate information on capacity and demand in calculating accessibility.
+Unlike other location- and person-based measures of accessibility, floating catchment area (FCA) methods incorporate information on capacity and demand in calculating accessibility.
 These measures have been widely employed in healthcare accessibility research and are typically calculated across two steps.
@@
 Specifically, we use a recent improvement to this approach that was achieved through a simple and intuitive balancing that addressed the effects of demand and service inflation from previous FCA approaches.
-In terms of spatial equity, we compare accessibility with and without the equity stations to assess the effect of the initiative, before reaggregating the data for further analysis using median total household income information from the census.
+We also conducted a comparative analysis with the two-step FCA method to show the benefits of the balanced FCA approach.
+In terms of equity, we compare accessibility with and without the equity stations to assess the effect of the added docking stations, before reaggregating the data for further analysis using median total household income information from the census.
 

### 2021-08-23 Bike-share-spatial-equity/Highlights.Rmd
Final draft of Highlights for submission
Revised Highlights - June 2021

--- 2021-03-26 Bike-share-spatial-equity/Highlights.Rmd
@@
 
-* Balanced floating catchment area methods are applied to measure accessibility to PBSP
+* Balanced floating catchment area method applied to measure accessibility to bikeshare
 
-* Contribution of equity stations to accessibility assessed at different thresholds
+* Contribution of stations added to increase equity assessed at different thresholds
 
-* Equity stations increase accessibility modestly for all income groups
+* Added stations increased accessibility and reduced inequities, albeit only modestly
 
-* Lowest income group has lowest levels of accessibility, even with equity stations
+* Lowest income group has lowest levels of accessibility, even with "equity" stations
 
-* Locations for new stations are identified to increase equitable access to PBSP
+* Location and size of docking stations matters for increasing equity in accessibility

### 2021-08-23 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Major revisions resubmitted June 2021

--- 2021-06-18 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
@@
 ---
-title: Examining equity in accessibility to bike share - a balanced floating catchment area approach 
+title: "Examining equity in accessibility to bike share: a balanced floating catchment area approach" 
 author:
@@
 HBS was launched in 2015 and currently has over 900 operational bicycles and 130 docking stations.
-An equity program, Everyone Rides Initiative, was implemented in 2018 which expanded the program by introducing twelve "equity" stations which expanded HBS to more disadvantaged neighbourhoods in the core service area.
+An equity program, Everyone Rides Initiative, was implemented in 2018 which expanded the program by introducing twelve "equity" stations to more disadvantaged neighbourhoods in the core service area.
 
 This paper examines accessibility to Hamilton Bike Share using a balanced floating catchment area (BFCA) approach.
-We also conduct a comparative analysis to the conventional two-step floating catchment area approach to highlight the benefits of this method which, to our knowledge, has not yet been used in the cycling literature.
-The paper also assesses the contribution of HBS's equity stations to reducing inequities in accessibility for different groups according to median total household income, and provides policy recommendations to further improve equity.
+We also conduct a comparative analysis to the conventional two-step floating catchment area (2SFCA) approach to highlight the benefits of this method which, to our knowledge, has not yet been used in the cycling literature.
+The paper also assesses the contribution of the equity stations to reducing inequities in accessibility for different groups according to median total household income, and provides policy recommendations to further improve equity.
 
@@
 
-2.1. Literature Review
+2. Literature Review
 ===================
 
-## 1.1. Public Bicycle Share Programs
+## 2.1. Public Bicycle Share Programs
 
@@
 
-## 1.2. Equity of PBSPs
+## 2.2. Equity of PBSPs
 
@@
 Offering more people the option of using sustainable and active transportation, particularly those who have lower socioeconomic status and might benefit the most, is a worthy policy goal for cities with PBSPs.
-However, exploring transportation equity by investigating where bicycle share stations are located, often using neighborhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
+However, exploring transportation equity by investigating where bike share stations are located, often using neighborhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
 Meaning that, stations may be lacking in certain neighborhoods but there may be stations accessible within a reasonable walking time.
@@
 The location and size of docking stations are two factors that are relevant to accessibility.
-Since the time or distance needed to reach a docking station decreases the potential of accessing the program, the location of stations matters.
+Since the time or distance needed to reach a docking station decreases the potential of accessing the program, the location matters.
 Kabra et al.
-[-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300 m of stations, which amounts to 2-4 minutes walking by an adult who does not have a disability.
-Similar to other public amenities affected by crowding, the utility of stations is also limited by the _maximum_ number of bicycles that they can hold (e.g., their size).
+[-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris, France comes from areas within 300 m of stations, which amounts to 2-4 minutes walking by an adult who does not have a disability.
+Similar to other public amenities affected by crowding, the utility of docking stations is also limited by the _maximum_ number of bicycles that they can hold (e.g., their size).
 Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts.
-For instance, Wang and Lindsey [-@wangNewBikeShare2019] investigated whether new or relocated bicycle share stations increased accessibility and use, which offered important insights to improve the performance of the program.
+For instance, Wang and Lindsey [-@wangNewBikeShare2019] investigated whether new or relocated bike share stations increased accessibility and use, which offered important insights to improve the performance of the program.
 
 Several approaches have been commonly used for measuring place-based accessibility, including cumulative opportunities, gravity, and utility-based measures [@handyMeasuringAccessibilityExploration1997].
-The gravity-based approach involves weighting destination opportunities, such as the quantity of bicycle share stations, by the time required to reach them from an origin using an impedance function [@handyMeasuringAccessibilityExploration1997; @kwanSpaceTimeIntegral1998].
+The gravity-based approach involves weighting destination opportunities, such as the quantity of bike share stations, by the time required to reach them from an origin using an impedance function [@handyMeasuringAccessibilityExploration1997; @kwanSpaceTimeIntegral1998].
 While such measures are suitable for capturing the potential for reaching destinations from a given location, they do not take demand or congestion effects into account which is an important consideration when calculating accessibility for amenities such as bike share stations.
@@
 When measuring accessibility, researchers have also taken different approaches when it comes to the aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones.
-Previous research on bicycle share equity has typically used a meso- or macro-level approach with aggregated data from entire neighborhoods or census tracts [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021].
-This is also true for studies examining correlates of bicycle share demand [@wangNewBikeShare2019].
+Previous research on bike share equity has typically used a meso- or macro-level approach with aggregated data from entire neighborhoods or census tracts [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021].
+This is also true for studies examining correlates of bike share demand [@wangNewBikeShare2019].
 Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals.
@@
 Hosford and Winters [-@hosfordWhoArePublic2018, pp.
-] acknowledge that "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas".
-Therefore, the core service area for the PBSP in Hamilton by default covers more of the less advantaged areas in the city.
-However, there is also a great deal of variation in income in the city center because of the local university and increasing gentrification.
+47] acknowledge that "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas".
+Therefore, the core service area for the PBSP in Hamilton by default covers more of the disadvantaged areas in the city.
+However, there is also a great deal of variation in income in the core service area because of the local university and increasing gentrification.
 Hosford and Winters [-@hosfordWhoArePublic2018] took a macro-level approach in their analysis by using dissemination areas _across_ the city as the unit of analysis.
@@
 
-Therefore, this paper contributes to the cycling literature by expanding the analysis of Hosford and Winters [-@hosfordWhoArePublic2018] to assess equity in accessibility to HBS and explore the contribution of the "equity" stations added to Hamilton's PBSP.
+Therefore, this paper also contributes to the cycling literature by expanding the analysis of Hosford and Winters [-@hosfordWhoArePublic2018] to assess equity in accessibility to HBS and explore the contribution of the equity stations added to Hamilton's PBSP.
 
@@
 The City of Hamilton undertook a large public engagement campaign to validate the locations of stations that had been selected and to crowdsource potential locations for additional locations [@hamiltonsobi2014].
-Most of the suggested locations were in the east end of the core service area that lack stations or in neighborhoods not serviced by the PBSP.
-The program was enthusiastically welcomed in the city in 2015 - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015], however inadequate coverage given the size of Hamilton Bike Share's service area was identified as a problem early on, and transportation planners noted that the small size (i.e., supply of bicycle racks) and low quantity of stations would lead to challenges in balancing the system [@hamiltonsobi2015].
+Most of the potential locations that were suggested were in the east end of the core service area that lack stations or in neighborhoods not serviced by the PBSP.
+The program was enthusiastically welcomed in the city in 2015 - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015], however inadequate coverage given the size of the program's service area was identified as a problem early on, and transportation planners noted that the small size (i.e., supply of bicycle racks) and low quantity of stations would lead to challenges in balancing the system [@hamiltonsobi2015].
 
@@
 
-Hamilton Bike Share's equity program, Everyone Rides Initiative (ERI), was launched in 2018 with the objective of reducing barriers that may prevent individuals from accessing the program.
-Additional bicycles and twelve "equity" stations were added to the core service area.
+Hamilton Bike Share Inc., the non-profit operator, launched an equity program, Everyone Rides Initiative (ERI), in 2018 with the objective of reducing barriers that may prevent individuals from accessing the program.
+Additional bicycles and twelve "equity" stations were added to the core service area in more disadvantaged neighbourhoods.
 The equity program also offers subsidized memberships to individuals who identify as low income, and complements this service with cycle skills education.
@@
 It would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.
-```{r}
-
-```
-
-```{r hamilton-and-sobi-service-area, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "The core service area of Hamilton Bike Share is outlined in blue.
+
+```{r hamilton-and-sobi-service-area, echo=FALSE, message=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "The core service area of Hamilton Bike Share is outlined in blue.
 Hamilton Census Metropolitan Area is shown in grey."}
@@
 
-```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bicycle share stations in Hamilton, Ontario.
-Conventional stations are in blue and equity (ERI) stations are in orange.
-The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in dark green.
+```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bike share docking stations in Hamilton, Ontario.
+Conventional stations are in purple and equity (ERI) stations are in orange.
+The service area of Hamilton Bike Share is outlined in blue and the city's downtown core is outlined in dark green.
 Hamilton Census Metropolitan Area is shown in grey."}
@@
           fill = NA,
-          color = "black") +
+          color = "blue") +
   geom_sf(data = community_downtown %>%
@@
   scale_fill_manual(name = "Hub Type",
-                    values = c("Conventional" = "blue", "ERI" = "orange")) +
+                    values = c("Conventional" = "purple", "ERI" = "orange")) +
   theme(panel.background = element_rect(fill = "white"),
@@
 
-Since walking trips to a bicycle share station likely happen at a level lower than even the smallest census geography, this requires a more disaggregated approach than the use of census geographies.
+Since walking trips to a docking station likely happen at a level lower than even the smallest census geography, this requires a more disaggregated approach than the use of census geographies.
 For this reason, we implemented our analysis parting from micro population zones to better reflect the friction of walking to a docking station, which is an important component of a bike share trip [@chenExploringEquityPerformance2019].
@@
 When interpolating the population at this high level of resolution, it is important to ensure that population numbers were not allocated to areas where people do not live in Hamilton (for example, to parks or large institutional buildings, etc.).
-To do so, we retrieved shapefiles for various geographic features [see Table \ref{tab:data-features}] from Open Hamilton.
+To do so, we retrieved shapefiles for various geographic features from Open Hamilton.
 Next, we removed these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population within the remaining area [see Figure \ref{fig:interpolated-population}].
 
-```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Interpolated population in the Hamilton Bike Share's core service area (outlined in black) and within 30 minutes of walking to the core service area.
+```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Interpolated population in Hamilton Bike Share's core service area (outlined in blue) and within 30 minutes of walking to the core service area.
 Each population cell is 50-by-50 m in size.
@@
           fill = NA,
-          colour = "black") +
+          colour = "blue") +
   geom_sf(data = community_downtown %>%
@@
 
-To calculate walking times from the centroid of the micro population cells to docking stations, we extracted OpenStreetMap data for Hamilton Bike Share's service area from [BBBike](https://download.bbbike.org/osm/bbbike/), an online cycle route planner that interfaces with OpenStreetMap.
-OpenStreetMap data provides the networks for calculating walking times from each population cell to nearby bicycle share stations, using a maximum walking distance of 10 km and walking time of 30 minutes as thresholds.
-A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the bicycle share stations within the maximum threshold.
+To calculate walking times from the centroid of the micro population cells to docking stations, we extracted OpenStreetMap data for HBS's service area from [BBBike](https://download.bbbike.org/osm/bbbike/), an online cycle route planner that interfaces with OpenStreetMap.
+OpenStreetMap data provides the networks for calculating walking times from each population cell to nearby docking stations, using a maximum walking distance of 10 km and walking time of 30 minutes as thresholds.
+A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the docking stations within the maximum threshold.
 This process provides a more realistic measure of the friction of reaching stations by taking infrastructure into account network travel times, rather than using the Euclidean distance from population cell to station.
@@
 All data for this research were accessed from publicly available Census of Canada sources, from OpenStreetMaps, and from Open Hamilton^[https://open.hamilton.ca/], a public online repository of data curated by the City of Hamilton.
+Median total household income statistics were drawn from the 2016 Canadian Census.
 
@@
 
-Consensus regarding the distance that individuals are willing to walk to access a bike share station is lacking, but the literature on walking behaviour can provide some guidelines to determine the thresholds for our sensitivity analysis.
-Previous studies have found that living within 250 m [@fullerUseNewPublic2011] and 300 m [@kabraBikeShareSystemsAccessibility2020] of a docking station is correlated with bicycle share use, while other research has found that walking trips are less than 600 m and rarely more than 1200 m [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 m [@larsenQuarterMileReexamining2010].
-Hamilton Bike Share will often depict a map at some docking stations to show the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are expected to walk.
-National Association of City Transportation Officials (NACTO) has a similar normative guide [@nactowalkingstation2015].
-
-In the present case, we found that accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at five minutes.
-Accessibility decreases substantially after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each station.
-
-For this reason, we experimented with various thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times from population cell to bike share stations: three minutes, five minutes, ten minutes, and fifteen minutes.
+Consensus regarding the distance that individuals are willing to walk to access a docking station is lacking, but the literature on walking behaviour provided some guidelines to determine the thresholds for our sensitivity analysis.
+Previous studies have found that living within 250 m [@fullerUseNewPublic2011] and 300 m [@kabraBikeShareSystemsAccessibility2020] of a docking station is correlated with bike share use, while other research has found that walking trips are less than 600 m and rarely more than 1200 m [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 m [@larsenQuarterMileReexamining2010].
+HBS will often depict a map at some docking stations to show the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are expected to walk.
+The National Association of City Transportation Officials (NACTO) has a similar normative guide [@nactowalkingstation2015].
+
+In the present case, we found that accessibility calculated using the BFCA method increased with a threshold between two and four minutes, but was then maximized at five minutes.
+Accessibility decreased substantially after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each station.
+
+We experimented with various thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times from population cell to docking stations: three minutes, five minutes, ten minutes, and fifteen minutes.
 We categorized these thresholds as minimum, average, maximum, and extreme, respectively.
@@
 Again, accessibility is fairly uniform, with the exception of one very small area.
+The equity stations noticeably increase accessibility in the east end of the core service area by filling gaps in PBSP coverage.
 
@@
 Differences in accessibility across the service area are now apparent, with users near the university and its adjacent neighborhoods, as well as neighborhoods north of the downtown area (the latter is outlined in green), having slightly higher accessibility.
-While the differences are modest, they are more apparent at this threshold than at shorter walking distances.
+While the differences are modest, they are more apparent at this threshold than at shorter walking distances, especially in the more disadvantaged neighbourhoods in the east end of the core service area where equity stations were added.
 
@@
 Users near the university and the neighborhoods north of the downtown area (the latter is outlined in green) have the highest accessibility, followed by those who live in the city's downtown area.
-Accessibility in the east end, where equity stations were implemented, of the core service area remains low.
+Accessibility in the east end, where equity stations were implemented, of the core service area remains lower than other areas.
 
@@
 The latter method does not allocate population and level of service proportionally, which means that there are often substantial amounts of multiple counting each.
-Figure \ref{fig:figure-10} shows the level of inflation and deflation at five minutes that results when the balanced and the conventional approaches are compared.
+Figure \ref{fig:figure-10} shows the level of inflation and deflation at five minutes that results when the balanced and the conventional FCA methods are compared.
 The inflation and deflation is defined as the ratio of the level of service and accessibility calculated using the conventional approach to the balanced approach.
@@
 This is evident for the equity stations in the east end of the core service area.
-Noticeably, the level of service is least deflated in the downtown area, likely due to the density of stations.
+Noticeably, the level of service is least deflated in the downtown area, likely due to the density of docking stations.
 Figure \ref{fig:figure-11} shows that accessibility is inflated by as much as 50% in many areas across the city.
@@
 
-Next, we wish to examine accessibility differences by income level.
-To this end, we took our micro-zones and imputed the median household income of their corresponding DA.
-Use of bivariate maps helps us to visualize the level of accessibility by level of income.
-Median total household income statistics were drawn from the 2016 Canadian Census.
-
+To examine whether accessibility to HBS was increased for groups with lower income, we estimated what level of accessibility accrues to how many people in different income strata.
+To this end, we took our micro population zones and aggregated the accessibility and population by the median total household income as imputed from the dissemination areas.
+A unique property of the BFCA method, which does not hold for the 2SFCA approach because of the inflation/deflation issues discussed above, is that accessibility and population data can be re-aggregated using income as an aggregator criterion, while preserving the total population and the level of service.
+This avoids demand and supply inflation, and also enables us to present findings in a way that is more intuitive to interpret.
 Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} depict bivariate choropleth maps that combine the spatial distribution of accessibility and median total household income, using tertiles for the coloring scheme.
-In each figure, the top panel is the system without the equity stations (i.e., the original system configuration), and the bottom panel is accessibility in the current system with equity stations.
-As expected, the analysis indicates that the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service area of stations.
-We found that stations added to HBS did indeed expand the spatial coverage of the system, leading to more balance across the core service area.
-By implementing equity stations in areas with lower median total household income in Hamilton, the PBSP attained greater horizontal equity by extending the spatial distribution of bicycles across the city.
-This is particularly evident at the minimum and average thresholds of three and five minutes, respectively, where the equity stations fill a number of gaps in program coverage.
 
@@
           color = "lightgrey") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") +
+  ggtitle("All stations") +
   bi_scale_fill(pal = "GrPink", 
                 dim = 3) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering) +
   theme(panel.background = element_rect(fill = "white"),
@@
           color = "lightgrey") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") +
+  ggtitle("Conventional") +
   bi_scale_fill(pal = "GrPink", 
                 dim = 3) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering) +
   theme(panel.background = element_rect(fill = "white"),
@@
         axis.ticks = element_blank(),
-        plot.margin = unit(c(0, 0, 0, 0), "cm"))
-```
-
-```{r figure-bi-map-threshold-3, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): with equity stations (top panel) and without equity stations (bottom panel)"}
-ggdraw() + draw_plot(bi_map_3_all, 0.00, 0.5, 1, 0.5) + 
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-bi-map-threshold-3, echo = FALSE, out.width="1.2\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income at the minimum threshold of three minutes with equity stations (top panel) and without equity stations (bottom panel)."}
+ggdraw() + 
+  draw_plot(bi_map_3_all, 0.00, 0.5, 1, 0.5) + 
   draw_plot(bi_map_3_conv, 0.00, 0.0, 1, 0.5) + 
-  draw_plot(legend, 0.06, 0.4, 0.2, 0.2) +
-
-  annotation_scale(location = "bl", width_hint = 0.4,
-                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
-  annotation_north_arrow(location = "t", which_north = "true", 
-        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
-        style = north_arrow_fancy_orienteering)
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
 ```
@@
         axis.ticks = element_blank(),
-        plot.margin = unit(c(0, 0, 0, 0), "cm"))
-```
-
-```{r figure-bi-map-threshold-5, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): with equity stations (top panel) and without equity stations (bottom panel)"}
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-bi-map-threshold-5, echo = FALSE, out.width="1.2\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income at the average threshold of five minutes with equity stations (top panel) and without equity stations (bottom panel)."}
 ggdraw() + draw_plot(bi_map_5_all, 0.00, 0.5, 1, 0.5) + 
   draw_plot(bi_map_5_conv, 0.00, 0.0, 1, 0.5) + 
-  draw_plot(legend, 0.06, 0.4, 0.2, 0.2) +
-
-annotation_scale(location = "bl", width_hint = 0.4,
-                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
-  annotation_north_arrow(location = "t", which_north = "true", 
-        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
-        style = north_arrow_fancy_orienteering)
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
 ```
@@
           fill = NA,
-          color = "lightgrey") +  bi_scale_fill(pal = "GrPink", 
+          color = "lightgrey") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") +
+  ggtitle("All stations") +
+  bi_scale_fill(pal = "GrPink", 
                 dim = 3) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering) +
   theme(panel.background = element_rect(fill = "white"),
@@
         axis.ticks = element_blank(),
-        plot.margin = unit(c(0, 0, 0, 0), "cm"))
-```
-
-```{r figure-bi-map-threshold-10, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): with equity stations (top panel) and without equity stations (bottom panel)"}
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-bi-map-threshold-10, echo = FALSE, out.width="1.2\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income at the maximum threshold of ten minutes with equity stations (top panel) and without equity stations (bottom panel)."}
 ggdraw() + draw_plot(bi_map_10_all, 0.00, 0.5, 1, 0.5) + 
   draw_plot(bi_map_10_conv, 0.00, 0.0, 1, 0.5) + 
-  draw_plot(legend, 0.06, 0.4, 0.2, 0.2) +
-
-annotation_scale(location = "bl", width_hint = 0.4,
-                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
-  annotation_north_arrow(location = "t", which_north = "true", 
-        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
-        style = north_arrow_fancy_orienteering)
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
 ```
@@
         axis.ticks = element_blank(),
-        plot.margin = unit(c(0, 0, 0, 0), "cm"))
-```
-
-```{r figure-bi-map-threshold-15, echo = FALSE, out.width="1\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): with equity stations (top panel) and without equity stations (bottom panel)"}
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-bi-map-threshold-15, echo = FALSE, out.width="1.2\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income at the extreme threshold of fifteen with equity stations (top panel) and without equity stations (bottom panel)."}
 ggdraw() + draw_plot(bi_map_15_all, 0.00, 0.5, 1, 0.5) + 
   draw_plot(bi_map_15_conv, 0.00, 0.0, 1, 0.5) + 
-  draw_plot(legend, 0.06, 0.4, 0.2, 0.2) +
-
-annotation_scale(location = "bl", width_hint = 0.4,
-                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
-  annotation_north_arrow(location = "t", which_north = "true", 
-        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
-        style = north_arrow_fancy_orienteering)
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
 ```
@@
 
-To examine whether equity was increased for groups with lower income, we need to estimate what level of accessibility accrues to how many people in different income strata.
-A unique property of the BFCA method, which does not hold for the conventional 2SFCA approach because of the inflation/deflation issues discussed above, is that accessibility and population data can be re-aggregated using income as an aggregator criterion, while preserving the total population and the level of service.
-This avoids demand and supply inflation, and also enables us to present findings in a way that is perhaps more intuitive to interpret.
-
-Therefore, we aggregated the accessibility and population by the median household income as imputed from the DAs (see Table \ref{tab:accessibility-income}).
-Similar to Hosford and Winters  [-@hosfordWhoArePublic2018], we found that Hamilton's PBSP already serviced a large proportion of the population in the bottom and second quintiles of median total household income.
-This may well be an artifact of the spatial socioeconomic and demographic profile of Hamilton, where the most dense parts of the city (where a PBSP is most easily launched) are also those with relatively lower incomes.
+Table \ref{tab:accessibility-income} presents accessibility levels by income strata.
+As expected, the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service area of docking stations.
+We found that stations added to HBS expanded the spatial coverage of HBS, leading to more balance across the core service area.
+This is particularly evident at the minimum and average thresholds of three and five minutes, respectively, where the equity stations fill a number of gaps in program coverage.
+
+In line with Hosford and Winters' [-@hosfordWhoArePublic2018] findings, we found that accessibility to HBS was high for individuals in the bottom and second quintiles of median total household income who live in the centre of the core service area.
+This may well be an artifact of the spatial socioeconomic and demographic profile of Hamilton, where the most dense parts of the city (where a PBSP is most easily launched) are also those with relatively lower incomes [@hosfordWhoArePublic2018].
 On the other hand, we found the levels of accessibility to Hamilton's PBSP are generally lower for populations in the bottom 20% of median total household income, compared to populations in the top 20%.
 
-The spatial mapping revealed that the addition of equity stations increased accessibility to HBS by growing the population serviced irrespective of the walking threshold.
-The largest gains were made for dissemination areas in the second 20%, where an additional 3,073 and 5,395 people could reach a bicycle share station within three and five minutes walk, respectively, after the addition of equity stations.
-However, we find that there were only small increases in the population in the bottom 20% of median total household income who are serviced by the equity stations, and the accessibility gains are also quite modest and smaller than for populations in the second and third quintiles of median total household income.
+Our analysis revealed that the addition of equity stations increased accessibility to HBS by growing the population serviced irrespective of the walking threshold.
+The largest gains were made for dissemination areas in the second 20% median total household income, where an additional 3,073 and 5,395 people could reach a docking station within three and five minutes walk, respectively, after the addition of equity stations.
+However, we found that there were only small increases in the population in the bottom 20% of median total household income who are serviced by the equity stations, and the accessibility gains are also quite modest and smaller than for populations in the second and third quintiles of median total household income.
 This suggests that inequities in accessibility to HBS persist according to income, albeit this depends on the walking time thresholds.
 With and without equity stations, people in the top 20% of income have the highest level of access at a threshold of ten and fifteen minutes.
-Although dissemination areas in the second 20% have the highest level of access by a significant amount at lower distance thresholds, the bottom 20%, who may benefit the most from the equity stations, have the lowest access at three minutes threshold and the second lowest access at all other thresholds.
+Although dissemination areas in the second 20% have the highest level of access by a significant amount at lower distance thresholds, the bottom 20%, who may benefit more from increased access to HBS, have the lowest accessibility at three minutes threshold and the second lowest access at all other thresholds.
 
@@
 Using disaggregated population data, we examined equity in accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
-Using a BFCA approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro scale which better reflects the friction [*is this the right word?*] of walking thresholds and avoids the "absorption of disparities", as articulated by Chen et al.
-[-@chenExploringEquityPerformance2019] and elsewhere in the transportation literature [@rowangouldIdentifyingEnvironmentalJustice2016].
-The BFCA method also considers demand and supply in the calculation of accessibility, which is important for services like PBSPs but not accounted for in other common accessibility models.
+The BFCA approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro scale which better reflects the scale at which walking takes place and avoids the "absorption of disparities", as articulated by Chen et al.
+[-@chenExploringEquityPerformance2019] and elsewhere in the transportation literature [@nationalacademiesofsciencesEffectiveMethodsEnvironmental2004; @rowangouldIdentifyingEnvironmentalJustice2016].
+Our method also considers demand and supply in the calculation of accessibility, which is important for services like PBSPs but not accounted for in other common accessibility models.
 Unlike the 2SFCA method, there was no inflation or deflation in our accessibility estimates because population and level of service were preserved (see Figure \ref{fig:figure-10} and Figure \ref{fig:figure-11}).
@@
 In this way, our paper has made an important contribution by applying an intuitive and useful approach to measure accessibility to a PBSP.
-This open and reproducible research could be replicated for additional transportation and statistical analysis using other individual-level data in Hamilton or revised for PBSPs in other North American cities.
+This open and reproducible research could be expanded for additional analysis using other individual-level data in Hamilton (e.g., age, gender, household size, education, etc.) or applied to PBSPs in other North American cities.
 
@@
 This reflects the normative guide advertised on some docking stations in Hamilton showing other stations within a five minute walk, as well as the directive of NACTO [@nactowalkingstation2015].
-We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}].
+We find that over 118,000 people can access a bike share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}].
 At a minimum threshold of three minutes, too few people can reach stations which leads to relatively high levels of service since there is little crowding for the stations.
@@
 
-The City of Hamilton has recognized from the launch of the PBSP that substantially more stations and bicycles are needed to service the area [@hamiltonsobi2015].
+The City of Hamilton has recognized from the launch of HBS that substantially more stations and bicycles are needed to service the area [@hamiltonsobi2015].
 With a service area of 40 sq.km, it is estimated that Hamilton should have between 380 and 440 stations instead of 130, and 1,500 bicycles instead of 900 [@hamiltonsobi2015].
@@
 Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in service were filled by the additional equity stations.
-The improvement in accessibility is most noticeable at the average and maximum thresholds in the east end of the city, which corresponds to the neighbourhoods where stations were added.
+The improvement in accessibility is most noticeable at the average and maximum thresholds in the east end of the core service area, which corresponds to the neighbourhoods where stations were added.
 However, the contribution of equity stations to reducing inequities in accessibility was modest.
@@
 While previous research found that neighborhoods with more disadvantage are better serviced by HBS, the authors used the Pampalon Deprivation Index to determine the level of disadvantage for dissemination areas _across_ the city not just within the core service area [@hosfordWhoArePublic2018].
-Instead, we use median total household income for each dissemination area _within_ the core service area.
-We conclude that Hamilton's PBSP, while by default located in areas with more deprivation compared to other cities, has disparities in accessibility between income groups.
-With equity stations, many areas with low median total household income in the east end of the service area continue to have low accessibility.
+Instead, we used median total household income for each dissemination area _within_ the core service area.
+We conclude that Hamilton's PBSP, while by default located in areas with more deprivation compared to other cities, continues to have disparities in accessibility between income groups.
+With equity stations, many areas with low median total household income in the east end of the service area continue to have zero or low accessibility at the average threshold of five minutes.
 At the maximum and extreme thresholds, the top 20% have the highest level of access to Hamilton Bike Share.
-These findings aligns with other studies from Tampa [@chenExploringEquityPerformance2019], Philadelphia [@caspiBikesharingPhiladelphiaLowerincome2019] and Seattle [@mooneyFreedomStationSpatial2019], which have found disparities in station location, annual trips, or access to bicycles, respectively, between levels of income and education.
+These findings align with other studies from Tampa [@chenExploringEquityPerformance2019], Philadelphia [@caspiBikesharingPhiladelphiaLowerincome2019] and Seattle [@mooneyFreedomStationSpatial2019], which have found disparities in station location, annual trips, or access to bicycles, respectively, between levels of income and education.
 With the addition of equity stations, there were large gains in accessibility for the second 20% at the average threshold, amounting to over 5,000 more people, but much smaller gains for the bottom 20% with only 326 more people who are able to access a HBS station.
@@
 This paper highlights the importance of both the location and size of docking stations for increasing equity in accessibility to PBSPs.
-Several studies have examined who can access PBSPs based on the location of stations [@babagoliExploringHealthSpatial2019; @qianBikesharingEquityDisadvantaged2020; @qianEnhancingEquitableService2020; @wangNewBikeShare2019], but the potential supply of bicycles at stations in disadvantaged communities is not considered in the literature.
-The size of stations (i.e., potential supply of bicycles) was also not identified by PBSP owners and operators as a factor that needed to be considered in equity efforts [@howlandCurrentEffortsMake2017].
+Several studies have examined who can access PBSPs based on the location of stations [among others, see @babagoliExploringHealthSpatial2019; @qianBikesharingEquityDisadvantaged2020; @qianEnhancingEquitableService2020; @wangNewBikeShare2019], but the size of docking stations in disadvantaged communities is not considered in the literature.
+Station size (i.e., potential maximum supply of bicycles) was also not identified by PBSP owners and operators as a factor that needed to be addressed in equity efforts [@howlandCurrentEffortsMake2017].
 PBSPs may not necessarily reduce inequities in access if stations are easy to reach for disadvantaged groups but offer only a small number of bicycles.
@@
 Babgoli et al.
-[-@babagoliExploringHealthSpatial2019] found a slight but not statistically significant increase in the proportion of neighborhoods with the highest levels of poverty with stations after the Citi Bike expansion in 2015.
+[-@babagoliExploringHealthSpatial2019] found a slight but not statistically significant increase in the proportion of neighborhoods with the highest levels of poverty that had stations after the Citi Bike expansion in 2015.
 Although the Citi Bike expansion was not specifically driven by a desire to reduce inequity in access, 16% of neighborhoods with the highest levels of poverty had stations compared to 12% before.
@@
 Our analysis has identified specific areas that have both low accessibility and low median total household income.
-Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} highlight potential locations for new equity stations to better accommodate groups with lower socioeconomic status.
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} highlight potential locations for new equity stations to better accommodate low income groups.
 Additional stations are needed to further reduce inequities in accessibility.
@@
 Docking stations added to target vertical equity should have an adequate supply of public bicycles, particularly more in areas that have a strong interest in adopting bike share.
-Finally, distance to stations is associated with bike share use [@fullerUseNewPublic2011; @wangNewBikeShare2019], especially when the trip starts at the residence [@fishmanBikeShareSynthesis2013], and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
+Finally, distance to stations is associated with bike share use [@fullerUseNewPublic2011; @wangNewBikeShare2019], especially when the trip starts at the residence [@fishmanBikeShareSynthesis2013].
+Distance can also be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
 To address this issue, transportation planners and PBSP operators should consult with groups and communities who are under-represented among bike share users to determine how far they are willing to walk to reach a docking station and to get their input on the location of new docking stations.
@@
 An additional limitation is the lack of publicly available information about the average number of bicycles at each station on a daily basis.
-HBS works to balance the number of bicycles across stations, but it is reasonable to expect that the number of bicycles will not match exactly the number of racks at every station.
+HBS works to balance the number of bicycles across the core service area, but it is reasonable to expect that the number of bicycles will not match exactly the number of racks at every station.
 Ideally, instead of number of bicycle racks as our measure of supply, we would have liked to use the average number of bicycles at stations, perhaps at different times during the day across many seasons.
-Should this data become available, it would be worthwhile to revisit the question to examine how well operation of the system (including balancing of bicycles across stations) works to maintain the nominal levels of accessibility examined in this paper.
+Should this data become available, it would be worthwhile to revisit the question to examine how well the operation of the system (including balancing of bicycles across stations) works to maintain the nominal levels of accessibility examined in this paper.
 
@@
 
-The addition of specific stations to improve equity within the PBSP in Hamilton had the net effect of increasing accessibility and reducing to some extent vertical inequities.
+The addition of specific stations to improve equity within the PBSP in Hamilton had the net effect of increasing accessibility and reducing vertical inequities to some extent.
 In particular, accessibility improved the most for those in the second 20% median total household income at all thresholds, but the gains were only modest for all income groups.
-Dissemination areas with the bottom 20% had the lowest accessibility at three minutes, and second lowest levels of accessibility at five, ten, and fifteen minutes.
+Dissemination areas in the bottom 20% had the lowest accessibility at three minutes, and second lowest levels of accessibility at five, ten, and fifteen minutes.
 Congestion effects were observed at higher thresholds, with accessibility decreasing significantly once the catchment area is increased to ten minutes walking.
@@
 Wang and Lindsey [-@wangNewBikeShare2019] have noted that there is a lack of research that examines how bike share users' behaviour changes as a result of program changes to station locations or improvements in accessibility.
-As such, a logical next step to this research is to examine whether Hamilton Bike Share's equity stations increased ridership or resulted in new memberships in areas that were previously under-serviced.
+As such, a logical next step to this research is to examine whether the equity stations increased ridership or resulted in new memberships in areas that were previously under-serviced.
 An examination of the types of trips undertaken by residents in these areas would also be informative, such as the study undertaken by Caspi and Norland [-@caspiBikesharingPhiladelphiaLowerincome2019] after bike share stations were implemented in low-income Philadelphia neighborhoods.
@@
 
-This research was completed using open software, and the authors wish to acknowledge the developers of the following `R` packages: `biscale` [@R-biscale], `cowplot` [@R-cowplot], `data.table` [@R-data.table], `disk.frame` [@R-disk.frame], `gdistance` [@R-gdistance], `gridExtra` [@R-gridExtra], `kableExtra` [@R-kableExtra], `knitr` [@R-knitr], `pycno` [@R-pycno], `r5r` [@R-r5r], `raster` [@R-raster], `rgdal` [@R-rgdal], `rticles` [@R-rticles], `sf` [@R-sf], `tidyverse` [@R-tidyverse], `tinytex` [@R-tinytex],
+We wish to thank the anonymous reviewers who provided constructive feedback to improve the quality of our paper.
+This research was completed using open software, and we also wish to acknowledge the developers of the following `R` packages: `biscale` [@R-biscale], `cowplot` [@R-cowplot], `data.table` [@R-data.table], `disk.frame` [@R-disk.frame], `gdistance` [@R-gdistance], `gridExtra` [@R-gridExtra], `kableExtra` [@R-kableExtra], `knitr` [@R-knitr], `pycno` [@R-pycno], `r5r` [@R-r5r], `raster` [@R-raster], `rgdal` [@R-rgdal], `rticles` [@R-rticles], `sf` [@R-sf], `tidyverse` [@R-tidyverse], `tinytex` [@R-tinytex],
 `units` [@R-units].

### 2021-08-23 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Major revisions resubmitted June 2021

--- 2021-06-18 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
@@
             pdfauthor={},
-            pdftitle={Examining equity in accessibility to bike share - a balanced floating catchment area approach},
+            pdftitle={Examining equity in accessibility to bike share: a balanced floating catchment area approach},
             colorlinks=false,
@@
 
-  \title{Examining equity in accessibility to bike share - a balanced floating
+  \title{Examining equity in accessibility to bike share: a balanced floating
 catchment area approach}
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{rm}\NormalTok{(}\DataTypeTok{list =} \KeywordTok{ls}\NormalTok{())}
-\end{Highlighting}
-\end{Shaded}
-
 \newpage
@@
 program, Everyone Rides Initiative, was implemented in 2018 which
-expanded the program by introducing twelve ``equity'' stations which
-expanded HBS to more disadvantaged neighbourhoods in the core service
-area.
+expanded the program by introducing twelve ``equity'' stations to more
+disadvantaged neighbourhoods in the core service area.
 
@@
 comparative analysis to the conventional two-step floating catchment
-area approach to highlight the benefits of this method which, to our
-knowledge, has not yet been used in the cycling literature. The paper
-also assesses the contribution of HBS's equity stations to reducing
+area (2SFCA) approach to highlight the benefits of this method which, to
+our knowledge, has not yet been used in the cycling literature. The
+paper also assesses the contribution of the equity stations to reducing
 inequities in accessibility for different groups according to median
@@
 \hypertarget{literature-review}{%
-\section{2.1. Literature Review}\label{literature-review}}
+\section{2. Literature Review}\label{literature-review}}
 
 \hypertarget{public-bicycle-share-programs}{%
-\subsection{1.1. Public Bicycle Share
+\subsection{2.1. Public Bicycle Share
 Programs}\label{public-bicycle-share-programs}}
@@
 \hypertarget{equity-of-pbsps}{%
-\subsection{1.2. Equity of PBSPs}\label{equity-of-pbsps}}
+\subsection{2.2. Equity of PBSPs}\label{equity-of-pbsps}}
 
@@
 for cities with PBSPs. However, exploring transportation equity by
-investigating where bicycle share stations are located, often using
+investigating where bike share stations are located, often using
 neighborhoods or census tracts as the geographical unit of analysis, can
@@
 docking station decreases the potential of accessing the program, the
-location of stations matters. Kabra et al.~(2020) found that the
-majority of bike share usage in Paris comes from areas within 300 m of
-stations, which amounts to 2-4 minutes walking by an adult who does not
-have a disability. Similar to other public amenities affected by
-crowding, the utility of stations is also limited by the \emph{maximum}
-number of bicycles that they can hold (e.g., their size). Accessibility
+location matters. Kabra et al.~(2020) found that the majority of bike
+share usage in Paris, France comes from areas within 300 m of stations,
+which amounts to 2-4 minutes walking by an adult who does not have a
+disability. Similar to other public amenities affected by crowding, the
+utility of docking stations is also limited by the \emph{maximum} number
+of bicycles that they can hold (e.g., their size). Accessibility
 analyses for PBSPs constitute a positive and evaluation-based approach
 that also has the potential to inform equity efforts. For instance, Wang
-and Lindsey (2019b) investigated whether new or relocated bicycle share
+and Lindsey (2019b) investigated whether new or relocated bike share
 stations increased accessibility and use, which offered important
@@
 approach involves weighting destination opportunities, such as the
-quantity of bicycle share stations, by the time required to reach them
-from an origin using an impedance function (Handy and Niemeier, 1997;
-Kwan, 1998). While such measures are suitable for capturing the
-potential for reaching destinations from a given location, they do not
-take demand or congestion effects into account which is an important
-consideration when calculating accessibility for amenities such as bike
-share stations.
+quantity of bike share stations, by the time required to reach them from
+an origin using an impedance function (Handy and Niemeier, 1997; Kwan,
+1998). While such measures are suitable for capturing the potential for
+reaching destinations from a given location, they do not take demand or
+congestion effects into account which is an important consideration when
+calculating accessibility for amenities such as bike share stations.
 
@@
 individual or household as the smallest unit of analysis or larger
-spatial zones. Previous research on bicycle share equity has typically
-used a meso- or macro-level approach with aggregated data from entire
+spatial zones. Previous research on bike share equity has typically used
+a meso- or macro-level approach with aggregated data from entire
 neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
@@
 are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
-also true for studies examining correlates of bicycle share demand (J.
-Wang and Lindsey, 2019b). Handy and Niemeier (1997) note that using
+also true for studies examining correlates of bike share demand (J. Wang
+and Lindsey, 2019b). Handy and Niemeier (1997) note that using
 disaggregated data in accessibility analyses provides a more accurate
@@
 Vancouver, Montreal, and Ottawa-Gatineau) where areas that are less
-deprived have greater access. Hosford and Winters (2018, pp.~)
+deprived have greater access. Hosford and Winters (2018, p. 47)
 acknowledge that ``Hamilton stands out in that the lower income
@@
 core service area for the PBSP in Hamilton by default covers more of the
-less advantaged areas in the city. However, there is also a great deal
-of variation in income in the city center because of the local
+disadvantaged areas in the city. However, there is also a great deal of
+variation in income in the core service area because of the local
 university and increasing gentrification. Hosford and Winters (2018)
@@
 
-Therefore, this paper contributes to the cycling literature by expanding
-the analysis of Hosford and Winters (2018) to assess equity in
-accessibility to HBS and explore the contribution of the ``equity''
-stations added to Hamilton's PBSP.
+Therefore, this paper also contributes to the cycling literature by
+expanding the analysis of Hosford and Winters (2018) to assess equity in
+accessibility to HBS and explore the contribution of the equity stations
+added to Hamilton's PBSP.
 
@@
 that had been selected and to crowdsource potential locations for
-additional locations (Hamilton, 2014). Most of the suggested locations
-were in the east end of the core service area that lack stations or in
-neighborhoods not serviced by the PBSP. The program was enthusiastically
-welcomed in the city in 2015 - within three weeks of launching, 10,000
-trips had been made (Hamilton, 2015b), however inadequate coverage given
-the size of Hamilton Bike Share's service area was identified as a
-problem early on, and transportation planners noted that the small size
-(i.e., supply of bicycle racks) and low quantity of stations would lead
-to challenges in balancing the system (Hamilton, 2015a).
+additional locations (Hamilton, 2014). Most of the potential locations
+that were suggested were in the east end of the core service area that
+lack stations or in neighborhoods not serviced by the PBSP. The program
+was enthusiastically welcomed in the city in 2015 - within three weeks
+of launching, 10,000 trips had been made (Hamilton, 2015b), however
+inadequate coverage given the size of the program's service area was
+identified as a problem early on, and transportation planners noted that
+the small size (i.e., supply of bicycle racks) and low quantity of
+stations would lead to challenges in balancing the system (Hamilton,
+2015a).
 
@@
 
-Hamilton Bike Share's equity program, Everyone Rides Initiative (ERI),
-was launched in 2018 with the objective of reducing barriers that may
-prevent individuals from accessing the program. Additional bicycles and
-twelve ``equity'' stations were added to the core service area. The
+Hamilton Bike Share Inc., the non-profit operator, launched an equity
+program, Everyone Rides Initiative (ERI), in 2018 with the objective of
+reducing barriers that may prevent individuals from accessing the
+program. Additional bicycles and twelve ``equity'' stations were added
+to the core service area in more disadvantaged neighbourhoods. The
 equity program also offers subsidized memberships to individuals who
@@
 
-\caption{The spatial distribution of bicycle share stations in Hamilton, Ontario.
-Conventional stations are in blue and equity (ERI) stations are in orange.
-The service area of Hamilton Bike Share is outlined in black and the city's downtown core is outlined in dark green.
+\caption{The spatial distribution of bike share docking stations in Hamilton, Ontario.
+Conventional stations are in purple and equity (ERI) stations are in orange.
+The service area of Hamilton Bike Share is outlined in blue and the city's downtown core is outlined in dark green.
 Hamilton Census Metropolitan Area is shown in grey.}\label{fig:sobi-stations-in-hamilton}
@@
 
-Since walking trips to a bicycle share station likely happen at a level
-lower than even the smallest census geography, this requires a more
+Since walking trips to a docking station likely happen at a level lower
+than even the smallest census geography, this requires a more
 disaggregated approach than the use of census geographies. For this
@@
 (for example, to parks or large institutional buildings, etc.). To do
-so, we retrieved shapefiles for various geographic features {[}see Table
-\ref{tab:data-features}{]} from Open Hamilton. Next, we removed these
-features from the PBSP core service area and used pycnophylactic
-interpolation to disaggregate and reallocate population within the
-remaining area {[}see Figure \ref{fig:interpolated-population}{]}.
+so, we retrieved shapefiles for various geographic features from Open
+Hamilton. Next, we removed these features from the PBSP core service
+area and used pycnophylactic interpolation to disaggregate and
+reallocate population within the remaining area {[}see Figure
+\ref{fig:interpolated-population}{]}.
 
@@
 
-\caption{Interpolated population in the Hamilton Bike Share's core service area (outlined in black) and within 30 minutes of walking to the core service area.
+\caption{Interpolated population in Hamilton Bike Share's core service area (outlined in blue) and within 30 minutes of walking to the core service area.
 Each population cell is 50-by-50 m in size.
@@
 To calculate walking times from the centroid of the micro population
-cells to docking stations, we extracted OpenStreetMap data for Hamilton
-Bike Share's service area from
+cells to docking stations, we extracted OpenStreetMap data for HBS's
+service area from
 \href{https://download.bbbike.org/osm/bbbike/}{BBBike}, an online cycle
@@
 provides the networks for calculating walking times from each population
-cell to nearby bicycle share stations, using a maximum walking distance
-of 10 km and walking time of 30 minutes as thresholds. A travel time
-matrix was created with the origins as the coordinates of the population
-cells and the destinations as the coordinates of the bicycle share
-stations within the maximum threshold. This process provides a more
-realistic measure of the friction of reaching stations by taking
-infrastructure into account network travel times, rather than using the
-Euclidean distance from population cell to station. Routing and travel
-time calculations were completed using the \texttt{R} package
-\texttt{r5r}, used for rapid realistic routing operations (Pereira et
-al., 2021).
+cell to nearby docking stations, using a maximum walking distance of 10
+km and walking time of 30 minutes as thresholds. A travel time matrix
+was created with the origins as the coordinates of the population cells
+and the destinations as the coordinates of the docking stations within
+the maximum threshold. This process provides a more realistic measure of
+the friction of reaching stations by taking infrastructure into account
+network travel times, rather than using the Euclidean distance from
+population cell to station. Routing and travel time calculations were
+completed using the \texttt{R} package \texttt{r5r}, used for rapid
+realistic routing operations (Pereira et al., 2021).
 
@@
 a public online repository of data curated by the City of Hamilton.
+Median total household income statistics were drawn from the 2016
+Canadian Census.
 
@@
 Consensus regarding the distance that individuals are willing to walk to
-access a bike share station is lacking, but the literature on walking
-behaviour can provide some guidelines to determine the thresholds for
-our sensitivity analysis. Previous studies have found that living within
-250 m (Fuller et al., 2011) and 300 m (Kabra et al., 2020) of a docking
-station is correlated with bicycle share use, while other research has
+access a docking station is lacking, but the literature on walking
+behaviour provided some guidelines to determine the thresholds for our
+sensitivity analysis. Previous studies have found that living within 250
+m (Fuller et al., 2011) and 300 m (Kabra et al., 2020) of a docking
+station is correlated with bike share use, while other research has
 found that walking trips are less than 600 m and rarely more than 1200 m
 (Millward et al., 2013) or a median distance of 650 m (Larsen et al.,
-2010). Hamilton Bike Share will often depict a map at some docking
-stations to show the locations of the other nearest stations within a
-five minute walk, which suggests that this is an average distance that
-people are expected to walk. National Association of City Transportation
-Officials (NACTO) has a similar normative guide (City Transportation
-Officials, 2015).
+2010). HBS will often depict a map at some docking stations to show the
+locations of the other nearest stations within a five minute walk, which
+suggests that this is an average distance that people are expected to
+walk. The National Association of City Transportation Officials (NACTO)
+has a similar normative guide (City Transportation Officials, 2015).
 
 In the present case, we found that accessibility calculated using the
-balanced FCA approach increases with a threshold between two and four
-minutes, but is then maximized at five minutes. Accessibility decreases
+BFCA method increased with a threshold between two and four minutes, but
+was then maximized at five minutes. Accessibility decreased
 substantially after eight minutes, which is intuitive given that demand
@@
 
-For this reason, we experimented with various thresholds by conducting a
-sensitivity analysis to calculate accessibility at different walking
-times from population cell to bike share stations: three minutes, five
-minutes, ten minutes, and fifteen minutes. We categorized these
-thresholds as minimum, average, maximum, and extreme, respectively. At
-each threshold, we compared accessibility between the current system and
-the original system to examine the contribution of the twelve equity
-stations. When considering the results reported below, it is important
-to remember that accessibility is technically a form of smoothing
-(O'Kelly and Horner, 2003, pp. 7--8): smaller thresholds produce less
-smoothing (which can result in ``spiky'' accessibility landscapes),
-while larger thresholds produce more smoothing and fewer spikes.
+We experimented with various thresholds by conducting a sensitivity
+analysis to calculate accessibility at different walking times from
+population cell to docking stations: three minutes, five minutes, ten
+minutes, and fifteen minutes. We categorized these thresholds as
+minimum, average, maximum, and extreme, respectively. At each threshold,
+we compared accessibility between the current system and the original
+system to examine the contribution of the twelve equity stations. When
+considering the results reported below, it is important to remember that
+accessibility is technically a form of smoothing (O'Kelly and Horner,
+2003, pp. 7--8): smaller thresholds produce less smoothing (which can
+result in ``spiky'' accessibility landscapes), while larger thresholds
+produce more smoothing and fewer spikes.
 
@@
 systems. Again, accessibility is fairly uniform, with the exception of
-one very small area.
+one very small area. The equity stations noticeably increase
+accessibility in the east end of the core service area by filling gaps
+in PBSP coverage.
 
@@
 differences are modest, they are more apparent at this threshold than at
-shorter walking distances.
+shorter walking distances, especially in the more disadvantaged
+neighbourhoods in the east end of the core service area where equity
+stations were added.
 
@@
 Accessibility in the east end, where equity stations were implemented,
-of the core service area remains low.
+of the core service area remains lower than other areas.
 
@@
 Figure \ref{fig:figure-10} shows the level of inflation and deflation at
-five minutes that results when the balanced and the conventional
-approaches are compared. The inflation and deflation is defined as the
+five minutes that results when the balanced and the conventional FCA
+methods are compared. The inflation and deflation is defined as the
 ratio of the level of service and accessibility calculated using the
@@
 of service is least deflated in the downtown area, likely due to the
-density of stations. Figure \ref{fig:figure-11} shows that accessibility
-is inflated by as much as 50\% in many areas across the city.
+density of docking stations. Figure \ref{fig:figure-11} shows that
+accessibility is inflated by as much as 50\% in many areas across the
+city.
 
@@
 
-Next, we wish to examine accessibility differences by income level. To
-this end, we took our micro-zones and imputed the median household
-income of their corresponding DA. Use of bivariate maps helps us to
-visualize the level of accessibility by level of income. Median total
-household income statistics were drawn from the 2016 Canadian Census.
-
-Figures \ref{fig:figure-bi-map-threshold-3},
+To examine whether accessibility to HBS was increased for groups with
+lower income, we estimated what level of accessibility accrues to how
+many people in different income strata. To this end, we took our micro
+population zones and aggregated the accessibility and population by the
+median total household income as imputed from the dissemination areas. A
+unique property of the BFCA method, which does not hold for the 2SFCA
+approach because of the inflation/deflation issues discussed above, is
+that accessibility and population data can be re-aggregated using income
+as an aggregator criterion, while preserving the total population and
+the level of service. This avoids demand and supply inflation, and also
+enables us to present findings in a way that is more intuitive to
+interpret. Figures \ref{fig:figure-bi-map-threshold-3},
 \ref{fig:figure-bi-map-threshold-5},
@@
 that combine the spatial distribution of accessibility and median total
-household income, using tertiles for the coloring scheme. In each
-figure, the top panel is the system without the equity stations (i.e.,
-the original system configuration), and the bottom panel is
-accessibility in the current system with equity stations. As expected,
-the analysis indicates that the extreme threshold of fifteen minutes is
+household income, using tertiles for the coloring scheme.
+
+\begin{figure}
+\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income at the minimum threshold of three minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-3}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income at the average threshold of five minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-5}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income at the maximum threshold of ten minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-10}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income at the extreme threshold of fifteen with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-15}
+\end{figure}
+
+Table \ref{tab:accessibility-income} presents accessibility levels by
+income strata. As expected, the extreme threshold of fifteen minutes is
 associated with the largest number of people who are within the assumed
-service area of stations. We found that stations added to HBS did indeed
-expand the spatial coverage of the system, leading to more balance
-across the core service area. By implementing equity stations in areas
-with lower median total household income in Hamilton, the PBSP attained
-greater horizontal equity by extending the spatial distribution of
-bicycles across the city. This is particularly evident at the minimum
-and average thresholds of three and five minutes, respectively, where
-the equity stations fill a number of gaps in program coverage.
-
-\begin{verbatim}
-## Using plotunit = 'm'
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in f(...): True north is not meaningful without coord_sf()
-\end{verbatim}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income (threshold: 3 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-3}
-\end{figure}
-
-\begin{verbatim}
-## Using plotunit = 'm'
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in f(...): True north is not meaningful without coord_sf()
-\end{verbatim}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income (threshold: 5 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-5}
-\end{figure}
-
-\begin{verbatim}
-## Using plotunit = 'm'
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in f(...): True north is not meaningful without coord_sf()
-\end{verbatim}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income (threshold: 10 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-10}
-\end{figure}
-
-\begin{verbatim}
-## Using plotunit = 'm'
-\end{verbatim}
-
-\begin{verbatim}
-## Warning in f(...): True north is not meaningful without coord_sf()
-\end{verbatim}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income (threshold: 15 min): with equity stations (top panel) and without equity stations (bottom panel)}\label{fig:figure-bi-map-threshold-15}
-\end{figure}
-
-To examine whether equity was increased for groups with lower income, we
-need to estimate what level of accessibility accrues to how many people
-in different income strata. A unique property of the BFCA method, which
-does not hold for the conventional 2SFCA approach because of the
-inflation/deflation issues discussed above, is that accessibility and
-population data can be re-aggregated using income as an aggregator
-criterion, while preserving the total population and the level of
-service. This avoids demand and supply inflation, and also enables us to
-present findings in a way that is perhaps more intuitive to interpret.
-
-Therefore, we aggregated the accessibility and population by the median
-household income as imputed from the DAs (see Table
-\ref{tab:accessibility-income}). Similar to Hosford and Winters (2018),
-we found that Hamilton's PBSP already serviced a large proportion of the
-population in the bottom and second quintiles of median total household
-income. This may well be an artifact of the spatial socioeconomic and
-demographic profile of Hamilton, where the most dense parts of the city
-(where a PBSP is most easily launched) are also those with relatively
-lower incomes. On the other hand, we found the levels of accessibility
-to Hamilton's PBSP are generally lower for populations in the bottom
-20\% of median total household income, compared to populations in the
-top 20\%.
-
-The spatial mapping revealed that the addition of equity stations
-increased accessibility to HBS by growing the population serviced
-irrespective of the walking threshold. The largest gains were made for
-dissemination areas in the second 20\%, where an additional 3,073 and
-5,395 people could reach a bicycle share station within three and five
-minutes walk, respectively, after the addition of equity stations.
-However, we find that there were only small increases in the population
-in the bottom 20\% of median total household income who are serviced by
-the equity stations, and the accessibility gains are also quite modest
-and smaller than for populations in the second and third quintiles of
-median total household income. This suggests that inequities in
-accessibility to HBS persist according to income, albeit this depends on
-the walking time thresholds. With and without equity stations, people in
-the top 20\% of income have the highest level of access at a threshold
-of ten and fifteen minutes. Although dissemination areas in the second
-20\% have the highest level of access by a significant amount at lower
-distance thresholds, the bottom 20\%, who may benefit the most from the
-equity stations, have the lowest access at three minutes threshold and
-the second lowest access at all other thresholds.
+service area of docking stations. We found that stations added to HBS
+expanded the spatial coverage of HBS, leading to more balance across the
+core service area. This is particularly evident at the minimum and
+average thresholds of three and five minutes, respectively, where the
+equity stations fill a number of gaps in program coverage.
+
+In line with Hosford and Winters' (2018) findings, we found that
+accessibility to HBS was high for individuals in the bottom and second
+quintiles of median total household income who live in the centre of the
+core service area. This may well be an artifact of the spatial
+socioeconomic and demographic profile of Hamilton, where the most dense
+parts of the city (where a PBSP is most easily launched) are also those
+with relatively lower incomes (Hosford and Winters, 2018). On the other
+hand, we found the levels of accessibility to Hamilton's PBSP are
+generally lower for populations in the bottom 20\% of median total
+household income, compared to populations in the top 20\%.
+
+Our analysis revealed that the addition of equity stations increased
+accessibility to HBS by growing the population serviced irrespective of
+the walking threshold. The largest gains were made for dissemination
+areas in the second 20\% median total household income, where an
+additional 3,073 and 5,395 people could reach a docking station within
+three and five minutes walk, respectively, after the addition of equity
+stations. However, we found that there were only small increases in the
+population in the bottom 20\% of median total household income who are
+serviced by the equity stations, and the accessibility gains are also
+quite modest and smaller than for populations in the second and third
+quintiles of median total household income. This suggests that
+inequities in accessibility to HBS persist according to income, albeit
+this depends on the walking time thresholds. With and without equity
+stations, people in the top 20\% of income have the highest level of
+access at a threshold of ten and fifteen minutes. Although dissemination
+areas in the second 20\% have the highest level of access by a
+significant amount at lower distance thresholds, the bottom 20\%, who
+may benefit more from increased access to HBS, have the lowest
+accessibility at three minutes threshold and the second lowest access at
+all other thresholds.
 
@@
 to Hamilton Bike Share, with a particular focus on assessing the
-contribution of the program's equity stations. Using a BFCA approach,
+contribution of the program's equity stations. The BFCA approach,
 combined with pycnophylactic interpolation, enabled us to measure
-accessibility on a micro scale which better reflects the friction
-{[}\emph{is this the right word?}{]} of walking thresholds and avoids
-the ``absorption of disparities'', as articulated by Chen et al.~(2019)
-and elsewhere in the transportation literature (Rowangould et al.,
-2016). The BFCA method also considers demand and supply in the
-calculation of accessibility, which is important for services like PBSPs
-but not accounted for in other common accessibility models. Unlike the
-2SFCA method, there was no inflation or deflation in our accessibility
+accessibility on a micro scale which better reflects the scale at which
+walking takes place and avoids the ``absorption of disparities'', as
+articulated by Chen et al.~(2019) and elsewhere in the transportation
+literature (National Academies of Sciences, 2004; Rowangould et al.,
+2016). Our method also considers demand and supply in the calculation of
+accessibility, which is important for services like PBSPs but not
+accounted for in other common accessibility models. Unlike the 2SFCA
+method, there was no inflation or deflation in our accessibility
 estimates because population and level of service were preserved (see
@@
 approach to measure accessibility to a PBSP. This open and reproducible
-research could be replicated for additional transportation and
-statistical analysis using other individual-level data in Hamilton or
-revised for PBSPs in other North American cities.
+research could be expanded for additional analysis using other
+individual-level data in Hamilton (e.g., age, gender, household size,
+education, etc.) or applied to PBSPs in other North American cities.
 
@@
 directive of NACTO (City Transportation Officials, 2015). We find that
-over 118,000 people can access a bicycle share station within a 15
-minute walk, which represents roughly 85\% of the total population in
-the core service area {[}see Table \ref{tab:accessibility-income}{]}. At
-a minimum threshold of three minutes, too few people can reach stations
+over 118,000 people can access a bike share station within a 15 minute
+walk, which represents roughly 85\% of the total population in the core
+service area {[}see Table \ref{tab:accessibility-income}{]}. At a
+minimum threshold of three minutes, too few people can reach stations
 which leads to relatively high levels of service since there is little
@@
 
-The City of Hamilton has recognized from the launch of the PBSP that
+The City of Hamilton has recognized from the launch of HBS that
 substantially more stations and bicycles are needed to service the area
@@
 accessibility is most noticeable at the average and maximum thresholds
-in the east end of the city, which corresponds to the neighbourhoods
-where stations were added. However, the contribution of equity stations
-to reducing inequities in accessibility was modest.
+in the east end of the core service area, which corresponds to the
+neighbourhoods where stations were added. However, the contribution of
+equity stations to reducing inequities in accessibility was modest.
 
@@
 dissemination areas \emph{across} the city not just within the core
-service area (Hosford and Winters, 2018). Instead, we use median total
+service area (Hosford and Winters, 2018). Instead, we used median total
 household income for each dissemination area \emph{within} the core
 service area. We conclude that Hamilton's PBSP, while by default located
-in areas with more deprivation compared to other cities, has disparities
-in accessibility between income groups. With equity stations, many areas
-with low median total household income in the east end of the service
-area continue to have low accessibility. At the maximum and extreme
+in areas with more deprivation compared to other cities, continues to
+have disparities in accessibility between income groups. With equity
+stations, many areas with low median total household income in the east
+end of the service area continue to have zero or low accessibility at
+the average threshold of five minutes. At the maximum and extreme
 thresholds, the top 20\% have the highest level of access to Hamilton
-Bike Share. These findings aligns with other studies from Tampa (Chen et
+Bike Share. These findings align with other studies from Tampa (Chen et
 al., 2019), Philadelphia (Caspi and Noland, 2019) and Seattle (Mooney et
@@
 Several studies have examined who can access PBSPs based on the location
-of stations (Babagoli et al., 2019; Qian and Jaller, 2020; Qian et al.,
-2020; J. Wang and Lindsey, 2019b), but the potential supply of bicycles
-at stations in disadvantaged communities is not considered in the
-literature. The size of stations (i.e., potential supply of bicycles)
+of stations (among others, see Babagoli et al., 2019; Qian and Jaller,
+2020; Qian et al., 2020; J. Wang and Lindsey, 2019b), but the size of
+docking stations in disadvantaged communities is not considered in the
+literature. Station size (i.e., potential maximum supply of bicycles)
 was also not identified by PBSP owners and operators as a factor that
-needed to be considered in equity efforts (Howland et al., 2017). PBSPs
+needed to be addressed in equity efforts (Howland et al., 2017). PBSPs
 may not necessarily reduce inequities in access if stations are easy to
@@
 significant increase in the proportion of neighborhoods with the highest
-levels of poverty with stations after the Citi Bike expansion in 2015.
-Although the Citi Bike expansion was not specifically driven by a desire
-to reduce inequity in access, 16\% of neighborhoods with the highest
-levels of poverty had stations compared to 12\% before. This leaves open
-the question as to whether the stations held enough bicycles to meet
-demand, and whether congestion effects in neighbourhoods with more
-poverty are reducing the potential of bike share trips.
+levels of poverty that had stations after the Citi Bike expansion in
+2015. Although the Citi Bike expansion was not specifically driven by a
+desire to reduce inequity in access, 16\% of neighborhoods with the
+highest levels of poverty had stations compared to 12\% before. This
+leaves open the question as to whether the stations held enough bicycles
+to meet demand, and whether congestion effects in neighbourhoods with
+more poverty are reducing the potential of bike share trips.
 
@@
 \ref{fig:figure-bi-map-threshold-15} highlight potential locations for
-new equity stations to better accommodate groups with lower
-socioeconomic status. Additional stations are needed to further reduce
-inequities in accessibility. This information will be useful to
-transportation planners in Hamilton and HBS operators. Our findings also
-suggest that PBSPs should consider increasing the size of stations in
-disadvantaged communities. Docking stations added to target vertical
-equity should have an adequate supply of public bicycles, particularly
-more in areas that have a strong interest in adopting bike share.
-Finally, distance to stations is associated with bike share use (Fuller
-et al., 2011; J. Wang and Lindsey, 2019b), especially when the trip
-starts at the residence (Fishman et al., 2013), and can be a barrier to
-using PBSPs (Fishman et al., 2014). To address this issue,
-transportation planners and PBSP operators should consult with groups
-and communities who are under-represented among bike share users to
-determine how far they are willing to walk to reach a docking station
-and to get their input on the location of new docking stations.
+new equity stations to better accommodate low income groups. Additional
+stations are needed to further reduce inequities in accessibility. This
+information will be useful to transportation planners in Hamilton and
+HBS operators. Our findings also suggest that PBSPs should consider
+increasing the size of stations in disadvantaged communities. Docking
+stations added to target vertical equity should have an adequate supply
+of public bicycles, particularly more in areas that have a strong
+interest in adopting bike share. Finally, distance to stations is
+associated with bike share use (Fuller et al., 2011; J. Wang and
+Lindsey, 2019b), especially when the trip starts at the residence
+(Fishman et al., 2013). Distance can also be a barrier to using PBSPs
+(Fishman et al., 2014). To address this issue, transportation planners
+and PBSP operators should consult with groups and communities who are
+under-represented among bike share users to determine how far they are
+willing to walk to reach a docking station and to get their input on the
+location of new docking stations.
 
@@
 about the average number of bicycles at each station on a daily basis.
-HBS works to balance the number of bicycles across stations, but it is
-reasonable to expect that the number of bicycles will not match exactly
-the number of racks at every station. Ideally, instead of number of
-bicycle racks as our measure of supply, we would have liked to use the
-average number of bicycles at stations, perhaps at different times
-during the day across many seasons. Should this data become available,
-it would be worthwhile to revisit the question to examine how well
-operation of the system (including balancing of bicycles across
+HBS works to balance the number of bicycles across the core service
+area, but it is reasonable to expect that the number of bicycles will
+not match exactly the number of racks at every station. Ideally, instead
+of number of bicycle racks as our measure of supply, we would have liked
+to use the average number of bicycles at stations, perhaps at different
+times during the day across many seasons. Should this data become
+available, it would be worthwhile to revisit the question to examine how
+well the operation of the system (including balancing of bicycles across
 stations) works to maintain the nominal levels of accessibility examined
@@
 The addition of specific stations to improve equity within the PBSP in
-Hamilton had the net effect of increasing accessibility and reducing to
-some extent vertical inequities. In particular, accessibility improved
-the most for those in the second 20\% median total household income at
-all thresholds, but the gains were only modest for all income groups.
-Dissemination areas with the bottom 20\% had the lowest accessibility at
-three minutes, and second lowest levels of accessibility at five, ten,
-and fifteen minutes. Congestion effects were observed at higher
-thresholds, with accessibility decreasing significantly once the
-catchment area is increased to ten minutes walking.
+Hamilton had the net effect of increasing accessibility and reducing
+vertical inequities to some extent. In particular, accessibility
+improved the most for those in the second 20\% median total household
+income at all thresholds, but the gains were only modest for all income
+groups. Dissemination areas in the bottom 20\% had the lowest
+accessibility at three minutes, and second lowest levels of
+accessibility at five, ten, and fifteen minutes. Congestion effects were
+observed at higher thresholds, with accessibility decreasing
+significantly once the catchment area is increased to ten minutes
+walking.
 
@@
 program changes to station locations or improvements in accessibility.
-As such, a logical next step to this research is to examine whether
-Hamilton Bike Share's equity stations increased ridership or resulted in
-new memberships in areas that were previously under-serviced. An
-examination of the types of trips undertaken by residents in these areas
-would also be informative, such as the study undertaken by Caspi and
-Norland (2019) after bike share stations were implemented in low-income
-Philadelphia neighborhoods. The bulk of cycling facilities that have
-been built in Hamilton to date are located in the core service area near
-the conventional stations. It would be worthwhile to explore the route
+As such, a logical next step to this research is to examine whether the
+equity stations increased ridership or resulted in new memberships in
+areas that were previously under-serviced. An examination of the types
+of trips undertaken by residents in these areas would also be
+informative, such as the study undertaken by Caspi and Norland (2019)
+after bike share stations were implemented in low-income Philadelphia
+neighborhoods. The bulk of cycling facilities that have been built in
+Hamilton to date are located in the core service area near the
+conventional stations. It would be worthwhile to explore the route
 choice of bike share trips departing or ending at the equity stations
@@
 
-This research was completed using open software, and the authors wish to
-acknowledge the developers of the following \texttt{R} packages:
-\texttt{biscale} (Prener et al., 2020), \texttt{cowplot} (Wilke, 2020),
+We wish to thank the anonymous reviewers who provided constructive
+feedback to improve the quality of our paper. This research was
+completed using open software, and we also wish to acknowledge the
+developers of the following \texttt{R} packages: \texttt{biscale}
+(Prener et al., 2020), \texttt{cowplot} (Wilke, 2020),
 \texttt{data.table} (Dowle and Srinivasan, 2021), \texttt{disk.frame}
@@
 
+\leavevmode\hypertarget{ref-nationalacademiesofsciencesEffectiveMethodsEnvironmental2004}{}%
+National Academies of Sciences, E., 2004. Effective Methods for
+Environmental Justice Assessment.
+doi:\href{https://doi.org/10.17226/13694}{10.17226/13694}
+
 \leavevmode\hypertarget{ref-nickkarSpatialtemporalGenderLand2019}{}%

### 2021-08-23 Bike-share-spatial-equity/Response-to-Reviewers-R2.Rmd
First stab at revisions after second decision.

---
title: "Response to Reviewers"
author: ""
date: "8/23/2021"
output: pdf_document
---

We would like to express our gratitude to the editors and anonymous reviewers for their time and attention to our submission.

## Reviewer #1:

I appreciate the authors' efforts to address my comments. There are some points that need further clarification.

\textcolor{blue}{Thank you for the opportunity to further clarify the paper.}

1. The BFCA introduces an element of congestion into accessibility calculations. You are talking about congestion at docking stations, right? If so, please specify. Also, this is a static approach without considering the time dimension. It essentially assumes that all travelers arrive at the
docking stations simultaneously, and thus one bicycle can be used only by one user. This is hardly the case in any real-world settings. How would you justify this?

\textcolor{blue}{Thank you for this comment. Floating catchment areas introduce an element of demand (or congestion). We clarified in the text that demand in our case is at the docking stations. You are right that the analysis is static in that it does not account for temporal variations in demand. Unfortunately, the location of bicycles throughout the day is not available publicly, and we are not able at the moment to implement a dynamic analysis. We note this as a direction for future research. With respect to the static analysis, it is true that not all travellers arrive at a station at the same time. However, as a standard, this is still a valid assumption from a systems design perspective. Contrast to standards in terms of area of parks per population: not everyone goes to the park at the same time, but calculating area of park per person is still a useful indicator of supply and demand. Or standards for health care provision: not everyone goes to an emergency room at the same time, but calculating ER beds per person is still a good indicator of supply and demand. In the case of accessibility to bike share systems discussing the capacity of the system in these terms may be surprising initially, but we'd argue this is because no standards have been developed for this kind of public service.}

2. In response to my previous comment 3, the authors noted that the data are not aggregated. In the abstract, they say "we then reaggregate the estimated accessibility for further analysis using census data". I am confused. I am also confused about the geographic units of analysis. Population cell is used in Section 4.1. Then in Section 4.2 population data is interpolated to smaller polygons that are 50-by-50 in size. I guess these are the same thing. Besides, there is something named micro-zones. Please check carefully throughout the manuscript and unify the terms.

\textcolor{blue}{We are sorry that this was unclear. In section 4.1 we mean population units in general (they could be census polygons, interpolated populatation, or any other carographic representation of the population. We change the text to say "population unit" to ensure that this is not confused with the population cells in section 4.2. These populations cells are the 50x50m polygons that we use to interpolate the population from census Dissemination Areas to smaller geographical units. For vertical equity analysis we use census income data at the level of Dissemination Areas. In figures 9-12 accessibility is seen at the level of the 50x50m population cells, and income is identical for all cells in the same DA. In other words, accessibility is still spatially disaggregated but income is for DAs. In Table 1 we do aggregate accessibility, but this aggregation is not by geography but rather by income level. A major advantage of the BFCA approach is that this kind of aggregation is possible while still preserving the population and level of service. We have edited the abstract and section 4.1 to clarify this. We also revised the text to make consistent use of the terms.}

3. Regarding the distance threshold, I guess you used it to compute w_ij via a distance decay function. But it is not clear how the distance-decay function is defined. Further, it is counterintuitive to me that the increase in the distance threshold decreases the accessibility. 
With the increase in the distance threshold, more people can reach the docking stations. Isn't this indicating an increase in the accessibility? When the demand surpasses the capacity, the number of people that can be served is bounded by the capacity of the docking station. But it will not decrease. Adding an upper bound to Pj may resolve this issue. But this brings me to the problem of why considering congestion at the very beginning? For highway systems, we need to consider congestion since the higher the volume, the longer the queuing delay one needs to pass the highway in the congested regime. However, are customers going to wait in a bikeshare system? A clearer motivation for using this method would be helpful.

\textcolor{blue}{Very good point. We assumed that the distance decay function would be apparent from the reference to thresholds. We now note that the distance decay is as follows:}

$$
\color{blue}{{w_{ij} = \begin{cases} 
      1 & \text{when } t_{ij}\leq \gamma \\
      0 & \text{otherwise} 
   \end{cases}}}
$$
\textcolor{blue}{where $\gamma$ is the threshold.}

\textcolor{blue}{With respect to the "counter-intuitive" result that accessibility declines when the distance threshold increases, we can illustrate the effect with a simple example. Imagine a station with a single bicycle. A single person is located at 3 minutes from that stations. Then, a single person is located at 4 minutes from that station. Finally, a third person is located at 5 minutes from the station. Suppose that the catchment of the station extends to a threshold of three minutes. The population within the catchment is 1, and the level of service is therefore 1 bicycle per person. Since the other two people in this simple system do not reach the station within three minutes they are not serviced. Hence, the person at 3 minutes enjoys a high level of accesibility, since it can reach a level of service of 1 bicycle per person. Next, suppose that the catchment of the station is 4 minutes. Now two people are in the service area of the station. The level of service of the station is 1/2 bicycles per person. These two people now collectively have lower accessibility, since they can reach only a lower level of service. Finally, suppose that the catchment of the station includes all three people: the level of service now is 1/3 bicycles per person, and collectively the accessibility is even lower. Now, perhaps the confusion is about what accessibility means in this framework. As this simple example shows, it means accessibility to level of service. When the catchment area is small, few people enjoy relatively high levels of accessibility; when the catchment area is large, more people enjoy accessibility, but only to lower levels of service. We discuss the motivation for using this approach in our response to your first point.}

4. Analysis shows that accessibility increase is modest especially for population in the bottom 20% of median total household income. How do you interpret this finding? Does that mean the equity stations fail their purposes?

\textcolor{blue}{Thank you for this question. In our view, the equity stations were a qualified success: they increased accessibility in general, and more people have higher accessibility than without the equity stations. This suggests that horizontal equity improved. As well, the accessibility differential for people in DAs with low incomes and people in DAs with higher incomes became smaller, but accessibility still was much lower for lower income populations even after the introduction of the equity stations. Hence, we believe that the gap in vertical equity was only modestly addressed. Please see the revised discussion.}

5. Eq. (4) and (5), I could not see the differences between the terms on the right hand side. If they are not same thing, why use two notation?

\textcolor{blue}{Thank you for pointing this out. There were typos in these formulas, and they should be as follows. In the first step:}
$$
\color{blue}{{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ij}}}}}
$$

\textcolor{blue}{and in the second step:}
$$
\color{blue}{{w_{ij}^{j} = \frac {w_{ij}}{\sum_{i = 1}^{n} {w_{ij}}}}}
$$

\textcolor{blue}{The difference is that in the first step the weights are row-standardized and in the second step they are column-standardized.}

\textcolor{blue}{Thank you again for the opportunity to clarify the above points.}

## Reviewer #2:

The authors have addressed my concerns

\textcolor{blue}{Thank you again for your suggestions to improve the paper.}

### 2021-08-23 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
First stab at revisions after second decision.

@@
     text: "Corresponding Author"
-abstract:
+abstract: |
 
@@
 In this case study, we investigate differentials in accessibility to stations using a balanced floating catchment area approach and compare accessibility with and without the equity stations.
-We analyze micro zones to better reflect walking to a station and conduct a sensitivity analysis at several walking time thresholds.
-We then reaggregate the estimated accessibility for further analysis using census data.
+We analyze population interpolated to small cells to better reflect walking to a station and conduct a sensitivity analysis at several walking time thresholds.
+We then reaggregate the estimated accessibility by income groups for further analysis.
 Our findings indicate that equity stations increased accessibility for the serviced population at every threshold examined, but the increase was relatively modest especially for population in the bottom 20% of median total household income.
@@
 All data were obtained from publicly available sources and organized in the form of a data package.
-Following best practices in spatial data science [@brunsdon2020opening], the code and data needed to reproduce, modify or extend the analysis are available for download.^[https://github.com/paezha/Accessibility-Sobi-Hamilton]
+Following best practices in spatial data science [@brunsdon2020opening], an open data product [@arribas2021open] along with the code needed to reproduce, modify or extend the analysis are available for download.^[https://github.com/paezha/Accessibility-Sobi-Hamilton]
 
@@
 As seen in the equation above, the population allocated to station $j$ is the weighted sum of the population in the region; a spatial weight $w_{ij}$ represents the friction that the population at $i$ faces when reaching station $j$, and is usually given by a distance-decay function, so that each station is assumed to service only a segment of the population within a limited geographical range.
-
-Next, the supply at each station (i.e., the maximum number of bicycle racks) is divided by its estimated service population within the established catchment area; this gives the level of service of station $j$ in bicycle racks per person:
+The level of service of station $j$ in bicycle racks per person is the supply at each station (i.e., the maximum number of bicycle racks) divided by the population within the established catchment area:
 \begin{equation}
@@
 
-Finally, the accessibility of population cell $i$ is calculated as the weighted sum of the level of service of all stations that can be reached from there according to the spatial weights:
+In the second step, the accessibility of population unit $i$ is calculated as the weighted sum of the level of service of all stations that can be reached from there according to the spatial weights:
 \begin{equation}
 \label{eq:FCA-accessibility}
-A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
+A_i = {\sum_{j = 1}^{J} L_j{w_{ij}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ij}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
 \end{equation}
 
-The balanced approach of Paez et al [-@paezDemandLevelService2019] replaces the spatial weights with normalized versions as follows:
+The balanced approach of Paez et al. [-@paezDemandLevelService2019] replaces the spatial weights with normalized versions as follows. In the first step, the population is weighted with:
 \begin{equation}
 \label{eq:spatial-weights}
-{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
+{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ij}}}}
 \end{equation}
 
-\noindent and:
+\noindent and in the second step, the level of service is weighted using:
 \begin{equation}
 \label{eq:spatial-weights-2}
-{w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
+{w_{ij}^{j} = \frac {w_{ij}}{\sum_{i = 1}^{n} {w_{ij}}}}
 \end{equation}
@@
 \label{eq:weights}
-\sum_{j = 1}^{J} {w^i_{ji}} = 1
+\sum_{j = 1}^{J} {w^i_{ij}} = 1
 \end{equation}
@@
 \label{eq:weights-2}
-\sum_{i = 1}^{n} {w^j_{ji}} = 1
+\sum_{i = 1}^{n} {w^j_{ij}} = 1
 \end{equation}
@@
 Since walking trips to a docking station likely happen at a level lower than even the smallest census geography, this requires a more disaggregated approach than the use of census geographies.
-For this reason, we implemented our analysis parting from micro population zones to better reflect the friction of walking to a docking station, which is an important component of a bike share trip [@chenExploringEquityPerformance2019].
+For this reason, we implemented our analysis parting from small population cells to better reflect the friction of walking to a docking station, which is an important component of a bike share trip [@chenExploringEquityPerformance2019].
 
@@
 
-To calculate walking times from the centroid of the micro population cells to docking stations, we extracted OpenStreetMap data for HBS's service area from [BBBike](https://download.bbbike.org/osm/bbbike/), an online cycle route planner that interfaces with OpenStreetMap.
+To calculate walking times from the centroid of the population cells to docking stations, we extracted OpenStreetMap data for HBS's service area from [BBBike](https://download.bbbike.org/osm/bbbike/), an online cycle route planner that interfaces with OpenStreetMap.
 OpenStreetMap data provides the networks for calculating walking times from each population cell to nearby docking stations, using a maximum walking distance of 10 km and walking time of 30 minutes as thresholds.
@@
 7-8]: smaller thresholds produce less smoothing (which can result in "spiky" accessibility landscapes), while larger thresholds produce more smoothing and fewer spikes.
+
+In the analysis we use the following distance decay function: 
+$$
+{w_{ij} = \begin{cases} 
+      1 & \text{when } t_{ij}\leq \gamma \\
+      0 & \text{otherwise} 
+   \end{cases}}
+$$
+\noindent where $\gamma$ is the relevant threshold. The weights are standardized as discussed above.
 
@@
 To examine whether accessibility to HBS was increased for groups with lower income, we estimated what level of accessibility accrues to how many people in different income strata.
-To this end, we took our micro population zones and aggregated the accessibility and population by the median total household income as imputed from the dissemination areas.
+To this end, we took our small population cells and aggregated the accessibility and population by the median total household income as imputed from the dissemination areas.
 A unique property of the BFCA method, which does not hold for the 2SFCA approach because of the inflation/deflation issues discussed above, is that accessibility and population data can be re-aggregated using income as an aggregator criterion, while preserving the total population and the level of service.

### 2021-08-30 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Revisions - August 2021

@@
     footnote: 
-  - name: Antonio Paez
+  - name: Antonio Páez
     email: paezha@mcmaster.ca
@@
   Public bicycle share programs (PBSPs) can play a role in advancing transportation equity if they make bicycling more accessible to disadvantaged populations.
-In Ontario, the City of Hamilton expanded their PBSP in 2018 by adding twelve "equity" stations with the explicit objective of increasing access for under-serviced neighbourhoods.
+In Ontario, Hamilton Bike Share expanded their program in 2018 by adding twelve "equity" stations with the explicit objective of increasing access for under-serviced neighbourhoods.
 In this case study, we investigate differentials in accessibility to stations using a balanced floating catchment area approach and compare accessibility with and without the equity stations.
@@
 
-The potential of public bicycle share programs (PBSPs) to increase bicycling levels is but one of many reasons for implementing such programs in urban areas [see @hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019].
+The potential of public bicycle share programs (PBSPs) to increase bicycling levels is but one of many reasons for implementing such programs in urban areas [see @fishmanBikeshareImpactActive2015; @hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019].
 As a healthy, inexpensive, and convenient form of public transportation, shared bicycles can encourage individuals to take up bicycling for short local trips or first and last mile trips to other public transportation instead of using personal vehicles.
@@
 Many PBSPs in North America now offer specific programs to address equity that primarily focus on removing cost barriers and increasing access for groups that are under-represented among existing users [@trec2019].
-The locations of docking stations is a common consideration for increasing equity [@howlandCurrentEffortsMake2017].
+The locations of docking stations is a common consideration for increasing access to improve equity [@howlandCurrentEffortsMake2017].
 Hamilton Bike Share (HBS), located in Hamilton, Ontario, was the only Canadian PBSP included in a North American scan of bike share equity programs [see @trec2019].
 HBS was launched in 2015 and currently has over 900 operational bicycles and 130 docking stations.
-An equity program, Everyone Rides Initiative, was implemented in 2018 which expanded the program by introducing twelve "equity" stations to more disadvantaged neighbourhoods in the core service area.
+An equity program, _Everyone Rides Initiative_, was implemented in 2018 which expanded the program by introducing twelve "equity" stations to more disadvantaged neighbourhoods in the core service area.
 
@@
 
-Public bicycle share programs have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users [@fishmanBikeshareReviewRecent2016].
-In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @ogilvieInequalitiesUsagePublic2012; @reillyGenderDisparitiesNew2020; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011].
+Public bicycle share programs have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users to date [@fishmanBikeshareReviewRecent2016].
+In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @ogilvieInequalitiesUsagePublic2012; @reillyGenderDisparitiesNew2020; @wangGenderGapGenerators2019; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011].
 However, one study found that bike share users in Washington, DC were more likely to be female [@buckAreBikeshareUsers2013], which suggests that the gender gap among bicyclists who use PBSPs is less disparate than the gap for personal bicycle use [@fishmanBikeshareReviewRecent2016].
@@
 
-The introduction of PBSPs has been accompanied by a flurry of research focusing on benefits from them.
+The introduction of PBSPs has been accompanied by a flurry of research focusing on who benefits from them.
 Many studies have examined differences in demographics or socioeconomic status between those who use or have access to PBSPs and those who don't, while other research has explored spatial inequities in where stations are located [e.g., @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @hullgrassoBikeShareEquity2020; @qianBikesharingEquityDisadvantaged2020; @qianBikeshareDestinationChoices2021; @smith2015exploring].
-In Chicago, Qian and Jaller [-@qianBikesharingEquityDisadvantaged2020] estimated ridership in the city's PBSP and found that a minority of bike share stations were located in disadvantaged communities, while annual members from such areas had a lower share of trips compared to other areas in the city.
+In Chicago, Qian and Jaller [-@qianBikesharingEquityDisadvantaged2020] estimated ridership in the city's PBSP and found that a minority of docking stations were located in disadvantaged communities, while annual members from such areas had a lower share of trips compared to other areas in the city.
 Similar results were found in Philadelphia.
@@
 Trips from stations in such areas were utilitarian (e.g., commuting to work), which points to the importance of ensuring equitable access [@caspiBikesharingPhiladelphiaLowerincome2019].
-In the case of Seattle, all neighborhoods had some level of access to dockless bicycles but those with higher incomes and more residents of higher education had more bikes [@mooneyFreedomStationSpatial2019].
+In the case of Seattle, all neighbourhoods were found to have some level of access to dockless bicycles but those with higher incomes and more residents of higher education had more bicycles [@mooneyFreedomStationSpatial2019].
 Babagoli et al.
-[-@babagoliExploringHealthSpatial2019] also found that neighborhoods in New York City with higher affluence had the greatest proportion of Citi Bike stations.
+[-@babagoliExploringHealthSpatial2019] also found that neighbourhoods in New York City with higher affluence had the greatest proportion of Citi Bike stations.
 
@@
 [-@chenExploringEquityPerformance2019] distinguish two types of equity related to bike share, horizontal and vertical, based on the work of Delbosc and Currie [@delboscUsingLorenzCurves2011].
-Horizontal equity leads to balanced or equal distribution of accessibility and costs for all similar groups, while vertical equity would involve greater and targeted access for under-represented or disadvantaged populations [see @chenExploringEquityPerformance2019].
+Horizontal equity leads to balanced or equal distribution of accessibility and costs for all similar groups, while vertical equity would involve greater and targeted access for under-represented or disadvantaged populations [@chenExploringEquityPerformance2019].
 Both are of interest to researchers and transportation planners since they are often linked in that advantage, or conversely disadvantage, has spatial patterns.
-Bike share equity programs, as identified by McNeil et al.
-[-@trec2019], can be considered efforts to improve vertical equity since they favour groups that have benefited less from PBSPs.
-
-On the whole, existing studies highlight the need for PBSPs to be more accessible for diverse populations in order to increase use beyond the "typical" users.
+Bike share equity programs [see @trec2019; @@howlandCurrentEffortsMake2017] can be considered efforts to improve vertical equity since they favour groups that have benefited less from PBSPs through the placement of stations or equitable fee structures.
+
+On the whole, existing studies highlight the need for PBSPs to be more accessible for equity populations in order to increase use beyond the "typical" users.
 This has been the focus of recent research [see, among others, @auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020].
 Offering more people the option of using sustainable and active transportation, particularly those who have lower socioeconomic status and might benefit the most, is a worthy policy goal for cities with PBSPs.
-However, exploring transportation equity by investigating where bike share stations are located, often using neighborhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
+However, exploring transportation equity by investigating where docking stations are located, often using neighborhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
 Meaning that, stations may be lacking in certain neighborhoods but there may be stations accessible within a reasonable walking time.
@@
 [-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris, France comes from areas within 300 m of stations, which amounts to 2-4 minutes walking by an adult who does not have a disability.
-Similar to other public amenities affected by crowding, the utility of docking stations is also limited by the _maximum_ number of bicycles that they can hold (e.g., their size).
+Other studies have adopted  [-@hosfordEvaluationImpactPublic2018; -@hosfordEvaluatingImpactImplementing2019] Similar to other public amenities affected by crowding, the utility of docking stations is also limited by the _maximum_ number of bicycles that they can hold (e.g., their size).
 Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts.
@@
 
-Several approaches have been commonly used for measuring place-based accessibility, including cumulative opportunities, gravity, and utility-based measures [@handyMeasuringAccessibilityExploration1997].
+Several approaches have been commonly used for measuring place-based accessibility, including cumulative opportunities, gravity, and utility-based measures [@geursAccessibilityEvaluationLanduse2004; @handyMeasuringAccessibilityExploration1997].
 The gravity-based approach involves weighting destination opportunities, such as the quantity of bike share stations, by the time required to reach them from an origin using an impedance function [@handyMeasuringAccessibilityExploration1997; @kwanSpaceTimeIntegral1998].
@@
 In contrast, floating catchment area (FCA) methods have been widely employed in health care accessibility research.
-The benefit of this approach is that the method incorporates information on capacity, demand, and congestion in calculating accessibility.
-This approach is more appropriate and informative than calculating provider-to-population ratios (PPR) that simply divide the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019].
+The benefit of this approach is that the method incorporates information on capacity and demand in calculating accessibility.
+Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] note "competition for activities with restricted capacity" should be taken into account for land-use components of accessibility.
+The FCA approach is more appropriate and informative than calculating provider-to-population ratios (PPR) that simply divide the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019].
 In particular, the Two-Step Floating Catchment Area (2SFCA) method [@luoMeasuresSpatialAccessibility2003; @radkeSpatialDecompositionsModeling2000] produces flexible catchment areas instead of using rigid boundaries like PPR.
@@
 Hamilton Bike Share Inc., the non-profit operator, launched an equity program, Everyone Rides Initiative (ERI), in 2018 with the objective of reducing barriers that may prevent individuals from accessing the program.
-Additional bicycles and twelve "equity" stations were added to the core service area in more disadvantaged neighbourhoods.
+Additional bicycles and twelve "equity" stations were added to the core service area in more disadvantaged and under-serviced neighbourhoods.
 The equity program also offers subsidized memberships to individuals who identify as low income, and complements this service with cycle skills education.
@@
 The BFCA method was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by earlier FCA methods [@paezDemandLevelService2019].
-Paez et al.
+Páez et al.
 [-@paezDemandLevelService2019] adjusted the impedance weights so that both supply and demand are proportionally allocated.
@@
 
-The balanced approach of Paez et al. [-@paezDemandLevelService2019] replaces the spatial weights with normalized versions as follows. In the first step, the population is weighted with:
+The balanced approach of Páez et al. [-@paezDemandLevelService2019] replaces the spatial weights with normalized versions as follows. In the first step, the population is weighted with:
 \begin{equation}
@@
 Consensus regarding the distance that individuals are willing to walk to access a docking station is lacking, but the literature on walking behaviour provided some guidelines to determine the thresholds for our sensitivity analysis.
-Previous studies have found that living within 250 m [@fullerUseNewPublic2011] and 300 m [@kabraBikeShareSystemsAccessibility2020] of a docking station is correlated with bike share use, while other research has found that walking trips are less than 600 m and rarely more than 1200 m [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 m [@larsenQuarterMileReexamining2010].
+Previous studies have found that living within 250 m [@fullerUseNewPublic2011] and 300 m [@kabraBikeShareSystemsAccessibility2020] of a docking station is correlated with bike share use.
+Some studies have defined "close proximity" to a docking station as 500 metres or less [@fullerImpactEvaluationPublic2013; @hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019], while other research has found that walking trips are less than 600 m and rarely more than 1200 m [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 m [@larsenQuarterMileReexamining2010].
 HBS will often depict a map at some docking stations to show the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are expected to walk.
@@
 
-Using disaggregated population data, we examined equity in accessibility to Hamilton Bike Share, with a particular focus on assessing the contribution of the program's equity stations.
+Using disaggregated population data, we examined equity in accessibility to Hamilton Bike Share (HBS), with a focus on assessing the contribution of the program's equity stations.
 The BFCA approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro scale which better reflects the scale at which walking takes place and avoids the "absorption of disparities", as articulated by Chen et al.
 [-@chenExploringEquityPerformance2019] and elsewhere in the transportation literature [@nationalacademiesofsciencesEffectiveMethodsEnvironmental2004; @rowangouldIdentifyingEnvironmentalJustice2016].
-Our method also considers demand and supply in the calculation of accessibility, which is important for services like PBSPs but not accounted for in other common accessibility models.
+Our method also considers potential demand and supply, as well as congestion effects, in the calculation of accessibility, which is important for services like PBSPs but not accounted for in other common accessibility models.
 Unlike the 2SFCA method, there was no inflation or deflation in our accessibility estimates because population and level of service were preserved (see Figure \ref{fig:figure-10} and Figure \ref{fig:figure-11}).
-This differentiates our analysis from similar papers exploring equity in PBSPs that use larger geographical units of analysis or that focus on station location instead of level of service.
-In this way, our paper has made an important contribution by applying an intuitive and useful approach to measure accessibility to a PBSP.
+This differentiates our analysis from similar papers exploring equity in PBSPs that use larger geographical units of analysis or that focus only on station location instead of level of service.
+In this way, our paper has made an important contribution by applying an intuitive and useful approach not previously used in the cycling literature to measure accessibility to a PBSP.
 This open and reproducible research could be expanded for additional analysis using other individual-level data in Hamilton (e.g., age, gender, household size, education, etc.) or applied to PBSPs in other North American cities.
@@
 This reflects the normative guide advertised on some docking stations in Hamilton showing other stations within a five minute walk, as well as the directive of NACTO [@nactowalkingstation2015].
-We find that over 118,000 people can access a bike share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}].
+We found that over 118,000 people can access a bike share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}].
 At a minimum threshold of three minutes, too few people can reach stations which leads to relatively high levels of service since there is little crowding for the stations.
@@
 With a service area of 40 sq.km, it is estimated that Hamilton should have between 380 and 440 stations instead of 130, and 1,500 bicycles instead of 900 [@hamiltonsobi2015].
-Reduced capacity within the system led to gaps in coverage in some areas of the city "with some areas not having the recommended station density of 300m between stations or 10 stations per square km" [@hamiltonsobi2015].
-Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in service were filled by the additional equity stations.
+The City acknowledged that reduced capacity within the system led to gaps in coverage in some areas of the city "with some areas not having the recommended station density of 300m between stations or 10 stations per square km" [@hamiltonsobi2015].
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in accessibility were filled by the additional equity stations implemented by Hamilton Bike Share.
 The improvement in accessibility is most noticeable at the average and maximum thresholds in the east end of the core service area, which corresponds to the neighbourhoods where stations were added.
@@
 
-The equity stations were added to HBS as a targeted initiative to increase access for disadvantaged communities.
+The equity stations were added to HBS as a targeted initiative to increase access for equity users in specific under-serviced neighbourhoods.
 However, we found that inequities persisted as evidenced by differences in accessibility according to income quintile (Table \ref{tab:accessibility-income}).
 While the addition of equity stations seems to modestly increase accessibility for all income groups at all thresholds, they did not increase accessibility substantially for any single income group.
-Most importantly, individuals in the bottom 20% of median total household income have the second lowest level of access to Hamilton Bike Share at most thresholds (average, maximum, and extreme).
+Most importantly, individuals in the bottom 20% of median total household income have the second lowest level of access to HBS at most thresholds (average, maximum, and extreme).
 At the minimum threshold, the bottom 20% have the lowest level of access.
-This suggests that the equity stations may not have sufficiently addressed vertical equity.
+This suggests that additional equity stations are needed to more sufficiently address vertical equity.
 While previous research found that neighborhoods with more disadvantage are better serviced by HBS, the authors used the Pampalon Deprivation Index to determine the level of disadvantage for dissemination areas _across_ the city not just within the core service area [@hosfordWhoArePublic2018].
 Instead, we used median total household income for each dissemination area _within_ the core service area.
-We conclude that Hamilton's PBSP, while by default located in areas with more deprivation compared to other cities, continues to have disparities in accessibility between income groups.
-With equity stations, many areas with low median total household income in the east end of the service area continue to have zero or low accessibility at the average threshold of five minutes.
-At the maximum and extreme thresholds, the top 20% have the highest level of access to Hamilton Bike Share.
-These findings align with other studies from Tampa [@chenExploringEquityPerformance2019], Philadelphia [@caspiBikesharingPhiladelphiaLowerincome2019] and Seattle [@mooneyFreedomStationSpatial2019], which have found disparities in station location, annual trips, or access to bicycles, respectively, between levels of income and education.
+We conclude that Hamilton's PBSP, while by default located in areas with more deprivation compared to other cities as noted by Hosford and Winters [-@hosfordWhoArePublic2018], continues to have disparities in accessibility between income groups.
+With equity stations, many areas with low median total household income in the east end of the service area have zero or low accessibility at the average threshold of five minutes.
+At the maximum and extreme thresholds, the top 20% have the highest level of access to HBS.
+These findings align broadly with other studies from Tampa [@chenExploringEquityPerformance2019], Philadelphia [@caspiBikesharingPhiladelphiaLowerincome2019] and Seattle [@mooneyFreedomStationSpatial2019], which have found disparities in station location, annual trips, or access to bicycles, respectively, between levels of income and education.
 With the addition of equity stations, there were large gains in accessibility for the second 20% at the average threshold, amounting to over 5,000 more people, but much smaller gains for the bottom 20% with only 326 more people who are able to access a HBS station.
@@
 This paper highlights the importance of both the location and size of docking stations for increasing equity in accessibility to PBSPs.
+Congestion effects for services like bike share that offer a potential maximum supply of bicycles (i.e., number of racks) at each docking station must be taken into account when measuring accessibility.
 Several studies have examined who can access PBSPs based on the location of stations [among others, see @babagoliExploringHealthSpatial2019; @qianBikesharingEquityDisadvantaged2020; @qianEnhancingEquitableService2020; @wangNewBikeShare2019], but the size of docking stations in disadvantaged communities is not considered in the literature.
-Station size (i.e., potential maximum supply of bicycles) was also not identified by PBSP owners and operators as a factor that needed to be addressed in equity efforts [@howlandCurrentEffortsMake2017].
-PBSPs may not necessarily reduce inequities in access if stations are easy to reach for disadvantaged groups but offer only a small number of bicycles.
+An exception is a study on bike share ridership in New York City that found "the number of docks at a bike share station is statistically significant and positively associated with trip attractions for both men and women" [@wangGenderGapGenerators2019, p.5].
+Station size was also not identified by PBSP owners and operators as a factor that needed to be addressed in equity efforts [@howlandCurrentEffortsMake2017].
+PBSPs may not necessarily reduce inequities in access if stations are easy to reach for equity users but offer only a small number of bicycles.
 Likewise, more people may not opt to use the program if the supply of bicycles available at nearby stations is insufficient to meet demand.
@@
 
-The following policy and planning implications can inform efforts for increasing equity in Hamilton Bike Share, and potentially other PBSPs in North America.
+The following policy and planning implications can inform efforts for increasing equity in HBS, and potentially other PBSPs in North America.
 Our analysis has identified specific areas that have both low accessibility and low median total household income.
-Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} highlight potential locations for new equity stations to better accommodate low income groups.
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} highlight potential locations for new equity stations to increase access for low-income groups.
 Additional stations are needed to further reduce inequities in accessibility.
 This information will be useful to transportation planners in Hamilton and HBS operators.
-Our findings also suggest that PBSPs should consider increasing the size of stations in disadvantaged communities.
-Docking stations added to target vertical equity should have an adequate supply of public bicycles, particularly more in areas that have a strong interest in adopting bike share.
+Our findings also suggest that PBSPs should consider increasing the size of stations in targeted communities.
+Notably, figure \ref{fig:sobi-stations-in-hamilton} highlights how the equity stations in Hamilton have a lower number of racks compared to docking stations in the city's downtown core.
+Docking stations added to increase vertical equity should have an adequate supply of public bicycles, particularly more in areas that have a strong interest in adopting bike share, to increase levels of service and potentially attract new users.
 Finally, distance to stations is associated with bike share use [@fullerUseNewPublic2011; @wangNewBikeShare2019], especially when the trip starts at the residence [@fishmanBikeShareSynthesis2013].
@@
 This paper did not examine or compare ridership data between conventional and equity stations.
-Therefore, further research is needed to determine whether the addition of equity stations encouraged more bicycling for disadvantaged groups living near them.
+Therefore, further research is needed to determine whether the addition of equity stations encouraged more bicycling for equity users living near them.
 Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighborhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of publicly available route and individual user data to conduct similar analyses for HBS.
@@
 Ideally, instead of number of bicycle racks as our measure of supply, we would have liked to use the average number of bicycles at stations, perhaps at different times during the day across many seasons.
-Should this data become available, it would be worthwhile to revisit the question to examine how well the operation of the system (including balancing of bicycles across stations) works to maintain the nominal levels of accessibility examined in this paper.
+Should this data become available, it would be worthwhile to revisit the study to examine how well the operation of the system (including balancing of bicycles across stations) works to maintain the nominal levels of accessibility examined in this paper.
 
@@
 Wang and Lindsey [-@wangNewBikeShare2019] have noted that there is a lack of research that examines how bike share users' behaviour changes as a result of program changes to station locations or improvements in accessibility.
-As such, a logical next step to this research is to examine whether the equity stations increased ridership or resulted in new memberships in areas that were previously under-serviced.
-An examination of the types of trips undertaken by residents in these areas would also be informative, such as the study undertaken by Caspi and Norland [-@caspiBikesharingPhiladelphiaLowerincome2019] after bike share stations were implemented in low-income Philadelphia neighborhoods.
-The bulk of cycling facilities that have been built in Hamilton to date are located in the core service area near the conventional stations.
+As such, a logical next step to this research is to examine whether the equity stations increased ridership or resulted in new memberships for equity users.
+An examination of the types of trips undertaken by residents in these areas would also be informative, such as the study undertaken by Caspi and Norland [-@caspiBikesharingPhiladelphiaLowerincome2019] after docking stations were implemented in low-income Philadelphia neighborhoods.
+The bulk of bicycle infrastructure that has been built in Hamilton to date is located in the core service area particularly near the conventional stations.
 It would be worthwhile to explore the route choice of bike share trips departing or ending at the equity stations and to identify factors that specifically influence trips from these stations, which would extend existing studies conducted by Scott and colleagues [@luUnderstandingBikeShare2018; @scottWhatFactorsInfluence2019; @scottRouteChoiceBike2021].
-This paper, combined with additional studies such as those conceptualized above, would serve as a valuable case study for Hamilton and other cities with PBSPs that wish to evaluate and address inequities in accessibility and transportation options in urban areas.
+Perceived accessibility using bike share according to low-income neighbourhoods could also be studied in the future.
+Lättman et al.
+[-@lattmanDevelopmentTestPerceived2016] have developed the _Perceived Accessibility Scale_ that would be useful for this purpose.
+This paper, combined with additional studies such as those conceptualized above, would serve as a valuable case study for Hamilton and other cities with PBSPs that wish to evaluate and address inequities in accessibility to active transportation options in urban areas.
 

### 2021-08-30 Bike-share-spatial-equity/Response-to-Reviewers-R2.Rmd
Update Response-to-Reviewers-R2.Rmd

@@
 Floating catchment areas introduce an element of demand (or congestion).
-We clarified in the text that demand in our case is at the docking stations.
+We clarified in the text that demand in our case is at the docking stations (see _page_).
 You are right that the analysis is static in that it does not account for temporal variations in demand.
-Unfortunately, the location of bicycles throughout the day is not available publicly, and we are not able at the moment to implement a dynamic analysis.
-We note this as a direction for future research.
-With respect to the static analysis, it is true that not all travellers arrive at a station at the same time.
+Unfortunately, the number of bicycles available at each docking station at various times throughout the day is not available publicly, and we are not able to implement a dynamic analysis at the moment.
+We note this as a direction for future research (see _page_).
+With respect to the static analysis, it is true that not all users arrive at a docking station at the same time.
 However, as a standard, this is still a valid assumption from a systems design perspective.
-Contrast to standards in terms of area of parks per population: not everyone goes to the park at the same time, but calculating area of park per person is still a useful indicator of supply and demand.
-Or standards for health care provision: not everyone goes to an emergency room at the same time, but calculating ER beds per person is still a good indicator of supply and demand.
-In the case of accessibility to bike share systems discussing the capacity of the system in these terms may be surprising initially, but we'd argue this is because no standards have been developed for this kind of public service.}
+Contrast to standards in terms of area of park per population: not everyone goes to the park at the same time, but calculating area of park per person is still a useful indicator of supply and demand.
+Similar standards for health care provision exist: not everyone goes to an emergency room at the same time, but calculating ER beds per person is still a good indicator of supply and demand.
+In the case of accessibility to public bicycle share programs, discussing the capacity of the system in these terms may be surprising initially, but we'd argue that this is because no standards have been developed for this kind of public service.}
 
@@
 \textcolor{blue}{We are sorry that this was unclear.
-In section 4.1 we mean population units in general (they could be census polygons, interpolated populatation, or any other carographic representation of the population.
-We change the text to say "population unit" to ensure that this is not confused with the population cells in section 4.2.
-These populations cells are the 50x50m polygons that we use to interpolate the population from census Dissemination Areas to smaller geographical units.
-For vertical equity analysis we use census income data at the level of Dissemination Areas.
-In figures 9-12 accessibility is seen at the level of the 50x50m population cells, and income is identical for all cells in the same DA.
-In other words, accessibility is still spatially disaggregated but income is for DAs.
-In Table 1 we do aggregate accessibility, but this aggregation is not by geography but rather by income level.
+In section 4.1, we mean population units in general; they could be census polygons, dissemination areas, interpolated populatation, or any other cartographic representation of the population.
+We have changed the text to say "population unit" (see _page_) to ensure that this is not confused with the population cells in section 4.2.
+These populations cells are the 50x50m polygons that we used to interpolate the population from census dissemination areas (DAs) to smaller geographical units.
+For vertical equity analysis, we used census income data at the level of DAs.
+In figures 9-12, accessibility is seen at the level of the 50x50m population cells, and income is identical for all cells in the same DA.
+In other words, accessibility is still spatially disaggregated but income is from the DAs.
+In Table 1, we did aggregate accessibility, but this aggregation is not by geography but rather by income level.
 A major advantage of the BFCA approach is that this kind of aggregation is possible while still preserving the population and level of service.
@@
 
-\textcolor{blue}{Very good point. We assumed that the distance decay function would be apparent from the reference to thresholds. We now note that the distance decay is as follows:}
+\textcolor{blue}{Very good point. We assumed that the distance decay function would be apparent from the reference to thresholds. We now note on _page_ that the distance decay is as follows:}
 
@@
 Imagine a station with a single bicycle.
-A single person is located at 3 minutes from that stations.
+A single person is located at 3 minutes from that station.
 Then, a single person is located at 4 minutes from that station.
@@
 Now two people are in the service area of the station.
-The level of service of the station is 1/2 bicycles per person.
+The level of service of the station is 1/2 bicycle per person.
 These two people now collectively have lower accessibility, since they can reach only a lower level of service.
-Finally, suppose that the catchment of the station includes all three people: the level of service now is 1/3 bicycles per person, and collectively the accessibility is even lower.
+Finally, suppose that the catchment of the station includes all three people: the level of service now is 1/3 bicycle per person, and collectively the accessibility is even lower.
 Now, perhaps the confusion is about what accessibility means in this framework.
@@
 When the catchment area is small, few people enjoy relatively high levels of accessibility; when the catchment area is large, more people enjoy accessibility, but only to lower levels of service.
-We discuss the motivation for using this approach in our response to your first point.}
+We discuss the motivation for using this approach in our response to your first point.
+Geurs and van Wee (2004), as cited in the paper, note that land-use components of accessibility measures take congestion effects into account.}
 

### 2021-08-30 Bike-share-spatial-equity/Response-to-Reviewers-R2.Rmd
Proof reading Elise's edits before resubmitting R2; created document with marked changes between versions

@@
 Floating catchment areas introduce an element of demand (or congestion).
-We clarified in the text that demand in our case is at the docking stations (see _page_).
+We clarified in the text that demand in our case is at the docking stations (see \emph{page}).
 You are right that the analysis is static in that it does not account for temporal variations in demand.
 Unfortunately, the number of bicycles available at each docking station at various times throughout the day is not available publicly, and we are not able to implement a dynamic analysis at the moment.
-We note this as a direction for future research (see _page_).
+We note this as a direction for future research (see \emph{page}).
 With respect to the static analysis, it is true that not all users arrive at a docking station at the same time.
@@
 In section 4.1, we mean population units in general; they could be census polygons, dissemination areas, interpolated populatation, or any other cartographic representation of the population.
-We have changed the text to say "population unit" (see _page_) to ensure that this is not confused with the population cells in section 4.2.
+We have changed the text to say "population unit" (see \emph{page}) to ensure that this is not confused with the population cells in section 4.2.
 These populations cells are the 50x50m polygons that we used to interpolate the population from census dissemination areas (DAs) to smaller geographical units.
@@
 
-\textcolor{blue}{Very good point. We assumed that the distance decay function would be apparent from the reference to thresholds. We now note on _page_ that the distance decay is as follows:}
+\textcolor{blue}{Very good point. We assumed that the distance decay function would be apparent from the reference to thresholds. We now note on \emph{page} that the distance decay is as follows:}
 

### 2021-08-30 Bike-share-spatial-equity/Bike-share-spatial-equity-R1.tex
Proof reading Elise's edits before resubmitting R2; created document with marked changes between versions

--- 2021-08-23 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
@@
 % Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
-  \title{Examining equity in accessibility to bike share: a balanced floating
-catchment area approach}
+  \title{Examining equity in accessibility to bike share: a balanced
+floating catchment area approach}
     \author[McMaster University]{Elise Desjardins\corref{1}}
@@
 spatial inequities in where stations are located (e.g., Hosford and
-Winters, 2018; Hull Grasso et al., 2020; Mooney et al., 2019; Qian and
-Jaller, 2021, 2020; Smith et al., 2015). In Chicago, Qian and Jaller
-(2020) estimated ridership in the city's PBSP and found that a minority
-of bike share stations were located in disadvantaged communities, while
-annual members from such areas had a lower share of trips compared to
-other areas in the city. Similar results were found in Philadelphia.
-Despite efforts to increase equity within the city's PBSP, census block
-groups with lower median income generated fewer trips (Caspi and Noland,
-2019). Trips from stations in such areas were utilitarian (e.g.,
-commuting to work), which points to the importance of ensuring equitable
-access (Caspi and Noland, 2019). In the case of Seattle, all
-neighborhoods had some level of access to dockless bicycles but those
-with higher incomes and more residents of higher education had more
-bikes (Mooney et al., 2019). Babagoli et al.~(2019) also found that
+Winters, 2018; Mooney et al., 2019; Qian and Jaller, 2021, 2020; Smith
+et al., 2015; \textbf{hullgrassoBikeShareEquity2020?}). In Chicago, Qian
+and Jaller (2020) estimated ridership in the city's PBSP and found that
+a minority of bike share stations were located in disadvantaged
+communities, while annual members from such areas had a lower share of
+trips compared to other areas in the city. Similar results were found in
+Philadelphia. Despite efforts to increase equity within the city's PBSP,
+census block groups with lower median income generated fewer trips
+(Caspi and Noland, 2019). Trips from stations in such areas were
+utilitarian (e.g., commuting to work), which points to the importance of
+ensuring equitable access (Caspi and Noland, 2019). In the case of
+Seattle, all neighborhoods had some level of access to dockless bicycles
+but those with higher incomes and more residents of higher education had
+more bikes (Mooney et al., 2019). Babagoli et al. (2019) also found that
 neighborhoods in New York City with higher affluence had the greatest
@@
 ``typical'' users. This has been the focus of recent research (see,
-among others, Auchincloss et al., 2020; Hull Grasso et al., 2020;
-MacArthur et al., 2020). Offering more people the option of using
-sustainable and active transportation, particularly those who have lower
-socioeconomic status and might benefit the most, is a worthy policy goal
-for cities with PBSPs. However, exploring transportation equity by
-investigating where bike share stations are located, often using
-neighborhoods or census tracts as the geographical unit of analysis, can
-ignore or miss the benefits that may be derived from adjacent zones.
-Meaning that, stations may be lacking in certain neighborhoods but there
-may be stations accessible within a reasonable walking time. This is
-where geographical accessibility becomes an important consideration.
+among others, Auchincloss et al., 2020; MacArthur et al., 2020;
+\textbf{hullgrassoBikeShareEquity2020?}). Offering more people the
+option of using sustainable and active transportation, particularly
+those who have lower socioeconomic status and might benefit the most, is
+a worthy policy goal for cities with PBSPs. However, exploring
+transportation equity by investigating where bike share stations are
+located, often using neighborhoods or census tracts as the geographical
+unit of analysis, can ignore or miss the benefits that may be derived
+from adjacent zones. Meaning that, stations may be lacking in certain
+neighborhoods but there may be stations accessible within a reasonable
+walking time. This is where geographical accessibility becomes an
+important consideration.
 
@@
 docking station decreases the potential of accessing the program, the
-location matters. Kabra et al.~(2020) found that the majority of bike
+location matters. Kabra et al. (2020) found that the majority of bike
 share usage in Paris, France comes from areas within 300 m of stations,
@@
 neighborhoods or census tracts (Babagoli et al., 2019; Mooney et al.,
-2019; Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), although there
+2019; Qian and Jaller, 2020; Wang and Lindsey, 2019a), although there
 are recent exceptions (Chen et al., 2019; Chen and Li, 2021). This is
-also true for studies examining correlates of bike share demand (J. Wang
+also true for studies examining correlates of bike share demand (Wang
 and Lindsey, 2019b). Handy and Niemeier (1997) note that using
 disaggregated data in accessibility analyses provides a more accurate
-estimate for individuals. Chen et al.~(2019, p. 530) are in favour of
+estimate for individuals. Chen et al. (2019, p. 530) are in favour of
 using disaggregated data, which they did in their recent analysis of
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/hamilton-and-sobi-service-area-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/hamilton-and-sobi-service-area-1}
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{Bike-share-spatial-equity_files/figure-latex/sobi-stations-in-hamilton-1}
+{\centering \includegraphics[width=1\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/sobi-stations-in-hamilton-1}
 
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/interpolated-population-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/interpolated-population-1}
 
@@
 All data for this research were accessed from publicly available Census
-of Canada sources, from OpenStreetMaps, and from Open Hamilton\footnote{https://open.hamilton.ca/},
-a public online repository of data curated by the City of Hamilton.
-Median total household income statistics were drawn from the 2016
-Canadian Census.
+of Canada sources, from OpenStreetMaps, and from Open
+Hamilton\footnote{https://open.hamilton.ca/}, a public online repository
+of data curated by the City of Hamilton. Median total household income
+statistics were drawn from the 2016 Canadian Census.
 
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-7-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-7-1}
 
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-8-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-8-1}
 
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-9-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-9-1}
 
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-10-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-10-1}
 
@@
 
-{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-11-1}
+{\centering \includegraphics[width=0.9\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-11-1}
 
@@
 \begin{figure}
-\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income at the minimum threshold of three minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-3}
+\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-bi-map-threshold-3-1} \caption{\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income at the minimum threshold of three minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-3}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income at the average threshold of five minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-5}
+\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-bi-map-threshold-5-1} \caption{\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income at the average threshold of five minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-5}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income at the maximum threshold of ten minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-10}
+\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-bi-map-threshold-10-1} \caption{\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income at the maximum threshold of ten minutes with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-10}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income at the extreme threshold of fifteen with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-15}
+\includegraphics[width=1.2\linewidth]{Bike-share-spatial-equity-R1_files/figure-latex/figure-bi-map-threshold-15-1} \caption{\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income at the extreme threshold of fifteen with equity stations (top panel) and without equity stations (bottom panel).}\label{fig:figure-bi-map-threshold-15}
 \end{figure}
@@
 accessibility on a micro scale which better reflects the scale at which
-walking takes place and avoids the ``absorption of disparities'', as
-articulated by Chen et al.~(2019) and elsewhere in the transportation
+walking takes place and avoids the ``absorption of disparities,'' as
+articulated by Chen et al. (2019) and elsewhere in the transportation
 literature (National Academies of Sciences, 2004; Rowangould et al.,
@@
 of stations (among others, see Babagoli et al., 2019; Qian and Jaller,
-2020; Qian et al., 2020; J. Wang and Lindsey, 2019b), but the size of
-docking stations in disadvantaged communities is not considered in the
+2020; Wang and Lindsey, 2019b;
+\textbf{qianEnhancingEquitableService2020?}), but the size of docking
+stations in disadvantaged communities is not considered in the
 literature. Station size (i.e., potential maximum supply of bicycles)
@@
 interest in adopting bike share. Finally, distance to stations is
-associated with bike share use (Fuller et al., 2011; J. Wang and
-Lindsey, 2019b), especially when the trip starts at the residence
-(Fishman et al., 2013). Distance can also be a barrier to using PBSPs
-(Fishman et al., 2014). To address this issue, transportation planners
-and PBSP operators should consult with groups and communities who are
+associated with bike share use (Fuller et al., 2011; Wang and Lindsey,
+2019b), especially when the trip starts at the residence (Fishman et
+al., 2013). Distance can also be a barrier to using PBSPs (Fishman et
+al., 2014). To address this issue, transportation planners and PBSP
+operators should consult with groups and communities who are
 under-represented among bike share users to determine how far they are
@@
 2021a), \texttt{pycno} (Brunsdon, 2014), \texttt{r5r} (Saraiva et al.,
-2021), \texttt{raster} (Hijmans, 2020), \texttt{rgdal} (Bivand et al.,
+2021), \texttt{raster} (Hijmans, 2021), \texttt{rgdal} (Bivand et al.,
 2021), \texttt{rticles} (Allaire et al., 2021), \texttt{sf} (Pebesma,
@@
 \leavevmode\hypertarget{ref-breyWantRideMy2017}{}%
-Brey, R., Castillo-Manzano, J.I., Castro-Nuño, M., 2017. ``I want to
-ride my bicycle'': Delimiting cyclist typologies. Applied Economics
-Letters 24, 549--552.
+Brey, R., Castillo-Manzano, J.I., Castro-Nuño, M., 2017. 'I want to ride
+my bicycle': Delimiting cyclist typologies. Applied Economics Letters
+24, 549--552.
 doi:\href{https://doi.org/10.1080/13504851.2016.1210760}{10.1080/13504851.2016.1210760}
@@
 Buck, D., Buehler, R., Happ, P., Rawls, B., Chung, P., Borecki, N.,
-2013. Are Bikeshare Users Different from Regular Cyclists?: A First Look
-at Short-Term Users, Annual Members, and Area Cyclists in the
-Washington, D.C., Region. Transportation Research Record 2387, 112--119.
+2013. Are bikeshare users different from regular cyclists?: A first look
+at short-term users, annual members, and area cyclists in the
+washington, d.c., region. Transportation Research Record 2387, 112--119.
 doi:\href{https://doi.org/10.3141/2387-13}{10.3141/2387-13}
@@
 \leavevmode\hypertarget{ref-R-raster}{}%
-Hijmans, R.J., 2020. Raster: Geographic data analysis and modeling.
+Hijmans, R.J., 2021. Raster: Geographic data analysis and modeling.
 
@@
 Howland, S., McNeil, N., Broach, J., Rankins, K., MacArthur, J., Dill,
-J., 2017. Current Efforts to Make Bikeshare More Equitable: Survey of
-System Owners and Operators. Transportation Research Record 2662,
+J., 2017. Current {Efforts} to {Make Bikeshare More Equitable}: {Survey}
+of {System Owners} and {Operators}. Transportation Research Record 2662,
 160--167. doi:\href{https://doi.org/10.3141/2662-18}{10.3141/2662-18}
 
-\leavevmode\hypertarget{ref-hullgrassoBikeShareEquity2020}{}%
-Hull Grasso, S., Barnes, P., Chavis, C., 2020. Bike Share Equity for
-Underrepresented Groups: Analyzing Barriers to System Usage in
-Baltimore, Maryland. Sustainability 12, 7600.
-doi:\href{https://doi.org/10.3390/su12187600}{10.3390/su12187600}
-
 \leavevmode\hypertarget{ref-kabraBikeShareSystemsAccessibility2020}{}%
-Kabra, A., Belavina, E., Girotra, K., 2020. Bike-Share Systems:
-Accessibility and Availability. Management Science 66, 3803--3824.
+Kabra, A., Belavina, E., Girotra, K., 2020. Bike-{Share Systems}:
+{Accessibility} and {Availability}. Management Science 66, 3803--3824.
 doi:\href{https://doi.org/10.1287/mnsc.2019.3407}{10.1287/mnsc.2019.3407}
@@
 \leavevmode\hypertarget{ref-nationalacademiesofsciencesEffectiveMethodsEnvironmental2004}{}%
-National Academies of Sciences, E., 2004. Effective Methods for
-Environmental Justice Assessment.
+National Academies of Sciences, E., 2004. Effective {Methods} for
+{Environmental Justice Assessment}.
 doi:\href{https://doi.org/10.17226/13694}{10.17226/13694}
@@
 
-\leavevmode\hypertarget{ref-qianEnhancingEquitableService2020}{}%
-Qian, X., Jaller, M., Niemeier, D., 2020. Enhancing equitable service
-level: Which can address better, dockless or dock-based Bikeshare
-systems? Journal of Transport Geography 86, 102784.
-doi:\href{https://doi.org/10.1016/j.jtrangeo.2020.102784}{10.1016/j.jtrangeo.2020.102784}
-
 \leavevmode\hypertarget{ref-radkeSpatialDecompositionsModeling2000}{}%
-Radke, J., Mu, L., 2000. Spatial Decompositions, Modeling and Mapping
-Service Regions to Predict Access to Social Programs. Geographic
-Information Sciences 6, 105--112.
+Radke, J., Mu, L., 2000. Spatial {Decompositions}, {Modeling} and
+{Mapping Service Regions} to {Predict Access} to {Social Programs}.
+Geographic Information Sciences 6, 105--112.
 doi:\href{https://doi.org/10.1080/10824000009480538}{10.1080/10824000009480538}

### 2021-08-30 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Proof reading Elise's edits before resubmitting R2; created document with marked changes between versions

@@
    \ead{cd.higgins@utoronto.ca} 
-    \author[McMaster University]{Antonio Paez}
+    \author[McMaster University]{Antonio Páez}
    \ead{paezha@mcmaster.ca} 
-      \address[McMaster University]{School of Earth, Environment \& Society, McMaster University, 1280 Main
-Street West, Hamilton, ON L8S4L8}
-    \address[University of Toronto Scarborough]{Department of Geography \& Planning, University of Toronto Scarborough,
-1265 Military Trail, Toronto, ON M1C1A4}
+      \address[McMaster University]{School of Earth, Environment \&
+Society, McMaster University, 1280 Main Street West, Hamilton, ON
+L8S4L8}
+    \address[University of Toronto Scarborough]{Department of Geography
+\& Planning, University of Toronto Scarborough, 1265 Military Trail,
+Toronto, ON M1C1A4}
       \cortext[1]{Corresponding Author}
@@
   transportation equity if they make bicycling more accessible to
-  disadvantaged populations. In Ontario, the City of Hamilton expanded
-  their PBSP in 2018 by adding twelve ``equity'' stations with the
+  disadvantaged populations. In Ontario, Hamilton Bike Share expanded
+  their program in 2018 by adding twelve ``equity'' stations with the
   explicit objective of increasing access for under-serviced
@@
   accessibility to stations using a balanced floating catchment area
-  approach and compare accessibility with and without the equity stations.
-  We analyze micro zones to better reflect walking to a station and
-  conduct a sensitivity analysis at several walking time thresholds. We
-  then reaggregate the estimated accessibility for further analysis using
-  census data. Our findings indicate that equity stations increased
-  accessibility for the serviced population at every threshold examined,
-  but the increase was relatively modest especially for population in the
-  bottom 20\% of median total household income.
+  approach and compare accessibility with and without the equity
+  stations. We analyze population interpolated to small cells to better
+  reflect walking to a station and conduct a sensitivity analysis at
+  several walking time thresholds. We then reaggregate the estimated
+  accessibility by income groups for further analysis. Our findings
+  indicate that equity stations increased accessibility for the serviced
+  population at every threshold examined, but the increase was
+  relatively modest especially for population in the bottom 20\% of
+  median total household income.
   \end{abstract}
@@
 bicycling levels is but one of many reasons for implementing such
-programs in urban areas (see Hosford et al., 2018, 2019). As a healthy,
-inexpensive, and convenient form of public transportation, shared
-bicycles can encourage individuals to take up bicycling for short local
-trips or first and last mile trips to other public transportation
-instead of using personal vehicles. These programs can also play a role
-in advancing transportation equity if they make bicycling more
-accessible to disadvantaged populations. Although PBSPs are available to
-the general public in over 800 cities worldwide (Fishman, 2016), and
-ought to be accessible to any individual who wishes to use them,
-research on PBSPs indicates that inequities persist with respect to who
-can use and access them.
+programs in urban areas (see Fishman et al., 2015; Hosford et al., 2019,
+2018). As a healthy, inexpensive, and convenient form of public
+transportation, shared bicycles can encourage individuals to take up
+bicycling for short local trips or first and last mile trips to other
+public transportation instead of using personal vehicles. These programs
+can also play a role in advancing transportation equity if they make
+bicycling more accessible to disadvantaged populations. Although PBSPs
+are available to the general public in over 800 cities worldwide
+(Fishman, 2016), and ought to be accessible to any individual who wishes
+to use them, research on PBSPs indicates that inequities persist with
+respect to who can use and access them.
 
@@
 (McNeil et al., 2019). The locations of docking stations is a common
-consideration for increasing equity (Howland et al., 2017). Hamilton
-Bike Share (HBS), located in Hamilton, Ontario, was the only Canadian
-PBSP included in a North American scan of bike share equity programs
-(see McNeil et al., 2019). HBS was launched in 2015 and currently has
-over 900 operational bicycles and 130 docking stations. An equity
-program, Everyone Rides Initiative, was implemented in 2018 which
-expanded the program by introducing twelve ``equity'' stations to more
-disadvantaged neighbourhoods in the core service area.
+consideration for increasing access to improve equity (Howland et al.,
+2017). Hamilton Bike Share (HBS), located in Hamilton, Ontario, was the
+only Canadian PBSP included in a North American scan of bike share
+equity programs (see McNeil et al., 2019). HBS was launched in 2015 and
+currently has over 900 operational bicycles and 130 docking stations. An
+equity program, \emph{Everyone Rides Initiative}, was implemented in
+2018 which expanded the program by introducing twelve ``equity''
+stations to more disadvantaged neighbourhoods in the core service area.
 
@@
 sources and organized in the form of a data package. Following best
-practices in spatial data science (Brunsdon and Comber, 2020), the code
-and data needed to reproduce, modify or extend the analysis are
-available for download.\footnote{https://github.com/paezha/Accessibility-Sobi-Hamilton}
+practices in spatial data science (Brunsdon and Comber, 2020), an open
+data product (Arribas/Bel et al., 2021) along with the code needed to
+reproduce, modify or extend the analysis are available for
+download.\footnote{https://github.com/paezha/Accessibility-Sobi-Hamilton}
 
@@
 Public bicycle share programs have been implemented in over 800 cities
-worldwide and a great deal has been learned about their typical users
-(Fishman, 2016). In many cities, males use bike share more than females
-(Brey et al., 2017; Nickkar et al., 2019; Ogilvie and Goodman, 2012;
-Reilly, S. M. Wang, et al., 2020; Winters et al., 2019) as do younger
-age cohorts (Brey et al., 2017; Buck et al., 2013; Fuller et al., 2011).
-However, one study found that bike share users in Washington, DC were
-more likely to be female (Buck et al., 2013), which suggests that the
-gender gap among bicyclists who use PBSPs is less disparate than the gap
-for personal bicycle use (Fishman, 2016). There is some evidence that
-bike share users are less likely to own a car (Buck et al., 2013;
-Reilly, Noyes, et al., 2020). However, the relationship between income
-or education and bike share use is less clear-cut. Stations in
-disadvantaged communities in Chicago have been found to generate most of
-the average annual trips (Qian and Jaller, 2020) and individuals from
-minority or lower socioeconomic status neighborhoods in
-Minneapolis-St.~Paul also used the city's PBSP more (J. Wang and
-Lindsey, 2019a). Similar findings were reported in London (Ogilvie and
-Goodman, 2012). Being university educated was a significant correlate of
-bike share use in Montreal, Canada (Fuller et al., 2011). Not
-coincidentally, financial savings have been found to motivate those on a
-low income to use bike share (Fishman, 2016).
+worldwide and a great deal has been learned about their typical users to
+date (Fishman, 2016). In many cities, males use bike share more than
+females (Brey et al., 2017; Nickkar et al., 2019; Ogilvie and Goodman,
+2012; Reilly et al., 2020b; Wang and Akar, 2019; Winters et al., 2019)
+as do younger age cohorts (Brey et al., 2017; Buck et al., 2013; Fuller
+et al., 2011). However, one study found that bike share users in
+Washington, DC were more likely to be female (Buck et al., 2013), which
+suggests that the gender gap among bicyclists who use PBSPs is less
+disparate than the gap for personal bicycle use (Fishman, 2016). There
+is some evidence that bike share users are less likely to own a car
+(Buck et al., 2013; Reilly et al., 2020a). However, the relationship
+between income or education and bike share use is less clear-cut.
+Stations in disadvantaged communities in Chicago have been found to
+generate most of the average annual trips (Qian and Jaller, 2020) and
+individuals from minority or lower socioeconomic status neighborhoods in
+Minneapolis-St.~Paul also used the city's PBSP more (Wang and Lindsey,
+2019a). Similar findings were reported in London (Ogilvie and Goodman,
+2012). Being university educated was a significant correlate of bike
+share use in Montreal, Canada (Fuller et al., 2011). Not coincidentally,
+financial savings have been found to motivate those on a low income to
+use bike share (Fishman, 2016).
 
@@
 The introduction of PBSPs has been accompanied by a flurry of research
-focusing on benefits from them. Many studies have examined differences
-in demographics or socioeconomic status between those who use or have
-access to PBSPs and those who don't, while other research has explored
-spatial inequities in where stations are located (e.g., Hosford and
-Winters, 2018; Hull Grasso et al., 2020; Mooney et al., 2019; Qian and
-Jaller, 2021, 2020; Smith et al., 2015). In Chicago, Qian and Jaller
-(2020) estimated ridership in the city's PBSP and found that a minority
-of bike share stations were located in disadvantaged communities, while
-annual members from such areas had a lower share of trips compared to
-other areas in the city. Similar results were found in Philadelphia.
-Despite efforts to increase equity within the city's PBSP, census block
-groups with lower median income generated fewer trips (Caspi and Noland,
-2019). Trips from stations in such areas were utilitarian (e.g.,
-commuting to work), which points to the importance of ensuring equitable
-access (Caspi and Noland, 2019). In the case of Seattle, all
-neighborhoods had some level of access to dockless bicycles but those
-with higher incomes and more residents of higher education had more
-bikes (Mooney et al., 2019). Babagoli et al.~(2019) also found that
-neighborhoods in New York City with higher affluence had the greatest
-proportion of Citi Bike stations.
-
-Chen et al.~(2019) distinguish two types of equity related to bike
+focusing on who benefits from them. Many studies have examined
+differences in demographics or socioeconomic status between those who
+use or have access to PBSPs and those who don't, while other research
+has explored spatial inequities in where stations are located (e.g.,
+Hosford and Winters, 2018; Hull Grasso et al., 2020; Mooney et al.,
+2019; Qian and Jaller, 2021, 2020; Smith et al., 2015). In Chicago, Qian
+and Jaller (2020) estimated ridership in the city's PBSP and found that
+a minority of docking stations were located in disadvantaged
+communities, while annual members from such areas had a lower share of
+trips compared to other areas in the city. Similar results were found in
+Philadelphia. Despite efforts to increase equity within the city's PBSP,
+census block groups with lower median income generated fewer trips
+(Caspi and Noland, 2019). Trips from stations in such areas were
+utilitarian (e.g., commuting to work), which points to the importance of
+ensuring equitable access (Caspi and Noland, 2019). In the case of
+Seattle, all neighbourhoods were found to have some level of access to
+dockless bicycles but those with higher incomes and more residents of
+higher education had more bicycles (Mooney et al., 2019). Babagoli et
+al. (2019) also found that neighbourhoods in New York City with higher
+affluence had the greatest proportion of Citi Bike stations.
+
+Chen et al. (2019) distinguish two types of equity related to bike
 share, horizontal and vertical, based on the work of Delbosc and Currie
@@
 vertical equity would involve greater and targeted access for
-under-represented or disadvantaged populations (see Chen et al., 2019).
-Both are of interest to researchers and transportation planners since
-they are often linked in that advantage, or conversely disadvantage, has
-spatial patterns. Bike share equity programs, as identified by McNeil et
-al.~(2019), can be considered efforts to improve vertical equity since
-they favour groups that have benefited less from PBSPs.
+under-represented or disadvantaged populations (Chen et al., 2019). Both
+are of interest to researchers and transportation planners since they
+are often linked in that advantage, or conversely disadvantage, has
+spatial patterns. Bike share equity programs (Howland et al., 2017; see
+McNeil et al., 2019) can be considered efforts to improve vertical
+equity since they favour groups that have benefited less from PBSPs
+through the placement of stations or equitable fee structures.
 
 On the whole, existing studies highlight the need for PBSPs to be more
-accessible for diverse populations in order to increase use beyond the
+accessible for equity populations in order to increase use beyond the
 ``typical'' users. This has been the focus of recent research (see,
@@
 for cities with PBSPs. However, exploring transportation equity by
-investigating where bike share stations are located, often using
+investigating where docking stations are located, often using
 neighborhoods or census tracts as the geographical unit of analysis, can
@@
 docking station decreases the potential of accessing the program, the
-location matters. Kabra et al.~(2020) found that the majority of bike
+location matters. Kabra et al. (2020) found that the majority of bike
 share usage in Paris, France comes from areas within 300 m of stations,
 which amounts to 2-4 minutes walking by an adult who does not have a
-disability. Similar to other public amenities affected by crowding, the
-utility of docking stations is also limited by the \emph{maximum} number
-of bicycles that they can hold (e.g., their size). Accessibility
-analyses for PBSPs constitute a positive and evaluation-based approach
-that also has the potential to inform equity efforts. For instance, Wang
-and Lindsey (2019b) investigated whether new or relocated bike share
-stations increased accessibility and use, which offered important
-insights to improve the performance of the program.
+disability. Other studies have adopted (2019, 2018) Similar to other
+public amenities affected by crowding, the utility of docking stations
+is also limited by the \emph{maximum} number of bicycles that they can
+hold (e.g., their size). Accessibility analyses for PBSPs constitute a
+positive and evaluation-based approach that also has the potential to
+inform equity efforts. For instance, Wang and Lindsey (2019b)
+investigated whether new or relocated bike share stations increased
+accessibility and use, which offered important insights to improve the
+performance of the program.
 
@@
 accessibility, including cumulative opportunities, gravity, and
-utility-based measures (Handy and Niemeier, 1997). The gravity-based
-approach involves weighting destination opportunities, such as the
-quantity of bike share stations, by the time required to reach them from
-an origin using an impedance function (Handy and Niemeier, 1997; Kwan,
-1998). While such measures are suitable for capturing the potential for
-reaching destinations from a given location, they do not take demand or
-congestion effects into account which is an important consideration when
-calculating accessibility for amenities such as bike share stations.
+utility-based measures (Geurs and van Wee, 2004; Handy and Niemeier,
+1997). The gravity-based approach involves weighting destination
+opportunities, such as the quantity of bike share stations, by the time
+required to reach them from an origin using an impedance function (Handy
+and Niemeier, 1997; Kwan, 1998). While such measures are suitable for
+capturing the potential for reaching destinations from a given location,
+they do not take demand or congestion effects into account which is an
+important consideration when calculating accessibility for amenities
+such as bike share stations.
 
@@
 employed in health care accessibility research. The benefit of this
-approach is that the method incorporates information on capacity,
-demand, and congestion in calculating accessibility. This approach is
-more appropriate and informative than calculating provider-to-population
-ratios (PPR) that simply divide the level of supply of a service (e.g.,
-the number of bicycle racks at a station) by the population who have
-access to the service (Paez et al., 2019). In particular, the Two-Step
-Floating Catchment Area (2SFCA) method (Luo and Wang, 2003; Radke and
-Mu, 2000) produces flexible catchment areas instead of using rigid
-boundaries like PPR. In the first step, a ratio of supply to demand at
-service locations is calculated, such as the number of beds at a
-hospital divided by the number of people within the catchment area of
-the hospital, weighted by the distance involved in reaching the
-facility. Next, these service level ratios are allocated back to the
-population centers and summarized as a measure of congested
+approach is that the method incorporates information on capacity and
+demand in calculating accessibility. Geurs and van Wee (2004) note
+``competition for activities with restricted capacity'' should be taken
+into account for land-use components of accessibility. The FCA approach
+is more appropriate and informative than calculating
+provider-to-population ratios (PPR) that simply divide the level of
+supply of a service (e.g., the number of bicycle racks at a station) by
+the population who have access to the service (Paez et al., 2019). In
+particular, the Two-Step Floating Catchment Area (2SFCA) method (Luo and
+Wang, 2003; Radke and Mu, 2000) produces flexible catchment areas
+instead of using rigid boundaries like PPR. In the first step, a ratio
+of supply to demand at service locations is calculated, such as the
+number of beds at a hospital divided by the number of people within the
+catchment area of the hospital, weighted by the distance involved in
+reaching the facility. Next, these service level ratios are allocated
+back to the population centers and summarized as a measure of congested
 accessibility. Thus, this model does a good job of considering potential
@@
 program. Additional bicycles and twelve ``equity'' stations were added
-to the core service area in more disadvantaged neighbourhoods. The
-equity program also offers subsidized memberships to individuals who
-identify as low income, and complements this service with cycle skills
-education. A comparable program can be found in Philadelphia (see Caspi
-and Noland, 2019).
+to the core service area in more disadvantaged and under-serviced
+neighbourhoods. The equity program also offers subsidized memberships to
+individuals who identify as low income, and complements this service
+with cycle skills education. A comparable program can be found in
+Philadelphia (see Caspi and Noland, 2019).
 
@@
 inflation that result from the overlapping catchment areas produced by
-earlier FCA methods (Paez et al., 2019). Paez et al.~(2019) adjusted the
+earlier FCA methods (Paez et al., 2019). Páez et al. (2019) adjusted the
 impedance weights so that both supply and demand are proportionally
@@
 function, so that each station is assumed to service only a segment of
-the population within a limited geographical range.
-
-Next, the supply at each station (i.e., the maximum number of bicycle
-racks) is divided by its estimated service population within the
-established catchment area; this gives the level of service of station
-\(j\) in bicycle racks per person: \begin{equation}
+the population within a limited geographical range. The level of service
+of station \(j\) in bicycle racks per person is the supply at each
+station (i.e., the maximum number of bicycle racks) divided by the
+population within the established catchment area: \begin{equation}
 \label{eq:level-of-service}
@@
 
-Finally, the accessibility of population cell \(i\) is calculated as the
-weighted sum of the level of service of all stations that can be reached
-from there according to the spatial weights: \begin{equation}
+In the second step, the accessibility of population unit \(i\) is
+calculated as the weighted sum of the level of service of all stations
+that can be reached from there according to the spatial weights:
+\begin{equation}
 \label{eq:FCA-accessibility}
-A_i = {\sum_{j = 1}^{J} L_j{w_{ji}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ji}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
+A_i = {\sum_{j = 1}^{J} L_j{w_{ij}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ij}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
 \end{equation}
 
-The balanced approach of Paez et al (2019) replaces the spatial weights
-with normalized versions as follows: \begin{equation}
+The balanced approach of Páez et al. (2019) replaces the spatial weights
+with normalized versions as follows. In the first step, the population
+is weighted with: \begin{equation}
 \label{eq:spatial-weights}
-{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
+{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ij}}}}
 \end{equation}
 
-\noindent and: \begin{equation}
+\noindent and in the second step, the level of service is weighted
+using: \begin{equation}
 \label{eq:spatial-weights-2}
-{w_{ij}^{j} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ji}}}}
+{w_{ij}^{j} = \frac {w_{ij}}{\sum_{i = 1}^{n} {w_{ij}}}}
 \end{equation}
@@
 disaggregated approach than the use of census geographies. For this
-reason, we implemented our analysis parting from micro population zones
+reason, we implemented our analysis parting from small population cells
 to better reflect the friction of walking to a docking station, which is
@@
 50-by-50 m in size. Pycnophylactic interpolation involves smoothing out
-the population from each DA while preserving total volume {[}see Figure
-\ref{fig:da-population}{]}. When interpolating the population at this
-high level of resolution, it is important to ensure that population
-numbers were not allocated to areas where people do not live in Hamilton
-(for example, to parks or large institutional buildings, etc.). To do
-so, we retrieved shapefiles for various geographic features from Open
-Hamilton. Next, we removed these features from the PBSP core service
-area and used pycnophylactic interpolation to disaggregate and
-reallocate population within the remaining area {[}see Figure
-\ref{fig:interpolated-population}{]}.
+the population from each DA while preserving total volume. When
+interpolating the population at this high level of resolution, it is
+important to ensure that population numbers are not allocated to areas
+where people do not live in Hamilton (for example, to parks or large
+institutional buildings, etc.). To do so, we retrieved shapefiles for
+various geographic features from Open Hamilton. Next, we removed these
+features from the PBSP core service area and used pycnophylactic
+interpolation to disaggregate and reallocate population within the
+remaining area {[}see Figure \ref{fig:interpolated-population}{]}.
 
@@
 
-To calculate walking times from the centroid of the micro population
-cells to docking stations, we extracted OpenStreetMap data for HBS's
-service area from
-\href{https://download.bbbike.org/osm/bbbike/}{BBBike}, an online cycle
-route planner that interfaces with OpenStreetMap. OpenStreetMap data
-provides the networks for calculating walking times from each population
-cell to nearby docking stations, using a maximum walking distance of 10
-km and walking time of 30 minutes as thresholds. A travel time matrix
-was created with the origins as the coordinates of the population cells
-and the destinations as the coordinates of the docking stations within
-the maximum threshold. This process provides a more realistic measure of
-the friction of reaching stations by taking infrastructure into account
-network travel times, rather than using the Euclidean distance from
-population cell to station. Routing and travel time calculations were
-completed using the \texttt{R} package \texttt{r5r}, used for rapid
-realistic routing operations (Pereira et al., 2021).
+To calculate walking times from the centroid of the population cells to
+docking stations, we extracted OpenStreetMap data for HBS's service area
+from \href{https://download.bbbike.org/osm/bbbike/}{BBBike}, an online
+cycle route planner that interfaces with OpenStreetMap. OpenStreetMap
+data provides the networks for calculating walking times from each
+population cell to nearby docking stations, using a maximum walking
+distance of 10 km and walking time of 30 minutes as thresholds. A travel
+time matrix was created with the origins as the coordinates of the
+population cells and the destinations as the coordinates of the docking
+stations within the maximum threshold. This process provides a more
+realistic measure of the friction of reaching stations by taking
+infrastructure into account network travel times, rather than using the
+Euclidean distance from population cell to station. Routing and travel
+time calculations were completed using the \texttt{R} package
+\texttt{r5r}, used for rapid realistic routing operations (Pereira et
+al., 2021).
 
@@
 m (Fuller et al., 2011) and 300 m (Kabra et al., 2020) of a docking
-station is correlated with bike share use, while other research has
+station is correlated with bike share use. Some studies have defined
+``close proximity'' to a docking station as 500 metres or less (Fuller
+et al., 2013; Hosford et al., 2019, 2018), while other research has
 found that walking trips are less than 600 m and rarely more than 1200 m
@@
 produce more smoothing and fewer spikes.
+
+In the analysis we use the following distance decay function: \[
+{w_{ij} = \begin{cases} 
+      1 & \text{when } t_{ij}\leq \gamma \\
+      0 & \text{otherwise} 
+   \end{cases}}
+\] \noindent where \(\gamma\) is the relevant threshold. The weights are
+standardized as discussed above.
 
@@
 lower income, we estimated what level of accessibility accrues to how
-many people in different income strata. To this end, we took our micro
-population zones and aggregated the accessibility and population by the
+many people in different income strata. To this end, we took our small
+population cells and aggregated the accessibility and population by the
 median total household income as imputed from the dissemination areas. A
@@
 Using disaggregated population data, we examined equity in accessibility
-to Hamilton Bike Share, with a particular focus on assessing the
-contribution of the program's equity stations. The BFCA approach,
-combined with pycnophylactic interpolation, enabled us to measure
-accessibility on a micro scale which better reflects the scale at which
-walking takes place and avoids the ``absorption of disparities'', as
-articulated by Chen et al.~(2019) and elsewhere in the transportation
-literature (National Academies of Sciences, 2004; Rowangould et al.,
-2016). Our method also considers demand and supply in the calculation of
-accessibility, which is important for services like PBSPs but not
-accounted for in other common accessibility models. Unlike the 2SFCA
-method, there was no inflation or deflation in our accessibility
-estimates because population and level of service were preserved (see
-Figure \ref{fig:figure-10} and Figure \ref{fig:figure-11}). This
-differentiates our analysis from similar papers exploring equity in
-PBSPs that use larger geographical units of analysis or that focus on
-station location instead of level of service. In this way, our paper has
-made an important contribution by applying an intuitive and useful
-approach to measure accessibility to a PBSP. This open and reproducible
-research could be expanded for additional analysis using other
-individual-level data in Hamilton (e.g., age, gender, household size,
-education, etc.) or applied to PBSPs in other North American cities.
+to Hamilton Bike Share (HBS), with a focus on assessing the contribution
+of the program's equity stations. The BFCA approach, combined with
+pycnophylactic interpolation, enabled us to measure accessibility on a
+micro scale which better reflects the scale at which walking takes place
+and avoids the ``absorption of disparities,'' as articulated by Chen et
+al. (2019) and elsewhere in the transportation literature (National
+Academies of Sciences, 2004; Rowangould et al., 2016). Our method also
+considers potential demand and supply, as well as congestion effects, in
+the calculation of accessibility, which is important for services like
+PBSPs but not accounted for in other common accessibility models. Unlike
+the 2SFCA method, there was no inflation or deflation in our
+accessibility estimates because population and level of service were
+preserved (see Figure \ref{fig:figure-10} and Figure
+\ref{fig:figure-11}). This differentiates our analysis from similar
+papers exploring equity in PBSPs that use larger geographical units of
+analysis or that focus only on station location instead of level of
+service. In this way, our paper has made an important contribution by
+applying an intuitive and useful approach not previously used in the
+cycling literature to measure accessibility to a PBSP. This open and
+reproducible research could be expanded for additional analysis using
+other individual-level data in Hamilton (e.g., age, gender, household
+size, education, etc.) or applied to PBSPs in other North American
+cities.
 
@@
 showing other stations within a five minute walk, as well as the
-directive of NACTO (City Transportation Officials, 2015). We find that
+directive of NACTO (City Transportation Officials, 2015). We found that
 over 118,000 people can access a bike share station within a 15 minute
@@
 Hamilton should have between 380 and 440 stations instead of 130, and
-1,500 bicycles instead of 900 (Hamilton, 2015a). Reduced capacity within
-the system led to gaps in coverage in some areas of the city ``with some
-areas not having the recommended station density of 300m between
-stations or 10 stations per square km'' (Hamilton, 2015a). Figures
-\ref{fig:figure-bi-map-threshold-3},
+1,500 bicycles instead of 900 (Hamilton, 2015a). The City acknowledged
+that reduced capacity within the system led to gaps in coverage in some
+areas of the city ``with some areas not having the recommended station
+density of 300m between stations or 10 stations per square km''
+(Hamilton, 2015a). Figures \ref{fig:figure-bi-map-threshold-3},
 \ref{fig:figure-bi-map-threshold-5},
 \ref{fig:figure-bi-map-threshold-10}, and
-\ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in service
-were filled by the additional equity stations. The improvement in
-accessibility is most noticeable at the average and maximum thresholds
-in the east end of the core service area, which corresponds to the
-neighbourhoods where stations were added. However, the contribution of
-equity stations to reducing inequities in accessibility was modest.
+\ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in
+accessibility were filled by the additional equity stations implemented
+by Hamilton Bike Share. The improvement in accessibility is most
+noticeable at the average and maximum thresholds in the east end of the
+core service area, which corresponds to the neighbourhoods where
+stations were added. However, the contribution of equity stations to
+reducing inequities in accessibility was modest.
 
 The equity stations were added to HBS as a targeted initiative to
-increase access for disadvantaged communities. However, we found that
-inequities persisted as evidenced by differences in accessibility
-according to income quintile (Table \ref{tab:accessibility-income}).
-While the addition of equity stations seems to modestly increase
-accessibility for all income groups at all thresholds, they did not
-increase accessibility substantially for any single income group. Most
-importantly, individuals in the bottom 20\% of median total household
-income have the second lowest level of access to Hamilton Bike Share at
-most thresholds (average, maximum, and extreme). At the minimum
+increase access for equity users in specific under-serviced
+neighbourhoods. However, we found that inequities persisted as evidenced
+by differences in accessibility according to income quintile (Table
+\ref{tab:accessibility-income}). While the addition of equity stations
+seems to modestly increase accessibility for all income groups at all
+thresholds, they did not increase accessibility substantially for any
+single income group. Most importantly, individuals in the bottom 20\% of
+median total household income have the second lowest level of access to
+HBS at most thresholds (average, maximum, and extreme). At the minimum
 threshold, the bottom 20\% have the lowest level of access. This
-suggests that the equity stations may not have sufficiently addressed
-vertical equity. While previous research found that neighborhoods with
-more disadvantage are better serviced by HBS, the authors used the
-Pampalon Deprivation Index to determine the level of disadvantage for
-dissemination areas \emph{across} the city not just within the core
-service area (Hosford and Winters, 2018). Instead, we used median total
-household income for each dissemination area \emph{within} the core
-service area. We conclude that Hamilton's PBSP, while by default located
-in areas with more deprivation compared to other cities, continues to
-have disparities in accessibility between income groups. With equity
+suggests that additional equity stations are needed to more sufficiently
+address vertical equity. While previous research found that
+neighborhoods with more disadvantage are better serviced by HBS, the
+authors used the Pampalon Deprivation Index to determine the level of
+disadvantage for dissemination areas \emph{across} the city not just
+within the core service area (Hosford and Winters, 2018). Instead, we
+used median total household income for each dissemination area
+\emph{within} the core service area. We conclude that Hamilton's PBSP,
+while by default located in areas with more deprivation compared to
+other cities as noted by Hosford and Winters (2018), continues to have
+disparities in accessibility between income groups. With equity
 stations, many areas with low median total household income in the east
-end of the service area continue to have zero or low accessibility at
-the average threshold of five minutes. At the maximum and extreme
-thresholds, the top 20\% have the highest level of access to Hamilton
-Bike Share. These findings align with other studies from Tampa (Chen et
-al., 2019), Philadelphia (Caspi and Noland, 2019) and Seattle (Mooney et
-al., 2019), which have found disparities in station location, annual
-trips, or access to bicycles, respectively, between levels of income and
-education. With the addition of equity stations, there were large gains
-in accessibility for the second 20\% at the average threshold, amounting
-to over 5,000 more people, but much smaller gains for the bottom 20\%
-with only 326 more people who are able to access a HBS station.
+end of the service area have zero or low accessibility at the average
+threshold of five minutes. At the maximum and extreme thresholds, the
+top 20\% have the highest level of access to HBS. These findings align
+broadly with other studies from Tampa (Chen et al., 2019), Philadelphia
+(Caspi and Noland, 2019) and Seattle (Mooney et al., 2019), which have
+found disparities in station location, annual trips, or access to
+bicycles, respectively, between levels of income and education. With the
+addition of equity stations, there were large gains in accessibility for
+the second 20\% at the average threshold, amounting to over 5,000 more
+people, but much smaller gains for the bottom 20\% with only 326 more
+people who are able to access a HBS station.
 
@@
 docking stations for increasing equity in accessibility to PBSPs.
-Several studies have examined who can access PBSPs based on the location
-of stations (among others, see Babagoli et al., 2019; Qian and Jaller,
-2020; Qian et al., 2020; J. Wang and Lindsey, 2019b), but the size of
-docking stations in disadvantaged communities is not considered in the
-literature. Station size (i.e., potential maximum supply of bicycles)
-was also not identified by PBSP owners and operators as a factor that
-needed to be addressed in equity efforts (Howland et al., 2017). PBSPs
-may not necessarily reduce inequities in access if stations are easy to
-reach for disadvantaged groups but offer only a small number of
+Congestion effects for services like bike share that offer a potential
+maximum supply of bicycles (i.e., number of racks) at each docking
+station must be taken into account when measuring accessibility. Several
+studies have examined who can access PBSPs based on the location of
+stations (among others, see Babagoli et al., 2019; Qian et al., 2020;
+Qian and Jaller, 2020; Wang and Lindsey, 2019b), but the size of docking
+stations in disadvantaged communities is not considered in the
+literature. An exception is a study on bike share ridership in New York
+City that found ``the number of docks at a bike share station is
+statistically significant and positively associated with trip
+attractions for both men and women'' (Wang and Akar, 2019, p. 5).
+Station size was also not identified by PBSP owners and operators as a
+factor that needed to be addressed in equity efforts (Howland et al.,
+2017). PBSPs may not necessarily reduce inequities in access if stations
+are easy to reach for equity users but offer only a small number of
 bicycles. Likewise, more people may not opt to use the program if the
 supply of bicycles available at nearby stations is insufficient to meet
-demand. Babgoli et al.~(2019) found a slight but not statistically
+demand. Babgoli et al. (2019) found a slight but not statistically
 significant increase in the proportion of neighborhoods with the highest
@@
 The following policy and planning implications can inform efforts for
-increasing equity in Hamilton Bike Share, and potentially other PBSPs in
-North America. Our analysis has identified specific areas that have both
-low accessibility and low median total household income. Figures
+increasing equity in HBS, and potentially other PBSPs in North America.
+Our analysis has identified specific areas that have both low
+accessibility and low median total household income. Figures
 \ref{fig:figure-bi-map-threshold-3},
@@
 \ref{fig:figure-bi-map-threshold-15} highlight potential locations for
-new equity stations to better accommodate low income groups. Additional
+new equity stations to increase access for low-income groups. Additional
 stations are needed to further reduce inequities in accessibility. This
@@
 HBS operators. Our findings also suggest that PBSPs should consider
-increasing the size of stations in disadvantaged communities. Docking
-stations added to target vertical equity should have an adequate supply
-of public bicycles, particularly more in areas that have a strong
-interest in adopting bike share. Finally, distance to stations is
-associated with bike share use (Fuller et al., 2011; J. Wang and
-Lindsey, 2019b), especially when the trip starts at the residence
-(Fishman et al., 2013). Distance can also be a barrier to using PBSPs
-(Fishman et al., 2014). To address this issue, transportation planners
-and PBSP operators should consult with groups and communities who are
-under-represented among bike share users to determine how far they are
-willing to walk to reach a docking station and to get their input on the
-location of new docking stations.
+increasing the size of stations in targeted communities. Notably, figure
+\ref{fig:sobi-stations-in-hamilton} highlights how the equity stations
+in Hamilton have a lower number of racks compared to docking stations in
+the city's downtown core. Docking stations added to increase vertical
+equity should have an adequate supply of public bicycles, particularly
+more in areas that have a strong interest in adopting bike share, to
+increase levels of service and potentially attract new users. Finally,
+distance to stations is associated with bike share use (Fuller et al.,
+2011; Wang and Lindsey, 2019b), especially when the trip starts at the
+residence (Fishman et al., 2013). Distance can also be a barrier to
+using PBSPs (Fishman et al., 2014). To address this issue,
+transportation planners and PBSP operators should consult with groups
+and communities who are under-represented among bike share users to
+determine how far they are willing to walk to reach a docking station
+and to get their input on the location of new docking stations.
 
@@
 to determine whether the addition of equity stations encouraged more
-bicycling for disadvantaged groups living near them. Other studies have
+bicycling for equity users living near them. Other studies have
 specifically looked at differences in trip type, frequency, or length
 among users from disadvantaged neighborhoods (Caspi and Noland, 2019;
-Qian and Jaller, 2020; J. Wang and Lindsey, 2019a), but our analysis is
+Qian and Jaller, 2020; Wang and Lindsey, 2019a), but our analysis is
 limited by the lack of publicly available route and individual user data
@@
 times during the day across many seasons. Should this data become
-available, it would be worthwhile to revisit the question to examine how
+available, it would be worthwhile to revisit the study to examine how
 well the operation of the system (including balancing of bicycles across
@@
 As such, a logical next step to this research is to examine whether the
-equity stations increased ridership or resulted in new memberships in
-areas that were previously under-serviced. An examination of the types
-of trips undertaken by residents in these areas would also be
-informative, such as the study undertaken by Caspi and Norland (2019)
-after bike share stations were implemented in low-income Philadelphia
-neighborhoods. The bulk of cycling facilities that have been built in
-Hamilton to date are located in the core service area near the
-conventional stations. It would be worthwhile to explore the route
-choice of bike share trips departing or ending at the equity stations
-and to identify factors that specifically influence trips from these
-stations, which would extend existing studies conducted by Scott and
-colleagues (Lu et al., 2018; Scott and Ciuro, 2019; Scott et al., 2021).
-This paper, combined with additional studies such as those
-conceptualized above, would serve as a valuable case study for Hamilton
-and other cities with PBSPs that wish to evaluate and address inequities
-in accessibility and transportation options in urban areas.
+equity stations increased ridership or resulted in new memberships for
+equity users. An examination of the types of trips undertaken by
+residents in these areas would also be informative, such as the study
+undertaken by Caspi and Norland (2019) after docking stations were
+implemented in low-income Philadelphia neighborhoods. The bulk of
+bicycle infrastructure that has been built in Hamilton to date is
+located in the core service area particularly near the conventional
+stations. It would be worthwhile to explore the route choice of bike
+share trips departing or ending at the equity stations and to identify
+factors that specifically influence trips from these stations, which
+would extend existing studies conducted by Scott and colleagues (Lu et
+al., 2018; Scott et al., 2021; Scott and Ciuro, 2019). Perceived
+accessibility using bike share according to low-income neighbourhoods
+could also be studied in the future. Lättman et al. (2016) have
+developed the \emph{Perceived Accessibility Scale} that would be useful
+for this purpose. This paper, combined with additional studies such as
+those conceptualized above, would serve as a valuable case study for
+Hamilton and other cities with PBSPs that wish to evaluate and address
+inequities in accessibility to active transportation options in urban
+areas.
 
@@
 
+\leavevmode\hypertarget{ref-arribas2021open}{}%
+Arribas/Bel, D., Green, M., Rowe, F., Singleton, A., 2021. Open data
+products: A framework for creating valuable analysis-ready data. Journal
+of Geographical Systems.
+doi:\href{https://doi.org/10.1007/s10109-021-00361-7}{10.1007/s10109-021-00361-7}
+
 \leavevmode\hypertarget{ref-auchinclossDesignBaselineDescription2020}{}%
@@
 \leavevmode\hypertarget{ref-fishmanBikeShareSynthesis2013}{}%
-Fishman, E., Washington, S., Haworth, N., 2013. Bike Share: A Synthesis
-of the Literature. Transport Reviews 33, 148--165.
+Fishman, E., Washington, S., Haworth, N., 2013. Bike {Share}: {A
+Synthesis} of the {Literature}. Transport Reviews 33, 148--165.
 doi:\href{https://doi.org/10.1080/01441647.2013.775612}{10.1080/01441647.2013.775612}
+
+\leavevmode\hypertarget{ref-fishmanBikeshareImpactActive2015}{}%
+Fishman, E., Washington, S., Haworth, N., 2015. Bikeshare's impact on
+active travel: {Evidence} from the {United States}, {Great Britain}, and
+{Australia}. Journal of Transport \& Health 2, 135--142.
+doi:\href{https://doi.org/10.1016/j.jth.2015.03.004}{10.1016/j.jth.2015.03.004}
 
@@
 Fuller, D., Gauvin, L., Kestens, Y., Daniel, M., Fournier, M., Morency,
-P., Drouin, L., 2011. Use of a New Public Bicycle Share Program in
-Montreal, Canada. American Journal of Preventive Medicine 41, 80--83.
+P., Drouin, L., 2011. Use of a {New Public Bicycle Share Program} in
+{Montreal}, {Canada}. American Journal of Preventive Medicine 41,
+80--83.
 doi:\href{https://doi.org/10.1016/j.amepre.2011.03.002}{10.1016/j.amepre.2011.03.002}
+
+\leavevmode\hypertarget{ref-fullerImpactEvaluationPublic2013}{}%
+Fuller, D., Gauvin, L., Kestens, Y., Daniel, M., Fournier, M., Morency,
+P., Drouin, L., 2013. Impact evaluation of a public bicycle share
+program on cycling: A case example of {BIXI} in {Montreal}, {Quebec}.
+American Journal of Public Health 103, e85--92.
+doi:\href{https://doi.org/10.2105/AJPH.2012.300917}{10.2105/AJPH.2012.300917}
+
+\leavevmode\hypertarget{ref-geursAccessibilityEvaluationLanduse2004}{}%
+Geurs, K.T., van Wee, B., 2004. Accessibility evaluation of land-use and
+transport strategies: Review and research directions. Journal of
+Transport Geography 12, 127--140.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2003.10.005}{10.1016/j.jtrangeo.2003.10.005}
 
@@
 \leavevmode\hypertarget{ref-hullgrassoBikeShareEquity2020}{}%
-Hull Grasso, S., Barnes, P., Chavis, C., 2020. Bike Share Equity for
-Underrepresented Groups: Analyzing Barriers to System Usage in
-Baltimore, Maryland. Sustainability 12, 7600.
+Hull Grasso, S., Barnes, P., Chavis, C., 2020. Bike share equity for
+underrepresented groups: Analyzing barriers to system usage in
+baltimore, maryland. Sustainability 12.
 doi:\href{https://doi.org/10.3390/su12187600}{10.3390/su12187600}
@@
 \leavevmode\hypertarget{ref-larsenQuarterMileReexamining2010}{}%
-Larsen, J., El-Geneidy, A., Yasmin, F., 2010. Beyond the Quarter Mile:
-Re-examining Travel Distances by Active Transportation. Canadian Journal
-of Urban Research 19, 70--88.
+Larsen, J., El-Geneidy, A., Yasmin, F., 2010. Beyond the {Quarter Mile}:
+{Re}-examining {Travel Distances} by {Active Transportation}. Canadian
+Journal of Urban Research 19, 70--88.
+
+\leavevmode\hypertarget{ref-lattmanDevelopmentTestPerceived2016}{}%
+Lättman, K., Olsson, L.E., Friman, M., 2016. Development and test of the
+{Perceived Accessibility Scale} ({PAC}) in public transport. Journal of
+Transport Geography 54, 257--263.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2016.06.015}{10.1016/j.jtrangeo.2016.06.015}
 
@@
 Qian, X., Jaller, M., Niemeier, D., 2020. Enhancing equitable service
-level: Which can address better, dockless or dock-based Bikeshare
+level: Which can address better, dockless or dock-based bikeshare
 systems? Journal of Transport Geography 86, 102784.
-doi:\href{https://doi.org/10.1016/j.jtrangeo.2020.102784}{10.1016/j.jtrangeo.2020.102784}
+doi:\url{https://doi.org/10.1016/j.jtrangeo.2020.102784}
 
 \leavevmode\hypertarget{ref-radkeSpatialDecompositionsModeling2000}{}%
-Radke, J., Mu, L., 2000. Spatial Decompositions, Modeling and Mapping
-Service Regions to Predict Access to Social Programs. Geographic
-Information Sciences 6, 105--112.
+Radke, J., Mu, L., 2000. Spatial {Decompositions}, {Modeling} and
+{Mapping Service Regions} to {Predict Access} to {Social Programs}.
+Geographic Information Sciences 6, 105--112.
 doi:\href{https://doi.org/10.1080/10824000009480538}{10.1080/10824000009480538}
@@
 Wang, J., Lindsey, G., 2019b. Do new bike share stations increase member
-use: A quasi-experimental study. Transportation Research Part A: Policy
-and Practice 121, 1--11.
+use: {A} quasi-experimental study. Transportation Research Part A:
+Policy and Practice 121, 1--11.
 doi:\href{https://doi.org/10.1016/j.tra.2019.01.004}{10.1016/j.tra.2019.01.004}
+
+\leavevmode\hypertarget{ref-wangGenderGapGenerators2019}{}%
+Wang, K., Akar, G., 2019. Gender gap generators for bike share
+ridership: {Evidence} from {Citi Bike} system in {New York City}.
+Journal of Transport Geography 76, 1--9.
+doi:\href{https://doi.org/10.1016/j.jtrangeo.2019.02.003}{10.1016/j.jtrangeo.2019.02.003}
 

### 2021-10-14 Bike-share-spatial-equity/Response-to-Reviewers-R3.Rmd
Revisions after reviewer comments R2

---
title: "Response to Reviewers"
author: ""
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: pdf_document
---

We would like to express our gratitude to the editors and anonymous reviewers for their time and attention to our submission.

## Reviewer #1:

This paper is ready for publication. I don't have any more comments.

\textcolor{blue}{Thank you for your generous feedback to improve this paper.}

## Reviewer #3:

Thank you for revising in response to my prior comments. Due to an editorial error, I missed a round of feedback so will keep my comments brief and mainly reiterate some prior points.

\textcolor{blue}{Thank you again for your suggestions to improve the paper.}

Because travel is derived from a demand for activity participation, I still do not believe that you are quantifying accessibility in this paper. No one gets to a bike share station consumes the service of bike share and ends their trip. The bike share is a means to an end. You overstate your position on p. 6, l. 20 when you claim that the docking stations are an amenity because they offer a transportation service. Again – people consume travel to reach destinations where they can undertake activities. This fact must be noted in the study limitations.

\textcolor{blue}{Words go here.}

Many of the maps are still very difficult to interpret. I previously suggested plotting the difference in accessibility (for example for Figures 4-6) to make the differences easy to discern. Currently it’s very difficult to detect any changes in the maps at all. I also suggested experimenting with a discrete scale. This might also help to make the differences easier to discern without needing an explicit map of difference. All units in the map legends need to be spelled out (e.g., “accessibility (racks/person)”) and use plain English with any acronyms defined (e.g., for los_inflation and accessibility_inflation). The discordance between the label for Figure 7 “deflation” and legend “inflation” should be resolved. Scale bar and north arrow only need to be shown once on the small multiples maps.

\textcolor{blue}{Words go here.}

### 2021-10-14 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Revisions after reviewer comments R2

@@
 Both are of interest to researchers and transportation planners since they are often linked in that advantage, or conversely disadvantage, has spatial patterns.
-Bike share equity programs [see @trec2019; @@howlandCurrentEffortsMake2017] can be considered efforts to improve vertical equity since they favour groups that have benefited less from PBSPs through the placement of stations or equitable fee structures.
+Bike share equity programs [see @trec2019; @howlandCurrentEffortsMake2017] can be considered efforts to improve vertical equity since they favour groups that have benefited less from PBSPs through the placement of stations or equitable fee structures.
 
@@
 These zonal values of the population were interpolated to smaller polygons that are 50-by-50 m in size.
-Pycnophylactic interpolation involves smoothing out the population from each DA while preserving total volume [see Figure \ref{fig:da-population}].
-When interpolating the population at this high level of resolution, it is important to ensure that population numbers were not allocated to areas where people do not live in Hamilton (for example, to parks or large institutional buildings, etc.).
+Pycnophylactic interpolation involves smoothing out the population from each DA while preserving total volume.
+When interpolating the population at this high level of resolution, it is important to ensure that population numbers are not allocated to areas where people do not live in Hamilton (for example, to parks or large institutional buildings, etc.).
 To do so, we retrieved shapefiles for various geographic features from Open Hamilton.
@@
           color = NA) + 
-  #geom_sf(data = population_50x50,
-  #        aes(color = population)) +
-   #scale_fill_gradient(colours = "YlOrRd") +
   geom_tile(data = cbind(st_drop_geometry(population_50x50), 
@@
                 fill = population)) +
-  scale_fill_distiller(palette = "YlOrRd", 
+  scale_fill_distiller(name = "Population",
+                       palette = "YlOrRd",
                        direction = 1) +
@@
           color = "gray") +
-  #geom_sf(data = accessibility %>%
-  #          dplyr::filter(threshold == "3 min"),
-  #        aes(color = accessibility)) +
-  #scale_color_distiller(palette = "PuRd", 
-  #                      direction = 1) +
   geom_tile(data = cbind(st_drop_geometry(accessibility_3min), st_coordinates(accessibility_3min)),
@@
           color = "dark green") +
-  # geom_sf(data =  sobi_hubs %>% 
-  #           dplyr::filter(hub_type == "ERI"), 
-  #         size = 1.5, shape = 21, 
-  #         fill = "green") +
   theme(panel.background = element_rect(fill = "white"),
@@
 
-### 5.1.2. Average Threshold
+### 5.1.2. Medium Threshold
 
@@
 
-```{r figure-7, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 5 minutes walk (average threshold) compared between current system with equity stations and the original system without equity stations."}
+```{r figure-7-a, include=FALSE}
 #accessibility plot for current system with equity stations at 5 minutes walk
 accessibility_5min <- accessibility %>% 
-  dplyr::filter(threshold == "5 min")
-
-ggplot() +
+  dplyr::filter(threshold == "5 min",
+                system == "All stations") %>%
+  mutate(accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.001, 0.02, 0.4, 8, Inf), 
+                             labels=c("<0.001", "0.001-0.02", "0.02-0.4", "0.4-8", ">8")),
+         accessibility = factor(accessibility,
+                                levels = c(">8", "0.4-8", "0.02-0.4", "0.001-0.02", "<0.001"),
+                                ordered = TRUE))
+
+acc_5_all <- ggplot() +
   geom_sf(data = population_50x50,
           color = "gray") +
-  #geom_sf(data = accessibility %>%
-  #          dplyr::filter(threshold == "5 min"),
-  #        aes(color = accessibility)) +
-  #scale_color_distiller(palette = "PuRd", 
-  #                      direction = 1) +
-  geom_tile(data = cbind(st_drop_geometry(accessibility_5min), st_coordinates(accessibility_5min)),
+  geom_tile(data = cbind(st_drop_geometry(accessibility_5min), 
+                         st_coordinates(accessibility_5min)),
             aes(x = X, y = Y, fill = accessibility)) +
-  scale_fill_distiller(palette = "PuRd", 
-                       direction = 1,
-                       trans = "log",
-                       breaks = c(0.001, 0.02, 0.4, 8)) +
+  scale_fill_brewer(name = "Accessibility\n(racks/person)",
+                    palette = "PuRd", 
+                    direction = -1) +
   geom_sf(data = sobi_service,
           fill = NA) +
-  facet_grid(system ~ threshold)  +
+  facet_grid(system ~ .)  +
   geom_sf(data = community_downtown %>%
@@
         axis.ticks = element_blank())  +
-  annotation_scale(location = "bl", width_hint = 0.4,
-                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
   annotation_north_arrow(location = "tr", which_north = "true", 
@@
         style = north_arrow_fancy_orienteering)
-```
-
-### 5.1.3. Maximum Threshold
+
+# Extract the legend
+acc_5_legend <- cowplot::get_legend(acc_5_all)
+```
+
+```{r figure-7-b, include=FALSE}
+#accessibility plot for current system with equity stations at 5 minutes walk
+accessibility_5min <- accessibility %>% 
+  dplyr::filter(threshold == "5 min",
+                system == "Conventional") %>%
+  mutate(accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.001, 0.02, 0.4, 8, Inf), 
+                             labels = c("<0.001", "0.001-0.02", "0.02-0.4", "0.4-8", ">8")),
+         accessibility = factor(accessibility,
+                                levels = c(">8", "0.4-8", "0.02-0.4", "0.001-0.02", "<0.001"),
+                                ordered = TRUE))
+
+acc_5_conv <- ggplot() +
+  geom_sf(data = population_50x50,
+          color = "gray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_5min),
+                         st_coordinates(accessibility_5min)),
+            aes(x = X, 
+                y = Y, 
+                fill = accessibility)) +
+  scale_fill_brewer(name = "Accessibility\n(racks/person)",
+                    palette = "PuRd", 
+                    direction = -1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ .)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"))
+```
+
+```{r figure-7, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 5 minutes walk (average threshold) compared between current system with equity stations and the original system without equity stations."}
+ggdraw() + 
+  draw_plot(acc_5_all + 
+              theme(legend.position = "none"), 
+            0.00, 0.5, 1, 0.5) + 
+  draw_plot(acc_5_conv + 
+              theme(legend.position = "none"), 0.00, 0.0, 1, 0.5) + 
+  draw_plot(acc_5_legend, 0.01, 0.4, 0.2, 0.2)
+```
+
+### 5.1.3. High Threshold
 
@@
 
-```{r figure-8, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) compared between current system with equity stations and the original system without equity stations."}
+```{r figure-8-a, include=FALSE}
 #accessibility plot for current system with equity stations at 10 minutes walk
 accessibility_10min <- accessibility %>% 
-  dplyr::filter(threshold == "10 min")
-
-ggplot() +
+  dplyr::filter(threshold == "10 min",
+                system == "All stations")  %>%
+  mutate(accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.00004, 0.0003, 0.002, Inf), 
+                             labels = c("<0.00004", "0.00004-0.0003", "0.0003-0.002", ">0.002")),
+         accessibility = factor(accessibility,
+                                levels = c( ">0.002", "0.0003-0.002", "0.00004-0.0003", "<0.00004"),
+                                ordered = TRUE))
+
+acc_10_all <- ggplot() +
   geom_sf(data = population_50x50,
           color = "lightgray") +
-  #geom_sf(data = accessibility %>%
-  #          dplyr::filter(threshold == "10 min"),
-  #        aes(color = accessibility)) +
-  #scale_color_distiller(palette = "PuRd", 
-  #                      direction = 1) +
-  geom_tile(data = cbind(st_drop_geometry(accessibility_10min), st_coordinates(accessibility_10min)),
-            aes(x = X, y = Y, fill = accessibility)) +
-  scale_fill_distiller(palette = "PuRd",
-                       direction = 1,
-                       trans = "log",
-                       breaks = c(0.00004, 0.0003, 0.002)) +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_10min), 
+                         st_coordinates(accessibility_10min)),
+            aes(x = X, 
+                y = Y, 
+                fill = accessibility)) +
+  scale_fill_brewer(name = "Accessibility\n(racks/person)",
+                    palette = "PuRd",
+                    direction = -1) +
   geom_sf(data = sobi_service,
           fill = NA) +
-  facet_grid(system ~ threshold)  +
+  facet_grid(system ~ .)  +
   geom_sf(data = community_downtown %>%
@@
         style = north_arrow_fancy_orienteering)
+
+# Extract the legend
+acc_10_legend <- cowplot::get_legend(acc_10_all)
+```
+
+```{r figure-8-b, include=FALSE}
+#accessibility plot for current system without equity stations at 10 minutes walk
+accessibility_10min <- accessibility %>% 
+  dplyr::filter(threshold == "10 min",
+                system == "Conventional")  %>%
+  mutate(accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.00004, 0.0003, 0.002, Inf), 
+                             labels = c("<0.00004", "0.00004-0.0003", "0.0003-0.002", ">0.002")),
+         accessibility = factor(accessibility,
+                                levels = c( ">0.002", "0.0003-0.002", "0.00004-0.0003", "<0.00004"),
+                                ordered = TRUE))
+
+acc_10_conv <- ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_10min), 
+                         st_coordinates(accessibility_10min)),
+            aes(x = X, 
+                y = Y, 
+                fill = accessibility)) +
+  scale_fill_brewer(name = "Accessibility\n(racks/person)",
+                    palette = "PuRd",
+                    direction = -1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ .)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"))
+```
+
+```{r figure-8, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) compared between current system with equity stations and the original system without equity stations."}
+ggdraw() + 
+  draw_plot(acc_10_all + 
+              theme(legend.position = "none"), 
+            0.00, 0.5, 1, 0.5) + 
+  draw_plot(acc_10_conv + 
+              theme(legend.position = "none"), 0.00, 0.0, 1, 0.5) + 
+  draw_plot(acc_10_legend, 0.01, 0.4, 0.2, 0.2)
 ```
@@
 
-```{r figure-9, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 15 minutes walk (extreme threshold) compared between current system with equity stations and the original system without equity stations."}
+```{r figure-9-a, include=FALSE}
 #accessibility plot for current system with equity stations at 15 minutes walk
 accessibility_15min <- accessibility %>% 
-  dplyr::filter(threshold == "15 min")
-
-ggplot() +
+  dplyr::filter(threshold == "15 min",
+                system == "All stations")  %>%
+  mutate(accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.000006, 0.00005, 0.0003, Inf), 
+                             labels = c("<0.000006", "0.000006-0.00005", "0.00005-0.0003", ">0.0003")),
+         accessibility = factor(accessibility,
+                                levels = c( ">0.0003", "0.00005-0.0003", "0.000006-0.00005", "<0.000006"),
+                                ordered = TRUE))
+
+acc_15_all <- ggplot() +
   geom_sf(data = population_50x50,
           color = "lightgray") +
-  #geom_sf(data = accessibility %>%
-  #          dplyr::filter(threshold == "15 min"),
-  #        aes(color = accessibility)) +
-  #scale_color_distiller(palette = "PuRd", 
-  #                      direction = 1) +
   geom_tile(data = cbind(st_drop_geometry(accessibility_15min), st_coordinates(accessibility_15min)),
             aes(x = X, y = Y, fill = accessibility)) +
-  scale_fill_distiller(palette = "PuRd", 
-                       direction = 1,
-                       trans = "log",
-                       breaks = c(0.000006, 0.00005, 0.0003)) +
+  scale_fill_brewer(name = "Accessibility\n(racks/person)",
+                    palette = "PuRd", 
+                    direction = -1) +
   geom_sf(data = sobi_service,
           fill = NA) +
-
-  facet_grid(system ~ threshold)  +
+  facet_grid(system ~ .)  +
   geom_sf(data = community_downtown %>%
@@
         style = north_arrow_fancy_orienteering)
+
+# Extract the legend
+acc_15_legend <- cowplot::get_legend(acc_15_all)
+```
+
+```{r figure-9-b, include=FALSE}
+#accessibility plot for current system with equity stations at 15 minutes walk
+accessibility_15min <- accessibility %>% 
+  dplyr::filter(threshold == "15 min",
+                system == "Conventional")  %>%
+  mutate(accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.000006, 0.00005, 0.0003, Inf), 
+                             labels = c("<0.000006", "0.000006-0.00005", "0.00005-0.0003", ">0.0003")),
+         accessibility = factor(accessibility,
+                                levels = c( ">0.0003", "0.00005-0.0003", "0.000006-0.00005", "<0.000006"),
+                                ordered = TRUE))
+
+acc_15_conv <- ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_15min), st_coordinates(accessibility_15min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_brewer(name = "Accessibility\n(racks/person)",
+                    palette = "PuRd", 
+                    direction = -1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ .)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"))
+```
+
+```{r figure-9, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 15 minutes walk (extreme threshold) compared between current system with equity stations and the original system without equity stations."}
+ggdraw() + 
+  draw_plot(acc_15_all + 
+              theme(legend.position = "none"), 
+            0.00, 0.5, 1, 0.5) + 
+  draw_plot(acc_15_conv + 
+              theme(legend.position = "none"), 0.00, 0.0, 1, 0.5) + 
+  draw_plot(acc_15_legend, 0.01, 0.4, 0.2, 0.2)
 ```
@@
             aes(size = los_inflation, color = los_inflation)) +
-  scale_color_distiller(palette = "PuRd", direction = 1) +
+  scale_color_distiller(name = "Deflation of Level of Service",
+                        palette = "PuRd", 
+                        direction = 1) +
+  guides(size = guide_legend(title = "")) +
   geom_sf(data = sobi_service,
@@
         axis.ticks = element_blank())  +
-  annotation_scale(location = "bl", width_hint = 0.4,
-                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
-  annotation_north_arrow(location = "tr", which_north = "true", 
-        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+  annotation_scale(location = "bl", 
+                   width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), 
+                   pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "tr", 
+                         which_north = "true", 
+        pad_x = unit(0.0, "in"), 
+        pad_y = unit(0.0, "in"),
         style = north_arrow_fancy_orienteering)
@@
             aes(x = X, y = Y, fill = accessibility_inflation)) +
-  scale_fill_distiller(palette = "PuRd", direction = 1) +
+  scale_fill_distiller(name = "Inflation of accessibility",
+                       palette = "PuRd", 
+                       direction = 1) +
   geom_sf(data = sobi_service,
@@
                 dim = 3) +
-  annotation_scale(location = "bl", width_hint = 0.4,
-                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  #annotation_scale(location = "bl", width_hint = 0.4,
+  #                 pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
   annotation_north_arrow(location = "tr", which_north = "true", 
@@
                    pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
-  annotation_north_arrow(location = "tr", which_north = "true", 
-        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
-        style = north_arrow_fancy_orienteering) +
+  #annotation_north_arrow(location = "tr", which_north = "true", 
+  #      pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+  #      style = north_arrow_fancy_orienteering) +
   theme(panel.background = element_rect(fill = "white"),
@@
 
+\textcolor{blue}{Finally, it is important to note that this study examined accessibility to docking stations, that is, the entry point to the bikeshare system.
+There is a literature that examines transportation modes from this perspective, including accessibility to transit [e.g.,zhaoforecasting2003; @olszewskiusing2005; zieltracomparative2011] and more recently accessibility to bikeshare systems [e.g., @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @caggianiequity2020; @duranfair2020].
+These focus of these studies is to assess who has access to a transportation service.
+However, as argued by @moniruzzamanaccessibility2012, usage of the system likely depends as well on the accessibility to destinations that the system generates.
+An extension of the present study would be to investigate accessibility by the bikeshare system, and perhaps compare that to usage patterns.}
+
 8. Conclusion

### 2021-10-14 Bike-share-spatial-equity/Bike-share-spatial-equity.tex
Revisions after reviewer comments R2

@@
 practices in spatial data science (Brunsdon and Comber, 2020), an open
-data product (Arribas/Bel et al., 2021) along with the code needed to
+data product (Arribas-Bel et al., 2021) along with the code needed to
 reproduce, modify or extend the analysis are available for
@@
 
-\hypertarget{average-threshold}{%
-\subsubsection{5.1.2. Average Threshold}\label{average-threshold}}
+\hypertarget{medium-threshold}{%
+\subsubsection{5.1.2. Medium Threshold}\label{medium-threshold}}
 
@@
 
-\hypertarget{maximum-threshold}{%
-\subsubsection{5.1.3. Maximum Threshold}\label{maximum-threshold}}
+\hypertarget{high-threshold}{%
+\subsubsection{5.1.3. High Threshold}\label{high-threshold}}
 
@@
 
+\textcolor{blue}{Finally, it is important to note that this study examined accessibility to docking stations, that is, the entry point to the bikeshare system.
+There is a literature that examines transportation modes from this perspective, including accessibility to transit [e.g.,zhaoforecasting2003; @olszewskiusing2005; zieltracomparative2011] and more recently accessibility to bikeshare systems [e.g., @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @caggianiequity2020; @duranfair2020].
+These focus of these studies is to assess who has access to a transportation service.
+However, as argued by @moniruzzamanaccessibility2012, usage of the system likely depends as well on the accessibility to destinations that the system generates.
+An extension of the present study would be to investigate accessibility by the bikeshare system, and perhaps compare that to usage patterns.}
+
 \hypertarget{conclusion}{%
@@
 \leavevmode\hypertarget{ref-arribas2021open}{}%
-Arribas/Bel, D., Green, M., Rowe, F., Singleton, A., 2021. Open data
-products: A framework for creating valuable analysis-ready data. Journal
+Arribas-Bel, D., Green, M., Rowe, F., Singleton, A., 2021. Open data
+products: A framework for creating valuable analysis ready data. Journal
 of Geographical Systems.
-doi:\href{https://doi.org/10.1007/s10109-021-00361-7}{10.1007/s10109-021-00361-7}
+doi:\href{https://doi.org/10.1007/s10109-021-00363-5}{10.1007/s10109-021-00363-5}
 

### 2021-10-14 Bike-share-spatial-equity/Untitled/Untitled.Rmd
Revised Manuscript - Round 2

---
title: "Exercise 1: Energy consumption trends"
author:
- name: My Name
  student_number: 123456789
subject: "ENVSOCTY 3LT3"
highlights: |
    Write a brief summary from your perspective of the highlights of working on this assignment.

    You can use more than one paragraph but remember to indent the paragraphs. Try to keep this at less than 200 (?) words.
threshold_concepts: 
# Write the concepts that in your opinion are threshold concepts in this exercise.
- threshold concept 1
- threshold concept 2 
- threshold concept 3
- threshold concept 4
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    # The project-template-default.tex file was heavily  adapted from Steven V. Miller's template for academic manuscripts. See:
    # http://svmiller.com/blog/2016/02/svm-r-markdown-manuscript/
    # https://github.com/svmiller/svm-r-markdown-templates/blob/master/svm-latex-ms.tex
    template: exercise-template-default.tex
always_allow_html: true
---

# Introduction

Transport is fraught with many issues including growth in demand for energy use, greenhouse gases, urban air pollution and noise, negative health impacts and road traffic accidents and fatalities. It is well argued that these problems are not typical to one region but are becoming global in nature. Similar trends have been observed in many places around the world, albeit there are still individual differences among various countries. The observed data suggest an emerging transport crisis that could potentially compromise our quality of life.

In this exercise, you will analyze trends pertaining to energy use and emissions in several countries. In so doing, you will gain a better understanding of emerging transportation trends and their impacts on quality of life.

* Readings:
    + Installation of R.pdf
    + 00 Overview of Data Analysis and Visualization Using R.Rmd (found under Labs/Assignments)
    + 01 Regression Analysis.Rmd (found under READINGS)
* Due date:	See course outline for due date.
* Materials: This R Notebook, and the following Excel file (found under Exercises)
    + `Energy and Emissions.xlsx`

# Instructions

Answers must be provided by typing your solutions using RStudio and this R Notebook file. Change the name of the file to `Exercise 1 (Name-Student Number).Rmd`

The variables in the data file are:

Variable | Definition
-- | --
Country	| Country name
Area | Surface land area of country in sq.km
Population | Population
PYear | Year of estimated population
GDPPC | GDP per capita (nominal) in US dollars
bblpd | barrels of oil per day
EYear | Year of estimated usage of barrels of oil per day
CO2_1995 | CO2 emissions in kilotonnes per year (1995)
CO2_2005 | CO2 emissions in kilotonnes per year (2005)
CO2_2015 | CO2 emissions in kilotonnes per year (2015)

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*. When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

# Preliminaries

Again, you want to clear the working space to make sure that you do not have extraneous items there when you begin your work. The command in R to clear the workspace is `rm` (for "remove"), followed by a list of items to be removed. To clear the workspace from _all_ objects, do the following:

```{r}
rm(list = ls())
```

Begin by loading the datasets that you will need for this exercise. They are available from the Readings section. Note that you also need the package `readxl`. Make sure that it is installed before loading it with `library()`.

```{r}
library(envsocty3LT3) # Course package
library(ggplot2)
```

# Context

Words go here.

# Questions

## Energy Use and Emission Trends (100 marks)

1.	What was the approximate population of Earth circa 2016? How many barrels of oil per day did this population consume? (10)

2.	What proportion of the world's oil is consumed by Canada and the US? Do these two countries have the same per capita levels of demand for energy use? Why do you think this is so? (15)

3.	Calculate the consumption of oil in barrels per day per 1000 people for the countries listed in the data frame. What type of pattern do you observe? Try to explain the pattern. (20)

Hint: rank countries according to their oil consumption per 1000 people. You can use the `order()` function to do this. A good explanation is at the following link:

http://www.statmethods.net/management/sorting.html.

4. Use simple regression analysis to assess the relationship in the dataset between a country's per-capita CO2 emissions (for 2015) and per-capita oil consumption. Start by creating a scatterplot of these two variables; then add labels to your plot (try changing the `size` of the text for clarity).

What does a linear regression say is the relationship between per-capita CO2 emissions and per-capita oil consumption? How significant is the relationship, according to a regression? Which countries are cleaner and which countries are less clean in their use of energy? (25)

5. How much CO2 would be produced, if every person in the world produced CO2 equal to that produced by a Canadian? What if every person in the world produced CO2 equal to that produced by a Chinese citizen in 2015? Compare both results to the actual CO2 production of the world in 2015 and use a linear regression to explain the relationship. (15)

6.	Using the data in the data frame, find the six largest CO2 emitters in 1995, and then calculate the increase or decrease in the amount of generated CO2 emissions for each of them in the time period 1995 to 2015. Compare the size of the numbers. Which of these countries saw the largest increase in emissions since 1995? Why do you think so? (10)

7.	Which country in the world had the highest CO2 emissions per square kilometer of land area in 2015? Which country has the lowest? (5)

### 2021-10-14 Bike-share-spatial-equity/Response-to-Reviewers-R2.Rmd
Proofreading and editing the response to reviewers before resubmitting

--- 2021-08-30 Bike-share-spatial-equity/Response-to-Reviewers-R2.Rmd
@@
 author: ""
-date: "8/23/2021"
+date: "8/27/2021"
 output: pdf_document
@@
 We would like to express our gratitude to the editors and anonymous reviewers for their time and attention to our submission.
+All substantive revisions to the manuscript are underlined in blue and highlighted in yellow.
 
@@
 Floating catchment areas introduce an element of demand (or congestion).
-We clarified in the text that demand in our case is at the docking stations (see _page_).
+We clarified in the text that demand in our case is at the docking stations.
 You are right that the analysis is static in that it does not account for temporal variations in demand.
 Unfortunately, the number of bicycles available at each docking station at various times throughout the day is not available publicly, and we are not able to implement a dynamic analysis at the moment.
-We note this as a direction for future research (see _page_).
+We note this as a direction for future research (see \emph{page 24}).
 With respect to the static analysis, it is true that not all users arrive at a docking station at the same time.
@@
 Similar standards for health care provision exist: not everyone goes to an emergency room at the same time, but calculating ER beds per person is still a good indicator of supply and demand.
-In the case of accessibility to public bicycle share programs, discussing the capacity of the system in these terms may be surprising initially, but we'd argue that this is because no standards have been developed for this kind of public service.}
+In the case of accessibility to public bicycle share programs, discussing the capacity of the system in these terms may be surprising initially, but we'd argue that this is because no standards have been developed for this kind of public transportation service.}
 
@@
 In section 4.1, we mean population units in general; they could be census polygons, dissemination areas, interpolated populatation, or any other cartographic representation of the population.
-We have changed the text to say "population unit" (see _page_) to ensure that this is not confused with the population cells in section 4.2.
-These populations cells are the 50x50m polygons that we used to interpolate the population from census dissemination areas (DAs) to smaller geographical units.
+We have changed the text to say "population unit" (see \emph{pages 14 and 21}) to ensure that this is not confused with the population cells in section 4.2.
+These populations cells are 50x50m polygons that we used to interpolate the population from census dissemination areas (DAs) to smaller geographical units.
 For vertical equity analysis, we used census income data at the level of DAs.
@@
 
-\textcolor{blue}{Very good point. We assumed that the distance decay function would be apparent from the reference to thresholds. We now note on _page_ that the distance decay is as follows:}
+\textcolor{blue}{Very good point. We assumed that the distance decay function would be apparent from the reference to thresholds. We now note on \emph{page 17} that the distance decay is as follows:}
 
@@
 \textcolor{blue}{Thank you for this question.
-In our view, the equity stations were a qualified success: they increased accessibility in general, and more people have higher accessibility than without the equity stations.
-This suggests that horizontal equity improved.
-As well, the accessibility differential for people in DAs with low incomes and people in DAs with higher incomes became smaller, but accessibility still was much lower for lower income populations even after the introduction of the equity stations.
-Hence, we believe that the gap in vertical equity was only modestly addressed.
+In our view, the equity stations were a qualified success: they increased accessibility in general, and more people have higher accessibility than without the equity stations because they filled important gaps in service coverage.
+As well, the accessibility differential for people in DAs with low incomes and people in DAs with higher incomes became smaller, but accessibility was still much lower for lower income populations even after the introduction of the equity stations.
+Hence, we believe that the gap in vertical equity was only modestly addressed and additional equity stations are needed to further increase access for equity users.
 Please see the revised discussion.}

### 2021-10-25 README.Rmd
Updated title in README.Rmd
Knitted README.Rmd after paper accepted by TRD

--- 2021-03-29 README.Rmd
@@
 
-# Accessibility-Sobi-Hamilton
+# Examining spatial equity of bike share: A balanced floating catchment area approach
 
@@
 
-The goal of Accessibility-Sobi-Hamilton is to ...
+Elise Desjardins (McMaster University)  
+Christopher D. Higgins (University of Toronto Scarborough)  
+Antonio Páez (McMaster University)
 
-What is special about using `README.Rmd` instead of just `README.md`? You can include R chunks like so:
+Forthcoming in *Transportation Research Part D: Transport and Environment*
 
-```{r cars}
-summary(cars)
+## Abstract
+
+Public bicycle share programs (PBSPs) can play a role in advancing transportation equity if they make bicycling more accessible to disadvantaged populations.
+In Ontario, Hamilton Bike Share expanded their program in 2018 by adding twelve "equity" stations with the explicit objective of increasing access for under-serviced neighbourhoods.
+In this case study, we investigate differentials in accessibility to stations using a balanced floating catchment area approach and compare accessibility with and without the equity stations.
+We analyze population interpolated to small cells to better reflect walking to a station and conduct a sensitivity analysis at several walking time thresholds.
+We then reaggregate the estimated accessibility by income groups for further analysis.
+Our findings indicate that equity stations increased accessibility for the serviced population at every threshold examined, but the increase was relatively modest especially for population in the bottom 20% of median total household income.
+
+## Keywords
+
+- Public Bike Sharing Programs  
+- Transport equity  
+- Horizontal equity  
+- Vertical equity  
+- Accessibility  
+- Balanced floating catchment areas  
+- Hamilton  
+- Ontario
+- Reproducible research
+
+```{r echo=FALSE}
+rm(list = ls())
 ```
 
-You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date.
-`devtools::build_readme()` is handy for this.
-You could also use GitHub Actions to re-render `README.Rmd` every time you push.
-An example workflow can be found here: <https://github.com/r-lib/actions/tree/master/examples>.
-
-You can also embed plots, for example:
-
-```{r pressure, echo = FALSE}
-plot(pressure)
+```{r install-data-package, include = FALSE}
+# Run only once if needed to install data package `sobiEquity`
+if (!require("sobiEquity", character.only = TRUE)) {
+      devtools::install_github("https://github.com/paezha/Accessibility-Sobi-Hamilton", subdir = "sobiEquity")
+  }
 ```
 
-In that case, don't forget to commit and push the resulting figure files, so they display on GitHub.
+```{r load-packages, include=FALSE}
+library(biscale)
+library(cowplot)
+library(data.table)
+library(disk.frame)
+library(dplyr)
+library(gdistance)
+library(ggmap)
+library(ggspatial)
+library(gridExtra)
+library(kableExtra)
+library(pycno)
+library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.htmllibrary(readr)
+library(raster)
+library(rgdal)
+library(sf)
+library(sobiEquity) # Data package
+library(tidyverse)
+library(tinytex)
+library(units)
+```
+
+```{r generate-package-bibliography, include=FALSE}
+knitr::write_bib(c(.packages(), "knitr", "rticles"), "packages.bib")
+```
+
+```{r invoke-data, include=FALSE}
+# The data is part of data package `sobiEquity` 
+data("community_downtown")
+data("hamilton_cma")
+data("hamilton_da_2016")
+data("population_50x50")
+data("sobi_service")
+data("sobi_hubs")
+data("ttm_walk")
+```
+
+```{r crop-das, include=FALSE}
+# Crop DAs using bounding box of service area
+sobi_service_bbox <- sobi_service %>% 
+  st_bbox() %>%
+  st_as_sfc()
+
+cropped_da <- hamilton_da_2016 %>%
+  st_crop(sobi_service_bbox) %>%
+  dplyr::select(DA, geometry)
+```
+
+\newpage
+
+1. Introduction
+===================
+
+The potential of public bicycle share programs (PBSPs) to increase bicycling levels is but one of many reasons for implementing such programs in urban areas [see @fishmanBikeshareImpactActive2015; @hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019].
+As a healthy, inexpensive, and convenient form of public transportation, shared bicycles can encourage individuals to take up bicycling for short local trips or first and last mile trips to other public transportation instead of using personal vehicles.
+These programs can also play a role in advancing transportation equity if they make bicycling more accessible to disadvantaged populations.
+Although PBSPs are available to the general public in over 800 cities worldwide [@fishmanBikeshareReviewRecent2016], and ought to be accessible to any individual who wishes to use them, research on PBSPs indicates that inequities persist with respect to who can use and access them.
+
+Many PBSPs in North America now offer specific programs to address equity that primarily focus on removing cost barriers and increasing access for groups that are under-represented among existing users [@trec2019].
+The locations of docking stations is a common consideration for increasing access to improve equity [@howlandCurrentEffortsMake2017].
+Hamilton Bike Share (HBS), located in Hamilton, Ontario, was the only Canadian PBSP included in a North American scan of bike share equity programs [see @trec2019].
+HBS was launched in 2015 and currently has over 900 operational bicycles and 130 docking stations.
+An equity program, _Everyone Rides Initiative_, was implemented in 2018 which expanded the program by introducing twelve "equity" stations to more disadvantaged neighbourhoods in the core service area.
+
+This paper examines accessibility to Hamilton Bike Share using a balanced floating catchment area (BFCA) approach.
+We also conduct a comparative analysis to the conventional two-step floating catchment area (2SFCA) approach to highlight the benefits of this method which, to our knowledge, has not yet been used in the cycling literature.
+The paper also assesses the contribution of the equity stations to reducing inequities in accessibility for different groups according to median total household income, and provides policy recommendations to further improve equity.
+
+This paper is an example of open and reproducible research that uses only open software for transportation and statistical analysis [@bivand2020progress; @lovelace2021open].
+All data were obtained from publicly available sources and organized in the form of a data package.
+Following best practices in spatial data science [@brunsdon2020opening], an open data product [@arribas2021open] along with the code needed to reproduce, modify or extend the analysis are available for download.^[https://github.com/paezha/Accessibility-Sobi-Hamilton]
+
+2. Literature Review
+===================
+
+## 2.1. Public Bicycle Share Programs
+
+Public bicycle share programs have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users to date [@fishmanBikeshareReviewRecent2016].
+In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @ogilvieInequalitiesUsagePublic2012; @reillyGenderDisparitiesNew2020; @wangGenderGapGenerators2019; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011].
+However, one study found that bike share users in Washington, DC were more likely to be female [@buckAreBikeshareUsers2013], which suggests that the gender gap among bicyclists who use PBSPs is less disparate than the gap for personal bicycle use [@fishmanBikeshareReviewRecent2016].
+There is some evidence that bike share users are less likely to own a car [@buckAreBikeshareUsers2013; @reillyNoncyclistsFrequentCyclists2020].
+However, the relationship between income or education and bike share use is less clear-cut.
+Stations in disadvantaged communities in Chicago have been found to generate most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighborhoods in Minneapolis-St.
+Paul also used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019].
+Similar findings were reported in London [@ogilvieInequalitiesUsagePublic2012].
+Being university educated was a significant correlate of bike share use in Montreal, Canada [@fullerUseNewPublic2011].
+Not coincidentally, financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016].
+
+## 2.2. Equity of PBSPs
+
+The introduction of PBSPs has been accompanied by a flurry of research focusing on who benefits from them.
+Many studies have examined differences in demographics or socioeconomic status between those who use or have access to PBSPs and those who don't, while other research has explored spatial inequities in where stations are located [e.g., @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @hullgrassoBikeShareEquity2020; @qianBikesharingEquityDisadvantaged2020; @qianBikeshareDestinationChoices2021; @smith2015exploring].
+In Chicago, Qian and Jaller [-@qianBikesharingEquityDisadvantaged2020] estimated ridership in the city's PBSP and found that a minority of docking stations were located in disadvantaged communities, while annual members from such areas had a lower share of trips compared to other areas in the city.
+Similar results were found in Philadelphia.
+Despite efforts to increase equity within the city's PBSP, census block groups with lower median income generated fewer trips [@caspiBikesharingPhiladelphiaLowerincome2019].
+Trips from stations in such areas were utilitarian (e.g., commuting to work), which points to the importance of ensuring equitable access [@caspiBikesharingPhiladelphiaLowerincome2019].
+In the case of Seattle, all neighbourhoods were found to have some level of access to dockless bicycles but those with higher incomes and more residents of higher education had more bicycles [@mooneyFreedomStationSpatial2019].
+Babagoli et al.
+[-@babagoliExploringHealthSpatial2019] also found that neighbourhoods in New York City with higher affluence had the greatest proportion of Citi Bike stations.
+
+Chen et al.
+[-@chenExploringEquityPerformance2019] distinguish two types of equity related to bike share, horizontal and vertical, based on the work of Delbosc and Currie [@delboscUsingLorenzCurves2011].
+Horizontal equity leads to balanced or equal distribution of accessibility and costs for all similar groups, while vertical equity would involve greater and targeted access for under-represented or disadvantaged populations [@chenExploringEquityPerformance2019].
+Both are of interest to researchers and transportation planners since they are often linked in that advantage, or conversely disadvantage, has spatial patterns.
+Bike share equity programs [see @trec2019; @howlandCurrentEffortsMake2017] can be considered efforts to improve vertical equity since they favour groups that have benefited less from PBSPs through the placement of stations or equitable fee structures.
+
+On the whole, existing studies highlight the need for PBSPs to be more accessible for equity populations in order to increase use beyond the "typical" users.
+This has been the focus of recent research [see, among others, @auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020].
+Offering more people the option of using sustainable and active transportation, particularly those who have lower socioeconomic status and might benefit the most, is a worthy policy goal for cities with PBSPs.
+However, exploring transportation equity by investigating where docking stations are located, often using neighborhoods or census tracts as the geographical unit of analysis, can ignore or miss the benefits that may be derived from adjacent zones.
+Meaning that, stations may be lacking in certain neighborhoods but there may be stations accessible within a reasonable walking time.
+This is where geographical accessibility becomes an important consideration.
+
+## 2.3. Accessibility Approaches
+
+Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved over the past century and has increasingly become linked with recent planning interests in prioritizing modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020].
+Beyond the utility derived from using shared bicycles to access destinations of value, docking stations themselves are amenities because they offer a transportation service.
+Therefore, the ease of reaching these stations, which are spread spatially in a given area, can affect use of the PBSP.
+
+The location and size of docking stations are two factors that are relevant to accessibility.
+Since the time or distance needed to reach a docking station decreases the potential of accessing the program, the location matters.
+Kabra et al.
+[-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris, France comes from areas within 300 m of stations, which amounts to 2-4 minutes walking by an adult who does not have a disability.
+Other studies have adopted  [-@hosfordEvaluationImpactPublic2018; -@hosfordEvaluatingImpactImplementing2019] Similar to other public amenities affected by crowding, the utility of docking stations is also limited by the _maximum_ number of bicycles that they can hold (e.g., their size).
+Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts.
+For instance, Wang and Lindsey [-@wangNewBikeShare2019] investigated whether new or relocated bike share stations increased accessibility and use, which offered important insights to improve the performance of the program.
+
+Several approaches have been commonly used for measuring place-based accessibility, including cumulative opportunities, gravity, and utility-based measures [@geursAccessibilityEvaluationLanduse2004; @handyMeasuringAccessibilityExploration1997].
+The gravity-based approach involves weighting destination opportunities, such as the quantity of bike share stations, by the time required to reach them from an origin using an impedance function [@handyMeasuringAccessibilityExploration1997; @kwanSpaceTimeIntegral1998].
+While such measures are suitable for capturing the potential for reaching destinations from a given location, they do not take demand or congestion effects into account which is an important consideration when calculating accessibility for amenities such as bike share stations.
+
+In contrast, floating catchment area (FCA) methods have been widely employed in health care accessibility research.
+The benefit of this approach is that the method incorporates information on capacity and demand in calculating accessibility.
+Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] note "competition for activities with restricted capacity" should be taken into account for land-use components of accessibility.
+The FCA approach is more appropriate and informative than calculating provider-to-population ratios (PPR) that simply divide the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019].
+In particular, the Two-Step Floating Catchment Area (2SFCA) method [@luoMeasuresSpatialAccessibility2003; @radkeSpatialDecompositionsModeling2000] produces flexible catchment areas instead of using rigid boundaries like PPR.
+In the first step, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hospital, weighted by the distance involved in reaching the facility.
+Next, these service level ratios are allocated back to the population centers and summarized as a measure of congested accessibility.
+Thus, this model does a good job of considering potential crowding or competition for services.
+However, overlapping catchment areas from the conventional FCA approach lead to inflation of population totals and deflation of service levels across a study area which generates inaccurate or misleading accessibility estimates [@paezDemandLevelService2019].
+While there have been many methodological innovations in FCA methods [e.g., @delamaterSpatialAccessibilitySuboptimally2013; @luoMeasuresSpatialAccessibility2003; @luoEnhancedTwostepFloating2009; @radkeSpatialDecompositionsModeling2000; @wanThreestepFloatingCatchment2012], a recent improvement to this approach was achieved through a simple and intuitive balancing of the impedance that addressed the effects of demand and service inflation found in earlier FCA approaches [see @paezDemandLevelService2019].
+This provides more useful information because it does not assume that people are limited to service within pre-defined boundaries [@paezDemandLevelService2019].
+
+When measuring accessibility, researchers have also taken different approaches when it comes to the aggregation of data, either by using the individual or household as the smallest unit of analysis or larger spatial zones.
+Previous research on bike share equity has typically used a meso- or macro-level approach with aggregated data from entire neighborhoods or census tracts [@babagoliExploringHealthSpatial2019; @mooneyFreedomStationSpatial2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], although there are recent exceptions [@chenExploringEquityPerformance2019; @chenUnobservedHeterogeneityTransportation2021].
+This is also true for studies examining correlates of bike share demand [@wangNewBikeShare2019].
+Handy and Niemeier [-@handyMeasuringAccessibilityExploration1997] note that using disaggregated data in accessibility analyses provides a more accurate estimate for individuals.
+Chen et al.
+[-@chenExploringEquityPerformance2019, pp.
+530] are in favour of using disaggregated data, which they did in their recent analysis of Tampa's PBSP, because they note "the use of aggregated data might hinder our understanding of the equity impacts since individual disparities are absorbed after aggregation".
+
+## 2.4 Previous Research
+
+Our analysis builds upon a previous study [@hosfordWhoArePublic2018], which found that areas in Hamilton with less advantage are better served by the city's PBSP compared to other Canadian cities (i.e., Toronto, Vancouver, Montreal, and Ottawa-Gatineau) where areas that are less deprived have greater access.
+Hosford and Winters [-@hosfordWhoArePublic2018, pp.
+47] acknowledge that "Hamilton stands out in that the lower income neighborhoods are located near the city center and wealthier neighborhoods are in the surrounding suburban areas".
+Therefore, the core service area for the PBSP in Hamilton by default covers more of the disadvantaged areas in the city.
+However, there is also a great deal of variation in income in the core service area because of the local university and increasing gentrification.
+Hosford and Winters [-@hosfordWhoArePublic2018] took a macro-level approach in their analysis by using dissemination areas _across_ the city as the unit of analysis.
+They did not focus specifically on the core service area and did not differentiate between conventional and equity stations.
+
+Therefore, this paper also contributes to the cycling literature by expanding the analysis of Hosford and Winters [-@hosfordWhoArePublic2018] to assess equity in accessibility to HBS and explore the contribution of the equity stations added to Hamilton's PBSP.
+
+3. Case Study {#sec:study}
+======================
+
+## 3.1. Original System
+
+The case study of this paper is the city of Hamilton, Ontario, Canada.
+Before June 2020, the program was known as Social Bicycles or SoBi Hamilton, but is now called Hamilton Bike Share (HBS).
+The core service area spans 40 sq.km of the city, although it was planned to be 20 sq.km [@hamiltonsobi2015], and roughly 138,000 people are within 30 minutes walking of a station [see Figure \ref{fig:hamilton-and-sobi-service-area}].
+This represents roughly one fifth of the total population of the Hamilton Census Metropolitan Area according to the 2016 Canadian Census.
+The City of Hamilton undertook a large public engagement campaign to validate the locations of stations that had been selected and to crowdsource potential locations for additional locations [@hamiltonsobi2014].
+Most of the potential locations that were suggested were in the east end of the core service area that lack stations or in neighborhoods not serviced by the PBSP.
+The program was enthusiastically welcomed in the city in 2015 - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015], however inadequate coverage given the size of the program's service area was identified as a problem early on, and transportation planners noted that the small size (i.e., supply of bicycle racks) and low quantity of stations would lead to challenges in balancing the system [@hamiltonsobi2015].
+
+## 3.2. Equity Initiative
+
+Hamilton Bike Share Inc., the non-profit operator, launched an equity program, Everyone Rides Initiative (ERI), in 2018 with the objective of reducing barriers that may prevent individuals from accessing the program.
+Additional bicycles and twelve "equity" stations were added to the core service area in more disadvantaged and under-serviced neighbourhoods.
+The equity program also offers subsidized memberships to individuals who identify as low income, and complements this service with cycle skills education.
+A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019].
+
+## 3.3. Current System
+
+As of June 2020, HBS has 900 bikes, 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].
+The core service area remains 40 sq.km.
+The program has twelve "equity" stations and 118 "conventional" stations (i.e., that were not explicitly added to address inequities in the program).
+The City of Hamilton has positioned stations between 300 and 600 m apart [@scottRouteChoiceBike2021], but anticipates that they will service those living within a 250 m buffer [@hamiltonsobi2015].
+The latter constitutes a normative statement: people ought to be able to access a station in less than 600 m if they live in the core service area, with most usage coming from 250 m around.
+However, it is not known how far people are actually willing to travel to reach a station.
+It would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.
+
+```{r hamilton-and-sobi-service-area, echo=FALSE, message=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "The core service area of Hamilton Bike Share is outlined in blue.
+Hamilton Census Metropolitan Area is shown in grey."}
+geo_context <- ggplot() + 
+  geom_sf(data = hamilton_da_2016,
+          color = NA,
+          fill = "gray") + 
+  geom_sf(data = sobi_service,
+          fill = NA,
+          colour = "blue") + 
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank()) +
+  #annotate("text", x = -80, y = 43.5, label = "Lake Ontario") +
+  annotation_scale(location = "bl", width_hint = 0.4) +
+  annotation_north_arrow(location = "tl", which_north = "true", 
+        pad_x = unit(0.25, "in"), pad_y = unit(0.25, "in"),
+        style = north_arrow_fancy_orienteering)
+
+# Inset map
+geo_context_inset <- get_stamenmap(bbox = c(left = -79,
+                                    bottom = 43,
+                                    right = -81,
+                                    top = 44),
+          maptype = "toner-lite", 
+          crop = FALSE,
+          zoom = 8)
+# plot map
+geo_context_inset <- ggmap(geo_context_inset)  +
+    coord_sf(crs = st_crs(4326))  +
+    #xlab("Longitude") + ylab("Latitude") +
+  annotation_scale(location = "bl", width_hint = 0.4) +
+  theme_minimal() +
+  theme(axis.text = element_blank(),
+        panel.border = element_rect(color = "black",
+                                    fill = NA),
+        axis.title = element_blank())
+
+#geo_context + geo_context_inset
+ggdraw() + 
+  draw_plot(geo_context, 0.0, 0.0, 0.8, 1) + 
+  draw_plot(geo_context_inset, 0.5, 0.4, 0.6, 0.6) +
+  draw_plot_label(x = c(.25,.5), y = c(0.45,.6), 
+                  label =  c("Hamilton Bikeshare \nCore Service Area",  "Lake Ontario"),
+                  hjust = 0, 
+                  size = 8) +
+  draw_line(x = c(0.34, 0.4),
+    y = c(0.43, 0.5),
+    color = "blue", 
+    size = 1,
+    lineend = "round",
+    linejoin = "mitre",
+    arrow = arrow(length = unit(0.1, "inches")))
+```
+
+```{r sobi-stations-in-hamilton, echo=FALSE, fig.align = 'center', out.width = "100%", fig.cap = "The spatial distribution of bike share docking stations in Hamilton, Ontario.
+Conventional stations are in purple and equity (ERI) stations are in orange.
+The service area of Hamilton Bike Share is outlined in blue and the city's downtown core is outlined in dark green.
+Hamilton Census Metropolitan Area is shown in grey."}
+ggplot() + 
+  geom_sf(data = cropped_da,
+          color = NA,
+          fill = "gray") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          size = 1,
+          color = "darkgreen") +
+  geom_sf(data = sobi_hubs %>% 
+            dplyr::filter(hub_status == "Active"), # Remove deactivated hubs
+          aes(size = RACKS_AMOU, fill = hub_type),
+          shape = 21) +
+  scale_size(name = "Number of racks", 
+             range = c(0.5, 3.5)) +
+  scale_fill_manual(name = "Hub Type",
+                    values = c("Conventional" = "purple", "ERI" = "orange")) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        legend.position = "bottom",
+        axis.ticks = element_blank()) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "tl", which_north = "true", 
+        pad_x = unit(0.25, "in"), pad_y = unit(0.20, "in"),
+        style = north_arrow_fancy_orienteering)
+```
+
+4. Methods and Data {#sec:methods}
+======================
+
+## 4.1. Balanced Floating Catchment Area
+
+The BFCA method was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by earlier FCA methods [@paezDemandLevelService2019].
+Páez et al.
+[-@paezDemandLevelService2019] adjusted the impedance weights so that both supply and demand are proportionally allocated.
+The result is a FCA method that balances the population and level of service by eliminating the over-counting of population and level of service that leads to distortions in demand and supply.
+
+The first step in the FCA method is to allocate the population to be serviced by each docking station:
+
+\begin{equation}
+\label{eq:population-allocation}
+P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}
+\end{equation}
+
+As seen in the equation above, the population allocated to station $j$ is the weighted sum of the population in the region; a spatial weight $w_{ij}$ represents the friction that the population at $i$ faces when reaching station $j$, and is usually given by a distance-decay function, so that each station is assumed to service only a segment of the population within a limited geographical range.
+The level of service of station $j$ in bicycle racks per person is the supply at each station (i.e., the maximum number of bicycle racks) divided by the population within the established catchment area:
+\begin{equation}
+\label{eq:level-of-service}
+L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
+\end{equation}
+
+In the second step, the accessibility of population unit $i$ is calculated as the weighted sum of the level of service of all stations that can be reached from there according to the spatial weights:
+\begin{equation}
+\label{eq:FCA-accessibility}
+A_i = {\sum_{j = 1}^{J} L_j{w_{ij}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ij}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
+\end{equation}
+
+The balanced approach of Páez et al. [-@paezDemandLevelService2019] replaces the spatial weights with normalized versions as follows. In the first step, the population is weighted with:
+\begin{equation}
+\label{eq:spatial-weights}
+{w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ij}}}}
+\end{equation}
+
+\noindent and in the second step, the level of service is weighted using:
+\begin{equation}
+\label{eq:spatial-weights-2}
+{w_{ij}^{j} = \frac {w_{ij}}{\sum_{i = 1}^{n} {w_{ij}}}}
+\end{equation}
+
+These weights satisfy the following properties:
+\begin{equation}
+\label{eq:weights}
+\sum_{j = 1}^{J} {w^i_{ij}} = 1
+\end{equation}
+
+\noindent and:
+\begin{equation}
+\label{eq:weights-2}
+\sum_{i = 1}^{n} {w^j_{ij}} = 1
+\end{equation}
+
+With these weights, accessibility can be calculated without risk of demand or supply inflation:
+\begin{equation}
+\label{eq:balanced-accessibility}
+A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}
+\end{equation}
+
+By allocating the population and level of service proportionally, this method preserves the values of the population and level of service since:
+\begin{equation}
+\label{eq:proportionality}
+\sum_{i=1}^n A_i = \sum_{j=1}^J L_j
+\end{equation}
+
+In fact, since the proportional allocation procedure means that any proportion of the population allocated to a station is never allocated to other stations, and conversely any level of service allocated to a population is never re-allocated elsewhere, this property is replicated for any level of aggregation.
+
+## 4.2. Pycnophylactic Interpolation
+
+Since walking trips to a docking station likely happen at a level lower than even the smallest census geography, this requires a more disaggregated approach than the use of census geographies.
+For this reason, we implemented our analysis parting from small population cells to better reflect the friction of walking to a docking station, which is an important component of a bike share trip [@chenExploringEquityPerformance2019].
+
+To obtain population at sub-census geography levels, we used pycnophylactic interpolation [@toblerSmoothPycnophylacticInterpolation1979].
+We obtained population data from the 2016 Census of Canada for dissemination areas (DA), which is the smallest publicly available census geography in Canada.
+These zonal values of the population were interpolated to smaller polygons that are 50-by-50 m in size.
+Pycnophylactic interpolation involves smoothing out the population from each DA while preserving total volume.
+When interpolating the population at this high level of resolution, it is important to ensure that population numbers are not allocated to areas where people do not live in Hamilton (for example, to parks or large institutional buildings, etc.).
+To do so, we retrieved shapefiles for various geographic features from Open Hamilton.
+Next, we removed these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population within the remaining area [see Figure \ref{fig:interpolated-population}].
+
+```{r interpolated-population, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Interpolated population in Hamilton Bike Share's core service area (outlined in blue) and within 30 minutes of walking to the core service area.
+Each population cell is 50-by-50 m in size.
+The downtown area is outlined in dark green."}
+ggplot() + 
+  geom_sf(data = cropped_da,
+          fill = "gray",
+          color = NA) + 
+  geom_tile(data = cbind(st_drop_geometry(population_50x50), 
+                         st_coordinates(population_50x50)),
+            aes(x = X, 
+                y = Y, 
+                fill = population)) +
+  scale_fill_distiller(name = "Population",
+                       palette = "YlOrRd",
+                       direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA,
+          colour = "blue") +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.title = element_blank(),
+        axis.ticks = element_blank()) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "tl", which_north = "true", 
+        pad_x = unit(0.25, "in"), pad_y = unit(0.20, "in"),
+        style = north_arrow_fancy_orienteering)
+```
+
+## 4.3. Travel Time Matrix
+
+To calculate walking times from the centroid of the population cells to docking stations, we extracted OpenStreetMap data for HBS's service area from [BBBike](https://download.bbbike.org/osm/bbbike/), an online cycle route planner that interfaces with OpenStreetMap.
+OpenStreetMap data provides the networks for calculating walking times from each population cell to nearby docking stations, using a maximum walking distance of 10 km and walking time of 30 minutes as thresholds.
+A travel time matrix was created with the origins as the coordinates of the population cells and the destinations as the coordinates of the docking stations within the maximum threshold.
+This process provides a more realistic measure of the friction of reaching stations by taking infrastructure into account network travel times, rather than using the Euclidean distance from population cell to station.
+Routing and travel time calculations were completed using the `R` package `r5r`, used for rapid realistic routing operations [@Pereira2021r5r].
+
+## 4.4. Data
+
+All data for this research were accessed from publicly available Census of Canada sources, from OpenStreetMaps, and from Open Hamilton^[https://open.hamilton.ca/], a public online repository of data curated by the City of Hamilton.
+Median total household income statistics were drawn from the 2016 Canadian Census.
+
+5. Results
+===================
+
+## 5.1. Accessibility by Distance Thresholds
+
+Consensus regarding the distance that individuals are willing to walk to access a docking station is lacking, but the literature on walking behaviour provided some guidelines to determine the thresholds for our sensitivity analysis.
+Previous studies have found that living within 250 m [@fullerUseNewPublic2011] and 300 m [@kabraBikeShareSystemsAccessibility2020] of a docking station is correlated with bike share use.
+Some studies have defined "close proximity" to a docking station as 500 metres or less [@fullerImpactEvaluationPublic2013; @hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019], while other research has found that walking trips are less than 600 m and rarely more than 1200 m [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 m [@larsenQuarterMileReexamining2010].
+HBS will often depict a map at some docking stations to show the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are expected to walk.
+The National Association of City Transportation Officials (NACTO) has a similar normative guide [@nactowalkingstation2015].
+
+In the present case, we found that accessibility calculated using the BFCA method increased with a threshold between two and four minutes, but was then maximized at five minutes.
+Accessibility decreased substantially after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each station.
+
+We experimented with various thresholds by conducting a sensitivity analysis to calculate accessibility at different walking times from population cell to docking stations: three minutes, five minutes, ten minutes, and fifteen minutes.
+We categorized these thresholds as minimum, average, maximum, and extreme, respectively.
+At each threshold, we compared accessibility between the current system and the original system to examine the contribution of the twelve equity stations.
+When considering the results reported below, it is important to remember that accessibility is technically a form of smoothing [@okelly2003aggregate, pp.
+7-8]: smaller thresholds produce less smoothing (which can result in "spiky" accessibility landscapes), while larger thresholds produce more smoothing and fewer spikes.
+
+In the analysis we use the following distance decay function: 
+$$
+{w_{ij} = \begin{cases} 
+      1 & \text{when } t_{ij}\leq \gamma \\
+      0 & \text{otherwise} 
+   \end{cases}}
+$$
+\noindent where $\gamma$ is the relevant threshold. The weights are standardized as discussed above.
+
+```{r join-population-to-travel-time-data, include=FALSE}
+ttm_walk <- left_join(ttm_walk, 
+                      population_50x50 %>%
+                        st_drop_geometry(),
+                      by = "UID")
+```
+
+```{r join-sobi-station-to-travel-time-data, include=FALSE}
+ttm_walk <- ttm_walk %>% 
+  left_join(sobi_hubs %>%
+              st_drop_geometry() %>%
+              dplyr::select(OBJECTID, 
+                            RACKS_AMOU),
+            by = c("OBJECTID" = "OBJECTID"))
+```
+
+```{r, include=FALSE}
+names(ttm_walk) <- c("UID", 
+                     "hub", 
+                     "travel_time",
+                     "hub_type",
+                     "hub_status",
+                     "population",
+                     "racks")
+```
+
+```{r accessibility-sensibility, eval=FALSE, include=FALSE}
+# Testing different thresholds to check how accessibility changes
+accessibility_test <- data.frame(threshold = c(3:15), 
+                                 accessibility = numeric(13))
+for(i in 1:13){
+  results_test <- ttm_walk %>%
+    dplyr::filter(hub_type == "Conventional" &
+                    hub_status == "Active") %>%
+    b2sfca(threshold = i +2 )
+  accessibility_test$accessibility[i] <- sum(results_test$accessibility$accessibility)
+}
+
+ggplot(accessibility_test, 
+       aes(x = threshold, y = accessibility)) +
+  geom_line()
+```
+
+<!-- Calculate accessibility for different thresholds -->
+
+```{r 3-min-conventional, include=FALSE}
+results_3min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 3)
+```
+
+```{r 3-min-all, include=FALSE}
+results_3min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 3)
+```
+
+```{r 3-min-sum-conventional, include=FALSE}
+sum(results_3min_conventional$los$los)
+sum(results_3min_conventional$accessibility$accessibility)
+```
+
+```{r 3-min-sum-all, include=FALSE}
+sum(results_3min_all$los$los)
+sum(results_3min_all$accessibility$accessibility)
+```
+
+```{r 5-min-conventional, include=FALSE}
+results_5min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 5)
+```
+
+```{r 5-min-all, include=FALSE}
+results_5min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 5)
+```
+
+```{r, 5-min-sum-conventional, include=FALSE}
+sum(results_5min_conventional$los$los)
+sum(results_5min_conventional$accessibility$accessibility)
+```
+
+```{r 5-min-sum-all, include=FALSE}
+sum(results_5min_all$los$los)
+sum(results_5min_all$accessibility$accessibility)
+```
+
+```{r 10-min-conventional, include=FALSE}
+results_10min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 10)
+```
+
+```{r 10-min-all, include=FALSE}
+results_10min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 10)
+```
+
+```{r 10-min-sum-conventional, include=FALSE}
+sum(results_10min_conventional$los$los)
+sum(results_10min_conventional$accessibility$accessibility)
+```
+
+```{r, 10-min-sum-all, include=FALSE}
+sum(results_10min_all$los$los)
+sum(results_10min_all$accessibility$accessibility)
+```
+
+```{r 15-min-conventional, include=FALSE}
+results_15min_conventional <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 15)
+```
+
+```{r 15-min-all, include=FALSE}
+results_15min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 15)
+```
+
+```{r 15-min-sum-conventional, include=FALSE}
+sum(results_15min_conventional$los$los)
+sum(results_15min_conventional$accessibility$accessibility)
+```
+
+```{r 15-min-sum-all, include=FALSE}
+sum(results_15min_all$los$los)
+sum(results_15min_all$accessibility$accessibility)
+```
+
+```{r, bind-all-results, include=FALSE}
+los <- rbind(data.frame(results_3min_all$los, system = "All stations", threshold = "3 min"),
+             data.frame(results_3min_conventional$los, system = "Conventional", threshold = "3 min"),
+             data.frame(results_5min_all$los, system = "All stations", threshold = "5 min"),
+             data.frame(results_5min_conventional$los, system = "Conventional", threshold = "5 min"),
+             data.frame(results_10min_all$los, system = "All stations", threshold = "10 min"),
+             data.frame(results_10min_conventional$los, system = "Conventional", threshold = "10 min"),
+             data.frame(results_15min_all$los, system = "All stations", threshold = "15 min"),
+             data.frame(results_15min_conventional$los, system = "Conventional", threshold = "15 min"))
+
+accessibility <- rbind(data.frame(results_3min_all$accessibility, system = "All stations", threshold = "3 min"),
+                       data.frame(results_3min_conventional$accessibility, system = "Conventional", threshold = "3 min"),
+                       data.frame(results_5min_all$accessibility, system = "All stations", threshold = "5 min"),
+                       data.frame(results_5min_conventional$accessibility, system = "Conventional", threshold = "5 min"),
+                       data.frame(results_10min_all$accessibility, system = "All stations", threshold = "10 min"),
+                       data.frame(results_10min_conventional$accessibility, system = "Conventional", threshold = "10 min"),
+                       data.frame(results_15min_all$accessibility, system = "All stations", threshold = "15 min"),
+                       data.frame(results_15min_conventional$accessibility, system = "Conventional", threshold = "15 min"))
+```
+
+```{r, add-geometry-to-results, include=FALSE}
+los <- los %>%
+  left_join(sobi_hubs %>%
+              dplyr::select(OBJECTID),
+            by = c("hub" = "OBJECTID")) %>%
+  st_as_sf()
+
+accessibility <- accessibility %>%
+  left_join(population_50x50 %>%
+              dplyr::select(UID),
+            by = c("UID" = "UID")) %>%
+  st_as_sf()
+```
+
+### 5.1.1. Minimum Threshold
+
+With a walking distance of three minutes, we found that the total level of service is 25.2 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+The addition of equity stations increases this slightly to 25.4 bicycles per person.
+This total level of service is allocated to the population to obtain the levels of accessibility, which turn out to be relatively uniform overall, with the exception of two small areas where accessibility is slightly higher.
+This high level of system-wide accessibility occurs because the population that can reach a docking station when travel time is three minutes or less is very limited, and accessibility is strongly shaped by a few locations that concentrate population and stations.
+For this reason, accessibility is highly concentrated in small areas around those stations.
+The map is not shown, since it is less informative, but can be recreated using the source code and data.
+
+```{r figure-6, eval = FALSE, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 3 minutes walk (minimum threshold) compared between current system with equity stations and the original system without equity stations."}
+#accessibility plot for current system with equity stations at 3 minutes walk
+accessibility_3min <- accessibility %>% 
+  dplyr::filter(threshold == "3 min")
+
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "gray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_3min), st_coordinates(accessibility_3min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_distiller(palette = "PuRd", direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ threshold)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
+
+```
+
+### 5.1.2. Medium Threshold
+
+With a walking distance of five minutes, we found that there are 68.6 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+With the addition of equity stations, there are now 68.8 bicycle racks per person.
+At this threshold, there are more bicycle racks per person than at the minimum threshold.
+System-wide accessibility has in fact increased: the population that can reach the stations has grown, but not to the the point that congestion effects begin to take place.
+Figure \ref{fig:figure-7} presents a comparison of accessibility between the systems.
+Again, accessibility is fairly uniform, with the exception of one very small area.
+The equity stations noticeably increase accessibility in the east end of the core service area by filling gaps in PBSP coverage.
+
+```{r figure-7-a, include=FALSE}
+#accessibility plot for current system with equity stations at 5 minutes walk
+accessibility_5min <- accessibility %>% 
+  dplyr::filter(threshold == "5 min",
+                system == "All stations") %>%
+  mutate(accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.001, 0.02, 0.4, 8, Inf), 
+                             labels=c("<0.001", "0.001-0.02", "0.02-0.4", "0.4-8", ">8")),
+         accessibility = factor(accessibility,
+                                levels = c(">8", "0.4-8", "0.02-0.4", "0.001-0.02", "<0.001"),
+                                ordered = TRUE))
+
+acc_5_all <- ggplot() +
+  geom_sf(data = population_50x50,
+          color = "gray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_5min), 
+                         st_coordinates(accessibility_5min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_brewer(name = "Accessibility\n(racks/person)",
+                    palette = "PuRd", 
+                    direction = -1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ .)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
+
+# Extract the legend
+acc_5_legend <- cowplot::get_legend(acc_5_all)
+```
+
+```{r figure-7-b, include=FALSE}
+#accessibility plot for current system with equity stations at 5 minutes walk
+accessibility_5min <- accessibility %>% 
+  dplyr::filter(threshold == "5 min",
+                system == "Conventional") %>%
+  mutate(accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.001, 0.02, 0.4, 8, Inf), 
+                             labels = c("<0.001", "0.001-0.02", "0.02-0.4", "0.4-8", ">8")),
+         accessibility = factor(accessibility,
+                                levels = c(">8", "0.4-8", "0.02-0.4", "0.001-0.02", "<0.001"),
+                                ordered = TRUE))
+
+acc_5_conv <- ggplot() +
+  geom_sf(data = population_50x50,
+          color = "gray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_5min),
+                         st_coordinates(accessibility_5min)),
+            aes(x = X, 
+                y = Y, 
+                fill = accessibility)) +
+  scale_fill_brewer(name = "Accessibility\n(racks/person)",
+                    palette = "PuRd", 
+                    direction = -1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ .)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"))
+```
+
+```{r figure-7, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 5 minutes walk (average threshold) compared between current system with equity stations and the original system without equity stations."}
+ggdraw() + 
+  draw_plot(acc_5_all + 
+              theme(legend.position = "none"), 
+            0.00, 0.5, 1, 0.5) + 
+  draw_plot(acc_5_conv + 
+              theme(legend.position = "none"), 0.00, 0.0, 1, 0.5) + 
+  draw_plot(acc_5_legend, 0.01, 0.4, 0.2, 0.2)
+```
+
+### 5.1.3. High Threshold
+
+With a walking distance of ten minutes, we found that there are 3.61 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+With the addition of equity stations, there are now 3.74 bicycles per person.
+Figure \ref{fig:figure-8} presents a comparison of accessibility between the systems.
+Differences in accessibility across the service area are now apparent, with users near the university and its adjacent neighborhoods, as well as neighborhoods north of the downtown area (the latter is outlined in green), having slightly higher accessibility.
+While the differences are modest, they are more apparent at this threshold than at shorter walking distances, especially in the more disadvantaged neighbourhoods in the east end of the core service area where equity stations were added.
+
+```{r figure-8-a, include=FALSE}
+#accessibility plot for current system with equity stations at 10 minutes walk
+accessibility_10min <- accessibility %>% 
+  dplyr::filter(threshold == "10 min",
+                system == "All stations")  %>%
+  mutate(accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.00004, 0.0003, 0.002, Inf), 
+                             labels = c("<0.00004", "0.00004-0.0003", "0.0003-0.002", ">0.002")),
+         accessibility = factor(accessibility,
+                                levels = c( ">0.002", "0.0003-0.002", "0.00004-0.0003", "<0.00004"),
+                                ordered = TRUE))
+
+acc_10_all <- ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_10min), 
+                         st_coordinates(accessibility_10min)),
+            aes(x = X, 
+                y = Y, 
+                fill = accessibility)) +
+  scale_fill_brewer(name = "Accessibility\n(racks/person)",
+                    palette = "PuRd",
+                    direction = -1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ .)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
+
+# Extract the legend
+acc_10_legend <- cowplot::get_legend(acc_10_all)
+```
+
+```{r figure-8-b, include=FALSE}
+#accessibility plot for current system without equity stations at 10 minutes walk
+accessibility_10min <- accessibility %>% 
+  dplyr::filter(threshold == "10 min",
+                system == "Conventional")  %>%
+  mutate(accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.00004, 0.0003, 0.002, Inf), 
+                             labels = c("<0.00004", "0.00004-0.0003", "0.0003-0.002", ">0.002")),
+         accessibility = factor(accessibility,
+                                levels = c( ">0.002", "0.0003-0.002", "0.00004-0.0003", "<0.00004"),
+                                ordered = TRUE))
+
+acc_10_conv <- ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_10min), 
+                         st_coordinates(accessibility_10min)),
+            aes(x = X, 
+                y = Y, 
+                fill = accessibility)) +
+  scale_fill_brewer(name = "Accessibility\n(racks/person)",
+                    palette = "PuRd",
+                    direction = -1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ .)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"))
+```
+
+```{r figure-8, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) compared between current system with equity stations and the original system without equity stations."}
+ggdraw() + 
+  draw_plot(acc_10_all + 
+              theme(legend.position = "none"), 
+            0.00, 0.5, 1, 0.5) + 
+  draw_plot(acc_10_conv + 
+              theme(legend.position = "none"), 0.00, 0.0, 1, 0.5) + 
+  draw_plot(acc_10_legend, 0.01, 0.4, 0.2, 0.2)
+```
+
+### 5.1.4. Extreme Threshold
+
+With a walking distance of fifteen minutes, we found that there are 2.44 bicycle racks per person system-wide in the original system configuration (i.e., without equity stations).
+With the addition of equity stations, there are now 2.55 bicycles per person.
+Figure \ref{fig:figure-9} presents a comparison of accessibility between the systems.
+Users near the university and the neighborhoods north of the downtown area (the latter is outlined in green) have the highest accessibility, followed by those who live in the city's downtown area.
+Accessibility in the east end, where equity stations were implemented, of the core service area remains lower than other areas.
+
+```{r figure-9-a, include=FALSE}
+#accessibility plot for current system with equity stations at 15 minutes walk
+accessibility_15min <- accessibility %>% 
+  dplyr::filter(threshold == "15 min",
+                system == "All stations")  %>%
+  mutate(accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.000006, 0.00005, 0.0003, Inf), 
+                             labels = c("<0.000006", "0.000006-0.00005", "0.00005-0.0003", ">0.0003")),
+         accessibility = factor(accessibility,
+                                levels = c( ">0.0003", "0.00005-0.0003", "0.000006-0.00005", "<0.000006"),
+                                ordered = TRUE))
+
+acc_15_all <- ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_15min), st_coordinates(accessibility_15min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_brewer(name = "Accessibility\n(racks/person)",
+                    palette = "PuRd", 
+                    direction = -1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ .)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
+
+# Extract the legend
+acc_15_legend <- cowplot::get_legend(acc_15_all)
+```
+
+```{r figure-9-b, include=FALSE}
+#accessibility plot for current system with equity stations at 15 minutes walk
+accessibility_15min <- accessibility %>% 
+  dplyr::filter(threshold == "15 min",
+                system == "Conventional")  %>%
+  mutate(accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.000006, 0.00005, 0.0003, Inf), 
+                             labels = c("<0.000006", "0.000006-0.00005", "0.00005-0.0003", ">0.0003")),
+         accessibility = factor(accessibility,
+                                levels = c( ">0.0003", "0.00005-0.0003", "0.000006-0.00005", "<0.000006"),
+                                ordered = TRUE))
+
+acc_15_conv <- ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_15min), st_coordinates(accessibility_15min)),
+            aes(x = X, y = Y, fill = accessibility)) +
+  scale_fill_brewer(name = "Accessibility\n(racks/person)",
+                    palette = "PuRd", 
+                    direction = -1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ .)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"))
+```
+
+```{r figure-9, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 15 minutes walk (extreme threshold) compared between current system with equity stations and the original system without equity stations."}
+ggdraw() + 
+  draw_plot(acc_15_all + 
+              theme(legend.position = "none"), 
+            0.00, 0.5, 1, 0.5) + 
+  draw_plot(acc_15_conv + 
+              theme(legend.position = "none"), 0.00, 0.0, 1, 0.5) + 
+  draw_plot(acc_15_legend, 0.01, 0.4, 0.2, 0.2)
+```
+
+## 5.2. Comparative Analysis
+
+To highlight the benefit of using the BFCA approach, we conducted a comparative analysis with the 2SFCA method.
+The latter method does not allocate population and level of service proportionally, which means that there are often substantial amounts of multiple counting each.
+Figure \ref{fig:figure-10} shows the level of inflation and deflation at five minutes that results when the balanced and the conventional FCA methods are compared.
+The inflation and deflation is defined as the ratio of the level of service and accessibility calculated using the conventional approach to the balanced approach.
+This effect has been documented in other studies [@chenMeasuringAccessibilityHealth2021; @paezDemandLevelService2019].
+Figure \ref{fig:figure-10} shows that in some locations the level of service is deflated by up to 80%.
+This is evident for the equity stations in the east end of the core service area.
+Noticeably, the level of service is least deflated in the downtown area, likely due to the density of docking stations.
+Figure \ref{fig:figure-11} shows that accessibility is inflated by as much as 50% in many areas across the city.
+
+```{r conventional-5-min-non-equity, include=FALSE}
+conventional_5min_non_equity <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  c2sfca(threshold = 5)
+```
+
+```{r conventional-5-min-all, include=FALSE}
+conventional_5min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  c2sfca(threshold = 5)
+```
+
+```{r conventional-5-min-sum-non-equity, include=FALSE}
+sum(conventional_5min_non_equity$los$los)
+sum(conventional_5min_non_equity$accessibility$accessibility)
+```
+
+```{r conventional-5-min-sum-all, include=FALSE}
+sum(conventional_5min_non_equity$los$los)
+sum(conventional_5min_non_equity$accessibility$accessibility)
+```
+
+```{r balanced-5-min-non-equity, include=FALSE}
+balanced_5min_non_equity <- ttm_walk %>%
+  dplyr::filter(hub_type == "Conventional" &
+                  hub_status == "Active") %>%
+  b2sfca(threshold = 5)
+```
+
+```{r balanced-5-min-all, include=FALSE}
+balanced_5min_all <- ttm_walk %>%
+  dplyr::filter(hub_status == "Active") %>%
+  b2sfca(threshold = 5)
+```
+
+```{r balanced-5-min-sum-non-equity, include=FALSE}
+sum(balanced_5min_non_equity$los$los)
+sum(balanced_5min_non_equity$accessibility$accessibility)
+```
+
+```{r balanced-5-min-sum-all, include=FALSE}
+sum(conventional_5min_all$los$los)
+sum(conventional_5min_all$accessibility$accessibility)
+```
+
+```{r los-inflation-5-min, include=FALSE}
+los_inflation <- data.frame(hub = conventional_5min_non_equity$los$hub, 
+                       los_inflation = conventional_5min_non_equity$los$los/balanced_5min_non_equity$los$los) %>% # Ratio of conventional los to balanced los
+  left_join(sobi_hubs %>%
+              dplyr::select(OBJECTID),
+            by = c("hub" = "OBJECTID")) %>%
+  st_as_sf()
+
+accessibility_inflation <- data.frame(UID = conventional_5min_non_equity$accessibility$UID,
+                            accessibility_inflation = conventional_5min_non_equity$accessibility$accessibility/balanced_5min_non_equity$accessibility$accessibility) %>%
+  left_join(population_50x50 %>%
+              dplyr::select(UID),
+            by = c("UID" = "UID")) %>%
+  st_as_sf()
+```
+
+```{r figure-10, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Deflation of level of service at 5 minutes walk (average threshold) in the current system with equity stations."}
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "gray") +
+  geom_sf(data = los_inflation,
+            aes(size = los_inflation, color = los_inflation)) +
+  scale_color_distiller(name = "Deflation of Level of Service",
+                        palette = "PuRd", 
+                        direction = 1) +
+  guides(size = guide_legend(title = "")) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", 
+                   width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), 
+                   pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "tr", 
+                         which_north = "true", 
+        pad_x = unit(0.0, "in"), 
+        pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
+```
+
+```{r figure-11, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Inflation of accessibility at 5 minutes walk (average threshold) in the current system with equity stations."}
+ggplot() +
+  geom_sf(data = population_50x50,
+          color = "gray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_inflation), st_coordinates(accessibility_inflation)),
+            aes(x = X, y = Y, fill = accessibility_inflation)) +
+  scale_fill_distiller(name = "Inflation of accessibility",
+                       palette = "PuRd", 
+                       direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", 
+                   width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), 
+                   pad_y = unit(0.2, "in"),) +
+  annotation_north_arrow(location = "tr", 
+                         which_north = "true", 
+        pad_x = unit(0.0, "in"), 
+        pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
+```
+
+## 5.3. Accessibility by Median Total Household Income
+
+To examine whether accessibility to HBS was increased for groups with lower income, we estimated what level of accessibility accrues to how many people in different income strata.
+To this end, we took our small population cells and aggregated the accessibility and population by the median total household income as imputed from the dissemination areas.
+A unique property of the BFCA method, which does not hold for the 2SFCA approach because of the inflation/deflation issues discussed above, is that accessibility and population data can be re-aggregated using income as an aggregator criterion, while preserving the total population and the level of service.
+This avoids demand and supply inflation, and also enables us to present findings in a way that is more intuitive to interpret.
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} depict bivariate choropleth maps that combine the spatial distribution of accessibility and median total household income, using tertiles for the coloring scheme.
+
+```{r, include=FALSE}
+# drop geometry
+test <- accessibility
+
+# join accessibility to population cells
+test <- test %>% 
+  st_join(population_50x50 %>% 
+            dplyr::select(-UID)) %>%
+  rename(population50x50 = population)
+```
+
+```{r, include=FALSE}
+test <- test %>%
+  st_join(hamilton_da_2016)
+```
+
+```{r, include=FALSE}
+agg_accessibility_da <- test %>%
+  group_by(DA, 
+           system,
+           threshold) %>%
+  summarize(accessibility = sum(accessibility),
+            median_total_income = first(median_total_income), 
+            .groups = "drop")
+```
+
+```{r figure-12, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "3 min", 
+                system == "All stations") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_3_all <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") +
+  ggtitle("All stations") +
+  bi_scale_fill(pal = "GrPink", 
+                dim = 3) +
+  #annotation_scale(location = "bl", width_hint = 0.4,
+  #                 pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+
+legend <- bi_legend(pal = "GrPink",
+                    dim = 3,
+                    xlab = "Higher accessibility ",
+                    ylab = "Higher income",
+                    size = 6) + 
+  theme(plot.background = element_rect(color = "lightgrey"))
+```
+
+```{r figure-13, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "3 min", 
+                system == "Conventional") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_3_conv <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") +
+  ggtitle("Conventional") +
+  bi_scale_fill(pal = "GrPink", 
+                dim = 3) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  #annotation_north_arrow(location = "tr", which_north = "true", 
+  #      pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+  #      style = north_arrow_fancy_orienteering) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-bi-map-threshold-3, echo = FALSE, out.width="1.2\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-3}Bivariate map of accessibility and income at the minimum threshold of three minutes with equity stations (top panel) and without equity stations (bottom panel)."}
+ggdraw() + 
+  draw_plot(bi_map_3_all, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_3_conv, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
+```
+
+```{r figure-14, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "5 min", 
+                system == "All stations") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_5_all <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") +
+  ggtitle("All stations") +
+  bi_scale_fill(pal = "GrPink", 
+                dim = 3) +
+  #annotation_scale(location = "bl", width_hint = 0.4,
+  #                 pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-15, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "5 min", 
+                system == "Conventional") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_5_conv <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") +
+  ggtitle("Conventional") +
+  bi_scale_fill(pal = "GrPink", 
+                dim = 3) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  #annotation_north_arrow(location = "tr", which_north = "true", 
+  #      pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+  #      style = north_arrow_fancy_orienteering) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-bi-map-threshold-5, echo = FALSE, out.width="1.2\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-5}Bivariate map of accessibility and income at the average threshold of five minutes with equity stations (top panel) and without equity stations (bottom panel)."}
+ggdraw() + draw_plot(bi_map_5_all, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_5_conv, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
+```
+
+```{r figure-16, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "10 min", 
+                system == "All stations") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_10_all <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") +
+  ggtitle("All stations") +
+  bi_scale_fill(pal = "GrPink", 
+                dim = 3) +
+  #annotation_scale(location = "bl", width_hint = 0.4,
+  #                 pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-17, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "10 min", 
+                system == "Conventional") %>%
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_10_conv <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") +
+  ggtitle("Conventional") +
+  bi_scale_fill(pal = "GrPink", 
+                dim = 3) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  #annotation_north_arrow(location = "tr", which_north = "true", 
+  #      pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+  #      style = north_arrow_fancy_orienteering) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-bi-map-threshold-10, echo = FALSE, out.width="1.2\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-10}Bivariate map of accessibility and income at the maximum threshold of ten minutes with equity stations (top panel) and without equity stations (bottom panel)."}
+ggdraw() + draw_plot(bi_map_10_all, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_10_conv, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
+```
+
+```{r figure-18, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "15 min", 
+                system == "All stations") %>%
+  drop_na() %>% 
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_15_all <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") +
+  ggtitle("All stations") +
+  bi_scale_fill(pal = "GrPink", 
+                dim = 3) +
+  #annotation_scale(location = "bl", width_hint = 0.4,
+  #                 pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-19, echo=FALSE, fig.align = 'center', out.width = "90%"}
+data_map <- agg_accessibility_da %>%
+  dplyr::filter(threshold == "15 min", 
+                system == "Conventional") %>%
+  drop_na() %>% 
+  bi_class(x = accessibility, 
+           y = median_total_income, 
+           style = "quantile", 
+           dim = 3) %>%
+  st_as_sf()
+
+bi_map_15_conv <- ggplot() +
+  geom_sf(data = data_map %>%
+            st_buffer(dist = 25, endCapStyle = "SQUARE"),
+          mapping = aes(fill = bi_class), 
+          color = NA,
+          show.legend = FALSE) +
+  geom_sf(data = cropped_da,
+          fill = NA,
+          color = "lightgrey") +
+  geom_sf(data = sobi_service, 
+          fill = NA,
+          color = "blue") +
+  ggtitle("Conventional") +
+  bi_scale_fill(pal = "GrPink", 
+                dim = 3) +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in")) +
+  #annotation_north_arrow(location = "tr", which_north = "true", 
+  #      pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+  #      style = north_arrow_fancy_orienteering) +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        plot.margin = unit(c(0, 0, 0, 0), "cm"),
+        plot.title = element_text(hjust=0.5))
+```
+
+```{r figure-bi-map-threshold-15, echo = FALSE, out.width="1.2\\linewidth", fig.cap="\\label{fig-bivariate-map-threshold-15}Bivariate map of accessibility and income at the extreme threshold of fifteen with equity stations (top panel) and without equity stations (bottom panel)."}
+ggdraw() + draw_plot(bi_map_15_all, 0.00, 0.5, 1, 0.5) + 
+  draw_plot(bi_map_15_conv, 0.00, 0.0, 1, 0.5) + 
+  draw_plot(legend, 0.06, 0.4, 0.2, 0.2)
+```
+
+```{r, include=FALSE}
+# Find the income quantiles of the population served by SoBi
+iq <- hamilton_da_2016 %>% 
+  dplyr::filter(sobi_service == "Yes") %>%
+  pull(median_total_income) %>%
+  quantile(c(0, 0.2, 0.4, 0.6, 0.8, 1), na.rm = TRUE)
+```
+
+```{r population-core-service-by-income, include=FALSE}
+# Find the population by income quantile
+population_by_income <- population_50x50 %>%
+  dplyr::select(-population) %>%
+  st_join(hamilton_da_2016) %>%
+  st_drop_geometry() %>%
+  distinct(DA,
+           .keep_all = TRUE) %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE)) %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population)) %>%
+  drop_na()
+```
+
+```{r income-all-3, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_all_3 <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "3 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_all_3 <- agg_accessibility_all_3 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-con-3, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_con_3 <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "3 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_con_3 <- agg_accessibility_con_3 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-all-5, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_all_5 <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "5 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_all_5 <- agg_accessibility_all_5 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-con-5, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_con_5 <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "5 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_con_5 <- agg_accessibility_con_5 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-all-10, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_all_10 <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_all_10 <- agg_accessibility_all_10 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-con-10, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_con_10 <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_con_10 <- agg_accessibility_con_10 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-all-15, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_all_15 <- test %>% 
+  dplyr::filter(system == "All stations", threshold == "15 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_all_15 <- agg_accessibility_all_15 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "All stations", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-con-15, include=FALSE}
+# Add a variable to identify the income 
+agg_accessibility_con_15 <- test %>% 
+  dplyr::filter(system == "Conventional", threshold == "15 min") %>%
+  mutate(income_quintile = case_when(median_total_income <= iq[2] ~ "Bottom 20%",
+                                     median_total_income > iq[2] & median_total_income <= iq[3] ~ "Second 20%",
+                                     median_total_income > iq[3] & median_total_income <= iq[4] ~ "Third 20%",
+                                     median_total_income > iq[4] & median_total_income <= iq[5] ~ "Fourth 20%",
+                                     median_total_income > iq[5] ~ "Top 20%"),
+         income_quintile = factor(income_quintile, 
+                                  levels = c("Bottom 20%", "Second 20%", "Third 20%", "Fourth 20%", "Top 20%"),
+                                  ordered = TRUE))
+
+inc_stats_con_15 <- agg_accessibility_con_15 %>%
+  st_drop_geometry() %>%
+  #dplyr::filter(system == "Conventional", threshold == "10 min") %>%
+  group_by(income_quintile) %>%
+  summarize(population = sum(population50x50),
+            accessibility = sum(accessibility))
+```
+
+```{r income-all-hubs, include=FALSE}
+all_hubs <- rbind(inc_stats_all_3,
+                  inc_stats_all_5,
+                  inc_stats_all_10,
+                  inc_stats_all_15) %>%
+  drop_na() # there is a very small number of population cells without income data
+
+all_hubs
+```
+
+```{r income-conventional-hubs, include=FALSE}
+conventional_hubs <- rbind(inc_stats_con_3,
+                           inc_stats_con_5,
+                           inc_stats_con_10,
+                           inc_stats_con_15) %>%
+  drop_na() # there is a very small number of population cells without income data
+
+conventional_hubs
+```
+
+Table \ref{tab:accessibility-income} presents accessibility levels by income strata.
+As expected, the extreme threshold of fifteen minutes is associated with the largest number of people who are within the assumed service area of docking stations.
+We found that stations added to HBS expanded the spatial coverage of HBS, leading to more balance across the core service area.
+This is particularly evident at the minimum and average thresholds of three and five minutes, respectively, where the equity stations fill a number of gaps in program coverage.
+
+In line with Hosford and Winters' [-@hosfordWhoArePublic2018] findings, we found that accessibility to HBS was high for individuals in the bottom and second quintiles of median total household income who live in the centre of the core service area.
+This may well be an artifact of the spatial socioeconomic and demographic profile of Hamilton, where the most dense parts of the city (where a PBSP is most easily launched) are also those with relatively lower incomes [@hosfordWhoArePublic2018].
+On the other hand, we found the levels of accessibility to Hamilton's PBSP are generally lower for populations in the bottom 20% of median total household income, compared to populations in the top 20%.
+
+Our analysis revealed that the addition of equity stations increased accessibility to HBS by growing the population serviced irrespective of the walking threshold.
+The largest gains were made for dissemination areas in the second 20% median total household income, where an additional 3,073 and 5,395 people could reach a docking station within three and five minutes walk, respectively, after the addition of equity stations.
+However, we found that there were only small increases in the population in the bottom 20% of median total household income who are serviced by the equity stations, and the accessibility gains are also quite modest and smaller than for populations in the second and third quintiles of median total household income.
+This suggests that inequities in accessibility to HBS persist according to income, albeit this depends on the walking time thresholds.
+With and without equity stations, people in the top 20% of income have the highest level of access at a threshold of ten and fifteen minutes.
+Although dissemination areas in the second 20% have the highest level of access by a significant amount at lower distance thresholds, the bottom 20%, who may benefit more from increased access to HBS, have the lowest accessibility at three minutes threshold and the second lowest access at all other thresholds.
+
+```{r accessibility-income, echo=FALSE}
+accessibility_income <- cbind(
+  conventional_hubs,
+  all_hubs[, 2:3])
+
+colnames(accessibility_income) <- c("iq", "population_conv", "accessibility_conv", "population_all", "accessibility_all")
+
+accessibility_income <- accessibility_income %>%
+  transmute(iq,
+            population = rep(population_by_income$population, 4),
+            population_conv = round(population_conv),
+            accessibility_conv = round(accessibility_conv, 3), 
+            population_all = round(population_all),
+            accessibility_all = round(accessibility_all, 3),
+            population_diff = population_all - population_conv, 
+            accessibility_diff = accessibility_all - accessibility_conv)
+
+kable(accessibility_income,
+      "html",
+      booktabs = TRUE,
+      col.names = c("Income Quintile",
+                    "Total Population",
+                    "Population",
+                    "Accessibility",
+                    "Population",
+                    "Accessibility",
+                    "Population",
+                    "Accessibility"),
+      align = c("l", "c", "c", "c", "c", "c", "c"),
+      caption = "\\label{tab:accessibility-by-income}Accessibility and population serviced by income quintile and between systems (with and without equity stations).
+Total population is the population by income quintile in the DAs that have any PBSP service at all.") %>%
+  kable_styling(latex_options = c("striped", "scale_down")) %>%
+  pack_rows("Threshold - 3 minutes", 1, 5, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Threshold - 5 minutes", 6, 10, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Threshold - 10 minutes", 11, 15, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  pack_rows("Threshold - 15 minutes", 16, 20, 
+            label_row_css = "background-color: #8f9498; color: #fff;") %>%
+  add_header_above(c(" " = 2, "Without Equity Stations" = 2, "With Equity Stations" = 2, "Difference" = 2)) %>%
+  #column_spec(2, color = "white",
+  #background = spec_color(accessibility_income$accessibility[1:40], end = 1)) %>%
+  footnote(general = " ",
+           alphabet = c("With equity stations = Hamilton Bike Share current system (118 conventional stations, 12 equity stations)", "Without equity stations = Hamilton Bike Share original system (118 conventional stations, no equity stations)"))
+```
+
+6. Discussion
+===================
+
+Using disaggregated population data, we examined equity in accessibility to Hamilton Bike Share (HBS), with a focus on assessing the contribution of the program's equity stations.
+The BFCA approach, combined with pycnophylactic interpolation, enabled us to measure accessibility on a micro scale which better reflects the scale at which walking takes place and avoids the "absorption of disparities", as articulated by Chen et al.
+[-@chenExploringEquityPerformance2019] and elsewhere in the transportation literature [@nationalacademiesofsciencesEffectiveMethodsEnvironmental2004; @rowangouldIdentifyingEnvironmentalJustice2016].
+Our method also considers potential demand and supply, as well as congestion effects, in the calculation of accessibility, which is important for services like PBSPs but not accounted for in other common accessibility models.
+Unlike the 2SFCA method, there was no inflation or deflation in our accessibility estimates because population and level of service were preserved (see Figure \ref{fig:figure-10} and Figure \ref{fig:figure-11}).
+This differentiates our analysis from similar papers exploring equity in PBSPs that use larger geographical units of analysis or that focus only on station location instead of level of service.
+In this way, our paper has made an important contribution by applying an intuitive and useful approach not previously used in the cycling literature to measure accessibility to a PBSP.
+This open and reproducible research could be expanded for additional analysis using other individual-level data in Hamilton (e.g., age, gender, household size, education, etc.) or applied to PBSPs in other North American cities.
+
+The sensitivity analysis revealed that accessibility to docking stations is maximized at five minutes and decreases significantly by eight minutes [see Table \ref{tab:accessibility-income}].
+This reflects the normative guide advertised on some docking stations in Hamilton showing other stations within a five minute walk, as well as the directive of NACTO [@nactowalkingstation2015].
+We found that over 118,000 people can access a bike share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}].
+At a minimum threshold of three minutes, too few people can reach stations which leads to relatively high levels of service since there is little crowding for the stations.
+However, accessibility is at its lowest after eight minutes whereby congestion effects due to increased potential demand kick in.
+Chen et al.
+[-@chenExploringEquityPerformance2019] also found that accessibility decreases as larger geographical units are used in the estimation, since more people can access the service.
+
+The City of Hamilton has recognized from the launch of HBS that substantially more stations and bicycles are needed to service the area [@hamiltonsobi2015].
+With a service area of 40 sq.km, it is estimated that Hamilton should have between 380 and 440 stations instead of 130, and 1,500 bicycles instead of 900 [@hamiltonsobi2015].
+The City acknowledged that reduced capacity within the system led to gaps in coverage in some areas of the city "with some areas not having the recommended station density of 300m between stations or 10 stations per square km" [@hamiltonsobi2015].
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} demonstrate how gaps in accessibility were filled by the additional equity stations implemented by Hamilton Bike Share.
+The improvement in accessibility is most noticeable at the average and maximum thresholds in the east end of the core service area, which corresponds to the neighbourhoods where stations were added.
+However, the contribution of equity stations to reducing inequities in accessibility was modest.
+
+The equity stations were added to HBS as a targeted initiative to increase access for equity users in specific under-serviced neighbourhoods.
+However, we found that inequities persisted as evidenced by differences in accessibility according to income quintile (Table \ref{tab:accessibility-income}).
+While the addition of equity stations seems to modestly increase accessibility for all income groups at all thresholds, they did not increase accessibility substantially for any single income group.
+Most importantly, individuals in the bottom 20% of median total household income have the second lowest level of access to HBS at most thresholds (average, maximum, and extreme).
+At the minimum threshold, the bottom 20% have the lowest level of access.
+This suggests that additional equity stations are needed to more sufficiently address vertical equity.
+While previous research found that neighborhoods with more disadvantage are better serviced by HBS, the authors used the Pampalon Deprivation Index to determine the level of disadvantage for dissemination areas _across_ the city not just within the core service area [@hosfordWhoArePublic2018].
+Instead, we used median total household income for each dissemination area _within_ the core service area.
+We conclude that Hamilton's PBSP, while by default located in areas with more deprivation compared to other cities as noted by Hosford and Winters [-@hosfordWhoArePublic2018], continues to have disparities in accessibility between income groups.
+With equity stations, many areas with low median total household income in the east end of the service area have zero or low accessibility at the average threshold of five minutes.
+At the maximum and extreme thresholds, the top 20% have the highest level of access to HBS.
+These findings align broadly with other studies from Tampa [@chenExploringEquityPerformance2019], Philadelphia [@caspiBikesharingPhiladelphiaLowerincome2019] and Seattle [@mooneyFreedomStationSpatial2019], which have found disparities in station location, annual trips, or access to bicycles, respectively, between levels of income and education.
+With the addition of equity stations, there were large gains in accessibility for the second 20% at the average threshold, amounting to over 5,000 more people, but much smaller gains for the bottom 20% with only 326 more people who are able to access a HBS station.
+
+This paper highlights the importance of both the location and size of docking stations for increasing equity in accessibility to PBSPs.
+Congestion effects for services like bike share that offer a potential maximum supply of bicycles (i.e., number of racks) at each docking station must be taken into account when measuring accessibility.
+Several studies have examined who can access PBSPs based on the location of stations [among others, see @babagoliExploringHealthSpatial2019; @qianBikesharingEquityDisadvantaged2020; @qianEnhancingEquitableService2020; @wangNewBikeShare2019], but the size of docking stations in disadvantaged communities is not considered in the literature.
+An exception is a study on bike share ridership in New York City that found "the number of docks at a bike share station is statistically significant and positively associated with trip attractions for both men and women" [@wangGenderGapGenerators2019, p.5].
+Station size was also not identified by PBSP owners and operators as a factor that needed to be addressed in equity efforts [@howlandCurrentEffortsMake2017].
+PBSPs may not necessarily reduce inequities in access if stations are easy to reach for equity users but offer only a small number of bicycles.
+Likewise, more people may not opt to use the program if the supply of bicycles available at nearby stations is insufficient to meet demand.
+Babgoli et al.
+[-@babagoliExploringHealthSpatial2019] found a slight but not statistically significant increase in the proportion of neighborhoods with the highest levels of poverty that had stations after the Citi Bike expansion in 2015.
+Although the Citi Bike expansion was not specifically driven by a desire to reduce inequity in access, 16% of neighborhoods with the highest levels of poverty had stations compared to 12% before.
+This leaves open the question as to whether the stations held enough bicycles to meet demand, and whether congestion effects in neighbourhoods with more poverty are reducing the potential of bike share trips.
+
+The following policy and planning implications can inform efforts for increasing equity in HBS, and potentially other PBSPs in North America.
+Our analysis has identified specific areas that have both low accessibility and low median total household income.
+Figures \ref{fig:figure-bi-map-threshold-3}, \ref{fig:figure-bi-map-threshold-5}, \ref{fig:figure-bi-map-threshold-10}, and \ref{fig:figure-bi-map-threshold-15} highlight potential locations for new equity stations to increase access for low-income groups.
+Additional stations are needed to further reduce inequities in accessibility.
+This information will be useful to transportation planners in Hamilton and HBS operators.
+Our findings also suggest that PBSPs should consider increasing the size of stations in targeted communities.
+Notably, figure \ref{fig:sobi-stations-in-hamilton} highlights how the equity stations in Hamilton have a lower number of racks compared to docking stations in the city's downtown core.
+Docking stations added to increase vertical equity should have an adequate supply of public bicycles, particularly more in areas that have a strong interest in adopting bike share, to increase levels of service and potentially attract new users.
+Finally, distance to stations is associated with bike share use [@fullerUseNewPublic2011; @wangNewBikeShare2019], especially when the trip starts at the residence [@fishmanBikeShareSynthesis2013].
+Distance can also be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014].
+To address this issue, transportation planners and PBSP operators should consult with groups and communities who are under-represented among bike share users to determine how far they are willing to walk to reach a docking station and to get their input on the location of new docking stations.
+
+7. Study Limitations
+===================
+
+This paper did not examine or compare ridership data between conventional and equity stations.
+Therefore, further research is needed to determine whether the addition of equity stations encouraged more bicycling for equity users living near them.
+Other studies have specifically looked at differences in trip type, frequency, or length among users from disadvantaged neighborhoods [@caspiBikesharingPhiladelphiaLowerincome2019; @qianBikesharingEquityDisadvantaged2020; @wangNeighborhoodSociodemographicCharacteristics2019], but our analysis is limited by the lack of publicly available route and individual user data to conduct similar analyses for HBS.
+
+An additional limitation is the lack of publicly available information about the average number of bicycles at each station on a daily basis.
+HBS works to balance the number of bicycles across the core service area, but it is reasonable to expect that the number of bicycles will not match exactly the number of racks at every station.
+Ideally, instead of number of bicycle racks as our measure of supply, we would have liked to use the average number of bicycles at stations, perhaps at different times during the day across many seasons.
+Should this data become available, it would be worthwhile to revisit the study to examine how well the operation of the system (including balancing of bicycles across stations) works to maintain the nominal levels of accessibility examined in this paper.
+
+\textcolor{blue}{Finally, it is important to note that this study examined accessibility to docking stations, that is, the entry point to the bikeshare system.
+There is a literature that examines transportation modes from this perspective, including accessibility to transit [e.g.,zhaoforecasting2003; @olszewskiusing2005; zieltracomparative2011] and more recently accessibility to bikeshare systems [e.g., @hosfordWhoArePublic2018; @mooneyFreedomStationSpatial2019; @caggianiequity2020; @duranfair2020].
+These focus of these studies is to assess who has access to a transportation service.
+However, as argued by @moniruzzamanaccessibility2012, usage of the system likely depends as well on the accessibility to destinations that the system generates.
+An extension of the present study would be to investigate accessibility by the bikeshare system, and perhaps compare that to usage patterns.}
+
+8. Conclusion
+===================
+
+The addition of specific stations to improve equity within the PBSP in Hamilton had the net effect of increasing accessibility and reducing vertical inequities to some extent.
+In particular, accessibility improved the most for those in the second 20% median total household income at all thresholds, but the gains were only modest for all income groups.
+Dissemination areas in the bottom 20% had the lowest accessibility at three minutes, and second lowest levels of accessibility at five, ten, and fifteen minutes.
+Congestion effects were observed at higher thresholds, with accessibility decreasing significantly once the catchment area is increased to ten minutes walking.
+
+Wang and Lindsey [-@wangNewBikeShare2019] have noted that there is a lack of research that examines how bike share users' behaviour changes as a result of program changes to station locations or improvements in accessibility.
+As such, a logical next step to this research is to examine whether the equity stations increased ridership or resulted in new memberships for equity users.
+An examination of the types of trips undertaken by residents in these areas would also be informative, such as the study undertaken by Caspi and Norland [-@caspiBikesharingPhiladelphiaLowerincome2019] after docking stations were implemented in low-income Philadelphia neighborhoods.
+The bulk of bicycle infrastructure that has been built in Hamilton to date is located in the core service area particularly near the conventional stations.
+It would be worthwhile to explore the route choice of bike share trips departing or ending at the equity stations and to identify factors that specifically influence trips from these stations, which would extend existing studies conducted by Scott and colleagues [@luUnderstandingBikeShare2018; @scottWhatFactorsInfluence2019; @scottRouteChoiceBike2021].
+Perceived accessibility using bike share according to low-income neighbourhoods could also be studied in the future.
+Lättman et al.
+[-@lattmanDevelopmentTestPerceived2016] have developed the _Perceived Accessibility Scale_ that would be useful for this purpose.
+This paper, combined with additional studies such as those conceptualized above, would serve as a valuable case study for Hamilton and other cities with PBSPs that wish to evaluate and address inequities in accessibility to active transportation options in urban areas.
+
+Acknowledgments
+===================
+
+We wish to thank the anonymous reviewers who provided constructive feedback to improve the quality of our paper.
+This research was completed using open software, and we also wish to acknowledge the developers of the following `R` packages: `biscale` [@R-biscale], `cowplot` [@R-cowplot], `data.table` [@R-data.table], `disk.frame` [@R-disk.frame], `gdistance` [@R-gdistance], `gridExtra` [@R-gridExtra], `kableExtra` [@R-kableExtra], `knitr` [@R-knitr], `pycno` [@R-pycno], `r5r` [@R-r5r], `raster` [@R-raster], `rgdal` [@R-rgdal], `rticles` [@R-rticles], `sf` [@R-sf], `tidyverse` [@R-tidyverse], `tinytex` [@R-tinytex],
+`units` [@R-units].
+
+References {#references .unnumbered}
+===================

### 2021-10-25 README.Rmd
Updated README.rmd to display equations

@@
 
-\begin{equation}
-\label{eq:population-allocation}
+$$
 P_j = {\sum_{i = 1}^{n} P_i{w_{ij}}}
-\end{equation}
+$$
 
@@
 The level of service of station $j$ in bicycle racks per person is the supply at each station (i.e., the maximum number of bicycle racks) divided by the population within the established catchment area:
-\begin{equation}
-\label{eq:level-of-service}
+$$
 L_j = \frac {S_j}{P_j} = \frac {S_j}{{\sum_{i = 1}^{n} P_i{w_{ij}}}}
-\end{equation}
+$$
 
 In the second step, the accessibility of population unit $i$ is calculated as the weighted sum of the level of service of all stations that can be reached from there according to the spatial weights:
-\begin{equation}
-\label{eq:FCA-accessibility}
+$$
 A_i = {\sum_{j = 1}^{J} L_j{w_{ij}}} = {\sum_{j = 1}^{J} \frac {S_j{w_{ij}}}{\sum_{i = 1}^{n} P_i{w_{ij}}}}
-\end{equation}
+$$
 
 The balanced approach of Páez et al. [-@paezDemandLevelService2019] replaces the spatial weights with normalized versions as follows. In the first step, the population is weighted with:
-\begin{equation}
-\label{eq:spatial-weights}
+$$
 {w_{ij}^{i} = \frac {w_{ij}}{\sum_{j = 1}^{J} {w_{ij}}}}
-\end{equation}
+$$
 
 \noindent and in the second step, the level of service is weighted using:
-\begin{equation}
-\label{eq:spatial-weights-2}
+$$
 {w_{ij}^{j} = \frac {w_{ij}}{\sum_{i = 1}^{n} {w_{ij}}}}
-\end{equation}
-
+$$
 These weights satisfy the following properties:
-\begin{equation}
-\label{eq:weights}
+$$
 \sum_{j = 1}^{J} {w^i_{ij}} = 1
-\end{equation}
+$$
 
 \noindent and:
-\begin{equation}
-\label{eq:weights-2}
+$$
 \sum_{i = 1}^{n} {w^j_{ij}} = 1
-\end{equation}
+$$
 
 With these weights, accessibility can be calculated without risk of demand or supply inflation:
-\begin{equation}
-\label{eq:balanced-accessibility}
+$$
 A_i = {\sum_{j = 1}^{J} \frac {S_j{w^j_{ij}}}{\sum_{i = 1}^{n} P_i{w^i_{ij}}}}
-\end{equation}
+$$
 
 By allocating the population and level of service proportionally, this method preserves the values of the population and level of service since:
-\begin{equation}
-\label{eq:proportionality}
+$$
 \sum_{i=1}^n A_i = \sum_{j=1}^J L_j
-\end{equation}
+$$
 

### 2022-12-05 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
Initialized renv for greater reproducibility, updated README with bibliographical info and created a map for sharing to Mastodon

@@
 
-```{r install-data-package, include = FALSE}
+```{r load-packages-1, include=FALSE}
+library(glue)
+library(here)
+```
+
+```{r install-data-package, include = FALSE, renv.ignore=TRUE}
 # Run only once if needed to install data package `sobiEquity`
 if (!require("sobiEquity", character.only = TRUE)) {
-      devtools::install_github("https://github.com/paezha/Accessibility-Sobi-Hamilton", subdir = "sobiEquity")
+      install.packages(glue(here::here(),'/sobiEquity_0.1.0.tar.gz'), repos=NULL, type='source')
   }
@@
 
-```{r load-packages, include=FALSE}
+```{r load-packages-2, include=FALSE}
 library(biscale)
@@
 library(data.table)
-library(disk.frame)
+#library(disk.frame)
 library(dplyr)
@@
 library(ggmap)
+library(ggplot2)
 library(ggspatial)
+library(glue)
 library(gridExtra)
+library(here)
 library(kableExtra)
 library(pycno)
-library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.htmllibrary(readr)
+#library(r5r) # the r5r package requires Java Development Kit version 11, which can be downloaded from https://www.oracle.com/java/technologies/javase-jdk11-downloads.htmllibrary(readr)
 library(raster)
-library(rgdal)
+#library(rgdal)
 library(sf)
 library(sobiEquity) # Data package
-library(tidyverse)
-library(tinytex)
+library(tidyr)
+#library(tinytex)
 library(units)
@@
 ```
+
+<!--
+
+MAPSTODON MONDAY DEC. 5, 2022
+
+```{r figure-8-a, include=FALSE}
+#accessibility plot for current system with equity stations at 10 minutes walk
+accessibility_10min <- accessibility %>% 
+  dplyr::filter(threshold == "10 min",
+                system == "All stations")  %>%
+  mutate(system = "With equity stations",
+         accessibility = accessibility * 1000, # Convert to racks per 1,000 person
+         accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.04, 0.3, 2.00, Inf), 
+                             labels = c("<0.04", "0.04-0.3", "0.3-2.00", ">2.00")),
+         accessibility = factor(accessibility,
+                                levels = c( ">2.00", "0.3-2.00", "0.04-0.3", "<0.04"),
+                                ordered = TRUE))
+
+acc_10_all <- ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_10min), 
+                         st_coordinates(accessibility_10min)),
+            aes(x = X, 
+                y = Y, 
+                fill = accessibility)) +
+  scale_fill_viridis_d(name = "Accessibility\n(racks/1,000 person)",
+                    option = "viridis",
+                    direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ .)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_north_arrow(location = "tr", which_north = "true", 
+        pad_x = unit(0.0, "in"), pad_y = unit(0.0, "in"),
+        style = north_arrow_fancy_orienteering)
+
+# Extract the legend
+acc_10_legend <- cowplot::get_legend(acc_10_all)
+```
+
+```{r figure-8-b, include=FALSE}
+#accessibility plot for current system without equity stations at 10 minutes walk
+accessibility_10min <- accessibility %>% 
+  dplyr::filter(threshold == "10 min",
+                system == "Conventional")  %>%
+  mutate(system = "Without equity stations",
+         accessibility = accessibility * 1000, # Convert to racks per 1,000 person
+         accessibility = cut(accessibility,
+                             breaks = c(-Inf, 0.04, 0.3, 2.00, Inf), 
+                             labels = c("<0.04", "0.04-0.3", "0.3-2.00", ">2.00")),
+         accessibility = factor(accessibility,
+                                levels = c( ">2.00", "0.3-2.00", "0.04-0.3", "<0.04"),
+                                ordered = TRUE))
+
+acc_10_conv <- ggplot() +
+  geom_sf(data = population_50x50,
+          color = "lightgray") +
+  geom_tile(data = cbind(st_drop_geometry(accessibility_10min), 
+                         st_coordinates(accessibility_10min)),
+            aes(x = X, 
+                y = Y, 
+                fill = accessibility)) +
+  scale_fill_viridis_d(name = "Accessibility\n(racks/1,000 person)",
+                    option = "viridis",
+                    direction = 1) +
+  geom_sf(data = sobi_service,
+          fill = NA) +
+  facet_grid(system ~ .)  +
+  geom_sf(data = community_downtown %>%
+            dplyr::filter(NAME == "Hamilton"),
+          fill = NA,
+          color = "dark green") +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text = element_blank(),
+        axis.ticks = element_blank())  +
+  annotation_scale(location = "bl", width_hint = 0.4,
+                   pad_x = unit(0.4, "in"), pad_y = unit(0.2, "in"))
+```
+
+```{r figure-8, echo=FALSE, fig.align = 'center', out.width = "90%", fig.cap = "Accessibility at 10 minutes walk (maximum threshold) compared between current system with equity stations and the original system without equity stations."}
+ggdraw() + 
+  draw_plot(acc_10_conv + 
+              theme(legend.position = "none"), 
+            0.00, 0.5, 1, 0.5) + 
+  draw_plot(acc_10_all + 
+              theme(legend.position = "none"), 0.00, 0.0, 1, 0.5) + 
+  draw_plot(acc_10_legend, 0.01, 0.4, 0.2, 0.2)
+
+ggsave("acc_10_mapstodon.jpeg",
+       width = 8,
+       height = 7)
+```
+
+-->
 

### 2022-12-05 README.Rmd
Initialized renv for greater reproducibility, updated README with bibliographical info and created a map for sharing to Mastodon

@@
 
-Forthcoming in *Transportation Research Part D: Transport and Environment*
+Desjardins, E., Higgins, C.D., Paez, A.
+(2022) Examining spatial equity of bike share: A balanced floating catchment area approach, Transportation Research Part D (102) 103091 (doi.org/10.1016/j.trd.2021.103091)
 
@@
 
-```{r echo=FALSE}
-rm(list = ls())
-```
-
-```{r install-data-package, include = FALSE}
+```{r load-packages-1, include=FALSE}
+library(glue)
+library(here)
+```
+
+```{r install-data-package, include = FALSE, renv.ignore=TRUE}
 # Run only once if needed to install data package `sobiEquity`
 if (!require("sobiEquity", character.only = TRUE)) {
-      devtools::install_github("https://github.com/paezha/Accessibility-Sobi-Hamilton", subdir = "sobiEquity")
+      install.packages(glue(here::here(),'/sobiEquity_0.1.0.tar.gz'), repos=NULL, type='source')
   }
